---
title: "Progression-free survival — five covariate imputations"
format: html
execute:
  freeze: false
---

Each adjusted analysis uses five MICE datasets, with 50 iterations and five
donors. PFS event status and its Nelson–Aalen cumulative hazard inform both
imputation models. Identifiers, randomized treatment, event status, and
follow-up times are not imputed. Age, gender, bilirubin, and LDH are auxiliary
covariates. Binary ECOG and categorical mGPS are derived from completed sources.

The analysis with a lung indicator uses proportional-odds imputation
for ordered ECOG and predictive mean matching for laboratories. The diagnosis
analysis replaces lung with a diagnosis random intercept in every incomplete
covariate's two-level predictive mean matching model. ECOG uses a numeric
working model with observed integer categories as donors in this second run.

```{r}
#| label: impute-pfs
#| include: false
library(tidyverse)
library(mice)
library(miceadds)

source(here::here("R", "analysis-config.R"))
source(here::here("R", "survival-helpers.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
artifact_dir <- here::here("artifacts", "secondary", "pfs")
figure_dir <- file.path(artifact_dir, "figures")

dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

prepared <- arrow::read_parquet(file.path(artifact_dir, "prepared.parquet")) |>
  mutate(diagnosis = factor(diagnosis), gender = factor(gender))

prepared$na_pfs <- mice::nelsonaalen(prepared, timevar = pfs_time_days, statusvar = event_pfs)

imputations <- list()

for (adjustment in c("os_adjusted", "diagnosis")) {
  imputation_data <- if (adjustment == "os_adjusted") {
    prepared |> select(-diagnosis) |> mutate(ecog_fstcnt = ordered(ecog_fstcnt))
  } else {
    prepared |> mutate(diagnosis_id = as.integer(diagnosis)) |> select(-diagnosis, -lung)
  }

  incomplete <- names(which(colSums(is.na(imputation_data)) > 0))

  methods <- mice::make.method(imputation_data)
  methods[] <- ""
  methods[incomplete] <- if (adjustment == "os_adjusted") "pmm" else "2l.pmm"

  if (adjustment == "os_adjusted") methods["ecog_fstcnt"] <- "polr"

  predictors <- mice::make.predictorMatrix(imputation_data)
  predictors[, c("id", "pfs_time_days")] <- 0
  predictors[!nzchar(methods), ] <- 0

  if (adjustment == "diagnosis") predictors[incomplete, "diagnosis_id"] <- -2

  input_hash <- digest::digest(list(imputation_data, methods, predictors, config$imputation$supporting))
  imputation_path <- file.path(artifact_dir, paste0("imputations-", adjustment, ".rds"))
  imputation <- if (!force && file.exists(imputation_path)) readRDS(imputation_path) else NULL

  if (is.null(imputation) || !identical(attr(imputation, "input_hash"), input_hash)) {
    imputation <- mice::mice(
      imputation_data, method = methods, predictorMatrix = predictors,
      m = 5L, maxit = 50L, donors = 5L, printFlag = FALSE,
      seed = analysis_seed(config, "supporting", offset = if (adjustment == "os_adjusted") 2200L else 2201L)
    )
    attr(imputation, "input_hash") <- input_hash
    saveRDS(imputation, imputation_path, compress = "xz")
  }

  completed_datasets <- lapply(1:5, function(index) {
    completed <- mice::complete(imputation, index) |>
      mutate(diagnosis = prepared$diagnosis, ecog_binary = derive_ecog_binary(ecog_fstcnt),
             mgps = factor(derive_mgps(albumin, c_reactive_protein), levels = 0:2))

    assert_ecog_binary_consistent(completed)
    assert_mgps_consistent(completed)

    stopifnot(!anyNA(completed), imputation$m == 5L,
              all(as.numeric(as.character(completed$ecog_fstcnt)) %in% 0:4),
              identical(completed$pfs_time_days, prepared$pfs_time_days),
              identical(completed$event_pfs, prepared$event_pfs))
    completed
  })

  if (adjustment == "diagnosis") {
    stopifnot(all(imputation$predictorMatrix[incomplete, "diagnosis_id"] == -2))
  }

  saveRDS(completed_datasets, file.path(artifact_dir, paste0("completed-", adjustment, ".rds")))
  imputations[[adjustment]] <- imputation

  png(file.path(figure_dir, paste0("imputation-traces-", adjustment, ".png")), width = 1500, height = 1200, res = 150)
  print(plot(imputation, layout = c(2, 5)))
  dev.off()

  png(file.path(figure_dir, paste0("imputation-distributions-", adjustment, ".png")), width = 1500, height = 900, res = 150)
  print(densityplot(imputation, ~albumin + bilirubin + c_reactive_protein + lactate_dehydrogenase))
  dev.off()
}
```

```{r}
#| label: fig-pfs-imputation-checks
#| fig-cap: "Imputation traces and laboratory distributions: Adjusted (fixed) followed by Adjusted (hierarchical)"
#| fig-alt: "Means and standard deviations across MICE iterations, and observed-versus-imputed laboratory densities, for each of the two five-imputation analyses."
knitr::include_graphics(file.path(figure_dir, c(
  "imputation-traces-os_adjusted.png", "imputation-distributions-os_adjusted.png",
  "imputation-traces-diagnosis.png", "imputation-distributions-diagnosis.png"
)))
```

```{r}
#| label: tbl-pfs-imputation-events
imap_dfr(imputations, function(imputation, adjustment) {
  if (is.null(imputation$loggedEvents)) return(tibble())
  imputation$loggedEvents |> count(dep, meth, out) |> mutate(adjustment = adjustment)
}) |> knitr::kable()
```
