# Secondary outcomes

Secondary outcomes are exploratory.  The report renders compact artifacts
when available and keeps each missing/blocked status visible.

```{r}
#| label: secondary-overview
#| results: asis

report_render_section("secondary", root = report_root)
```

## Progression-free survival

PFS uses randomized ITT from randomization, with earliest recorded progression
or death as the event.  The Bayesian proportional-hazards model's primary
contrast is the hazard ratio.

```{r}
#| label: pfs-results
#| results: asis

report_render_section("pfs", root = report_root)
```

## Survival rates

Six- and 12-month rates are derived from OS posterior artifacts.  Later
horizons are shown only when the follow-up sufficiency gate is available.

```{r}
#| label: survival-rate-results
#| results: asis

report_render_section("survival_rates", root = report_root)
```

## Best supportive care

Time to BSC starts at first-line treatment initiation.  A `bsc_date` is
retained for every participant, and death before BSC is censored at death in
the cause-specific analysis.

```{r}
#| label: bsc-results
#| results: asis

report_render_section("bsc", root = report_root)
```

## Blood products

Blood-product rates use observed positive exposure and the adjusted Bayesian
negative-binomial model.  Missing event quantities or unknown product types
are not recoded as zero.

```{r}
#| label: blood-product-results
#| results: asis

report_render_section("blood_products", root = report_root)
```

## Tissue biopsy

The biopsy outcome counts unique biopsy days and uses the adjusted count model
with treatment, linear age, ECOG 0/1/2/3+, and observed survival-time offset.

```{r}
#| label: tissue-biopsy-results
#| results: asis

report_render_section("tissue_biopsy", root = report_root)
```

## Imaging

The total imaging count uses the same adjusted model.  CT, MRI, and PET-CT
summaries are descriptive.  Admissions and ER visits have no standalone model
because they are already represented by TAOOH states.

```{r}
#| label: imaging-results
#| results: asis

report_render_section("imaging", root = report_root)
```
