---
title: "QoL death-inclusive Markov — imputation"
subtitle: "Independent long-format imputation using 8 completed datasets"
format: html
execute:
  freeze: false
---

This test-cycle workflow creates eight independently completed longitudinal
datasets. Baseline variables are first imputed in one patient-level MICE
object with eight completed datasets. Each completed baseline dataset is then
joined to the weekly data before living QoL states 1--7 are imputed. The
previous-state predictor is based on the completed baseline state through the
first observed follow-up QoL value; thereafter, the most recent observed
follow-up value is carried forward. One observed transition to death is added
deterministically after imputation; death is never a MICE target.

Every cached baseline fit is checked for the required number of imputations and
iterations before reuse. Every completed longitudinal dataset is checked for
complete and unique patient-week keys, full patient coverage, valid outcome
states, and exact agreement with the observed death template.

```{r}
#| label: setup-death-inclusive-longitudinal-qol-imputation
#| include: false

source(here::here("R", "analysis-config.R"))
library(tidyverse)

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
n_imputations <- 8L
indices <- seq_len(n_imputations)

input_path <- here::here(
  "artifacts", "primary", "quality-of-life",
  "death-inclusive-longitudinal-prepared.parquet"
)
output_dir <- here::here(
  "artifacts", "supporting", "qol-death-inclusive-longitudinal"
)
imputation_dir <- file.path(output_dir, "imputations")
imputation_fit_dir <- file.path(output_dir, "imputation-fits")
prepared_path <- file.path(output_dir, "prepared.parquet")
death_template_path <- file.path(output_dir, "death-template.parquet")
baseline_imputation_path <- file.path(output_dir, "baseline-imputations.rds")
run_summary_path <- file.path(output_dir, "imputation-run-summary.parquet")

dir.create(imputation_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(imputation_fit_dir, recursive = TRUE, showWarnings = FALSE)

key_string <- function(data) {
  paste(data$id, data$week, sep = "\r")
}
```

## Prepare the weekly imputation data

```{r}
#| label: prepare-death-inclusive-longitudinal-qol-imputation
#| include: false

prepared_full <- arrow::read_parquet(input_path) |>
  as.data.frame() |>
  arrange(id, week)

required_columns <- c(
  "id", "tx", "week", "q30", "baseline_q30", "event_death",
  "survival_time_days", "death_week", "ecog_fstcnt", "diagnosis_code"
)
if (!all(required_columns %in% names(prepared_full))) {
  stop("The prepared longitudinal QoL input is missing required columns.")
}
if (anyNA(prepared_full$id) || anyDuplicated(prepared_full[c("id", "week")])) {
  stop("The prepared longitudinal QoL input requires complete, unique patient-week keys.")
}

patient_survival <- prepared_full |>
  distinct(id, survival_time_days, event_death)

patient_survival$na_est <- mice::nelsonaalen(
  patient_survival,
  timevar = survival_time_days,
  statusvar = event_death
)

mean_followup_qol <- prepared_full |>
  filter(!is.na(q30), q30 %in% 1:7) |>
  summarise(mean_observed_q30 = mean(as.numeric(q30)), .by = id)

week_basis <- Hmisc::rcspline.eval(1:26, nk = 4, inclx = TRUE) |>
  as.data.frame()
colnames(week_basis) <- c("week", "week_rcs_1", "week_rcs_2")

prepared_full <- prepared_full |>
  left_join(patient_survival |> select(id, na_est), by = "id") |>
  left_join(week_basis, by = "week") |>
  mutate(
    gender = factor(gender),
    age_scaled = (age - mean(age, na.rm = TRUE)) / sd(age, na.rm = TRUE),
    ecog_fstcnt = factor(
      if_else(ecog_fstcnt == 4, 3, ecog_fstcnt),
      ordered = TRUE
    ),
    plan_fstcnt_coded = fct_lump_n(factor(plan_fstcnt_coded), n = 4),
    diagnosis_code = factor(diagnosis_code),
    stage_binary = factor(stage_binary),
    baseline_q30 = factor(baseline_q30, levels = 1:7, ordered = TRUE)
  )

# Only confirmed living weeks enter MICE. An observed response in a partial
# censoring week is retained. Exactly one observed transition to death is kept
# separately and reattached after living QoL has been completed.
prepared <- prepared_full |>
  filter(
    (event_death == 1L & week < death_week) |
      (event_death == 0L &
        (week * 7L <= survival_time_days | !is.na(q30)))
  ) |>
  select(-y_death_inclusive)

death_template <- prepared_full |>
  filter(event_death == 1L, week == death_week) |>
  select(-y_death_inclusive)

expected_death_ids <- prepared_full |>
  filter(event_death == 1L) |>
  distinct(id)
if (
  anyNA(death_template$id) ||
    anyDuplicated(death_template[c("id", "week")]) ||
    nrow(death_template) != nrow(expected_death_ids) ||
    !setequal(death_template$id, expected_death_ids$id)
) {
  stop("The death template must contain exactly one observed death row per decedent.")
}

arrow::write_parquet(prepared, prepared_path, compression = "zstd")
arrow::write_parquet(death_template, death_template_path, compression = "zstd")
```

## Impute patient-level variables

```{r}
#| label: prepare-death-inclusive-qol-baseline-imputation
#| include: false

baseline_variables <- c(
  "age_scaled", "gender", "ecog_fstcnt", "diagnosis_code", "stage_binary",
  "plan_fstcnt_coded", "albumin", "c_reactive_protein", "baseline_q30"
)
baseline_targets <- c(
  "ecog_fstcnt", "diagnosis_code", "plan_fstcnt_coded", "albumin",
  "c_reactive_protein", "baseline_q30", "mean_observed_q30"
)

baseline_data <- prepared_full |>
  distinct(id, .keep_all = TRUE) |>
  select(id, tx, all_of(baseline_variables), event_death, na_est) |>
  left_join(mean_followup_qol, by = "id")

cached_baseline_imputations <- if (file.exists(baseline_imputation_path)) {
  tryCatch(readRDS(baseline_imputation_path), error = function(error) NULL)
} else {
  NULL
}
baseline_fit_is_current <-
  inherits(cached_baseline_imputations, "mids") &&
  identical(as.integer(cached_baseline_imputations$m), n_imputations) &&
  identical(
    as.integer(cached_baseline_imputations$iteration),
    as.integer(config$imputation$supporting$maxit)
  ) &&
  identical(nrow(cached_baseline_imputations$data), nrow(baseline_data)) &&
  identical(as.character(cached_baseline_imputations$data$id), baseline_data$id)

run_baseline_imputation <- force || !baseline_fit_is_current
```

```{r}
#| label: run-death-inclusive-qol-baseline-imputation
#| include: false

if (run_baseline_imputation) {
  methods <- mice::make.method(baseline_data)
  methods[setdiff(names(methods), baseline_targets)] <- ""

  predictors <- mice::make.predictorMatrix(baseline_data)
  predictors[, "id"] <- 0L

  baseline_imputations <- mice::mice(
    baseline_data,
    m = n_imputations,
    maxit = config$imputation$supporting$maxit,
    method = methods,
    predictorMatrix = predictors,
    donors = config$imputation$supporting$donors,
    seed = analysis_seed(config, "supporting", offset = 500L),
    printFlag = TRUE
  )

  saveRDS(baseline_imputations, baseline_imputation_path, compress = "xz")
} else {
  baseline_imputations <- cached_baseline_imputations
}

if (!identical(as.integer(baseline_imputations$m), n_imputations)) {
  stop("The baseline MICE object does not contain the eight required datasets.")
}

completed_baseline_variables <- setdiff(
  baseline_targets,
  "mean_observed_q30"
)
```

## Configure longitudinal QoL imputation

```{r}
#| label: configure-death-inclusive-qol-longitudinal-imputation
#| include: false

# This is a fixed auxiliary history derived only from observed follow-up QoL.
# Each observed value becomes available from the following week and is carried
# forward. Completed baseline QoL supplies the history through the first
# observed follow-up value after the history is joined within each imputation.
history_proxy <- prepared_full |>
  select(id, week, q30) |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(yprev_imputation = lag(q30)) |>
  fill(yprev_imputation, .direction = "down") |>
  ungroup() |>
  select(id, week, yprev_imputation)

# These predictors intentionally include the Nelson--Aalen estimate and death
# indicator because both are informative about missing follow-up QoL.
q30_predictors <- c(
  "tx", "age_scaled", "gender", "ecog_fstcnt", "diagnosis_code",
  "stage_binary", "plan_fstcnt_coded", "albumin", "c_reactive_protein",
  "baseline_q30", "event_death", "na_est", "week", "week_rcs_1",
  "week_rcs_2", "yprev_imputation"
)

analysis_baseline_variables <- c(
  "tx", "age", "gender", "diagnosis_code", "stage_binary", "ecog_fstcnt",
  "plan_fstcnt_coded", "albumin", "c_reactive_protein"
)
```

## Impute weekly QoL and add death

```{r}
#| label: run-death-inclusive-qol-longitudinal-imputation
#| include: false

run_summaries <- vector("list", length(indices))

for (position in seq_along(indices)) {
  index <- indices[[position]]
  completed_path <- file.path(
    imputation_dir,
    sprintf("imputation-%02d.parquet", index)
  )
  long_imputation_path <- file.path(
    imputation_fit_dir,
    sprintf("long-imputation-%02d.rds", index)
  )

  if (
    !force &&
      !run_baseline_imputation &&
      file.exists(completed_path) &&
      file.exists(long_imputation_path)
  ) {
    existing_completed <- arrow::read_parquet(completed_path) |>
      as.data.frame()
    existing_fit <- tryCatch(
      readRDS(long_imputation_path),
      error = function(error) NULL
    )
    existing_is_current <-
      inherits(existing_fit, "mids") &&
      identical(as.integer(existing_fit$m), 1L) &&
      identical(
        as.integer(existing_fit$iteration),
        as.integer(config$imputation$supporting$maxit)
      ) &&
      !anyNA(existing_completed$id) &&
      !anyNA(existing_completed$y) &&
      !anyDuplicated(existing_completed[c("id", "week")]) &&
      setequal(existing_completed$id, prepared_full$id) &&
      setequal(
        existing_completed$id[existing_completed$y == 8L],
        death_template$id
      )
    if (existing_is_current) {
      run_summaries[[position]] <- tibble(
        imputation = index,
        rows = nrow(existing_completed),
        participants = n_distinct(existing_completed$id),
        imputed_q30 = nrow(existing_fit$imp$q30),
        deaths = sum(existing_completed$y == 8L),
        logged_events = if (is.null(existing_fit$loggedEvents)) 0L else nrow(existing_fit$loggedEvents),
        reused = TRUE
      )
      next
    }
  }

  completed_baseline <- mice::complete(baseline_imputations, index) |>
    as.data.frame() |>
    select(id, all_of(completed_baseline_variables))
  if (
    anyNA(completed_baseline$id) ||
      anyDuplicated(completed_baseline$id) ||
      !setequal(completed_baseline$id, baseline_data$id)
  ) {
    stop("Baseline completion ", index, " does not preserve all patient IDs.")
  }

  imputation_data <- prepared |>
    select(-all_of(completed_baseline_variables)) |>
    left_join(completed_baseline, by = "id", relationship = "many-to-one") |>
    left_join(history_proxy, by = c("id", "week"), relationship = "one-to-one") |>
    mutate(
      yprev_imputation = coalesce(
        yprev_imputation,
        as.numeric(as.character(baseline_q30))
      ),
      yprev_imputation = factor(yprev_imputation, levels = 1:7),
      q30 = factor(q30, levels = 1:7, ordered = TRUE),
      tx = as.numeric(as.character(tx))
    )

  if (anyNA(imputation_data$id) || anyDuplicated(imputation_data[c("id", "week")])) {
    stop("Longitudinal imputation ", index, " has invalid patient-week keys before MICE.")
  }
  missing_predictors <- q30_predictors[
    vapply(imputation_data[q30_predictors], anyNA, logical(1))
  ]
  if (length(missing_predictors) > 0L) {
    stop(
      "Longitudinal imputation ", index,
      " has missing values in completed predictors: ",
      paste(missing_predictors, collapse = ", ")
    )
  }

  methods <- mice::make.method(imputation_data)
  methods[] <- ""
  methods["q30"] <- "polr"

  predictors <- mice::make.predictorMatrix(imputation_data)
  predictors[,] <- 0L
  predictors["q30", q30_predictors] <- 1L

  long_imputation <- tryCatch(
    mice::mice(
      imputation_data,
      m = 1L,
      maxit = config$imputation$supporting$maxit,
      method = methods,
      predictorMatrix = predictors,
      seed = analysis_seed(config, "supporting", index, offset = 550L),
      polr.to.loggedEvents = TRUE,
      printFlag = TRUE
    ),
    error = function(error) {
      stop(
        "Longitudinal MICE fit failed for imputation ", index, ": ",
        conditionMessage(error),
        call. = FALSE
      )
    }
  )

  living_completed <- mice::complete(long_imputation, 1L) |>
    as.data.frame() |>
    mutate(
      q30 = as.integer(as.character(q30)),
      y = q30
    )
  if (
    anyNA(living_completed$id) ||
      anyNA(living_completed$y) ||
      anyDuplicated(living_completed[c("id", "week")]) ||
      !setequal(key_string(living_completed), key_string(prepared))
  ) {
    stop("Longitudinal MICE completion ", index, " did not preserve living patient-week rows.")
  }

  death_rows <- death_template |>
    select(-all_of(completed_baseline_variables)) |>
    left_join(completed_baseline, by = "id", relationship = "many-to-one") |>
    mutate(q30 = NA_integer_, y = 8L)
  if (
    anyNA(death_rows$id) ||
      nrow(death_rows) != nrow(death_template) ||
      !setequal(key_string(death_rows), key_string(death_template))
  ) {
    stop("Imputation ", index, " did not preserve the observed death template.")
  }

  completed <- bind_rows(living_completed, death_rows) |>
    arrange(id, week) |>
    group_by(id) |>
    mutate(
      yprev = coalesce(
        lag(y),
        as.integer(as.character(first(baseline_q30)))
      ),
      yprev = factor(yprev, levels = 1:7)
    ) |>
    ungroup() |>
    select(
      id, y, q30, yprev, week, baseline_q30,
      all_of(analysis_baseline_variables)
    )

  expected_keys <- bind_rows(
    prepared |> select(id, week),
    death_template |> select(id, week)
  )
  if (
    anyNA(completed$id) ||
      anyNA(completed$y) ||
      anyDuplicated(completed[c("id", "week")]) ||
      !setequal(key_string(completed), key_string(expected_keys)) ||
      !setequal(completed$id, prepared_full$id) ||
      !setequal(completed$id[completed$y == 8L], death_template$id) ||
      any(!completed$y %in% 1:8)
  ) {
    stop("Completed imputation ", index, " failed the patient, week, outcome, or death checks.")
  }

  saveRDS(long_imputation, long_imputation_path, compress = "xz")
  arrow::write_parquet(completed, completed_path, compression = "zstd")

  run_summaries[[position]] <- tibble(
    imputation = index,
    rows = nrow(completed),
    participants = n_distinct(completed$id),
    imputed_q30 = nrow(long_imputation$imp$q30),
    deaths = sum(completed$y == 8L),
    logged_events = if (is.null(long_imputation$loggedEvents)) 0L else nrow(long_imputation$loggedEvents),
    reused = FALSE
  )
}

run_summary <- bind_rows(run_summaries)
if (!identical(run_summary$imputation, indices)) {
  stop("The imputation run summary is incomplete or out of order.")
}
arrow::write_parquet(run_summary, run_summary_path, compression = "zstd")
```
