---
title: "QoL death-inclusive Markov — imputation"
subtitle: "Independent long-format imputation using 50 completed datasets"
format: html
execute:
  freeze: false
---

Baseline variables are first imputed in one patient-level dataset using the
other baseline variables, survival information, and mean observed follow-up
QoL. Each completed baseline dataset is then joined to the weekly data before
living QoL states 1--7 are imputed. The previous-state predictor is based on
the completed baseline state through the first observed follow-up QoL value;
thereafter, the most recent observed follow-up value is carried forward. One
observed transition to death is added after imputation; death is never a MICE
target.

```{r}
#| label: setup-death-inclusive-longitudinal-qol-imputation
#| include: false

source(here::here("R", "analysis-config.R"))
library(tidyverse)

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
indices <- seq_len(config$imputation$main$m)

input_path <- here::here(
  "artifacts", "primary", "quality-of-life",
  "death-inclusive-longitudinal-prepared.parquet"
)
output_dir <- here::here(
  "artifacts", "supporting", "qol-death-inclusive-longitudinal"
)
imputation_dir <- file.path(output_dir, "imputations")
imputation_fit_dir <- file.path(output_dir, "imputation-fits")
prepared_path <- file.path(output_dir, "prepared.parquet")
baseline_imputation_path <- file.path(output_dir, "baseline-imputations.rds")

dir.create(imputation_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(imputation_fit_dir, recursive = TRUE, showWarnings = FALSE)
```

## Prepare the weekly imputation data

```{r}
#| label: prepare-death-inclusive-longitudinal-qol-imputation
#| include: false

prepared_full <- arrow::read_parquet(input_path) |>
  as.data.frame() |>
  arrange(id, week)

patient_survival <- prepared_full |>
  distinct(id, survival_time_days, event_death)

patient_survival$na_est <- mice::nelsonaalen(
  patient_survival,
  timevar = survival_time_days,
  statusvar = event_death
)

mean_followup_qol <- prepared_full |>
  filter(!is.na(q30), q30 %in% 1:7) |>
  summarise(mean_observed_q30 = mean(as.numeric(q30)), .by = id)

week_basis <- Hmisc::rcspline.eval(1:26, nk = 4, inclx = TRUE)
week_basis <- as.data.frame(week_basis)
colnames(week_basis) <- c("week", "week_rcs_1", "week_rcs_2")

prepared_full <- prepared_full |>
  left_join(patient_survival |> select(id, na_est), by = "id") |>
  left_join(week_basis, by = "week") |>
  mutate(
    gender = factor(gender),
    age_scaled = (age - mean(age, na.rm = TRUE)) / sd(age, na.rm = TRUE),
    ecog_fstcnt = factor(
      if_else(ecog_fstcnt == 4, 3, ecog_fstcnt),
      ordered = TRUE
    ),
    plan_fstcnt_coded = fct_lump_n(factor(plan_fstcnt_coded), n = 4),
    diagnosis_code = factor(diagnosis_code),
    stage_binary = factor(stage_binary), 
    baseline_q30 = factor(baseline_q30, ordered = TRUE)
  )

# Complete weeks are retained until death or
# censoring; an observed response in a partial censoring week is also retained.
prepared <- prepared_full |>
  filter(
    (event_death == 1L & week < death_week) |
      (event_death == 0L &
        (week * 7L <= survival_time_days | !is.na(q30)))
  ) |>
  select(-y_death_inclusive)

death_template <- prepared_full |>
  filter(event_death == 1L, week == death_week) |>
  select(-y_death_inclusive)

arrow::write_parquet(prepared, prepared_path, compression = "zstd")
```

## Impute patient-level variables

```{r}
#| label: prepare-death-inclusive-qol-baseline-imputation
#| include: false

baseline_variables <- c(
  "age_scaled", "gender", "ecog_fstcnt", "diagnosis_code", "stage_binary",
  "plan_fstcnt_coded", "albumin", "c_reactive_protein", "baseline_q30"
)
baseline_targets <- c(
  "ecog_fstcnt", "diagnosis_code", "plan_fstcnt_coded", "albumin",
  "c_reactive_protein", "baseline_q30", "mean_observed_q30"
)

baseline_data <- prepared_full |>
  distinct(id, .keep_all = TRUE) |>
  select(id, tx, all_of(baseline_variables), event_death, na_est) |>
  left_join(mean_followup_qol, by = "id")

run_baseline_imputation <- force || !file.exists(baseline_imputation_path)
```

```{r}
#| label: run-death-inclusive-qol-baseline-imputation
#| include: false

if (run_baseline_imputation) {
  methods <- mice::make.method(baseline_data)
  # methods[setdiff(names(methods), baseline_targets)] <- ""

  predictors <- mice::make.predictorMatrix(baseline_data)
  predictors[, "id"] <- 0L

  baseline_imputations <- mice::mice(
    baseline_data,
    m = config$imputation$main$m,
    maxit = config$imputation$supporting$maxit,
    method = methods,
    predictorMatrix = predictors,
    donors = config$imputation$supporting$donors,
    seed = analysis_seed(config, "supporting", offset = 500L),
    printFlag = TRUE
  )

  saveRDS(baseline_imputations, baseline_imputation_path, compress = "xz")
}

baseline_imputations <- readRDS(baseline_imputation_path)
completed_baseline_variables <- setdiff(
  baseline_targets,
  "mean_observed_q30"
)
```

## Configure longitudinal QoL imputation

```{r}
#| label: configure-death-inclusive-qol-longitudinal-imputation
#| include: false

# This is a fixed auxiliary history derived only from observed follow-up QoL.
# Each observed value becomes available from the following week and is carried
# forward. Completed baseline QoL supplies the history through the first
# observed follow-up value after the history is joined within each imputation.
history_proxy <- prepared_full |>
  select(id, week, q30) |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(yprev_imputation = lag(q30)) |>
  fill(yprev_imputation, .direction = "down") |>
  ungroup() |>
  select(id, week, yprev_imputation)

# including post baline na-est and event_death, because they are also part of the markov model, even though functional mismatch
q30_predictors <- c(
  "tx", "age_scaled", "gender", "ecog_fstcnt", "diagnosis_code",
  "stage_binary", "plan_fstcnt_coded", "albumin", "c_reactive_protein",
  "baseline_q30", "event_death", "na_est", "week", "week_rcs_1",
  "week_rcs_2", "yprev_imputation"
)

analysis_baseline_variables <- c(
  "tx", "age", "gender", "diagnosis_code", "stage_binary", "ecog_fstcnt",
  "plan_fstcnt_coded", "albumin", "c_reactive_protein"
)
```

## Impute weekly QoL and add death

```{r}
#| label: run-death-inclusive-qol-longitudinal-imputation
#| include: false

for (index in indices) {
  completed_path <- file.path(
    imputation_dir,
    sprintf("imputation-%02d.parquet", index)
  )
  long_imputation_path <- file.path(
    imputation_fit_dir,
    sprintf("long-imputation-%02d.rds", index)
  )

  if (
    !force &&
      !run_baseline_imputation &&
      file.exists(completed_path) &&
      file.exists(long_imputation_path)
  ) next

  completed_baseline <- mice::complete(baseline_imputations, index) |>
    as.data.frame() |>
    select(id, all_of(completed_baseline_variables))

  imputation_data <- prepared |>
    select(-all_of(completed_baseline_variables)) |>
    left_join(completed_baseline, by = "id") |>
    left_join(history_proxy, by = c("id", "week")) |>
    mutate(
      yprev_imputation = coalesce(
        yprev_imputation,
        as.numeric(as.character(baseline_q30))
      ),
      yprev_imputation = factor(yprev_imputation, levels = 1:7),
      q30 = factor(q30, levels = 1:7, ordered = TRUE),
      tx = as.numeric(as.character(tx))
    )

  methods <- mice::make.method(imputation_data)
  methods[] <- ""
  methods["q30"] <- "polr"

  predictors <- mice::make.predictorMatrix(imputation_data)
  predictors[,] <- 0L
  predictors["q30", q30_predictors] <- 1L

  long_imputation <- mice::mice(
    imputation_data,
    m = 1L,
    maxit = config$imputation$supporting$maxit,
    method = methods,
    predictorMatrix = predictors,
    seed = analysis_seed(config, "supporting", index, offset = 550L),
    polr.to.loggedEvents = TRUE,
    printFlag = TRUE
  )

  saveRDS(long_imputation, long_imputation_path, compress = "xz")

  living_completed <- mice::complete(long_imputation, 1L) |>
    as.data.frame() |>
    mutate(
      q30 = as.integer(as.character(q30)),
      y = q30
    )

  death_rows <- death_template |>
    select(-all_of(completed_baseline_variables)) |>
    left_join(completed_baseline, by = "id") |>
    mutate(q30 = NA_integer_, y = 8L)

  completed <- bind_rows(living_completed, death_rows) |>
    arrange(id, week) |>
    group_by(id) |>
    mutate(
      yprev = coalesce(
        lag(y),
        as.integer(as.character(first(baseline_q30)))
      ),
      yprev = factor(yprev, levels = 1:7)
    ) |>
    ungroup() |>
    select(
      id, y, q30, yprev, week, baseline_q30,
      all_of(analysis_baseline_variables)
    )

  arrow::write_parquet(completed, completed_path, compression = "zstd")
}
```
