---
title: "QoL death-inclusive Markov — diagnostics and description"
subtitle: "Imputation diagnostics, posterior diagnostics, and raw QoL trajectories"
format: html
execute:
  freeze: false
---

The observed QoL assessments are sparse and irregular rather than weekly.
Consequently, empirical weekly state-occupancy, transition, correlation, and
variogram plots would describe the observation schedule as much as the outcome
process and are not produced. Diagnostics instead assess the actual MICE
chains and fitted posterior distributions. A separate descriptive figure shows
the raw observed QoL values at their actual assessment days without treating
unobserved weeks as observed states.

## Raw observed QoL trajectories

```{r}
#| label: describe-raw-longitudinal-qol
#| include: false

source(here::here("R", "analysis-config.R"))
source(here::here("R", "markov-helpers.R"))
library(tidyverse)

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
indices <- seq_len(config$imputation$main$m)
output_dir <- here::here(
  "artifacts", "supporting", "qol-death-inclusive-longitudinal"
)
imputation_dir <- file.path(output_dir, "imputations")
imputation_fit_dir <- file.path(output_dir, "imputation-fits")
model_dir <- file.path(output_dir, "models")
diagnostic_dir <- file.path(output_dir, "diagnostics")
stan_dx_dir <- file.path(diagnostic_dir, "stan-dx")
imputation_diagnostic_dir <- file.path(diagnostic_dir, "imputation")
figure_dir <- file.path(output_dir, "figures")
result_dir <- here::here(
  "results", "supporting", "qol-death-inclusive-longitudinal"
)
dir.create(diagnostic_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(stan_dx_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(imputation_diagnostic_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)

cohort <- arrow::read_parquet(here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
))
qol <- arrow::read_parquet(here::here(
  "analysis", "00-data", "results", "analysis-qol.parquet"
))
baseline_qol <- cohort |>
  transmute(
    id,
    assessment_id = NA_character_,
    day = 0,
    q30 = baseline_q30
  ) |>
  filter(!is.na(q30))
followup_qol <- qol |>
  filter(
    !is.na(q30),
    day > 0,
    day <= config$analysis$horizon_days
  ) |>
  select(id, assessment_id, day, q30)
raw_qol <- bind_rows(baseline_qol, followup_qol) |>
  left_join(cohort |> select(id, tx), by = "id") |>
  distinct(id, assessment_id, day, q30, .keep_all = TRUE) |>
  arrange(id, day) |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Intervention"))
  )

raw_trajectory_plot <- ggplot(
  raw_qol,
  aes(x = day, y = q30, group = id)
) +
  geom_line(colour = "grey35", alpha = 0.20, linewidth = 0.45) +
  geom_point(
    colour = "black",
    fill = "grey70",
    shape = 21,
    size = 1.8,
    stroke = 0.35,
    alpha = 0.50,
    position = position_jitter(width = 0, height = 0.035)
  ) +
  facet_wrap(vars(tx), nrow = 1) +
  scale_x_continuous(
    breaks = c(0, 28, 56, 84, 112, 140, 168, 182),
    limits = c(0, config$analysis$horizon_days)
  ) +
  scale_y_continuous(breaks = 1:7, limits = c(0.8, 7.2)) +
  labs(
    x = "Day since first contact",
    y = "QoL state (1 = excellent, 7 = very poor)"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )

ggsave(
  file.path(figure_dir, "raw-qol-trajectories.png"),
  raw_trajectory_plot,
  width = 12,
  height = 6.5,
  dpi = 300
)
```

```{r}
#| label: fig-death-inclusive-qol-raw-trajectories
#| fig-cap: "Raw observed longitudinal QoL values by randomized group"
#| fig-alt: "Two faceted panels show observed QoL states from 1, excellent, to 7, very poor, over 182 days. Semi-transparent lines connect repeated observations from the same participant and points mark actual questionnaire dates; gaps contain no assumed weekly observations."
#| fig-width: 12
#| fig-height: 6.5
#| echo: false

raw_trajectory_plot
```

## Imputation diagnostics

```{r}
#| label: save-death-inclusive-qol-imputation-diagnostics
#| include: false

prepared <- arrow::read_parquet(file.path(output_dir, "prepared.parquet")) |>
  as.data.frame()
baseline_imputations <- readRDS(file.path(output_dir, "baseline-imputations.rds"))
baseline_trace_variables <- intersect(
  c("baseline_q30", "ecog_fstcnt", "mean_observed_q30"),
  names(baseline_imputations$data)
)

baseline_trace_plot <- plot(baseline_imputations)
grDevices::png(
  file.path(imputation_diagnostic_dir, "baseline-traces.png"),
  width = 1800,
  height = 1000,
  res = 180
)
print(baseline_trace_plot)
grDevices::dev.off()

baseline_density_plot <- mice::densityplot(
  baseline_imputations)

grDevices::png(
  file.path(imputation_diagnostic_dir, "baseline-observed-vs-imputed.png"),
  width = 1800,
  height = 1000,
  res = 180
)
print(baseline_density_plot)
grDevices::dev.off()

long_trace_data <- purrr::map_dfr(indices, function(index) {
  fit <- readRDS(file.path(
    imputation_fit_dir,
    sprintf("long-imputation-%02d.rds", index)
  ))
  tibble(
    imputation = factor(index),
    iteration = seq_len(config$imputation$supporting$maxit),
    mean_imputed_qol = as.numeric(fit$chainMean["q30", , 1L])
  )
})
long_trace_plot <- ggplot(
  long_trace_data,
  aes(x = iteration, y = mean_imputed_qol, group = imputation)
) +
  geom_line(colour = "#D95F02", alpha = 0.30, linewidth = 0.45) +
  labs(
    x = "MICE iteration",
    y = "Mean imputed QoL state"
  ) +
  theme_bw() +
  theme(legend.position = "none")

long_density_data <- bind_rows(
  tibble(
    source = "Observed",
    imputation = NA_integer_,
    value = prepared$q30[!is.na(prepared$q30)]
  ),
  purrr::map_dfr(indices, function(index) {
    completed <- arrow::read_parquet(file.path(
      imputation_dir,
      sprintf("imputation-%02d.parquet", index)
    )) |>
      filter(y != 8L) |>
      select(id, week, q30)
    prepared |>
      filter(is.na(q30)) |>
      select(id, week) |>
      inner_join(completed, by = c("id", "week")) |>
      transmute(
        source = "Imputed",
        imputation = index,
        value = q30
      )
  })
)
long_density_plot <- ggplot() +
  geom_density(
    data = filter(long_density_data, source == "Imputed"),
    aes(x = value, group = imputation),
    colour = "#D95F02",
    alpha = 0.18,
    linewidth = 0.45,
    adjust = 1.4
  ) +
  geom_density(
    data = filter(long_density_data, source == "Observed"),
    aes(x = value),
    colour = "black",
    linewidth = 1.1,
    adjust = 1.4
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(1, 7)) +
  labs(
    x = "QoL state",
    y = "Density"
  ) +
  theme_bw()

ggsave(
  file.path(imputation_diagnostic_dir, "long-imputation-traces.png"),
  long_trace_plot,
  width = 8,
  height = 5,
  dpi = 180
)
ggsave(
  file.path(imputation_diagnostic_dir, "long-observed-vs-imputed.png"),
  long_density_plot,
  width = 8,
  height = 5,
  dpi = 180
)
```

## Posterior diagnostics

```{r}
#| label: collect-death-inclusive-qol-model-diagnostics
#| include: false

diagnostics <- purrr::map_dfr(indices, function(index) {
  path <- file.path(diagnostic_dir, sprintf("diagnostics-%02d.parquet", index))
  if (!file.exists(path)) stop("Missing longitudinal QoL diagnostic artifact: ", path)
  arrow::read_parquet(path)
})
arrow::write_parquet(
  diagnostics,
  file.path(result_dir, "diagnostics.parquet"),
  compression = "zstd"
)
```

One complete `stanDx()` text/`stanDxplot()` pair is saved for every fitted
imputation. This replaces separate posterior `mcmc_trace()` figures.

```{r}
#| label: save-death-inclusive-qol-stan-diagnostics
#| include: false

for (index in indices) {
  stan_dx_text_path <- file.path(stan_dx_dir, sprintf("stan-dx-%02d.txt", index))
  stan_dx_plot_path <- file.path(stan_dx_dir, sprintf("stan-dx-%02d.png", index))
  if (
    !force &&
      file.exists(stan_dx_text_path) &&
      file.exists(stan_dx_plot_path)
  ) next

  model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", index)))
  writeLines(capture.output(rmsb::stanDx(model)), stan_dx_text_path)
  stan_dx_plot <- rmsb::stanDxplot(model)
  ggsave(
    stan_dx_plot_path,
    stan_dx_plot,
    width = 11,
    height = max(6, 1.05 * length(unique(stan_dx_plot$data$param))),
    dpi = 180,
    limitsize = FALSE
  )
}
```
