---
title: "QoL death-inclusive Markov — estimand"
subtitle: "Difference in weeks with good or better QoL over 26 weeks"
format: html
execute:
  freeze: false
---

Both treatment paths are predicted for every participant in each completed
dataset. State occupancy probabilities and the difference in time spent in QoL
states 1–3 are calculated with `avg_sops()` and `avg_comparisons()`. Two hundred
posterior draws from each of 50 imputations are pooled with equal weight. Death
is state 8 and absorbing; positive differences in weeks in states 1–3 favour
intervention.

```{r}
#| label: derive-death-inclusive-longitudinal-qol-estimand
#| include: false

source(here::here("R", "analysis-config.R"))
library(tidyverse)

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
indices <- seq_len(config$imputation$main$m)
output_dir <- here::here(
  "artifacts", "supporting", "qol-death-inclusive-longitudinal"
)
imputation_dir <- file.path(output_dir, "imputations")
model_dir <- file.path(output_dir, "models")
estimand_dir <- file.path(output_dir, "estimand")
figure_dir <- file.path(output_dir, "figures")
result_dir <- here::here(
  "results", "supporting", "qol-death-inclusive-longitudinal"
)
dir.create(estimand_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)

for (index in indices) {
  data_path <- file.path(
    imputation_dir,
    sprintf("imputation-%02d.parquet", index)
  )
  model_path <- file.path(model_dir, sprintf("model-%02d.rds", index))
  sop_summary_path <- file.path(
    estimand_dir,
    sprintf("sops-summary-%02d.parquet", index)
  )
  sop_draws_path <- file.path(
    estimand_dir,
    sprintf("sops-draws-%02d.parquet", index)
  )
  comparison_summary_path <- file.path(
    estimand_dir,
    sprintf("comparisons-summary-%02d.parquet", index)
  )
  comparison_draws_path <- file.path(
    estimand_dir,
    sprintf("comparisons-draws-%02d.parquet", index)
  )
  estimand_paths <- c(
    sop_summary_path,
    sop_draws_path,
    comparison_summary_path,
    comparison_draws_path
  )
  if (!force && all(file.exists(estimand_paths))) next

  data <- arrow::read_parquet(data_path) |>
    as.data.frame() |>
    mutate(
      y = factor(y, levels = 1:8, ordered = TRUE),
      yprev = factor(yprev, levels = 1:8)
    )
  model <- readRDS(model_path)
  observed <- data |>
    group_by(id) |>
    slice_head(n = 1L) |>
    ungroup() |>
    select(id, tx, diagnosis_code, ecog_fstcnt, yprev)
  estimand_seed <- analysis_seed(config, "supporting", index, offset = 600L)

  sops <- markov.misc::avg_sops(
    model = model,
    newdata = observed,
    variables = list(tx = c(0, 1)),
    times = 1:26,
    y_levels = 1:8,
    absorb = 8L,
    p_var = "yprev",
    id_var = "id",
    time_var = "week",
    include_re = FALSE,
    n_draws = config$markov$post_estimation_draws,
    seed = estimand_seed,
    posterior_summary = "mean",
    return_draws = TRUE
  )
  comparisons <- markov.misc::avg_comparisons(
    model = model,
    newdata = observed,
    variables = list(tx = c(0, 1)),
    estimand = "time_in_state",
    state_sets = 1:3,
    times = 1:26,
    y_levels = 1:8,
    absorb = 8L,
    p_var = "yprev",
    id_var = "id",
    time_var = "week",
    include_re = FALSE,
    n_draws = config$markov$post_estimation_draws,
    seed = estimand_seed,
    posterior_summary = "mean",
    return_draws = TRUE
  )
  sop_draws <- attr(sops, "draws")
  comparison_draws <- attr(comparisons, "draws")
  if (is.null(sop_draws) || is.null(comparison_draws)) {
    stop("Longitudinal QoL post-estimation results must retain posterior draws.")
  }

  arrow::write_parquet(
    as.data.frame(sops),
    sop_summary_path,
    compression = "zstd"
  )
  arrow::write_parquet(
    as.data.frame(sop_draws),
    sop_draws_path,
    compression = "zstd"
  )
  arrow::write_parquet(
    as.data.frame(comparisons),
    comparison_summary_path,
    compression = "zstd"
  )
  arrow::write_parquet(
    as.data.frame(comparison_draws),
    comparison_draws_path,
    compression = "zstd"
  )
}

comparison_draws <- purrr::map_dfr(seq_along(indices), function(position) {
  index <- indices[[position]]
  draws <- arrow::read_parquet(file.path(
    estimand_dir,
    sprintf("comparisons-draws-%02d.parquet", index)
  )) |>
    as.data.frame()
  estimate_column <- if ("estimate" %in% names(draws)) {
    "estimate"
  } else {
    "comparison"
  }
  if (!estimate_column %in% names(draws)) {
    stop("Could not identify the longitudinal QoL treatment contrast.")
  }
  draws |>
    mutate(
      imputation = index,
      draw = .data[[estimate_column]]
    ) |>
    select(
      imputation,
      any_of("draw_id"),
      draw,
      any_of(c("contrast", "comparison", "term", "state"))
    )
})

summary <- comparison_draws |>
  summarise(
    estimand = "difference in weeks in QoL states 1-3, death included",
    posterior_mean = mean(draw),
    posterior_median = median(draw),
    q025 = quantile(draw, 0.025, names = FALSE),
    q975 = quantile(draw, 0.975, names = FALSE),
    probability_benefit = mean(draw > 0),
    imputations = n_distinct(imputation),
    posterior_draws_per_imputation = config$markov$post_estimation_draws,
    computationally_reduced = FALSE
  )

arrow::write_parquet(
  comparison_draws,
  file.path(result_dir, "draws.parquet"),
  compression = "zstd"
)
arrow::write_parquet(
  summary,
  file.path(result_dir, "summary.parquet"),
  compression = "zstd"
)
```

The model-derived state occupancy curves pool equal numbers of posterior draws
from all 50 completed datasets. They are not empirical weekly SOPs: each curve
is a posterior prediction from the fitted Markov models.

```{r}
#| label: plot-death-inclusive-longitudinal-qol-estimands
#| include: false

sop_draws <- purrr::map_dfr(indices, function(index) {
  arrow::read_parquet(file.path(
    estimand_dir,
    sprintf("sops-draws-%02d.parquet", index)
  )) |>
    mutate(imputation = index)
})

state_labels <- c(
  "Excellent", "Very good", "Good", "Average",
  "Below average", "Poor", "Very poor", "Dead"
)
pooled_sop_summary <- sop_draws |>
  mutate(
    state = factor(state, levels = 1:8, labels = state_labels),
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Intervention"))
  ) |>
  group_by(tx, time, state) |>
  summarise(
    conf.low = quantile(estimate, 0.025, names = FALSE),
    conf.high = quantile(estimate, 0.975, names = FALSE),
    estimate = mean(estimate),
    .groups = "drop"
  )
arrow::write_parquet(
  pooled_sop_summary,
  file.path(result_dir, "state-occupancy-summary.parquet"),
  compression = "zstd"
)

model_sop_plot <- suppressMessages(
  markov.misc::plot_sops(
    pooled_sop_summary,
    facet_var = NULL,
    time_var = "time",
    estimate_var = "estimate",
    show_uncertainty = TRUE,
    linetype_var = "tx",
    ribbon_alpha = 0.25
  ) +
    scale_colour_brewer(palette = "Dark2") +
    scale_fill_brewer(palette = "Dark2")
) +
  labs(
    x = "Week",
    y = "Posterior state occupancy probability",
    colour = "QoL/death state",
    fill = "QoL/death state",
    linetype = "Randomized group"
  ) +
  theme_bw()

comparison_posterior_plot <- ggplot(
  comparison_draws,
  aes(x = draw)
) +
  ggdist::stat_dots(side = "bottom") +
  ggdist::stat_halfeye(
    .width = c(0.66, 0.95),
    point_interval = ggdist::median_qi
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(
    x = paste0(
      "Difference in weeks with good or better QoL over 26 weeks ",
      "(intervention - control)"
    ),
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

ggsave(
  file.path(figure_dir, "model-state-occupancy-pooled.png"),
  model_sop_plot,
  width = 10,
  height = 6,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "good-qol-time-difference-posterior.png"),
  comparison_posterior_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

```{r}
#| label: fig-death-inclusive-qol-model-sops
#| fig-cap: "Pooled model-derived QoL/death state occupancy probabilities by randomized group"
#| fig-alt: "Posterior state occupancy curves over 26 weeks for seven ordered QoL states and death. Colors distinguish states, line types distinguish randomized groups, and shaded ribbons show 95 percent credible intervals."
#| fig-width: 10
#| fig-height: 6
#| echo: false

model_sop_plot
```

```{r}
#| label: fig-death-inclusive-qol-comparison-posterior
#| fig-cap: "Posterior distribution of the treatment difference in weeks with good or better QoL"
#| fig-alt: "A half-eye density above a posterior dot plot of the intervention-minus-control difference in weeks spent in QoL states 1 through 3 over 26 weeks. A dashed vertical line marks no difference."
#| fig-width: 8
#| fig-height: 5
#| echo: false

comparison_posterior_plot
```
