---
title: "QoL death-inclusive Markov — model diagnostics"
subtitle: "Posterior sampling quality across 8 completed datasets"
format:
  html:
    toc: true
execute:
  freeze: false
---

This document contains only posterior sampling diagnostics for the eight fitted
Markov models. Imputation diagnostics are reported separately. The configured
R-hat, effective sample size, divergence, and treedepth thresholds are applied
to every model, with E-BFMI checked when available; a complete `stanDx()` text
report and `stanDxplot()` are also retained for each imputation.

```{r}
#| label: collect-death-inclusive-qol-model-diagnostics
#| include: false

source(here::here("R", "analysis-config.R"))
source(here::here("R", "markov-helpers.R"))
library(tidyverse)

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
n_imputations <- 8L
indices <- seq_len(n_imputations)
output_dir <- here::here(
  "artifacts", "supporting", "qol-death-inclusive-longitudinal"
)
model_dir <- file.path(output_dir, "models")
diagnostic_dir <- file.path(output_dir, "diagnostics", "model")
stan_dx_dir <- file.path(diagnostic_dir, "stan-dx")
figure_dir <- file.path(output_dir, "figures", "model-diagnostics")
result_dir <- here::here(
  "results", "supporting", "qol-death-inclusive-longitudinal",
  "model-diagnostics"
)
dir.create(diagnostic_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(stan_dx_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)

diagnostics <- purrr::map_dfr(indices, function(index) {
  path <- file.path(
    diagnostic_dir,
    sprintf("diagnostics-%02d.parquet", index)
  )
  if (!file.exists(path)) {
    stop("Missing longitudinal QoL model diagnostic artifact: ", path)
  }
  arrow::read_parquet(path) |>
    as.data.frame()
})
fit_attempt_path <- file.path(diagnostic_dir, "fit-attempts-all.parquet")
if (!file.exists(fit_attempt_path)) {
  stop("Missing longitudinal QoL fit-attempt audit log: ", fit_attempt_path)
}
fit_attempts <- arrow::read_parquet(fit_attempt_path) |>
  as.data.frame()
final_fit_attempts <- fit_attempts |>
  slice_max(attempt, n = 1L, with_ties = FALSE, by = imputation)
if (
  !setequal(fit_attempts$imputation, indices) ||
    any(!final_fit_attempts$diagnostics_passed)
) {
  stop("The final fit attempt did not pass for every imputation.")
}

finite_min <- function(value) {
  if (any(is.finite(value))) min(value[is.finite(value)]) else NA_real_
}
finite_max <- function(value) {
  if (any(is.finite(value))) max(value[is.finite(value)]) else NA_real_
}

diagnostic_summary <- diagnostics |>
  summarise(
    max_rhat = finite_max(rhat),
    min_ess_bulk = finite_min(ess_bulk),
    min_ess_tail = finite_min(ess_tail),
    divergent_transitions = first(divergent_transitions),
    treedepth_exceeded = first(treedepth_exceeded),
    min_bfmi = first(min_bfmi),
    .by = imputation
  ) |>
  mutate(
    passed =
      max_rhat <= config$diagnostics$max_rhat &
      min_ess_bulk >= config$diagnostics$min_ess_bulk &
      min_ess_tail >= config$diagnostics$min_ess_tail &
      divergent_transitions <= config$diagnostics$max_divergent_transitions &
      treedepth_exceeded <= config$diagnostics$max_treedepth_exceeded &
      (is.na(min_bfmi) | min_bfmi >= config$diagnostics$min_bfmi)
  )
if (nrow(diagnostic_summary) != n_imputations || any(!diagnostic_summary$passed)) {
  stop("At least one longitudinal QoL Markov model failed posterior diagnostics.")
}

arrow::write_parquet(
  diagnostics,
  file.path(result_dir, "parameter-diagnostics.parquet"),
  compression = "zstd"
)
arrow::write_parquet(
  diagnostic_summary,
  file.path(result_dir, "model-diagnostic-summary.parquet"),
  compression = "zstd"
)
```

```{r}
#| label: tbl-death-inclusive-qol-model-diagnostics
#| tbl-cap: "Worst posterior sampling diagnostic within each fitted imputation"
#| echo: false

diagnostic_summary |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  knitr::kable()
```

```{r}
#| label: tbl-death-inclusive-qol-fit-attempts
#| tbl-cap: "Model fit attempts and recorded reasons for refitting"
#| echo: false

fit_attempts |>
  transmute(
    imputation,
    attempt,
    iterations_per_chain,
    retained_draws_per_chain,
    diagnostics_passed,
    reason = refit_reason
  ) |>
  knitr::kable()
```

```{r}
#| label: save-death-inclusive-qol-stan-diagnostics
#| include: false

for (index in indices) {
  stan_dx_text_path <- file.path(
    stan_dx_dir,
    sprintf("stan-dx-%02d.txt", index)
  )
  stan_dx_plot_path <- file.path(
    stan_dx_dir,
    sprintf("stan-dx-%02d.png", index)
  )
  if (
    !force &&
      file.exists(stan_dx_text_path) &&
      file.exists(stan_dx_plot_path)
  ) next

  model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", index)))
  writeLines(capture.output(rmsb::stanDx(model)), stan_dx_text_path)
  stan_dx_plot <- rmsb::stanDxplot(model)
  ggsave(
    stan_dx_plot_path,
    stan_dx_plot,
    width = 11,
    height = max(6, 1.05 * length(unique(stan_dx_plot$data$param))),
    dpi = 180,
    limitsize = FALSE
  )
}
```

```{r}
#| label: derive-death-inclusive-qol-diagnostic-overview
#| include: false

rhat_plot <- diagnostics |>
  filter(is.finite(rhat)) |>
  ggplot(aes(x = factor(imputation), y = rhat)) +
  geom_hline(
    yintercept = config$diagnostics$max_rhat,
    linetype = "dashed",
    colour = "#D95F02"
  ) +
  geom_violin(fill = "grey85", colour = "grey45", scale = "width") +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  labs(x = "Imputation", y = "R-hat") +
  theme_bw()

ess_plot <- diagnostics |>
  select(imputation, ess_bulk, ess_tail) |>
  pivot_longer(c(ess_bulk, ess_tail), names_to = "metric", values_to = "ess") |>
  filter(is.finite(ess)) |>
  mutate(metric = recode(metric, ess_bulk = "Bulk ESS", ess_tail = "Tail ESS")) |>
  ggplot(aes(x = factor(imputation), y = ess, fill = metric)) +
  geom_hline(
    yintercept = config$diagnostics$min_ess_bulk,
    linetype = "dashed",
    colour = "#D95F02"
  ) +
  geom_boxplot(position = position_dodge(width = 0.75), outlier.alpha = 0.25) +
  labs(x = "Imputation", y = "Effective sample size", fill = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(
  file.path(figure_dir, "rhat-by-imputation.png"),
  rhat_plot,
  width = 9,
  height = 5,
  dpi = 220
)
ggsave(
  file.path(figure_dir, "ess-by-imputation.png"),
  ess_plot,
  width = 9,
  height = 5,
  dpi = 220
)
```

```{r}
#| label: fig-death-inclusive-qol-posterior-diagnostics
#| fig-cap: "Posterior convergence and effective sample size across eight imputations"
#| fig-subcap:
#|   - "Parameter-level R-hat distributions; the dashed line is the configured maximum"
#|   - "Parameter-level bulk and tail effective sample sizes; the dashed line is the configured minimum"
#| fig-alt: "Two panels summarize posterior diagnostics for each of eight Markov models. R-hat values remain below the configured threshold and effective sample sizes remain above the configured minimum."
#| layout-ncol: 1
#| fig-width: 9
#| fig-height: 5
#| echo: false

rhat_plot
ess_plot
```
