---
title: "QoL death-inclusive Markov — state occupancy and estimand"
subtitle: "Difference in weeks with good-or-better QoL over 26 weeks"
format:
  html:
    toc: true
execute:
  freeze: false
---

Both treatment paths are predicted for every participant in each of eight
completed datasets. Posterior state occupancy probability (SOP) draws are
pooled with equal weight across imputations. The same draws are reported for
the original seven QoL states plus death and for the SAP grouping: Good
(states 1--3), Average (state 4), Bad (states 5--7), and Death (state 8).

The estimand is calculated directly from the matched SOP draws. Within each
posterior draw and treatment path, probabilities for states 1--3 are summed
over all 26 weeks. The intervention-minus-control difference is then formed
within that same draw. Positive values favour intervention. Because this is an
eight-imputation test cycle rather than the final 50-imputation run, outputs are
explicitly flagged as computationally reduced.

```{r}
#| label: derive-death-inclusive-longitudinal-qol-sops
#| include: false

source(here::here("R", "analysis-config.R"))
library(tidyverse)

config <- read_analysis_config()
# force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
force <- TRUE
n_imputations <- 8L
indices <- seq_len(n_imputations)
output_dir <- here::here(
  "artifacts", "supporting", "qol-death-inclusive-longitudinal"
)
imputation_dir <- file.path(output_dir, "imputations")
model_dir <- file.path(output_dir, "models")
estimand_dir <- file.path(output_dir, "estimand")
figure_dir <- file.path(output_dir, "figures", "estimand")
result_dir <- here::here(
  "results", "supporting", "qol-death-inclusive-longitudinal",
  "estimand"
)
dir.create(estimand_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)

for (index in indices) {
  data_path <- file.path(
    imputation_dir,
    sprintf("imputation-%02d.parquet", index)
  )
  model_path <- file.path(model_dir, sprintf("model-%02d.rds", index))
  sop_summary_path <- file.path(
    estimand_dir,
    sprintf("sops-summary-%02d.parquet", index)
  )
  sop_draws_path <- file.path(
    estimand_dir,
    sprintf("sops-draws-%02d.parquet", index)
  )
  if (!force && file.exists(sop_summary_path) && file.exists(sop_draws_path)) {
    next
  }
  if (!file.exists(data_path) || !file.exists(model_path)) {
    stop("Imputation and model artifacts are required for estimand ", index, ".")
  }

  data <- arrow::read_parquet(data_path) |>
    as.data.frame() |>
    mutate(
      y = factor(y, levels = 1:8, ordered = TRUE),
      yprev = factor(yprev, levels = 1:7),
      ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE),
      diagnosis_code = factor(diagnosis_code)
    )
  model <- readRDS(model_path)
  observed <- data |>
    slice_head(n = 1L, by = id) |>
    # Retain a numeric week column so mostr resolves the time axis as
    # continuous rather than mistaking restricted-cubic-spline knots for
    # factor levels when `newdata` omits the fitted time variable.
    select(id, tx, week, diagnosis_code, ecog_fstcnt, yprev)
  estimand_seed <- analysis_seed(config, "supporting", index, offset = 600L)

  sops <- mostr::avg_sops(
    model = model,
    # newdata = observed,
    variables = list(tx = c(0, 1)),
    times = 1:26,
    y_levels = 1:8,
    absorb = 8L,
    p_var = "yprev",
    id_var = "id",
    time_var = "week",
    include_re = TRUE,
    n_draws = config$markov$post_estimation_draws,
    seed = estimand_seed,
    posterior_summary = "median",
    return_draws = TRUE
  )
  sop_draws <- attr(sops, "draws")
  if (is.null(sop_draws)) {
    stop("Longitudinal QoL SOP results must retain posterior draws.")
  }
  sop_draws <- as.data.frame(sop_draws)
  required_sop_columns <- c("draw_id", "estimate", "time", "state", "tx")
  if (!all(required_sop_columns %in% names(sop_draws))) {
    stop("Longitudinal QoL SOP draws lack matched draw, state, time, or treatment columns.")
  }
  if (
    n_distinct(sop_draws$draw_id) != config$markov$post_estimation_draws ||
      anyNA(sop_draws[required_sop_columns])
  ) {
    stop("Longitudinal QoL SOP draws are incomplete for imputation ", index, ".")
  }

  arrow::write_parquet(
    as.data.frame(sops),
    sop_summary_path,
    compression = "zstd"
  )
  arrow::write_parquet(
    sop_draws,
    sop_draws_path,
    compression = "zstd"
  )
}
```

## Pool and integrate matched SOP draws

```{r}
#| label: pool-death-inclusive-longitudinal-qol-sops
#| include: false

ungrouped_sop_draws <- purrr::map_dfr(indices, function(index) {
  arrow::read_parquet(file.path(
    estimand_dir,
    sprintf("sops-draws-%02d.parquet", index)
  )) |>
    as.data.frame() |>
    mutate(
      imputation = index,
      state = as.integer(as.character(state)),
      tx = as.integer(as.character(tx))
    )
})

draw_sums <- ungrouped_sop_draws |>
  summarise(
    probability = sum(estimate),
    .by = c(imputation, draw_id, tx, time)
  )
if (any(abs(draw_sums$probability - 1) > 1e-6)) {
  stop("At least one posterior SOP draw does not sum to one across states.")
}

state_labels <- c(
  "Excellent", "Very good", "Good", "Average",
  "Below average", "Poor", "Very poor", "Death"
)
ungrouped_sop_summary <- ungrouped_sop_draws |>
  mutate(
    state = factor(state, levels = 1:8, labels = state_labels),
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Intervention"))
  ) |>
  summarise(
    conf.low = quantile(estimate, 0.025, names = FALSE),
    conf.high = quantile(estimate, 0.975, names = FALSE),
    estimate = median(estimate),
    .by = c(tx, time, state)
  )

grouped_sop_draws <- ungrouped_sop_draws |>
  mutate(
    state = case_when(
      state %in% 1:3 ~ "Good",
      state == 4L ~ "Average",
      state %in% 5:7 ~ "Bad",
      state == 8L ~ "Death",
      TRUE ~ NA_character_
    )
  ) |>
  summarise(
    estimate = sum(estimate),
    .by = c(imputation, draw_id, tx, time, state)
  )
if (anyNA(grouped_sop_draws$state)) {
  stop("A QoL/death state could not be assigned to the SAP grouping.")
}

grouped_sop_summary <- grouped_sop_draws |>
  mutate(
    state = factor(state, levels = c("Good", "Average", "Bad", "Death")),
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Intervention"))
  ) |>
  summarise(
    conf.low = quantile(estimate, 0.025, names = FALSE),
    conf.high = quantile(estimate, 0.975, names = FALSE),
    estimate = median(estimate),
    .by = c(tx, time, state)
  )

time_in_group_draws <- grouped_sop_draws |>
  summarise(
    weeks = sum(estimate),
    .by = c(imputation, draw_id, tx, state)
  )
comparison_draws <- time_in_group_draws |>
  filter(state == "Good") |>
  select(imputation, draw_id, tx, weeks) |>
  pivot_wider(names_from = tx, values_from = weeks, names_prefix = "tx_") |>
  mutate(
    draw = tx_1 - tx_0,
    estimand = "Difference in weeks with good-or-better QoL"
  )
if (
  nrow(comparison_draws) !=
    n_imputations * config$markov$post_estimation_draws ||
    anyNA(comparison_draws$draw)
) {
  stop("The integrated good-QoL contrast must retain one matched contrast per draw.")
}

estimand_summary <- comparison_draws |>
  summarise(
    estimand = first(estimand),
    posterior_mean = mean(draw),
    posterior_median = median(draw),
    q025 = quantile(draw, 0.025, names = FALSE),
    q975 = quantile(draw, 0.975, names = FALSE),
    probability_benefit = mean(draw > 0),
    horizon_weeks = 26L,
    imputations = n_distinct(imputation),
    posterior_draws_per_imputation = config$markov$post_estimation_draws,
    computationally_reduced = TRUE
  )

outputs <- list(
  "state-occupancy-ungrouped-draws.parquet" = ungrouped_sop_draws,
  "state-occupancy-ungrouped-summary.parquet" = ungrouped_sop_summary,
  "state-occupancy-grouped-draws.parquet" = grouped_sop_draws,
  "state-occupancy-grouped-summary.parquet" = grouped_sop_summary,
  "time-in-group-draws.parquet" = time_in_group_draws,
  "good-qol-difference-draws.parquet" = comparison_draws,
  "good-qol-difference-summary.parquet" = estimand_summary
)
purrr::iwalk(outputs, function(data, filename) {
  arrow::write_parquet(
    as.data.frame(data),
    file.path(result_dir, filename),
    compression = "zstd"
  )
})
```

```{r}
#| label: tbl-death-inclusive-qol-estimand
#| tbl-cap: "Posterior summary of the intervention-minus-control difference in weeks with good-or-better QoL"
#| echo: false

estimand_summary |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  knitr::kable()
```

## State occupancy probabilities

The ungrouped display preserves all seven QoL states and death. The grouped
display implements the SAP categories without refitting or otherwise changing
the posterior draws.

```{r}
#| label: plot-death-inclusive-longitudinal-qol-sops
#| include: false

validate_sop_intervals <- function(data, label) {
  required <- c("estimate", "conf.low", "conf.high", "time", "state", "tx")
  if (
    !all(required %in% names(data)) ||
      anyNA(data[required]) ||
      any(data$conf.low > data$estimate) ||
      any(data$conf.high < data$estimate) ||
      all(abs(data$conf.high - data$conf.low) < .Machine$double.eps)
  ) {
    stop(label, " SOP summaries do not contain valid posterior intervals.")
  }
  invisible(data)
}

validate_sop_intervals(ungrouped_sop_summary, "Ungrouped")
validate_sop_intervals(grouped_sop_summary, "Grouped")

ungrouped_sop_plot <- ggplot(
  ungrouped_sop_summary,
  aes(
    x = time,
    y = estimate,
    colour = state,
    fill = state,
    linetype = tx,
    group = interaction(state, tx)
  )
) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    alpha = 0.20,
    colour = NA,
    linewidth = 0
  ) +
  geom_line(linewidth = 0.75) +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    x = "Week",
    y = "Posterior state occupancy probability",
    colour = "QoL/death state",
    fill = "QoL/death state",
    linetype = "Randomized group"
  ) +
  theme_bw()

grouped_sop_colours <- c(
  Good = "#1B9E77",
  Average = "#7570B3",
  Bad = "#D95F02",
  Death = "#1F1F1F"
)
grouped_sop_plot <- ggplot(
  grouped_sop_summary,
  aes(
    x = time,
    y = estimate,
    colour = state,
    fill = state,
    linetype = tx,
    group = interaction(state, tx)
  )
) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    alpha = 0.25,
    colour = NA,
    linewidth = 0
  ) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = grouped_sop_colours) +
  scale_fill_manual(values = grouped_sop_colours) +
  labs(
    x = "Week",
    y = "Posterior state occupancy probability",
    colour = "Grouped state",
    fill = "Grouped state",
    linetype = "Randomized group"
  ) +
  theme_bw()

comparison_posterior_plot <- ggplot(
  comparison_draws,
  aes(x = draw)
) +
  ggdist::stat_dots(side = "bottom") +
  ggdist::stat_halfeye(
    .width = c(0.66, 0.95),
    point_interval = ggdist::median_qi
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(
    x = paste0(
      "Difference in weeks with good-or-better QoL over 26 weeks ",
      "(intervention - control)"
    ),
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

estimand_plots <- list(
  "model-state-occupancy-ungrouped.png" = ungrouped_sop_plot,
  "model-state-occupancy-grouped.png" = grouped_sop_plot,
  "good-qol-time-difference-posterior.png" = comparison_posterior_plot
)
purrr::iwalk(estimand_plots, function(plot, filename) {
  ggsave(
    file.path(figure_dir, filename),
    plot,
    width = if (grepl("occupancy", filename)) 10 else 8,
    height = if (grepl("occupancy", filename)) 6 else 5,
    dpi = 300
  )
})
```

```{r}
#| label: fig-death-inclusive-qol-model-sops-ungrouped
#| fig-cap: "Pooled model-derived QoL/death state occupancy probabilities by randomized group"
#| fig-alt: "Posterior state occupancy curves over 26 weeks for seven ordered QoL states and death. Colours distinguish states, line types distinguish randomized groups, and ribbons show 95 percent credible intervals."
#| fig-width: 10
#| fig-height: 6
#| echo: false

ungrouped_sop_plot
```

```{r}
#| label: fig-death-inclusive-qol-model-sops-grouped
#| fig-cap: "Pooled model-derived state occupancy probabilities grouped as Good, Average, Bad, and Death"
#| fig-alt: "Posterior state occupancy curves over 26 weeks for Good quality of life states one to three, Average state four, Bad states five to seven, and Death state eight. Line types identify randomized groups and ribbons show 95 percent credible intervals."
#| fig-width: 10
#| fig-height: 6
#| echo: false

grouped_sop_plot
```

```{r}
#| label: fig-death-inclusive-qol-comparison-posterior
#| fig-cap: "Posterior distribution of the treatment difference in weeks with good-or-better QoL"
#| fig-alt: "A half-eye density above a posterior dot plot of the intervention-minus-control difference in weeks spent in QoL states one through three over 26 weeks. Thick and thin intervals show the 66 and 95 percent credible intervals, and a dashed line marks zero."
#| fig-width: 8
#| fig-height: 5
#| echo: false

comparison_posterior_plot
```
