---
title: CONSORT Flowchart
---

## Setup

```{r}
library(REDCapR)
library(tidyverse)
library(consort)
library(glue)
```

## Download REDCap data


```{r}
data <- REDCapR::redcap_read(redcap_uri = Sys.getenv("redcap_fxdb_url"), token = Sys.getenv("liqplat_api"), raw_or_label = "label")
data <- data$data
```

## Select Screening Columns

```{r}
d <- data |>
  filter(is.na(redcap_repeat_instrument)) |>
  select(
    pat_id_liqplat,
    setting_liqplat,
    solidtu_liqplat,
    elig_pre_solid_tumor       = solidtu_liqplat,     # Prerequisite: Solid Tumor
    elig_4_brain_tumor         = eligcrit_4_liqplat,  # Crit 4: Primary Brain Tumor?
    elig_1_not_resectable      = eligcrit_1_liqplat,  # Crit 1: Tumor not resectable
    elig_2_tx_indicated        = eligcrit_2_liqplat,  # Crit 2: Anti-cancer tx indicated
    elig_5_prior_meta_tx       = eligcrit_5_liqplat,  # Crit 5: Prior tx for metastatic disease
    elig_3_consent_signed      = eligcrit_3_liqplat,  # Crit 3: General consent signed
    eligible_liqplat,
    lung_liqplat,
    radiother_liqplat,
    adt_liqplat,
    adt_date_liqplat,
    othertrial_liqplat,
    randomselect_liqplat,
    select_date_liqplat,
    screenclose_liqplat___1,
    treatstartcheck_liqplat,
    treatstartdate_liqplat,
    treatmentstart_ctdna_date,
    diagnosischeck_liqplat,
    ctdnadatechecked_1_liqplat,
    secop,
    screening_limit,              # Indicates weekly enrollment limit reached
    screening_complete,
    icf_reg                       # Informed consent for SAT acceptance
  )
```

## Hierarchy for Screening

Exclusion criteria (applied in order):

1. If consent not explicitly signed → screened out
2. If enrolled in other trial & not randomized → screened out (competing trial)
3. If no solid tumor (explicit "No") → screened out
4. If brain tumor (explicit "Yes") → screened out
5. If tumor resectable (explicit "No" on "not resectable") → screened out
6. If treatment not indicated (explicit "No") → screened out
7. If prior treatment for metastatic disease (explicit "Yes") → screened out
8. **If any eligibility criterion is "Unclear" or NA and not randomized → screened out**
9. **If screening limit reached (weekly cap) and not randomized → screened out**

**Trial Design Note:** The study randomly selected up to 3 eligible patients per week for invitation to the single-arm trial (SAT). Those selected were offered the ctDNA-guided intervention (`randomselect_liqplat == "ctDNA + SOC"`), which they could accept (`icf_reg == "Yes"`) or decline (`icf_reg == "No"`). The comparison groups are:

- **Intervention arm:** All patients selected for SAT invitation (regardless of acceptance)
- **Control arm:** Eligible patients not selected for SAT invitation (`randomselect_liqplat == "SOC"`)

## Apply Screening Hierarchy

```{r}
# Apply screening hierarchy to classify patients
d_consort <- d |>
  mutate(
    # Check for "Unclear" or NA in any eligibility criterion
    has_unclear_criterion = (elig_pre_solid_tumor == "Unclear" | is.na(elig_pre_solid_tumor)) |
                            (elig_4_brain_tumor == "Unclear" | is.na(elig_4_brain_tumor)) |
                            (elig_1_not_resectable == "Unclear" | is.na(elig_1_not_resectable)) |
                            (elig_2_tx_indicated == "Unclear" | is.na(elig_2_tx_indicated)) |
                            (elig_5_prior_meta_tx == "Unclear" | is.na(elig_5_prior_meta_tx)) |
                            (elig_3_consent_signed == "Unclear" | is.na(elig_3_consent_signed)),

    # Apply exclusion hierarchy
    screen_out_reason = case_when(
      # Explicit exclusion criteria
      elig_3_consent_signed == "No consent given" ~ "No consent signed",
      othertrial_liqplat == "Yes" & is.na(randomselect_liqplat) ~ "Enrolled in competing trial",
      elig_pre_solid_tumor == "No" ~ "No solid tumor",
      elig_4_brain_tumor == "Yes" ~ "Primary brain tumor",
      elig_1_not_resectable == "No" ~ "Tumor resectable",
      elig_2_tx_indicated == "No" ~ "Treatment not indicated",
      elig_5_prior_meta_tx == "Yes" ~ "Prior treatment for metastatic disease",

      # Screen out those with unclear eligibility who weren't randomized
      has_unclear_criterion & is.na(randomselect_liqplat) ~ "Unclear eligibility",

      # Screen out those who hit the screening limit
      screening_limit == TRUE & is.na(randomselect_liqplat) ~ "Screening limit reached",

      TRUE ~ NA_character_
    ),
    screened_out = !is.na(screen_out_reason),
    random_selection = !is.na(randomselect_liqplat),

    # Classify final status
    status = case_when(
      screened_out ~ "Screened out",
      random_selection ~ "Randomized",
      TRUE ~ "Eligible but not randomized"
    )
  )

# Create summary counts for CONSORT
consort_summary <- list(
  total_assessed = nrow(d_consort),
  screened_out = sum(d_consort$screened_out),
  eligible = sum(!d_consort$screened_out),
  randomized = sum(d_consort$random_selection),
  eligible_not_randomized = sum(!d_consort$screened_out & !d_consort$random_selection),

  # SAT-specific counts
  invited_to_sat = sum(d_consort$randomselect_liqplat == "ctDNA + SOC", na.rm = TRUE),
  accepted_sat = sum(d_consort$randomselect_liqplat == "ctDNA + SOC" &
                     d_consort$icf_reg == "Yes", na.rm = TRUE),
  declined_sat = sum(d_consort$randomselect_liqplat == "ctDNA + SOC" &
                     d_consort$icf_reg == "No", na.rm = TRUE),
  control_group = sum(d_consort$randomselect_liqplat == "SOC", na.rm = TRUE)
)

# Display summary
cat("CONSORT Flowchart Counts:\n")
cat(strrep("=", 60), "\n")
cat("Total patients assessed:", consort_summary$total_assessed, "\n")
cat("Screened out:", consort_summary$screened_out, "\n")
cat("Eligible:", consort_summary$eligible, "\n")
cat("  - Selected for randomization:", consort_summary$randomized, "\n")
cat("    • Invited to SAT (ctDNA + SOC):", consort_summary$invited_to_sat, "\n")
cat("      - Accepted:", consort_summary$accepted_sat, "\n")
cat("      - Declined:", consort_summary$declined_sat, "\n")
cat("    • Control (SOC):", consort_summary$control_group, "\n")
cat("  - Eligible but not randomized:", consort_summary$eligible_not_randomized, "\n\n")

cat("Screen-out reasons breakdown:\n")
d_consort |>
  filter(screened_out) |>
  count(screen_out_reason, sort = TRUE) |>
  mutate(pct = scales::percent(n / sum(n), accuracy = 0.1))
```

**Note:** There are `r consort_summary$eligible_not_randomized` patients who passed all screening criteria but were not selected for randomization. Among the `r consort_summary$invited_to_sat` patients invited to the SAT, `r consort_summary$accepted_sat` accepted and `r consort_summary$declined_sat` declined the intervention.

## Create CONSORT Flowchart

```{r}
#| label: fig-consort-screening
#| fig-cap: "CONSORT flowchart for screening and randomization phase"
#| fig-width: 10
#| fig-height: 10

# Create detailed exclusion data frame
exclusions <- d_consort |>
  filter(screened_out) |>
  count(screen_out_reason, sort = TRUE) |>
  rename(reason = screen_out_reason, count = n)

# Format exclusion reasons for display with bullet points
exclusion_text <- exclusions |>
  mutate(text = glue("• {reason} (n={count})")) |>
  pull(text) |>
  paste(collapse = "\n")

# Build the CONSORT diagram
consort_diagram <- consort::add_box(
  txt = glue("Patients assessed for eligibility\n(n = {consort_summary$total_assessed})")
) |>
  consort::add_side_box(
    txt = glue("Excluded (n = {consort_summary$screened_out}):\n{exclusion_text}")
  ) |>
  consort::add_box(
    txt = glue("Eligible\n(n = {consort_summary$eligible})")
  ) |>
  consort::add_side_box(
    txt = glue("Eligible, but already 3 patients \nselected for invitation that week\n(n = {consort_summary$eligible_not_randomized})")
  ) |>
  consort::add_box(
    txt = glue("Random Selection\n(n = {consort_summary$randomized})")
  ) |>
  consort::add_split(
    txt = c(
      as.character(glue("Allocated to SAT\n(n = {consort_summary$invited_to_sat})\n
      \nNot invited: {consort_summary$invited_to_sat - consort_summary$accepted_sat - consort_summary$declined_sat}
      \nAccepted: {consort_summary$accepted_sat}
      \nDeclined: {consort_summary$declined_sat}")),
      as.character(glue("Allocated to Usual Care\n(n = {consort_summary$control_group})"))
    )
  )

# Display the diagram
plot(consort_diagram)

# Save the diagram as SVG and PNG
svg(here::here("0-main-analysis", "img", "consort_flowchart.svg"), width = 10, height = 10)
plot(consort_diagram)
dev.off()

png(here::here("0-main-analysis", "img", "consort_flowchart.png"), width = 10, height = 10, units = "in", res = 300)
plot(consort_diagram)
dev.off()
```

## Summary

### Screening Results

- **Total assessed:** `r consort_summary$total_assessed`
- **Excluded:** `r consort_summary$screened_out` (`r scales::percent(consort_summary$screened_out / consort_summary$total_assessed, accuracy = 0.1)`)
- **Eligible:** `r consort_summary$eligible` (`r scales::percent(consort_summary$eligible / consort_summary$total_assessed, accuracy = 0.1)`)

### Randomization Results

Of the `r consort_summary$eligible` eligible patients:

- **Randomized:** `r consort_summary$randomized` (`r scales::percent(consort_summary$randomized / consort_summary$eligible, accuracy = 0.1)`)
  - **SAT arm (invited for ctDNA + SOC):** `r consort_summary$invited_to_sat`
    - Accepted intervention: `r consort_summary$accepted_sat` (`r scales::percent(consort_summary$accepted_sat / consort_summary$invited_to_sat, accuracy = 0.1)`)
    - Declined intervention: `r consort_summary$declined_sat` (`r scales::percent(consort_summary$declined_sat / consort_summary$invited_to_sat, accuracy = 0.1)`)
  - **Control arm (SOC only):** `r consort_summary$control_group`
- **Not selected:** `r consort_summary$eligible_not_randomized`

### Exclusion Reasons

```{r}
#| label: tbl-exclusions
#| tbl-cap: "Reasons for exclusion from trial"

d_consort |>
  filter(screened_out) |>
  count(screen_out_reason, sort = TRUE) |>
  mutate(
    pct = n / sum(n),
    pct_text = scales::percent(pct, accuracy = 0.1)
  ) |>
  select(
    `Exclusion Reason` = screen_out_reason,
    `N` = n,
    `%` = pct_text
  ) |>
  knitr::kable()
```

**Note:** The "Unclear eligibility" category includes patients with incomplete or ambiguous responses to eligibility criteria who were not ultimately randomized. These cases require database review and clarification before final database lock.
