---
title: "Ordinal Model Simulation (No Random Effects) using `ordinal` Package"
---

## 1. Setup

This script adapts the original simulation to use the `ordinal` package for cumulative link models (`clm`), replacing the `VGAM` package.

```{r setup}
library(ordinal)
library(arrow)
library(tidyverse)
library(furrr)
library(glue)
library(fs)
library(here)
```

```{r}
source(here("functions", "jackknife_mcse.R"))
```

## 2. Simulation Function

The `run_simulation_iteration` function is updated to use `ordinal::clm` for model fitting.

```{r sim-function}
run_simulation_iteration <- function(iter_num, full_data, model_formula, seed = 123, sample_size = 250) {
    # This function runs one full simulation iteration.
    
    set.seed(seed + iter_num)

    # Sample patient IDs for treatment and standard of care groups
    all_ids <- unique(full_data$id)
    tx_ids <- sample(all_ids, round(0.5 * sample_size), replace = FALSE)
    soc_ids <- sample(setdiff(all_ids, tx_ids), round(0.5 * sample_size), replace = FALSE)

    data_for_model <- full_data |> 
        filter(id %in% c(tx_ids, soc_ids)) |>
        mutate(
            tx = if_else(id %in% tx_ids, "1", "0"),
            tx = factor(tx, levels = c("0", "1"))
        )
    
    if (nrow(data_for_model) > 0) {
        model <- tryCatch({ 
            # Use clm from the ordinal package
            ordinal::clm(
                formula = model_formula,
                data = data_for_model)
            
        }, error = function(e) {
            message(glue("ERROR in iteration {iter_num}: {e$message}"))
            return(NULL)
        })
        
        # Extract results if the model converged successfully
        if (!is.null(model)) {
            summary_df <- coef(summary(model)) |> 
                data.frame(check.names = TRUE) |> # Use check.names to match original col name format
                rownames_to_column("term")
            return(summary_df)
        }
    }
    return(NULL)
}
```

## 3. Simulation Specifications

```{r sim-specs}
N_SIMULATIONS <- 1000
N_CORES <- 8
future::plan(future.callr::callr, workers = N_CORES)
```

## 4. Run Simulations

This loop iterates through different effect sizes, loads the appropriate data, and runs the simulation using the updated function.

```{r run-sims}
# Define output and input paths
OUTPUT_PATH <- here("9-markov-tests", "output", "ordinal_no_re")
DATA_PATH <- here("9-markov-tests", "data")

effect_sizes <- seq(-1.5, 1.5, by = 0.5)

# Define the main model formula
main_formula <- ordered(y) ~ tx + time + yprev

# Main simulation loop
for (effect in effect_sizes) {
    effect_str <- sprintf("%.1f", effect)
    message(glue("\n--- Running simulation for x_effect = {effect_str} ---\n"))
    
    INPUT_FILE <- fs::path(DATA_PATH, glue("simulation_x_effect_{effect_str}.parquet"))
    RESULTS_FILE <- fs::path(OUTPUT_PATH, glue("results_{effect_str}.parquet"))

    df_full <- arrow::read_parquet(INPUT_FILE) |> 
        filter(time > 0, yprev != 5) |>
        rename(id = patient_id, y = state) |> 
        mutate(
            y = factor(y, ordered = TRUE, levels = 1:5),
            yprev = factor(yprev, ordered = FALSE)
        )
    
    # Run simulations in parallel
    results <- furrr::future_map_dfr(
        .x = 1:N_SIMULATIONS,
        .f = ~ run_simulation_iteration(
            iter_num = .x,
            full_data = df_full,
            model_formula = main_formula,
            sample_size = 250,
            seed = 123
        ),
        .options = furrr_options(seed = TRUE),
        .progress = TRUE,
        .id = "iter"
    )

    arrow::write_parquet(results, RESULTS_FILE)
}
```

## 5. Summarise Simulation Results

```{r summarise-results}
result_files <- fs::dir_ls(OUTPUT_PATH, glob = "*.parquet")

process_simulation_results <- function(file_path) {
    filename <- fs::path_file(file_path)
    x_effect_strength <- as.numeric(stringr::str_extract(filename, "-?\\d+\\.\\d+"))
    
    sim_data <- arrow::read_parquet(file_path) |>
        filter(term == "tx1")
    
    if(nrow(sim_data) == 0) return(NULL)
    
    # The column Pr(>|z|) is converted to Pr...z.. by data.frame()
    type_i_indicators <- sim_data$Pr...z.. < 0.05
    type_i_error_rate <- mean(type_i_indicators, na.rm = TRUE)
    type_i_error_mcse <- jackknife_mcse(estimates = type_i_indicators)
    
    estimates <- sim_data$Estimate
    bias <- mean(estimates, na.rm = TRUE)
    bias_mcse <- jackknife_mcse(estimates = estimates)

    tibble::tibble(
        `X Effect Strength` = x_effect_strength,
        `Sample Size` = 250,
        `Type I Error (MCSE)` = sprintf("%.3f (%.3f)", type_i_error_rate, type_i_error_mcse),
        `Bias (MCSE)` = sprintf("%.3f (%.3f)", bias, bias_mcse),
        `Simulations` = nrow(sim_data)
    )
}

summary_table <- purrr::map_dfr(result_files, process_simulation_results) |>
    arrange(`X Effect Strength`)

knitr::kable(
    summary_table,
    caption = "Ordinal Model (clm): Comparison of Type I Error and Bias for Varying Strengths of an Omitted Covariate (ordinal::clm)"
)
