---
title: "Ordinal Model Simulation (With Random Effects) using `ordinal` Package"
---

## 1. Setup

This script introduces a random intercept at the patient level using the `ordinal::clmm` function.

```{r setup}
library(ordinal)
library(arrow)
library(tidyverse)
library(furrr)
library(glue)
library(fs)
library(here)
```

```{r}
source(here("functions", "jackknife_mcse.R"))
```

## 2. Simulation Function

The simulation function now uses `ordinal::clmm` and includes a random intercept in the model formula.

```{r sim-function-re}
run_simulation_iteration_re <- function(iter_num, full_data, model_formula, seed = 123, sample_size = 250) {
    # This function runs one full simulation iteration with a random effect.
    
    set.seed(seed + iter_num)

    all_ids <- unique(full_data$id)
    tx_ids <- sample(all_ids, round(0.5 * sample_size), replace = FALSE)
    soc_ids <- sample(setdiff(all_ids, tx_ids), round(0.5 * sample_size), replace = FALSE)

    data_for_model <- full_data |> 
        filter(id %in% c(tx_ids, soc_ids)) |>
        mutate(
            tx = if_else(id %in% tx_ids, "1", "0"),
            tx = factor(tx, levels = c("0", "1")),
            id = factor(id) # clmm requires the grouping variable to be a factor
        )
    
    if (nrow(data_for_model) > 0) {
        model <- tryCatch({ 
            # Use clmm for mixed-effects model
            ordinal::clmm(
                formula = model_formula,
                data = data_for_model)
            
        }, error = function(e) {
            message(glue("ERROR in iteration {iter_num}: {e$message}"))
            return(NULL)
        })
        
        if (!is.null(model)) {
            summary_df <- coef(summary(model)) |> 
                data.frame(check.names = TRUE) |>
                rownames_to_column("term")
            return(summary_df)
        }
    }
    return(NULL)
}
```

## 3. Simulation Specifications

```{r sim-specs-re}
N_SIMULATIONS <- 1000
N_CORES <- 8
future::plan(future.callr::callr, workers = N_CORES)
```

## 4. Run Simulations

```{r run-sims-re}
OUTPUT_PATH_RE <- here("9-markov-tests", "output", "ordinal_with_re")
DATA_PATH <- here("9-markov-tests", "data")

effect_sizes <- seq(-1, 1, by = 0.5)
effect_sizes <- effect_sizes[-3] #removing 0 because this causes convergence issues (random intercept isn't absorbing anything)

# Define the model formula with a random intercept for 'id'
main_formula_re <- ordered(y) ~ tx + time + yprev + (1|id)

for (effect in effect_sizes) {
    effect_str <- sprintf("%.1f", effect)
    message(glue("\n--- Running RE simulation for x_effect = {effect_str} ---\n"))
    
    INPUT_FILE <- fs::path(DATA_PATH, glue("simulation_x_effect_{effect_str}.parquet"))
    RESULTS_FILE_RE <- fs::path(OUTPUT_PATH_RE, glue("results_{effect_str}.parquet"))

    df_full <- arrow::read_parquet(INPUT_FILE) |> 
        filter(time > 0, yprev != 5) |>
        rename(id = patient_id, y = state) |> 
        mutate(
            y = factor(y, ordered = TRUE, levels = 1:5),
            yprev = factor(yprev, ordered = FALSE)
        )
    
    results <- furrr::future_map_dfr(
        .x = 1:N_SIMULATIONS,
        .f = ~ run_simulation_iteration_re(
            iter_num = .x,
            full_data = df_full,
            model_formula = main_formula_re,
            sample_size = 250,
            seed = 123
        ),
        .options = furrr_options(seed = TRUE),
        .progress = TRUE,
        .id = "iter"
    )

    arrow::write_parquet(results, RESULTS_FILE_RE)
}
```

```{r}
# m <- clmm(ordered(y) ~ tx + time + yprev + (1|id), data = df_full |> filter(id < 250) |> mutate(id = factor(id)))
# m2 <- clm(ordered(y) ~ tx + time + yprev, data = df_full |> filter(id < 250))
```

## 5. Summarise Simulation Results

Cave: I'm removing iterations that likely had convergence issues (z-value missing or >10). The estimates for tx from iterations with these very large z-values were very close to 0, but the SEs were estimated close to 0.

The random intercepts do seem to absorb a lot of the unkown outcome hetergeneity, leading to lower type I error rates.

```{r summarise-results-re}
result_files <- fs::dir_ls(OUTPUT_PATH_RE, glob = "*.parquet")

# This function is identical to the non-RE version but is included for completeness.
process_simulation_results_re <- function(file_path) {
    filename <- fs::path_file(file_path)
    x_effect_strength <- as.numeric(stringr::str_extract(filename, "-?\\d+\\.\\d+"))
    
    sim_data <- arrow::read_parquet(file_path) |>
        filter(term == "tx1") |> 
        filter(!is.na(Pr...z..)) |> 
        filter(z.value < 10) # Remove extreme z-values that likely indicate convergence issues
        
    if(nrow(sim_data) == 0) return(NULL)

    type_i_indicators <- sim_data$Pr...z.. < 0.05
    type_i_error_rate <- mean(type_i_indicators, na.rm = TRUE)
    type_i_error_mcse <- jackknife_mcse(estimates = type_i_indicators)
    
    estimates <- sim_data$Estimate
    bias <- mean(estimates, na.rm = TRUE)
    bias_mcse <- jackknife_mcse(estimates = estimates)

    tibble::tibble(
        `X Effect Strength` = x_effect_strength,
        `Sample Size` = 250,
        `Type I Error (MCSE)` = sprintf("%.3f (%.3f)", type_i_error_rate, type_i_error_mcse),
        `Bias (MCSE)` = sprintf("%.3f (%.3f)", bias, bias_mcse),
        `Simulations` = nrow(sim_data)
    )
}

summary_table_re <- purrr::map_dfr(result_files, process_simulation_results_re) |>
    arrange(`X Effect Strength`)

knitr::kable(
    summary_table_re,
    caption = "Ordinal Model (clmm with Random Intercept): Comparison of Type I Error and Bias"
)

create_bias_table <- function(file_path) {
    filename <- fs::path_file(file_path)
    
    # Extract the omitted covariate's effect strength from the filename
    x_effect_strength <- as.numeric(stringr::str_extract(filename, "-?\\d+\\.\\d+"))
    
    sim_data <- arrow::read_parquet(file_path) |> 
        filter(!is.na(Pr...z..))
    
    # Calculate Bias and its MCSE (true effect is 0)
    estimates <- sim_data |> 
        pivot_wider(
            names_from = term, 
            values_from = Estimate)
        

    bias_time <- mean(estimates$time, na.rm = TRUE) - log(0.95)
    bias_yprev2 <- mean(estimates$`yprev2`, na.rm = TRUE) - log(10)
    bias_yprev3 <- mean(estimates$`yprev3`, na.rm = TRUE) - log(3)
    bias_yprev4 <- mean(estimates$`yprev4`, na.rm = TRUE) - log(40)
    bias_tx <- mean(estimates$`tx1`, na.rm = TRUE) - log(1)

    bias_time_mcse <- jackknife_mcse(estimates = na.omit(estimates$time))
    bias_yprev2_mcse <- jackknife_mcse(estimates = na.omit(estimates$`yprev2`))
    bias_yprev3_mcse <- jackknife_mcse(estimates = na.omit(estimates$`yprev3`))
    bias_yprev4_mcse <- jackknife_mcse(estimates = na.omit(estimates$`yprev4`))
    bias_tx_mcse <- jackknife_mcse(estimates = na.omit(estimates$`tx1`))

    # Return a tibble with formatted columns
    tibble::tibble(
        `X Effect Strength` = x_effect_strength,
        `Sample Size` = 250,
        `'tx' Bias (MCSE)` = sprintf("%.3f (%.3f)", bias_tx, bias_tx_mcse),
        `'time' Bias (MCSE)` = sprintf("%.3f (%.3f)", bias_time, bias_time_mcse),
        `'yprev=2' Bias (MCSE)` = sprintf("%.3f (%.3f)", bias_yprev2, bias_yprev2_mcse),
        `'yprev=3' Bias (MCSE)` = sprintf("%.3f (%.3f)", bias_yprev3, bias_yprev3_mcse),
        `'yprev=4' Bias (MCSE)` = sprintf("%.3f (%.3f)", bias_yprev2, bias_yprev4_mcse),
        `Simulations` = nrow(sim_data)/9
    )
}

bias_table <- purrr::map_dfr(result_files, create_bias_table) |>
    arrange(`X Effect Strength`)

knitr::kable(
    bias_table,
    caption = "Bias of all parameters for Varying Strengths of an Omitted Covariate (ordinal::clmm)"
)
