---
Title: Calculating unconditional SOPs from second-order Markov model
---

```{r}
library(tidyverse)
library(data.table)
library(Hmisc)
library(zoo)
library(arrow)
library(splines)
library(brms)
library(rms)
library(rmsb)
```

## Fit model to use

```{r}
df_ypprev <- read_parquet("./data/hosp_long_weeks_two_lags.parquet") |> 
    mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE),
        yprev  = factor(yprev,  levels = 1:4, ordered = FALSE),
        ypprev = factor(ypprev, levels = 1:4, ordered = FALSE),
        y = ordered(y)
    ) |> 
      select(pat_id, tx, y, yprev, ypprev, week, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)

d <- df_ypprev
REFIT <- FALSE
```

```{r}
fit_second_order <- orm(
  y ~ tx + rcs(week, 4) + yprev + ypprev + week %ia% yprev + week %ia% ypprev + ecog_fstcnt + albumin,
  data = d
)
model_path_cluster <- "./2-hosp/models/fit_second_order_b_cluster.rds"
model_path_nocuster <- "./2-hosp/models/fit_second_order_b.rds"
model_path_cluster_no_int <- "./2-hosp/models/fit_second_order_b_cluster_no_int.rds"

if (REFIT) {
  Sys.setenv(RSTUDIO = 1)
    
    dd <- datadist(d)
    options(datadist = 'dd')
    options(mc.cores = 4)
    
  fit_second_order_b_cluster <- blrm(
    y ~ tx + rcs(week, 4) + yprev + ypprev + week %ia% yprev + week %ia% ypprev + ecog_fstcnt + albumin + cluster(pat_id),
    data = d,
    refresh = 5,
    iter = 2000,
    chains = 4,
    seed = 1234,
    method = "sampling",
    backend = "cmdstan"
  )

  fit_second_order_b_cluster_no_int <- blrm(
    y ~ tx + rcs(week, 4) + yprev + ypprev + ecog_fstcnt + albumin + cluster(pat_id),
    data = d,
    refresh = 5,
    iter = 2000,
    chains = 4,
    seed = 1234,
    method = "sampling",
    backend = "cmdstan"
  )

  fit_second_order_b <- blrm(
    y ~ tx + rcs(week, 4) + yprev + ypprev + week %ia% yprev + week %ia% ypprev + ecog_fstcnt + albumin,
    data = d,
    refresh = 5,
    iter = 2000,
    chains = 4,
    seed = 1234,
    method = "sampling",
    backend = "cmdstan"
  )
  Sys.setenv(RSTUDIO = 0)
  saveRDS(fit_second_order_b_cluster, file = model_path_cluster)
  saveRDS(fit_second_order_b, file = model_path_nocuster)
  saveRDS(fit_second_order_b_cluster_no_int, file = model_path_cluster_no_int)
} else {
  # Load the models from disk if they exist
  if (file.exists(model_path_cluster)) {
    fit_second_order_b_cluster <- readRDS(model_path_cluster)
  } else {
    warning("Model file not found: ", model_path_cluster)
    fit_second_order_b_cluster <- NULL
  }
 
  if (file.exists(model_path_nocuster)) {
    fit_second_order_b <- readRDS(model_path_nocuster)
  } else {
    warning("Model file not found: ", model_path_nocuster)
    fit_second_order_b <- NULL
  }

  if (file.exists(model_path_cluster_no_int)) {
    fit_second_order_b_cluster_no_int <- readRDS(model_path_cluster_no_int)
  } else {
    warning("Model file not found: ", model_path_cluster_no_int)
    fit_second_order_b_cluster_no_int <- NULL
  }
}

```

# Fit the same model with brms

```{r}
Sys.setenv(RSTUDIO = 1)

priors <- c(
  # Priors for Intercepts/Cutpoints
  prior(normal(1, 0.5), class = "Intercept", coef = 1), # Corresponds to y>=2
  prior(normal(1.5, 0.5), class = "Intercept", coef = 2), # Corresponds to y>=3
  prior(normal(2, 0.5), class = "Intercept", coef = 3), # Corresponds to y>=4
  prior(normal(5, 0.5), class = "Intercept", coef = 4)
)

fit_brms <- brm(
  formula = bf(
    y ~ tx + ns(week, df = 4) + yprev + ypprev +
        ecog_fstcnt + albumin + (1 | pat_id)
  ),
  data = d,
  family = cumulative("logit"),
  prior = priors,
  iter = 2000,
  chains = 4,
  cores = 4,
  seed = 1234,
  refresh = 5,
  backend = "cmdstanr",
  file = "./2-hosp/models/fit_second_order_brms_cluster.rds",
  file_refit = "on_change"
)
Sys.setenv(RSTUDIO = 0)

marginaleffects::predictions(fit_brms, newdata = d[10,], type = "response", re_formula = NULL)
```

## Predictions conditional on random effects

In think we need to include the random intercept to correctly perform g-computation. Therefore using the custom function here.
I think that at the moment this 

```{r}
library(posterior)

fit <- fit_second_order_b_cluster
sf  <- rmsb::stanGet(fit)

# 1) collect all parameter names and locate gamma[j] columns
all_pars <- colnames(as_draws_matrix(sf$draws()))
gcols    <- grep("^gamma\\[", all_pars, value = TRUE)

# 2) stack all chains and build an n_draws x J matrix of random intercepts
G <- as_draws_matrix(sf$draws(variables = gcols))  # rows = draws, cols = patients

# 3) per-patient summaries
per_pat_median <- apply(G, 2, median)

all(per_pat_median == fit$gammas) #Is TRUE !

newdata_row <- d[1,]
```


```{r}
predict_blrm_manual <- function(fit, newdata, include_re = FALSE) {
  # --- 0) Prerequisite Checks ---
  if (!is.data.frame(newdata)) {
    stop("`newdata` must be a data frame.")
  }
  if (include_re && !"pat_id" %in% colnames(newdata)) {
    stop("`newdata` must contain a 'pat_id' column when include_re=TRUE")
  }
  
  N_new <- nrow(newdata)

  # --- 1) Get the design matrix for the new data ---
  X <- rms::predictrms(fit, newdata = newdata, type = 'x')

  # --- 2) Extract posterior draws and model information ---
  draws <- fit$draws
  ndraws <- nrow(draws)
  ns <- fit$non.slopes
  K <- ns + 1
  
  intercept_draws <- draws[, 1:ns, drop = FALSE]
  beta_draws <- draws[, (ns + 1):ncol(draws), drop = FALSE]

  # --- 3) Calculate the fixed-effect linear predictor (eta_fixed) ---
  eta_fixed <- beta_draws %*% t(X)

  # --- 4) Get the random effect draws (u_draws) if requested ---
  u_draws <- matrix(0, nrow = ndraws, ncol = N_new)
  
  if (include_re) {
    if (length(unique(newdata$pat_id)) > 1) {
      stop("`newdata` contains multiple unique patient IDs. This version is restricted to one at a time.")
    }
    if (!requireNamespace("posterior", quietly = TRUE)) {
      stop("The 'posterior' package is required to extract random effect draws.")
    }
    sf <- rmsb::stanGet(fit)
    all_gamma_draws <- posterior::as_draws_matrix(sf$draws(variables = "gamma"))
    patient_ids <- newdata$pat_id
    u_draws <- all_gamma_draws[, patient_ids, drop = FALSE]
  }

  # --- 5) Calculate the total linear predictor (eta) ---
  stopifnot(all(dim(eta_fixed) == dim(u_draws)))
  eta <- eta_fixed + u_draws

  # --- 6) Calculate probabilities for each observation and each draw ---
  p_cat_draws_list <- vector("list", N_new) 
  
  for (i in 1:N_new) {
    eta_i <- eta[, i]
    s_ge <- plogis(sweep(intercept_draws, 1, eta_i, "+"))
    
    p_cat_draws <- matrix(NA_real_, nrow = ndraws, ncol = K)
    p_cat_draws[, 1] <- 1 - s_ge[, 1]
    if (K > 2) {
      for (kk in 2:(K - 1)) {
        p_cat_draws[, kk] <- s_ge[, kk - 1] - s_ge[, kk]
      }
    }
    p_cat_draws[, K] <- s_ge[, K - 1]
    colnames(p_cat_draws) <- paste0("y=", 1:K)
    
    p_cat_draws_list[[i]] <- p_cat_draws
  }

  # --- 7) NEW: Reshape draws into a single 3D array ---
  draws_array <- array(
    data = NA_real_,
    dim = c(ndraws, N_new, K)
  )

  for (i in 1:N_new) {
    draws_array[, i, ] <- p_cat_draws_list[[i]]
  }

  dimnames(draws_array) <- list(
    NULL,
    rownames(newdata),
    colnames(p_cat_draws_list[[1]])
  )

  return(draws_array)
}
```


Check whether prediction when ignoring random effects match the predict function in rmsb:

```{r}
# 1. Get the prediction from your manual function
pred_manual <- predict_blrm_manual(
  fit_second_order_b_cluster, 
  newdata_row, 
  include_re = FALSE
)

# 2. Get the prediction from the built-in function
pred_builtin <- predict(
  fit_second_order_b_cluster, 
  newdata = newdata_row, 
  type = "fitted.ind", 
  posterior.summary = "all"
)

# 3. First, check if the dimensions are identical
dim(pred_manual)
dim(pred_builtin)

# 4. Then, check if the numeric values are nearly equal
all.equal(pred_manual, pred_builtin)
```

## Functions

The `predict()` function for `orm()` and `blrm()` return a set of cumulative probabilities, such as $P(Y\geq1)$, $P(Y\geq2)$, etc. To get the actual probability for each individual state, $P(Y=k)$, we need a helper function. The `convert_to_states` function below transforms the model's output into a standard state probability vector.

```{r}
convert_to_states <- function(cumulative_probs) {
    if (!is.matrix(cumulative_probs)) {
        cumulative_probs <- t(as.matrix(cumulative_probs))
    }
    
    n_rows <- nrow(cumulative_probs)
    n_states <- ncol(cumulative_probs) + 1
    
    # Initialize the matrix for state probabilities
    state_probs <- matrix(NA, nrow = n_rows, ncol = n_states)
    
    # P(y=1)
    state_probs[, 1] <- 1 - cumulative_probs[, 1]
    
    # P(y=2) to P(y=4)
    for (j in 2:(n_states - 1)) {
        state_probs[, j] <- cumulative_probs[, j - 1] - cumulative_probs[, j]
    }
    
    # P(y=5)
    state_probs[, n_states] <- cumulative_probs[, n_states - 1]
    
    return(state_probs)
}
```

## Theory

The model gives us the conditional probability of being in a state l at time t, given the states at the two previous time points: 
$$
P(X_t=l \mid X_{t-1}=j,X_{t-2}=h)
$$

Our goal is to calculate the unconditional state occupancy probability, $P(X_t=l)$, which represents the overall probability of being in state l at time t, regardless of the specific path taken to get there (but still conditional on the two states before randomization). We can find the unconditional probability of an event by summing the probabilities of all possible ways it can happen (Total Law of Probability). In our case, this means summing over all possible histories (h, j) from the previous two time steps:

$$P(X_t = l) = \sum_{h=1}^{M} \sum_{j=1}^{M} \underbrace{P(X_t=l \ | \ X_{t-1}=j, X_{t-2}=h)}_{\text{The Model's "Prediction"}} \times \underbrace{P(X_{t-1}=j, X_{t-2}=h)}_{\text{The "Weight" of the History}}$$

How do we find the "weight," which is the joint probability of a specific history occurring? We can't get the joint probability $P(X_{t-1}=j, X_{t-2}=h)$ directly from the model. Instead, we have to calculate it iteratively. This is possible thanks to the Chain Rule of Probability: $$P(A, B, C) = P(C | B, A) \times P(B, A)$$

Translating this to our problem, the probability of a three-step path is: $$P(X_t=l, X_{t-1}=j, X_{t-2}=h) = P(X_t=l \ | \ X_{t-1}=j, X_{t-2}=h) \times P(X_{t-1}=j, X_{t-2}=h)$$

By summing over $X_{t-2} = h$, we calculate the joint probability of the next state $X_t = l$ and lag = 1 $X_{t-1} = j$:

$$P(X_t=l, X_{t-1}=j) = \sum_{h=1}^{M} P(X_t=l \ | \ X_{t-1}=j, X_{t-2}=h) \times P(X_{t-1}=j, X_{t-2}=h)$$

1.  We start with a known history at t=0, so the joint probability of the observed $j_{obs}$ and $h_{obs}$ and states for $X_{t-1}$ and $X_{t-2}$ is $1.0$

2.  We use this known probability as the "weight" to calculate the joint probabilities for t=1.

    -   $P(X_{t=1} = l, X_{t-1} = j\_{obs}) = P(X_{t=1}  |  X_{t-1}=j_{obs}, X_{t-2}=h_{obs})$
    -   So for $t=1$ we have $M$ non-zero joint probabilities, one for each possible state at time $t=1$, because $j$ is fixed to the observed state at $t=0$.

3.  We then use the joint probabilities from $t = 1$ as the weights to calculate the joint probabilities for $t=2$.

    -   $P(X_{t=2} = l, X_{t-1} = j) = \sum_{h=1}^{M} P(X_{t=2}  |  X_{t-1}=j, X_{t-2}=h) \times P(X_{t-1}=j, X_{t-2}=h)$
    -   Now for $t=2$ we have $M \times M$ joint probabilities, one for each combination of possible states at times $t=1$ and $t=2$.

4.  ...and so on. The output of one step becomes the input for the next.

## Calculate Unconditional Second-Order State Occupation Probabilities

Now we implement this recursive logic in R.

### Setup and Initial Conditions

First, we define our parameters and create two data structures:

-   P_joint: A 3D array to store the joint probability "weights" $P(X_{t-1}, X_t)$ at each time step.
-   P_uncond: A matrix to store our final answer, the unconditional probabilities $P(X_t)$ for each state over time.

We initialize the process with a known starting history for a specific individual. This means at $t=0$, the joint probability of that specific history is $1$, and all others are $0$.


```{r}
# --- Parameters ---
M <- 5              # Number of states
T_horizon <- 26     # How far into the future we want to predict
absorbing_states <- c(5) # Define State 5 as absorbing

baseline_data <- d[1,]

# --- Initial Conditions ---
initial_state_t_minus_1 <- as.numeric(baseline_data[, "ypprev"])
initial_state_t_0 <- as.numeric(baseline_data[, "yprev"])


# --- Data Structures to Store Results ---
# Joint probability array: P(X_{t-1}, X_t)
P_joint <- array(0, dim = c(M, M, T_horizon + 1),
                 dimnames = list(paste0("t-1=", 1:M),
                                 paste0("t=", 1:M),
                                 paste0("time=", 0:T_horizon)))

# Unconditional probability matrix: P(X_t)
P_uncond <- matrix(0, nrow = M, ncol = T_horizon + 1,
                   dimnames = list(paste0("State_", 1:M),
                                   paste0("t=", 0:T_horizon)))

# Set the known probability at t=0
P_joint[initial_state_t_minus_1, initial_state_t_0, 1] <- 1.0
P_uncond[initial_state_t_0, 1] <- 1.0
```


The main loop below executes the chain reaction. At each time step `t`, it iterates through all possible histories `(h, j)` from the two preceding steps.

A crucial feature is the handling of absorbing states (e.g., death or discharge). Before getting a prediction, we check if the current state `(j)` is an absorbing one. If it is, we manually set the probability of staying in that state to 1.

A Note on "Impossible" Paths: This logic also handles impossible paths. For example, a path segment `... -> 5 -> 4` is impossible if State 5 is absorbing. Our code correctly identifies this because the history_weight for this path will be zero. The absorbing state rule applied at the previous step would have set the probability of the `5 -> 4` transition to zero. The line `if (history_weight == 0) { next }` prunes these impossible branches from the calculation, so the initially assigned probability of $0$ remains unchanged.

### Looping through time steps

```{r}
t1 <- Sys.time()
# Loop through each future time step, from t=1 to our horizon
for (t in 1:T_horizon) {
  
  # Get the joint probabilities from the PREVIOUS step
  prev_P_joint <- P_joint[,, t]
  
  # A temporary matrix for the new joint probabilities
  current_P_joint <- matrix(0, nrow = M, ncol = M)
  
  # Loop over all possible states at t-2 (h) and t-1 (j)
  for (h in 1:M) {
    for (j in 1:M) {
      
      history_weight <- prev_P_joint[h, j]
      if (history_weight == 0) {
        next
      }
      
      # 1. Initialize a placeholder vector for the transition probabilities.
      conditional_probs <- rep(0, M)

      # 2. Check if the state at t-1 (j) is absorbing.
      if (j %in% absorbing_states) {
        # 3a. If so, apply the deterministic rule.
        conditional_probs[j] <- 1.0 
      } else {
        # 3b. If not, call the model to get the stochastic prediction.
        conditional_probs <- predict(
          fit_second_order, 
          newdata = list(
              tx = baseline_data$tx,
              week = t,
              yprev = factor(j, levels = c(1:M)[-absorbing_states], ordered = FALSE),
              ypprev = factor(h, levels = c(1:M)[-absorbing_states], ordered = FALSE),
              ecog_fstcnt = baseline_data$ecog_fstcnt,
              albumin = baseline_data$albumin),
          type = "fitted.ind")
      }
      
      # Loop over all possible outcomes at time t (l)
      for (l in 1:M) {        
        path_prob <- conditional_probs[l] * history_weight
        # Add this path's probability to the new joint probability
        current_P_joint[j, l] <- current_P_joint[j, l] + path_prob
      }
    }
  }

  P_joint[,, t + 1] <- current_P_joint
  P_uncond[, t + 1] <- colSums(current_P_joint)
}
t2 <- Sys.time()
t2 - t1

P_uncond_slow <- P_uncond
P_joint_slow <- P_joint
```

### Optimized Code

Now we let Gemini Pro 2.5 have a go at optimizing this code:

```{r}
t1 <- Sys.time()
# Create a static grid of all possible histories (h, j)
history_grid <- expand.grid(
  ypprev = factor(1:M, levels = 1:M),
  yprev = factor(1:M, levels = 1:M)
)

# Add the static baseline covariates to the grid
prediction_data <- cbind(
  baseline_data[, c("tx", "ecog_fstcnt", "albumin")],
  history_grid
)

# Loop through each future time step
for (t in 1:T_horizon) {
  
  # Create a placeholder matrix, initialized to zero.
  # This correctly handles impossible paths (e.g., 5->4) whose
  # probabilities should be 0 but aren't explicitly calculated.
  cond_probs_matrix <- matrix(0, nrow = M * M, ncol = M)
  
  # A history is "predictable" ONLY IF BOTH t-2 and t-1 are NOT absorbing states.
  predictable_idx <- which(
      !(history_grid$ypprev %in% absorbing_states) & 
      !(history_grid$yprev %in% absorbing_states)
  )
  
  # A history is "determined" if the state at t-1 is absorbing.
  absorbing_idx <- which(history_grid$yprev %in% absorbing_states)
  
  # Manually set probabilities for absorbing states: P(j -> j) = 1
  if (length(absorbing_idx) > 0) {
    absorbing_state <- as.numeric(as.character(history_grid$yprev[absorbing_idx]))
    rows_and_cols <- cbind(absorbing_idx, absorbing_state)
    # The matrix was initialized to 0, so we just need to place the 1s
    cond_probs_matrix[rows_and_cols] <- 1.0
  }
  
  # Run the model ONLY on the subset of histories that are truly predictable.
  if (length(predictable_idx) > 0) {
    prediction_data_subset <- prediction_data[predictable_idx, ]
    prediction_data_subset$week <- t # Set the current time
    
    cond_probs_batch <- predict(fit_second_order, newdata = prediction_data_subset, type = "fitted.ind")
    
    # Place the results from the model into the correct rows of our main matrix
    cond_probs_matrix[predictable_idx, ] <- cond_probs_batch
  }
  
  cond_probs_array <- array(
    cond_probs_matrix,
    dim = c(M, M, M),
    dimnames = list(h = 1:M, j = 1:M, l = 1:M)
  )
  
  prev_P_joint <- P_joint[,, t]
  
  weighted_probs <- cond_probs_array * array(prev_P_joint, dim = dim(cond_probs_array))
  
  current_P_joint <- apply(weighted_probs, c(2, 3), sum)
  
  P_joint[,, t + 1] <- current_P_joint
  P_uncond[, t + 1] <- colSums(current_P_joint)
}
t2 <- Sys.time()

t2 - t1 # 10 times faster!
```

```{r}
# The `all.equal` function checks for near-equality, which is important
# for floating-point arithmetic.
# `stopifnot` will throw an error if the comparison is FALSE.
stopifnot(
  all.equal(P_joint_slow, P_joint, check.attributes = FALSE)
)

stopifnot(
  all.equal(P_uncond_slow, P_uncond, check.attributes = FALSE)
)

print("The optimized code produces identical results.")
```

The optimized code from Gemini is 10 times faster and produces identical results

## Code for RMSB package

We want to extend this to work with the Bayesian model fit using the `rmsb` package. The main difference is that we now have multiple posterior draws. We will calculate the unconditional probabilities for each draw separately, resulting in a 3D array where one dimension corresponds to the different draws.

```{r}
# --- Parameters ---
# Use the number of posterior draws from the fitted model
n_draws <- nrow(fit_second_order_b$draws)

# --- Initial Conditions ---
initial_state_t_minus_1 <- as.numeric(baseline_data[, "ypprev"])
initial_state_t_0 <- as.numeric(baseline_data[, "yprev"])

# --- Data Structures to Store Results (with an added dimension for draws) ---
# Joint probability array: P(X_{t-1}, X_t) for each draw
P_joint_bayes <- array(0, dim = c(n_draws, M, M, T_horizon + 1),
                       dimnames = list(paste0("draw_", 1:n_draws),
                                       paste0("t-1=", 1:M),
                                       paste0("t=", 1:M),
                                       paste0("time=", 0:T_horizon)))

# Unconditional probability matrix: P(X_t) for each draw
P_uncond_bayes <- array(0, dim = c(n_draws, M, T_horizon + 1),
                        dimnames = list(paste0("draw_", 1:n_draws),
                                        paste0("State_", 1:M),
                                        paste0("t=", 0:T_horizon)))

# Set the known probability at t=0 for all draws
# The starting history is known with certainty (probability = 1.0)
P_joint_bayes[, initial_state_t_minus_1, initial_state_t_0, 1] <- 1.0
P_uncond_bayes[, initial_state_t_0, 1] <- 1.0
```



```{r}
# Create the static grid for batch predictions (same as before)
history_grid <- expand.grid(
  ypprev = factor(1:M, levels = 1:M), # h = X_{t-2}
  yprev = factor(1:M, levels = 1:M)   # h = X_{t-1}
)
prediction_data <- cbind(
  baseline_data[, c("pat_id", "tx", "ecog_fstcnt", "albumin")],
  history_grid
)

for (t in 1:T_horizon) {
  
  # Identify predictable histories (where neither t-1 nor t-2 is absorbing)
  predictable_idx <- which(
      !(history_grid$ypprev %in% absorbing_states) & 
      !(history_grid$yprev %in% absorbing_states)
  )

  n_predictable <- length(predictable_idx)
  
  # Subset the data for prediction
  prediction_data_subset <- prediction_data[predictable_idx, ]
  prediction_data_subset$week <- t
  
  # Predict to get the object containing draws
  draws <- predict_blrm_manual(fit_second_order_b_cluster, prediction_data_subset, include_re = TRUE) 
  
  # dimension 1 = 4000 draws
  # dimension 2 = X_{t-1} and X_{t-2} combinations (rows from prediction_data_subset)
  # dimension 3 = cumulative probabilities (4 columns for 5 states)
  state_probs_all_draws <- do.call(rbind, lapply(1:dim(draws)[1], function(d) draws[d,,]))
  # if we have 16 prior scenarios, every 16 rows is one draw, one for each scenario
  
  # ASSEMBLE TRANSITION PROBABILITIES FOR EACH DRAW
  # --------------------------------------------------
  
  # Create a 4D array to hold all transition probabilities: P(X_t=l | X_{t-2}=h, X_{t-1}=j)
  # Dims: (draw, h, j, l)
  # So [, 1, 1, 1] give all draws for P(l = 1 | h = 1, j = 1)
  # [, 1, 1, ] for all draws for P(l | h = 1, j = 1) (rows must sum to 1)
  cond_probs_4d_array <- array(0, dim = c(n_draws, M, M, M))
  
  # Loop through each draw to populate the array
  for (d in 1:n_draws) {
    # Create a placeholder matrix for this specific draw
    # each row is combination of (h, j), each column is l (rows must sum to 1)
    # this is for one draw only
    cond_probs_matrix_d <- matrix(0, nrow = M*M, ncol = M)
    
    # Handle absorbing states (P(j->j) = 1 if j is absorbing)
    absorbing_idx <- which(history_grid$yprev %in% absorbing_states)
    if (length(absorbing_idx) > 0) {
      absorbing_state <- as.numeric(as.character(history_grid$yprev[absorbing_idx]))
      rows_and_cols <- cbind(absorbing_idx, absorbing_state)
      cond_probs_matrix_d[rows_and_cols] <- 1.0
    }
    
    # Get the state probabilities for this draw from the big matrix
    # e.g., if 16 predictable histories, then we select rows 1:16 for draw 16
    start_row <- (d - 1) * n_predictable + 1
    end_row <- d * n_predictable
    draw_d_probs <- state_probs_all_draws[start_row:end_row, ]
    
    # Place them into the correct rows
    cond_probs_matrix_d[predictable_idx, ] <- draw_d_probs
    
    # Reshape and store in the final 4D array
    cond_probs_4d_array[d, , , ] <- array(cond_probs_matrix_d, dim = c(M, M, M))
  }
  
  # Get joint probabilities from the previous step for all draws: P(X_{t-2}, X_{t-1})
  prev_P_joint_all_draws <- P_joint_bayes[,,, t]
  current_P_joint_all_draws <- array(0, dim = c(n_draws, M, M))
  
  # Loop through draws
  for (d in 1:n_draws) {
    # This is P(X_{t-2}=h, X_{t-1}=j) for draw d
    prev_P_joint_d <- prev_P_joint_all_draws[d,,]
    # P(X_t = l | X_{t-1} = j, X_{t-2} = h) for this draw
    cond_probs_d <- cond_probs_4d_array[d,,,]
    
    # This is the multiplication inside the summation: P(X_t=l|h,j) * P(h,j)
    weighted_probs <- cond_probs_d * array(prev_P_joint_d, dim = dim(cond_probs_d))
    # This is the summation over h (the state at t-2), resulting in P(X_{t-1}=j, X_t=l)
    current_P_joint_d <- apply(weighted_probs, c(2, 3), sum)
    
    # Store results for this draw
    current_P_joint_all_draws[d,,] <- current_P_joint_d
    P_uncond_bayes[d, , t + 1] <- colSums(current_P_joint_d)
  }
  
  # Store the full joint probability array for the next time step
  P_joint_bayes[,,, t + 1] <- current_P_joint_all_draws
}
```

```{r}
# SOPs for day 26 conditional on baseline state only
P_uncond_bayes[, , 26]
```

## Turning this into a function

```{r}
#' Calculate Second-Order State Occupation Probabilities (Generalized)
#'
#' This function calculates the unconditional state occupation probabilities for a
#' second-order Markov model. It is generalized to work with any set of predictors
#' provided in the baseline data.
#'
#' @param model_fit The fitted model object (from rmsb), which must contain
#'   a `draws` component.
#' @param baseline_data A single-row data frame containing the initial states and
#'   all baseline predictor values required by the model.
#' @param col_yprev A string with the column name for the state at t-1 (e.g., "yprev").
#' @param col_ypprev A string with the column name for the state at t-2 (e.g., "ypprev").
#' @param col_time A string with the column name that represents time in the model (e.g., "week").
#' @param time_horizon A numeric vector specifying the time points for calculation (e.g., `1:26`).
#' @param n_states An integer for the number of levels in the ordinal outcome.
#' @param absorbing_states A numeric vector listing the absorbing states.
#'
#' @return A list containing two arrays:
#'   \item{unconditional}{A 3D array `(draws, states, time)` with the unconditional state probabilities.}
#' @export

sop_markov_second_order <- function(model_fit,
                                    baseline_data,
                                    col_yprev,
                                    col_ypprev,
                                    col_time,
                                    time_horizon,
                                    n_states,
                                    absorbing_states,
                                    include_re = FALSE) {

  # --- Parameters ---
  n_draws <- nrow(model_fit$draws)
  M <- n_states
  T_horizon <- max(time_horizon)

  # --- Initial Conditions ---
  initial_state_t_minus_1 <- as.numeric(baseline_data[[col_ypprev]])
  initial_state_t_0 <- as.numeric(baseline_data[[col_yprev]])

  # --- Data Structures to Store Results ---
  P_joint_bayes <- array(0, dim = c(n_draws, M, M, T_horizon + 1),
                         dimnames = list(paste0("draw_", 1:n_draws),
                                         paste0("t-1=", 1:M),
                                         paste0("t=", 1:M),
                                         paste0("time=", 0:T_horizon)))

  P_uncond_bayes <- array(0, dim = c(n_draws, M, T_horizon + 1),
                          dimnames = list(paste0("draw_", 1:n_draws),
                                          paste0("State_", 1:M),
                                          paste0("t=", 0:T_horizon)))

  # Set the known probability at t=0
  P_joint_bayes[, initial_state_t_minus_1, initial_state_t_0, 1] <- 1.0
  P_uncond_bayes[, initial_state_t_0, 1] <- 1.0

  # --- Create the static grid for batch predictions ---
  history_grid <- expand.grid(
    h = factor(1:M, levels = 1:M), # X_{t-2}
    j = factor(1:M, levels = 1:M)  # X_{t-1}
  )
  names(history_grid) <- c(col_ypprev, col_yprev)

  # --- Dynamically create the full prediction dataset ---
  # Identify static covariates by excluding time-varying/history columns
  cols_to_exclude <- c(col_yprev, col_ypprev, col_time)
  static_covariate_names <- setdiff(names(baseline_data), cols_to_exclude)
  static_covariates <- baseline_data[, static_covariate_names, drop = FALSE]

  # Combine static covariates with the grid of all possible histories
  prediction_data <- cbind(static_covariates, history_grid)

  # --- Main Iteration Loop ---
  for (t in time_horizon) {
    # Identify predictable histories
    predictable_idx <- which(
      !(prediction_data[[col_ypprev]] %in% absorbing_states) &
      !(prediction_data[[col_yprev]] %in% absorbing_states)
    )
    n_predictable <- length(predictable_idx)

    # Subset the data for prediction and set the current time
    prediction_data_subset <- prediction_data[predictable_idx, ]
    prediction_data_subset[[col_time]] <- t

    # Predict to get the draws.
    draws <- predict_blrm_manual(model_fit, prediction_data_subset, include_re = include_re)
    state_probs_all_draws <- do.call(rbind, lapply(1:dim(draws)[1], function(d) draws[d, , ]))

    # --- Assemble transition probabilities for each draw ---
    cond_probs_4d_array <- array(0, dim = c(n_draws, M, M, M))

    for (d in 1:n_draws) {
      cond_probs_matrix_d <- matrix(0, nrow = M * M, ncol = M)
      
      # Handle absorbing states
      absorbing_idx <- which(prediction_data[[col_yprev]] %in% absorbing_states)
      if (length(absorbing_idx) > 0) {
        absorbing_state <- as.numeric(as.character(prediction_data[[col_yprev]][absorbing_idx]))
        rows_and_cols <- cbind(absorbing_idx, absorbing_state)
        cond_probs_matrix_d[rows_and_cols] <- 1.0
      }
      
      # Get state probabilities for the current draw
      start_row <- (d - 1) * n_predictable + 1
      end_row <- d * n_predictable
      draw_d_probs <- state_probs_all_draws[start_row:end_row, ]
      
      # Place them into the correct rows
      cond_probs_matrix_d[predictable_idx, ] <- draw_d_probs
      
      # Reshape and store in the final 4D array
      cond_probs_4d_array[d, , , ] <- array(cond_probs_matrix_d, dim = c(M, M, M))
    }

    # --- Update Probabilities ---
    prev_P_joint_all_draws <- P_joint_bayes[, , , t]
    current_P_joint_all_draws <- array(0, dim = c(n_draws, M, M))

    for (d in 1:n_draws) {
      prev_P_joint_d <- prev_P_joint_all_draws[d, , ]
      cond_probs_d <- cond_probs_4d_array[d, , , ]
      
      # Law of Total Probability: P(A,B) = P(A|B)P(B)
      weighted_probs <- cond_probs_d * array(prev_P_joint_d, dim = dim(cond_probs_d))
      # Marginalize over the t-2 state (h) to get P(X_{t-1}, X_t)
      current_P_joint_d <- apply(weighted_probs, c(2, 3), sum)
      
      current_P_joint_all_draws[d, , ] <- current_P_joint_d
      
      # Marginalize over the t-1 state (j) to get P(X_t)
      uncond_probs_t <- colSums(current_P_joint_d)
      # Normalize to prevent floating point error accumulation
      P_uncond_bayes[d, , t + 1] <- uncond_probs_t
      # P_uncond_bayes[d, , t + 1] <- uncond_probs_t / sum(uncond_probs_t)
    }
    
    P_joint_bayes[, , , t + 1] <- current_P_joint_all_draws
  }

  return(P_uncond_bayes)
}
```

```{r}
#' Calculate Second-Order State Occupation Probabilities (Generalized)
#'
#' This function calculates the unconditional state occupation probabilities for a
#' second-order Markov model. It is generalized to work with any set of predictors
#' provided in the baseline data.
#'
#' @param model_fit The fitted model object (from rmsb), which must contain
#'   a `draws` component.
#' @param baseline_data A single-row data frame containing the initial states and
#'   all baseline predictor values required by the model.
#' @param col_yprev A string with the column name for the state at t-1 (e.g., "yprev").
#' @param col_ypprev A string with the column name for the state at t-2 (e.g., "ypprev").
#' @param col_time A string with the column name that represents time in the model (e.g., "week").
#' @param time_horizon A numeric vector specifying the time points for calculation (e.g., `1:26`).
#' @param n_states An integer for the number of levels in the ordinal outcome.
#' @param absorbing_states A numeric vector listing the absorbing states.
#'
#' @return A list containing two arrays:
#'   \item{unconditional}{A 3D array `(draws, states, time)` with the unconditional state probabilities.}
#' @export

sop_markov_second_order_2 <- function(model_fit,
                                    baseline_data,
                                    col_yprev,
                                    col_ypprev,
                                    col_time,
                                    time_horizon,
                                    n_states,
                                    absorbing_states,
                                    include_re = FALSE) {

  # --- Auto-detect model type and get number of draws ---
  if (inherits(model_fit, "brmsfit")) {
    detected_type <- "brms"
    if (!requireNamespace("marginaleffects", quietly = TRUE) || !requireNamespace("posterior", quietly = TRUE)) {
      stop("Packages 'marginaleffects' and 'posterior' are required for brms models.")
    }
    n_draws <- posterior::ndraws(model_fit)
  } else if (inherits(model_fit, "blrm")) {
    detected_type <- "rmsb"
    n_draws <- nrow(model_fit$draws)
  } else {
    stop("Unsupported model object class. Please provide a model from 'brms' (class 'brmsfit') or 'rmsb' (class 'blrm').")
  }
  
  # --- Parameters ---
  M <- n_states
  T_horizon <- max(time_horizon)

  # --- Initial Conditions ---
  initial_state_t_minus_1 <- as.numeric(baseline_data[[col_ypprev]])
  initial_state_t_0 <- as.numeric(baseline_data[[col_yprev]])

  # --- Data Structures to Store Results ---
  P_joint_bayes <- array(0, dim = c(n_draws, M, M, T_horizon + 1),
                         dimnames = list(paste0("draw_", 1:n_draws),
                                         paste0("t-1=", 1:M),
                                         paste0("t=", 1:M),
                                         paste0("time=", 0:T_horizon)))

  P_uncond_bayes <- array(0, dim = c(n_draws, M, T_horizon + 1),
                          dimnames = list(paste0("draw_", 1:n_draws),
                                          paste0("State_", 1:M),
                                          paste0("t=", 0:T_horizon)))

  # Set the known probability at t=0
  P_joint_bayes[, initial_state_t_minus_1, initial_state_t_0, 1] <- 1.0
  P_uncond_bayes[, initial_state_t_0, 1] <- 1.0

  # --- Create the static grid for batch predictions ---
  history_grid <- expand.grid(
    h = factor(1:M, levels = 1:M), # X_{t-2}
    j = factor(1:M, levels = 1:M)  # X_{t-1}
  )
  names(history_grid) <- c(col_ypprev, col_yprev)

  # --- Dynamically create the full prediction dataset ---
  # Identify static covariates by excluding time-varying/history columns
  cols_to_exclude <- c(col_yprev, col_ypprev, col_time)
  static_covariate_names <- setdiff(names(baseline_data), cols_to_exclude)
  static_covariates <- baseline_data[, static_covariate_names, drop = FALSE]

  # Combine static covariates with the grid of all possible histories
  prediction_data <- cbind(static_covariates, history_grid)

  # --- Main Iteration Loop ---
  for (t in time_horizon) {
    # Identify predictable histories
    predictable_idx <- which(
      !(prediction_data[[col_ypprev]] %in% absorbing_states) &
      !(prediction_data[[col_yprev]] %in% absorbing_states)
    )
    n_predictable <- length(predictable_idx)

    # Subset the data for prediction and set the current time
    prediction_data_subset <- prediction_data[predictable_idx, ]
    prediction_data_subset[[col_time]] <- t


    pred_array_3d <- NULL

    if (detected_type == "brms") {
      # refactor previous states to remove absorbing states
      prediction_data_subset[[col_yprev]] <- factor(prediction_data_subset[[col_yprev]], levels = c(1:M)[-absorbing_states])
      prediction_data_subset[[col_ypprev]] <- factor(prediction_data_subset[[col_ypprev]], levels = c(1:M)[-absorbing_states])  
      re_arg <- if (include_re) NULL else NA
      brms_pred_long <- marginaleffects::predictions(model_fit, 
                                                     newdata = prediction_data_subset, 
                                                     type = "response", 
                                                     re_formula = re_arg) |>
                        marginaleffects::posterior_draws()
      
      pred_array_3d <- array(0, dim = c(n_draws, n_predictable, M))
      indices <- cbind(brms_pred_long$drawid, 
                       brms_pred_long$rowid, 
                       as.numeric(brms_pred_long$group))
      pred_array_3d[indices] <- brms_pred_long$draw
      
    } else { # detected_type == "rmsb"
      pred_array_3d <- predict_blrm_manual(model_fit, prediction_data_subset, include_re = include_re)
    }

    state_probs_all_draws <- do.call(rbind, lapply(1:dim(pred_array_3d)[1], function(d) pred_array_3d[d, , ]))

    # --- Assemble transition probabilities for each draw ---
    cond_probs_4d_array <- array(0, dim = c(n_draws, M, M, M))

    for (d in 1:n_draws) {
      cond_probs_matrix_d <- matrix(0, nrow = M * M, ncol = M)
      
      # Handle absorbing states
      absorbing_idx <- which(prediction_data[[col_yprev]] %in% absorbing_states)
      if (length(absorbing_idx) > 0) {
        absorbing_state <- as.numeric(as.character(prediction_data[[col_yprev]][absorbing_idx]))
        rows_and_cols <- cbind(absorbing_idx, absorbing_state)
        cond_probs_matrix_d[rows_and_cols] <- 1.0
      }
      
      # Get state probabilities for the current draw
      start_row <- (d - 1) * n_predictable + 1
      end_row <- d * n_predictable
      draw_d_probs <- state_probs_all_draws[start_row:end_row, ]
      
      # Place them into the correct rows
      cond_probs_matrix_d[predictable_idx, ] <- draw_d_probs
      
      # Reshape and store in the final 4D array
      cond_probs_4d_array[d, , , ] <- array(cond_probs_matrix_d, dim = c(M, M, M))
    }

    # --- Update Probabilities ---
    prev_P_joint_all_draws <- P_joint_bayes[, , , t]
    current_P_joint_all_draws <- array(0, dim = c(n_draws, M, M))

    for (d in 1:n_draws) {
      prev_P_joint_d <- prev_P_joint_all_draws[d, , ]
      cond_probs_d <- cond_probs_4d_array[d, , , ]
      
      # Law of Total Probability: P(A,B) = P(A|B)P(B)
      weighted_probs <- cond_probs_d * array(prev_P_joint_d, dim = dim(cond_probs_d))
      # Marginalize over the t-2 state (h) to get P(X_{t-1}, X_t)
      current_P_joint_d <- apply(weighted_probs, c(2, 3), sum)
      
      current_P_joint_all_draws[d, , ] <- current_P_joint_d
      
      # Marginalize over the t-1 state (j) to get P(X_t)
      uncond_probs_t <- colSums(current_P_joint_d)
      # Normalize to prevent floating point error accumulation
      P_uncond_bayes[d, , t + 1] <- uncond_probs_t
      # P_uncond_bayes[d, , t + 1] <- uncond_probs_t / sum(uncond_probs_t)
    }
    
    P_joint_bayes[, , , t + 1] <- current_P_joint_all_draws
  }

  return(P_uncond_bayes)
}
```


```{r}
# Example usage of the function
results <- sop_markov_second_order(
  model_fit = fit_second_order_b_cluster_no_int,
  baseline_data = baseline_data,
  col_yprev = "yprev",
  col_ypprev = "ypprev",
  col_time = "week",
  time_horizon = 1:26,
  n_states = 5,
  absorbing_states = c(5),
  include_re = TRUE
)

results[, , 2] == P_uncond_bayes[, , 2] # results from loop and function match!
```

# Comparing Version 1 and 2 of the function
```{r}
results_1 <- sop_markov_second_order(
  model_fit = fit_second_order_b_cluster,
  baseline_data = baseline_data,
  col_yprev = "yprev",
  col_ypprev = "ypprev",
  col_time = "week",
  time_horizon = 1:26,
  n_states = 5,
  absorbing_states = c(5),
  include_re = TRUE
)

results_2 <- sop_markov_second_order_2(
  model_fit = fit_second_order_b_cluster_no_int,
  baseline_data = baseline_data,
  col_yprev = "yprev",
  col_ypprev = "ypprev",
  col_time = "week",
  time_horizon = 1:26,
  n_states = 5,
  absorbing_states = c(5),
  include_re = FALSE
)

results_2_brms <- sop_markov_second_order_2(
  model_fit = fit_brms,
  baseline_data = baseline_data,
  col_yprev = "yprev",
  col_ypprev = "ypprev",
  col_time = "week",
  time_horizon = 1:26,
  n_states = 5,
  absorbing_states = c(5),
  include_re = FALSE
)

all.equal(results_1, results_2, check.attributes = FALSE)
```

## Visualize Results
```{r}
library(reshape2)

# Convert the 3D array to a long data frame
results_df <- melt(results_2, 
                   varnames = c("draw", "state", "time"), 
                   value.name = "probability")

# Clean up the data frame and exclude t=0
results_df_clean <- results_df |> 
  mutate(time = as.numeric(sub("t=", "", time))) |> 
  filter(time > 0)

ggplot(results_df_clean, aes(x = time, y = probability, fill = state)) +
  ggdist::stat_lineribbon(
    linetype = "solid",
    linewidth = 0.6,
    alpha = 0.5,
    .width = c(0.95)) +  
  scale_fill_brewer(palette = "Dark2") +  
  labs(
    title = "State Occupancy Probabilities Over Time",
    subtitle = "Median and 95% Credible Bands",
    x = "Time (weeks)",
    y = "Probability"
  ) +
  theme_minimal()
```

```{r}
# Convert the 3D array to a long data frame
results_df <- melt(results_2_brms, 
                   varnames = c("draw", "state", "time"), 
                   value.name = "probability")

# Clean up the data frame and exclude t=0
results_df_clean <- results_df |> 
  mutate(time = as.numeric(sub("t=", "", time))) |> 
  filter(time > 0)

ggplot(results_df_clean, aes(x = time, y = probability, fill = state)) +
  ggdist::stat_lineribbon(
    linetype = "solid",
    linewidth = 0.6,
    alpha = 0.5,
    .width = c(0.95)) +  
  scale_fill_brewer(palette = "Dark2") +  
  labs(
    title = "State Occupancy Probabilities Over Time",
    subtitle = "Median and 95% Credible Bands",
    x = "Time (weeks)",
    y = "Probability"
  ) +
  theme_minimal()
```

```{r}
# Convert the 3D array to a long data frame
results_df <- melt(results_1, 
                   varnames = c("draw", "state", "time"), 
                   value.name = "probability")

# Clean up the data frame and exclude t=0
results_df_clean <- results_df |> 
  mutate(time = as.numeric(sub("t=", "", time))) |> 
  filter(time > 0)

ggplot(results_df_clean, aes(x = time, y = probability, fill = state)) +
  ggdist::stat_lineribbon(
    linetype = "solid",
    linewidth = 0.6,
    alpha = 0.5,
    .width = c(0.95)) +  
  scale_fill_brewer(palette = "Dark2") +  
  labs(
    title = "State Occupancy Probabilities Over Time",
    subtitle = "Median and 95% Credible Bands",
    x = "Time (weeks)",
    y = "Probability"
  ) +
  theme_minimal()
```