---
title: "Varying the Strength of a single continuous covariates"
---

## Setup

```{r}
library(tidyverse)
library(VGAM)
```

## Simulate baseline data

```{r}
N <- 1e5
tx <- sample(c(0, 1), size = N, replace = TRUE)
X <- rnorm(N)
yprev <- sample(c("1", "2", "3", "4"), N, replace = TRUE, prob = c(0.8, 0.1, 0.05, 0.05))

X <- tibble(tx, X, yprev) |> 
    mutate(id = row_number())
```

## Simulation Function

```{r}
# intercepts
ints <- c(-6, -3, -2.5, -2)
extra <- c(
    X = log(3), 
    `yprev=2` = log(10), 
    `yprev=3` = log(3), 
    `yprev=4` = log(40),
    week      = log(0.95)
    )
```

```{r}
g <- function(yprev, t, tx_val, x_val, parameter = 0, extra_params) {
    
    # 1. Treatment Effect
    tx_effect <- parameter * tx_val

    # 2. Effect of covariate X
    cov_effect <- extra_params["X"] * x_val

    # 3. Effect of time
    time_effect <- extra_params["week"] * t

    # 4. Previous state effect
    yprev_effect <- 0

    # Create the name to look up in the 'extra' vector, e.g., "yprev=2"
    yprev_coef_name <- paste0('yprev=', yprev)
    if (yprev_coef_name %in% names(extra_params)) {
        yprev_effect <- extra_params[[yprev_coef_name]]
    }

    # Sum all effects to get the linear predictor
    lp <- as.numeric(tx_effect + cov_effect + time_effect + yprev_effect)
    
    return(lp)
}
```

## Apply simulation function

This is the section that needs to be expanded to efficiently simulate data for every patient (row) in X. This should be done as efficiently as possible (wihtout parallelization). The data should be in long format, where the data of X or kept and duplicated across rows for every patient. We might also want to assign seuqnetial IDS to patients.

```{r}
times <- 1:28

# Create a matrix to store the state of each patient (row) at each time (col).
# Initialize with the baseline state (at time 0).
state_matrix <- matrix(as.numeric(X$yprev), nrow = N, ncol = length(times) + 1,
                       dimnames = list(X$id, 0:28))

# Define the treatment effect parameter
parameter <- 0

for (t in times) {
    # Get the state of all patients from the previous time point.
    # For t=1, this gets column 1 (time 0). For t=2, it gets column 2 (time 1), etc.
    yprev <- state_matrix[, t]

    # Identify patients not yet in the absorbing state (5). These are 'active'.
    active_idx <- which(yprev != 5)
    
    # If all patients are in the absorbing state, no need to continue.
    if (length(active_idx) == 0) {
      # Fill remaining time steps for any absorbed patients and break.
      if (t <= max(times)) {
        state_matrix[, (t+1):ncol(state_matrix)] <- 5
      }
      break
    }
    
    # --- Calculation for active patients only ---
    
    # a. Get the previous state and baseline covariate data for active patients.
    yprev_active <- yprev[active_idx]
    X_active <- X[active_idx, ]
    
    # b. Apply the 'g' function to each active patient to get their linear predictor.
    lp <- mapply(
        FUN = g,
        yprev = yprev_active,
        tx_val = X_active$tx,
        x_val = X_active$X,
        MoreArgs = list(t = t, parameter = parameter, extra_params = extra)
    )

    # c. Calculate state probabilities from the linear predictor.
    # `outer` creates a matrix of thresholds (4 intercepts x n_active patients).
    thresholds_matrix <- outer(ints, lp, FUN = "+")
    
    # `plogis` computes the cumulative probabilities P(Y<=j).
    # Transpose to get a matrix where each row is a patient.
    cum_probs <- t(plogis(thresholds_matrix))
    
    # Calculate the probability for each state: P(Y=j) = P(Y<=j) - P(Y<=j-1).
    prob_matrix <- cbind(cum_probs, 1) - cbind(0, cum_probs)
    
    # d. Sample the new state for each active patient.
    # Note: Y is reverse coded, so we sample from 5 to 1, and NOT from 1 to 5.
    y_new_active <- apply(prob_matrix, 1, function(p) {
      sample(5:1, size = 1, prob = p)
    })

    # --- Update the main state matrix ---
    
    # First, carry forward the previous state for everyone pateints for the new time step t.
    # This is needed for absorbed patients
    state_matrix[, t+1] <- yprev
    # Then, update the states for the active patients with their newly sampled state.
    state_matrix[active_idx, t+1] <- y_new_active
}

long_data_forward <- as_tibble(state_matrix, rownames = "id") |> 
  pivot_longer(
    cols = -id,
    names_to = "time",
    values_to = "state"
  ) |> 
  mutate(
    id = as.integer(id),
    time = as.integer(time)
  ) |> 
  left_join(X |> rename(yprev_initial = yprev), by = "id") |> 
  group_by(id) |> 
  mutate(yprev = factor(lag(state, n = 1))) |> 
  filter(time > 0)
```


```{r}
long_data_forward |> 
    group_by(time, state) |> 
    summarise(n = n(), .groups = "drop") |>
    group_by(time) |>
    mutate(
        sop = n / sum(n),
        state = factor(state)) |>
    ggplot(aes(x = time, y = sop, group = state, color = state)) +
    geom_line() +
    theme_minimal()
```

## Fit model

```{r}
d <- long_data_forward |> 
    filter(yprev != "5")

m <- vglm(
        state ~ tx + X + yprev + time, 
        family = cumulative(reverse = TRUE, parallel = TRUE), 
        data = d
    )

m2 <- vglm(
        state ~ tx + yprev + time, 
        family = cumulative(reverse = TRUE, parallel = TRUE), 
        data = d
    )
```