---
title: "Tissue biopsy — data and imputation"
format: html
execute:
  echo: true
  warning: false
  message: false
---

The outcome is the number of unique days on which a tissue sample was taken
during the first 182 days after randomization (REDCap
`n_solid_biopsy_analyses`, exported as `unique_biopsy_days`). Exposure is the
time each patient was observed during that period, from randomization to
`follow_up_end_date`, which already includes the 182-day censoring and death.

```{r}
#| label: setup

library(tidyverse)
library(mice)

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")

cohort_path <- here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
)
result_dir <- here::here("results", "secondary", "tissue-biopsy")
figure_dir <- file.path(result_dir, "figures")
imputation_path <- file.path(result_dir, "imputation.rds")
completed_path <- file.path(result_dir, "completed-datasets.rds")

dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
```

## Data

The SAP uses ECOG with 0 as reference and 3 for ECOG 3 or higher.

```{r}
#| label: biopsy-data

cohort <- arrow::read_parquet(cohort_path)

biopsy_data <- cohort |>
  mutate(
    exposure_days = as.numeric(follow_up_end_date - randomization_date),
    ecog_group = case_when(
      ecog_fstcnt >= 3 ~ "3+",
      .default = as.character(ecog_fstcnt)
    ),
    ecog_group = factor(
      ecog_group, levels = c("0", "1", "2", "3+"), ordered = TRUE
    )
  ) |>
  select(
    id, tx,
    unique_biopsy_days,
    exposure_days,
    age, ecog_group
  )

stopifnot(!anyNA(biopsy_data$unique_biopsy_days))
```

Patients whose follow-up ended on the day of randomization have no time at
risk and cannot contribute to a rate, so they are excluded.

```{r}
#| label: exclude-zero-exposure

n_zero_exposure <- sum(biopsy_data$exposure_days == 0)

biopsy_data <- biopsy_data |>
  filter(exposure_days > 0)

arrow::write_parquet(
  biopsy_data,
  file.path(result_dir, "analysis-data.parquet"),
  compression = "zstd"
)

n_zero_exposure
```

## Descriptive summary

```{r}
#| label: tbl-biopsy-descriptive
#| tbl-cap: "Tissue biopsy days by randomized group"

descriptive_summary <- biopsy_data |>
  mutate(group = if_else(tx == 1, "Invitation group", "External control")) |>
  summarise(
    patients = n(),
    patients_biopsied = sum(unique_biopsy_days > 0),
    biopsy_days = sum(unique_biopsy_days),
    exposure_days = sum(exposure_days),
    .by = group
  ) |>
  mutate(n_excluded_zero_exposure = n_zero_exposure)

arrow::write_parquet(
  descriptive_summary,
  file.path(result_dir, "descriptive-summary.parquet"),
  compression = "zstd"
)

descriptive_summary |>
  knitr::kable()
```

## Imputation of baseline covariates

Only ECOG is missing.

```{r}
#| label: missing-covariates

biopsy_data |>
  summarise(across(c(age, ecog_group), \(x) sum(is.na(x))))
```

As for the other supporting analyses, we create the supporting number of MICE
datasets, with the number of datasets, iterations, and donors taken from the
shared imputation settings in `config/analysis.yml`. The imputation model
contains only the variables of the analysis model: ECOG, age, randomized
group, the outcome, and the log exposure. ECOG uses proportional odds
imputation.

```{r}
#| label: impute-covariates

imputation_data <- biopsy_data |>
  mutate(log_exposure = log(exposure_days)) |>
  select(tx, unique_biopsy_days, log_exposure, age, ecog_group)

methods <- make.method(imputation_data)
methods[] <- ""
methods["ecog_group"] <- "polr"

if (force || !file.exists(imputation_path)) {
  imputation <- mice(
    imputation_data,
    method = methods,
    m = config$imputation$supporting$m,
    maxit = config$imputation$supporting$maxit,
    donors = config$imputation$supporting$donors,
    seed = analysis_seed(config, "supporting", offset = 2700L),
    printFlag = FALSE
  )

  saveRDS(imputation, imputation_path, compress = "xz")
}

imputation <- readRDS(imputation_path)

c(m = imputation$m, iterations = imputation$iteration)
```

```{r}
#| label: fig-imputation-traces
#| fig-cap: "MICE trace plots for imputed ECOG"

plot(imputation)
```

## Completed datasets

```{r}
#| label: completed-datasets

completed_datasets <- complete(imputation, action = "all") |>
  map(\(data) {
    data |>
      mutate(
        exposure_days = exp(log_exposure),
        ecog_group = factor(ecog_group, ordered = FALSE)
      ) |>
      select(-log_exposure)
  })

saveRDS(completed_datasets, completed_path, compress = "xz")
```
