---
title: "Tissue biopsy — marginal rates, rate ratio, and rate difference"
format: html
execute:
  echo: true
  warning: false
  message: false
---

For each model and imputation, every patient's expected number of biopsy
days is predicted under both randomized groups with the exposure set to
30 days, and these predictions are averaged over all patients. This gives the
marginal rate per 30 days in each group, their ratio, and their difference.
The first `post_estimation_draws` posterior draws of each imputation are then
stacked. The unadjusted model is summarized the same way, using the observed
data.

```{r}
#| label: setup

library(tidyverse)

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()

result_dir <- here::here("results", "secondary", "tissue-biopsy")
figure_dir <- file.path(result_dir, "figures")
biopsy_draws_path <- file.path(result_dir, "draws.parquet")
biopsy_summary_path <- file.path(result_dir, "summary.parquet")

biopsy_data <- arrow::read_parquet(file.path(result_dir, "analysis-data.parquet"))
completed_datasets <- readRDS(file.path(result_dir, "completed-datasets.rds"))

biopsy_fits <- tibble(adjustment = c("adjusted", "unadjusted")) |>
  mutate(fits = map(adjustment, \(adjustment) {
    readRDS(file.path(result_dir, paste0("model-", adjustment, ".rds")))
  }))

rate_days <- 30

adjustment_labels <- c(
  adjusted = "Adjusted (prespecified)",
  unadjusted = "Unadjusted"
)
```

## Marginal rates

```{r}
#| label: marginal-rates

predict_marginal_rates <- function(fit, data) {
  fit <- marginaleffects::set_modeldata(fit, data)

  counterfactual_data <- bind_rows(
    data |> mutate(tx = 0),
    data |> mutate(tx = 1)
  ) |>
    mutate(exposure_days = rate_days)

  marginaleffects::avg_predictions(
    fit, newdata = counterfactual_data, by = "tx", type = "response"
  ) |>
    marginaleffects::get_draws() |>
    mutate(drawid = as.integer(as.character(drawid))) |>
    filter(drawid <= config$imputation$supporting$post_estimation_draws) |>
    select(tx, drawid, draw)
}

predict_model_rates <- function(fits, adjustment) {
  prediction_data <- if (adjustment == "adjusted") {
    completed_datasets
  } else {
    list(biopsy_data)
  }

  map2(fits, prediction_data, predict_marginal_rates) |>
    list_rbind(names_to = "imputation")
}

rate_draws <- biopsy_fits |>
  mutate(draws = map2(fits, adjustment, predict_model_rates)) |>
  select(adjustment, draws) |>
  unnest(draws)
```

```{r}
#| label: contrasts

biopsy_draws <- rate_draws |>
  pivot_wider(names_from = tx, values_from = draw, names_prefix = "rate_tx") |>
  mutate(
    rate_ratio = rate_tx1 / rate_tx0,
    rate_difference = rate_tx1 - rate_tx0
  )

arrow::write_parquet(biopsy_draws, biopsy_draws_path, compression = "zstd")
```

## Summary

`probability_lower_rate` is the posterior probability that the invitation
group has a lower biopsy-day rate than the external control.

```{r}
#| label: tbl-biopsy-summary
#| tbl-cap: "Posterior marginal biopsy-day rates per 30 days, rate ratio, and rate difference (intervention - control)"

biopsy_summary <- biopsy_draws |>
  pivot_longer(
    c(rate_tx0, rate_tx1, rate_ratio, rate_difference),
    names_to = "quantity", values_to = "draw"
  ) |>
  summarise(
    posterior_median = median(draw),
    lower_95 = quantile(draw, 0.025),
    upper_95 = quantile(draw, 0.975),
    probability_lower_rate = case_when(
      quantity[1] == "rate_ratio" ~ mean(draw < 1),
      quantity[1] == "rate_difference" ~ mean(draw < 0)
    ),
    .by = c(adjustment, quantity)
  )

arrow::write_parquet(
  biopsy_summary, biopsy_summary_path, compression = "zstd"
)

biopsy_summary |>
  knitr::kable(digits = 3)
```

## Figures

The adjusted (prespecified) and unadjusted estimates are shown in separate
panels.

```{r}
#| label: plot-setup

posterior_theme <- theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

plot_draws <- biopsy_draws |>
  mutate(adjustment = factor(
    adjustment_labels[adjustment], levels = adjustment_labels
  ))
```

```{r}
#| label: fig-biopsy-rate-ratio
#| fig-cap: "Posterior marginal rate ratio of biopsy days (intervention / control). Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "Two panels, adjusted and unadjusted, each with a half-eye density above a posterior dot plot of the marginal rate ratio of biopsy days, intervention over control, on a log scale, with thick and thin intervals for the 66 and 95 percent credible intervals and a dashed vertical line at a ratio of one."
#| fig-width: 8
#| fig-height: 7

rate_ratio_plot <- ggplot(plot_draws, aes(x = rate_ratio)) +
  ggdist::stat_dots(side = "bottom") +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  scale_x_log10() +
  facet_wrap(~ adjustment, ncol = 1) +
  labs(x = "Rate ratio of biopsy days (intervention / control)", y = NULL) +
  posterior_theme

ggsave(
  file.path(figure_dir, "biopsy-rate-ratio.png"),
  rate_ratio_plot, width = 8, height = 7, dpi = 300
)

rate_ratio_plot
```

```{r}
#| label: fig-biopsy-rate-difference
#| fig-cap: "Posterior marginal difference in biopsy days per 30 days (intervention - control). Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "Two panels, adjusted and unadjusted, each with a half-eye density above a posterior dot plot of the marginal difference in biopsy days per 30 days of follow-up, intervention minus control, with thick and thin intervals for the 66 and 95 percent credible intervals and a dashed vertical line at zero."
#| fig-width: 8
#| fig-height: 7

rate_difference_plot <- ggplot(plot_draws, aes(x = rate_difference)) +
  ggdist::stat_dots(side = "bottom") +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ adjustment, ncol = 1) +
  labs(
    x = "Difference in biopsy days per 30 days (intervention - control)",
    y = NULL
  ) +
  posterior_theme

ggsave(
  file.path(figure_dir, "biopsy-rate-difference.png"),
  rate_difference_plot, width = 8, height = 7, dpi = 300
)

rate_difference_plot
```

```{r}
#| label: fig-biopsy-marginal-rates
#| fig-cap: "Posterior marginal rate of biopsy days per 30 days by randomized group. Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "A grid of posterior half-eye densities above dot plots of the marginal number of biopsy days per 30 days of follow-up, with rows for the adjusted and unadjusted models and columns for the external control and the invitation group, each with thick and thin intervals for the 66 and 95 percent credible intervals."
#| fig-width: 10
#| fig-height: 7

marginal_rates_plot <- rate_draws |>
  mutate(
    adjustment = factor(adjustment_labels[adjustment], levels = adjustment_labels),
    group = if_else(tx == 1, "Invitation group", "External control")
  ) |>
  ggplot(aes(x = draw)) +
  ggdist::stat_dots(side = "bottom") +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  facet_grid(adjustment ~ group) +
  labs(x = "Biopsy days per 30 days", y = NULL) +
  posterior_theme

ggsave(
  file.path(figure_dir, "biopsy-marginal-rates.png"),
  marginal_rates_plot, width = 10, height = 7, dpi = 300
)

marginal_rates_plot
```
