---
title: "LIQPLAT baseline population table"
format: html
execute:
  echo: false
  warning: false
  message: false
---

Table 1 describes the randomized population by invitation assignment, irrespective
of whether an invitation was accepted. Continuous variables are summarized by
median [Q1, Q3]; categorical variables by n (%), using the full column denominator.
Missing values remain visible; no multiple imputation is used.

```{r}
#| label: baseline-table-data
#| include: false

source(here::here("R", "analysis-config.R"))
source(here::here("R", "survival-helpers.R"))
source(here::here("R", "baseline-table-helpers.R"))
config <- read_analysis_config()
# Read the current V5 cohort, including the newly exported baseline Q29.
cohort <- arrow::read_parquet(here::here(
  config_get(config, "paths", "private_data"), "cohort_follow_up.parquet"
)) |> as.data.frame()

required <- c("id", "tx", "gender", "age", "ecog_fstcnt", "baseline_q29",
              "baseline_q30", "diagnosis", "albumin", "c_reactive_protein", "stage_binary")

stopifnot(all(required %in% names(cohort)), !anyDuplicated(cohort$id),
          all(cohort$tx %in% 0:1),
          all(is.na(cohort$gender) | cohort$gender %in% c("female", "male")),
          all(is.na(cohort$ecog_fstcnt) | cohort$ecog_fstcnt %in% 0:4),
          all(is.na(cohort$stage_binary) | cohort$stage_binary %in% 0:1))

cohort$group <- factor(cohort$tx, levels = c(1, 0),
  labels = c("Selected for SAT invitation", "Usual care"))

cohort$gender <- factor(cohort$gender, levels = c("female", "male"), labels = c("Female", "Male"))

cohort$ecog_fstcnt <- factor(cohort$ecog_fstcnt, levels = 0:4)

qol <- arrow::read_parquet(here::here(
  config_get(config, "paths", "private_data"), "qol.parquet"
)) |> as.data.frame()

cohort <- complete_table_baseline_qol(cohort, qol)

cohort$ql2 <- baseline_ql2(cohort$baseline_q29, cohort$baseline_q30)

cohort$diagnosis <- snomed_to_diagnosis(as.character(cohort$diagnosis))

if (any(grepl("^SNM_", cohort$diagnosis))) stop("Add labels for unmapped diagnosis codes.")

cohort$diagnosis <- factor(cohort$diagnosis)
cohort$mgps <- factor(derive_mgps(cohort$albumin, cohort$c_reactive_protein), levels = 0:2)

cohort$stage_binary <- factor(cohort$stage_binary, levels = 0:1,
  labels = c("Locally advanced", "Metastasised"))
  
labels <- c(gender = "Gender", age = "Age (years)", ecog_fstcnt = "ECOG at first contact",
  ql2 = "Global health status / QoL (QL2, 0–100)", diagnosis = "Diagnosis",
  mgps = "Modified Glasgow Prognostic Score (mGPS)", stage_binary = "Disease stage")

for (variable in names(labels)) attr(cohort[[variable]], "label") <- labels[[variable]]

footnote <- paste(
  "PROMs are considered up to 14 days after first contact.",
  "QL2: higher scores indicate better QoL.",
  "Per EORTC scoring, one available item is sufficient; both missing gives a missing score.",
  "mGPS uses baseline CRP and albumin; missing either gives a missing score.",
  "Percentages include missing observations in the denominator."
)

baseline_table <- table1::table1(
  ~ gender + age + ecog_fstcnt + ql2 + diagnosis + mgps + stage_binary | group,
  data = cohort, overall = "Overall", render.continuous = c(. = "Median [Q1, Q3]"),
  render.categorical = "FREQ (PCT%)", topclass = "Rtable1-zebra",
  footnote = footnote
)

result_dir <- here::here("analysis", "01-population", "results")
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)
saveRDS(baseline_table, file.path(result_dir, "baseline-table1.rds"))

# Store aggregate display cells, never participant-level records, for other formats.
arrow::write_parquet(as.data.frame(baseline_table),
  file.path(result_dir, "baseline-characteristics.parquet"))
```

```{r}
#| label: tbl-baseline-table1
#| tbl-cap: "Baseline characteristics by randomized invitation assignment."

baseline_table
```

QL2 follows the [EORTC QLQ-C30 scoring manual](https://qol.eortc.org/app/uploads/sites/2/2018/02/SCmanual.pdf).
Dedicated baseline values take priority. For each missing item, the closest
nonmissing questionnaire response within 14 days before or after first contact
is used, including eligible post-contact measurements. Ties use the earlier
questionnaire date, then assessment ID, as in the primary QoL analysis.
Selection is item-specific, so Q29 and Q30 can originate from different
questionnaires when item availability differs.
The broad diagnosis categories use the labels from the original project Table 1,
without restricting the display to the most common diagnoses or using OncoTree.

```{r}
#| results: asis

cat(sprintf("QL2 uses both baseline items for %d participants and one item for %d; %d have neither item.\n",
  sum(!is.na(cohort$baseline_q29) & !is.na(cohort$baseline_q30)),
  sum(xor(is.na(cohort$baseline_q29), is.na(cohort$baseline_q30))),
  sum(is.na(cohort$baseline_q29) & is.na(cohort$baseline_q30))))
```
