---
title: "Overall survival — adjusted censoring analysis"
subtitle: "Observed and assumed administrative-only censoring"
format: html
---

- Exploratory: censoring is the event; deaths are censored.
- Administrative-only (assumed): no recorded death means alive, censored at end of potential follow-up.
- Missing binary ECOG and mGPS replaced by pooled modes; uncertainty not propagated.

```{r}
#| label: setup
#| include: false

library(rstanarm)
library(survival)
library(posterior)
library(ggsurvfit)
library(knitr)
library(here)
library(arrow)
library(yaml)
library(tidyverse)
source(here("R", "plot-style.R"))

config <- read_yaml(here("config", "analysis.yml"))
source(here("R", "covariates.R"))
source(here("R", "survival.R"))

results_dir <- here(
  "analysis", "02-overall-survival", "06-censoring", "results",
  "adjusted-censoring"
)
pattern_dirs <- list(
  observed = file.path(results_dir, "observed"),
  administrative = file.path(results_dir, "administrative")
)
for (pattern_dir in pattern_dirs) {
  dir.create(
    file.path(pattern_dir, "models"),
    recursive = TRUE,
    showWarnings = FALSE
  )
  dir.create(
    file.path(pattern_dir, "figures"),
    recursive = TRUE,
    showWarnings = FALSE
  )
}
```

## Censoring data and modal replacement

```{r}
#| label: prepare-data
#| include: false

cohort <- read_parquet(
  here("data", "private", "cohort_follow_up.parquet")
)

stopifnot(
  all(
    c(
      "id", "tx", "survival_time_days_unrestricted",
      "status_death_unrestricted", "potential_follow_up_days", "ecog_fstcnt",
      "stage_binary", "diagnosis", "albumin", "c_reactive_protein"
    ) %in% names(cohort)
  ),
  !anyDuplicated(cohort$id),
  !anyNA(cohort[c(
    "id", "tx", "survival_time_days_unrestricted",
    "status_death_unrestricted", "potential_follow_up_days", "stage_binary",
    "diagnosis"
  )]),
  all(cohort$tx %in% 0:1),
  all(cohort$status_death_unrestricted %in% 0:1),
  all(cohort$stage_binary %in% 0:1),
  all(is.finite(cohort$survival_time_days_unrestricted)),
  all(cohort$survival_time_days_unrestricted >= 0),
  all(is.finite(cohort$potential_follow_up_days)),
  all(
    cohort$potential_follow_up_days >= cohort$survival_time_days_unrestricted
  ),
  all(na.omit(cohort$ecog_fstcnt) %in% 0:4)
)

# A plain data.frame (not a tibble), as used by the original fits and by
# rstanarm's prediction code.
censoring_data <- cohort |>
  transmute(
    id,
    tx,
    follow_up_days = survival_time_days_unrestricted,
    censoring_event = as.integer(1L - status_death_unrestricted),
    ecog_binary = derive_ecog_binary(ecog_fstcnt),
    stage_binary,
    diagnosis = factor(diagnosis),
    mgps = derive_mgps(albumin, c_reactive_protein)
  ) |>
  as.data.frame()

# table() orders the numeric categories, so ties select the lowest category.
modal_replacements <- map(c("ecog_binary", "mgps"), \(variable) {
  frequencies <- table(censoring_data[[variable]])
  stopifnot(length(frequencies) > 0)

  tibble(
    variable = variable,
    missing = sum(is.na(censoring_data[[variable]])),
    replacement = as.numeric(names(frequencies)[which.max(frequencies)])
  )
}) |>
  list_rbind()

ecog_mode <- modal_replacements$replacement[1]
mgps_mode <- modal_replacements$replacement[2]

censoring_data <- censoring_data |>
  mutate(
    ecog_binary = if_else(is.na(ecog_binary), ecog_mode, ecog_binary),
    mgps = factor(if_else(is.na(mgps), mgps_mode, mgps), levels = 0:2)
  )

stopifnot(!anyNA(censoring_data))

write_parquet(
  modal_replacements,
  file.path(results_dir, "modal-replacements.parquet")
)
```

```{r}
#| label: censoring-patterns
#| include: false

# Only follow-up time changes; the censoring indicator stays as is.
administrative_data <- censoring_data |>
  mutate(
    follow_up_days = if_else(
      cohort$status_death_unrestricted == 1L,
      cohort$survival_time_days_unrestricted,
      cohort$potential_follow_up_days
    )
  )

patterns <- list(
  observed = censoring_data,
  administrative = administrative_data
)

# The participant censored at day zero contributes no positive-time
# likelihood, but stays in the curves and the standardization population.
fitting_data <- map(patterns, \(data) filter(data, follow_up_days > 0))

max_follow_up <- max(cohort$potential_follow_up_days)
```

```{r}
#| label: tbl-modal-replacements
#| tbl-cap: "Pooled modal replacement of missing binary ECOG and mGPS"
#| echo: false

kable(modal_replacements)
```

## Fitting

```{r}
#| label: fit-models
#| include: false

models <- imap(fitting_data, \(data, pattern) {
  stopifnot(
    all(
      levels(patterns[[pattern]]$diagnosis) %in% as.character(data$diagnosis)
    )
  )

  model_path <- file.path(pattern_dirs[[pattern]], "models", "model-01.rds")

  if (!file.exists(model_path)) {
    model <- stan_surv(
      Surv(follow_up_days, censoring_event) ~
        tx + ecog_binary + stage_binary + mgps + (1 | diagnosis),
      data = data,
      basehaz = "ms",
      basehaz_ops = list(df = 5),
      # Deliberately broad treatment prior, autoscaling off.
      prior = normal(
        location = rep(0, 5),
        scale = c(5, rep(2.5, 4)),
        autoscale = FALSE
      ),
      prior_intercept = normal(0, 20, autoscale = FALSE),
      prior_aux = dirichlet(1),
      prior_covariance = decov(),
      chains = 4L,
      cores = 4L,
      iter = 2000L,
      warmup = 1000L,
      adapt_delta = 0.99,
      seed = 4401L,
      refresh = 100L
    )

    saveRDS(model, model_path, compress = "xz")
  }

  readRDS(model_path)
})
```

## Sampling diagnostics

```{r}
#| label: diagnostics
#| include: false

parameters <- map(models, \(model) {
  summarise_draws(as_draws_array(model)) |>
    as.data.frame()
})

diagnostics <- imap(models, \(model, pattern) {
  sampler <- rstan::get_sampler_params(model$stanfit, inc_warmup = FALSE)

  tibble(
    fits = 1L,
    multiple_imputations = 0L,
    mice_maxit = NA_integer_,
    randomized = nrow(patterns[[pattern]]),
    fitted = nrow(fitting_data[[pattern]]),
    max_rhat = max(parameters[[pattern]]$rhat, na.rm = TRUE),
    min_ess_bulk = min(parameters[[pattern]]$ess_bulk, na.rm = TRUE),
    min_ess_tail = min(parameters[[pattern]]$ess_tail, na.rm = TRUE),
    divergences = sum(map_dbl(sampler, \(x) sum(x[, "divergent__"]))),
    treedepth_hits = sum(map_dbl(sampler, \(x) sum(x[, "treedepth__"] >= 10))),
    min_bfmi = min(map_dbl(sampler, \(x) {
      mean(diff(x[, "energy__"])^2) / var(x[, "energy__"])
    })),
    chains = model$stanfit@sim$chains,
    iterations = model$stanfit@sim$iter,
    warmup = model$stanfit@sim$warmup
  )
})

iwalk(diagnostics, \(pattern_diagnostics, pattern) {
  write_parquet(
    pattern_diagnostics,
    file.path(pattern_dirs[[pattern]], "adjusted-diagnostics.parquet")
  )
  write_parquet(
    parameters[[pattern]],
    file.path(pattern_dirs[[pattern]], "adjusted-parameter-summary.parquet")
  )

  stopifnot(
    pattern_diagnostics$max_rhat <= 1.01,
    pattern_diagnostics$min_ess_bulk >= 400,
    pattern_diagnostics$min_ess_tail >= 400,
    pattern_diagnostics$divergences == 0,
    pattern_diagnostics$treedepth_hits == 0,
    pattern_diagnostics$min_bfmi >= 0.3
  )
})
```

## Censoring hazard ratio and cohort summary

- HR above 1: faster censoring with ctDNA.

```{r}
#| label: hazard-ratio
#| include: false

adjusted_summaries <- map(models, \(model) {
  hr <- exp(as.matrix(model, pars = "tx"))[, 1]

  tibble(
    median = median(hr),
    lower_95 = quantile(hr, 0.025),
    upper_95 = quantile(hr, 0.975),
    probability_higher_censoring = mean(hr > 1)
  )
})

cohort_summaries <- map(patterns, \(data) {
  data |>
    summarise(
      participants = n(),
      censoring_events = sum(censoring_event),
      deaths = sum(1L - censoring_event),
      zero_follow_up = sum(follow_up_days == 0),
      .by = tx
    )
})

iwalk(adjusted_summaries, \(summary, pattern) {
  write_parquet(
    summary,
    file.path(
      pattern_dirs[[pattern]],
      "adjusted-hazard-ratio-summary.parquet"
    )
  )
  write_parquet(
    cohort_summaries[[pattern]],
    file.path(pattern_dirs[[pattern]], "adjusted-cohort-summary.parquet")
  )
})
```

## Standardized censoring curves

```{r}
#| label: censoring-curves
#| include: false

curve_summaries <- imap(models, \(model, pattern) {
  curves <- posterior_standardized_survival_draws(
    model,
    patterns[[pattern]],
    horizon = max_follow_up,
    n_points = 200L,
    n_draws = 1000L,
    seed = 4402L
  )

  # Probabilities in [0, 1] and non-increasing within each draw.
  monotone <- curves |>
    group_by(tx, draw) |>
    summarise(ok = all(diff(survival) <= 1e-10), .groups = "drop") |>
    pull(ok)
  stopifnot(all(curves$survival >= 0 & curves$survival <= 1), all(monotone))

  curves |>
    summarise(
      median = median(survival),
      lower = quantile(survival, 0.025),
      upper = quantile(survival, 0.975),
      .by = c(tx, time)
    )
})
```

## Prediction checks

```{r}
#| label: prediction-checks
#| include: false

checks <- imap(models, \(model, pattern) {
  newdata <- patterns[[pattern]] |>
    mutate(tx = 1)

  individual <- posterior_survfit_compatible(
    model,
    newdata = newdata,
    times = 182,
    extrapolate = FALSE,
    standardise = FALSE,
    return_matrix = TRUE,
    draws = 100L,
    seed = 4403L
  )[[1]]

  standardized <- posterior_survfit_compatible(
    model,
    newdata = newdata,
    times = 182,
    extrapolate = FALSE,
    standardise = TRUE,
    return_matrix = TRUE,
    draws = 100L,
    seed = 4403L
  )[[1]]

  # One row per diagnosis, otherwise identical.
  contrast_data <- newdata[rep(1L, nlevels(newdata$diagnosis)), ]
  contrast_data$diagnosis <- factor(
    levels(newdata$diagnosis),
    levels = levels(newdata$diagnosis)
  )

  diagnosis_predictions <- posterior_survfit_compatible(
    model,
    newdata = contrast_data,
    times = 182,
    extrapolate = FALSE,
    standardise = FALSE,
    return_matrix = TRUE,
    draws = 100L,
    seed = 4403L
  )[[1]]

  pattern_checks <- tibble(
    maximum_averaging_error = max(
      abs(rowMeans(individual) - as.numeric(standardized))
    ),
    maximum_diagnosis_contrast = max(
      apply(diagnosis_predictions, 1, \(x) diff(range(x)))
    )
  )

  stopifnot(
    pattern_checks$maximum_averaging_error < 1e-10,
    pattern_checks$maximum_diagnosis_contrast > 1e-8
  )

  pattern_checks
})

iwalk(curve_summaries, \(curve_summary, pattern) {
  write_parquet(
    curve_summary,
    file.path(pattern_dirs[[pattern]], "adjusted-censoring-curves.parquet")
  )
  write_parquet(
    checks[[pattern]],
    file.path(pattern_dirs[[pattern]], "adjusted-prediction-checks.parquet")
  )
})
```

## Figures

```{r}
#| label: build-plots
#| include: false

reverse_plots <- imap(patterns, \(plot_data, pattern) {
  km <- survfit2(
    Surv(follow_up_days, censoring_event) ~ tx_plot,
    data = mutate(plot_data, tx_plot = group_factor(tx)),
    conf.type = "log-log"
  )

  # Reverse Kaplan–Meier (dashed) and adjusted censoring curves (solid).
  reverse_plot <- plot_km_model(
    km,
    curve_summaries[[pattern]],
    x_max = max_follow_up,
    y_label = "Probability of remaining uncensored",
    event_label = "Censoring"
  )

  save_figure(
    ggsurvfit_build(reverse_plot),
    file.path(
      pattern_dirs[[pattern]],
      "figures",
      "adjusted-reverse-kaplan-meier.png"
    ),
    height = 6.5
  )

  reverse_plot
})

trace_plots <- imap(models, \(model, pattern) {
  trace_plot <- trace_draws(
    model,
    c("tx", "ecog_binary", "stage_binary", "mgps1", "mgps2")
  ) |>
    plot_traces(ncol = 2)

  save_figure(
    trace_plot,
    file.path(
      pattern_dirs[[pattern]],
      "figures",
      "adjusted-censoring-traces.png"
    ),
    height = 4.5
  )

  trace_plot
})
```

## Frequentist Cox models and reverse Kaplan–Meier landmarks

```{r}
#| label: cox-and-landmarks
#| include: false

cox_models <- map(patterns, \(data) {
  coxph(Surv(follow_up_days, censoring_event) ~ tx, data = data)
})

cox_results <- imap(cox_models, \(cox, pattern) {
  cox_summary <- summary(cox)

  tibble(
    pattern,
    hazard_ratio = cox_summary$conf.int[1, "exp(coef)"],
    lower_95 = cox_summary$conf.int[1, "lower .95"],
    upper_95 = cox_summary$conf.int[1, "upper .95"],
    p_value = cox_summary$coefficients[1, "Pr(>|z|)"],
    participants = nrow(patterns[[pattern]]),
    censoring_events = sum(patterns[[pattern]]$censoring_event)
  )
}) |>
  list_rbind()

cox_ph_results <- imap(cox_models, \(cox, pattern) {
  ph_test <- cox.zph(cox)$table

  tibble(
    pattern,
    term = rownames(ph_test),
    chisq = ph_test[, "chisq"],
    df = ph_test[, "df"],
    p_value = ph_test[, "p"]
  )
}) |>
  list_rbind()

landmarks <- imap(patterns, \(data, pattern) {
  km <- survfit(Surv(follow_up_days, censoring_event) ~ tx, data = data)
  landmark <- summary(km, times = c(182, 365, 547, 730), extend = TRUE)

  tibble(
    pattern,
    tx = as.integer(sub("tx=", "", landmark$strata)),
    time = landmark$time,
    estimate = landmark$surv
  )
}) |>
  list_rbind()

hazard_ratios <- imap(adjusted_summaries, \(summary, pattern) {
  mutate(summary, pattern = pattern, .before = 1)
}) |>
  list_rbind()

diagnostics_comparison <- imap(diagnostics, \(pattern_diagnostics, pattern) {
  mutate(pattern_diagnostics, pattern = pattern, .before = 1)
}) |>
  list_rbind()

write_parquet(
  hazard_ratios,
  file.path(results_dir, "adjusted-comparison.parquet")
)
write_parquet(
  diagnostics_comparison,
  file.path(results_dir, "diagnostics-comparison.parquet")
)
write_parquet(cox_results, file.path(results_dir, "cox-comparison.parquet"))
write_parquet(
  cox_ph_results,
  file.path(results_dir, "cox-ph-comparison.parquet")
)
write_parquet(
  landmarks,
  file.path(results_dir, "reverse-km-landmarks.parquet")
)
```

## Results

```{r}
#| label: tbl-adjusted-censoring-cohort
#| tbl-cap: "Participants, censoring events, deaths and zero follow-up by randomized group (0: usual care; 1: invitation)"
#| echo: false

imap(cohort_summaries, \(summary, pattern) {
  mutate(summary, pattern = pattern, .before = 1)
}) |>
  list_rbind() |>
  kable()
```

```{r}
#| label: tbl-adjusted-censoring-hr
#| tbl-cap: "Adjusted censoring hazard ratio: invitation versus usual care"
#| echo: false

kable(hazard_ratios, digits = 3)
```

- Similar adjusted curves do not establish non-informative censoring.

```{r}
#| label: fig-adjusted-reverse-km
#| fig-cap: "Observed censoring. Dashed: empirical reverse Kaplan–Meier curves. Solid: standardized adjusted censoring curves with 95% credible bands. Marks represent deaths; table events represent censoring."
#| fig-alt: "Empirical and covariate-adjusted reverse-survival curves for the two randomized groups, with risk counts and cumulative censoring."
#| echo: false

reverse_plots$observed
```

```{r}
#| label: fig-scenarios-administrative-censoring
#| fig-cap: "Assumed administrative-only censoring: all participants without recorded deaths are assumed alive to the cutoff, including those censored within 14 days. Dashed lines are reverse Kaplan–Meier estimates; solid lines are adjusted predictions with 95% credible bands."
#| fig-alt: "Reverse Kaplan–Meier and adjusted curves after extending censored patients to the cutoff, with numbers at risk and cumulative censoring."
#| echo: false

reverse_plots$administrative
```

```{r}
#| label: tbl-censoring-cox-comparison
#| tbl-cap: "Unadjusted Cox models for the censoring hazard, their proportional-hazards tests, and reverse Kaplan–Meier landmarks"
#| echo: false

kable(cox_results, digits = 3)
kable(cox_ph_results, digits = 3)
kable(landmarks, digits = 3)
```

## Diagnostics and actual sampling settings

```{r}
#| label: tbl-adjusted-censoring-diagnostics
#| tbl-cap: "Sampling diagnostics and prediction checks for both adjusted censoring models"
#| echo: false

kable(diagnostics_comparison, digits = 3)

imap(checks, \(pattern_checks, pattern) {
  mutate(pattern_checks, pattern = pattern, .before = 1)
}) |>
  list_rbind() |>
  kable(digits = 10)
```

```{r}
#| label: fig-adjusted-censoring-traces
#| fig-cap: "Posterior chains for the fixed effects of the observed-censoring model"
#| echo: false

trace_plots$observed
```

```{r}
#| label: fig-administrative-censoring-traces
#| fig-cap: "Posterior chains for the fixed effects of the administrative-only censoring model"
#| echo: false

trace_plots$administrative
```
