---
title: "Overall survival with unrestricted follow-up — invitation-acceptance comparisons"
subtitle: "Descriptive Kaplan–Meier curves"
format: html
execute:
  freeze: false
---

These descriptive plots compare the control group with all intervention
participants and, separately, with intervention participants who accepted or
did not accept the invitation. Invitation acceptance is a post-randomization
characteristic, so the restricted comparisons do not estimate a causal effect
of accepting the invitation. Intervention participants with missing acceptance
status are retained in the randomized-group plot and omitted from both
acceptance-restricted plots.

```{r}
#| label: build-supporting-os-unrestricted-invitation-km
#| include: false
library(tidyverse)

input_path <- here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
)
figure_dir <- here::here(
  "artifacts", "supporting", "os-unrestricted", "figures"
)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

prepared <- arrow::read_parquet(input_path) |>
  dplyr::mutate(
    tx_plot = factor(
      tx,
      levels = 0:1,
      labels = c("Control", "Intervention")
    )
  )

max_follow_up <- max(
  prepared$survival_time_days_unrestricted,
  na.rm = TRUE
)
risk_times <- seq(0, max_follow_up, by = 100)

randomized_km_fit <- ggsurvfit::survfit2(
  survival::Surv(
    survival_time_days_unrestricted,
    status_death_unrestricted
  ) ~ tx_plot,
  data = prepared,
  conf.type = "log-log"
)

randomized_km_plot <- ggsurvfit::ggsurvfit(
  randomized_km_fit,
  linewidth = 0.55,
  linetype = 2
) +
  ggsurvfit::add_censor_mark(size = 1.8) +
  ggplot2::scale_x_continuous(breaks = risk_times) +
  ggplot2::coord_cartesian(
    xlim = c(0, max_follow_up),
    ylim = c(0, 1)
  ) +
  ggplot2::labs(
    x = "Days since randomization",
    y = "Survival probability",
    colour = NULL,
    fill = NULL
  ) +
  ggplot2::theme_bw() +
  ggsurvfit::add_risktable(
    times = risk_times,
    risktable_stats = c("n.risk", "cum.event"),
    stats_label = list(n.risk = "At risk", cum.event = "Events")
  )

accepted <- prepared |>
  dplyr::filter(
    tx == 0 | (tx == 1 & invitation_accepted == 1)
  ) |>
  dplyr::mutate(
    acceptance_plot = factor(
      tx,
      levels = 0:1,
      labels = c("Control", "Intervention")
    )
  )

accepted_km_fit <- ggsurvfit::survfit2(
  survival::Surv(
    survival_time_days_unrestricted,
    status_death_unrestricted
  ) ~ acceptance_plot,
  data = accepted,
  conf.type = "log-log"
)

accepted_km_plot <- ggsurvfit::ggsurvfit(
  accepted_km_fit,
  linewidth = 0.55,
  linetype = 2
) +
  ggsurvfit::add_censor_mark(size = 1.8) +
  ggplot2::scale_x_continuous(breaks = risk_times) +
  ggplot2::coord_cartesian(
    xlim = c(0, max_follow_up),
    ylim = c(0, 1)
  ) +
  ggplot2::labs(
    x = "Days since randomization",
    y = "Survival probability",
    colour = NULL,
    fill = NULL
  ) +
  ggplot2::theme_bw() +
  ggsurvfit::add_risktable(
    times = risk_times,
    risktable_stats = c("n.risk", "cum.event"),
    stats_label = list(n.risk = "At risk", cum.event = "Events")
  )

not_accepted <- prepared |>
  dplyr::filter(
    tx == 0 | (tx == 1 & (invitation_accepted == 0 | is.na(invitation_accepted == 0)))
  ) |>
  dplyr::mutate(
    acceptance_plot = factor(
      tx,
      levels = 0:1,
      labels = c("Control", "Intervention")
    )
  )

not_accepted_km_fit <- ggsurvfit::survfit2(
  survival::Surv(
    survival_time_days_unrestricted,
    status_death_unrestricted
  ) ~ acceptance_plot,
  data = not_accepted,
  conf.type = "log-log"
)

not_accepted_km_plot <- ggsurvfit::ggsurvfit(
  not_accepted_km_fit,
  linewidth = 0.55,
  linetype = 2
) +
  ggsurvfit::add_censor_mark(size = 1.8) +
  ggplot2::scale_x_continuous(breaks = risk_times) +
  ggplot2::coord_cartesian(
    xlim = c(0, max_follow_up),
    ylim = c(0, 1)
  ) +
  ggplot2::labs(
    x = "Days since randomization",
    y = "Survival probability",
    colour = NULL,
    fill = NULL
  ) +
  ggplot2::theme_bw() +
  ggsurvfit::add_risktable(
    times = risk_times,
    risktable_stats = c("n.risk", "cum.event"),
    stats_label = list(n.risk = "At risk", cum.event = "Events")
  )

ggplot2::ggsave(
  file.path(figure_dir, "kaplan-meier-randomized-groups.png"),
  ggsurvfit::ggsurvfit_build(randomized_km_plot),
  width = 8,
  height = 6.5,
  dpi = 300
)
ggplot2::ggsave(
  file.path(figure_dir, "kaplan-meier-control-vs-accepted.png"),
  ggsurvfit::ggsurvfit_build(accepted_km_plot),
  width = 8,
  height = 6.5,
  dpi = 300
)
ggplot2::ggsave(
  file.path(figure_dir, "kaplan-meier-control-vs-not-accepted.png"),
  ggsurvfit::ggsurvfit_build(not_accepted_km_plot),
  width = 8,
  height = 6.5,
  dpi = 300
)
```

```{r}
#| label: fig-supporting-os-unrestricted-randomized-km
#| fig-cap: "Kaplan–Meier estimates using unrestricted follow-up in the randomized groups"
#| fig-alt: "Kaplan–Meier survival curves for all control and all intervention participants, with censoring marks and at-risk and cumulative-event counts every 100 days."
#| fig-width: 8
#| fig-height: 6.5
#| echo: false

randomized_km_plot
```

```{r}
#| label: fig-supporting-os-unrestricted-accepted-km
#| fig-cap: "Kaplan–Meier estimates using unrestricted follow-up for controls and intervention participants who accepted the invitation"
#| fig-alt: "Kaplan–Meier survival curves for all control participants and intervention participants recorded as having accepted the invitation, with censoring marks and at-risk and cumulative-event counts every 100 days."
#| fig-width: 8
#| fig-height: 6.5
#| echo: false

accepted_km_plot
```

```{r}
#| label: fig-supporting-os-unrestricted-not-accepted-km
#| fig-cap: "Kaplan–Meier estimates using unrestricted follow-up for controls and intervention participants who did not accept the invitation"
#| fig-alt: "Kaplan–Meier survival curves for all control participants and intervention participants recorded as not having accepted the invitation, with censoring marks and at-risk and cumulative-event counts every 100 days."
#| fig-width: 8
#| fig-height: 6.5
#| echo: false

not_accepted_km_plot
```
