---
title: "Unrestricted follow-up — adjusted censoring analysis"
format: html
execute:
  freeze: false
---

This exploratory extension treats censoring as the event and death as a censored observation. It uses the five-degree-of-freedom M-spline baseline hazard and diagnosis random-intercept structure of the supporting OS model, with binary ECOG, stage, and categorical mGPS. Missing binary ECOG and mGPS are replaced by their respective pooled modes, without using treatment-specific modes or multiple imputation. Ties select the lowest category. The uncertainty introduced by these deterministic replacements is not propagated.

The treatment log hazard ratio has a proper Normal(0, 5) prior with automatic scaling disabled. It is deliberately broad: its central 95% interval on the hazard-ratio scale is approximately 0.000055 to 18,000. Other fixed effects use Normal(0, 2.5), the intercept Normal(0, 20), the baseline spline weights Dirichlet(1), and the diagnosis variance uses `decov()`.

```{r}
#| label: fit-adjusted-censoring
#| include: false
library(tidyverse)
source(here::here("R", "survival-helpers.R"))
source(here::here("R", "censoring-helpers.R"))
source(here::here("R", "analysis-config.R"))
config <- read_analysis_config()
artifact_dir <- here::here("artifacts", "supporting", "os-unrestricted-censoring")
result_dir <- here::here("results", "supporting", "os-unrestricted-censoring")
cohort <- arrow::read_parquet(here::here(config_get(config, "paths", "private_data"), "cohort_follow_up.parquet"))
prepared <- prepare_censoring_data(cohort)
result <- analyze_adjusted_censoring(prepared$data, artifact_dir, result_dir,
                                    max_follow_up = max(cohort$potential_follow_up_days))
arrow::write_parquet(prepared$replacements, file.path(result_dir, "modal-replacements.parquet"))
```

## Missing covariates and fitting population

```{r}
knitr::kable(prepared$replacements)
knitr::kable(result$cohort_summary)
```

The pooled modes use all randomized patients. The patient censored at day zero is retained in the empirical curves and standardization population, but excluded from the positive-time survival likelihood. The continuous model cannot represent a point mass of censoring at day zero. Predictions average over the same patients under each assignment and retain fitted diagnosis effects.

## Results

```{r}
#| label: tbl-adjusted-censoring-hr
#| tbl-cap: "Adjusted censoring hazard ratio: invitation versus usual care"
knitr::kable(result$adjusted_summary, digits = 3)
```

An HR above one indicates more rapid censoring in the invitation group. This is conditional on baseline covariates and diagnosis effects. The standardized curves are marginal over the trial population; neither a change in the HR after adjustment nor overlap of curves proves that patient characteristics explain differential censoring. Differences can also reflect non-collapsibility and modeling assumptions. The curves estimate the reverse-survival distribution, not the probability of being alive and under observation; deaths are censored for this diagnostic. Similar adjusted curves would not establish non-informative censoring.

```{r}
#| label: fig-adjusted-reverse-km
#| fig-cap: "Dashed: empirical reverse Kaplan–Meier curves. Solid: standardized adjusted censoring curves with 95% credible bands. Marks represent deaths; table events represent censoring."
#| fig-alt: "Empirical and covariate-adjusted reverse-survival curves for the two randomized groups, with risk counts and cumulative censoring."
result$reverse_plot
```

## Diagnostics and actual sampling settings

```{r}
knitr::kable(result$diagnostics, digits = 3)
knitr::kable(result$checks, digits = 10)
```

```{r}
#| label: fig-adjusted-censoring-traces
#| fig-cap: "Posterior chains for the adjusted censoring model's fixed effects"
result$trace_plot
```
