---
title: "Overall survival censoring scenarios — preparation"
subtitle: "Best and worst cases without covariates or imputation"
format: html
execute:
  freeze: false
---

Both scenarios use unrestricted follow-up and apply the same assumption to
both randomized groups. Observed deaths retain their recorded death times.

- **Best case:** participants censored more than 14 days before their potential
  follow-up ends survive to that endpoint, when they are censored.
- **Worst case:** those same participants die on their recorded censoring day,
  within the week in which they were censored.

Participants without a recorded death who were censored within 14 days of
potential follow-up, including exactly 14 days, are assumed alive to that
endpoint in both scenarios and censored there. Their follow-up is extended
to the existing cutoff; the cutoff itself does not move. This assumption
uses preserved elapsed durations; shifted calendar dates are not linked or
reconstructed. Observed analyses keep their original follow-up times.

Only randomized treatment is included in the models. No covariates are read
or imputed. The original day-zero times remain in the Kaplan–Meier estimates;
day-zero deaths use 0.5 days in the continuous-time model so these participants
remain in its likelihood. Day-zero censoring contributes no follow-up to that
likelihood.

```{r}
#| label: prepare-supporting-os-censoring-scenarios
#| include: false
library(tidyverse)

source(here::here("R", "survival-helpers.R"))

input_path <- here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
)
artifact_dir <- here::here(
  "artifacts", "supporting", "os-censoring-scenarios"
)
result_dir <- here::here(
  "results", "supporting", "os-censoring-scenarios"
)
prepared_path <- file.path(artifact_dir, "scenarios.parquet")
dir.create(artifact_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)

cohort <- arrow::read_parquet(input_path, col_select = c(
  "id", "tx", "survival_time_days_unrestricted",
  "status_death_unrestricted", "potential_follow_up_days"
))
prepared <- prepare_os_censoring_scenarios(cohort)
if (!file.exists(prepared_path) ||
    !identical(prepared, arrow::read_parquet(prepared_path, mmap = FALSE))) {
  arrow::write_parquet(prepared, prepared_path, compression = "zstd")
}

scenario_counts <- prepared |>
  dplyr::summarise(
    participants = dplyr::n(),
    observed_deaths = sum(observed_death),
    originally_censored = sum(observed_death == 0),
    early_censored = sum(observed_death == 0 & !assumed_alive_14_days),
    assumed_alive_14_days = sum(assumed_alive_14_days),
    assumed_deaths = sum(event_death - observed_death),
    scenario_deaths = sum(event_death),
    scenario_censored = sum(event_death == 0),
    day_zero_deaths = sum(survival_time_days == 0 & event_death == 1),
    zero_follow_up_censored = sum(fitting_time_days == 0),
    .by = c(scenario, tx)
  )
arrow::write_parquet(
  scenario_counts, file.path(result_dir, "scenario-counts.parquet"),
  compression = "zstd"
)
```

```{r}
#| label: tbl-supporting-os-censoring-scenarios-counts
#| tbl-cap: "Observed and assumed deaths in each censoring scenario (0: control; 1: intervention)"
#| echo: false

knitr::kable(scenario_counts)
```
