---
title: "Overall survival censoring scenarios — results"
subtitle: "Unadjusted survival curves and treatment hazard ratios"
format: html
execute:
  freeze: false
---

Participants without a recorded death who were censored within 14 days of
potential follow-up (inclusive) are assumed alive to that endpoint and
censored there in both scenarios. For participants censored earlier, the
best case assumes survival to the cutoff and the worst case assigns death
on their recorded censoring day, within the censoring week. Recorded deaths
retain their times. These are scenarios for survival outcomes, not bounds
on the treatment hazard ratio. The data cutoff itself is unchanged.

Each model includes randomized treatment only, without covariates or
imputation. Survival curves use 1,000 posterior draws per scenario; hazard
ratios use all retained posterior draws. The scenarios are summarized
separately. Credible intervals describe posterior uncertainty conditional on
each assumed endpoint.

```{r}
#| label: derive-supporting-os-censoring-scenarios-results
#| include: false
library(tidyverse)

source(here::here("R", "analysis-config.R"))
source(here::here("R", "survival-helpers.R"))

config <- read_analysis_config()
n_draws <- 1000L
artifact_dir <- here::here(
  "artifacts", "supporting", "os-censoring-scenarios"
)
model_dir <- file.path(artifact_dir, "models")
draw_dir <- file.path(artifact_dir, "posterior-draws")
result_dir <- here::here(
  "results", "supporting", "os-censoring-scenarios"
)
figure_dir <- file.path(artifact_dir, "figures")
dir.create(draw_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

prepared <- arrow::read_parquet(file.path(artifact_dir, "scenarios.parquet")) |>
  dplyr::mutate(
    tx_plot = factor(tx, levels = 0:1, labels = c("Control", "Intervention"))
  )
scenario_counts <- arrow::read_parquet(file.path(result_dir, "scenario-counts.parquet"))
scenarios <- c("best-case", "worst-case")
scenario_labels <- c("Best case", "Worst case")
curve_list <- hr_list <- sampling_list <- vector("list", length(scenarios))

for (index in seq_along(scenarios)) {
  scenario_name <- scenarios[[index]]
  model_path <- file.path(model_dir, paste0(scenario_name, "-model.rds"))
  if (file.mtime(model_path) < file.mtime(file.path(artifact_dir, "scenarios.parquet"))) {
    stop("Scenario data changed; rerun 02-fitting.qmd before deriving results.")
  }
  model <- readRDS(model_path)
  scenario_data <- prepared |>
    dplyr::filter(scenario == scenario_name)
  # Treatment is the only predictor: one representative row per prediction
  # gives the same curve as averaging over every participant.
  curve_list[[index]] <- posterior_standardized_survival_draws(
    model, data = model$data[1L, , drop = FALSE],
    horizon = max(scenario_data$survival_time_days),
    n_points = 200L, n_draws = n_draws,
    seed = analysis_seed(config, "supporting", index, offset = 1500L)
  ) |>
    dplyr::mutate(scenario = scenario_name)
  coefficient_draws <- as.matrix(model, pars = "tx")
  hr_list[[index]] <- tibble::tibble(
    scenario = scenario_name,
    draw = seq_len(nrow(coefficient_draws)),
    log_hazard_ratio = coefficient_draws[, 1L],
    hazard_ratio = exp(log_hazard_ratio)
  )
  sampler <- rstan::get_sampler_params(model$stanfit, inc_warmup = FALSE)
  sampling_list[[index]] <- tibble::tibble(
    scenario = scenario_name,
    participants_fitted = nrow(model$data),
    chains = model$stanfit@sim$chains,
    iterations = model$stanfit@sim$iter,
    warmup = model$stanfit@sim$warmup,
    posterior_draws = nrow(coefficient_draws),
    max_rhat = max(model$stan_summary[, "Rhat"], na.rm = TRUE),
    min_ess = min(model$stan_summary[, "n_eff"], na.rm = TRUE),
    divergences = sum(vapply(sampler, function(x) sum(x[, "divergent__"]), numeric(1)))
  )
}

curves <- dplyr::bind_rows(curve_list)
hr_draws <- dplyr::bind_rows(hr_list)
sampling_summary <- dplyr::bind_rows(sampling_list)
curve_summary <- curves |>
  dplyr::summarise(
    median = median(survival),
    lower = quantile(survival, 0.025, names = FALSE),
    upper = quantile(survival, 0.975, names = FALSE),
    .by = c(scenario, tx, time)
  )
hr_summary <- hr_draws |>
  dplyr::summarise(
    median = median(hazard_ratio),
    ci_66_low = quantile(hazard_ratio, 0.17, names = FALSE),
    ci_66_high = quantile(hazard_ratio, 0.83, names = FALSE),
    ci_95_low = quantile(hazard_ratio, 0.025, names = FALSE),
    ci_95_high = quantile(hazard_ratio, 0.975, names = FALSE),
    probability_benefit = mean(hazard_ratio < 1),
    posterior_draws = dplyr::n(),
    .by = scenario
  )

arrow::write_parquet(curves, file.path(draw_dir, "scenario-curves.parquet"), compression = "zstd")
arrow::write_parquet(hr_draws, file.path(draw_dir, "scenario-hazard-ratios.parquet"), compression = "zstd")
arrow::write_parquet(curve_summary, file.path(result_dir, "scenario-survival-curve-summary.parquet"), compression = "zstd")
arrow::write_parquet(hr_summary, file.path(result_dir, "scenario-hazard-ratio-summary.parquet"), compression = "zstd")
arrow::write_parquet(sampling_summary, file.path(result_dir, "scenario-sampling-summary.parquet"), compression = "zstd")
```

```{r}
#| label: build-supporting-os-censoring-scenarios-plots
#| include: false

max_follow_up <- max(prepared$survival_time_days)
risk_times <- seq(0, max_follow_up, by = 100)
survival_plots <- vector("list", length(scenarios))
for (index in seq_along(scenarios)) {
  scenario_name <- scenarios[[index]]
  scenario_data <- prepared |>
    dplyr::filter(scenario == scenario_name)
  plot_curves <- curve_summary |>
    dplyr::filter(scenario == scenario_name) |>
    dplyr::mutate(tx_plot = factor(tx, levels = 0:1, labels = c("Control", "Intervention")))
  km_fit <- ggsurvfit::survfit2(
    survival::Surv(survival_time_days, event_death) ~ tx_plot,
    data = scenario_data, conf.type = "log-log"
  )
  survival_plots[[index]] <- ggsurvfit::ggsurvfit(km_fit, linewidth = 0.55, linetype_aes = FALSE, linetype = 2) +
    ggplot2::geom_ribbon(
      data = plot_curves,
      ggplot2::aes(x = time, ymin = lower, ymax = upper, fill = tx_plot),
      alpha = 0.16, inherit.aes = FALSE
    ) +
    ggplot2::geom_line(
      data = plot_curves, ggplot2::aes(x = time, y = median, colour = tx_plot),
      linewidth = 0.9, inherit.aes = FALSE
    ) +
    ggsurvfit::add_censor_mark(size = 1.8) +
    ggplot2::scale_x_continuous(breaks = risk_times) +
    ggplot2::coord_cartesian(xlim = c(0, max_follow_up), ylim = c(0, 1)) +
    ggplot2::labs(
      title = scenario_labels[[index]], x = "Days since randomization",
      y = "Survival probability", colour = NULL, fill = NULL
    ) +
    ggplot2::theme_bw() +
    ggsurvfit::add_risktable(
      times = risk_times, risktable_stats = c("n.risk", "cum.event"),
      stats_label = list(n.risk = "At risk", cum.event = "Events")
    )
  ggplot2::ggsave(
    file.path(figure_dir, paste0(scenario_name, "-survival.png")),
    ggsurvfit::ggsurvfit_build(survival_plots[[index]]),
    width = 8, height = 6.5, dpi = 300
  )
}

hr_plot <- hr_draws |>
  dplyr::mutate(scenario = factor(scenario, levels = rev(scenarios), labels = rev(scenario_labels))) |>
  ggplot2::ggplot(ggplot2::aes(x = hazard_ratio, y = scenario)) +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  ggplot2::geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  ggplot2::scale_x_log10() +
  ggplot2::labs(x = "Unadjusted hazard ratio, intervention versus control", y = NULL) +
  ggplot2::theme_bw()
ggplot2::ggsave(
  file.path(figure_dir, "scenario-hazard-ratio.png"), hr_plot,
  width = 8, height = 5, dpi = 300
)
```

```{r}
#| label: tbl-supporting-os-censoring-scenarios-results-counts
#| tbl-cap: "Participants, observed deaths, and scenario assumptions (0: control; 1: intervention)"
#| echo: false
knitr::kable(scenario_counts)
```

```{r}
#| label: tbl-supporting-os-censoring-scenarios-hr
#| tbl-cap: "Unadjusted posterior treatment hazard ratios by scenario; probability of benefit is Pr(HR < 1)"
#| echo: false
knitr::kable(hr_summary, digits = 3)
```

```{r}
#| label: fig-supporting-os-censoring-scenarios-best-case
#| fig-cap: "Best case: participants without recorded deaths are assumed alive to the cutoff"
#| fig-alt: "Dashed Kaplan–Meier and solid unadjusted model curves by randomized group, with 95 percent credible bands, censoring marks, and at-risk and event counts."
#| fig-width: 8
#| fig-height: 6.5
#| echo: false
survival_plots[[1L]]
```

```{r}
#| label: fig-supporting-os-censoring-scenarios-worst-case
#| fig-cap: "Worst case: patients censored within 14 days are assumed alive to the cutoff; earlier censoring becomes death"
#| fig-alt: "Dashed Kaplan–Meier and solid unadjusted model curves by randomized group, with 95 percent credible bands. Event counts include deaths assigned on censoring days."
#| fig-width: 8
#| fig-height: 6.5
#| echo: false
survival_plots[[2L]]
```

```{r}
#| label: fig-supporting-os-censoring-scenarios-hr
#| fig-cap: "Unadjusted posterior treatment hazard ratios under the two censoring scenarios"
#| fig-alt: "Separate half-eye distributions for the best- and worst-case intervention-versus-control hazard ratios, with medians, 66 and 95 percent credible intervals, and a reference line at one."
#| fig-width: 8
#| fig-height: 5
#| echo: false
hr_plot
```

```{r}
#| label: tbl-supporting-os-censoring-scenarios-sampling
#| tbl-cap: "Saved model sampling settings and diagnostics by scenario"
#| echo: false
knitr::kable(sampling_summary, digits = 3)
```
