---
title: "Unrestricted overall survival — Adjusted (hierarchical) imputation"
format: html
execute:
  freeze: false
---

This sensitivity analysis replaces the lung indicator with diagnosis random
intercepts. Five MICE datasets use two-level predictive mean matching for
each incomplete baseline covariate, with diagnosis as the grouping variable.
The unrestricted death indicator and Nelson–Aalen cumulative hazard inform
imputation. ECOG uses a numeric working model with observed categories as
donors; binary ECOG and mGPS are derived from the completed measurements.

As in Adjusted (fixed), the randomized cohort enters imputation and
standardization. Participants with zero follow-up cannot enter the survival
likelihood. Endpoint times, death status, treatment, and diagnosis are not imputed.

```{r}
#| label: impute-os-hierarchical
#| include: false
library(tidyverse)
library(mice)
library(miceadds)

source(here::here("R", "analysis-config.R"))
source(here::here("R", "survival-helpers.R"))
config <- read_analysis_config()

artifact_dir <- here::here("artifacts", "supporting", "os-unrestricted-hierarchical")

dir.create(artifact_dir, recursive = TRUE, showWarnings = FALSE)

prepared <- arrow::read_parquet(here::here("analysis", "00-data", "results", "analysis-cohort.parquet")) |>
  select(id, tx, survival_time_days_unrestricted, status_death_unrestricted,
    diagnosis, ecog_fstcnt, stage_binary, age, gender, albumin, bilirubin,
    c_reactive_protein, lactate_dehydrogenase) |>
  mutate(
    diagnosis = factor(diagnosis), 
    diagnosis_id = as.integer(diagnosis), 
    gender = factor(gender))

stopifnot(!anyNA(prepared$diagnosis), !anyDuplicated(prepared$id),
          all(prepared$survival_time_days_unrestricted >= 0))

prepared$na_est <- nelsonaalen(prepared,
  timevar = survival_time_days_unrestricted, statusvar = status_death_unrestricted)

imputation_data <- prepared |> select(-diagnosis)

incomplete <- names(which(colSums(is.na(imputation_data)) > 0))

methods <- make.method(imputation_data)
methods[] <- ""
methods[incomplete] <- "2l.pmm"

predictors <- make.predictorMatrix(imputation_data)
predictors[, c("id", "survival_time_days_unrestricted")] <- 0
predictors[!nzchar(methods), ] <- 0

predictors[incomplete, "diagnosis_id"] <- -2

input_hash <- digest::digest(
  list(
    imputation_data, 
    methods, 
    predictors, 
    m = config$imputation$supporting$m, 
    maxit = config$imputation$supporting$maxit, 
    donors = config$imputation$supporting$donors))

imputation_path <- file.path(artifact_dir, "imputation.rds")

imputation <- if (file.exists(imputation_path)) readRDS(imputation_path) else NULL

if (identical(Sys.getenv("LIQPLAT_FORCE"), "1") || is.null(imputation) ||
    !identical(attr(imputation, "input_hash"), input_hash)) {

  imputation <- mice(
    imputation_data, 
    method = methods, 
    predictorMatrix = predictors,
    m = config$imputation$supporting$m, 
    maxit = config$imputation$supporting$maxit, 
    donors = config$imputation$supporting$donors, 
    seed = 4101L, 
    printFlag = FALSE)

  attr(imputation, "input_hash") <- input_hash
  saveRDS(imputation, imputation_path, compress = "xz")
}

completed_datasets <- lapply(config$imputation$supporting$indices, function(index) {

  completed <- complete(imputation, index) |>
    mutate(diagnosis = prepared$diagnosis, ecog_binary = derive_ecog_binary(ecog_fstcnt),
      mgps = factor(derive_mgps(albumin, c_reactive_protein), levels = 0:2))

  assert_ecog_binary_consistent(completed)
  assert_mgps_consistent(completed)

  stopifnot(!anyNA(completed), all(completed$ecog_fstcnt %in% 0:4), imputation$m == config$imputation$supporting$m,
    identical(completed$survival_time_days_unrestricted, prepared$survival_time_days_unrestricted),
    identical(completed$status_death_unrestricted, prepared$status_death_unrestricted),
    all(imputation$predictorMatrix[incomplete, "diagnosis_id"] == -2))
    
  completed
})

saveRDS(completed_datasets, file.path(artifact_dir, "completed-datasets.rds"))
```

```{r}
#| label: fig-os-hierarchical-imputation-traces
#| fig-cap: "Adjusted (hierarchical): means and standard deviations across five imputations and 50 iterations"
#| fig-alt: "Trace plots for imputed ECOG and laboratory values, with separate traces for each of five imputations."
#| fig-width: 10
#| fig-height: 8
plot(imputation, layout = c(2, 5))
```

```{r}
#| label: fig-os-hierarchical-imputation-distributions
#| fig-cap: "Adjusted (hierarchical): observed and imputed laboratory distributions"
#| fig-alt: "Observed and imputed densities for albumin, bilirubin, CRP, and LDH."
#| fig-width: 10
#| fig-height: 6
densityplot(imputation, ~albumin + bilirubin + c_reactive_protein + lactate_dehydrogenase)
```

```{r}
#| label: tbl-os-hierarchical-imputation-events
if (!is.null(imputation$loggedEvents)) {
  imputation$loggedEvents |> count(dep, meth, out) |> knitr::kable()
}
```
