---
title: "Baseline ctDNA detection"
date: 2026-08-24
format: html
execute:
  echo: true
  warning: false
  message: false
---

Baseline is the earliest valid dated ctDNA sample before treatment starts. If
no treatment start is recorded, the earliest valid dated sample is used.

```{r}
#| label: baseline-detection
library(dplyr)

samples_path <- here::here(
  "analysis", "00-data", "results", "analysis-ctdna-samples.parquet"
)
cohort_path <- here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
)
result_dir <- here::here("analysis", "04-implementation", "results")
summary_path <- file.path(result_dir, "baseline-detection-summary.parquet")
patient_path <- file.path(result_dir, "baseline-detection-patient-status.parquet")
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)

samples <- arrow::read_parquet(samples_path)
cohort <- arrow::read_parquet(cohort_path)

selected_ids <- cohort |>
  filter(tx == 1) |>
  pull(id)

# Earliest valid sample taken before treatment start (or any valid sample if no
# treatment start is recorded); ties on date are broken by sample_id.
baseline <- samples |>
  filter(id %in% selected_ids) |>
  left_join(select(cohort, id, treatment_start_date), by = "id") |>
  filter(
    !is.na(sample_date),
    valid_ctdna_result %in% TRUE,
    is.na(treatment_start_date) | sample_date < treatment_start_date
  ) |>
  arrange(id, sample_date, sample_id) |>
  distinct(id, .keep_all = TRUE) |>
  mutate(ctdna_detected = n_mut > 0)

baseline_summary <- tibble(
  n_selected = length(selected_ids),
  n_accepted = sum(cohort$invitation_accepted, na.rm = TRUE),
  n_valid_baseline = nrow(baseline),
  # n_no_qualifying_sample = sum(is.na(detection)),
  n_detected = sum(baseline$ctdna_detected),
  n_not_detected = n_valid_baseline - n_detected,
  detection_proportion_valid_baseline = n_detected / n_valid_baseline
) |> 
  data.frame()

arrow::write_parquet(baseline_summary, summary_path, compression = "zstd")
```

```{r}
#| label: baseline-summary
#| tbl-cap: "Baseline ctDNA detection summary"
#| echo: false

baseline_summary |>
  knitr::kable()
```

Among selected patients with a valid baseline sample, detection is modeled with
a Bayesian logistic model for histological diagnosis, as specified in the SAP.
Deviating from the SAP's fixed diagnosis coefficients, diagnosis enters as a
random intercept because several diagnoses include only one to three patients.
The probability of detection is averaged over the observed baseline rows,
including each patient's own diagnosis intercept (`re_formula = NULL`).

```{r}
#| label: baseline-detection-model

source(here::here("R", "analysis-config.R"))
source(here::here("R", "baseline-table-helpers.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
figure_dir <- file.path(result_dir, "figures")
detection_model_path <- here::here(
  "analysis", "04-implementation", "models", "baseline-detection.rds"
)
model_summary_path <- file.path(
  result_dir, "baseline-detection-model-summary.parquet"
)
detection_draws_path <- file.path(result_dir, "baseline-detection-draws.parquet")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(detection_model_path), recursive = TRUE, showWarnings = FALSE)

if (force || any(!file.exists(c(
  model_summary_path, detection_draws_path, detection_model_path
)))) {
  model_data <- baseline |>
    left_join(select(cohort, id, diagnosis), by = "id") |>
    mutate(ctdna_detected = as.integer(ctdna_detected)) |>
    select(ctdna_detected, diagnosis)

  detection_fit <- brms::brm(
    ctdna_detected ~ 1 + (1 | diagnosis),
    data = model_data,
    family = brms::bernoulli("logit"),
    iter = 2000L,
    chains = 4L,
    cores = 1L,
    seed = analysis_seed(config, "supporting", offset = 3L),
    refresh = 0L
  )

  draws <- marginaleffects::avg_predictions(
    detection_fit, newdata = model_data, type = "response", re_formula = NULL
  ) |>
    marginaleffects::get_draws() |>
    select(draw)

  detection_model_summary <- data.frame(
    outcome = "baseline_ctdna_detection_probability",
    posterior_mean = mean(draws$draw),
    posterior_median = median(draws$draw),
    lower_95 = quantile(draws$draw, 0.025),
    upper_95 = quantile(draws$draw, 0.975),
    n_model = nrow(model_data),
    n_diagnoses = n_distinct(model_data$diagnosis),
    stringsAsFactors = FALSE
  )

  saveRDS(detection_fit, detection_model_path, compress = "xz")
  arrow::write_parquet(draws, detection_draws_path, compression = "zstd")
  arrow::write_parquet(
    detection_model_summary, model_summary_path, compression = "zstd"
  )
}
detection_model_summary <- arrow::read_parquet(model_summary_path) |>
  as.data.frame()
```

```{r}
#| label: detection-model-summary
#| tbl-cap: "Posterior probability of ctDNA detection at baseline"
#| echo: false

detection_model_summary |>
  knitr::kable()
```

```{r}
#| label: fig-detection-posterior
#| fig-cap: "Posterior probability of ctDNA detection at baseline. Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "A half-eye density above a posterior dot plot of the probability of ctDNA detection at baseline among selected patients with a valid baseline sample, in percent, with thick and thin intervals for the 66 and 95 percent credible intervals."
#| fig-width: 8
#| fig-height: 5

detection_draws <- arrow::read_parquet(detection_draws_path) |>
  as.data.frame()

detection_posterior_plot <- ggplot2::ggplot(
  detection_draws, ggplot2::aes(x = 100 * draw)
) +
  ggdist::stat_dots(side = "bottom") +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  ggplot2::labs(x = "Probability of ctDNA detection at baseline, %", y = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor.y = ggplot2::element_blank()
  )
ggplot2::ggsave(file.path(figure_dir, "baseline-detection-posterior.png"),
  detection_posterior_plot, width = 8, height = 5, dpi = 300)
detection_posterior_plot
```

Predictions by diagnosis are averaged over the observed rows within each
diagnosis using `by = "diagnosis"`. Diagnoses with few patients are pulled
toward the overall detection probability by the random intercept.

```{r}
#| label: detection-diagnosis-predictions

detection_fit <- readRDS(detection_model_path)
detection_fit <- marginaleffects::set_modeldata(
  detection_fit, detection_fit$data
)

diagnosis_counts <- detection_fit$data |>
  count(diagnosis, name = "n_patients")

diagnosis_draws <- marginaleffects::avg_predictions(
  detection_fit, by = "diagnosis", type = "response", re_formula = NULL
) |>
  marginaleffects::get_draws() |>
  select(diagnosis, drawid, draw)

diagnosis_summary <- diagnosis_draws |>
  group_by(diagnosis) |>
  summarise(
    posterior_mean = mean(draw),
    posterior_median = median(draw),
    lower_95 = quantile(draw, 0.025),
    upper_95 = quantile(draw, 0.975),
    .groups = "drop"
  ) |>
  left_join(diagnosis_counts, by = "diagnosis") |>
  mutate(diagnosis = snomed_to_diagnosis(diagnosis)) |>
  arrange(desc(posterior_median))

arrow::write_parquet(
  diagnosis_draws,
  file.path(result_dir, "baseline-detection-diagnosis-draws.parquet"),
  compression = "zstd"
)

arrow::write_parquet(
  diagnosis_summary,
  file.path(result_dir, "baseline-detection-diagnosis-summary.parquet"),
  compression = "zstd"
)
```

```{r}
#| label: tbl-detection-diagnosis-summary
#| tbl-cap: "Posterior probability of ctDNA detection at baseline by diagnosis"
#| echo: false

diagnosis_summary |>
  knitr::kable()
```

```{r}
#| label: fig-detection-diagnosis-predictions
#| fig-cap: "Posterior probability of ctDNA detection at baseline by diagnosis, ordered by posterior median; n is the number of patients with a valid baseline sample. Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "Posterior half-eye densities of the probability of ctDNA detection at baseline, in percent, one row per histological diagnosis ordered by posterior median, with thick and thin intervals for the 66 and 95 percent credible intervals."
#| fig-width: 8
#| fig-height: 9

diagnosis_plot_draws <- diagnosis_draws |>
  left_join(diagnosis_counts, by = "diagnosis") |>
  mutate(
    diagnosis_label = paste0(
      snomed_to_diagnosis(diagnosis), " (n = ", n_patients, ")"
    ),
    diagnosis_label = forcats::fct_reorder(diagnosis_label, draw, .fun = median)
  )

diagnosis_predictions_plot <- ggplot2::ggplot(
  diagnosis_plot_draws, ggplot2::aes(x = 100 * draw, y = diagnosis_label)
) +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  ggplot2::labs(x = "Probability of ctDNA detection at baseline, %", y = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor.y = ggplot2::element_blank()
  )
ggplot2::ggsave(file.path(figure_dir, "baseline-detection-diagnosis-predictions.png"),
  diagnosis_predictions_plot, width = 8, height = 9, dpi = 300)
diagnosis_predictions_plot
```

## Ordinal model for the number of mutations

As an additional analysis, the number of mutations in the baseline sample is
modeled with a Bayesian proportional odds (cumulative logit) model with the
same diagnosis random intercept. Each observed mutation count is one ordered
category, so the model makes no assumption about the shape of the count
distribution. If some diagnoses shed more ctDNA, their whole count
distribution should shift upwards, which the proportional odds model uses
through the full ordinal outcome rather than only its zero category. From the
posterior category probabilities $p_k$, averaged over the observed baseline
rows with each patient's own diagnosis intercept, we derive the probability of
detection, $1 - p_0$, and the expected number of mutations, $\sum_k k \, p_k$.

```{r}
#| label: baseline-mutations-ordinal-model

ordinal_model_path <- here::here(
  "analysis", "04-implementation", "models", "baseline-mutations-ordinal.rds"
)
ordinal_draws_path <- file.path(
  result_dir, "baseline-mutations-ordinal-draws.parquet"
)
ordinal_summary_path <- file.path(
  result_dir, "baseline-mutations-ordinal-summary.parquet"
)
ordinal_diagnosis_draws_path <- file.path(
  result_dir, "baseline-mutations-ordinal-diagnosis-draws.parquet"
)
ordinal_diagnosis_summary_path <- file.path(
  result_dir, "baseline-mutations-ordinal-diagnosis-summary.parquet"
)
ordinal_paths <- c(
  ordinal_model_path, ordinal_draws_path, ordinal_summary_path,
  ordinal_diagnosis_draws_path, ordinal_diagnosis_summary_path
)

# Per-draw derived quantities from the category probabilities.
summarise_mutation_draws <- function(category_draws, by = character()) {
  category_draws |>
    mutate(n_mut = as.numeric(as.character(group))) |>
    summarise(
      detection_probability = 1 - sum(draw[n_mut == 0]),
      expected_mutations = sum(n_mut * draw),
      .by = dplyr::all_of(c(by, "drawid"))
    ) |>
    tidyr::pivot_longer(
      c(detection_probability, expected_mutations),
      names_to = "quantity", values_to = "draw"
    )
}

summarise_posterior <- function(draws, by) {
  draws |>
    summarise(
      posterior_mean = mean(draw),
      posterior_median = median(draw),
      lower_95 = quantile(draw, 0.025),
      upper_95 = quantile(draw, 0.975),
      .by = dplyr::all_of(by)
    )
}

if (force || any(!file.exists(ordinal_paths))) {
  ordinal_data <- baseline |>
    left_join(select(cohort, id, diagnosis), by = "id") |>
    mutate(n_mut = factor(n_mut, levels = sort(unique(n_mut)), ordered = TRUE)) |>
    select(n_mut, diagnosis)

  ordinal_fit <- brms::brm(
    n_mut ~ 1 + (1 | diagnosis),
    data = ordinal_data,
    family = brms::cumulative("logit"),
    iter = 2000L,
    chains = 4L,
    cores = 1L,
    seed = analysis_seed(config, "supporting", offset = 4L),
    refresh = 0L
  )

  ordinal_draws <- marginaleffects::avg_predictions(
    ordinal_fit, newdata = ordinal_data, type = "response", re_formula = NULL
  ) |>
    marginaleffects::get_draws() |>
    select(group, drawid, draw) |>
    summarise_mutation_draws()

  ordinal_diagnosis_draws <- marginaleffects::avg_predictions(
    ordinal_fit, newdata = ordinal_data, by = "diagnosis",
    type = "response", re_formula = NULL
  ) |>
    marginaleffects::get_draws() |>
    select(group, diagnosis, drawid, draw) |>
    summarise_mutation_draws("diagnosis")

  ordinal_summary <- ordinal_draws |>
    summarise_posterior("quantity") |>
    mutate(n_model = nrow(ordinal_data))

  ordinal_diagnosis_summary <- ordinal_diagnosis_draws |>
    summarise_posterior(c("quantity", "diagnosis")) |>
    left_join(
      count(ordinal_data, diagnosis, name = "n_patients"), by = "diagnosis"
    ) |>
    mutate(diagnosis = snomed_to_diagnosis(diagnosis)) |>
    arrange(quantity, desc(posterior_median))

  saveRDS(ordinal_fit, ordinal_model_path, compress = "xz")
  arrow::write_parquet(ordinal_draws, ordinal_draws_path, compression = "zstd")
  arrow::write_parquet(
    ordinal_summary, ordinal_summary_path, compression = "zstd"
  )
  arrow::write_parquet(
    ordinal_diagnosis_draws, ordinal_diagnosis_draws_path, compression = "zstd"
  )
  arrow::write_parquet(
    ordinal_diagnosis_summary, ordinal_diagnosis_summary_path,
    compression = "zstd"
  )
}
ordinal_draws <- arrow::read_parquet(ordinal_draws_path)
ordinal_summary <- arrow::read_parquet(ordinal_summary_path)
ordinal_diagnosis_draws <- arrow::read_parquet(ordinal_diagnosis_draws_path)
ordinal_diagnosis_summary <- arrow::read_parquet(ordinal_diagnosis_summary_path)
```

```{r}
#| label: tbl-ordinal-summary
#| tbl-cap: "Posterior probability of detection and expected number of mutations at baseline from the proportional odds model"
#| echo: false

ordinal_summary |>
  knitr::kable()
```

```{r}
#| label: fig-ordinal-posterior
#| fig-cap: "Posterior probability of ctDNA detection and expected number of mutations at baseline from the proportional odds model. Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "Two panels of half-eye densities above posterior dot plots from the proportional odds model: the probability of ctDNA detection at baseline in percent, and the expected number of mutations in the baseline sample, each with thick and thin intervals for the 66 and 95 percent credible intervals."
#| fig-width: 8
#| fig-height: 7

ordinal_quantity_labels <- c(
  detection_probability = "Probability of ctDNA detection at baseline, %",
  expected_mutations = "Expected number of mutations at baseline"
)

ordinal_posterior_plot <- ordinal_draws |>
  mutate(draw = if_else(quantity == "detection_probability", 100 * draw, draw)) |>
  ggplot2::ggplot(ggplot2::aes(x = draw)) +
  ggdist::stat_dots(side = "bottom") +
  ggdist::stat_halfeye(
    .width = c(0.66, 0.95), point_interval = ggdist::median_qi,
    normalize = "panels"
  ) +
  ggplot2::facet_wrap(
    ~ quantity, ncol = 1, scales = "free_x",
    labeller = ggplot2::as_labeller(ordinal_quantity_labels)
  ) +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor.y = ggplot2::element_blank()
  )
ggplot2::ggsave(file.path(figure_dir, "baseline-mutations-ordinal-posterior.png"),
  ordinal_posterior_plot, width = 8, height = 7, dpi = 300)
ordinal_posterior_plot
```

```{r}
#| label: tbl-ordinal-diagnosis-summary
#| tbl-cap: "Posterior probability of detection and expected number of mutations at baseline by diagnosis from the proportional odds model"
#| echo: false

ordinal_diagnosis_summary |>
  knitr::kable()
```

```{r}
#| label: fig-ordinal-diagnosis-predictions
#| fig-cap: "Posterior expected number of mutations at baseline by diagnosis from the proportional odds model, ordered by posterior median; n is the number of patients with a valid baseline sample. Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "Posterior half-eye densities of the expected number of mutations in the baseline sample, one row per histological diagnosis ordered by posterior median, with thick and thin intervals for the 66 and 95 percent credible intervals."
#| fig-width: 8
#| fig-height: 9

ordinal_diagnosis_plot <- ordinal_diagnosis_draws |>
  filter(quantity == "expected_mutations") |>
  left_join(diagnosis_counts, by = "diagnosis") |>
  mutate(
    diagnosis_label = paste0(
      snomed_to_diagnosis(diagnosis), " (n = ", n_patients, ")"
    ),
    diagnosis_label = forcats::fct_reorder(diagnosis_label, draw, .fun = median)
  ) |>
  ggplot2::ggplot(ggplot2::aes(x = draw, y = diagnosis_label)) +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  ggplot2::labs(x = "Expected number of mutations at baseline", y = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor.y = ggplot2::element_blank()
  )
ggplot2::ggsave(
  file.path(figure_dir, "baseline-mutations-ordinal-diagnosis-predictions.png"),
  ordinal_diagnosis_plot, width = 8, height = 9, dpi = 300
)
ordinal_diagnosis_plot
```
