---
title: "ctDNA alteration actionability"
date: 2026-08-20
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

Mutation, CNV, and fusion records are retained in a long actionability table.
Exact sensitivity and resistance levels are counted without ranking or
collapsing source labels. Alteration counts exclude CHIP records. Patient
proportions use all patients with at least one valid ctDNA result as the
denominator, allow a patient to occupy multiple levels, and explicitly report
no actionable finding when no non-CHIP alteration has recorded evidence.

```{r}
#| label: setup
#| include: false

source(here::here("analysis", "_module-common.R"), local = FALSE)
liqplat_source_public_helpers()
config <- read_analysis_config()
force <- module_force()
stage_id <- "implementation_actionability"
result_dir <- module_result_dir(file.path("analysis", "04-implementation"))
checkpoint_path <- module_checkpoint_path(stage_id)
long_path <- file.path(result_dir, "actionability-long.parquet")
level_path <- file.path(result_dir, "actionability-level-summary.parquet")
patient_path <- file.path(result_dir, "actionability-patient-summary.parquet")
summary_path <- file.path(result_dir, "actionability-summary.parquet")
figure_path <- file.path(result_dir, "actionability-levels.svg")
alterations_path <- file.path(module_result_dir(file.path("analysis", "00-data")),
                              "analysis-alterations.parquet")
ctdna_path <- file.path(module_result_dir(file.path("analysis", "00-data")),
                        "analysis-ctdna-samples.parquet")
if (!file.exists(alterations_path) || !file.exists(ctdna_path)) {
  stop("Actionability analysis requires the built ctDNA and alteration tables.")
}
input_paths <- c(alterations_path, ctdna_path,
                 private_domain_path(config, "ctdna_samples"),
                 private_domain_path(config, "alterations"),
                 schema_domain_path(config, "ctdna_samples"),
                 schema_domain_path(config, "alterations"))
input_paths <- input_paths[file.exists(input_paths)]
output_paths <- c(long_path, level_path, patient_path, summary_path, figure_path)
decision <- checkpoint_decision(
  checkpoint_path, stage_id, input_paths = input_paths,
  output_paths = output_paths,
  code_paths = here::here("analysis", "04-implementation", "06-actionability.qmd"),
  config = config, force = force
)

if (identical(decision, "skip")) {
  actionability_summary <- read_result_table(summary_path)
  actionability_levels <- read_result_table(level_path)
} else {
  alterations <- read_result_table(alterations_path)
  ctdna <- read_result_table(ctdna_path)
  require_public_columns(
    ctdna, c("id", "sample_id", "valid_ctdna_result"),
    "valid ctDNA-result denominator"
  )
  require_public_columns(
    alterations,
    c("id", "alteration_id", "sensitivity", "resistance", "chip_suspicion",
      "valid_ctdna_result"),
    "actionability analysis"
  )
  type_col <- intersect(c("alteration_type", "alteration_class", "variant_type"),
                        names(alterations))[1L]
  if (is.na(type_col)) {
    stop("Actionability requires an alteration_type column identifying mutation, CNV, or fusion.")
  }
  valid_values <- alterations$valid_ctdna_result
  valid_ctdna <- !is.na(ctdna$valid_ctdna_result) &
    as.logical(ctdna$valid_ctdna_result) & !is.na(ctdna$id)
  if (!any(valid_ctdna)) {
    stop("Actionability is pending because no valid ctDNA-result patient is recorded.")
  }
  chip <- validate_binary(alterations$chip_suspicion, "chip_suspicion")
  valid <- !is.na(valid_values) & as.logical(valid_values) &
    !is.na(alterations$id)
  valid_patient_ids <- unique(ctdna$id[valid_ctdna])
  denominator <- length(valid_patient_ids)
  type_values <- tolower(trimws(as.character(alterations[[type_col]])))
  type_values[type_values %in% c("snv", "snp", "indel", "small_variant")] <- "mutation"
  type_values[type_values %in% c("copy_number", "copy number", "copy-number")] <- "cnv"
  type_values[type_values %in% c("structural_variant", "structural variant")] <- "fusion"
  if (nrow(alterations) > 0L &&
      !any(type_values %in% c("mutation", "cnv", "fusion"), na.rm = TRUE)) {
    stop("Actionability alteration_type has no mutation, CNV, or fusion records.")
  }
  sensitivity <- trimws(as.character(alterations$sensitivity))
  resistance <- trimws(as.character(alterations$resistance))
  long <- rbind(
    data.frame(
      id = alterations$id,
      alteration_id = alterations$alteration_id,
      alteration_type = type_values,
      actionability_type = "sensitivity",
      actionability_level = sensitivity,
      recorded_exact_level = !is.na(sensitivity) & nzchar(sensitivity),
      non_chip = !is.na(chip) & chip == 0,
      valid_ctdna_result = valid_values,
      stringsAsFactors = FALSE
    ),
    data.frame(
      id = alterations$id,
      alteration_id = alterations$alteration_id,
      alteration_type = type_values,
      actionability_type = "resistance",
      actionability_level = resistance,
      recorded_exact_level = !is.na(resistance) & nzchar(resistance),
      non_chip = !is.na(chip) & chip == 0,
      valid_ctdna_result = valid_values,
      stringsAsFactors = FALSE
    )
  )
  long <- long[!is.na(long$valid_ctdna_result) &
                 as.logical(long$valid_ctdna_result) & long$non_chip &
                 long$alteration_type %in% c("mutation", "cnv", "fusion"), ,
               drop = FALSE]
  long_evidence <- long[long$recorded_exact_level, , drop = FALSE]
  level_key <- paste(long_evidence$id, long_evidence$alteration_id,
                     long_evidence$actionability_type,
                     long_evidence$actionability_level, sep = "\r")
  long_unique <- long_evidence[!duplicated(level_key), , drop = FALSE]
  level_groups <- split(
    seq_len(nrow(long_unique)),
    interaction(long_unique$actionability_type, long_unique$actionability_level,
                drop = TRUE, lex.order = TRUE)
  )
  if (nrow(long_unique) == 0L) {
    actionability_levels <- data.frame(
      actionability_type = character(), actionability_level = character(),
      alteration_count = integer(), patient_count = integer(),
      patient_proportion = numeric(), stringsAsFactors = FALSE
    )
  } else {
    actionability_levels <- do.call(rbind, lapply(level_groups, function(index) {
      values <- long_unique[index, , drop = FALSE]
      patient_count <- length(unique(values$id))
      data.frame(
        actionability_type = values$actionability_type[[1L]],
        actionability_level = values$actionability_level[[1L]],
        alteration_count = nrow(values),
        patient_count = patient_count,
        patient_proportion = if (denominator == 0L) NA_real_ else
          patient_count / denominator,
        stringsAsFactors = FALSE
      )
    }))
    rownames(actionability_levels) <- NULL
  }
  patient_rows <- lapply(valid_patient_ids, function(identifier) {
    patient_rows <- valid & alterations$id == identifier & !is.na(chip) & chip == 0 &
      type_values %in% c("mutation", "cnv", "fusion")
    sensitivity_evidence <- any(is_recorded_actionability_evidence(
      alterations$sensitivity[patient_rows]
    ))
    resistance_evidence <- any(is_recorded_actionability_evidence(
      alterations$resistance[patient_rows]
    ))
    data.frame(
      id = identifier,
      any_sensitivity = sensitivity_evidence,
      any_resistance = resistance_evidence,
      any_actionable = sensitivity_evidence || resistance_evidence,
      neither = !sensitivity_evidence && !resistance_evidence,
      no_actionable_finding = !sensitivity_evidence && !resistance_evidence,
      stringsAsFactors = FALSE
    )
  })
  patient_summary <- if (length(patient_rows) == 0L) {
    data.frame(
      id = alterations$id[FALSE], any_sensitivity = logical(),
      any_resistance = logical(), any_actionable = logical(),
      neither = logical(),
      no_actionable_finding = logical(), stringsAsFactors = FALSE
    )
  } else {
    do.call(rbind, patient_rows)
  }
  write_module_table(long, long_path)
  write_module_table(actionability_levels, level_path)
  write_module_table(patient_summary, patient_path)
  actionability_summary <- data.frame(
    denominator_valid_ctdna_patients = nrow(patient_summary),
    patients_any_sensitivity = sum(patient_summary$any_sensitivity),
    patients_any_resistance = sum(patient_summary$any_resistance),
    patients_any_actionable = sum(patient_summary$any_actionable),
    patients_neither = sum(patient_summary$neither),
    patients_no_actionable_finding = sum(patient_summary$no_actionable_finding),
    stringsAsFactors = FALSE
  )
  write_module_table(actionability_summary, summary_path)
  if (!requireNamespace("ggplot2", quietly = TRUE)) {
    stop("Package 'ggplot2' is required for the actionability figure.")
  }
  plot <- ggplot2::ggplot(
    actionability_levels,
    ggplot2::aes(x = actionability_level, y = alteration_count,
                 fill = actionability_type)
  ) +
    ggplot2::geom_col(position = "dodge") +
    ggplot2::coord_flip() +
    ggplot2::labs(x = "Exact recorded level", y = "Alteration count", fill = NULL) +
    ggplot2::theme_minimal()
  assert_artifact_path(figure_path)
  ggplot2::ggsave(figure_path, plot = plot, width = 8, height = 5, units = "in")
  finalize_module_checkpoint(
    stage_id = stage_id,
    checkpoint_path = checkpoint_path,
    input_paths = input_paths,
    output_paths = output_paths,
    config = config,
    metadata = list(data_lock = config_get(config, "analysis", "data_lock"),
                    status = "built", denominator = "patients_with_valid_ctdna"),
    code_paths = here::here("analysis", "04-implementation", "06-actionability.qmd")
  )
}
```

```{r}
#| label: actionability-summary
#| tbl-cap: "Patient-level actionability summary"

actionability_summary |>
  knitr::kable()
```

```{r}
#| label: actionability-levels
#| tbl-cap: "Exact alteration-level actionability counts"

actionability_levels |>
  knitr::kable()
```
