---
title: "ctDNA alteration actionability"
date: 2026-08-24
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

Mutation, CNV, and fusion records are retained in a long actionability table.
Exact sensitivity and resistance levels are counted without collapsing source
labels. The published OncoKB order is displayed and explained, but it is not
converted to a numeric analysis variable. All seven current levels are shown,
including zero-count levels, and any additional exact recorded strings are
retained as separate rows. Actionability is repeated under two assumptions for
ctDNA-only mutations whose buffy-coat sequencing is missing. Mutations found in
both tissue and buffy coat are kept outside both assumptions as possible
germline variants. Patient proportions use all patients with at least one valid
ctDNA result as the denominator. No regression model is fitted.

```{r}
#| label: setup
#| include: false

source(here::here("R", "actionability.R"))
source(here::here("R", "data-helpers.R"))
result_dir <- here::here("analysis", "04-implementation", "results")
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)
long_path <- file.path(result_dir, "actionability-long.parquet")
level_path <- file.path(result_dir, "actionability-level-summary.parquet")
patient_path <- file.path(result_dir, "actionability-patient-summary.parquet")
summary_path <- file.path(result_dir, "actionability-summary.parquet")
qc_path <- file.path(result_dir, "actionability-level-completeness.parquet")
alterations_path <- here::here(
  "analysis", "00-data", "results", "analysis-alterations.parquet"
)
ctdna_path <- here::here(
  "analysis", "00-data", "results", "analysis-ctdna-samples.parquet"
)
alterations <- arrow::read_parquet(alterations_path) |>
  as.data.frame()
ctdna <- arrow::read_parquet(ctdna_path) |>
  as.data.frame()
valid_values <- alterations$valid_ctdna_result
valid_ctdna <- !is.na(ctdna$valid_ctdna_result) &
  as.logical(ctdna$valid_ctdna_result) & !is.na(ctdna$id)
if (!any(valid_ctdna)) {
  stop("Actionability is pending because no valid ctDNA-result patient is recorded.")
}
chip_classification <- classify_chip_status(alterations)
valid_patient_ids <- unique(ctdna$id[valid_ctdna])
denominator <- length(valid_patient_ids)
type_values <- tolower(trimws(as.character(alterations$alteration_type)))
type_values[type_values %in% c("snv", "snp", "indel", "small_variant")] <- "mutation"
type_values[type_values %in% c("copy_number", "copy number", "copy-number")] <- "cnv"
type_values[type_values %in% c("structural_variant", "structural variant")] <- "fusion"
if (nrow(alterations) > 0L &&
    !any(type_values %in% c("mutation", "cnv", "fusion"), na.rm = TRUE)) {
  stop("Actionability alteration_type has no mutation, CNV, or fusion records.")
}
assumptions <- c(
  "Assume ctDNA-only unknowns are not CHIP",
  "Assume ctDNA-only unknowns are CHIP"
)
  record_index <- rep(seq_len(nrow(alterations)), times = length(assumptions))
  assumption <- rep(assumptions, each = nrow(alterations))
  scenario_type <- type_values[record_index]
  scenario_chip_class <- chip_classification[record_index]
  scenario_unknown_is_chip <- assumption == assumptions[[2L]]
  non_chip <- ifelse(
    scenario_type == "mutation",
    scenario_chip_class %in% c(
      "definite_non_chip", "tissue_positive_buffy_missing"
    ) | (!scenario_unknown_is_chip &
           scenario_chip_class == "ctdna_only_unknown"),
    scenario_type %in% c("cnv", "fusion")
  )
  scenario_records <- data.frame(
    id = alterations$id[record_index],
    alteration_id = alterations$alteration_id[record_index],
    assumption = assumption,
    alteration_type = scenario_type,
    actionability_sensitivity = trimws(as.character(
      alterations$sensitivity[record_index]
    )),
    actionability_resistance = trimws(as.character(
      alterations$resistance[record_index]
    )),
    onkb_version = alterations$onkb_version[record_index],
    onkb_updated_date = alterations$onkb_updated_date[record_index],
    valid_ctdna_result = valid_values[record_index],
    non_chip = non_chip,
    stringsAsFactors = FALSE
  )
  long <- rbind(
    data.frame(
      id = scenario_records$id,
      alteration_id = scenario_records$alteration_id,
      assumption = scenario_records$assumption,
      alteration_type = scenario_records$alteration_type,
      actionability_type = "sensitivity",
      actionability_level = scenario_records$actionability_sensitivity,
      onkb_version = scenario_records$onkb_version,
      onkb_updated_date = scenario_records$onkb_updated_date,
      recorded_exact_level = !is.na(scenario_records$actionability_sensitivity) &
        nzchar(scenario_records$actionability_sensitivity),
      non_chip = scenario_records$non_chip,
      valid_ctdna_result = scenario_records$valid_ctdna_result,
      stringsAsFactors = FALSE
    ),
    data.frame(
      id = scenario_records$id,
      alteration_id = scenario_records$alteration_id,
      assumption = scenario_records$assumption,
      alteration_type = scenario_records$alteration_type,
      actionability_type = "resistance",
      actionability_level = scenario_records$actionability_resistance,
      onkb_version = scenario_records$onkb_version,
      onkb_updated_date = scenario_records$onkb_updated_date,
      recorded_exact_level = !is.na(scenario_records$actionability_resistance) &
        nzchar(scenario_records$actionability_resistance),
      non_chip = scenario_records$non_chip,
      valid_ctdna_result = scenario_records$valid_ctdna_result,
      stringsAsFactors = FALSE
    )
  )
long <- long[!is.na(long$valid_ctdna_result) &
               as.logical(long$valid_ctdna_result) & long$non_chip &
               long$alteration_type %in% c("mutation", "cnv", "fusion"), ,
             drop = FALSE]
  evidence_flag <- is_recorded_actionability_evidence(long$actionability_level)
  long_evidence <- long[evidence_flag, , drop = FALSE]
  long_non_evidence <- long[!evidence_flag, , drop = FALSE]
  if (nrow(long_non_evidence) == 0L) {
    actionability_qc <- data.frame(
      assumption = character(), evidence_direction = character(),
      recorded_value = character(),
      alteration_records_n = integer(), patients_n = integer(),
      stringsAsFactors = FALSE
    )
  } else {
    qc_value <- trimws(as.character(long_non_evidence$actionability_level))
    qc_value[is.na(qc_value)] <- "<missing>"
    qc_value[!is.na(qc_value) & !nzchar(qc_value)] <- "<blank>"
    qc_groups <- split(
      seq_len(nrow(long_non_evidence)),
      interaction(long_non_evidence$assumption,
                  long_non_evidence$actionability_type, qc_value,
                  drop = TRUE, lex.order = TRUE)
    )
    actionability_qc <- do.call(rbind, lapply(qc_groups, function(index) {
      values <- long_non_evidence[index, , drop = FALSE]
      data.frame(
        assumption = values$assumption[[1L]],
        evidence_direction = values$actionability_type[[1L]],
        recorded_value = qc_value[[index[[1L]]]],
        alteration_records_n = nrow(values),
        patients_n = length(unique(values$id)),
        stringsAsFactors = FALSE
      )
    }))
    rownames(actionability_qc) <- NULL
  }
  level_key <- paste(long_evidence$assumption, long_evidence$id,
                     long_evidence$alteration_id,
                     long_evidence$actionability_type,
                     long_evidence$actionability_level, sep = "\r")
  long_unique <- long_evidence[!duplicated(level_key), , drop = FALSE]
  current_levels <- data.frame(
    actionability_type = c(rep("sensitivity", 5L), rep("resistance", 2L)),
    actionability_level = c("1", "2", "3A", "3B", "4", "R1", "R2"),
    onkb_category = c(
      "Standard care", "Standard care", "Investigational",
      "Investigational", "Hypothetical", "Standard-care resistance",
      "Investigational resistance"
    ),
    within_direction_position = c(
      "1 (strongest)", "2", "3", "4", "5 (weakest)",
      "1 (strongest)", "2"
    ),
    level_meaning = c(
      "FDA-recognized biomarker predictive of response to an FDA-approved drug in the same indication",
      "Standard-care biomarker recommended by the NCCN or other professional guidelines and predictive of response to an FDA-approved drug in the same indication",
      "Compelling clinical evidence supports the biomarker as predictive of response to a drug in the same indication",
      "Standard-care or investigational biomarker predictive of response to an FDA-approved or investigational drug in another indication",
      "Compelling biological evidence supports the biomarker as predictive of response to a drug",
      "Standard-care biomarker predictive of resistance to an FDA-approved drug in the same indication",
      "Compelling clinical evidence supports the biomarker as predictive of resistance to a drug"
    ),
    stringsAsFactors = FALSE
  )
  level_groups <- split(
    seq_len(nrow(long_unique)),
    interaction(long_unique$assumption, long_unique$actionability_type,
                long_unique$actionability_level,
                drop = TRUE, lex.order = TRUE)
  )
  if (nrow(long_unique) == 0L) {
    actionability_levels <- data.frame(
      assumption = character(), actionability_type = character(),
      actionability_level = character(),
      onkb_category = character(), within_direction_position = character(),
      level_meaning = character(), onkb_versions = character(),
      onkb_update_range = character(), mutation_count = integer(),
      mutation_patient_count = integer(), cnv_count = integer(),
      cnv_patient_count = integer(), fusion_count = integer(),
      fusion_patient_count = integer(), alteration_count = integer(),
      patient_count = integer(),
      denominator_valid_ctdna_patients = integer(),
      patient_proportion = numeric(), stringsAsFactors = FALSE
    )
  } else {
    actionability_levels <- do.call(rbind, lapply(level_groups, function(index) {
      values <- long_unique[index, , drop = FALSE]
      patient_count <- length(unique(values$id))
      level_code <- toupper(trimws(as.character(values$actionability_level[[1L]])))
      level_code <- sub("^LEVEL[ _-]*", "", level_code)
      definition_row <- current_levels[current_levels$actionability_level == level_code &
                                         current_levels$actionability_type ==
                                           values$actionability_type[[1L]], ,
                                       drop = FALSE]
      known_current_level <- nrow(definition_row) == 1L
      recorded_versions <- sort(unique(trimws(as.character(values$onkb_version))))
      recorded_versions <- recorded_versions[
        !is.na(recorded_versions) & nzchar(recorded_versions)
      ]
      update_dates <- as.Date(values$onkb_updated_date)
      update_dates <- update_dates[!is.na(update_dates)]
      data.frame(
        assumption = values$assumption[[1L]],
        actionability_type = values$actionability_type[[1L]],
        actionability_level = values$actionability_level[[1L]],
        onkb_category = if (known_current_level) {
          definition_row$onkb_category[[1L]]
        } else {
          "Version-specific or non-canonical recorded value"
        },
        within_direction_position = if (known_current_level) {
          definition_row$within_direction_position[[1L]]
        } else {
          NA_character_
        },
        level_meaning = if (known_current_level) {
          definition_row$level_meaning[[1L]]
        } else {
          "Review the exact recorded value against the recorded OncoKB version"
        },
        onkb_versions = if (length(recorded_versions) == 0L) {
          NA_character_
        } else {
          paste(recorded_versions, collapse = ", ")
        },
        onkb_update_range = if (length(update_dates) == 0L) {
          NA_character_
        } else {
          paste(format(range(update_dates), "%Y-%m-%d"), collapse = " to ")
        },
        mutation_count = sum(values$alteration_type == "mutation"),
        mutation_patient_count = length(unique(
          values$id[values$alteration_type == "mutation"]
        )),
        cnv_count = sum(values$alteration_type == "cnv"),
        cnv_patient_count = length(unique(
          values$id[values$alteration_type == "cnv"]
        )),
        fusion_count = sum(values$alteration_type == "fusion"),
        fusion_patient_count = length(unique(
          values$id[values$alteration_type == "fusion"]
        )),
        alteration_count = nrow(values),
        patient_count = patient_count,
        denominator_valid_ctdna_patients = denominator,
        patient_proportion = if (denominator == 0L) NA_real_ else
          patient_count / denominator,
        stringsAsFactors = FALSE
      )
    }))
    rownames(actionability_levels) <- NULL
  }
  canonical_levels <- do.call(rbind, lapply(assumptions, function(value) {
    data.frame(assumption = value, current_levels, stringsAsFactors = FALSE)
  }))
  canonical_keys <- paste(canonical_levels$assumption,
                          canonical_levels$actionability_type,
                          canonical_levels$actionability_level, sep = "\r")
  observed_level_codes <- toupper(trimws(actionability_levels$actionability_level))
  observed_level_codes <- sub("^LEVEL[ _-]*", "", observed_level_codes)
  observed_keys <- paste(
    actionability_levels$assumption,
    actionability_levels$actionability_type,
    observed_level_codes, sep = "\r"
  )
  missing_current <- canonical_levels[!canonical_keys %in% observed_keys, , drop = FALSE]
  if (nrow(missing_current) > 0L) {
    zero_rows <- data.frame(
      assumption = missing_current$assumption,
      actionability_type = missing_current$actionability_type,
      actionability_level = paste0("LEVEL_", missing_current$actionability_level),
      onkb_category = missing_current$onkb_category,
      within_direction_position = missing_current$within_direction_position,
      level_meaning = missing_current$level_meaning,
      onkb_versions = NA_character_, onkb_update_range = NA_character_,
      mutation_count = 0L, mutation_patient_count = 0L,
      cnv_count = 0L, cnv_patient_count = 0L,
      fusion_count = 0L, fusion_patient_count = 0L,
      alteration_count = 0L, patient_count = 0L,
      denominator_valid_ctdna_patients = denominator,
      patient_proportion = 0,
      stringsAsFactors = FALSE
    )
    actionability_levels <- rbind(actionability_levels, zero_rows)
  }
  output_level_codes <- toupper(trimws(actionability_levels$actionability_level))
  output_level_codes <- sub("^LEVEL[ _-]*", "", output_level_codes)
  output_keys <- paste(actionability_levels$actionability_type,
                       output_level_codes, sep = "\r")
  output_keys <- paste(actionability_levels$assumption, output_keys, sep = "\r")
  canonical_position <- match(output_keys, canonical_keys)
  output_order <- order(
    match(actionability_levels$assumption, assumptions),
    is.na(canonical_position), canonical_position,
    actionability_levels$actionability_type,
    actionability_levels$actionability_level,
    method = "radix", na.last = TRUE
  )
  actionability_levels <- actionability_levels[output_order, , drop = FALSE]
  rownames(actionability_levels) <- NULL
  patient_summary <- do.call(rbind, lapply(assumptions, function(value) {
    patient_rows <- lapply(valid_patient_ids, function(identifier) {
      rows <- scenario_records$assumption == value &
        !is.na(scenario_records$valid_ctdna_result) &
        as.logical(scenario_records$valid_ctdna_result) &
        !is.na(scenario_records$id) & scenario_records$id == identifier &
        scenario_records$non_chip &
        scenario_records$alteration_type %in% c("mutation", "cnv", "fusion")
      records <- scenario_records[rows, , drop = FALSE]
      sensitivity_evidence <- any(is_recorded_actionability_evidence(
        records$actionability_sensitivity
      ))
      resistance_evidence <- any(is_recorded_actionability_evidence(
        records$actionability_resistance
      ))
      data.frame(
        assumption = value,
        id = identifier,
        any_sensitivity = sensitivity_evidence,
        any_resistance = resistance_evidence,
        any_actionable = sensitivity_evidence || resistance_evidence,
        no_actionable_finding = !sensitivity_evidence && !resistance_evidence,
        stringsAsFactors = FALSE
      )
    })
    do.call(rbind, patient_rows)
  }))
  arrow::write_parquet(long, long_path, compression = "zstd")
  arrow::write_parquet(actionability_levels, level_path, compression = "zstd")
  arrow::write_parquet(patient_summary, patient_path, compression = "zstd")
  arrow::write_parquet(actionability_qc, qc_path, compression = "zstd")
  actionability_summary <- do.call(rbind, lapply(assumptions, function(value) {
    patients <- patient_summary[patient_summary$assumption == value, , drop = FALSE]
    counts <- c(
      sum(patients$any_sensitivity),
      sum(patients$any_resistance),
      sum(patients$any_actionable),
      sum(patients$any_sensitivity & patients$any_resistance),
      sum(patients$no_actionable_finding)
    )
    data.frame(
      assumption = value,
      patient_category = c(
        "Any sensitivity evidence",
        "Any resistance evidence",
        "Any sensitivity or resistance evidence",
        "Both sensitivity and resistance evidence",
        "No actionable finding"
      ),
      patient_count = counts,
      denominator_valid_ctdna_patients = denominator,
      patient_proportion = if (denominator == 0L) NA_real_ else
        counts / denominator,
      stringsAsFactors = FALSE
    )
  }))
  arrow::write_parquet(actionability_summary, summary_path, compression = "zstd")
```

```{r}
#| label: actionability-summary
#| tbl-cap: "Patient-level actionability summary"

actionability_summary |>
  dplyr::select(assumption, dplyr::everything()) |>
  knitr::kable()
```

```{r}
#| label: actionability-alteration-counts
#| tbl-cap: "Non-CHIP alteration records by exact OncoKB therapeutic level"

actionability_levels |>
  dplyr::select(
    assumption,
    evidence_direction = actionability_type,
    exact_recorded_level = actionability_level,
    onkb_category,
    position_within_direction = within_direction_position,
    level_meaning,
    recorded_oncokb_versions = onkb_versions,
    recorded_oncokb_update_range = onkb_update_range,
    mutation_records_n = mutation_count,
    cnv_records_n = cnv_count,
    fusion_records_n = fusion_count,
    all_alteration_records_n = alteration_count
  ) |>
  knitr::kable()
```

```{r}
#| label: actionability-completeness
#| tbl-cap: "Missing and non-evidence OncoKB level values retained for completeness checks"

actionability_qc |>
  dplyr::select(assumption, dplyr::everything()) |>
  knitr::kable()
```

```{r}
#| label: actionability-patient-prevalence
#| tbl-cap: "Patients with at least one non-CHIP alteration at each exact OncoKB therapeutic level"

actionability_levels |>
  dplyr::select(
    assumption,
    evidence_direction = actionability_type,
    exact_recorded_level = actionability_level,
    patients_with_mutation_n = mutation_patient_count,
    patients_with_cnv_n = cnv_patient_count,
    patients_with_fusion_n = fusion_patient_count,
    patients_with_any_alteration_n = patient_count,
    validly_tested_patients_n = denominator_valid_ctdna_patients,
    any_alteration_patient_proportion = patient_proportion
  ) |>
  knitr::kable()
```
