---
title: "Best supportive care — endpoint preparation"
format: html
execute:
  freeze: false
---

Time starts at randomization. The main supplementary BSC
analysis estimates the **cause-specific hazard of switching to BSC**: death
before BSC ends follow-up and is coded as censoring. The composite analysis
counts the first of BSC or death as an event. BSC and death on the same day
count as BSC in the cause-specific analysis and as one composite event.

Both unadjusted Bayesian proportional-hazards models contain only randomized
group (`tx`). They use the OS five-df M-spline baseline and treatment prior.
Missing adjustment covariates do not remove participants from these models.

The data cutoff is **2026-09-05**, without a 182-day cap. Exported dates are
shifted for privacy. The private export's `potential_follow_up_days` encodes
the cutoff (or earlier death) on this shifted scale; unrestricted observed
follow-up supplies earlier censoring. The primary `follow_up_end_date` is
already capped at 182 days and is not used here. A recorded BSC date can
extend an earlier last-known-alive record, but never past the exported cutoff
or death. Later BSC dates do not count as observed events.

BSC times are the difference between `bsc_date` and `randomization_date`.
Unrestricted survival time and status supply death or censoring times.
Treatment-start dates are not needed. Participants with BSC before
randomization or no positive follow-up remain in the endpoint audit but cannot
enter these time-to-event models. Event dates and follow-up times are not imputed.

Run this document, then [imputation](01-imputation.qmd),
[model fitting](02-fitting.qmd), and [results](03-results.qmd).

```{r}
#| label: prepare-bsc
#| include: false
library(tidyverse)

source(here::here("R", "analysis-config.R"))
source(here::here("R", "data-helpers.R"))
source(here::here("R", "bsc-helpers.R"))

config <- read_analysis_config()
artifact_dir <- here::here("artifacts", "secondary", "bsc")
result_dir <- here::here("results", "secondary", "bsc")

dir.create(artifact_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)

cohort <- arrow::read_parquet(here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
))
endpoint <- derive_bsc_endpoints(cohort, config$analysis$data_lock)

arrow::write_parquet(endpoint, file.path(result_dir, "endpoint.parquet"))

prepared <- endpoint |>
  filter(bsc_in_analysis) |>
  select(
    id, tx, time_to_bsc_days, event_bsc, event_death_before_bsc,
    event_bsc_or_death, ecog_fstcnt, stage_binary, lung, age, gender,
    albumin, bilirubin, c_reactive_protein, lactate_dehydrogenase
  )

if (nrow(prepared) == 0L) stop("No participants have analyzable BSC follow-up.")

arrow::write_parquet(prepared, file.path(artifact_dir, "prepared.parquet"))

endpoint_audit <- endpoint |>
  count(tx, bsc_exclusion_reason, name = "participants")
event_counts <- prepared |>
  summarise(
    participants = n(), bsc_events = sum(event_bsc),
    death_before_bsc = sum(event_death_before_bsc),
    composite_events = sum(event_bsc_or_death),
    censored_without_either = sum(event_bsc_or_death == 0),
    .by = tx
  )
arrow::write_parquet(endpoint_audit, file.path(result_dir, "endpoint-audit.parquet"))
arrow::write_parquet(event_counts, file.path(result_dir, "event-counts.parquet"))
```

```{r}
#| label: tbl-bsc-endpoint-audit
#| tbl-cap: "BSC analysis population and exclusions"
#| echo: false
knitr::kable(endpoint_audit)
```

```{r}
#| label: tbl-bsc-event-counts
#| tbl-cap: "Observed BSC and competing death events by randomized group"
#| echo: false
knitr::kable(event_counts)
```

