---
title: "Best supportive care — fixed and hierarchical imputations"
format: html
execute:
  freeze: false
---

Five MICE datasets supply each adjusted composite and cause-specific model.
The Adjusted (fixed) branch follows OS: ordered ECOG is imputed by proportional
odds, laboratory values by predictive mean matching, and binary ECOG and mGPS
are derived passively from their completed source variables. The Adjusted
(hierarchical) branch replaces lung cancer with a diagnosis-level random
intercept and uses two-level predictive mean matching for every incomplete
covariate, including a numeric working model for ECOG.

MICE uses fully conditional specification (FCS): each incomplete covariate is
imputed in turn from a model conditional on the current values of the other
predictors. Both event indicators (BSC and death before BSC) and both
cause-specific Nelson–Aalen cumulative hazards inform these models, following
the approximate competing-risks FCS approach described by @Bonneville2022
[sec. 4.1]. Competing deaths inform missing covariates even though they are
treated as censoring in the cause-specific BSC analysis.

This is not substantive-model-compatible FCS (SMC-FCS), which incorporates
the survival likelihood directly into imputation [@Bonneville2022, sec. 4.2].
The hazard predictors provide approximate compatibility; they do not ensure
compatibility with the full Bayesian survival model or its derived covariates.
The same five completed datasets within each branch are used for both
endpoints. Treatment, identifiers, endpoint times, event indicators, and
hazard estimates are never imputed. Only participants with analyzable observed
endpoint times enter MICE. The two branches intentionally retain their
separate cached MICE objects because their predictor matrices and adjustment
estimands differ.

```{r}
#| label: impute-bsc
#| include: false
library(tidyverse)
library(mice)

source(here::here("R", "analysis-config.R"))
source(here::here("R", "survival-helpers.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
artifact_dir <- here::here("artifacts", "secondary", "bsc")
imputation_path <- file.path(artifact_dir, "imputations.rds")
prepared <- arrow::read_parquet(file.path(artifact_dir, "prepared.parquet")) |>
  mutate(
    ecog_binary = derive_ecog_binary(ecog_fstcnt),
    mgps = derive_mgps(albumin, c_reactive_protein),
    ecog_fstcnt = ordered(ecog_fstcnt),
    gender = factor(gender)
  )
prepared$na_bsc <- mice::nelsonaalen(
  prepared, timevar = time_to_bsc_days, statusvar = event_bsc
)
prepared$na_death <- mice::nelsonaalen(
  prepared, timevar = time_to_bsc_days, statusvar = event_death_before_bsc
)
prepared <- prepared |>
  relocate(ecog_binary, mgps, .after = last_col())

methods <- mice::make.method(prepared)
predictors <- mice::make.predictorMatrix(prepared)
fixed_columns <- c(
  "id", "tx", "time_to_bsc_days", "event_bsc", "event_death_before_bsc",
  "event_bsc_or_death", "na_bsc", "na_death"
)
methods[fixed_columns] <- ""
methods["ecog_fstcnt"] <- "polr"
methods["ecog_binary"] <- "~I(derive_ecog_binary(ecog_fstcnt))"
methods["mgps"] <- "~I(derive_mgps(albumin, c_reactive_protein))"
predictors[fixed_columns, ] <- 0
# The composite indicator is the sum of the two cause indicators.
predictors[, c("id", "time_to_bsc_days", "event_bsc_or_death", "ecog_binary", "mgps")] <- 0

input_hash <- digest::digest(list(prepared, methods, predictors, config$imputation$supporting))
imputation <- if (!force && file.exists(imputation_path)) readRDS(imputation_path) else NULL
if (is.null(imputation) || !identical(attr(imputation, "bsc_input_hash"), input_hash)) {
  imputation <- mice::mice(
    prepared, method = methods, predictorMatrix = predictors,
    m = config$imputation$supporting$m, maxit = config$imputation$supporting$maxit,
    donors = config$imputation$supporting$donors,
    seed = analysis_seed(config, "supporting", offset = 1700L),
    printFlag = FALSE
  )
  attr(imputation, "bsc_input_hash") <- input_hash
  saveRDS(imputation, imputation_path, compress = "xz")
}
for (index in config$imputation$supporting$indices) {
  completed <- mice::complete(imputation, index)
  assert_ecog_binary_consistent(completed)
  assert_mgps_consistent(completed)
  stopifnot(!anyNA(completed[c("tx", "ecog_binary", "stage_binary", "lung", "mgps")]))
}
```

```{r}
#| label: fig-bsc-imputation-traces
#| fig-cap: "MICE traces for the five BSC covariate imputations"
#| fig-width: 10
#| fig-height: 8
#| echo: false
plot(imputation, layout = c(2, 5))
```

```{r}
#| label: fig-bsc-imputation-distributions
#| fig-cap: "Observed and imputed laboratory measurements"
#| fig-width: 10
#| fig-height: 6
#| echo: false
densityplot(imputation, ~ albumin + bilirubin + c_reactive_protein + lactate_dehydrogenase)
```

```{r}
#| label: tbl-bsc-imputation-events
#| tbl-cap: "Logged MICE events, if any"
#| echo: false
if (!is.null(imputation$loggedEvents)) {
  imputation$loggedEvents |>
    count(dep, meth, out, name = "logged_events") |>
    knitr::kable()
}
```

```{r}
#| label: impute-bsc-diagnosis
#| include: false
library(miceadds)

diagnosis_artifact_dir <- here::here("artifacts", "secondary", "bsc-diagnosis")
dir.create(diagnosis_artifact_dir, recursive = TRUE, showWarnings = FALSE)
diagnoses <- arrow::read_parquet(here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
)) |>
  select(id, diagnosis)
diagnosis_prepared <- arrow::read_parquet(file.path(artifact_dir, "prepared.parquet")) |>
  select(-lung) |>
  left_join(diagnoses, by = "id", relationship = "one-to-one") |>
  mutate(
    diagnosis = factor(diagnosis), diagnosis_id = as.integer(diagnosis),
    gender = factor(gender)
  )
stopifnot(!anyNA(diagnosis_prepared$diagnosis), nrow(diagnosis_prepared) > 0L,
          !anyDuplicated(diagnosis_prepared$id))
diagnosis_prepared$na_bsc <- mice::nelsonaalen(
  diagnosis_prepared, timevar = time_to_bsc_days, statusvar = event_bsc
)
diagnosis_prepared$na_death <- mice::nelsonaalen(
  diagnosis_prepared, timevar = time_to_bsc_days, statusvar = event_death_before_bsc
)
diagnosis_imputation_data <- diagnosis_prepared |> select(-diagnosis)
diagnosis_methods <- mice::make.method(diagnosis_imputation_data)
diagnosis_incomplete <- names(which(colSums(is.na(diagnosis_imputation_data)) > 0))
diagnosis_methods[] <- ""
diagnosis_methods[diagnosis_incomplete] <- "2l.pmm"
diagnosis_predictors <- mice::make.predictorMatrix(diagnosis_imputation_data)
diagnosis_predictors[, c("id", "time_to_bsc_days", "event_bsc_or_death")] <- 0
diagnosis_predictors[!names(diagnosis_methods) %in% diagnosis_incomplete, ] <- 0
diagnosis_predictors[diagnosis_incomplete, "diagnosis_id"] <- -2
diagnosis_input_hash <- digest::digest(list(
  diagnosis_imputation_data, diagnosis_methods, diagnosis_predictors, config$imputation$supporting
))
diagnosis_imputation_path <- file.path(diagnosis_artifact_dir, "imputation.rds")
diagnosis_imputation <- if (!force && file.exists(diagnosis_imputation_path)) {
  readRDS(diagnosis_imputation_path)
} else NULL
if (is.null(diagnosis_imputation) ||
    !identical(attr(diagnosis_imputation, "input_hash"), diagnosis_input_hash)) {
  diagnosis_imputation <- mice::mice(
    diagnosis_imputation_data, method = diagnosis_methods,
    predictorMatrix = diagnosis_predictors, m = config$imputation$supporting$m, maxit = config$imputation$supporting$maxit, donors = config$imputation$supporting$donors,
    seed = 2701L, printFlag = FALSE
  )
  attr(diagnosis_imputation, "input_hash") <- diagnosis_input_hash
  saveRDS(diagnosis_imputation, diagnosis_imputation_path)
}
diagnosis_completed_datasets <- lapply(config$imputation$supporting$indices, function(index) {
  completed <- mice::complete(diagnosis_imputation, index) |>
    mutate(
      diagnosis = diagnosis_prepared$diagnosis,
      ecog_binary = derive_ecog_binary(ecog_fstcnt),
      mgps = factor(derive_mgps(albumin, c_reactive_protein), levels = 0:2)
    )
  assert_ecog_binary_consistent(completed)
  assert_mgps_consistent(completed)
  stopifnot(!anyNA(completed), all(completed$ecog_fstcnt %in% 0:4),
            diagnosis_imputation$m == config$imputation$supporting$m,
            all(diagnosis_imputation$predictorMatrix[
              diagnosis_incomplete, "diagnosis_id"
            ] == -2))
  completed
})
saveRDS(
  diagnosis_completed_datasets,
  file.path(diagnosis_artifact_dir, "completed-datasets.rds")
)
```

```{r}
#| label: fig-bsc-diagnosis-imputation-traces
#| fig-cap: "MICE traces for the Adjusted (hierarchical) BSC imputations"
#| fig-alt: "Means and standard deviations of imputed ECOG and laboratory values over 50 MICE iterations, with a separate trace for each of five diagnosis-level imputations."
#| fig-width: 10
#| fig-height: 8
#| echo: false
plot(diagnosis_imputation, layout = c(2, 5))
```

```{r}
#| label: fig-bsc-diagnosis-imputation-distributions
#| fig-cap: "Observed and imputed laboratory measurements for the Adjusted (hierarchical) BSC branch"
#| fig-alt: "Density curves compare observed laboratory measurements with each of the five completed diagnosis-level imputations for albumin, bilirubin, CRP, and LDH."
#| fig-width: 10
#| fig-height: 6
#| echo: false
densityplot(
  diagnosis_imputation,
  ~ albumin + bilirubin + c_reactive_protein + lactate_dehydrogenase
)
```

```{r}
#| label: tbl-bsc-diagnosis-imputation-events
#| tbl-cap: "Logged MICE events for the Adjusted (hierarchical) branch, if any"
#| echo: false
if (!is.null(diagnosis_imputation$loggedEvents)) {
  diagnosis_imputation$loggedEvents |>
    count(dep, meth, out, name = "logged_events") |>
    knitr::kable()
}
```
