---
title: "Best supportive care — unadjusted and adjusted model fitting"
format: html
execute:
  freeze: false
---

The two BSC endpoints are fitted together for three adjustment approaches:
Unadjusted, Adjusted (fixed), and Adjusted (hierarchical). The cause-specific
endpoint censors death before BSC; the composite endpoint counts the first of
BSC or death. Existing fitted models are reused when their input hashes match.

Adjusted (fixed) includes binary ECOG, stage, lung cancer, and categorical mGPS.
Adjusted (hierarchical) replaces lung cancer with `(1 || diagnosis)`, retaining
ECOG, stage, and mGPS. The diagnosis random intercept has one variance
parameter and no correlation parameter.

```{r}
#| label: fit-bsc
#| include: false
library(tidyverse)

source(here::here("R", "analysis-config.R"))
source(here::here("R", "survival-helpers.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
artifact_dir <- here::here("artifacts", "secondary", "bsc")
model_dir <- file.path(artifact_dir, "models")
imputations <- readRDS(file.path(artifact_dir, "imputations.rds"))
prepared <- arrow::read_parquet(file.path(artifact_dir, "prepared.parquet"))
stopifnot(imputations$m == config$imputation$supporting$m, nrow(prepared) > 0L)
dir.create(model_dir, recursive = TRUE, showWarnings = FALSE)

for (endpoint_name in c("cause_specific", "composite")) {
  unadjusted_data <- prepared |>
    transmute(
      id, tx, time_to_bsc_days,
      event = if (endpoint_name == "cause_specific") event_bsc else event_bsc_or_death
    )
  if (sum(unadjusted_data$event) == 0L) stop("No events for ", endpoint_name, ".")
  model_path <- file.path(model_dir, paste0("unadjusted-", endpoint_name, ".rds"))
  input_hash <- digest::digest(unadjusted_data)
  model <- if (!force && file.exists(model_path)) readRDS(model_path) else NULL
  if (is.null(model) || !identical(attr(model, "bsc_input_hash"), input_hash)) {
    model <- rstanarm::stan_surv(
      formula = survival::Surv(time_to_bsc_days, event) ~ tx,
      data = unadjusted_data,
      basehaz = "ms", basehaz_ops = list(df = 5),
      prior = rstanarm::normal(0, 0.55),
      prior_intercept = rstanarm::normal(0, 20),
      prior_aux = rstanarm::dirichlet(1),
      chains = 4L, cores = 4L, iter = 2000L, warmup = 1000L,
      refresh = 0L,
      seed = analysis_seed(config, "supporting", offset = if (
        endpoint_name == "cause_specific"
      ) 1600L else 1601L)
    )
    attr(model, "bsc_input_hash") <- input_hash
    saveRDS(model, model_path, compress = "xz")
  }

  for (index in config$imputation$supporting$indices) {
    completed <- mice::complete(imputations, index)
    assert_ecog_binary_consistent(completed)
    assert_mgps_consistent(completed)
    fitting_data <- completed |>
      transmute(
        id, tx, time_to_bsc_days,
        event = if (endpoint_name == "cause_specific") event_bsc else event_bsc_or_death,
        ecog_binary, stage_binary, lung, mgps = factor(mgps, levels = 0:2)
      )
    stopifnot(!anyNA(fitting_data), all(fitting_data$time_to_bsc_days > 0))
    model_path <- file.path(
      model_dir, sprintf("adjusted-%s-%02d.rds", endpoint_name, index)
    )
    input_hash <- digest::digest(fitting_data)
    model <- if (!force && file.exists(model_path)) readRDS(model_path) else NULL
    if (!is.null(model) && identical(attr(model, "bsc_input_hash"), input_hash)) next

    model <- rstanarm::stan_surv(
      formula = survival::Surv(time_to_bsc_days, event) ~
        tx + ecog_binary + stage_binary + lung + mgps,
      data = fitting_data,
      basehaz = "ms", basehaz_ops = list(df = 5),
      prior = rstanarm::normal(location = rep(0, 6), scale = c(0.55, rep(2.5, 5))),
      prior_intercept = rstanarm::normal(0, 20),
      prior_aux = rstanarm::dirichlet(1),
      chains = 4L, cores = 4L, iter = 2000L, warmup = 1000L,
      refresh = 0L,
      seed = analysis_seed(config, "supporting", index, offset = if (
        endpoint_name == "cause_specific"
      ) 1800L else 1900L)
    )
    attr(model, "bsc_input_hash") <- input_hash
    saveRDS(model, model_path, compress = "xz")
  }
}
```

```{r}
#| label: fit-bsc-hierarchical
#| include: false
hierarchical_artifact_dir <- here::here(
  "artifacts", "secondary", "bsc-diagnosis"
)
hierarchical_model_dir <- hierarchical_artifact_dir
completed_datasets <- readRDS(file.path(
  hierarchical_artifact_dir, "completed-datasets.rds"
))
hierarchical_imputation <- readRDS(file.path(
  hierarchical_artifact_dir, "imputation.rds"
))
hierarchical_input_hash <- attr(hierarchical_imputation, "input_hash")
stopifnot(length(completed_datasets) == config$imputation$supporting$m, hierarchical_imputation$m == config$imputation$supporting$m)

for (index in config$imputation$supporting$indices) {
  completed <- completed_datasets[[index]]
  assert_ecog_binary_consistent(completed)
  assert_mgps_consistent(completed)
  for (endpoint_name in c("cause_specific", "composite")) {
    fitting_data <- completed |>
      mutate(event = if (endpoint_name == "cause_specific") event_bsc else event_bsc_or_death)
    stopifnot(!anyNA(fitting_data), all(fitting_data$time_to_bsc_days > 0))
    model_path <- file.path(
      hierarchical_model_dir, sprintf("%s-%02d.rds", endpoint_name, index)
    )
    model_hash <- digest::digest(list(
      hierarchical_input_hash, fitting_data, "diagnosis-intercept-v1"
    ))
    model <- if (!force && file.exists(model_path)) readRDS(model_path) else NULL
    if (!is.null(model) && identical(attr(model, "input_hash"), model_hash)) next

    model <- rstanarm::stan_surv(
      formula = survival::Surv(time_to_bsc_days, event) ~
        tx + ecog_binary + stage_binary + mgps + (1 || diagnosis),
      data = fitting_data, basehaz = "ms", basehaz_ops = list(df = 5),
      prior = rstanarm::normal(location = rep(0, 5), scale = c(0.55, rep(2.5, 4))),
      prior_intercept = rstanarm::normal(0, 20),
      prior_aux = rstanarm::dirichlet(1),
      prior_covariance = rstanarm::decov(),
      chains = 4L, cores = 4L, iter = 2000L, warmup = 1000L,
      adapt_delta = 0.99, refresh = 0L,
      seed = index + if (endpoint_name == "cause_specific") 2800L else 2900L
    )
    attr(model, "input_hash") <- model_hash
    saveRDS(model, model_path, compress = "xz")
  }
}
```
