---
title: "Best supportive care — unadjusted and adjusted results"
format: html
execute:
  freeze: false
---

The cause-specific hazard ratio compares the instantaneous rate of BSC among
patients who are still alive and have not switched to BSC. The composite hazard
ratio concerns the first occurrence of BSC or death. Neither is a
subdistribution hazard ratio, and censoring deaths in the cause-specific
analysis does not estimate the cumulative probability of switching to BSC.

The reported hazard ratios are exponentiated treatment coefficients. Unadjusted
models contain only randomized group. Adjusted (fixed) models condition on
ECOG, stage, lung cancer, and mGPS; Adjusted (hierarchical) models replace lung
cancer with a fitted diagnosis random intercept and condition on that fitted
effect. The adjusted HRs are conditional estimands, and the three adjustment
structures therefore target different HRs. Pooling draws across imputations
does not marginalize an HR over patients. Standardized curves and day-182
survival differences are marginal over the observed analysis cohort, retaining
fitted diagnosis effects for the hierarchical models.

```{r}
#| label: summarize-bsc
#| include: false
library(tidyverse)

source(here::here("R", "analysis-config.R"))
source(here::here("R", "survival-helpers.R"))

config <- read_analysis_config()
artifact_dir <- here::here("artifacts", "secondary", "bsc")
hierarchical_artifact_dir <- here::here(
  "artifacts", "secondary", "bsc-diagnosis"
)
model_dir <- file.path(artifact_dir, "models")
result_dir <- here::here("results", "secondary", "bsc")
figure_dir <- file.path(artifact_dir, "figures")
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

prepared <- arrow::read_parquet(file.path(artifact_dir, "prepared.parquet"))
max_follow_up <- max(prepared$time_to_bsc_days)
post_draws <- as.integer(config$imputation$supporting$post_estimation_draws)
adjustment_labels <- c(
  unadjusted = "Unadjusted",
  fixed = "Adjusted (fixed)",
  hierarchical = "Adjusted (hierarchical)"
)
endpoint_labels <- c(
  cause_specific = "Cause-specific BSC",
  composite = "BSC or death"
)

fits <- list()
for (endpoint_name in names(endpoint_labels)) {
  for (adjustment in names(adjustment_labels)) {
    indices <- if (adjustment == "unadjusted") 0L else config$imputation$supporting$indices
    for (index in indices) {
      model_path <- if (adjustment == "hierarchical") {
        file.path(
          hierarchical_artifact_dir,
          sprintf("%s-%02d.rds", endpoint_name, index)
        )
      } else if (adjustment == "unadjusted") {
        file.path(model_dir, paste0("unadjusted-", endpoint_name, ".rds"))
      } else {
        file.path(
          model_dir, sprintf("adjusted-%s-%02d.rds", endpoint_name, index)
        )
      }
      model <- readRDS(model_path)
      fits[[paste(endpoint_name, adjustment, index, sep = "-")]] <- list(
        model = model,
        data = model$data,
        endpoint = endpoint_name,
        adjustment = adjustment,
        imputation = index
      )
    }
  }
}

diagnostics <- purrr::map_dfr(fits, function(fit) {
  parameters <- posterior::summarise_draws(posterior::as_draws_array(fit$model))
  sampler <- rstan::get_sampler_params(fit$model$stanfit, inc_warmup = FALSE)
  bfmi <- vapply(sampler, function(x) {
    mean(diff(x[, "energy__"])^2) / var(x[, "energy__"])
  }, numeric(1))
  tibble(
    endpoint = fit$endpoint,
    adjustment = unname(adjustment_labels[fit$adjustment]),
    imputation = fit$imputation,
    max_rhat = max(parameters$rhat, na.rm = TRUE),
    min_ess_bulk = min(parameters$ess_bulk, na.rm = TRUE),
    min_ess_tail = min(parameters$ess_tail, na.rm = TRUE),
    divergences = sum(vapply(
      sampler, function(x) sum(x[, "divergent__"]), numeric(1)
    )),
    treedepth_hits = sum(vapply(
      sampler, function(x) sum(x[, "treedepth__"] >= 10), numeric(1)
    )),
    min_bfmi = min(bfmi)
  )
}) |>
  mutate(
    diagnostics_passed = max_rhat <= config$diagnostics$max_rhat &
      min_ess_bulk >= config$diagnostics$min_ess_bulk &
      min_ess_tail >= config$diagnostics$min_ess_tail &
      divergences <= config$diagnostics$max_divergent_transitions &
      treedepth_hits <= config$diagnostics$max_treedepth_exceeded &
      min_bfmi >= config$diagnostics$min_bfmi
  )
stopifnot(
  nrow(diagnostics) == 22L,
  all(diagnostics$diagnostics_passed)
)

hr_draws <- purrr::map_dfr(fits, function(fit) {
  log_hr <- as.numeric(as.matrix(fit$model, pars = "tx"))
  if (fit$imputation > 0L) {
    seed <- if (fit$adjustment == "fixed") {
      analysis_seed(config, "supporting", fit$imputation, offset = 2000L)
    } else {
      3000L + fit$imputation
    }
    set.seed(seed)
    log_hr <- sample(log_hr, post_draws)
  }
  if (fit$adjustment == "hierarchical") {
    stopifnot(
      identical(fit$model$cnms$diagnosis, "(Intercept)"),
      length(grep("^Sigma", colnames(as.matrix(fit$model)))) == 1L
    )
  }
  tibble(
    endpoint = fit$endpoint,
    adjustment = unname(adjustment_labels[fit$adjustment]),
    imputation = fit$imputation,
    draw = seq_along(log_hr),
    log_hazard_ratio = log_hr,
    hazard_ratio = exp(log_hr)
  )
})
hr_summary <- hr_draws |>
  summarise(
    median = median(hazard_ratio),
    lower_95 = quantile(hazard_ratio, 0.025),
    upper_95 = quantile(hazard_ratio, 0.975),
    probability_hr_below_one = mean(hazard_ratio < 1),
    imputations = sum(unique(imputation) > 0),
    posterior_draws = n(),
    .by = c(endpoint, adjustment)
  ) |>
  left_join(
    diagnostics |>
      summarise(
        diagnostics_passed = all(diagnostics_passed),
        .by = c(endpoint, adjustment)
      ),
    by = c("endpoint", "adjustment")
  )
stopifnot(
  nrow(hr_summary) == 6L,
  all(count(hr_draws, endpoint, adjustment, imputation)$n ==
        ifelse(count(hr_draws, endpoint, adjustment, imputation)$imputation == 0L,
               4000L, post_draws))
)

curves <- purrr::map_dfr(fits, function(fit) {
  posterior_standardized_survival_draws(
    fit$model, fit$data, horizon = max_follow_up, n_points = 200L,
    n_draws = post_draws, seed = 3000L + fit$imputation
  ) |>
    mutate(
      endpoint = fit$endpoint,
      adjustment = unname(adjustment_labels[fit$adjustment]),
      imputation = fit$imputation
    )
})
stopifnot(
  min(curves$time) == 0,
  max(curves$time) == max_follow_up,
  all(is.finite(curves$survival)),
  all(dplyr::between(curves$survival, 0, 1))
)

checks <- purrr::map_dfr(fits, function(fit) {
  newdata <- fit$data |> mutate(tx = 1)
  individual <- posterior_survfit_compatible(
    fit$model, newdata = newdata, times = 182, extrapolate = FALSE,
    standardise = FALSE, return_matrix = TRUE, draws = 100L,
    seed = 3001L + fit$imputation
  )[[1]]
  standardized <- posterior_survfit_compatible(
    fit$model, newdata = newdata, times = 182, extrapolate = FALSE,
    standardise = TRUE, return_matrix = TRUE, draws = 100L,
    seed = 3001L + fit$imputation
  )[[1]]
  error <- max(abs(rowMeans(individual) - as.numeric(standardized)))
  diagnosis_range <- NA_real_
  if (fit$adjustment == "hierarchical") {
    contrast_data <- newdata[rep(1L, nlevels(newdata$diagnosis)), ]
    contrast_data$diagnosis <- factor(
      levels(newdata$diagnosis), levels = levels(newdata$diagnosis)
    )
    predictions <- posterior_survfit_compatible(
      fit$model, newdata = contrast_data, times = 182, extrapolate = FALSE,
      standardise = FALSE, return_matrix = TRUE, draws = 100L,
      seed = 3001L + fit$imputation
    )[[1]]
    diagnosis_range <- max(apply(predictions, 1, function(x) diff(range(x))))
    stopifnot(diagnosis_range > 1e-8)
  }
  stopifnot(error < 1e-10)
  tibble(
    endpoint = fit$endpoint,
    adjustment = unname(adjustment_labels[fit$adjustment]),
    imputation = fit$imputation,
    maximum_averaging_error = error,
    maximum_diagnosis_contrast = diagnosis_range
  )
})

landmark_draws <- purrr::map_dfr(fits, function(fit) {
  predictions <- lapply(0:1, function(treatment) {
    newdata <- fit$data |> mutate(tx = treatment)
    as.numeric(posterior_survfit_compatible(
      fit$model, newdata = newdata, times = 182, extrapolate = FALSE,
      standardise = TRUE, return_matrix = TRUE, draws = post_draws,
      seed = 3000L + fit$imputation
    )[[1]])
  })
  contrast <- predictions[[2]] - predictions[[1]]
  stopifnot(length(contrast) == post_draws, sd(contrast) > 0)
  tibble(
    endpoint = fit$endpoint,
    adjustment = unname(adjustment_labels[fit$adjustment]),
    imputation = fit$imputation,
    draw = seq_len(post_draws),
    control = predictions[[1]],
    intervention = predictions[[2]],
    contrast = contrast
  )
})
stopifnot(
  nrow(landmark_draws) == 22L * post_draws,
  all(count(landmark_draws, endpoint, adjustment, imputation)$n == post_draws)
)
landmarks <- landmark_draws |>
  summarise(
    control = median(control),
    intervention = median(intervention),
    difference = median(contrast),
    lower_95 = quantile(contrast, 0.025),
    upper_95 = quantile(contrast, 0.975),
    probability_difference_positive = mean(contrast > 0),
    .by = c(endpoint, adjustment)
  )

curve_summary <- curves |>
  summarise(
    median = median(survival),
    lower_95 = quantile(survival, 0.025),
    upper_95 = quantile(survival, 0.975),
    .by = c(endpoint, adjustment, tx, time)
  )
arrow::write_parquet(hr_draws, file.path(result_dir, "hazard-ratio-draws.parquet"))
arrow::write_parquet(hr_summary, file.path(result_dir, "summary.parquet"))
arrow::write_parquet(landmark_draws, file.path(result_dir, "day-182-draws.parquet"))
arrow::write_parquet(landmarks, file.path(result_dir, "day-182-summary.parquet"))
arrow::write_parquet(checks, file.path(result_dir, "prediction-checks.parquet"))
arrow::write_parquet(diagnostics, file.path(result_dir, "diagnostics.parquet"))
arrow::write_parquet(
  curves, file.path(artifact_dir, "standardized-survival-draws.parquet")
)
```

```{r}
#| label: tbl-bsc-hazard-ratios
#| tbl-cap: "Intervention-versus-control hazard ratios for BSC and BSC-or-death"
#| echo: false
hr_summary |>
  mutate(
    Endpoint = unname(endpoint_labels[endpoint]),
    Adjustment = adjustment
  ) |>
  select(Endpoint, Adjustment, median, lower_95, upper_95,
         probability_hr_below_one) |>
  knitr::kable(digits = 3)
```

```{r}
#| label: fig-bsc-hazard-ratios
#| fig-cap: "Unadjusted, Adjusted (fixed), and Adjusted (hierarchical) BSC hazard ratios with 95% credible intervals"
#| fig-alt: "Treatment hazard ratios for cause-specific BSC and BSC or death, comparing unadjusted, fixed-covariate, and diagnosis-random-intercept models."
#| fig-width: 9
#| fig-height: 5
#| echo: false
hr_plot <- hr_summary |>
  mutate(
    endpoint = factor(unname(endpoint_labels[endpoint]),
                      levels = unname(endpoint_labels)),
    adjustment = factor(adjustment, levels = unname(adjustment_labels))
  ) |>
  ggplot(aes(x = median, y = adjustment, xmin = lower_95, xmax = upper_95)) +
  geom_pointrange(orientation = "y") +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ endpoint) +
  scale_x_log10() +
  labs(x = "Hazard ratio, intervention versus control", y = NULL) +
  theme_bw()
ggsave(file.path(figure_dir, "hazard-ratios.png"), hr_plot,
       width = 9, height = 5, dpi = 150)
hr_plot
```

The standardized cause-specific curve is **net survival** for the BSC hazard;
it is not the probability of remaining alive and BSC-free when death competes,
and it is not a cumulative-incidence curve. The standardized composite curve
estimates the probability of remaining alive and BSC-free. The same distinction
applies to the day-182 survival probabilities and their intervention-minus-
control contrasts below.

```{r}
#| label: tbl-bsc-day-182
#| tbl-cap: "Day-182 standardized survival and intervention-minus-control difference"
#| echo: false
landmarks |>
  mutate(
    Outcome = unname(endpoint_labels[endpoint]),
    Adjustment = adjustment
  ) |>
  select(Outcome, Adjustment, control, intervention, difference, lower_95,
         upper_95, probability_difference_positive) |>
  knitr::kable(digits = 3)
```

```{r}
#| label: fig-bsc-standardized-survival
#| fig-cap: "Standardized BSC survival by endpoint, adjustment, and randomized group"
#| fig-alt: "Standardized cause-specific BSC net survival and BSC-or-death survival curves for control and intervention under unadjusted, fixed-covariate, and diagnosis-random-intercept adjustment."
#| fig-width: 12
#| fig-height: 7
#| echo: false
plot_data <- curve_summary |>
  mutate(
    endpoint = factor(unname(endpoint_labels[endpoint]),
                      levels = unname(endpoint_labels)),
    adjustment = factor(adjustment, levels = unname(adjustment_labels)),
    tx = factor(tx, levels = 0:1, labels = c("Control", "Intervention"))
  )
survival_plot <- ggplot(
  plot_data, aes(time, median, colour = tx, fill = tx)
) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), alpha = 0.15, colour = NA) +
  geom_line() +
  facet_grid(endpoint ~ adjustment) +
  scale_x_continuous(
    breaks = seq(0, max_follow_up, by = 100)
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Days from randomization", y = "Standardized survival",
       colour = NULL, fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(file.path(figure_dir, "standardized-survival.png"), survival_plot,
       width = 12, height = 7, dpi = 150)
survival_plot
```

Observed curves are shown separately from the model-derived standardized
curves. The composite curve estimates the probability of remaining alive and
BSC-free. The cause-specific curve is the Nelson–Aalen cumulative hazard for
BSC with death censored; it is not a cumulative-incidence curve.

```{r}
#| label: fig-bsc-composite-km
#| fig-cap: "Unadjusted Kaplan-Meier probability of remaining alive and BSC-free"
#| fig-alt: "Kaplan-Meier curves for the randomized groups from randomization until BSC or death, with censor marks and numbers at risk."
#| fig-width: 8
#| fig-height: 6
#| echo: false
observed <- prepared |>
  mutate(tx_label = factor(tx, levels = 0:1, labels = c("Control", "Intervention")))
composite_fit <- ggsurvfit::survfit2(
  survival::Surv(time_to_bsc_days, event_bsc_or_death) ~ tx_label,
  data = observed
)
composite_plot <- ggsurvfit::ggsurvfit(composite_fit) +
  ggsurvfit::add_confidence_interval() +
  ggsurvfit::add_censor_mark() +
  scale_x_continuous(breaks = seq(0, max_follow_up, by = 100)) +
  ggsurvfit::add_risktable(
    times = seq(0, max_follow_up, by = 100),
    risktable_stats = c("n.risk", "cum.event"),
    stats_label = list(n.risk = "At risk", cum.event = "Events")
  ) +
  labs(x = "Days since randomization", y = "Alive and BSC-free probability")
ggsave(
  file.path(figure_dir, "composite-kaplan-meier.png"),
  ggsurvfit::ggsurvfit_build(composite_plot), width = 8, height = 6, dpi = 150
)
composite_plot
```

```{r}
#| label: fig-bsc-cumulative-hazard
#| fig-cap: "Unadjusted cumulative cause-specific hazard of BSC, censoring death before BSC"
#| fig-alt: "Nelson–Aalen cumulative BSC hazard curves by randomized group; death before BSC is treated as censoring."
#| fig-width: 8
#| fig-height: 5
#| echo: false
cause_fit <- survival::survfit(
  survival::Surv(time_to_bsc_days, event_bsc) ~ tx_label, data = observed
)
hazard_data <- tibble(
  time = cause_fit$time,
  cumulative_hazard = cause_fit$cumhaz,
  group = sub(
    "tx_label=", "", rep(names(cause_fit$strata), cause_fit$strata), fixed = TRUE
  )
) |>
  bind_rows(tibble(
    time = 0, cumulative_hazard = 0, group = levels(observed$tx_label)
  )) |>
  arrange(group, time)
hazard_plot <- ggplot(hazard_data, aes(time, cumulative_hazard, colour = group)) +
  geom_step() +
  labs(x = "Days since randomization", y = "Cumulative cause-specific BSC hazard",
       colour = NULL) +
  theme_bw()
ggsave(file.path(figure_dir, "cause-specific-hazard.png"), hazard_plot,
       width = 8, height = 5, dpi = 150)
hazard_plot
```

```{r}
#| label: tbl-bsc-diagnostics
#| tbl-cap: "Posterior diagnostics and prediction checks for all BSC models"
#| echo: false
knitr::kable(diagnostics, digits = 3)
knitr::kable(checks, digits = 10)
```
