---
title: Long Ordinal - Data imputation
format: 
  typst:
    papersize: a4
    margin:
      x: 2cm
      y: 2.5cm
    toc: false
    columns: 1
    tbl-cap-location: bottom
---

## Setup

```{r}
#| output: false
library(here)
library(tidyverse)
library(mice)
library(miceadds)
library(arrow)
library(data.table)
```

## Load data

These are historic data from patients of the Division of Medical Oncology at USB, which had their first appointment from 2023 to mid 2025, i.e., before LIQPLAT started. The data are not publicly available.

```{r}
set.seed(1234)
df_raw <- read_parquet(here("data", "long_data.parquet"))
```

### Simplify some baseline varaibles

```{r}
# remove observations after day 182
# lump factor variables according to frequency to have a maximum of 5 levels

# create random group allocation in historical data
tx <- sample(unique(df_raw$pat_id), size = 0.5 * length(unique(df$pat_id)), replace = FALSE)

df_raw <- df_raw |> 
  mutate(tx = if_else(pat_id %in% tx, "1", "0"))


imputation_vars <- c(
  "pat_id",
  "tx",
  "gender",
  "ecog_fstcnt",
  "diagnosis",
  "age", 
  "albumin",
  "bilirubin", 
  "c_reactive_protein",	
  "lactate_dehydrogenase",
  "day", 
  "in_hospital", 
  "unplanned_stay",
  "er_visit",
  "death"
)

df <- df_raw |> 
  select(all_of(imputation_vars)) |> 
  mutate(
    diagnosis = fct_lump_n(diagnosis, n = 4),
    diagnosis = factor(as.integer(diagnosis)),    
    ecog_fstcnt = factor(ecog_fstcnt, levels = 1:3, ordered = TRUE), #in liqplat might be enough patients with 4 levels
    in_hospital = case_when(
        in_hospital == 1 ~ TRUE, 
        TRUE ~ FALSE),
    unplanned_stay = case_when(
        unplanned_stay == 1 ~ TRUE, 
        TRUE ~ FALSE),
    er_visit = case_when(
        er_visit == 1 ~ TRUE, 
        TRUE ~ FALSE)
    ) |> 
  filter(day < 183) |>
  group_by(pat_id) |>
  mutate(
    survival_time = max(day),
    died = if_else(any(death == 1), 1, 0)
  ) |> 
  ungroup() |> 
  arrange(pat_id, day)
```

### Add Nelson-Aalen-estimates, "Randomize"

Nelson-Aalen yields slightly better predictions than raw survival time, see: https://stefvanbuuren.name/fimd/sec-toomany.html

```{r}
na_df <- df |> 
    select(pat_id, survival_time, died) |> 
    unique()

# nelason alen estimator
na_est <- cbind(na_df$pat_id, nelsonaalen(na_df, timevar = survival_time, statusvar = died)) |> 
    data.table() |> 
    rename(pat_id = V1, na_est = V2)

df <- df |> 
    left_join(na_est, by = "pat_id") |> 
    select(-survival_time)
```

## Variables for final analysis

When analyzing the trial data, we will use the following predictors for imputation and modeling:

For reasons explain in:
Mathur, Maya B., Willi Zhang, and Ilya Shpitser. 2025. “Imputation without nightMARs: Graphical Criteria for Valid Imputation of Missing Data.” https://doi.org/10.31219/osf.io/zqne9_v2.

We restrict the imputation model to only include variables included in the analysis model and variables that causally affect either the incomplete variables or the missingness indicators, i.e., we do not include post-baseline variables that are not part of the analysis model.


**Demographics and baseline clinical:**

-   `age` - age at enrollment

-   `gender` - patient sex

-   `ecog_fstcnt` - ECOG performance status (ordered factor: 0-4)

-   `diagnosis` - primary cancer diagnosis (lumped to 5 most frequent categories)

**Laboratory values:**

-   `albumin` - serum albumin (g/L)

-   `c_reactive_protein` - CRP (mg/L)

-   `bilirubin` - serum bilirubin (µmol/L)

-   `lactate_dehydrogenase` - LDH (U/L)

**Treatment assignment:**

-   `tx` - treatment group allocation

**Longitudinal outcome:**

-   `day` - day after random selection

-   `in_hospital` - indicator for being in hospital at a given timepoint

-   `unplanned_stay` - indicator for the hospital stay to be unplanned

-   `er_visit` - indicator for emergency room visit at that timepoint

-   `na_est` - Nelson-Aalen cumulative hazard estimate

-   `died` - death indicator

## Multiple imputation

Baseline values will be imputed at the patient (cluster level).

```{r}
#| output: false
md.pattern(df)

init <- mice(df, maxit = 0)
pred_matrix <- init$predictorMatrix
meth <- init$method

# Use patient ID as clustering variable
pred_matrix[, "pat_id"] <- -2

# use mean matching at the cluster level for baseline observations
# this ensures only one ecog per patient per imputation
meth["age"] <- "2lonly.pmm"
meth["diagnosis"] <- "2lonly.pmm"
meth["gender"] <- "2lonly.pmm"
meth["ecog_fstcnt"] <- "2lonly.pmm"
meth["albumin"] <- "2lonly.pmm"
meth["bilirubin"] <- "2lonly.pmm" 
meth["c_reactive_protein"] <- "2lonly.pmm"
meth["lactate_dehydrogenase"] <- "2lonly.pmm"
meth["died"] <- "2lonly.pmm"
meth["na_est"] <- "2lonly.pmm"

# These are not missing in our dummy data but mnight have some missingness in the trial data
meth["in_hospital"] <- "2l.pmm"
meth["unplanned_stay"] <- "2l.pmm"
meth["er_visit"] <- "2l.pmm"


imputed_data <- mice(df,
                     m = 5, # will set to 50
                     maxit = 20, #will set to 50
                     predictorMatrix = pred_matrix,
                     method = meth,
                     seed = 1234)
```

## Diagnostics

The diagnostics look fine to me.

```{r}
plot(imputed_data)
densityplot(imputed_data)
densityplot(imputed_data, ~ ecog_fstcnt)
```

## Select one imputation

For the analysis of the trial data we'd of course use all the imputed datasets and stack the posterior draws. However, for our simulations this is unfeasible, and we shall thus only use one complete dataset for the setup. I will use the data imputed in long format for the time being.

```{r}
complete_data <- complete(imputed_data, 4)
```

## Export

```{r}
write_parquet(complete_data, sink = here("data", "hosp_long_complete.parquet"))
```