---
Title: Type I Error Rate with Poisson Regression
---

## Setup
```{r}
# Load Libraries
library(arrow)
library(tidyverse)
library(furrr)
library(glue)
library(fs)
library(here)
library(brms)
library(rmsb)
```

## Load data

```{r}
set.seed(1234)
refit_model <- FALSE

# Load full dataset
# Collapse staet 3 and 4
df <- read_parquet(here("data", "hosp_long_weeks_anon.parquet")) |> 
    mutate(
        y = if_else(y == 4, 3, y),
        y = if_else(y == 5, 4, y),
        yprev = if_else(yprev == "4", "3", yprev),
        yprev = factor(yprev),
        tx = factor(tx)) |> 
    select(id, tx, y, yprev, time = week)


surv_time <- df |> group_by(id) |> 
    mutate(surv_time = n()) |> 
    select(id, surv_time) |> 
    distinct()

# weeks spent with at least one day in state 3 or 2 (hospital or ER)
count_df <- df |> 
    group_by(id, tx) |> 
    count(y) |> 
    # complete each group with n = 0 if y is missing
    complete(y = 0:4, fill = list(n = 0)) |>
    filter(y == 3 | y == 2) |> # weeks in hospital
    summarise(n = sum(n)) |> 
    left_join(surv_time, by = "id") |>
    ungroup()

if(refit_model){
    model <- brm(
        n | rate(surv_time) ~ tx,
        data = count_df, 
        family = negbinomial(),
        prior = c(
            prior(normal(0, 100), class = "b")
        ),
        iter = 2000,
        seed = 1234,
        cores = 4,
        file = here("2-hosp", "models", "model_neg_binom.rds")
    )

} else {
    model <- readRDS(here("2-hosp", "models", "model_neg_binom.rds"))
}
```

I'm really not sure what estimand we are targeting, when we sum the days not at home (and not dead), and add a offset for the survival_time. For each patient, the prediction based on the model would be the number of days not at home, considering his survival time (at assuming a constant rate of the events). So is it just the same as summing SOP state 2 and 3 over time / the probabilities? The predicted value is the exp(log(linpred)) * survival time.

## Simulation function

```{r}
run_simulation_iteration <- function(iter_num, base_path, full_data, model) {
    # This function runs one full simulation iteration.
    
    # --- 1. Setup for the current iteration ---
    set.seed(123 + iter_num)
    iter_path <- file.path(base_path, glue("run_{iter_num}"))
    if (!dir.exists(iter_path)) {
        dir.create(iter_path, recursive = TRUE)
    }

    # Resample from data with no effect
    tx_ids <- full_data |> 
        distinct(id) |> 
        sample_frac(2/3) |> pull(id)

    data_for_model <- full_data |> 
        mutate(
            tx = if_else(id %in% tx_ids, "1", "0"),
            tx = factor(tx, levels = c(0, 1)))
    
    rm(full_data)


    # --- 5. Save the data used for model fitting ---
    saveRDS(data_for_model, file.path(iter_path, "data_for_model.rds"))


    # --- 6. Fit the Model and Save ---

    # we'll stick to evaluating type I error rate via OR of tx for the time being, then we don't need to marginalize
    # if we move to computing cluster, maybe do marginalization
    
    if (nrow(data_for_model) > 0) {
        model <- tryCatch({        
            update(model, 
              newdata = data_for_model, 
              cores = 1)            
        }, error = function(e) {
            message(glue("ERROR in iteration {iter_num}: {e$message}"))
            
            saveRDS(data_for_model, file.path(iter_path, "problematic_data.rds"))
            
            return(NULL)
        })
        
        # --- 7. Save the results ONLY if the model ran successfully ---
        if (!is.null(model)) {
            # Save the MCMC draws to a separate file
            saveRDS(brms::as_draws_df(model), file.path(iter_path, "model_draws.rds"))

            # Predicted difference in the number of weeks in state 3
            # Predictions when everyone is treated and untreated, get draws, take individual level differences, group by draw, average over all patient, then takes the median for point estimate
            ATE_draws <- marginaleffects::avg_comparisons(model, variables = "tx", type = "response") |> 
                marginaleffects::get_draws()

            saveRDS(ATE_draws, file.path(iter_path, "ATE_draws.rds"))

            # Save rhat values for all params
            rhat_values <- brms::rhat(model) |> 
                data.frame() |> 
                rownames_to_column() |> 
                pivot_wider(names_from = rowname, values_from = brms..rhat.model.)

            saveRDS(rhat_values, file.path(iter_path, "rhat_values.rds"))
            
            # save ess ratios for all parameters
            ess_ratios <- brms::neff_ratio(model) |> 
                data.frame() |> 
                rownames_to_column() |> 
                pivot_wider(names_from = rowname, values_from = brms..neff_ratio.model.)

            saveRDS(ess_ratios, file.path(iter_path, "ess_ratios.rds"))
        }
    }
    # Return a status message for progress tracking
    return(glue("Iteration {iter_num} completed."))
}
```

```{r}
check_completed_simulations <- function(output_path, total_sims) {
    completed_files <- fs::dir_ls(
        path = output_path, 
        glob = "*/model_draws.rds", 
        recurse = TRUE
    )

    completed_iters <- integer(0) 
    if (length(completed_files) > 0) {
        completed_iters <- stringr::str_extract(completed_files, "(?<=run_)\\d+") |>
            as.integer() |>
            na.omit() |>
            unique()
    }

    all_possible_iters <- 1:total_sims
    iters_to_run <- setdiff(all_possible_iters, completed_iters)

    return(list(
        iters_to_run = iters_to_run,
        n_to_run = length(iters_to_run),
        n_completed = length(completed_iters)
    ))
}
```

## Simulations

Setup
```{r}
N_SIMULATIONS <- 1008
N_CORES <- 8
future::plan(future.callr::callr, workers = N_CORES)
```

Parallelized Simulation
```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-resampling-H0-neg-binom")
sim_status <- check_completed_simulations(OUTPUT_PATH, N_SIMULATIONS)


if (sim_status$n_to_run > 0) {
    message(glue("Starting {sim_status$n_to_run} new simulations on {N_CORES} cores..."))
    
    future::plan(future.callr::callr, workers = N_CORES)
    
    # --- Run Simulations in Parallel ---
    furrr::future_walk(
        .x = sim_status$iters_to_run,
        .f = ~run_simulation_iteration(
            iter_num = .x,
            base_path = OUTPUT_PATH,
            full_data = count_df,
            model = model
        ),
        .options = furrr_options(seed = TRUE),
        .progress = TRUE
    )
    
    message(glue("Finished {sim_status$n_to_run} simulations."))
} else {
    message("All simulations were already complete. No new runs started.")
}

message("Simulation process finished.")
```