---
Title: Results of Simulations
---

In this document we summarise and visualize the results of the simulations from step 6.

## Setup

### Load packages

```{r}
library(fs)
library(tidyverse)
library(arrow)
library(survival)
library(patchwork)
```

### Set Threshold for P(Benefit)

```{r}
threshold <- 0.95
```

### Functions

```{r}
source("./functions/jackknife_mcse.R")
source("./functions/summarise-posterior.R")
source("./functions/summarise-iterations.R")
source("./functions/plot-rhat.R")
source("./functions/plot-p-benefit.R")
source("./functions/plot-posterior-distribution.R")
source("./functions/load-draws.R")
source("./functions/load-rhat-values.R")
```

## Markov Models

### H0

#### Load draws + diagnostics
```{r}
# H0 misspecified, no clustering
path <- "./2-hosp/output/sim-H0-misspec/"
misspec_h0 <- load_draws(path)
rhat_misspec_h0 <- load_rhat_values(path = path)

path <- "./2-hosp/output/sim-H0-correct-spec/"
true_h0 <- load_draws(path)
rhat_true_h0 <- load_rhat_values(path = path)

path <- "./2-hosp/output/sim-H0-cluster-misspec/"
misspec_cluster_h0 <- load_draws(path)
rhat_misspec_clsuter_h0 <- load_rhat_values(path = path)
```

#### Convergence diagnostics
```{r}
rhat_plot_misspec_h0 <- rhat_misspec_h0 |> 
    plot_rhat(y = `beta.1.`)

rhat_plot_true_h0 <- rhat_true_h0 |> 
    plot_rhat(y = `beta.1.`)

rhat_plot_misspec_cluster_h0 <- rhat_misspec_clsuter_h0 |> 
    plot_rhat(y = `beta.1.`)

rhat_plot_misspec_h0 + rhat_plot_true_h0 + rhat_plot_misspec_cluster_h0

# ggsave(filename = "./img/fig-rhat-markov.svg", width = 10, height = 5)
```

#### Plot probability of benefit

```{r}
p1 <- misspec_h0 |> 
    plot_p_benefit(param = `tx=1`)

p2 <- true_h0 |> 
    plot_p_benefit(param = `tx=1`)

p3 <- misspec_cluster_h0 |> 
    plot_p_benefit(param = `tx=1`)

p1 + p2 + p3

# ggsave(filename = "./img/fig-pbenefit-markov.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
p1 <- misspec_h0 |> 
  plot_posterior_distribution(x_variable = `tx=1`) +
  labs(title = "Misspecified, no random intercept per patient.")

p2 <- true_h0 |> 
  plot_posterior_distribution(x_variable = `tx=1`) +
  labs(title = "Analysis model = data generating model")

p1 + p2

# ggsave(filename = "./img/fig-posteriors-markov.svg", width = 10, height = 5)
```

```{r}
summary_misspec_h0 <- misspec_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = `tx=1`, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_true_h0 <- true_h0 |>
    group_by(iter) |> 
    summarise_posterior(
        param_name = `tx=1`, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_misspec_cluster_h0 <- misspec_cluster_h0 |>
    group_by(iter) |> 
    summarise_posterior(
        param_name = `tx=1`, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

# create dataframe with for comparison
all_summaries <- bind_rows(
  'true_model' = summary_true_h0,
  'misspec_model' = summary_misspec_h0,
  'misspec_cluster_model' = summary_misspec_cluster_h0,
  .id = 'model_type'
)

# all_summaries |> tinytable::tt()

formatted_summary <- all_summaries |> 
  # transmute creates new columns and drops the old ones
  transmute(
    Model = model_type,
    # sprintf() formats the numbers and combines them into one string
    `Point Estimate` = sprintf("%.3f (%.3f)", mean_point_estimate, mcse_point_estimate),
    `Rejection Rate` = sprintf("%.3f (%.3f)", prob_rejection, mcse_prob_rejection),
    Bias             = sprintf("%.3f (%.3f)", mean_bias, mcse_bias),
    Coverage         = sprintf("%.3f (%.3f)", mean_coverage, mcse_coverage),
    `Empirical SE`   = sprintf("%.3f (%.3f)", emp_SE, mcse_emp_SE),
    `Model SE`       = sprintf("%.3f (%.3f)", mean_modSE, mcse_modSE)
  )

formatted_summary |> knitr::kable()

saveRDS(all_summaries, file = "./2-hosp/output/sim-summary.rds")
```

### HA

#### Load draws + diagnostics
```{r}
# H0 misspecified, no clustering
path <- "./2-hosp/output/sim-HA-09/"
ha_09 <- load_draws(path)
rhat_ha_09 <- load_rhat_values(path = path)

path <- "./2-hosp/output/sim-HA-08/"
ha_08 <- load_draws(path)
rhat_ha_08 <- load_rhat_values(path = path)

path <- "./2-hosp/output/sim-HA-07/"
ha_07 <- load_draws(path)
rhat_ha_07 <- load_rhat_values(path = path)
```

#### Convergence

```{r}
rhat_09 <- rhat_ha_09 |> 
    plot_rhat(y = `beta.1.`)

rhat_08 <- rhat_ha_08 |> 
    plot_rhat(y = `beta.1.`)

rhat_07 <- rhat_ha_07 |> 
    plot_rhat(y = `beta.1.`)

rhat_09 + rhat_08
```

#### Power

```{r}
summary_ha_09 <- ha_09 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = `tx`, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

length(unique(ha_09$iter))
summary_ha_09

summary_ha_08 <- ha_08 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = `tx`, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

length(unique(ha_08$iter))
summary_ha_08

setdiff(1:1000, unique(ha_08$iter))
```

## Count Regression

### Resampling H0

#### Load draws + diagnostics
```{r}
# H0 no interaction
path <- "./2-hosp/output/sim-resampling-H0-neg-binom/"
neg_binom_h0 <- load_draws(path)
rhat_neg_binom_h0 <- load_rhat_values(path = path)

ate_neg_binom_h0 <- load_draws(path = path, glob = "*/ATE_draws.rds")
```

#### Convergence diagnostics
```{r}
rhat_plot_neg_binom_h0 <- rhat_neg_binom_h0 |> 
    plot_rhat(y = b_tx1)

# ggsave(filename = "./img/fig-rhat-markov.svg", width = 10, height = 5)
```

### Plot probability of benefit

```{r}
p1 <- neg_binom_h0 |> 
    plot_p_benefit(param = b_tx1)

p2 <- ate_neg_binom_h0 |>
    plot_p_benefit(param = draw)

p1 + p2

# ggsave(filename = "./img/fig-pbenefit-markov.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_neg_binom_h0 <- neg_binom_h0 |> 
  plot_posterior_distribution(x_variable = b_tx1) + 
  scale_x_continuous(limits = c(-1, 1))

post_neg_binom_h0

ate_post_neg_binom_h0 <- ate_neg_binom_h0 |> 
  plot_posterior_distribution(x_variable = draw)

# ggsave(filename = "./img/fig-posteriors-markov.svg", width = 10, height = 5)
```

```{r}
summary_neg_binom_h0 <- neg_binom_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx1, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()
```

## DRS 26

Not actually 26 categories, but 1 to max(number of days not at home) + 1 worse for people who died. In our resample dataset that 15 categories.

### Resampling H0

#### Load draws + diagnostics
```{r}
# H0 rmsb (default dirichlet prior)
path <- "./2-hosp/output/sim-resampling-H0-drs26-rmsb/"
drs26_rmsb_h0 <- load_draws(path)
rhat_drs26_rmsb_h0 <- load_rhat_values(path = path)

# H0 brms (default priors)
path <- "./2-hosp/output/sim-resampling-H0-drs26-brms/"
drs26_brms_h0 <- load_draws(path)
rhat_drs26_brms_h0 <- load_rhat_values(path = path)
```

#### Convergence diagnostics
```{r}
rhat_plot_drs26_rmsb_h0 <- rhat_drs26_rmsb_h0 |> 
    plot_rhat(y = `tx=1`)

rhat_plot_drs26_rmsb_h0

rhat_plot_drs26_brms_h0 <- rhat_drs26_brms_h0 |> 
    plot_rhat(y = b_tx1)

rhat_plot_drs26_brms_h0

# ggsave(filename = "./img/fig-rhat-markov.svg", width = 10, height = 5)
```

### Plot probability of benefit

```{r}
p1 <- drs26_rmsb_h0 |> 
    plot_p_benefit(param = `tx=1`)

p1

p2 <- drs26_brms_h0 |>
    plot_p_benefit(param = b_tx1)
p2

# ggsave(filename = "./img/fig-pbenefit-markov.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_drs26_rmsb_h0 <- drs26_rmsb_h0 |> 
  plot_posterior_distribution(x_variable = `tx=1`)

post_drs26_rmsb_h0

post_drs26_brms_h0 <- drs26_brms_h0 |> 
  plot_posterior_distribution(x_variable = b_tx1)

post_drs26_brms_h0

# ggsave(filename = "./img/fig-posteriors-markov.svg", width = 10, height = 5)
```

```{r}
summary_drs26_rmsb_h0 <- drs26_rmsb_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = `tx=1`, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_drs26_brms_h0 <- drs26_brms_h0 |>
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx1, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()
```

