---
Title: Primary Estimand
---

## Setup

```{r}
library(furrr)
library(tidyverse)
library(data.table)
library(arrow)
library(rmsb)
library(here)
library(Hmisc)
library(patchwork)
```

## Load model

We use simulated data with constant OR of 0.8.

```{r}
set.seed(1234)
df_full <- read_parquet(here("2-hosp", "sim-data", "ha_08.parquet")) |> 
    select(id = pat_id, tx, y, yprev, ypprev, week, ecog_fstcnt, diagnosis)

tx_id <- sample(unique(df_full$id[df_full$tx == 1]), 180, replace = FALSE)
soc_id <- sample(unique(df_full$id[df_full$tx == 0]), 90, replace = FALSE)

df <- df_full |> 
  filter(id %in% c(tx_id, soc_id))

# we need to change the patient id to be sequential starting from 1
# otherwise the random effects gammas won't match
df <- df |> 
  arrange(id, week) |> 
  mutate(id = as.integer(factor(id)))
```

## Fit model

```{r}
if(FALSE){
  options(mc.cores = 6)
  model <-       
      blrm(
              formula = y  ~ tx + rcs(week, 4) + yprev + ypprev + week %ia% yprev + week %ia% ypprev + ecog_fstcnt + diagnosis + cluster(id),
              data = df,
              refresh = 5,
              iter = 2000,
              chains = 4,
              seed = 1234,
              loo = FALSE,
              method = "sampling",
              backend = "cmdstan"
          )
}
```

## Functions

### Predictions conditional on random effects

```{r}
predict_blrm_manual <- function(fit, newdata, include_re = FALSE) {
  # --- 0) Prerequisite Checks ---
  if (!is.data.frame(newdata)) {
    stop("`newdata` must be a data frame.")
  }
  if (include_re && !"id" %in% colnames(newdata)) {
    stop("`newdata` must contain a 'id' column when include_re=TRUE")
  }
  
  N_new <- nrow(newdata)

  # --- 1) Get the design matrix for the new data ---
  X <- rms::predictrms(fit, newdata = newdata, type = 'x')

  # --- 2) Extract posterior draws and model information ---
  draws <- fit$draws
  ndraws <- nrow(draws)
  ns <- fit$non.slopes
  K <- ns + 1
  
  intercept_draws <- draws[, 1:ns, drop = FALSE]
  beta_draws <- draws[, (ns + 1):ncol(draws), drop = FALSE]

  # --- 3) Calculate the fixed-effect linear predictor (eta_fixed) ---
  eta_fixed <- beta_draws %*% t(X)

  # --- 4) Get the random effect draws (u_draws) if requested ---
  u_draws <- matrix(0, nrow = ndraws, ncol = N_new)
  
  if (include_re) {
    if (length(unique(newdata$id)) > 1) {
      stop("`newdata` contains multiple unique patient IDs. This version is restricted to one at a time.")
    }
    if (!requireNamespace("posterior", quietly = TRUE)) {
      stop("The 'posterior' package is required to extract random effect draws.")
    }
    sf <- rmsb::stanGet(fit)
    all_gamma_draws <- posterior::as_draws_matrix(sf$draws(variables = "gamma"))
    patient_ids <- newdata$id
    u_draws <- all_gamma_draws[, patient_ids, drop = FALSE]
  }

  # --- 5) Calculate the total linear predictor (eta) ---
  stopifnot(all(dim(eta_fixed) == dim(u_draws)))
  eta <- eta_fixed + u_draws

  # --- 6) Calculate probabilities for each observation and each draw ---
  p_cat_draws_list <- vector("list", N_new) 
  
  for (i in 1:N_new) {
    eta_i <- eta[, i]
    s_ge <- plogis(sweep(intercept_draws, 1, eta_i, "+"))
    
    p_cat_draws <- matrix(NA_real_, nrow = ndraws, ncol = K)
    p_cat_draws[, 1] <- 1 - s_ge[, 1]
    if (K > 2) {
      for (kk in 2:(K - 1)) {
        p_cat_draws[, kk] <- s_ge[, kk - 1] - s_ge[, kk]
      }
    }
    p_cat_draws[, K] <- s_ge[, K - 1]
    colnames(p_cat_draws) <- paste0("y=", 1:K)
    
    p_cat_draws_list[[i]] <- p_cat_draws
  }

  # --- 7) Reshape draws into a single 3D array ---
  draws_array <- array(
    data = NA_real_,
    dim = c(ndraws, N_new, K)
  )

  for (i in 1:N_new) {
    draws_array[, i, ] <- p_cat_draws_list[[i]]
  }

  dimnames(draws_array) <- list(
    NULL,
    rownames(newdata),
    colnames(p_cat_draws_list[[1]])
  )

  return(draws_array)
}
```

### Calculate SOPs

```{r}
#' Calculate Second-Order State Occupation Probabilities
#'
#' This function calculates the unconditional state occupation probabilities for a
#' second-order Markov model.
#'
#' @param model_fit The fitted model object (rmsb or brms)
#' @param baseline_data A single-row data frame containing the initial states and
#'   all baseline predictor values required by the model.
#' @param col_yprev A string with the column name for the state at t-1 (e.g., "yprev").
#' @param col_ypprev A string with the column name for the state at t-2 (e.g., "ypprev").
#' @param col_time A string with the column name that represents time in the model (e.g., "week").
#' @param times A numeric vector specifying the time points for calculation (e.g., `1:26`).
#' @param n_states An integer for the number of levels in the ordinal outcome.
#' @param absorbing_states A numeric vector listing the absorbing states.
#'
#' @return A list containing two arrays:
#'   \item{unconditional}{A 3D array `(draws, states, time)` with the unconditional state probabilities.}
#' @export

sop_markov_second_order <- function(model_fit,
                                    baseline_data,
                                    col_yprev,
                                    col_ypprev,
                                    col_time,
                                    times,
                                    n_states,
                                    absorbing_states,
                                    include_re = FALSE) {

  # --- Detect model type and get number of draws ---
  if (inherits(model_fit, "brmsfit")) {
    detected_type <- "brms"
    if (!requireNamespace("marginaleffects", quietly = TRUE) || !requireNamespace("posterior", quietly = TRUE)) {
      stop("Packages 'marginaleffects' and 'posterior' are required for brms models.")
    }
    n_draws <- posterior::ndraws(model_fit)
  } else if (inherits(model_fit, "blrm")) {
    detected_type <- "rmsb"
    n_draws <- nrow(model_fit$draws)
  } else {
    stop("Unsupported model object class. Please provide a model from 'brms' (class 'brmsfit') or 'rmsb' (class 'blrm').")
  }
  
  # --- Parameters ---
  M <- n_states
  T_horizon <- max(times)

  # --- Initial Conditions ---
  initial_state_t_minus_1 <- as.numeric(baseline_data[[col_ypprev]])
  initial_state_t_0 <- as.numeric(baseline_data[[col_yprev]])

  # --- Data Structures to Store Results ---
  P_joint_bayes <- array(0, dim = c(n_draws, M, M, T_horizon + 1),
                         dimnames = list(paste0("draw_", 1:n_draws),
                                         paste0("t-1=", 1:M),
                                         paste0("t=", 1:M),
                                         paste0("time=", 0:T_horizon)))

  P_uncond_bayes <- array(0, dim = c(n_draws, M, T_horizon + 1),
                          dimnames = list(paste0("draw_", 1:n_draws),
                                          paste0("State_", 1:M),
                                          paste0("t=", 0:T_horizon)))

  # Set the known probability at t=0
  P_joint_bayes[, initial_state_t_minus_1, initial_state_t_0, 1] <- 1.0
  P_uncond_bayes[, initial_state_t_0, 1] <- 1.0

  # --- Create the static grid for batch predictions ---
  history_grid <- expand.grid(
    h = factor(1:M, levels = 1:M), # X_{t-2}
    j = factor(1:M, levels = 1:M)  # X_{t-1}
  )
  names(history_grid) <- c(col_ypprev, col_yprev)

  # --- Dynamically create the full prediction dataset ---
  # Identify static covariates by excluding time-varying/history columns
  cols_to_exclude <- c(col_yprev, col_ypprev, col_time)
  static_covariate_names <- setdiff(names(baseline_data), cols_to_exclude)
  static_covariates <- baseline_data[, static_covariate_names, drop = FALSE]

  # Combine static covariates with the grid of all possible histories
  prediction_data <- cbind(static_covariates, history_grid)

  # --- Main Iteration Loop ---
  for (t in times) {
    # Identify predictable histories
    predictable_idx <- which(
      !(prediction_data[[col_ypprev]] %in% absorbing_states) &
      !(prediction_data[[col_yprev]] %in% absorbing_states)
    )
    n_predictable <- length(predictable_idx)

    # Subset the data for prediction and set the current time
    prediction_data_subset <- prediction_data[predictable_idx, ]
    prediction_data_subset[[col_time]] <- t


    pred_array_3d <- NULL

    if (detected_type == "brms") {
      # refactor previous states to remove absorbing states
      prediction_data_subset[[col_yprev]] <- factor(prediction_data_subset[[col_yprev]], levels = c(1:M)[-absorbing_states])
      prediction_data_subset[[col_ypprev]] <- factor(prediction_data_subset[[col_ypprev]], levels = c(1:M)[-absorbing_states])  
      re_arg <- if (include_re) NULL else NA
      brms_pred_long <- marginaleffects::predictions(model_fit, 
                                                     newdata = prediction_data_subset, 
                                                     type = "response", 
                                                     re_formula = re_arg) |>
                        marginaleffects::posterior_draws()
      
      pred_array_3d <- array(0, dim = c(n_draws, n_predictable, M))
      indices <- cbind(brms_pred_long$drawid, 
                       brms_pred_long$rowid, 
                       as.numeric(brms_pred_long$group))
      pred_array_3d[indices] <- brms_pred_long$draw
      
    } else { # detected_type == "rmsb"
    # dimension 1 = 4000 draws
    # dimension 2 = X_{t-1} and X_{t-2} combinations (rows from prediction_data_subset)
    # dimension 3 = cumulative probabilities (e.g., 4 columns for 5 states)
      pred_array_3d <- predict_blrm_manual(model_fit, prediction_data_subset, include_re = include_re)
    }

    state_probs_all_draws <- do.call(rbind, lapply(1:dim(pred_array_3d)[1], function(d) pred_array_3d[d, , ]))

    # --- Assemble transition probabilities for each draw ---
    # Create a 4D array to hold all transition probabilities: P(X_t=l | X_{t-2}=h, X_{t-1}=j)
    # Dims: (draw, h, j, l)
    # So [, 1, 1, 1] give all draws for P(l = 1 | h = 1, j = 1)
    # [, 1, 1, ] for all draws for P(l | h = 1, j = 1) (rows must sum to 1)
    cond_probs_4d_array <- array(0, dim = c(n_draws, M, M, M))

    for (d in 1:n_draws) {
      # Create a placeholder matrix for this specific draw
      # each row is combination of (h, j), each column is l (rows must sum to 1)
      # this is for one draw only
      cond_probs_matrix_d <- matrix(0, nrow = M * M, ncol = M)
      
      # Handle absorbing states (P(j->j) = 1 if j is absorbing)
      absorbing_idx <- which(prediction_data[[col_yprev]] %in% absorbing_states)
      if (length(absorbing_idx) > 0) {
        absorbing_state <- as.numeric(as.character(prediction_data[[col_yprev]][absorbing_idx]))
        rows_and_cols <- cbind(absorbing_idx, absorbing_state)
        cond_probs_matrix_d[rows_and_cols] <- 1.0
      }
      
      # Get state probabilities for the current draw
      start_row <- (d - 1) * n_predictable + 1
      end_row <- d * n_predictable
      draw_d_probs <- state_probs_all_draws[start_row:end_row, ]
      
      # Place them into the correct rows
      cond_probs_matrix_d[predictable_idx, ] <- draw_d_probs
      
      # Reshape and store in the final 4D array
      cond_probs_4d_array[d, , , ] <- array(cond_probs_matrix_d, dim = c(M, M, M))
    }

    # --- Update Probabilities ---
    # Get joint probabilities from the previous step for all draws: P(X_{t-2}, X_{t-1})
    prev_P_joint_all_draws <- P_joint_bayes[, , , t]
    current_P_joint_all_draws <- array(0, dim = c(n_draws, M, M))

    for (d in 1:n_draws) {
      # This is P(X_{t-2}=h, X_{t-1}=j) for draw d
      prev_P_joint_d <- prev_P_joint_all_draws[d, , ]
      # P(X_t = l | X_{t-1} = j, X_{t-2} = h) for this draw
      cond_probs_d <- cond_probs_4d_array[d, , , ]
      
      # Law of Total Probability: P(X_t=l|h,j) * P(h,j)
      weighted_probs <- cond_probs_d * array(prev_P_joint_d, dim = dim(cond_probs_d))
      # Marginalize over the t-2 state (h) to get P(X_{t-1}, X_t)
      # = Summation over h (the state at t-2), resulting in P(X_{t-1}=j, X_t=l)
      current_P_joint_d <- apply(weighted_probs, c(2, 3), sum)
      
      current_P_joint_all_draws[d, , ] <- current_P_joint_d
      
      # Marginalize over the t-1 state (j) to get P(X_t)
      uncond_probs_t <- colSums(current_P_joint_d)
      # Consider normalizing  to prevent floating point error accumulation (not doing this for now)
      P_uncond_bayes[d, , t + 1] <- uncond_probs_t
      # P_uncond_bayes[d, , t + 1] <- uncond_probs_t / sum(uncond_probs_t)
    }
    
    P_joint_bayes[, , , t + 1] <- current_P_joint_all_draws
  }

  return(P_uncond_bayes)
}
```

### Marginalize over whole population

```{r}
write_SOP <- function(i, tx, baseline_df, path, model){
  
  row <- baseline_df[i, ]
  
  sops <-
  sop_markov_second_order(
    model_fit = model,
    baseline_data = data.frame(id = row$id, tx = tx, ecog_fstcnt = row$ecog_fstcnt, diagnosis = row$diagnosis, yprev = row$yprev, ypprev = row$ypprev),
    col_yprev = "yprev",
    col_ypprev = "ypprev",
    col_time = "week",
    times = 1:26,
    n_states = 5,
    absorbing_states = c(5),
    include_re = TRUE
  )
  
  sops <- reshape2::melt(sops, 
                   varnames = c("draw", "state", "time"), 
                   value.name = "probability")
  sops <- sops |> 
    mutate(
        time  = as.integer(sub("t=", "", time)),
        draw  = as.integer(sub("draw_", "", draw)),
        state = as.factor(sub("State_", "", state)),
        tx    = tx,
        i     = i ) |> 
    filter(time > 0)
  
  folder <- file.path(path, glue::glue("/marginalized_sop_{tx}"), glue::glue("/msop_{i}.parquet"))

  # Create the directory if it doesn't exist
  dir_to_create <- dirname(folder)
  if (!dir.exists(dir_to_create)) {
    dir.create(dir_to_create, recursive = TRUE)
  }
  
  arrow::write_parquet(
    x = sops,
    sink = folder
    )
}
```

## Marginalized SOPs

### Generate SOPs

```{r}
generate_SOPS <- FALSE
```

```{r}
path <- here("2-hosp", "output", "primary-estimand")
future::plan(future::multisession, workers=6)

baseline_df <- df |> 
    filter(week == 1) |> 
    select(id, week, ecog_fstcnt, diagnosis, yprev, ypprev) |> 
    distinct() |> 
    arrange(id)

# Control
if (generate_SOPS) {
    furrr::future_walk(1:nrow(baseline_df), \(x) write_SOP(x, 0, baseline_df, path, model = model),
                       .progress=TRUE,
                       .options = furrr_options(seed=TRUE))
}
# Treatment
if (generate_SOPS) {
    furrr::future_walk(1:nrow(baseline_df), \(x) write_SOP(x, 1, baseline_df, path, model = model),
                       .progress=TRUE,
                       .options = furrr_options(seed=TRUE))
}
```

### Read marginalized SOPs

```{r}
files <- fs::dir_ls(here("2-hosp", "output", "primary-estimand", "marginalized_sop_0"), glob = "*.parquet")

soc_df <- map(files, \(x) arrow::read_parquet(x), .progress=TRUE) |> rbindlist()

# Average SOPs over covariate settings for each state, day, and MCMC draw
soc_df <- soc_df |> 
  group_by(state, time, draw, tx) |>
  summarise(sop = mean(probability), .groups = "drop") |> 
  mutate(stat = as.factor(state))

files <- fs::dir_ls(here("2-hosp", "output", "primary-estimand", "marginalized_sop_1"), glob = "*.parquet")
sat_df <- map(files, \(x) arrow::read_parquet(x), .progress=TRUE) |> rbindlist()
sat_df <- sat_df |> 
  group_by(state, time, draw, tx) |>
  summarise(sop = mean(probability), .groups = "drop") |> 
  mutate(stat = as.factor(state))

marg_df <- rbind(soc_df, sat_df)
```

### Visualize Results

```{r}
plot_model <- marg_df |>
  mutate(tx = factor(tx)) |> 
  ggplot() +
  aes(x = time, y = sop) +
  ggdist::stat_lineribbon(aes(fill = state, linetype = tx),
                          linewidth = 0.6,
                          alpha = 0.5,
                          .width = c(0.95)) +
  scale_color_brewer(palette = "Dark2", guide="none") +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Study Week",
   y = "SOP",
   subtitle = "Marginalized state occupancy probabilities",
   fill = "State",
   linetype = "Treatment")
```

### Empirical SOPs


```{r}
df_forward <- read_parquet(here("2-hosp", "sim-data", "ha_08_forward.parquet")) |> 
  filter(pat_id %in% c(tx_id, soc_id))

empirical_sop <- df_forward |>
  group_by(tx, week) |> 
  count(y) |> 
  mutate(
    sop = n / sum(n),
    state = as.factor(y),
    tx = factor(tx)
  ) |> 
  ungroup()

plot_empirical <- empirical_sop |> 
  ggplot() +
  aes(x = week, y = sop) +
  geom_line(aes(color = state, linetype = tx), linewidth = 1) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Study Week",
       y = "SOP",
       subtitle = "Empirical state occupancy probabilities",
       color = "State",
       linetype = "Treatment") +
  theme(legend.position = "")

```

We can se that some empirical lines cross, but we KNOW that the treatment effect is constant and proportional, so this is just noise.
```{r}
plot_empirical + plot_model
```

## Primary Estimand

Difference is in the mean time spent out of hospital and alive (y = 1). 
```{r}
estimand_df <- marg_df |> 
  filter(state == "1") |> 
  group_by(tx, draw) |> 
  summarise(total_time = sum(sop), .groups = "drop") |> 
  pivot_wider(names_from = tx, values_from = total_time, names_prefix = "tx_") |> 
  mutate(diff = tx_1 - tx_0)

# plot
estimand_df |> 
  ggplot() +
  aes(x = diff) +
  ggdist::stat_halfeye() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Difference in mean weeks spent out of hospital and alive (SAT - SoC)",
       y = "Density",
       subtitle = "Posterior distribution of the primary estimand")
```