---
Title : Model for Simulation
---

Examining first vs. second order markov modoel and interactnig yprev with time or not.

## Setup

```{r}
library(tidyverse)
library(data.table)
library(arrow)
library(rms)
library(rmsb)
library(furrr)
library(VGAM)
library(Hmisc)
library(ordinal)
library(marginaleffects)
```

## Load data

```{r}
df <- read_parquet("./data/hosp_long_weeks.parquet") |> 
    mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE)
    ) |> 
      select(pat_id, tx, y, yprev, week, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)

df_ypprev <- read_parquet("./data/hosp_long_weeks_two_lags.parquet") |> 
    mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE),
        y = ordered(y)
    ) |> 
      select(pat_id, tx, y, yprev, ypprev, week, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)

df_daily <- read_parquet("./data/hosp_long_daily.parquet") |> 
  mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE),
        y = case_when(
          y == 4 ~ 3,
          y == 5 ~ 4,
          TRUE ~ y),
        yprev = case_when(
          yprev == "4" ~ "3",
          TRUE ~ yprev),
        ypprev = case_when(
          ypprev == "4" ~ "3",
          TRUE ~ ypprev)   
    ) |> 
      select(pat_id, tx, y, yprev, ypprev, day, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)
```

## Weekly States

### VGAM constrained

```{r}
d <- as.data.table(df_ypprev)
week_spl_m <- rcs(d$week, 4)

d$week_linear <- week_spl_m[, 1]
d$week_nonlinear_1 <- week_spl_m[, 2]
d$week_nonlinear_2 <- week_spl_m[, 3]

M <- 4

cons <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

# PO model, first order -------------------------------------------------------------------
fit_po <- vglm(
  y ~ tx + yprev + week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = TRUE, reverse = TRUE),
  data = d, 
  constraints = NULL
)

# pPO for linear time, first order ---------------------------------------------------------
fit_non_po <- vglm(
  y ~ tx + yprev + week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons
)

# Doesn't converge well even with constraints
AIC(fit_po); AIC(fit_non_po)

# --- non-PO with interaction of yprev and linear time ---
# constraints when we add interaction
cons_lin_int <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),
  "yprev:week_linear"  = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_non_po_lin_int <- vglm(
  y ~ tx + yprev * week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_lin_int
)

fit_po_lin_int <- vglm(
  y ~ tx + yprev * week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = TRUE, reverse = TRUE),
  data = d
)



# interaction of yprev with time improves fit by a lot
# non proportional effect of time does not --> we can likely drop this
AIC(fit_non_po); AIC(fit_po_lin_int); AIC(fit_non_po_lin_int)


# --- Second order, full PO vs partial PO ---

# constraints when we add interaction
cons_second <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "ypprev"            = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

# Doesn't converge well even with constraints
fit_second_non_po <- vglm(
  y ~ tx + yprev + ypprev + week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_second
)

fit_second_po <- vglm(
  y ~ tx + yprev + ypprev + week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = TRUE, reverse = TRUE),
  data = d
)

# second order fit is substantially better
AIC(fit_po); AIC(fit_second_po)

# second order with interaction of yprev and linear time -------------------------------------------
cons_second_lin_int <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "ypprev"            = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),
  "yprev:week_linear"  = cbind(PO_effect = 1),
  "ypprev:week_linear" = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_second_non_po_lin_int <- vglm(
  y ~ tx + (yprev + ypprev) * week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_second_lin_int
)

fit_second_po_lin_int <- vglm(
  y ~ tx + (yprev + ypprev) * week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = TRUE, reverse = TRUE),
  data = d
)

# strong improvement with interaction of yprev and ypprev with time
# non-proportional effect of time still doesn't improve fit much
AIC(fit_second_po); AIC(fit_second_po_lin_int); AIC(fit_second_non_po_lin_int)

# second order model, no interaction with time, but interaction between yprev and ypprev ----------------------
cons_second_yprev_int <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "ypprev"            = cbind(PO_effect = 1),
  "yprev:ypprev"      = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_second_non_po_yprev_int <- vglm(
  y ~ tx + yprev * ypprev + week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_second_yprev_int
)
# Results in intersecting linear predictors
```


Second order fit seems to be better than first order fit, interaction of yprev and ypprev with previous time improves fit by a lot. Non-linear modelling of time doesn't improve fit much.  --> stick to full PO model with interaction of yprev and ypprev with linear time component.