---
Title : Model for Simulation
---

Examining first vs. second order markov modoel and interactnig yprev with time or not.

## Setup

```{r}
library(tidyverse)
library(data.table)
library(arrow)
library(rms)
library(rmsb)
library(furrr)
library(VGAM)
library(Hmisc)
library(ordinal)
library(marginaleffects)
```

## Load data

```{r}
df <- read_parquet("./data/hosp_long_weeks.parquet") |> 
    mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE)
    ) |> 
      select(pat_id, tx, y, yprev, week, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)

df_ypprev <- read_parquet("./data/hosp_long_weeks_two_lags.parquet") |> 
    mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE),
        y = ordered(y)
    ) |> 
      select(pat_id, tx, y, yprev, ypprev, week, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)

df_daily <- read_parquet("./data/hosp_long_daily.parquet") |> 
  mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE),
        y = case_when(
          y == 4 ~ 3,
          y == 5 ~ 4,
          TRUE ~ y),
        yprev = case_when(
          yprev == "4" ~ "3",
          TRUE ~ yprev),
        ypprev = case_when(
          ypprev == "4" ~ "3",
          TRUE ~ ypprev)   
    ) |> 
      select(pat_id, tx, y, yprev, ypprev, day, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)
```

## Weekly States

### VGAM constrained

```{r}
d <- as.data.table(na.omit(df_ypprev))
week_spl_m <- rcs(d$week, 4)

d$week_linear <- week_spl_m[, 1]
d$week_nonlinear_1 <- week_spl_m[, 2]
d$week_nonlinear_2 <- week_spl_m[, 3]

M <- 4

cons <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

# PO model, first order -------------------------------------------------------------------
fit_po <- vglm(
  y ~ tx + yprev + week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = TRUE, reverse = TRUE),
  data = d
)

# pPO for linear time, first order ---------------------------------------------------------
fit_non_po <- vglm(
  y ~ tx + yprev + week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons
)

AIC(fit_po); AIC(fit_non_po)

# non-PO with interaction of yprev and linear time ------------------------------------------
# constraints when we add interaction
cons_lin_int <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),
  "yprev:week_linear"  = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_non_po_lin_int <- vglm(
  y ~ tx + yprev * week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_lin_int
)

# improves fit by a lot
AIC(fit_non_po); AIC(fit_non_po_lin_int)


# Second order, partial PO ------------------------------------------------------------------

# constraints when we add interaction
cons_second <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "ypprev"            = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_second_non_po <- vglm(
  y ~ tx + yprev + ypprev + week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_second
)

# second order fit is substantially better
AIC(fit_non_po); AIC(fit_second_non_po)

# second order with interaction of yprev and linear time -------------------------------------------
cons_second_lin_int <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "ypprev"            = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),
  "yprev:week_linear"  = cbind(PO_effect = 1),
  "ypprev:week_linear" = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_second_non_po_lin_int <- vglm(
  y ~ tx + (yprev + ypprev) * week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_second_lin_int
)

# dramatic improvement in fit
AIC(fit_second_non_po); AIC(fit_second_non_po_lin_int)

# second order model, no interaction with time, but interaction between yprev and ypprev ----------------------
cons_second_yprev_int <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "ypprev"            = cbind(PO_effect = 1),
  "yprev:ypprev"      = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_second_non_po_yprev_int <- vglm(
  y ~ tx + yprev * ypprev + week_linear + week_nonlinear_1 + week_nonlinear_2,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_second_yprev_int
)
# Results in intersecting linear predictors
```


Second order fit seems to be better than first order fit, interaction of yprev and ypprev with previous time improves fit by a lot. 
Interaction between yprev and ypprev results in intersecting linear predictors.

### Bootstrapping CIs

```{r}
# Ensure the patient_id column is keyed for maximum data.table speed
setkey(d, pat_id)

M <- length(levels(d$y))-1 # NUmber of thresholds

# Fit the original model to get the point estimate
week_spl_m <- rcs(d$week, 4)
d$week_linear <- week_spl_m[, 1]
d$week_nonlinear_1 <- week_spl_m[, 2]
d$week_nonlinear_2 <- week_spl_m[, 3]

cons_second_lin_int <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"        = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                 = cbind(PO_effect = 1),
  "yprev"              = cbind(PO_effect = 1),
  "ypprev"             = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),
  "yprev:week_linear"  = cbind(PO_effect = 1),
  "ypprev:week_linear" = cbind(PO_effect = 1),
  "ecog_fstcnt"        = cbind(PO_effect = 1),
  "diagnosis"          = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_second_non_po_lin_int <- vglm(
  y ~ tx + (yprev + ypprev) * week_linear + week_nonlinear_1 + week_nonlinear_2 + ecog_fstcnt + diagnosis,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_second_lin_int
)

original_coefs <- coef(fit_second_non_po_lin_int)
num_coefs <- length(original_coefs)

# This function is designed to be called repeatedly inside a loop.
run_single_bootstrap_all <- function(original_data, unique_ids, constraints, n_coef) {
  
  boot_ids <- sample(unique_ids, size = length(unique_ids), replace = TRUE)
  
  boot_d <- original_data[J(boot_ids), allow.cartesian = TRUE]

  week_spl_m_boot <- tryCatch(rcs(boot_d$week, 4), error = function(e) NULL)
  
  if (is.null(week_spl_m_boot)) return(rep(NA_real_, n_coef))
  
  boot_d$week_linear      <- week_spl_m_boot[, 1]
  boot_d$week_nonlinear_1 <- week_spl_m_boot[, 2]
  boot_d$week_nonlinear_2 <- week_spl_m_boot[, 3]
  
  boot_coef_vector <- tryCatch({
    fit_boot <- vglm(
      y ~ tx + (yprev + ypprev) * week_linear + week_nonlinear_1 + week_nonlinear_2 + ecog_fstcnt + diagnosis,
      family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
      data = boot_d,
      constraints = constraints
    )

    coef(fit_boot)
  }, error = function(e) {
    rep(NA_real_, n_coef)
  })
  
  return(boot_coef_vector)
}

B <- 1000
patient_ids <- unique(d$pat_id)

t1 <- Sys.time()
boot_results_matrix <- sapply(1:B, function(i) {
  if (i %% 100 == 0) cat("..", i)
  run_single_bootstrap_all(
    original_data = d, 
    unique_ids = patient_ids, 
    constraints = cons_second_lin_int,
    n_coef = num_coefs
  )
})
t2 <- Sys.time()
cat("Bootstrapping took", round(difftime(t2, t1, units = "mins"), 2), "minutes\n")

# Check for failed iterations (any column with all NAs)
failed_boots <- sum(apply(boot_results_matrix, 2, function(col) all(is.na(col))))
ci_percentile_matrix <- t(apply(boot_results_matrix, 1, quantile, probs = c(0.025, 0.975), na.rm = TRUE))

# Combine original estimates with CIs into a single data frame for printing
results_df <- data.frame(
  Parameter = names(original_coefs),
  Estimate = original_coefs,
  CI_Lower = ci_percentile_matrix[, 1],
  CI_Upper = ci_percentile_matrix[, 2]
)
row.names(results_df) <- NULL # Clean up row names
```

Bootstrapping SEs takes longer than fitting a Bayesian model. 





### Ordinal package
```{r}
# PO for time
fit_first_po <- clm(
  formula = y ~ rcs(week, 4) + yprev,
  link = "logit",
  data = na.omit(df_ypprev) |> mutate(y = factor(y, ordered = TRUE))
)

# PO for time with linear interaction
fit_first_po_lin_int <- clm(
  formula = y ~ rcs(week, 4) + yprev + yprev %ia% week,
  link = "logit",
  data = na.omit(df_ypprev) |> mutate(y = factor(y, ordered = TRUE))
)

AIC(fit_first_po); AIC(fit_first_po_lin_int)

# Better fit with linear interaction

# Non-PO for time for linear term only
fit_first_nonpo <- clm(
  formula = y ~ rcs(week, 4) + yprev,
  nominal = ~ week,
  link = "logit",
  data = na.omit(df_ypprev) |> mutate(y = factor(y, ordered = TRUE))
)

AIC(fit_first_po); AIC(fit_first_nonpo)

# Fit with non-PO for time is a lot better --> keep that

# Non-PO for time with linear interaction
fit_first_nonpo_lin_int <- clm(
  formula = y ~ rcs(week, 4) + yprev + yprev %ia% week,
  nominal = ~ week,
  link = "logit",
  data = na.omit(df_ypprev) |> mutate(y = factor(y, ordered = TRUE))
)

AIC(fit_first_nonpo); AIC(fit_first_nonpo_lin_int)
# Fit for interaction with time is marginally better, not sure if that's enough to matter

# Remove interaction and add second lag
fit_second_nonpo <- clm(
  formula = y ~ rcs(week, 4) + yprev + ypprev,
  nominal = ~ week,
  link = "logit",
  data = na.omit(df_ypprev) |> mutate(y = factor(y, ordered = TRUE))
)
AIC(fit_first_nonpo); AIC(fit_second_nonpo)

# fit with second lag is substantially better than first order fit

fit_second_nonpo_yprev_int <- clm(
  formula = y ~ rcs(week, 4) + yprev + ypprev,
  nominal = ~ week,
  link = "logit",
  data = na.omit(df_ypprev) |> mutate(y = factor(y, ordered = TRUE))
)

AIC(fit_second_nonpo); AIC(fit_second_nonpo_yprev_int)

# Add interaction to to second order fit
fit_second_nonpo_lin_int <- clm(
  formula = y ~ rcs(week, 4) + yprev + ypprev + yprev %ia% week + ypprev %ia% week,
  nominal = ~ week,
  link = "logit",
  data = na.omit(df_ypprev) |> mutate(y = factor(y, ordered = TRUE))
)

AIC(fit_second_nonpo); AIC(fit_second_nonpo_lin_int)
# interaction again improves fit

# add random intercept and check SD
fit_second_nonpo_ri <- clmm2(
  location = y ~ rcs(week, 4) + yprev + ypprev + yprev %ia% week + ypprev %ia% week,
  nominal = ~ week,
  random = id,
  link = "logistic",
  data = na.omit(df_ypprev) |> mutate(y = factor(y, ordered = TRUE), id = factor(pat_id)),
  control = clmm2.control(
    method = "nlminb")
)

fit_second_nonpo_ri
```

```{r}
model_list <- list(
  "fit_first_po" = fit_first_po,
  "fit_first_po_lin_int" = fit_first_po_lin_int,
  "fit_first_nonpo" = fit_first_nonpo,
  "fit_first_nonpo_lin_int" = fit_first_nonpo_lin_int,
  "fit_second_nonpo" = fit_second_nonpo,
  "fit_second_nonpo_lin_int" = fit_second_nonpo_lin_int
)

results_df <- data.frame(
  Step = 1:length(model_list),
  Description = c(
      "Baseline 1st-order PO",
      "PO w/ time interaction",
      "1st-order Non-PO (NPO)",
      "NPO w/ time interaction",
      "2nd-order NPO",
      "2nd-order NPO w/ interactions"
  ),
  AIC = sapply(model_list, AIC)
)

results_df <- data.frame(
  Description = c(
      "Baseline 1st-order PO",
      "PO w/ time interaction",
      "1st-order Non-PO for time (NPO)",
      "NPO w/ time interaction",
      "2nd-order NPO",
      "2nd-order NPO w/ interactions"
  ),
  AIC = sapply(model_list, AIC)
)

results_df |> knitr::kable()
```

## Daily States

```{r}
fit_first <- vglm(
    ordered(y)  ~ rcs(day, 4) + yprev,
    cumulative(reverse = TRUE, parallel = FALSE ~  rcs(day, 4)),
    data = na.omit(df_daily))

fit_second <- vglm(
    ordered(y)  ~ rcs(day, 4) + yprev + ypprev,
    cumulative(reverse = TRUE, parallel = FALSE ~  rcs(day, 4)),
    data = na.omit(df_daily))

lrtest(fit_second, fit_first)

AIC(fit_first); AIC(fit_second)
```

- Also for daily data second order fit seems to be better