---
Title : Model for Simulation
---

## Setup

```{r}
library(furrr)
library(tidyverse)
library(data.table)
library(arrow)
library(rms)
library(rmsb)
```

## Load data

```{r}
df_ypprev <- read_parquet("./data/hosp_long_weeks_two_lags.parquet") |> 
    mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE),
        yprev  = factor(yprev,  levels = 1:4, ordered = FALSE),
        ypprev = factor(ypprev, levels = 1:4, ordered = FALSE),
        y = ordered(y)
    ) |> 
      select(pat_id, tx, y, yprev, ypprev, week, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)

d <- df_ypprev
```

Do we want to fit the model?
```{r}
fit <- FALSE
```

## Model fitting

We fit a model without a treatment parameter. We will randomly assign a treatment effect during the simulation of the dataset.
We will use the same for model for HA, but simulate a treatment effect during the simulation of the dataset.

```{r}
if (fit) {
    Sys.setenv(RSTUDIO = 1)
    
    dd <- datadist(d)
    options(datadist = 'dd')
    options(mc.cores = 4)
    
  fit_h0 <- blrm(
    y ~ rcs(week, 5) + yprev + ypprev + week %ia% yprev + week %ia% ypprev + ecog_fstcnt + gender + diagnosis + rcs(albumin, 3) + rcs(c_reactive_protein, 3) + rcs(age,3),
    data = d,
    refresh = 5,
    iter = 2000,
    chains = 4,
    seed = 1234,
    method = "sampling",
    backend = "cmdstan"
  )
  saveRDS(fit_h0, "./2-hosp/models/fit-hosp-null.rds")


  Sys.setenv(RSTUDIO = 0)
} else{
  fit_h0 <- readRDS("./2-hosp/models/fit-hosp-null.rds")
}

fit_h0

# stanDx(fit_h0)
# stanDxplot(fit_h0)
```

## Simulations function

### Simulate baseline data

```{r}
N <- 1000
baseline_data <- d |> 
  filter(week == 1) |> 
  select(yprev, ypprev, ecog_fstcnt, gender, diagnosis, age, albumin, c_reactive_protein) |> 
  slice_sample(n = N, replace = TRUE) |> 
  mutate(
    pat_id = row_number(),
    tx = sample(c(0, 1), size = N, replace = TRUE),
    x = rnorm(N) # add patient specific offset
    ) |> 
  data.frame()

# x_effects <- seq(-1, 1, by = 1)
times <- 1:26
# intercept and parameters
ints <- fit_h0$draws[, 1:4] |> apply(2, median)
#reverse the order for easier use later
ints <- rev(ints)
```

### Simulation Function

```{r}
predict_new_y <- function(model, t, data, parameter = 0, x_effect = 0, ints) {
    data$week <- as.integer(t)
    lp <- predict(model, newdata = data, type = "lp")[1]$linear.predictors
    
    lp <- lp + parameter * data$tx + x_effect * data$x
    cut_points <- outer(ints, lp, FUN = "+")

    cumulative_probs <- t(plogis(cut_points))
    prob_matrix <- cbind(cumulative_probs, 1) - cbind(0, cumulative_probs)
    colnames(prob_matrix) <- 1:ncol(prob_matrix)

    y_new_active <- apply(prob_matrix, 1, function(p) {
            sample(5:1, size = 1, prob = p)
        })

    return(y_new_active)
}
```

There's still a mistake. Transitions are almost not happening. Need to debug more.
```{r}
y <- matrix(NA, ncol = 26, nrow = N)
# Loop over each time point (week)
for(i in 1:26) { # Loop over the full 26 weeks
  
  active_patients_idx <- which(baseline_data$yprev != "5")
  inactive_patients_idx <- which(baseline_data$yprev == "5")
  
  # For inactive patients, carry the '5' forward
  if (length(inactive_patients_idx) > 0) {
    y[inactive_patients_idx, i] <- 5
  }
  
  # For active patients, run the prediction function
  if (length(active_patients_idx) > 0) {
    # Subset data for only the active patients
    active_data <- baseline_data[active_patients_idx, ]
    
    # Predict new outcomes for the active group
    y_new_active <- predict_new_y(
      model = fit_h0,
      t = i,
      data = active_data,
      parameter = 0,
      x_effect = 0,
      ints = ints
    )
    
    # Assign the new outcomes to the correct rows in the main 'y' matrix
    y[active_patients_idx, i] <- y_new_active
  }
  
  # 4. Update history variables for ALL patients for the next iteration
  # ypprev gets the old value of yprev
  baseline_data$ypprev <- baseline_data$yprev
  # yprev gets the new value from the current week's simulation
  baseline_data$yprev <- as.character(y[, i])
}
```

## Apply simulation function

Create baseline data
```{r}
X_base <- data.frame(
    tx = sample(c(0, 1), size = N, replace = TRUE),
    X = rnorm(N),
    yprev = sample(c("1", "2", "3", "4"), N, replace = TRUE,  prob = c(0.8, 0.1, 0.05, 0.05))
) |>
mutate(patient_id = row_number())
```

Generate datasets

```{r}
for (current_x_effect in x_effects) {
    
    cat(paste0("--- Starting simulation for X effect = ", current_x_effect, " ---\n"))

    # a. Set the model parameters for the current scenario.
    extra <- c(
        `yprev=2` = log(10), 
        `yprev=3` = log(3), 
        `yprev=4` = log(40),
        week      = log(0.95)
    )
    extra["X"] <- current_x_effect

    # b. Initialize the matrix to store patient states over time.
    state_matrix <- matrix(
        as.numeric(X_base$yprev), 
        nrow = N, 
        ncol = length(times) + 1,
        dimnames = list(X_base$patient_id, 0:28)
    )

    # c. Run the simulation over the time points.
    for (t in times) {
        yprev <- state_matrix[, t]
        active_idx <- which(yprev != 5)
        
        if (length(active_idx) == 0) {
          if (t <= max(times)) {
            state_matrix[, (t + 1):ncol(state_matrix)] <- 5
          }
          break
        }
        
        yprev_active <- yprev[active_idx]
        X_active <- X_base[active_idx, ]
        
        lp <- mapply(
            FUN = g,
            yprev = yprev_active,
            tx_val = X_active$tx,
            x_val = X_active$X,
            MoreArgs = list(t = t, parameter = parameter, extra_params = extra)
        )
        
        thresholds_matrix <- outer(ints, lp, FUN = "+")
        cum_probs <- t(plogis(thresholds_matrix))
        prob_matrix <- cbind(cum_probs, 1) - cbind(0, cum_probs)
        
        y_new_active <- apply(prob_matrix, 1, function(p) {
            sample(5:1, size = 1, prob = p)
        })
        
        state_matrix[, t + 1] <- yprev
        state_matrix[active_idx, t + 1] <- y_new_active
    }

    # d. Format the results into a long data frame.
    final_data <- as.data.frame(state_matrix) |>
        tibble::rownames_to_column(var = "patient_id") |>
        tidyr::pivot_longer(
            cols = -patient_id,
            names_to = "time",
            values_to = "state"
        ) |>
        mutate(
            patient_id = as.integer(patient_id),
            time = as.integer(time)
        ) |>
        left_join(X_base |> rename(yprev_initial = yprev), by = "patient_id") |>
        arrange(patient_id, time) |>
        group_by(patient_id) |>
        mutate(yprev = lag(state)) |>
        ungroup()

    # e. Save the final data to a Parquet file.
    file_name <- sprintf("simulation_x_effect_%.1f.parquet", current_x_effect)
    write_parquet(final_data, file_name)
    
    cat(paste0("Saved data to '", file_name, "'\n\n"))
}

cat("--- All simulations complete. ---\n")

```


