---
Title: Simulation of different scenarios
---

## Setup

```{r}
library(tidyverse)
library(data.table)
library(tidybayes)
library(rmsb)
library(rms)
library(Hmisc)
library(arrow)
library(VGAM)
library(patchwork)

set.seed(1234)
```

## Load model and dataset

```{r}
df <- read_parquet("./data/hosp_long_weeks.parquet")  |> 
    mutate(ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE))

model <- readRDS("./2-hosp/models/model_hosp_pol.rds")

```

## Extract median of posterior draws

```{r}
draws <- as.data.table(model$draws) 

coef <- draws |> 
  summarise_all(median)

coef
```

## Baseline-Covariates

```{r}
N <- 250000 # Simulation size

baseline_w <- df |> 
    filter(week == 1) |> 
    select(pat_id, yprev,ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)

i <- sample(1:nrow(baseline_w), N, replace = TRUE)
sim_baseline_data <- baseline_w[i, ] |> 
  mutate(pat_id = row_number())

init <- sim_baseline_data$yprev # initial states to be passed to g function

baseline_formula <- ~ ecog_fstcnt + diagnosis + age + gender + albumin + c_reactive_protein
X_baseline <- model.matrix(baseline_formula, data = sim_baseline_data)[, -1] # Baseline values dummy coded, remove intercept

tx <- sample(0:1, N, replace = TRUE) # Sample treatment assignment
X <- cbind(X_baseline, tx) 
```

## PO tx effect, constant over time

- Proportional Part: Effects that are constant across all k-1 logits (e.g., yprev, gap, baseline covariates, pol(week,3)).
- Non-Proportional Part: The week x f(y) term, which varies with the outcome level k. For our model cppo = function(y) y, this term is k * week * \gamma_{week}.

```{r}
ints <- unlist(coef[1:4])
extra <- unlist(coef[- (1:5)]) #remove tx coef as well

names(extra)[names(extra) == "ecog_fstcnt=2"] <- "ecog_fstcnt2"
names(extra)[names(extra) == "ecog_fstcnt=3"] <- "ecog_fstcnt3"
names(extra)[names(extra) == "diagnosis=2"] <- "diagnosis2"
names(extra)[names(extra) == "diagnosis=3"] <- "diagnosis3"
names(extra)[names(extra) == "diagnosis=4"] <- "diagnosis4"
names(extra)[names(extra) == "diagnosis=5"] <- "diagnosis5"
names(extra)[names(extra) == "plan_fstcnt_coded=2"] <- "plan_fstcnt_coded2"
names(extra)[names(extra) == "plan_fstcnt_coded=3"] <- "plan_fstcnt_coded3"
names(extra)[names(extra) == "plan_fstcnt_coded=4"] <- "plan_fstcnt_coded4"
names(extra)[names(extra) == "plan_fstcnt_coded=5"] <- "plan_fstcnt_coded5"
names(extra)[names(extra) == "gender=1"] <- "gender1" # 1 = male

g <- function(yprev, t, gap, X, parameter = 0, extra) {

  # X is a vector for one patient, containing baseline covariates and tx (which we want to remove)
  # yprev is the previous state (character)
  # t is the current week (numeric)
  # gap is assumed to be 1 for this simulation
  gap <- 1 

  # 1. Treatment Effect
  # Treatment effect stable over time
  # The treatment indicator 'tx' is the last element of the X vector.
  tx_value <- X[length(X)]
  tx_effect <- parameter * tx_value

  # 2. Baseline effect
  # X contains the non-treatment variables. We match its names to the 'extra' vector.
  X_nontx <- X[-length(X)] # X is a single vector. We remove tx which is as the end of this vector
  baseline_coef_names <- names(X_nontx)

  baseline_effect <- sum(extra[baseline_coef_names] * X_nontx)

  # 3.  time effect (proportional part) from pol(week, 2)
  time_effect <- extra[["week"]] * t + extra[["week^2"]] * (t^2) + extra[["week^3"]] * (t^3)

  # This is if yprev is equal to 1 (the reference group)
  yprev_effect <- 0
  yprev_week_effect <- 0
  yprev_week_effect_2 <- 0
  yprev_week_effect_3 <- 0

  # 4. Previous state effect
  # standalone effects
  yprev_coef_name <- paste0('yprev=', yprev)
  if (yprev_coef_name %in% names(extra)) {
      yprev_effect <- extra[[yprev_coef_name]]
  }
  # interaction with time (week), linear
  yprev_week_coef_name <- paste0('week * yprev=', yprev)
  if (yprev_week_coef_name %in% names(extra)) {
      yprev_week_effect <- extra[[yprev_week_coef_name]] * t
  }

  # interaction with time (week), 2nd order polynomial
  yprev_week_coef_name_2 <- paste0('week^2 * yprev=', yprev)
  if (yprev_week_coef_name_2 %in% names(extra)) {
      yprev_week_effect_2 <- extra[[yprev_week_coef_name_2]] * (t^2)
  }

  # interaction with time (week), 3rd order polynomial
  yprev_week_coef_name_3 <- paste0('week^3 * yprev=', yprev)
  if (yprev_week_coef_name_3 %in% names(extra)) {
      yprev_week_effect_3 <- extra[[yprev_week_coef_name_3]] * (t^3)
  }

  # Sum of all proportional effects
  lin_pred_prop <- tx_effect + baseline_effect + time_effect + yprev_effect + yprev_week_effect + yprev_week_effect_2 + yprev_week_effect_3

  # 5. Non-proportional effect of time
  # From cppo=function(y) y, the effect for P(Y>=k) is k * week * gamma
  # The columns of the final lp correspond to k = 2, 3, ..., 5
  nonprop_gamma <- extra[['week x f(y)']]
  k_levels <- 2:5 
  lin_pred_nonprop <- k_levels * t * nonprop_gamma

  # 6. Final linear predictor (vector of length 4)
  # This is the sum of the proportional part (a scalar) and non-proportional part (a vector)
  lp <- lin_pred_prop + lin_pred_nonprop
    
  return(lp)

}

formals(g)$extra <- extra
```

## G-Function for PO tx effect, 0 at t = 0, linearly increasing over time

- The functions is specifically for a simulation over 26 weeks of follow-up (hardcoded)
- the parameter value passed to the function logodds is the parameter at week 26

```{r}
g2 <- function(yprev, t, gap, X, parameter = 0, extra) {

  # X is a vector for one patient, containing baseline covariates and tx (which we want to remove)
  # yprev is the previous state (character)
  # t is the current week (numeric)
  # gap is assumed to be 1 for this simulation
  gap <- 1 

  # 1. Treatment Effect
  # The treatment linearly increases on the logit scale from time 0 to time 26
  # at week 26 logodds of treatment = the passed parameter value
  tx_value <- X[length(X)]
  tx_effect <- parameter / 26 * t * tx_value 

  # 2. Baseline effect
  # X contains the non-treatment variables. We match its names to the 'extra' vector.
  X_nontx <- X[-length(X)] # X is a single vector. We remove tx which is as the end of this vector
  baseline_coef_names <- names(X_nontx)

  baseline_effect <- sum(extra[baseline_coef_names] * X_nontx)

  # 3.  time effect (proportional part) from pol(week, 2)
  time_effect <- extra[["week"]] * t + extra[["week^2"]] * (t^2) + extra[["week^3"]] * (t^3)

  # This is if yprev is equal to 1 (the reference group)
  yprev_effect <- 0
  yprev_week_effect <- 0
  yprev_week_effect_2 <- 0
  yprev_week_effect_3 <- 0

  # 4. Previous state effect
  # standalone effects
  yprev_coef_name <- paste0('yprev=', yprev)
  if (yprev_coef_name %in% names(extra)) {
      yprev_effect <- extra[[yprev_coef_name]]
  }
  # interaction with time (week), linear
  yprev_week_coef_name <- paste0('week * yprev=', yprev)
  if (yprev_week_coef_name %in% names(extra)) {
      yprev_week_effect <- extra[[yprev_week_coef_name]] * t
  }

  # interaction with time (week), 2nd order polynomial
  yprev_week_coef_name_2 <- paste0('week^2 * yprev=', yprev)
  if (yprev_week_coef_name_2 %in% names(extra)) {
      yprev_week_effect_2 <- extra[[yprev_week_coef_name_2]] * (t^2)
  }

  # interaction with time (week), 3rd order polynomial
  yprev_week_coef_name_3 <- paste0('week^3 * yprev=', yprev)
  if (yprev_week_coef_name_3 %in% names(extra)) {
      yprev_week_effect_3 <- extra[[yprev_week_coef_name_3]] * (t^3)
  }

  # Sum of all proportional effects
  lin_pred_prop <- tx_effect + baseline_effect + time_effect + yprev_effect + yprev_week_effect + yprev_week_effect_2 + yprev_week_effect_3

  # 5. Non-proportional effect of time
  # From cppo=function(y) y, the effect for P(Y>=k) is k * week * gamma
  # The columns of the final lp correspond to k = 2, 3, ..., 8
  nonprop_gamma <- extra[['week x f(y)']]
  k_levels <- 2:5 
  lin_pred_nonprop <- k_levels * t * nonprop_gamma

  # 6. Final linear predictor (vector of length 7)
  # This is the sum of the proportional part (a scalar) and non-proportional part (a vector)
  lp <- lin_pred_prop + lin_pred_nonprop
    
  return(lp)

}

formals(g2)$extra <- extra
```

## G-Function for constant nonPO Tx Effect

- Assume that effect on mortality is smaller than effect on planned / unplanned hospitalizations

```{r}
g3 <- function(yprev, t, gap, X, parameter = 0, extra) {
  gap <- 1 

  # 1. Treatment Effect
  tx_value <- X[length(X)]
  tx_effect <- parameter * tx_value

  # 2. Baseline effect
  X_nontx <- X[-length(X)] # X is a single vector. We remove tx which is as the end of this vector
  baseline_coef_names <- names(X_nontx)

  baseline_effect <- sum(extra[baseline_coef_names] * X_nontx)

  # 3.  time effect (proportional part) from pol(week, 2)
  time_effect <- extra[["week"]] * t + extra[["week^2"]] * (t^2) + extra[["week^3"]] * (t^3)

  # This is if yprev is equal to 1 (the reference group)
  yprev_effect <- 0
  yprev_week_effect <- 0
  yprev_week_effect_2 <- 0
  yprev_week_effect_3 <- 0

  # 4. Previous state effect
  # standalone effects
  yprev_coef_name <- paste0('yprev=', yprev)
  if (yprev_coef_name %in% names(extra)) {
      yprev_effect <- extra[[yprev_coef_name]]
  }
  # interaction with time (week), linear
  yprev_week_coef_name <- paste0('week * yprev=', yprev)
  if (yprev_week_coef_name %in% names(extra)) {
      yprev_week_effect <- extra[[yprev_week_coef_name]] * t
  }

  # interaction with time (week), 2nd order polynomial
  yprev_week_coef_name_2 <- paste0('week^2 * yprev=', yprev)
  if (yprev_week_coef_name_2 %in% names(extra)) {
      yprev_week_effect_2 <- extra[[yprev_week_coef_name_2]] * (t^2)
  }

  # interaction with time (week), 3rd order polynomial
  yprev_week_coef_name_3 <- paste0('week^3 * yprev=', yprev)
  if (yprev_week_coef_name_3 %in% names(extra)) {
      yprev_week_effect_3 <- extra[[yprev_week_coef_name_3]] * (t^3)
  }

  # Sum of all proportional effects
  lin_pred_prop <- baseline_effect + time_effect + yprev_effect + yprev_week_effect + yprev_week_effect_2 + yprev_week_effect_3

  # 5. Non-proportional effect of time
  nonprop_gamma <- extra[['week x f(y)']]
  k_levels <- 2:5 
  lin_pred_nonprop <- k_levels * t * nonprop_gamma

  # 6. Final linear predictor (vector of length 7)
  # This is the sum of the proportional part (a scalar) and non-proportional part (a vector)
  lp <- lin_pred_prop + lin_pred_nonprop + c(rep(tx_effect, 3), 0) # tx not applied to last threshold
    
  return(lp)

}

formals(g3)$extra <- extra
```

## Simulate dataset with PO for treatment

- Odds of 0.8 of moving to a higher (worse state)

```{r}
# Run the simulation
carried_forward <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:5, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 5,
                  intercepts = ints, 
                  g = g, 
                  parameter = log(0.8),
                  carry = TRUE)

not_carried_forward <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:5, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 5,
                  intercepts = ints, 
                  g = g, 
                  parameter = log(0.8),
                  carry = FALSE)
```

### Check if OR of 0.8 is recovered

```{r}
sample <- not_carried_forward |> 
    select(id, time, y, yprev, tx) |> 
    left_join(sim_baseline_data |> select(-yprev), by = c("id" = "pat_id"))

h <- vglm(ordered(y) ~ yprev * poly(time, 3) + ecog_fstcnt + diagnosis + tx,
                cumulative(reverse=TRUE, parallel=FALSE ~ poly(time, 3)),
                data=sample)

h
```

### Calculate the SOPs
```{r}
sops <- carried_forward |> 
  group_by(time, tx) |> 
  count(y) |> 
  reframe(
    y = y,
    sop = n / sum(n))
```

### Calculate the difference in weeks spent at home
```{r}
# Difference in mean days with excellent, very good or good quality of life
weeks_at_home <- sops |> 
  filter(y == 1) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_at_home$mean_weeks[2] - weeks_at_home$mean_weeks[1]

# difference in weeks at home or planned hospitalization
weeks_wo_emerg <- sops |> 
  filter(y == 1 | y == 2) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_wo_emerg$mean_weeks[2] - weeks_wo_emerg$mean_weeks[1]

# difference in RMST
rmst <- sops |> 
  filter(y != 5) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

rmst$mean_weeks[2] - rmst$mean_weeks[1]
```

In this population an OR of 0.8 for a proportional treatment effect leads to a
-  1.24 week increase in time spent at home 
-  1.08 weeks more without an emergency / unplanned hospitalization
-  0.7 week increase in the restricted mean survival time at 6 months.


## Simulate dataset with increasing tx effect

- Odds of 0.7 of moving to a worse state at week 26, linear increase until the on the logit scale
- I choose OR of 0.7, because 0.8 results in a very small difference. I think MCID should be at least 1 week more at home.

```{r}
# Run the simulation
carried_forward_non_const <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:5, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 5,
                  intercepts = ints, 
                  g = g2, 
                  parameter = log(0.6),
                  carry = TRUE)

not_carried_forward_non_const  <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:5, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 5,
                  intercepts = ints, 
                  g = g2, 
                  parameter = log(0.6),
                  carry = FALSE)
```

### Examine OR

What happens if we don't fit interaction between tx and time ?

```{r}
sample_non_const <- not_carried_forward_non_const |> 
    select(id, time, y, yprev, tx) |> 
    left_join(sim_baseline_data |> select(-yprev), by = c("id" = "pat_id"))

h_non_const <- vglm(ordered(y) ~ tx + yprev * poly(time, 3) + tx:time + ecog_fstcnt + diagnosis,
                cumulative(reverse=TRUE, parallel=FALSE ~ poly(time, 3)),
                data=sample_non_const)

h_non_const
```

The OR is estimated as 1 but with the interaction term included is calculated as close to 0.7 for week 26. 

### Calculate the SOPs
```{r}
sops_non_const <- carried_forward_non_const |> 
  group_by(time, tx) |> 
  count(y) |> 
  reframe(
    y = y,
    sop = n / sum(n))
```

### Calculate the difference in weeks spent at home

This results in differences of only days.

```{r}
# Difference in mean days with excellent, very good or good quality of life
weeks_at_home <- sops_non_const |> 
  filter(y == 1) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_at_home$mean_weeks[2] - weeks_at_home$mean_weeks[1]

# difference in weeks at home or planned hospitalization
weeks_wo_emerg <- sops_non_const |> 
  filter(y == 1 | y == 2) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_wo_emerg$mean_weeks[2] - weeks_wo_emerg$mean_weeks[1]

# difference in RMST
rmst <- sops_non_const |> 
  filter(y != 5) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

rmst$mean_weeks[2] - rmst$mean_weeks[1]
```

- 1.04 more weeks at home
- 0.88 more weeks without emergency / unplanned
- 0.54 longer RMST

## Simulate dataset with nonPO for treatment

```{r}
carried_forward_nonPO <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:5, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 5,
                  intercepts = ints, 
                  g = g3, 
                  parameter = log(0.8),
                  carry = TRUE)

not_carried_forward_nonPO <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:5, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 5,
                  intercepts = ints, 
                  g = g3, 
                  parameter = log(0.8),
                  carry = FALSE)
```

### Examine OR

```{r}
sample_nonPO <- not_carried_forward_nonPO |> 
    select(id, time, y, yprev, tx) |> 
    left_join(sim_baseline_data |> select(-yprev), by = c("id" = "pat_id"))

h_nonPO <- vglm(ordered(y) ~ tx + yprev * poly(time, 3) + ecog_fstcnt + diagnosis,
                cumulative(reverse=TRUE, parallel=FALSE ~ tx + poly(time, 3)),
                data=sample_nonPO)

h_nonPO
```

As expected, the treatment effect is close to 0 (OR = 1) for threshold 4, and close to log(0.8) for thresholds 1 to 3.

### Calculate the SOPs
```{r}
sops_nonPO <- carried_forward_nonPO |> 
  group_by(time, tx) |> 
  count(y) |> 
  reframe(
    y = y,
    sop = n / sum(n))
```

### Calculate the difference in weeks spent at home

```{r}
# Difference in mean days with excellent, very good or good quality of life
weeks_at_home <- sops_nonPO |> 
  filter(y == 1) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_at_home$mean_weeks[2] - weeks_at_home$mean_weeks[1]

# difference in weeks at home or planned hospitalization
weeks_wo_emerg <- sops_nonPO |> 
  filter(y == 1 | y == 2) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_wo_emerg$mean_weeks[2] - weeks_wo_emerg$mean_weeks[1]

# difference in RMST
rmst <- sops_nonPO |> 
  filter(y != 5) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

rmst$mean_weeks[2] - rmst$mean_weeks[1]
```

- 0.97 weeks more at home (was 1.2 with PO)
- 0.79 weeks more w/o emergency / unplanned
- 0.34 weeks longer RMST (was 0.7 with PO)

## Create and export images

### State Occupancy Probabilities

#### Constant OR, PO
```{r}
sops_line <- carried_forward |> 
  group_by(time, tx) |> 
  count(y) |> 
  mutate(
    total = sum(n),
    sop   = n / total,
    tx = factor(tx)) |>
  ungroup() |> 
  ggplot(aes(x = time, y = sop, color = y)) +
  geom_line(aes(linetype = tx)) +
  scale_color_viridis_d(end = 0.9) +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Study Week",
   y = "SOP",
   subtitle = "Empirical state occupancy probabilities",
   linetype = "Treatment",
   color = "State")

sops_stacked <- carried_forward |>  
  ggplot(aes(x = time, fill = y)) +
  geom_bar(position = "fill")  +
  scale_fill_viridis_d(end = 0.9) +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.2)) +
  facet_wrap(~tx) +
  labs(x = "Study Week",
   y = "Cumulative Proportion",
   subtitle = "Distribution of ordinal states over time",
   color = "State")

sops_line / sops_stacked +
  plot_annotation(tag_level = 'A')

ggsave(filename = "./img/fig-hosp-sim-data-sop-PO-constant.svg", width = 12, height = 10)

```

#### Non-Constant OR, PO
```{r}
sops_line <- carried_forward_non_const |> 
  group_by(time, tx) |> 
  count(y) |> 
  mutate(
    total = sum(n),
    sop   = n / total,
    tx = factor(tx)) |>
  ungroup() |> 
  ggplot(aes(x = time, y = sop, color = y)) +
  geom_line(aes(linetype = tx)) +
  scale_color_viridis_d(end = 0.9) +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Study Week",
   y = "SOP",
   subtitle = "Empirical state occupancy probabilities",
   linetype = "Treatment",
   color = "State")

sops_stacked <- carried_forward_non_const |>  
  ggplot(aes(x = time, fill = y)) +
  geom_bar(position = "fill")  +
  scale_fill_viridis_d(end = 0.9) +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.2)) +
  facet_wrap(~tx) +
  # coord_cartesian(ylim = c(0, 0.4)) +
  labs(x = "Study Week",
   y = "Cumulative Proportion",
   subtitle = "Distribution of ordinal states over time",
   color = "State")

sops_line / sops_stacked +
  plot_annotation(tag_level = 'A')

ggsave(filename = "./img/fig-hosp-sim-data-sop-non-const.svg", width = 12, height = 10)
```

#### Constant OR, non-PO

```{r}
sops_line <- carried_forward_nonPO |> 
  group_by(time, tx) |> 
  count(y) |> 
  mutate(
    total = sum(n),
    sop   = n / total,
    tx = factor(tx)) |>
  ungroup() |> 
  ggplot(aes(x = time, y = sop, color = y)) +
  geom_line(aes(linetype = tx)) +
  scale_color_viridis_d(end = 0.9) +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Study Week",
   y = "SOP",
   subtitle = "Empirical state occupancy probabilities",
   linetype = "Treatment",
   color = "State")

sops_stacked <- carried_forward_nonPO |>  
  ggplot(aes(x = time, fill = y)) +
  geom_bar(position = "fill")  +
  scale_fill_viridis_d(end = 0.9) +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.2)) +
  facet_wrap(~tx) +
  # coord_cartesian(ylim = c(0, 0.4)) +
  labs(x = "Study Week",
   y = "Cumulative Proportion",
   subtitle = "Distribution of ordinal states over time",
   color = "State")

sops_line / sops_stacked +
  plot_annotation(tag_level = 'A')

ggsave(filename = "./img/fig-hosp-sim-data-sop-const-nonPO.svg", width = 12, height = 10)
```

### Individual trajectories

```{r}
set.seed(4)
ids <- sample$id |> unique() |> sample(20) 

fig_tile_random <- not_carried_forward |> 
  filter(id %in% ids) |> 
  mutate(id = as.factor(as.numeric(as.factor((id))))) |> 
  ggplot(aes(y = id, x = time)) +
  geom_tile(mapping = aes(fill = y), width = 0.95, height = 0.4) +
  scale_fill_viridis_d(end = 0.9) +
  scale_x_continuous(breaks = 1:26)

ggsave(filename = "./img/fig-hosp-sim-tile.svg", width = 12, height = 6)
```

## Export simulataed datasets

```{r}
write_parquet(sample, "./2-hosp/sim-data/sim_data_po_const.parquet")
write_parquet(sample_non_const, "./2-hosp/sim-data/sim_data_po_non_const.parquet")
write_parquet(sample_nonPO, "./2-hosp/sim-data/sim_data_nonpo_const.parquet")
```
