---
Title: "Calculating Marginal State Occupancy Probabilities from a Markov Model"
---

And also simulating patient trajectories. Special focus on second-order Markov models.

## Setup

```{r}
library(tidyverse)
library(VGAM)
library(data.table)
library(arrow)
library(rms)
library(zoo)
```

## Fit model to use

```{r}
df_ypprev <- read_parquet("./data/hosp_long_weeks_two_lags.parquet") |> 
    mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE),
        y = ordered(y)
    ) |> 
      select(pat_id, tx, y, yprev, ypprev, week, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)

d <- as.data.table(na.omit(df_ypprev))
```

```{r}
week_spl_m <- rcs(d$week, 4)
knots <- attr(week_spl_m, "parms")

d$week_linear <- week_spl_m[, 1]
d$week_nonlinear_1 <- week_spl_m[, 2]
d$week_nonlinear_2 <- week_spl_m[, 3]

M <- 4 # n thresholds

cons_lin_int <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),
  "yprev:week_linear"  = cbind(PO_effect = 1),
  "ecog_fstcnt"       = cbind(PO_effect = 1),
  "albumin"           = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_non_po_lin_int <- vglm(
  y ~ tx + yprev * week_linear + week_nonlinear_1 + week_nonlinear_2 + ecog_fstcnt + albumin,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_lin_int
)

cons_second_lin_int <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "tx"                = cbind(PO_effect = 1),
  "yprev"             = cbind(PO_effect = 1),
  "ypprev"            = cbind(PO_effect = 1),
  "week_nonlinear_1"   = cbind(PO_effect = 1),
  "week_nonlinear_2"   = cbind(PO_effect = 1),
  "yprev:week_linear"  = cbind(PO_effect = 1),
  "ypprev:week_linear" = cbind(PO_effect = 1),
  "ecog_fstcnt"       = cbind(PO_effect = 1),
  "albumin"           = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "week_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_second_non_po_lin_int <- vglm(
  y ~ tx + (yprev + ypprev) * week_linear + week_nonlinear_1 + week_nonlinear_2 + ecog_fstcnt + albumin,
  family = cumulative(parallel = FALSE ~ week_linear, reverse = TRUE),
  data = d,
  constraints = cons_second_lin_int
)
```

## Calculating State Occupancy Probabilities (SOPs)

### First order model

Generalized loop using matrix multiplications

```{r}
row <- d[1, ] |> 
  select(-ypprev)

ylevels <- factor(1:5, levels = 1:5)
absorb <- 5
times <- 3:26
steps <- length(ylevels) - 1 # N non-abosrbing

# State vector for beginning of loop (week 2) (S_2)
state_vector <- matrix(predict(fit_non_po_lin_int, newdata = row, type = "response"))

# Create a list to store all the results, starting with the initial state
results_list <- list(state_vector)

for(i in times) {

  # create newdata for this day
  # Now we wan to do this for all possible yprev states and days from 3 to 26
  newdata_i <- expand.grid(
    week = i,
    yprev = factor(1:steps, levels = 1:steps)
  ) |> 
    mutate(tx = row$tx, 
           ecog_fstcnt = row$ecog_fstcnt,
           albumin = row$albumin)

  spline_matrix_i <- rcs(newdata_i$week, parms = knots)  
  colnames(spline_matrix_i) <- c("week_linear", "week_nonlinear_1", "week_nonlinear_2")
  newdata_i <- cbind(
    newdata_i,
    week_linear = spline_matrix_i[, 1],
    week_nonlinear_1 = spline_matrix_i[, 2],
    week_nonlinear_2 = spline_matrix_i[, 3]
  )
  predictions_i <- predict(fit_non_po_lin_int, newdata = newdata_i, 
                               type = "response")

  absorbing_row <- rep(0, length(ylevels))
  absorbing_row[absorb] <- 1
  transition_matrix_i <- t(rbind(predictions_i, absorbing_row))

  state_vector <- transition_matrix_i %*% state_vector

  results_list <- append(results_list, list(state_vector))
}

sop_matrix_wide <- do.call(cbind, results_list)
sop_matrix_final <- t(sop_matrix_wide)

colnames(sop_matrix_final) <- paste("State", 1:5)
rownames(sop_matrix_final) <- c(2, times) 
```

### Second order model

#### Some theory

We need to track the **joint probability** of two-step histories, rather than relying on marginal probabilities which would incorrectly assume independence.

To find the probability of being in any state `i` at week `t`, we must sum the probabilities of all possible two-week histories that could lead to that state:

$$
P(Y_t=i) = \sum_{j} \sum_{k} P(Y_t=i | Y_{t-1}=j, Y_{t-2}=k) \times P(Y_{t-1}=j, Y_{t-2}=k)
$$

This equation has two key components:

1.  **Transition Probability**: `$P(Y_t=i | Y_{t-1}=j, Y_{t-2}=k)$`. This is the conditional probability of moving to state `i` given the immediate history. This is what our `vglm` model provides via the `predict()` function.

2.  **Joint Probability**: `$P(Y_{t-1}=j, Y_{t-2}=k)$`. This is the probability that the specific two-week history occurred. This is the crucial piece of information that we must calculate at each step and carry forward to the next.

------------------------------------------------------------------------

##### The Step-by-Step Implementation

The process begins with a known history and then enters a loop, where the full probability distribution from one week is used to generate the distribution for the next.

###### 1. Initialization (e.g., Calculating Week 2)

We begin with a new patient whose history is known. For example, we might know they were in **State 1 at Week 0** and **State 1 at Week 1**.

-   **First Prediction**: To predict the state probabilities for Week 2, the inputs are certain (`ypprev=1`, `yprev=1`). A call to `predict()` gives us the marginal probability vector for Week 2, which we can call `pi_week_2`.

-   **First Joint Probability Matrix**: We then construct the first joint probability matrix, `joint_prob_prev`. This matrix represents `$P(Y_2, Y_1)$` and is essential for kicking off the iterative loop to calculate Week 3.

    -   The matrix dimensions are: **Rows = State at Week 1 (**$Y_1$) and **Columns = State at Week 2 (**$Y_2$).
    -   Since we know with 100% certainty that the patient was in State 1 at Week 1, only the first **row** of this matrix can contain non-zero values. We populate this row with the `pi_week_2` vector. All other rows are zero.

###### 2. The Iterative Loop (Weeks 3 and Beyond)

The loop proceeds for each subsequent week, performing the same steps.

###### Step A: Construct a Complete "Transition Rulebook"

For each week, we build a matrix of transition probabilities (`transition_probs_w`) that contains the outcome for every two-week history.

1.  **Create a Grid of Histories**: A grid (`newdata_w`) of all possible `(ypprev, yprev)` combinations is generated. If there are 5 states, this results in a 5x5=25 row grid representing every possible history.

2.  **Fill the Rulebook**: The `transition_probs_w` matrix (25 rows, 5 columns) is populated using the following hierarchy of rules:

    -   **Rule 1 (Absorbing State)**: If a history ends with the patient in the absorbing state (`yprev = 5`), the outcome is certain: they remain in state 5. The corresponding row in the rulebook is set to `[0, 0, 0, 0, 1]`.
    -   **Rule 2 (Model Prediction)**: Next, we identify histories the model can handle. These are histories consisting **only of transient states** (i.e., `yprev != 5` AND `ypprev != 5`). The `predict()` function is called on this subset of the data, and the results are placed in the corresponding rows of the rulebook.
    -   **Rule 3 (Impossible Histories)**: Histories that are logically impossible (e.g., `ypprev=5, yprev=3`) are not covered by the rules above. Their transition probabilities are left as `[0, 0, 0, 0, 0]`. As the simulation progresses, the calculated joint probability of these histories occurring will be zero, so they will not contribute to the final result.

###### Step B: Calculate the New Probabilities

The **history probabilities** (`joint_prob_prev`) and the **transition rulebook** (`transition_probs_w`) are combined.

1.  **Calculate the New Joint Matrix**: We compute the next joint probability matrix, `joint_prob_new` (representing `$P(Y_t, Y_{t-1})$`). Each cell `(j, i)` in this new matrix is calculated via a dot product:

    `joint_prob_new[j, i] = sum(vector_of_transition_rules * vector_of_history_probabilities)`

    -   The `vector_of_history_probabilities` is column `j` from `joint_prob_prev`. It gives the probability of all histories that ended in state `j` at time `t-1`.
    -   The `vector_of_transition_rules` is extracted from our rulebook. It contains the probabilities of transitioning *to* state `i` *from* state `j`, given all possible states at `t-2`.

2.  **Calculate the Marginal Probabilities**: Once the `joint_prob_new` matrix is complete, the final state occupancy probabilities for week `t` (`pi_w`) are simply the **column sums** of this matrix. The total probability of being in state `i` at week `t` is the sum of all paths that arrived at state `i`.

###### Step C: Update for the Next Iteration

The process concludes by preparing for the next week: the newly calculated `joint_prob_new` becomes the `joint_prob_prev` for the next cycle, carrying the complete, updated historical information forward.

Exmple script:

```{r}
row <- d[1, ]

# Define constants
N_STATES <- 5
ABS_STATE <- 5 # The absorbing state
WEEKS_TO_PREDICT <- 3:26 # we predict for week 2 outside the loop
ALL_STATES <- factor(1:N_STATES)


# Predict probabilities for week 2
pred_week_2 <- predict(
    fit_second_non_po_lin_int,
    newdata = row, # Predict only for non-absorbing
    type = "response"
)

# State probability vector for week 2
pi_week_2 <- pred_week_2[1, ]

# Store results
occupancy_probs <- list()
occupancy_probs[["2"]] <- pi_week_2

# Initialize the first JOINT probability matrix: P(Y_2, Y_1)
# Y_1 is known to be `initial_state_week_1`.
# So, the joint probability is 0 everywhere except for the column of Y_1.
# ypprev refers to week 1

# Initialize an empty 5x5 matrix for P(Y₂, Y₁)
# Rows are Y₁ (state at week 1)
# Columns are Y₂ (state at week 2)
joint_prob_prev <- matrix(0, nrow = N_STATES, ncol = N_STATES,
                          dimnames = list(paste0("Y₁_state_", ALL_STATES), 
                                          paste0("Y₂_state_", ALL_STATES)))

joint_prob_prev[row$yprev, ] <- pi_week_2

for (w in WEEKS_TO_PREDICT) {
  
  # Create a grid for all possible previous states (ypprev, yprev)
  newdata_w <- expand.grid(
    ypprev = ALL_STATES,
    yprev = ALL_STATES,
    week = w,
    tx = row$tx
  )
  
  # Add spline terms
  spline_matrix_w <- rcs(newdata_w$week, parms = knots)
  colnames(spline_matrix_w) <- c("week_linear", "week_nonlinear_1", "week_nonlinear_2")
  newdata_w <- cbind(
    newdata_w,
    week_linear = spline_matrix_w[, 1],
    week_nonlinear_1 = spline_matrix_w[, 2],
    week_nonlinear_2 = spline_matrix_w[, 3]
  )

  # Initialize a full transition probability matrix for all 25 histories
  transition_probs_w <- matrix(0, nrow = nrow(newdata_w), ncol = N_STATES)

  # Apply the absorbing state rule: if yprev was 5, the next state must be 5.
  absorbing_rows <- newdata_w$yprev == ABS_STATE
  transition_probs_w[absorbing_rows, ABS_STATE] <- 1

  # Apply the model prediction rule: if the entire history is transient.
  # Identify rows where BOTH yprev and ypprev are not the absorbing state.
  model_predict_rows <- (newdata_w$yprev != ABS_STATE) & (newdata_w$ypprev != ABS_STATE)
  
  if (any(model_predict_rows)) {
    pred_transient <- predict(fit_second_non_po_lin_int,
                              newdata = newdata_w[model_predict_rows, ],
                              type = "response")
    
    # Place the predictions into the correct rows of our full transition matrix
    transition_probs_w[model_predict_rows, ] <- pred_transient
  }

  # Calculate the new joint and marginal probabilities ---
  
  joint_prob_new <- matrix(0, nrow = N_STATES, ncol = N_STATES,
                           dimnames = list(paste0("yprev_", ALL_STATES), paste0("ynext_", ALL_STATES)))

  for (j in 1:N_STATES) { 
    # j = yprev 
    # i = ynext
    for (i in 1:N_STATES) { 
      trans_probs_fixed_j <- transition_probs_w[newdata_w$yprev == j, i]
      prev_joint_probs_fixed_j <- joint_prob_prev[, j]
      joint_prob_new[j, i] <- sum(trans_probs_fixed_j * prev_joint_probs_fixed_j)
    }
  }
  
  pi_w <- colSums(joint_prob_new)
  
  if (abs(sum(pi_w) - 1) > 1e-6) {
    warning(paste("Probabilities do not sum to 1 at week", w))
  }
  occupancy_probs[[as.character(w)]] <- pi_w
  joint_prob_prev <- joint_prob_new
}

final_probs <- do.call(rbind, occupancy_probs)
```

#### Function for Second Order VGAM Model

Gemini opimized the function using arrays instead of nested loops. The results are the same as the script above, but faster. However, the math is a bit above my head!

```{r}
# Make sure you have the required packages installed and loaded
# install.packages(c("VGAM", "Hmisc"))
library(VGAM)
library(Hmisc)

predict_markov_sops <- function(model,
                                baseline_data,
                                n_states,
                                times,
                                time_var_name,
                                yprev_name,
                                ypprev_name,
                                absorbing_state = NULL,
                                knots = NULL) {

  # --- 1. Input Validation and Setup ---
  if (!inherits(model, "vglm")) {
    stop("This function currently only supports 'vglm' model objects from the VGAM package.")
  }
  if (nrow(baseline_data) != 1) {
    stop("'baseline_data' must be a single row (e.g., a data.frame or tibble).")
  }
  
  # Check that the specified history columns exist in the baseline data
  if (!all(c(yprev_name, ypprev_name) %in% names(baseline_data))) {
    stop("'baseline_data' must contain the specified 'yprev_name' and 'ypprev_name' columns.")
  }
  
  # Ensures the prediction times are consecutive (e.g., 2, 3, 4, 5)
  if (length(times) > 1 && !all(diff(sort(times)) == 1)) {
    stop("'times' must be a sequence of consecutive integers with no gaps (e.g., 2:26).")
  }

  if(is.null(baseline_data[[time_var_name]])){
    stop(paste0("The 'baseline_data' must contain the time variable '", time_var_name, "'."))
  }

  # Checks if the time in the baseline data is t-1 relative to the first prediction.
  first_pred_time <- min(times)
  expected_baseline_time <- first_pred_time
  if (baseline_data[[time_var_name]] != expected_baseline_time) {
      stop(paste0("Warning: The value of '", time_var_name, "' in 'baseline_data' is ",
                     baseline_data[[time_var_name]], ", but it should be ",
                     expected_baseline_time, " (first prediction time)."))
  }

  # Extract the response variable name from the model formula
  y_name <- all.vars(formula(model))[1]
  
  # Define state space
  all_states <- factor(1:n_states)

  # --- 2. Initialization Step (First Prediction) ---

  # The first prediction is for the minimum time specified
  first_pred_time <- min(times)
  baseline_data <- data.frame(baseline_data)
  initial_pred_data <- baseline_data
  initial_pred_data[[time_var_name]] <- first_pred_time
  
  # Handle splines if 'knots' are provided
  # THIS STILL NEEDS TO BE OPTIMIZED, because for now week_linear and week_nonlinear are hardcoded
  spline_var_names <- grep("week_linear|week_nonlinear", all.vars(formula(model)), value = TRUE)
  if (!is.null(knots)) {
      spline_values <- rcs(initial_pred_data[[time_var_name]], parms = knots)
      spline_preds <- as.data.frame(matrix(spline_values, ncol = length(spline_var_names)))
      names(spline_preds) <- spline_var_names
      initial_pred_data <- cbind(initial_pred_data, spline_preds)
  }
  
  # Predict the state probabilities for the first time point
  pred_first_step <- predict(model, newdata = initial_pred_data, type = "response")
  pi_first_step <- pred_first_step[1, ]
  
  # Initialize results list
  occupancy_probs <- list()
  occupancy_probs[[as.character(first_pred_time)]] <- pi_first_step

  # Initialize the first JOINT probability matrix: P(Y_t, Y_{t-1})
  # Initialize P(Y_1, Y_2) with Y_1 on ROWS and Y_2 on COLUMNS
  joint_prob_prev <- matrix(0, nrow = n_states, ncol = n_states)
  yprev_state <- baseline_data[[yprev_name]]
  joint_prob_prev[yprev_state, ] <- pi_first_step
  
  # --- 3. Iterative Loop ---

  # Loop over the remaining times
  iter_times <- times[times > first_pred_time]

  for (t in iter_times) {
    
    # A. Create a grid of all possible two-step histories
    # Extract baseline covariates that don't change over time
    static_covars <- setdiff(names(baseline_data), c(y_name, yprev_name, ypprev_name, time_var_name, spline_var_names))
    
    # Create a named list to ensure expand.grid uses the correct column names
    grid_list <- list(all_states, all_states)
    names(grid_list) <- c(ypprev_name, yprev_name)
    newdata_t <- expand.grid(grid_list)
    newdata_t[[time_var_name]] <- t
    
    # Add static covariates from baseline data
    if (length(static_covars) > 0) {
        newdata_t <- cbind(newdata_t, baseline_data[, static_covars, drop = FALSE])
    }

    # Recalculate spline terms for the current time point 't'
    if (!is.null(knots)) {
      spline_matrix_t <- rcs(newdata_t[[time_var_name]], parms = knots)
      spline_matrix_t <- as.data.frame(matrix(spline_matrix_t, ncol = length(spline_var_names)))
      names(spline_matrix_t) <- spline_var_names
      newdata_t <- cbind(newdata_t, spline_matrix_t)
    }

    # B. Build the "Transition Rulebook" for time 't'
    transition_probs_t <- matrix(0, nrow = nrow(newdata_t), ncol = n_states)

    # Rule 1: Handle histories ending in the absorbing state.
    if (!is.null(absorbing_state)) {
      absorbing_rows <- newdata_t[[yprev_name]] == absorbing_state
      transition_probs_t[absorbing_rows, absorbing_state] <- 1
    }

    # Rule 2: Predict ONLY for histories where BOTH previous states are transient.
    model_predict_rows <- (newdata_t[[yprev_name]] != absorbing_state) & (newdata_t[[ypprev_name]] != absorbing_state)


    # Rule 2: Use the model to predict transitions for all other histories
    if (any(model_predict_rows)) {
      pred_transient <- predict(model,
                                newdata = newdata_t[model_predict_rows, ],
                                type = "response")
      transition_probs_t[model_predict_rows, ] <- pred_transient
    }

    initial_array <- array(t(transition_probs_t), dim = c(n_states, n_states, n_states))
    transition_array <- aperm(initial_array, c(2, 3, 1))
    joint_prob_new <- matrix(0, nrow = n_states, ncol = n_states)
    
    for (i in 1:n_states) {
        T_i <- transition_array[, , i]
        joint_prob_new[, i] <- colSums(T_i * joint_prob_prev)
    }
    
    pi_t <- colSums(joint_prob_new)
    
    if (abs(sum(pi_t) - 1) > 1e-6) {
      warning(paste("Probabilities do not sum to 1 at time", t))
    }
    
    occupancy_probs[[as.character(t)]] <- pi_t
    joint_prob_prev <- joint_prob_new
  }

  # --- 4. Format and Return Output ---
  final_probs <- do.call(rbind, occupancy_probs)
  rownames(final_probs) <- names(occupancy_probs)
  colnames(final_probs) <- paste0("State_", 1:n_states)
  
  return(as.data.frame(final_probs))
}
```

#### Apply function to all patients and average

```{r}
N_STATES <- 5
ABS_STATE <- 5
WEEKS_TO_PREDICT <- 2:26
splines <- rcs(d$week, 4)
knots <- attr(splines, "parms")


baseline_patients_t1 <- d |> 
  filter(week == 2) |> 
  select(pat_id, week, yprev, ypprev, ecog_fstcnt, albumin) |> 
  mutate(tx = factor(1, levels = c(0, 1)))

baseline_patients_t0 <- d |>
  filter(week == 2) |> 
  select(pat_id, week, yprev, ypprev, ecog_fstcnt, albumin) |> 
  mutate(tx = factor(0, levels = c(0, 1)))


list_of_patients_1 <- split(baseline_patients_t1, baseline_patients_t1$pat_id)

all_predictions_raw_1 <- map_dfr(list_of_patients_1, ~ predict_markov_sops(
    model = fit_second_non_po_lin_int,
    baseline_data = .x,
    n_states = N_STATES,
    times = WEEKS_TO_PREDICT,
    time_var_name = "week",
    yprev_name = "yprev",
    ypprev_name = "ypprev",
    absorbing_state = ABS_STATE,
    knots = knots
  ), .id = "pat_id")

list_of_patients_0 <- split(baseline_patients_t0, baseline_patients_t0$pat_id)

all_predictions_raw_0 <- map_dfr(list_of_patients_0, ~ predict_markov_sops(
    model = fit_second_non_po_lin_int,
    baseline_data = .x, 
    n_states = N_STATES,
    times = WEEKS_TO_PREDICT,
    time_var_name = "week",
    yprev_name = "yprev",
    ypprev_name = "ypprev",
    absorbing_state = ABS_STATE,
    knots = knots
  ), .id = "pat_id")


all_predictions_tidy_1 <- all_predictions_raw_1 |> 
  # 1. Turn rownames (which are the weeks) into a column
  rownames_to_column(var = "week") |> 
  group_by(pat_id) |> 
  mutate(week = row_number()) |> 
  ungroup()

all_predictions_tidy_0 <- all_predictions_raw_0 |> 
  # 1. Turn rownames (which are the weeks) into a column
  rownames_to_column(var = "week") |> 
  group_by(pat_id) |> 
  mutate(week = row_number()) |> 
  ungroup()

marginalized_sops <- bind_rows(
  all_predictions_tidy_1 |> mutate(tx = "1"),
  all_predictions_tidy_0 |> mutate(tx = "0")
) |>  
    pivot_longer(
      cols = starts_with("State_"),
      names_to = "state",
      names_prefix = "State_",
      values_to = "probability"
    ) |>
  group_by(tx, week, state) |> 
  summarise(sop = mean(probability), .groups = "drop")
```

##### Visualize SOPS

```{r}
marginalized_sops |> 
  ggplot(aes(x = week, y = sop, color = state, linetype = tx)) +
  geom_line() +
  scale_color_viridis_d(end = 0.9, name = "State") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 1)) +
  labs(
    title = "Marginalized State Occupancy Probabilities Over Time",
    x = "Week",
    y = "Probability"
  )
```

## Simulating patient trajectories

First create baseline data and create spline matrix

```{r}
knots <- attr(week_spl_m, "parms")

N <- 1e5
baseline_data  <- d |>
  filter(week == 2) |> 
  select(pat_id, ecog_fstcnt, diagnosis, albumin, yprev, ypprev) |> 
  slice_sample(n = N, replace = TRUE) |> 
  mutate(
    pat_id = row_number(),
    tx     = factor(sample(c(1, 0), N, replace = TRUE)),
    week    = 2
  ) |> 
    data.frame()
```

Simulation function

```{r}
simulate_trajectories <- function(model, baseline_data, knots, start_week = 2, end_week = 26) {

  # List to store the results from each week
  weekly_results <- list()

  # Initialize with the cohort of patients active at the start
  active_patients_data <- baseline_data

  # Loop through each week, processing all active patients at once
  for (w in start_week:end_week) {

    # If no patients are left, stop the simulation
    if (nrow(active_patients_data) == 0) {
      break
    }

    # Ensure factor levels are consistent for prediction
    active_patients_data$yprev <- factor(active_patients_data$yprev, levels = 1:4)
    active_patients_data$ypprev <- factor(active_patients_data$ypprev, levels = 1:4)

    # Recalculate spline basis functions for the entire active cohort
    new_spline_matrix <- rcs(active_patients_data$week, parms = knots)
    colnames(new_spline_matrix) <- c("week_linear", "week_nonlinear_1", "week_nonlinear_2")
    
    active_patients_data$week_linear <- new_spline_matrix[, 1]
    active_patients_data$week_nonlinear_1 <- new_spline_matrix[, 2]
    active_patients_data$week_nonlinear_2 <- new_spline_matrix[, 3]
    
    # Predict probabilities for ALL active patients. This returns a matrix.
    probs_matrix <- predict(model, newdata = active_patients_data, type = "response")
    
    # For each patient (row in the matrix), sample a new outcome 'y'
    y_new_vector <- apply(probs_matrix, 1, function(p) {
      # Normalize probabilities just in case of floating point issues
      p_norm <- pmax(0, p) / sum(pmax(0, p))
      sample(1:5, size = 1, prob = p_norm)
    })

    # Add the new outcomes to the data for the current week
    active_patients_data$y <- y_new_vector

    # Store the results for the current week
    weekly_results[[as.character(w)]] <- active_patients_data

    # --- Prepare for the next week ---
    # Filter to keep only patients who are not in the absorbing state (y=5)
    active_patients_data <- active_patients_data |>
      filter(y != 5)

    # If there are still active patients, update their state for the next week
    if (nrow(active_patients_data) > 0) {
      active_patients_data <- active_patients_data |>
        mutate(
          ypprev = yprev,
          yprev = y,
          week = w + 1
        )
    }
  }

  # Combine the weekly results into a single data frame
  final_trajectories <- data.table::rbindlist(weekly_results)

  # Return as a standard data.frame
  return(as.data.frame(final_trajectories))
}
```

```{r}
simulated_data <- simulate_trajectories(
  model = fit_second_non_po_lin_int,
  baseline_data = baseline_data,
  knots = knots,
  start_week = 2,
  end_week = 26
)
```

Dataset with death carried forward

```{r}
setDT(simulated_data)

# Create a complete grid of all patients for all weeks using data.table's fast CJ function
all_pat_ids <- unique(baseline_data$pat_id)
complete_grid <- CJ(pat_id = all_pat_ids, week = 2:26)

# Perform a fast join of the simulated data onto the complete grid
completed_dt <- simulated_data[complete_grid, on = .(pat_id, week)]

# Identify the columns that need to be filled
cols_to_fill <- setdiff(names(completed_dt), c("pat_id", "week"))

# Use the fast `na.locf` (Last Observation Carried Forward) by patient group
completed_dt[, (cols_to_fill) := lapply(.SD, zoo::na.locf, na.rm = FALSE), by = pat_id, .SDcols = cols_to_fill]

# Convert back to a standard data frame
sim_df_forward <- as.data.frame(completed_dt)
```

Plot

```{r}
sim_df_forward |> 
  mutate(
    y = factor(y, levels = 1:5),
    tx = factor(tx, levels = c(0, 1))
  ) |>
  group_by(week, y, tx) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(week, tx) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = week, y = prop, color = factor(y), linetype = tx)) +
  geom_line() +
  scale_color_viridis_d(end = 0.9, name = "State") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 1)) +
  labs(
    title = "Simulated State Occupancy Over Time",
    x = "Week",
    y = "Proportion in State"
  ) +
  theme(
    legend.position = "right"
  )
```

# Old code
Getting SOPs for lag = 1. Example for weeks 2 to 3.

```{r}
matrix(predictions[1:steps, ], nrow = steps, ncol = length(ylevels))

row <- d[1, ]

ylevels <- factor(1:5, levels = 1:5)
absorb <- 5
times <- 3:26
steps <- length(ylevels) - 1 # N non-abosrbing

# Day 2 is predicted from known history
state_vector_week_2 <- predict(fit_non_po_lin_int, newdata = row, type = "response")
state_vector_week_2 <- matrix(state_vector_week_2)
state_vector_week_2

# Day 3 is predicted from half known history
transition_matrix_day_3 <- matrix(predictions[1:4, ], nrow = 4, ncol = 5)
transition_matrix_day_3 <- t(rbind(transition_matrix_day_3, c(0, 0, 0, 0, 1)))
transition_matrix_day_3

state_vector_week_3 <- transition_matrix_day_3 %*% state_vector_week_2

transition_matrix_day_4 <- matrix(predictions[5:8, ], nrow = 4, ncol = 5)
transition_matrix_day_4 <- t(rbind(transition_matrix_day_4, c(0, 0, 0, 0, 1)))
transition_matrix_day_4

state_vector_week_4 <- transition_matrix_day_4 %*% state_vector_week_3
````