---
Title: Simulation of constant OR, effect of yprev x time linear interaction
---

## Setup

```{r}
library(tidyverse)
library(data.table)
library(tidybayes)
library(rmsb)
library(rms)
library(Hmisc)
library(arrow)
library(VGAM)
library(patchwork)

set.seed(1234)
```

## Load model and dataset

```{r}
df <- read_parquet("./data/hosp_long_weeks.parquet")  |> 
    mutate(ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE))

model <- readRDS("./2-hosp/models/model_hosp_pol_lin_x.rds")

```

## Extract median of posterior draws

```{r}
draws <- as.data.table(model$draws) 

coef <- draws |> 
  summarise_all(median)

coef
```

## Baseline-Covariates

```{r}
N <- 20000 # Simulation size

baseline_w <- df |> 
    filter(week == 1) |> 
    select(pat_id, yprev,ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)

i <- sample(1:nrow(baseline_w), N, replace = TRUE)
sim_baseline_data <- baseline_w[i, ] |> 
  mutate(pat_id = row_number())

init <- sim_baseline_data$yprev # initial states to be passed to g function

baseline_formula <- ~ ecog_fstcnt + diagnosis + age + gender + albumin + c_reactive_protein
X_baseline <- model.matrix(baseline_formula, data = sim_baseline_data)[, -1] # Baseline values dummy coded, remove intercept

tx <- sample(0:1, N, replace = TRUE) # Sample treatment assignment
X <- cbind(X_baseline, tx) 
```

## PO tx effect, constant over time

- Proportional Part: Effects that are constant across all k-1 logits (e.g., yprev, gap, baseline covariates, pol(week,3)).
- Non-Proportional Part: The week x f(y) term, which varies with the outcome level k. For our model cppo = function(y) y, this term is k * week * \gamma_{week}.

```{r}
ints <- unlist(coef[1:4])
extra <- unlist(coef[- (1:5)]) #remove tx coef as well

names(extra)[names(extra) == "ecog_fstcnt=2"] <- "ecog_fstcnt2"
names(extra)[names(extra) == "ecog_fstcnt=3"] <- "ecog_fstcnt3"
names(extra)[names(extra) == "diagnosis=2"] <- "diagnosis2"
names(extra)[names(extra) == "diagnosis=3"] <- "diagnosis3"
names(extra)[names(extra) == "diagnosis=4"] <- "diagnosis4"
names(extra)[names(extra) == "diagnosis=5"] <- "diagnosis5"
names(extra)[names(extra) == "plan_fstcnt_coded=2"] <- "plan_fstcnt_coded2"
names(extra)[names(extra) == "plan_fstcnt_coded=3"] <- "plan_fstcnt_coded3"
names(extra)[names(extra) == "plan_fstcnt_coded=4"] <- "plan_fstcnt_coded4"
names(extra)[names(extra) == "plan_fstcnt_coded=5"] <- "plan_fstcnt_coded5"
names(extra)[names(extra) == "gender=1"] <- "gender1" # 1 = male

g <- function(yprev, t, gap, X, parameter = 0, extra) {

  # X is a vector for one patient, containing baseline covariates and tx (which we want to remove)
  # yprev is the previous state (character)
  # t is the current week (numeric)
  # gap is assumed to be 1 for this simulation
  gap <- 1 

  # 1. Treatment Effect
  # Treatment effect stable over time
  # The treatment indicator 'tx' is the last element of the X vector.
  tx_value <- X[length(X)]
  tx_effect <- parameter * tx_value

  # 2. Baseline effect
  # X contains the non-treatment variables. We match its names to the 'extra' vector.
  X_nontx <- X[-length(X)] # X is a single vector. We remove tx which is as the end of this vector
  baseline_coef_names <- names(X_nontx)

  baseline_effect <- sum(extra[baseline_coef_names] * X_nontx)

  # 3.  time effect (proportional part) from pol(week, 2)
  time_effect <- extra[["week"]] * t + extra[["week^2"]] * (t^2) + extra[["week^3"]] * (t^3)

  # This is if yprev is equal to 1 (the reference group)
  yprev_effect <- 0
  yprev_week_effect <- 0

  # 4. Previous state effect
  # standalone effects
  yprev_coef_name <- paste0('yprev=', yprev)
  if (yprev_coef_name %in% names(extra)) {
      yprev_effect <- extra[[yprev_coef_name]]
  }

  # interaction with time (week), linear
  yprev_week_coef_name <- paste0('yprev=', yprev, ' * week')
  if (yprev_week_coef_name %in% names(extra)) {
      yprev_week_effect <- extra[[yprev_week_coef_name]] * t
  }

  # Sum of all proportional effects
  lin_pred_prop <- tx_effect + baseline_effect + time_effect + yprev_effect + yprev_week_effect

  # 5. Non-proportional effect of time
  # From cppo=function(y) y, the effect for P(Y>=k) is k * week * gamma
  # The columns of the final lp correspond to k = 2, 3, ..., 5
  nonprop_gamma <- extra[['week x f(y)']]
  k_levels <- 2:5 
  lin_pred_nonprop <- k_levels * t * nonprop_gamma

  # 6. Final linear predictor (vector of length 4)
  # This is the sum of the proportional part (a scalar) and non-proportional part (a vector)
  lp <- lin_pred_prop + lin_pred_nonprop
    
  return(lp)

}

formals(g)$extra <- extra
```

## Simulate dataset with PO for treatment

- Odds of 0.8 of moving to a higher (worse state)

```{r}
# Run the simulation
carried_forward <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:5, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 5,
                  intercepts = ints, 
                  g = g, 
                  parameter = log(0.8),
                  carry = TRUE)

not_carried_forward <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:5, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 5,
                  intercepts = ints, 
                  g = g, 
                  parameter = log(0.8),
                  carry = FALSE)
```

### Calculate the SOPs
```{r}
sops <- carried_forward |> 
  group_by(time, tx) |> 
  count(y) |> 
  reframe(
    y = y,
    sop = n / sum(n))
```

### Calculate the difference in weeks spent at home
```{r}
# Difference in mean days with excellent, very good or good quality of life
weeks_at_home <- sops |> 
  filter(y == 1) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_at_home$mean_weeks[2] - weeks_at_home$mean_weeks[1]

# difference in weeks at home or planned hospitalization
weeks_wo_emerg <- sops |> 
  filter(y == 1 | y == 2) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_wo_emerg$mean_weeks[2] - weeks_wo_emerg$mean_weeks[1]

# difference in RMST
rmst <- sops |> 
  filter(y != 5) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

rmst$mean_weeks[2] - rmst$mean_weeks[1]
```

In this population an OR of 0.8 for a proportional treatment effect leads to a
-  XX week increase in time spent at home 
-  XX weeks more without an emergency / unplanned hospitalization
-  XX week increase in the restricted mean survival time at 6 months.

## Export data

```{r}
write_parquet(carried_forward, "./2-hosp/sim-data/sim_data_linear_interaction_forward.parquet")
```