---
Title: Examining correlation of sim data
---

Have to replace this simdata with the sim data from our analysis model.
Because right now it's actually from the data generating model, which contains far more variables.

## Setup

```{r}
library(arrow)
library(Hmisc)
library(tidyverse)
library(data.table)
```

## Load data
```{r}
sim_data <- read_parquet("./2-hosp/sim-data/h0_forward.parquet") |> 
    rename(
        id = pat_id, 
        time = week)
real_data <- read_parquet("./data/hosp_long_weeks_two_lags.parquet") |> 
    filter(week > 0) |> 
    rename(
        id = pat_id, 
        time = week)
```

## Carry Death Forward

```{r}
df_state_5 <-
    real_data |>
    group_by(id) |>
    mutate(
      y = as.integer(y),
      yprev = as.integer(yprev),
      death_time = ifelse(any(y == 5), time[y == 5], NA_real_)
    ) |>
    filter(y == 5) |>
    complete(time = death_time:26, fill = list(y = 5L, yprev = 5L)) |>
    arrange(id, time) |>
    fill(everything())

df_state_1_to_4 <-
    real_data |>
    mutate(y = as.integer(y), yprev = as.integer(yprev)) |>
    filter(y != 5) |>
    arrange(id, time)

real_forward <-
    bind_rows(df_state_1_to_4, df_state_5) |>
    mutate(y = as.ordered(y), yprev = as.factor(yprev)) |>
    arrange(id, time)
```

## Examine Correlation

### Real data
```{r}
corr_actual <-
    real_forward |> 
    select(id, time, y) |>
    arrange(time) |>
    mutate(y = as.integer(y)) |>
    pivot_wider(names_from = time, values_from = y) |>
    ungroup() |>
    select(-id) |>
    cor(method="spearman", use="pairwise.complete.obs")

corr_actual |>
  as.data.frame() |>
  rownames_to_column() |>
  pivot_longer(-rowname) |>
  mutate(rowname = as.integer(rowname), name = as.integer(name)) |>
  filter(rowname < name) |> #remove tiles below diagonal, because redundant
  mutate(rowname = as.factor(rowname), name = as.factor(name)) |>
  ggplot() + 
    aes(x=rowname, y=name, fill=value) +
    geom_tile() +
    geom_text(mapping=aes(x=rowname, y=name, label=round(value, 2)),
              color="black",
              size = 3) +
    scale_fill_viridis_c(limits=c(-0.01,1), n.breaks = 10) +
    expand_limits(value = c(0, 1)) +
    # cowplot::theme_half_open(font_size=FONT_SIZE,
    #           font_family = "Source Sans Pro") +
    theme(plot.title = element_text(hjust = 0.5),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    theme(axis.line = element_blank(),
          axis.ticks = element_blank()) +
    theme(legend.key.height = unit(0.8, "in")) +
    labs(x="Study week",
         y="Study week",
         fill="",
         subtitle = "Actual Data")

ggsave(filename = "./img/fig-emp-cor-heat-hosp-weekly.svg", height = 10, width = 11)
```

### Simulated Data based on model

```{r}
corr_sim <-
    sim_data |> 
    select(id, time, y) |>
    arrange(time) |>
    mutate(y = as.integer(y)) |>
    pivot_wider(names_from = time, values_from = y) |>
    ungroup() |>
    select(-id) |>
    cor(method="spearman", use="pairwise.complete.obs")

corr_sim |>
  as.data.frame() |>
  rownames_to_column() |>
  pivot_longer(-rowname) |>
  mutate(rowname = as.integer(rowname), name = as.integer(name)) |>
  filter(rowname < name) |> #remove tiles below diagonal, because redundant
  mutate(rowname = as.factor(rowname), name = as.factor(name)) |>
  ggplot() + 
    aes(x=rowname, y=name, fill=value) +
    geom_tile() +
    geom_text(mapping=aes(x=rowname, y=name, label=round(value, 2)),
              color="black",
              size = 3) +
    scale_fill_viridis_c(limits=c(-0.01,1), n.breaks = 10) +
    expand_limits(value = c(0, 1)) +
    # cowplot::theme_half_open(font_size=FONT_SIZE,
    #           font_family = "Source Sans Pro") +
    theme(plot.title = element_text(hjust = 0.5),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    theme(axis.line = element_blank(),
          axis.ticks = element_blank()) +
    theme(legend.key.height = unit(0.8, "in")) +
    labs(x="Study week",
         y="Study week",
         fill="",
         subtitle = "Simulated Data (no interaction)")

ggsave(filename = "./img/fig-sim-cor-heat.svg", height = 10, width = 11)
```
### Absolute differences between correlation matrices

Means of absolute differences, stratified by the day involved.
```{r}
ad <- abs(corr_actual - corr_sim)
m <- round(apply(ad, 1, mean), 2) |> data.table() |> 
    rename(abs_diff = V1)

m |> 
  rownames_to_column("rownames") |>
  pivot_longer(names_to = "model", values_to = "value", -rownames) |>
  pivot_wider(names_from = rownames, values_from = value) |> 
  knitr::kable()


m |> summarise_all(sum)
m |> summarise_all(mean)
```

### Variograms

```{r}
Hmisc::plotCorrM(corr_actual)[[2]] +
    scale_y_continuous(limits = c(-.1, 1))

ggsave(filename = "./img/fig-hosp-weekly-variogram-emp.svg", height = 10, width = 11)

Hmisc::plotCorrM(corr_no)[[2]] + 
    scale_y_continuous(limits = c(-.1, 1))

ggsave(filename = "./img/fig-hosp-weekly-variogram-no-int.svg", height = 10, width = 11)

Hmisc::plotCorrM(corr_lin)[[2]] + 
    scale_y_continuous(limits = c(-.1, 1))

ggsave(filename = "./img/fig-hosp-weekly-variogram-lin-int.svg", height = 10, width = 11)

Hmisc::plotCorrM(corr_full)[[2]] + 
    scale_y_continuous(limits = c(-.1, 1))

ggsave(filename = "./img/fig-hosp-weekly-variogram-full-int.svg", height = 10, width = 11)
```


