---
Title: "Calculating Marginal State Occupancy Probabilities from a Markov Model"
---

And also simulating patient trajectories. Special focus on second-order Markov models.

## Setup

```{r}
library(tidyverse)
library(VGAM)
library(data.table)
library(arrow)
library(rms)
library(zoo)
library(here)
```

## Prepare Data

```{r}
d <- read_parquet("./data/hosp_long_weeks.parquet") |> 
    mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE),
        y = ordered(y)
    ) |> 
      select(pat_id, tx, y, yprev, time = week, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein) |> 
      as.data.table()
```

## Fit simplest model - H0, PO for everyting, no covariates

```{r}
time_spl_m <- rcs(d$time, 4)
knots <- attr(time_spl_m, "parms")

d$time_linear <- time_spl_m[, 1]
d$time_nonlinear_1 <- time_spl_m[, 2]
d$time_nonlinear_2 <- time_spl_m[, 3]

M <- 4 # n thresholds

cons <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "yprev"              = cbind(PO_effect = 1),
  "time_linear"        = cbind(PO_effect = 1),
  "time_nonlinear_1"   = cbind(PO_effect = 1),
  "time_nonlinear_2"   = cbind(PO_effect = 1)

  # The linear component of the spline gets the special constraint
#   "time_linear" = cbind(
#     baseline_effect = 1,
#     linear_trend    = 0:(M - 1)
#   )
)

fit_po <- vglm(
  y ~ yprev + time_linear + time_nonlinear_1 + time_nonlinear_2,
  family = cumulative(parallel = TRUE, reverse = TRUE),
  data = d,
  constraints = cons
)
```

### Simulating patient trajectories

First create baseline data and create spline matrix

```{r}
knots <- attr(time_spl_m, "parms")

N <- 5e4
baseline_data  <- d |>
  filter(time == 1) |> 
  select(pat_id, time, yprev, ecog_fstcnt, diagnosis, age, albumin) |> 
  slice_sample(n = N, replace = TRUE) |> 
  mutate(
    pat_id = row_number()
  ) |> 
    data.frame()
```

Simulation function for first order model

```{r}
simulate_trajectories <- function(model, baseline_data, knots, start_week = 1, end_week = 26) {

  set.seed(123)
  
  # List to store the results from each week
  weekly_results <- list()

  # Initialize with the cohort of patients active at the start
  active_patients_data <- baseline_data

  # Loop through each week, processing all active patients at once
  for (w in start_week:end_week) {

    # If no patients are left, stop the simulation
    if (nrow(active_patients_data) == 0) {
      break
    }

    # Ensure factor levels are consistent for prediction
    active_patients_data$yprev <- factor(active_patients_data$yprev, levels = 1:4)

    # Recalculate spline basis functions for the entire active cohort
    new_spline_matrix <- rcs(active_patients_data$time, parms = knots)
    colnames(new_spline_matrix) <- c("time_linear", "time_nonlinear_1", "time_nonlinear_2")
    
    active_patients_data$time_linear <- new_spline_matrix[, 1]
    active_patients_data$time_nonlinear_1 <- new_spline_matrix[, 2]
    active_patients_data$time_nonlinear_2 <- new_spline_matrix[, 3]
    
    # Predict probabilities for ALL active patients. This returns a matrix.
    probs_matrix <- predict(model, newdata = active_patients_data, type = "response")
    
    # For each patient (row in the matrix), sample a new outcome 'y'
    y_new_vector <- apply(probs_matrix, 1, function(p) {
      # Normalize probabilities just in case of floating point issues
      p_norm <- pmax(0, p) / sum(pmax(0, p))
      sample(1:5, size = 1, prob = p_norm)
    })

    # Add the new outcomes to the data for the current week
    active_patients_data$y <- y_new_vector

    # Store the results for the current week
    weekly_results[[as.character(w)]] <- active_patients_data

    # --- Prepare for the next week ---
    # Filter to keep only patients who are not in the absorbing state (y=5)
    active_patients_data <- active_patients_data |>
      filter(y != 5)

    # If there are still active patients, update their state for the next week
    if (nrow(active_patients_data) > 0) {
      active_patients_data <- active_patients_data |>
        mutate(
          yprev = y,
          time = w + 1
        )
    }
  }

  # Combine the weekly results into a single data frame
  final_trajectories <- data.table::rbindlist(weekly_results)

  # Return as a standard data.frame
  return(as.data.frame(final_trajectories))
}
```

```{r}
simulated_data <- simulate_trajectories(
  model = fit_po,
  baseline_data = baseline_data,
  knots = knots,
  start_week = 1,
  end_week = 26
)

write_parquet(simulated_data, here("2-hosp", "sim-data", "H0-po.parquet"))
```

Dataset with death carried forward

```{r}
setDT(simulated_data)

# Create a complete grid of all patients for all weeks using data.table's fast CJ function
all_pat_ids <- unique(baseline_data$pat_id)
complete_grid <- CJ(pat_id = all_pat_ids, time = 2:26)

# Perform a fast join of the simulated data onto the complete grid
completed_dt <- simulated_data[complete_grid, on = .(pat_id, time)]

# Identify the columns that need to be filled
cols_to_fill <- setdiff(names(completed_dt), c("pat_id", "time"))

# Use the fast `na.locf` (Last Observation Carried Forward) by patient group
completed_dt[, (cols_to_fill) := lapply(.SD, zoo::na.locf, na.rm = FALSE), by = pat_id, .SDcols = cols_to_fill]

# Convert back to a standard data frame
sim_df_forward <- as.data.frame(completed_dt)
```

Plot

```{r}
sim_df_forward |> 
  group_by(pat_id) |> 
  mutate(tx = sample(c("0", "1"), size = 1, replace = FALSE)) |> 
  ungroup() |> 
  mutate(
    y = factor(y, levels = 1:5)
  ) |>
  group_by(time, y, tx) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(time, tx) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = time, y = prop, color = factor(y), linetype = tx)) +
  geom_line() +
  scale_color_viridis_d(end = 0.9, name = "State") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(2, 26, by = 2)) +
  labs(
    title = "Simulated State Occupancy Over Time",
    x = "Week",
    y = "Proportion in State"
  ) +
  theme(
    legend.position = "right"
  )
```

## H0, PO for everyting, with covariates

```{r}
cons <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "yprev"              = cbind(PO_effect = 1),
  "time_linear"        = cbind(PO_effect = 1),
  "time_nonlinear_1"   = cbind(PO_effect = 1),
  "time_nonlinear_2"   = cbind(PO_effect = 1),
  "ecog_fstcnt"        = cbind(PO_effect = 1),
  "diagnosis"          = cbind(PO_effect = 1),
  "age"   = cbind(PO_effect = 1),
  "albumin"   = cbind(PO_effect = 1)
)

fit_po <- vglm(
  y ~ yprev + time_linear + time_nonlinear_1 + time_nonlinear_2 + ecog_fstcnt + diagnosis + age + albumin,
  family = cumulative(parallel = TRUE, reverse = TRUE),
  data = d,
  constraints = cons
)
```

### Simulating patient trajectories

```{r}
simulated_data <- simulate_trajectories(
  model = fit_po,
  baseline_data = baseline_data,
  knots = knots,
  start_week = 1,
  end_week = 26
)

write_parquet(simulated_data, here("2-hosp", "sim-data", "H0-po-wi-cov.parquet"))
```

Dataset with death carried forward

```{r}
setDT(simulated_data)

# Create a complete grid of all patients for all weeks using data.table's fast CJ function
all_pat_ids <- unique(baseline_data$pat_id)
complete_grid <- CJ(pat_id = all_pat_ids, time = 2:26)

# Perform a fast join of the simulated data onto the complete grid
completed_dt <- simulated_data[complete_grid, on = .(pat_id, time)]

# Identify the columns that need to be filled
cols_to_fill <- setdiff(names(completed_dt), c("pat_id", "time"))

# Use the fast `na.locf` (Last Observation Carried Forward) by patient group
completed_dt[, (cols_to_fill) := lapply(.SD, zoo::na.locf, na.rm = FALSE), by = pat_id, .SDcols = cols_to_fill]

# Convert back to a standard data frame
sim_df_forward <- as.data.frame(completed_dt)
```

Plot

```{r}
sim_df_forward |> 
  group_by(pat_id) |> 
  mutate(tx = sample(c("0", "1"), size = 1, replace = FALSE)) |> 
  ungroup() |> 
  mutate(
    y = factor(y, levels = 1:5)
  ) |>
  group_by(time, y, tx) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(time, tx) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = time, y = prop, color = factor(y), linetype = tx)) +
  geom_line() +
  scale_color_viridis_d(end = 0.9, name = "State") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(2, 26, by = 2)) +
  labs(
    title = "Simulated State Occupancy Over Time",
    x = "Week",
    y = "Proportion in State"
  ) +
  theme(
    legend.position = "right"
  )
```

## Fit model - H0, non-PO for time with linear constraint, no covariates

```{r}
time_spl_m <- rcs(d$time, 4)
knots <- attr(time_spl_m, "parms")

d$time_linear <- time_spl_m[, 1]
d$time_nonlinear_1 <- time_spl_m[, 2]
d$time_nonlinear_2 <- time_spl_m[, 3]

M <- 4 # n thresholds

cons <- list(
  # Intercepts of course have separate parameters
  "(Intercept)"     = diag(M),

  # Proportional odds terms now get "column of ones" matrix
  "yprev"              = cbind(PO_effect = 1),
  "time_nonlinear_1"   = cbind(PO_effect = 1),
  "time_nonlinear_2"   = cbind(PO_effect = 1),

  # The linear component of the spline gets the special constraint
  "time_linear" = cbind(
    baseline_effect = 1,
    linear_trend    = 0:(M - 1)
  )
)

fit_nonpo <- vglm(
  y ~ yprev + time_linear + time_nonlinear_1 + time_nonlinear_2,
  family = cumulative(parallel = FALSE ~ time_linear, reverse = TRUE),
  data = d,
  constraints = cons
)
```

### Simulating patient trajectories

Simulation function for first order model with linear devation from nonPO for time

```{r}
simulated_data <- simulate_trajectories(
  model = fit_nonpo,
  baseline_data = baseline_data,
  knots = knots,
  start_week = 1,
  end_week = 26
)

write_parquet(simulated_data, here("2-hosp", "sim-data", "H0-nonpo-time-lin-constraint.parquet"))
```

Dataset with death carried forward

```{r}
setDT(simulated_data)

# Create a complete grid of all patients for all weeks using data.table's fast CJ function
all_pat_ids <- unique(baseline_data$pat_id)
complete_grid <- CJ(pat_id = all_pat_ids, time = 2:26)

# Perform a fast join of the simulated data onto the complete grid
completed_dt <- simulated_data[complete_grid, on = .(pat_id, time)]

# Identify the columns that need to be filled
cols_to_fill <- setdiff(names(completed_dt), c("pat_id", "time"))

# Use the fast `na.locf` (Last Observation Carried Forward) by patient group
completed_dt[, (cols_to_fill) := lapply(.SD, zoo::na.locf, na.rm = FALSE), by = pat_id, .SDcols = cols_to_fill]

# Convert back to a standard data frame
sim_df_forward <- as.data.frame(completed_dt)
```

Plot

```{r}
sim_df_forward |> 
  group_by(pat_id) |> 
  mutate(tx = sample(c("0", "1"), size = 1, replace = FALSE)) |> 
  ungroup() |> 
  mutate(
    y = factor(y, levels = 1:5)
  ) |>
  group_by(time, y, tx) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(time, tx) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = time, y = prop, color = factor(y), linetype = tx)) +
  geom_line() +
  scale_color_viridis_d(end = 0.9, name = "State") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(2, 26, by = 2)) +
  labs(
    title = "Simulated State Occupancy Over Time",
    x = "Week",
    y = "Proportion in State"
  ) +
  theme(
    legend.position = "right"
  )
