---
Title: Examining Type I Error with Resampling - No Interaction time x yrev
---

## Setup
```{r}
# Load Libraries
library(Hmisc)
library(rmsb)
library(arrow)
library(tidyverse)
library(data.table)
library(furrr)
library(glue)
library(fs)
library(here)
```

## Load data

```{r}
set.seed(1234)

# Load full dataset
# If we reasign the treatment again and again, emergency room visits are a bit too rare and mess up transitions
# We therefore collapse staet 3 and 4
df <- read_parquet(here("data", "hosp_long_weeks_anon.parquet")) |> 
    mutate(
        y = if_else(y == 4, 3, y),
        y = if_else(y == 5, 4, y),
        yprev = if_else(yprev == "4", "3", yprev),
        yprev = factor(yprev)) |> 
    select(id, tx, y, yprev, time = week)
```

## Functions

```{r}
run_simulation_iteration <- function(iter_num, base_path, full_data, model_formula, prior_spec = NULL) {
    # This function runs one full simulation iteration.
    
    # --- 1. Setup for the current iteration ---
    set.seed(123 + iter_num)
    iter_path <- file.path(base_path, glue("run_{iter_num}"))
    if (!dir.exists(iter_path)) {
        dir.create(iter_path, recursive = TRUE)
    }

    # Resample from data with no effect
    tx_ids <- full_data |> 
        distinct(id) |> 
        sample_frac(2/3) |> pull(id)

    data_for_model <- full_data |> 
        mutate(
            tx = if_else(id %in% tx_ids, "1", "0"),
            tx = factor(tx, levels = c(0, 1)))
    
    rm(full_data)


    # --- 5. Save the data used for model fitting ---
    saveRDS(data_for_model, file.path(iter_path, "data_for_model.rds"))


    # --- 6. Fit the Model and Save ---

    # we'll stick to evaluating type I error rate via OR of tx for the time being, then we don't need to marginalize
   
    if (nrow(data_for_model) > 0) {
        options(mc.cores = 1)
        dd <- datadist(data_for_model)
        options(datadist = 'dd')

        # Forcefully place the 'dd' object into the global environment of the worker.
        assign("dd", dd, envir = .GlobalEnv)
        on.exit(rm(dd, envir = .GlobalEnv))

        blrm_args <- list(
            formula = model_formula,
            data = data_for_model,
            ppo = ~ time,
            cppo = function(y) y,
            refresh = 0,
            iter = 2000,
            chains = 4,
            seed = 1234,
            loo = FALSE,
            method = "sampling"
        )
        
        if (!is.null(prior_spec)) {
            blrm_args$pcontrast <- prior_spec
        }

        model <- tryCatch({
            do.call(blrm, blrm_args)
        }, error = function(e) {
            message(glue("ERROR in iteration {iter_num}: {e$message}"))
            
            saveRDS(data_for_model, file.path(iter_path, "problematic_data.rds"))
            
            return(NULL)
        })
        
        # --- 7. Save the results ONLY if the model ran successfully ---
        if (!is.null(model)) {
            # Save the full model object
            saveRDS(model, file.path(iter_path, "model_object.rds"))
            
            # Save just the MCMC draws to a separate file
            saveRDS(as.data.table(model$draws), file.path(iter_path, "model_draws.rds"))

            rhat_values <- stanDx(model) |> 
                data.frame() |> 
                rownames_to_column()  |> 
                select(-n_eff) |> 
                pivot_wider(names_from = rowname, values_from = Rhat)

            saveRDS(rhat_values, file.path(iter_path, "rhat_values.rds"))

            ess_ratios <- stanDx(model) |> 
                data.frame() |> 
                rownames_to_column()  |> 
                select(-Rhat) |> 
                pivot_wider(names_from = rowname, values_from = n_eff) |> 
                mutate_all(~ . / (model$iter * model$chains))

             saveRDS(ess_ratios, file.path(iter_path, "ess_ratios.rds"))
        }
    }
    # Return a status message for progress tracking
    return(glue("Iteration {iter_num} completed."))
}
```

```{r}
check_completed_simulations <- function(output_path, total_sims) {
    completed_files <- fs::dir_ls(
        path = output_path, 
        glob = "*/model_draws.rds", 
        recurse = TRUE
    )

    completed_iters <- integer(0) 
    if (length(completed_files) > 0) {
        completed_iters <- stringr::str_extract(completed_files, "(?<=run_)\\d+") |>
            as.integer() |>
            na.omit() |>
            unique()
    }

    all_possible_iters <- 1:total_sims
    iters_to_run <- setdiff(all_possible_iters, completed_iters)

    return(list(
        iters_to_run = iters_to_run,
        n_to_run = length(iters_to_run),
        n_completed = length(completed_iters)
    ))
}
```
## Run Simulations
```{r}
N_SIMULATIONS <- 100
N_CORES <- 8
future::plan(future.callr::callr, workers = N_CORES)
```

### No yprev x time interaction
```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-resampling-H0-no-int")
sim_status <- check_completed_simulations(OUTPUT_PATH, N_SIMULATIONS)

if (sim_status$n_to_run > 0) {
    message(glue("Starting {sim_status$n_to_run} new simulations on {N_CORES} cores..."))
    
    formula <- y  ~ tx + yprev + rcs(time, 4)
    
    # --- Run Simulations in Parallel ---
    furrr::future_walk(
        .x = sim_status$iters_to_run,
        .f = ~run_simulation_iteration(
            iter_num = .x,
            base_path = OUTPUT_PATH,
            full_data = df,
            model_formula = formula
        ),
        .options = furrr_options(seed = TRUE),
        .progress = FALSE
    )
    
    message(glue("Finished {sim_status$n_to_run} simulations."))
} else {
    message("All simulations were already complete. No new runs started.")
}
```

### Linear nteraction time x yprev

```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-resampling-H0-linear-int")
sim_status <- check_completed_simulations(OUTPUT_PATH, N_SIMULATIONS)

if (sim_status$n_to_run > 0) {
    message(glue("Starting {sim_status$n_to_run} new simulations on {N_CORES} cores..."))

    formula <- y  ~ tx + yprev + rcs(time, 4) + time %ia% yprev
    
    # --- Run Simulations in Parallel ---
    furrr::future_walk(
        .x = sim_status$iters_to_run,
        .f = ~run_simulation_iteration(
            iter_num = .x,
            base_path = OUTPUT_PATH,
            full_data = df,
            model_formula = formula
        ),
        .options = furrr_options(seed = TRUE),
        .progress = FALSE
    )
    
    message(glue("Finished {sim_status$n_to_run} simulations."))
} else {
    message("All simulations were already complete. No new runs started.")
}
```

### Full Interaction time x yprev

```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-resampling-H0-full-int")
sim_status <- check_completed_simulations(OUTPUT_PATH, N_SIMULATIONS)

if (sim_status$n_to_run > 0) {
    message(glue("Starting {sim_status$n_to_run} new simulations on {N_CORES} cores..."))
    
    formula <- y  ~ tx + yprev * rcs(time, 4)
    
    # --- Run Simulations in Parallel ---
    furrr::future_walk(
        .x = sim_status$iters_to_run,
        .f = ~run_simulation_iteration(
            iter_num = .x,
            base_path = OUTPUT_PATH,
            full_data = df,
            model_formula = formula
        ),
        .options = furrr_options(seed = TRUE),
        .progress = FALSE
    )
    
    message(glue("Finished {sim_status$n_to_run} simulations."))
} else {
    message("All simulations were already complete. No new runs started.")
}
```

### No Interaction, but regularizing prior

I want to test if a regularizing prior impact the frequentist type I error rate, as I could imaginge that the default prior of Normal(0, 100) could cause some issues. 

Note: After 71 iterations it does not seem to change a lot, slightly lower bias, but impact seems neglibible. 

```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-resampling-H0-no-int-regularizing")
sim_status <- check_completed_simulations(OUTPUT_PATH, N_SIMULATIONS)

if (sim_status$n_to_run > 0) {
    message(glue("Starting {sim_status$n_to_run} new simulations on {N_CORES} cores..."))
    
    formula <- y  ~ tx + yprev + rcs(time, 4)
    
    # Define the regularizing prior
    prior_tx <- list(
      c1 = list(tx = "1"),
      contrast = expression(c1),
      mu = 0,
      sd = 2
    )
    
    # --- Run Simulations in Parallel ---
    furrr::future_walk(
        .x = sim_status$iters_to_run,
        .f = ~run_simulation_iteration(
            iter_num = .x,
            base_path = OUTPUT_PATH,
            full_data = df,
            model_formula = formula,
            prior_spec = prior_tx
        ),
        .options = furrr_options(seed = TRUE),
        .progress = FALSE
    )
    
    message(glue("Finished {sim_status$n_to_run} simulations."))
} else {
    message("All simulations were already complete. No new runs started.")
}
```