---
Title: Examining Type I Error with Resampling - No Interaction time x yrev
---

## Setup
```{r}
# Load Libraries
library(Hmisc)
library(rmsb)
library(arrow)
library(tidyverse)
library(data.table)
library(furrr)
library(glue)
library(fs)
library(here)

source(here("functions", "check-completed-simulations.R"))
source(here("functions", "bayes-hosp-run-markov-simulation-iteration.R"))
```

## Load data

```{r}
set.seed(1234)

# Load full dataset
# If we reasign the treatment again and again, emergency room visits are a bit too rare and mess up transitions
# We therefore collapse staet 3 and 4
df <- read_parquet(here("data", "hosp_long_weeks_anon.parquet")) |> 
    mutate(
        y = if_else(y == 4, 3, y),
        y = if_else(y == 5, 4, y),
        yprev = if_else(yprev == "4", "3", yprev),
        yprev = factor(yprev)) |> 
    select(id, tx, y, yprev, time = week)
```

## Run Simulations
```{r}
N_SIMULATIONS <- 1205
N_CORES <- 8
future::plan(future.callr::callr, workers = N_CORES)
```

### No yprev x time interaction
```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-resampling-H0-no-int")
sim_status <- check_completed_simulations(OUTPUT_PATH, N_SIMULATIONS)

if (sim_status$n_to_run > 0) {
    message(glue("Starting {sim_status$n_to_run} new simulations on {N_CORES} cores..."))
    
    formula <- y  ~ tx + yprev + rcs(time, 4)
    
    # --- Run Simulations in Parallel ---
    furrr::future_walk(
        .x = sim_status$iters_to_run,
        .f = ~run_simulation_iteration(
            iter_num = .x,
            base_path = OUTPUT_PATH,
            full_data = df,
            model_formula = formula
        ),
        .options = furrr_options(seed = TRUE, packages = c("rmsb", "data.table", "tidyverse")),
        .progress = FALSE
    )
    
    message(glue("Finished {sim_status$n_to_run} simulations."))
} else {
    message("All simulations were already complete. No new runs started.")
}
```

### Linear nteraction time x yprev

```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-resampling-H0-linear-int")
sim_status <- check_completed_simulations(OUTPUT_PATH, N_SIMULATIONS)

if (sim_status$n_to_run > 0) {
    message(glue("Starting {sim_status$n_to_run} new simulations on {N_CORES} cores..."))

    formula <- y  ~ tx + yprev + rcs(time, 4) + time %ia% yprev
    
    # --- Run Simulations in Parallel ---
    furrr::future_walk(
        .x = sim_status$iters_to_run,
        .f = ~run_simulation_iteration(
            iter_num = .x,
            base_path = OUTPUT_PATH,
            full_data = df,
            model_formula = formula
        ),
        .options = furrr_options(seed = TRUE, packages = c("rmsb", "data.table", "tidyverse")),
        .progress = FALSE
    )
    
    message(glue("Finished {sim_status$n_to_run} simulations."))
} else {
    message("All simulations were already complete. No new runs started.")
}
```

### Full Interaction time x yprev

```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-resampling-H0-full-int")
sim_status <- check_completed_simulations(OUTPUT_PATH, N_SIMULATIONS)

if (sim_status$n_to_run > 0) {
    message(glue("Starting {sim_status$n_to_run} new simulations on {N_CORES} cores..."))
    
    formula <- y  ~ tx + yprev * rcs(time, 4)
    
    # --- Run Simulations in Parallel ---
    furrr::future_walk(
        .x = sim_status$iters_to_run,
        .f = ~run_simulation_iteration(
            iter_num = .x,
            base_path = OUTPUT_PATH,
            full_data = df,
            model_formula = formula
        ),
        .options = furrr_options(seed = TRUE, packages = c("rmsb", "data.table", "tidyverse")),
        .progress = FALSE
    )
    
    message(glue("Finished {sim_status$n_to_run} simulations."))
} else {
    message("All simulations were already complete. No new runs started.")
}
```

### No Interaction, but regularizing prior

I want to test if a regularizing prior impact the frequentist type I error rate, as I could imaginge that the default prior of Normal(0, 100) could cause some issues. 

Note: After 71 iterations it does not seem to change a lot, slightly lower bias, but impact seems neglibible. 

```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-resampling-H0-no-int-regularizing")
sim_status <- check_completed_simulations(OUTPUT_PATH, N_SIMULATIONS)

if (sim_status$n_to_run > 0) {
    message(glue("Starting {sim_status$n_to_run} new simulations on {N_CORES} cores..."))
    
    formula <- y  ~ tx + yprev + rcs(time, 4)
    
    # Define the regularizing prior
    prior_tx <- list(
      c1 = list(tx = "1"),
      contrast = expression(c1),
      mu = 0,
      sd = 2
    )
    
    # --- Run Simulations in Parallel ---
    furrr::future_walk(
        .x = sim_status$iters_to_run,
        .f = ~run_simulation_iteration(
            iter_num = .x,
            base_path = OUTPUT_PATH,
            full_data = df,
            model_formula = formula,
            prior_spec = prior_tx
        ),
        .options = furrr_options(seed = TRUE, packages = c("rmsb", "data.table", "tidyverse")),
        .progress = FALSE
    )
    
    message(glue("Finished {sim_status$n_to_run} simulations."))
} else {
    message("All simulations were already complete. No new runs started.")
}
```