---
Title: Results of Simulations
---

In this document we summarise and visualize the results of the simulations from step 6.

## Setup

### Load packages

```{r}
library(fs)
library(tidyverse)
library(arrow)
library(survival)
library(patchwork)
```

### Set Threshold for P(Benefit)

```{r}
threshold <- 0.95
```

### Functions

```{r}
source("./functions/jackknife_mcse.R")
source("./functions/summarise-posterior.R")
source("./functions/summarise-iterations.R")
source("./functions/plot-rhat.R")
source("./functions/plot-p-benefit.R")
source("./functions/plot-posterior-distribution.R")
source("./functions/load-draws.R")
source("./functions/load-rhat-values.R")
```

## Markov Models

### Resampling H0

#### Load draws + diagnostics
```{r}
# H0 no interaction
path <- "./2-hosp/output/sim-resampling-H0-no-int/"
no_int_h0 <- load_draws(path)
rhat_no_int_h0 <- load_rhat_values(path = path)

# H0 linear interaction
path <- "./2-hosp/output/sim-resampling-H0-linear-int/"
linear_int_h0 <- load_draws(path)
rhat_linear_int_h0 <- load_rhat_values(path = path)

# H0 full interaction
path <- "./2-hosp/output/sim-resampling-H0-full-int/"
full_int_h0 <- load_draws(path)
rhat_full_int_h0 <- load_rhat_values(path = path)
```

#### Convergence diagnostics
```{r}
rhat_plot_no_int_h0 <- rhat_no_int_h0 |> 
    plot_rhat(y = `tx=1`)

rhat_plot_linear_int_h0 <- rhat_linear_int_h0 |>
    plot_rhat(y = `tx=1`)

rhat_plot_full_int_h0 <- rhat_full_int_h0 |>
    plot_rhat(y = `tx=1`)

# ggsave(filename = "./img/fig-rhat-markov.svg", width = 10, height = 5)
```

### Plot probability of benefit

```{r}
p1 <- no_int_h0 |> 
    plot_p_benefit(param = `tx=1`)

p2 <- linear_int_h0 |>
    plot_p_benefit(param = `tx=1`)

p3 <- full_int_h0 |>
    plot_p_benefit(param = `tx=1`)

p1 + p2 + p3

# ggsave(filename = "./img/fig-pbenefit-markov.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_no_int_h0 <- no_int_h0 |> 
  plot_posterior_distribution(x_variable = `tx=1`)

post_linear_int_h0 <- linear_int_h0 |>
  plot_posterior_distribution(x_variable = `tx=1`)

post_full_int_h0 <- full_int_h0 |>
  plot_posterior_distribution(x_variable = `tx=1`)

post_no_int_h0 + post_linear_int_h0 + post_full_int_h0

# ggsave(filename = "./img/fig-posteriors-markov.svg", width = 10, height = 5)
```

```{r}
summary_no_int_h0 <- no_int_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = `tx=1`, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_linear_int_h0 <- linear_int_h0 |>
    group_by(iter) |> 
    summarise_posterior(
        param_name = `tx=1`, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_full_int_h0 <- full_int_h0 |>
    group_by(iter) |> 
    summarise_posterior(
        param_name = `tx=1`, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

# rm(markov_h0, markov_ha)
```

## Count Regression

### Resampling H0

#### Load draws + diagnostics
```{r}
# H0 no interaction
path <- "./2-hosp/output/sim-resampling-H0-neg-binom/"
neg_binom_h0 <- load_draws(path)
rhat_neg_binom_h0 <- load_rhat_values(path = path)

ate_neg_binom_h0 <- load_draws(path = path, glob = "*/ATE_draws.rds")
```

#### Convergence diagnostics
```{r}
rhat_plot_neg_binom_h0 <- rhat_neg_binom_h0 |> 
    plot_rhat(y = b_tx1)

# ggsave(filename = "./img/fig-rhat-markov.svg", width = 10, height = 5)
```

### Plot probability of benefit

```{r}
p1 <- neg_binom_h0 |> 
    plot_p_benefit(param = b_tx1)

p2 <- ate_neg_binom_h0 |>
    plot_p_benefit(param = draw)

p1 + p2

# ggsave(filename = "./img/fig-pbenefit-markov.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_neg_binom_h0 <- neg_binom_h0 |> 
  plot_posterior_distribution(x_variable = b_tx1) + 
  scale_x_continuous(limits = c(-1, 1))

post_neg_binom_h0

ate_post_neg_binom_h0 <- ate_neg_binom_h0 |> 
  plot_posterior_distribution(x_variable = draw)

# ggsave(filename = "./img/fig-posteriors-markov.svg", width = 10, height = 5)
```

```{r}
summary_neg_binom_h0 <- neg_binom_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx1, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()
```