---
Title: Examining Type I Error with Resampling - No Interaction time x yrev FREQUENTIST
---

In this document I am investigating how different specifications of the interaction with time and previous state affect the type I error rate, and whether a second order model fits the data better than a first order model.

## Setup
```{r}
# Load Libraries
library(Hmisc)
library(ggplot2)
library(arrow)
library(tidyverse)
library(data.table)
library(furrr)
library(glue)
library(fs)
library(here)
library(rms)
library(VGAM)
```

```{r}
source("./functions/jackknife_mcse.R")

simulate <- FALSE
```

## Load data

```{r}
set.seed(1234)

# If we reasign the treatment again and again, emergency room visits are a bit too rare and mess up transitions. We therefore collapse staet 3 and 4. 
df <- read_parquet(here("data", "hosp_long_weeks_anon.parquet")) |> 
    mutate(
        y = if_else(y == 4, 3, y),
        y = if_else(y == 5, 4, y),
        yprev = if_else(yprev == "4", "3", yprev),
        yprev = factor(yprev)) |> 
    select(id, tx, y, yprev, time = week)

df_ypprev <- read_parquet(here("data", "hosp_long_weeks_two_lags.parquet")) |> 
    mutate(
        y = if_else(y == 4, 3, y),
        y = if_else(y == 5, 4, y),
        yprev = if_else(yprev == "4", "3", yprev),
        yprev = factor(yprev)) |> 
    select(id = pat_id, tx, y, yprev, ypprev, time = week)
```

## Simulation function

```{r}
run_simulation_iteration <- function(iter_num, base_path, full_data, main_formula, cumulative_arg) {
    # This function runs one full simulation iteration.
    
    # --- 1. Setup for the current iteration ---
    set.seed(123 + iter_num)

    tx_ids <- full_data |> 
        distinct(id) |> 
        sample_frac(2/3) |> pull(id)

    data_for_model <- full_data |> 
        mutate(
            tx = if_else(id %in% tx_ids, "1", "0"),
            tx = factor(tx, levels = c(0, 1)))
    
    rm(full_data)
    
    if (nrow(data_for_model) > 0) {
        model <- tryCatch({ 
            vglm(
                main_formula,
                cumulative_arg,
                data = data_for_model)
            
        }, error = function(e) {
            message(glue("ERROR in iteration {iter_num}: {e$message}"))
            return(NULL)
        })
        
        # --- 7. Save the results ONLY if the model ran successfully ---
        if (!is.null(model)) {
            summary_df <- summary(model)@coef3 |> data.frame() |> 
                rownames_to_column("term")

            return(summary_df)
        }
    }
    return(NULL)
}
```

Resampling WITH replacement
```{r}
run_simulation_iteration_boot <- function(iter_num, base_path, full_data, main_formula, cumulative_arg) {

    set.seed(123 + iter_num)

    unique_ids <- unique(full_data$id)
    
    resampled_ids <- sample(unique_ids, size = 500, replace = TRUE)
    
    resampled_map <- tibble(id = resampled_ids) |> 
        mutate(new_id = row_number())
        
    bootstrapped_data <- resampled_map |> 
        left_join(full_data, by = "id")

    # here 1:1 randomization for lower variance
    tx_ids <- bootstrapped_data |> 
        distinct(new_id) |> 
        sample_frac(0.5) |> 
        pull(new_id)

    # Create the final data for the model with the 'tx' (treatment) column.
    data_for_model <- bootstrapped_data |> 
        mutate(
            tx = if_else(new_id %in% tx_ids, "1", "0"),
            tx = factor(tx, levels = c(0, 1))
        )
    
    # Clean up to save memory
    rm(full_data, bootstrapped_data, resampled_map)
    
    # --- 6. Run the model ---
    if (nrow(data_for_model) > 0) {
        model <- tryCatch({
            vglm(
                main_formula,
                cumulative_arg,
                data = data_for_model
            )
        }, error = function(e) {
            message(glue("ERROR in iteration {iter_num}: {e$message}"))
            return(NULL)
        })
        
        # --- 7. Save the results ONLY if the model ran successfully ---
        if (!is.null(model)) {
            summary_df <- summary(model)@coef3 |> 
                data.frame() |> 
                tibble::rownames_to_column("term")
                
            return(summary_df)
        }
    }
    return(NULL)
}
```

Using ordinal package


## Run Parallelized Simulation

```{r}
N_SIMULATIONS <- 1000
N_CORES <- 8
```

### Sim without interaction
```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-resampling-H0-freq")
RESULTS_FILE_NO <- "sim-resampling-H0-no-int-freq.parquet"
RESULTS_FILE_NO_BOOT <- "sim-resampling-H0-no-int-freq-boot.parquet"

if(simulate){
    main_formula <- ordered(y) ~ tx + rcs(time, 4) + yprev
    cumulative_arg <- cumulative(reverse = TRUE, parallel = FALSE ~ rcs(time, 4))

    future::plan(future.callr::callr, workers = N_CORES)

    # --- Run Simulations in Parallel ---
    results_non <- furrr::future_map_dfr(
        .x = 1:N_SIMULATIONS,
        .f = ~ run_simulation_iteration(
            iter_num = .x,
            full_data = df,
            main_formula = main_formula,
            cumulative_arg = cumulative_arg
        ),
        .options = furrr_options(seed = TRUE, packages = "rms"),
        .progress = TRUE,
        .id = "iter" # Automatically adds an iteration column
    )

    results_non_boot <- furrr::future_map_dfr(
        .x = 1:N_SIMULATIONS,
        .f = ~ run_simulation_iteration_boot(
            iter_num = .x,
            full_data = df,
            main_formula = main_formula,
            cumulative_arg = cumulative_arg
        ),
        .options = furrr_options(seed = TRUE, packages = "rms"),
        .progress = TRUE,
        .id = "iter" # Automatically adds an iteration column
    )

    write_parquet(results, here(OUTPUT_PATH, RESULTS_FILE_NO))
    write_parquet(results_non_boot, here(OUTPUT_PATH, RESULTS_FILE_NO_BOOT))
}
```

### Sim with linear Interaction

```{r}
RESULTS_FILE_LIN <- "sim-resampling-H0-linear-int-freq.parquet"

if(simulate){
    main_formula <- ordered(y)  ~ tx + rcs(time, 4) + yprev + time %ia% yprev
    cumulative_arg <- cumulative(reverse = TRUE, parallel = FALSE ~  rcs(time, 4))

    # --- Run Simulations in Parallel ---
    results_int <- furrr::future_map_dfr(
        .x = 1:N_SIMULATIONS,
        .f = ~ run_simulation_iteration(
            iter_num = .x,
            full_data = df,
            main_formula = main_formula,
            cumulative_arg = cumulative_arg
        ),
        .options = furrr_options(seed = TRUE, packages = "rms"),
        .progress = TRUE,
        .id = "iter" # Automatically adds an iteration column
    )

    write_parquet(results_int, here(OUTPUT_PATH, RESULTS_FILE_LIN))
}
```

### Sim Second Order, no interaction

```{r}
RESULTS_FILE_SECOND <- "sim-resampling-H0-second-order.parquet"
if(simulate){
    main_formula <- ordered(y)  ~ tx + rcs(time, 4) + yprev + ypprev
    cumulative_arg <- cumulative(reverse = TRUE, parallel = FALSE ~  rcs(time, 4))


    # --- Run Simulations in Parallel ---
    results_second <- furrr::future_map_dfr(
        .x = 1:N_SIMULATIONS,
        .f = ~ run_simulation_iteration(
            iter_num = .x,
            full_data = df_ypprev,
            main_formula = main_formula,
            cumulative_arg = cumulative_arg
        ),
        .options = furrr_options(seed = TRUE, packages = "rms"),
        .progress = TRUE,
        .id = "iter"
    )

    write_parquet(results_second, here(OUTPUT_PATH, RESULTS_FILE_SECOND))
}
```

## Type I Error Rate

### RE-Randomizing, no replacement
```{r}
# Load the simulation results
results_non <- read_parquet(here(OUTPUT_PATH, RESULTS_FILE_NO))
results_lin <- read_parquet(here(OUTPUT_PATH, RESULTS_FILE_LIN))
results_second <- read_parquet(here(OUTPUT_PATH, RESULTS_FILE_SECOND))

# --- Calculations for Non-linear Simulation ---
p_values_non <- results_non |> filter(term == "tx1") |>
    pull(`Pr...z..`)
type_I_error_non <- mean(p_values_non < 0.05)
mcse_non <- jackknife_mcse(p_values_non < 0.05)

# --- Calculations for Linear Simulation ---
p_values_lin <- results_lin |> filter(term == "tx1") |>
    pull(`Pr...z..`)
type_I_error_lin <- mean(p_values_lin < 0.05)
mcse_lin <- jackknife_mcse(p_values_lin < 0.05)

# --- Calculations for Second-order Simulation ---
p_values_second <- results_second |> filter(term == "tx1") |>
    pull(`Pr...z..`)
type_I_error_second <- mean(p_values_second < 0.05)
mcse_second <- jackknife_mcse(p_values_second < 0.05)

# --- Create and format the results table ---
results_summary <- tibble(
  Simulation = c("Non-linear", "Linear", "Second-order"),
  `Type I Error (MCSE)` = c(
    paste0(sprintf("%.3f", type_I_error_non), " (", sprintf("%.3f", mcse_non), ")"),
    paste0(sprintf("%.3f", type_I_error_lin), " (", sprintf("%.3f", mcse_lin), ")"),
    paste0(sprintf("%.3f", type_I_error_second), " (", sprintf("%.3f", mcse_second), ")")
  )
)

knitr::kable(results_summary, caption = "Type I Error Rates with Monte Carlo Standard Errors (MCSE)")

```

### With Bootstrapping, 500 N

```{r}
results_non_boot <- read_parquet(here(OUTPUT_PATH, RESULTS_FILE_NO_BOOT))

p_values_non_boot <- results_non_boot |> filter(term == "tx1") |> 
    pull(`Pr...z..`)

mean(p_values_non_boot < 0.05)

jackknife_mcse(p_values_non_boot < 0.05)
```