---
Title: Examining Type I Error based on Sim Data
---

## Setup
```{r}
# Load Libraries
library(ggplot2)
library(arrow)
library(tidyverse)
library(furrr)
library(glue)
library(fs)
library(here)
library(rms)
library(VGAM)
```

```{r}
source(here("functions", "jackknife_mcse.R"))
source(here("functions", "check-completed-simulations.R"))
source(here("functions", "bayes-hosp-run-markov-simulation-iteration.R"))
```

## Load data

- underlying truths: 
- first order markov
- time follows third order polynomial
- treatment effect on y = 0
- yprev interacts fully with time

```{r}
set.seed(1234)

# Load full dataset
# underlying bayesian model 
# y  ~ tx + pol(week, 3) + yprev + ecog_fstcnt + diagnosis + age + gender + albumin + c_reactive_protein,

df_full <- read_parquet(here("2-hosp", "sim-data", "sim_data_no_interaction.parquet")) |> 
    mutate(
        y = as.numeric(as.character(y)),
        y = if_else(y == 4, 3, y),
        y = if_else(y == 5, 4, y),
        yprev = if_else(yprev == "4", "3", yprev),
        yprev = factor(yprev)) |> 
    filter(tx == "0") |> 
    select(id, tx, y, yprev, time)


# model used for df_full_no_int (vglm model)

# vm <- vglm(
#   ordered(y) ~ tx + week + I(week^2) + I(week^3) + yprev + ecog_fstcnt + diagnosis,
#   cumulative(reverse = TRUE, parallel = FALSE ~ week + I(week^2) + week + I(week^3)),
#   data = df
#   )

# states 3 and 4 were already collapsed during model fitting
df_full_no_int <- read_parquet(here("2-hosp", "sim-data", "sim_data_vglm.parquet")) |> 
    select(id, tx, y, yprev, time)
```

## Simulation function

```{r}
run_simulation_iteration <- function(iter_num, base_path, full_data, model_formula, cumulative_arg, seed = 123, sample_size = 270) {
    # This function runs one full simulation iteration.
    
    # --- 1. Setup for the current iteration ---
    set.seed(seed + iter_num)


    tx_ids <- full_data |> 
        distinct(id) |> 
        pull(id) |> sample(round(0.5* sample_size), replace = FALSE)

    soc_ids <- full_data |> 
        distinct(id) |> 
        pull(id) |> 
        setdiff(tx_ids) |> 
        sample(round(0.5* sample_size), replace = FALSE)

    data_for_model <- full_data |> 
        filter(id %in% c(tx_ids, soc_ids)) |>
        mutate(
            tx = if_else(id %in% tx_ids, "1", "0"),
            tx = factor(tx, levels = c(0, 1)))
    
    rm(full_data)
    
    if (nrow(data_for_model) > 0) {
        model <- tryCatch({ 
            vglm(
                model_formula,
                cumulative_arg,
                data = data_for_model)
            
        }, error = function(e) {
            message(glue("ERROR in iteration {iter_num}: {e$message}"))
            
            return(NULL)
        })
        
        # --- 7. Save the results ONLY if the model ran successfully ---
        if (!is.null(model)) {
            summary_df <- summary(model)@coef3 |> data.frame() |> 
                rownames_to_column("term")

            return(summary_df)
        }
    }
    return(NULL)
}
```

## Type I Error for data w/o interaction

### Large N
```{r}
N_SIMULATIONS <- 1000
OUTPUT_PATH <- here("2-hosp", "output", "sim-simdata-H0-freq")
RESULTS_FILE_LARGE <- "sim-simdata-H0-no-int-1k.parquet"
N_CORES <- 8

# define model specs
main_formula <- ordered(y) ~ tx + poly(time, 3) + yprev
cumulative_formula <- cumulative(reverse = TRUE, parallel = FALSE ~ poly(time, 3))

future::plan(future.callr::callr, workers = N_CORES)

# --- Run Simulations in Parallel ---
results <- furrr::future_map_dfr(
    .x = 1:N_SIMULATIONS,
    .f = ~ run_simulation_iteration(
        iter_num = .x,
        full_data = df_full_no_int,
        model_formula = main_formula,
        cumulative_arg = cumulative_formula,
        sample_size = 1000,
        seed = 888
    ),
    .options = furrr_options(seed = TRUE),
    .progress = TRUE,
    .id = "iter" # Automatically adds an iteration column
)

write_parquet(results, here(OUTPUT_PATH, RESULTS_FILE))
```

### Small N

```{r}
RESULTS_FILE_SMALL <- "sim-simdata-H0-no-int-250.parquet"

results <- furrr::future_map_dfr(
    .x = 1:N_SIMULATIONS,
    .f = ~ run_simulation_iteration(
        iter_num = .x,
        full_data = df_full_no_int,
        model_formula = main_formula,
        cumulative_arg = cumulative_formula,
        sample_size = 250,
        seed = 888
    ),
    .options = furrr_options(seed = TRUE),
    .progress = TRUE,
    .id = "iter" # Automatically adds an iteration column
)

write_parquet(results, here(OUTPUT_PATH, RESULTS_FILE_SMALL))
```


```{r}
# Read and stack all draw files into a single data frame
results_l_n <- read_parquet(here(OUTPUT_PATH, RESULTS_FILE_LARGE))
results_s_n <- read_parquet(here(OUTPUT_PATH, RESULTS_FILE_SMALL))

p_values_l <- results_l_n |> filter(term == "tx1") |> 
    pull(`Pr...z..`)

p_values_s <- results_s_n |> filter(term == "tx1") |> 
    pull(`Pr...z..`)

hist(p_values_l, breaks = 200, main = "Large N (1000)")
hist(p_values_s, breaks = 200, main = "Small N (250)")

mean(p_values_l < 0.05)
jackknife_mcse(p_values_l < 0.05)
mean(p_values_s < 0.05)
jackknife_mcse(p_values_s < 0.05)
```

## Bayesian Simulation

Function to run a single iteration

```{r}
N_SIMULATIONS <- 500
N_CORES <- 28
future::plan(future.callr::callr, workers = N_CORES)
```

## Small n (250)

The model is correclty specified, i.e., correctly models time (uses rcs(time, 4)) instead of pol(time, 3) which was used during data generation but that really shouldn't matter, and correctly does not include an interaction. It just does not include some predictors that were used to generate the data. The only thing that's not 100% as in the data generating model, is that levels 3 and 4 have been collapsed into one level. 

```{r}
OUTPUT_PATH <- here("2-hosp", "output", "sim-simdata-H0-no-int-bayes-250-correct")
sim_status <- check_completed_simulations(OUTPUT_PATH, N_SIMULATIONS)

if (sim_status$n_to_run > 0) {
    message(glue("Starting {sim_status$n_to_run} new simulations on {N_CORES} cores..."))
    
    formula <- y  ~ tx + yprev + rcs(time, 4)
    
    # --- Run Simulations in Parallel ---
    furrr::future_walk(
        .x = sim_status$iters_to_run,
        .f = ~run_simulation_iteration(
            iter_num = .x,
            base_path = OUTPUT_PATH,
            full_data = df_full,
            model_formula = formula,
            sample_size = 250
        ),
        .options = furrr_options(seed = TRUE, packages = c("rmsb", "data.table", "tidyverse")),
        .progress = TRUE
    )
    
    message(glue("Finished {sim_status$n_to_run} simulations."))
} else {
    message("All simulations were already complete. No new runs started.")
}
```

