
## Count Regression

### Resampling H0

#### Load draws + diagnostics
```{r}
# H0 no interaction
path <- "./2-hosp/output/sim-resampling-H0-neg-binom/"
neg_binom_h0 <- load_draws(path)
rhat_neg_binom_h0 <- load_rhat_values(path = path)

ate_neg_binom_h0 <- load_draws(path = path, glob = "*/ATE_draws.rds")
```

#### Convergence diagnostics
```{r}
rhat_plot_neg_binom_h0 <- rhat_neg_binom_h0 |> 
    plot_rhat(y = b_tx1)

# ggsave(filename = "./img/fig-rhat-markov.svg", width = 10, height = 5)
```

### Plot probability of benefit

```{r}
p1 <- neg_binom_h0 |> 
    plot_p_benefit(param = b_tx1)

p2 <- ate_neg_binom_h0 |>
    plot_p_benefit(param = draw)

p1 + p2

# ggsave(filename = "./img/fig-pbenefit-markov.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_neg_binom_h0 <- neg_binom_h0 |> 
  plot_posterior_distribution(x_variable = b_tx1) + 
  scale_x_continuous(limits = c(-1, 1))

post_neg_binom_h0

ate_post_neg_binom_h0 <- ate_neg_binom_h0 |> 
  plot_posterior_distribution(x_variable = draw)

# ggsave(filename = "./img/fig-posteriors-markov.svg", width = 10, height = 5)
```

```{r}
summary_neg_binom_h0 <- neg_binom_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx1, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()
```

## DRS 26

Not actually 26 categories, but 1 to max(number of days not at home) + 1 worse for people who died.

### H0

#### Load draws + diagnostics
```{r}
# H0 rmsb (default dirichlet prior)
path <- "./2-hosp/output/sim-drs26-h0/"
drs26_h0 <- load_draws(path) |> 
    mutate(`tx=1` = -`tx=1`) # change sign to match interpretation of other models
rhat_drs26_h0 <- load_rhat_values(path = path)
```

#### Convergence diagnostics
```{r}
rhat_plot_drs26_h0 <- rhat_drs26_h0 |> 
    plot_rhat(y = `tx=1`)
rhat_plot_drs26_h0
# ggsave(filename = "./img/fig-rhat-markov.svg", width = 10, height = 5)
```

#### Plot probability of benefit

```{r}
p1 <- drs26_h0 |> 
    plot_p_benefit(param = `tx=1`)

p1

# ggsave(filename = "./img/fig-pbenefit-markov.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_drs26_rmsb_h0 <- drs26_h0 |> 
  plot_posterior_distribution(x_variable = `tx=1`)

# ggsave(filename = "./img/fig-posteriors-markov.svg", width = 10, height = 5)
```

```{r}
summary_drs26_h0 <- drs26_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = `tx=1`, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()
```

### HA - Markov OR = 0.8

#### Load draws + diagnostics
```{r}
# H0 rmsb (default dirichlet prior)
path <- "./2-hosp/output/sim-drs26-HA-08/"
drs26_ha_08 <- load_draws(path) |> 
    mutate(tx = -tx) # change sign to match logodds <0 = benefit
rhat_drs26_ha_08 <- load_rhat_values(path = path)
```

#### Convergence diagnostics
```{r}
rhat_plot_drs26_ha_08 <- rhat_drs26_ha_08 |> 
    plot_rhat(y = tx)
rhat_plot_drs26_ha_08
# ggsave(filename = "./img/fig-rhat-markov.svg", width = 10, height = 5)
```

#### Plot probability of benefit

```{r}
# average p(benefit) over iterations
drs26_ha_08 |> 
    group_by(iter) |> 
    summarise(
        p_benefit = mean(tx < 0)
    ) |> 
    summarise(
        mean_p_benefit = mean(p_benefit),
        sd_p_benefit = sd(p_benefit),
        n = n()
    )

p1 <- drs26_ha_08 |> 
    plot_p_benefit(param = tx)

p1

# ggsave(filename = "./img/fig-pbenefit-markov.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_drs26_ha_08 <- drs26_ha_08 |> 
  plot_posterior_distribution(x_variable = tx)

post_drs26_ha_07 <- drs26_ha_07 |> 
  plot_posterior_distribution(x_variable = tx)


# ggsave(filename = "./img/fig-posteriors-markov.svg", width = 10, height = 5)
```

```{r}
summary_drs26_ha_08 <- drs26_ha_08 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = tx, 
        true_param_value = log(1), #true param unclear
        threshold = 0.95) |> 
    summarise_iterations()

summary_drs26_ha_08
```
