---
title: "Simulation Study: Handling Differential Missingness in Markov QoL Models"
format:
  typst:
    papersize: a4
    margin:
      x: 2cm
      y: 2.5cm
    toc: true
    columns: 1
    tbl-cap-location: bottom
---

## Problem Statement

In longitudinal QoL studies with death as an absorbing state, we face **differential missingness**:

- **Death (state 8)**: Fully observed at exact time of occurrence
- **QoL states (1-7)**: Sparsely observed (~15-20% of timepoints)

This creates biased state occupancy probability (SOP) estimates because the cumulative incidence of death is incorrectly estimated when unobserved QoL times are treated as missing.

**Approach:** Multiple Imputation (MI) to impute plausible QoL values constrained to 1-7.

**Key insight**: We cannot use `Ocens()` for this problem because intervals are only valid for the outcome at each timepoint, but we also need valid `yprev` values for the Markov model. Imputation addresses both issues.

## Setup

```{r}
#| output: false
#| label: setup
library(here)
library(tidyverse)
library(arrow)
library(patchwork)
library(markov.misc)
library(data.table)
library(Hmisc)
library(rms)
library(VGAM)
library(mice)
library(miceadds)
library(furrr)
library(future.callr)

set.seed(12345)

# Configuration
N_PATIENTS <- 270
N_IMPUTATIONS <- 5
# N_BOOTSTRAP <- 200
N_SIM_ITERATIONS <- 32
SAMPLE_PERCENTAGE <- 0.20
FU <- 26
N_WORKERS <- 8
```

## Data Generating Process

### Fit Model to Historical Data

```{r}
#| label: fit-dgp-model
data <- read_parquet(here("data", "qol_historic.parquet")) |>
  mutate(
    yprev = factor(yprev, ordered = FALSE),
    ecog_fstcnt = factor(ecog_fstcnt)
  )

# Fit proportional odds model (basis for DGP)
m_dgp <- vglm(
  ordered(y) ~ pol(week, 3) + yprev + ecog_fstcnt + diagnosis,
  family = cumulative(reverse = TRUE, parallel = TRUE),
  data = data
)

# Extract intercepts (add death state intercept)
ints <- rev(c(coef(m_dgp)[1:6], -5))
extra_params <- coef(m_dgp)[-c(1:6)]
```

### Define Linear Predictor Function

```{r}
#| label: lp-function
g <- function(yprev, t, tx, ecog_fstcnt, diagnosis, parameter, extra_params) {
  tx_effect <- parameter * as.numeric(as.character(tx))
  time_effect <- extra_params["pol(week, 3)week"] * t +
    extra_params["pol(week, 3)week^2"] * t^2 +
    extra_params["pol(week, 3)week^3"] * t^3

  yprev_effect <- rep(0, length(yprev))
  for (state in unique(yprev)) {
    if (state != 1) {
      state_idx <- which(yprev == state)
      yprev_coef_name <- paste0("yprev", state)
      if (yprev_coef_name %in% names(extra_params)) {
        yprev_effect[state_idx] <- extra_params[[yprev_coef_name]]
      }
    }
  }

  diagnosis_effect <- rep(0, length(diagnosis))
  for (diag in unique(diagnosis)) {
    if (diag != levels(diagnosis)[1]) {
      diag_idx <- which(diagnosis == diag)
      diag_coef_name <- paste0("diagnosis", diag)
      if (diag_coef_name %in% names(extra_params)) {
        diagnosis_effect[diag_idx] <- extra_params[[diag_coef_name]]
      }
    }
  }

  ecog_effect <- ifelse(ecog_fstcnt == 2, extra_params["ecog_fstcnt2"], 0)
  as.numeric(tx_effect + time_effect + yprev_effect + diagnosis_effect + ecog_effect)
}
```

### Create Baseline Data

```{r}
#| label: baseline-data
baseline_data <- data |>
  select(pat_id, baseline_q30, ecog_fstcnt, diagnosis, tx) |>
  distinct() |>
  mutate(
    yprev = factor(baseline_q30, ordered = TRUE),
    week = 1
  ) |>
  slice_sample(n = N_PATIENTS, replace = TRUE) |>
  mutate(id = row_number()) |>
  select(-pat_id, -week)
```

## Core Functions

### 1. Simulate Complete Trajectories

```{r}
#| label: simulate-trajectories
simulate_complete_data <- function(baseline_data, tx_effect = log(1), seed = NULL) {
  sim_trajectories_markov(
    baseline_data = baseline_data,
    follow_up_time = FU,
    intercepts = ints,
    lp_function = g,
    extra_params = extra_params,
    parameter = tx_effect,
    absorbing_states = 8,
    seed = seed,
    covariate_names = c("tx", "ecog_fstcnt", "diagnosis")
  )
}
```

### 2. Apply Differential Missingness

```{r}
#| label: apply-missingness
apply_differential_missingness <- function(df, sample_pct = SAMPLE_PERCENTAGE) {
  # Death is ALWAYS observed
  death_obs <- df |> filter(y == 8)

  # Baseline (time = 1) always observed
  baseline <- df |>
    filter(time == 1) |>
    mutate(observed = TRUE)

  # Sample QoL observations (sparse)
  qol_obs <- df |>
    filter(y != 8, time > 1) |>
    mutate(observed = runif(n()) < sample_pct)

  # Combine: keep structure but mark missingness
  bind_rows(baseline, qol_obs, death_obs |> mutate(observed = TRUE)) |>
    arrange(id, time) |>
    mutate(
      y_observed = if_else(observed, y, NA_real_),
      y_true = y  # Keep true value for validation
    )
}
```

### 3. Multiple Imputation for QoL

The key insight: we need to impute **both** `y` (outcome) and reconstruct `yprev` (lagged state) from the imputed trajectory. We also need to preserve the baseline observation at time=1 with its original yprev.

```{r}
#| label: impute-qol
impute_qol_trajectory <- function(df, m_imp = N_IMPUTATIONS) {
  # Convert to data.table for speed
  dt <- as.data.table(df)

  # 1. Extract death times efficiently
  death_info <- dt[y == 8, .(death_time = min(time)), by = id]

  # 2. Get baseline covariates (first row per patient)
  baseline_covars <- dt[time == 1, .(id, baseline_q30, ecog_fstcnt, diagnosis, tx)]

  # 3. Create full grid efficiently using data.table
  ids <- unique(dt$id)
  full_grid <- CJ(id = ids, time = 1:FU)
  full_grid <- death_info[full_grid, on = "id"]
  full_grid <- baseline_covars[full_grid, on = "id"]

  # Mark censored and death rows
  full_grid[, `:=`(
    censored = !is.na(death_time) & time > death_time,
    is_death_week = !is.na(death_time) & time == death_time
  )]
  full_grid <- full_grid[censored == FALSE]

  # 4. Merge observations
  obs_dt <- dt[, .(id, time, y_observed)]
  imp_prep <- obs_dt[full_grid, on = c("id", "time")]

  # Define target and proxies
  imp_prep[, `:=`(
    y_target = fifelse(is_death_week, 8, y_observed),
    death_time_safe = fifelse(is.na(death_time), FU + 1L, death_time)
  )]

  # LOCF for proxy (within each patient)
  imp_prep[, y_locf := y_observed]
  imp_prep[time == 1 & is.na(y_locf), y_locf := baseline_q30]
  setorder(imp_prep, id, time)
  imp_prep[, y_locf := nafill(y_locf, type = "locf"), by = id]
  imp_prep[, y_prev_proxy := shift(y_locf, n = 1, type = "lag"), by = id]
  imp_prep[time == 1, y_prev_proxy := baseline_q30]

  # 5. Prepare MICE data (exclude death weeks)
  mice_df <- imp_prep[is_death_week == FALSE, .(
    id, time,
    y = y_target,
    y_prev_proxy,
    death_time = death_time_safe,
    baseline_q30, ecog_fstcnt, diagnosis, tx
  )]
  mice_df[, `:=`(
    ecog_fstcnt = as.numeric(ecog_fstcnt),
    diagnosis = as.numeric(diagnosis)
  )]
  mice_df <- as.data.frame(mice_df)

  # 6. Configure MICE
  init <- mice(mice_df, maxit = 0, printFlag = FALSE)
  pred <- init$predictorMatrix
  meth <- init$method
  meth[] <- ""
  meth["y"] <- "2l.pmm"
  pred[, "id"] <- -2

  # 7. Impute (single call, most expensive step)
  imp <- mice(
    mice_df,
    m = m_imp,
    maxit = 10,
    method = meth,
    predictorMatrix = pred,
    printFlag = FALSE
  )

  # 8. Prepare death rows once
  death_rows <- as.data.frame(imp_prep[is_death_week == TRUE, .(
    id, time, y = y_target, baseline_q30, ecog_fstcnt, diagnosis, tx
  )])

  # 9. Process imputations efficiently
  process_imputation <- function(i) {
    completed_qol <- complete(imp, i)
    completed_qol$ecog_fstcnt <- factor(completed_qol$ecog_fstcnt)
    completed_qol$diagnosis <- factor(completed_qol$diagnosis)
    completed_qol <- completed_qol[, c("id", "time", "y", "baseline_q30",
                                        "ecog_fstcnt", "diagnosis", "tx")]

    full_traj <- rbind(completed_qol, death_rows)
    full_traj <- full_traj[order(full_traj$id, full_traj$time), ]

    # Use data.table for fast lag computation
    dt_traj <- as.data.table(full_traj)
    dt_traj[, `:=`(
      yprev = shift(y, n = 1, type = "lag", fill = first(baseline_q30)),
      timeprev = shift(time, n = 1, type = "lag", fill = 0L)
    ), by = id]
    dt_traj[, `:=`(
      yprev = factor(yprev, levels = 1:7, ordered = FALSE),
      gap = time - timeprev,
      imp_id = i
    )]

    as_tibble(dt_traj[!is.na(y)])
  }

  bind_rows(lapply(1:m_imp, process_imputation))
}
```

### 4. Estimation via vglm

For a single dataset (complete case or one imputation), fit the model and compute SOPs.

```{r}
#| label: fit-vglm-model
fit_vglm_single <- function(df, ylevels = factor(1:8), absorb = 8) {
  # Filter out transitions from death (can't predict from absorbing state)
  model_df <- df |>
    filter(as.numeric(as.character(yprev)) != 8) |>
    mutate(
      yprev = factor(yprev, levels = 1:7, ordered = FALSE),
      y = as.numeric(y),
      tx = as.numeric(as.character(tx))
    )

  # Fit proportional odds model using RCS in formula
  m <- vglm(
    y ~ tx + rms::rcs(time, 4) + yprev + ecog_fstcnt + diagnosis,
    data = model_df,
    family = cumulative(reverse = TRUE, parallel = TRUE)
  )

  # Compute standardized SOPs
  varnames <- list(tvarname = "time", pvarname = "yprev", id = "id", tx = "tx", gap = "gap")

  sops <- standardize_sops(
    model = m,
    data = model_df,
    times = 1:FU,
    ylevels = ylevels,
    absorb = absorb,
    varnames = varnames
  )

  list(model = m, sops = sops)
}

#' Fit model to MI data: fit separately to each imputation, then average SOPs
fit_vglm_mi <- function(mi_df, ylevels = factor(1:8), absorb = 8) {
  imp_ids <- unique(mi_df$imp_id)

  # Fit model to each imputed dataset
  results <- lapply(imp_ids, function(i) {
    fit_vglm_single(mi_df |> filter(imp_id == i), ylevels, absorb)
  })

  # Average SOPs across imputations (Rubin's rules for point estimates)
  sop_tx_list <- lapply(results, function(r) r$sops$sop_tx)
  sop_ctrl_list <- lapply(results, function(r) r$sops$sop_ctrl)

  avg_sop_tx <- Reduce(`+`, sop_tx_list) / length(sop_tx_list)
  avg_sop_ctrl <- Reduce(`+`, sop_ctrl_list) / length(sop_ctrl_list)

  list(
    sops = list(sop_tx = avg_sop_tx, sop_ctrl = avg_sop_ctrl),
    n_imputations = length(results)
  )
}

#' Extract time-in-state estimates from SOPs
extract_time_in_state <- function(sops) {
  if (is.null(sops)) return(NULL)

  # Time in each state = sum of SOPs across all timepoints
  time_tx <- colSums(sops$sop_tx)
  time_ctrl <- colSums(sops$sop_ctrl)

  tibble(
    state = 1:length(time_tx),
    time_tx = time_tx,
    time_ctrl = time_ctrl,
    diff = time_tx - time_ctrl
  )
}
```


### 5. Compute True Time-in-State from Complete Data

```{r}
#| label: true-time-in-state
compute_true_time_in_state <- function(complete_df) {
  # Empirical SOPs from complete data
  sops_by_tx <- complete_df |>
    group_by(tx, time) |>
    count(y) |>
    mutate(sop = n / sum(n)) |>
    ungroup()

  # Time in each state
  time_in_state <- sops_by_tx |>
    group_by(tx, y) |>
    summarise(time_in_state = sum(sop), .groups = "drop") |>
    pivot_wider(names_from = tx, values_from = time_in_state, names_prefix = "tx_") |>
    rename(state = y) |>
    mutate(
      time_tx = tx_1,
      time_ctrl = tx_0,
      diff = time_tx - time_ctrl
    ) |>
    select(state, time_tx, time_ctrl, diff)

  time_in_state
}
```

## Simulation Framework

### Single Iteration

```{r}
#| label: single-iteration
run_single_iteration <- function(iter, tx_effect, seed) {
  set.seed(seed + iter)

  # 1. Generate complete data
  complete_df <- simulate_complete_data(baseline_data, tx_effect = tx_effect, seed = seed + iter)

  # 2. Compute true time in each state (from complete data)
  true_time <- compute_true_time_in_state(complete_df)

  # 3. Apply differential missingness
  df_with_miss <- apply_differential_missingness(complete_df)

  # 4. Multiple Imputation
  mi_df <- impute_qol_trajectory(df_with_miss, m_imp = N_IMPUTATIONS)
  mi_fit <- fit_vglm_mi(mi_df)
  mi_time <- extract_time_in_state(mi_fit$sops)

  # Compile results (long format by state)
  tibble(
    iter = iter,
    state = 1:8,
    true_tx = true_time$time_tx,
    true_ctrl = true_time$time_ctrl,
    true_diff = true_time$diff,
    mi_tx = mi_time$time_tx,
    mi_ctrl = mi_time$time_ctrl,
    mi_diff = mi_time$diff
  )
}
```

### Run Full Simulation

```{r}
#| label: run-simulation
run_simulation <- function(n_iter, tx_effect, scenario_name, seed = 12345) {
  message(glue::glue("Running {scenario_name}: {n_iter} iterations"))

  plan(callr, workers = N_WORKERS)

  results <- future_map_dfr(1:n_iter, function(i) {
    run_single_iteration(i, tx_effect, seed)
  }, .options = furrr_options(
    seed = TRUE,
    packages = c("tidyverse", "data.table", "mice", "miceadds", "VGAM", "markov.misc")))

  plan(sequential)

  results |> mutate(scenario = scenario_name, tx_effect = tx_effect)
}
```

## Demonstration

### Generate Example Data

```{r}
#| label: demo-data
# H0: No treatment effect
d_h0 <- simulate_complete_data(baseline_data, tx_effect = log(1), seed = 1234)

# Visualize complete trajectories
p1 <- plot_sops(d_h0, time_var = "time", geom = "line", linetype_var = "tx", facet_var = NULL) +
  coord_cartesian(ylim = c(0, 0.35)) +
  labs(title = "Complete Data (H0)")

p1

# Apply missingness
d_miss <- apply_differential_missingness(d_h0)

# Summary of missingness
miss_summary <- d_miss |>
  group_by(y_true == 8) |>
  summarise(
    n_total = n(),
    n_observed = sum(observed),
    pct_observed = mean(observed) * 100
  )

miss_summary
```

### Compare Methods

```{r}
#| label: compare-methods
# True values from complete data
true_time <- compute_true_time_in_state(d_h0)

# MI approach (fit to each imputation separately, then average)
mi_data <- impute_qol_trajectory(d_miss, m_imp = 20)

mi_fit <- fit_vglm_mi(mi_data)
mi_time <- extract_time_in_state(mi_fit$sops)
```

### Bias Comparison (All States)

```{r}
#| label: bias-comparison
# Combine results
comparison <- true_time |>
  rename(true_tx = time_tx, true_ctrl = time_ctrl, true_diff = diff) |>
  left_join(
    mi_time |> rename(mi_tx = time_tx, mi_ctrl = time_ctrl, mi_diff = diff),
    by = "state"
  ) |>
  mutate(
    mi_bias_tx = mi_tx - true_tx,
    mi_bias_ctrl = mi_ctrl - true_ctrl,
    mi_bias_diff = mi_diff - true_diff
  )

# Print comparison
comparison |>
  select(state, true_tx, mi_tx, mi_bias_tx) |>
  knitr::kable(digits = 3, caption = "Time in State (Treatment Arm)")

comparison |>
  select(state, true_ctrl, mi_ctrl, mi_bias_ctrl) |>
  knitr::kable(digits = 3, caption = "Time in State (Control Arm)")

comparison |>
  select(state, true_diff, mi_diff, mi_bias_diff) |>
  knitr::kable(digits = 3, caption = "Treatment Effect (Difference in Weeks)")
```

### Visualize SOPs

```{r}
#| label: viz-sops
#| fig-width: 12
#| fig-height: 8

# Prepare SOP data for plotting
prepare_sop_plot_data <- function(sops, method) {
  bind_rows(
    as_tibble(sops$sop_tx) |>
      mutate(time = 1:FU, tx = "Treatment") |>
      pivot_longer(-c(time, tx), names_to = "state", values_to = "sop"),
    as_tibble(sops$sop_ctrl) |>
      mutate(time = 1:FU, tx = "Control") |>
      pivot_longer(-c(time, tx), names_to = "state", values_to = "sop")
  ) |>
    mutate(method = method)
}

# Prepare empirical SOPs from complete data
prepare_empirical_sops <- function(complete_df) {
  complete_df |>
    group_by(tx, time) |>
    count(y) |>
    mutate(sop = n / sum(n)) |>
    ungroup() |>
    mutate(
      state = as.character(y),
      tx = if_else(tx == 1, "Treatment", "Control"),
      method = "True (Complete Data)"
    ) |>
    select(time, tx, state, sop, method)
}

if (exists("mi_fit") && exists("d_h0") && !is.null(mi_fit)) {
  sop_data <- bind_rows(
    prepare_empirical_sops(d_h0),
    prepare_sop_plot_data(mi_fit$sops, "Multiple Imputation")
  ) |>
    mutate(method = factor(method, levels = c("True (Complete Data)", "Multiple Imputation")))

  ggplot(sop_data, aes(x = time, y = sop, color = state, linetype = tx)) +
    geom_line(linewidth = 0.8) +
    facet_wrap(~method, nrow = 1) +
    scale_color_brewer(palette = "Dark2") +
    labs(
      title = "State Occupancy Probabilities: True vs. Estimated",
      x = "Study Week", y = "Probability",
      color = "State", linetype = "Treatment"
    ) +
    theme_minimal()
}
```

## Run Simulation Study

```{r}
#| label: full-simulation
#| eval: false

# Define scenarios
scenarios <- list(
  H0 = log(1)      # Null: no effect
  # HA1 = log(0.8),   # Small effect
  # HA2 = log(0.6)    # Moderate effect
)

# Run simulations (this takes time!)
all_results <- map2_dfr(
  scenarios, names(scenarios),
  ~run_simulation(n_iter = N_SIM_ITERATIONS, tx_effect = .x, scenario_name = .y)
)

# Save results
write_parquet(all_results, here("1-qol", "output", "diff_miss_sim_results.parquet"))
```

## Analyze Results

```{r}
#| label: analyze-results
#| eval: false

results <- read_parquet(here("1-qol", "output", "diff_miss_sim_results.parquet"))

# Bias by state and scenario
bias_summary <- results |>
  mutate(
    mi_bias_tx = mi_tx - true_tx,
    mi_bias_ctrl = mi_ctrl - true_ctrl,
    mi_bias_diff = mi_diff - true_diff
  ) |>
  group_by(scenario, state) |>
  summarise(
    true_mean_diff = mean(true_diff, na.rm = TRUE),
    mi_mean_bias = mean(mi_bias_diff, na.rm = TRUE),
    mi_rmse = sqrt(mean(mi_bias_diff^2, na.rm = TRUE)),
    .groups = "drop"
  )

bias_summary

# Visualization: Bias in differences by state
results |>
  mutate(mi_bias = mi_diff - true_diff) |>
  ggplot(aes(x = factor(state), y = mi_bias)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~scenario) +
  labs(
    title = "Bias in Time-in-State Difference by State (Multiple Imputation)",
    x = "State", y = "Bias (Estimated - True)"
  ) +
  theme_minimal()
```

### Combined Time-in-State Visualization

```{r}
#| label: combined-time-in-state-viz
#| eval: false
#| fig-width: 12
#| fig-height: 8

# Create long format for combined visualization
combined_bias <- results |>
  mutate(
    mi_bias_tx = mi_tx - true_tx,
    mi_bias_ctrl = mi_ctrl - true_ctrl
  ) |>
  pivot_longer(
    cols = c(mi_bias_tx, mi_bias_ctrl),
    names_to = "arm",
    values_to = "bias"
  ) |>
  mutate(arm = if_else(arm == "mi_bias_tx", "Treatment", "Control"))

# Faceted boxplot
ggplot(combined_bias, aes(x = factor(state), y = bias)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_grid(arm ~ scenario) +
  labs(
    title = "MI Bias in Time-in-State Estimates by Treatment Arm and Scenario",
    x = "State", y = "Bias (Weeks)"
  ) +
  theme_minimal()
```
