---
title: "QoL unadjusted landmark — descriptive summaries"
subtitle: "Observed, imputed, and unadjusted model-derived state distributions"
format: html
execute:
  freeze: false
---

This stage applies the primary QoL descriptive check to the unadjusted
supporting analysis. It uses the same prepared data and primary completed
datasets, restricted to imputations 1–5. Missing outcomes remain an explicit
grey category in the raw-data panel. The model-derived bars use the median of
the stacked posterior state-probability draws across the five imputations.

```{r}
#| label: describe-supporting-qol-unadjusted
#| include: false

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
prepared <- arrow::read_parquet(here::here(
  "artifacts", "primary", "quality-of-life", "prepared.parquet"
))
imputations <- readRDS(here::here(
  "artifacts", "primary", "quality-of-life", "imputations.rds"
))
indices <- config$imputation$supporting$indices
model_dir <- here::here(
  "artifacts", "supporting", "qol-unadjusted-landmark", "models"
)
result_dir <- here::here(
  "results", "supporting", "qol-unadjusted-landmark", "descriptive"
)
figure_dir <- here::here(
  "artifacts", "supporting", "qol-unadjusted-landmark", "figures"
)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

state_levels <- as.character(1:8)
display_levels <- c(state_levels, "Missing")

raw_summary <- prepared |>
  dplyr::group_by(tx) |>
  dplyr::summarise(
    n = dplyr::n(),
    deaths_by_day_182 = sum(event_death == 1L, na.rm = TRUE),
    baseline_qol_observed = sum(!is.na(baseline_q30)),
    baseline_qol_missing = sum(is.na(baseline_q30)),
    landmark_qol_observed = sum(!is.na(landmark_q30)),
    landmark_qol_missing = sum(is.na(landmark_q30)),
    composite_observed = sum(!is.na(y_landmark)),
    composite_missing = sum(is.na(y_landmark)),
    .groups = "drop"
  )

raw_counts <- prepared |>
  dplyr::mutate(
    state = dplyr::if_else(
      is.na(y_landmark),
      "Missing",
      as.character(y_landmark)
    ),
    state = factor(state, levels = display_levels)
  ) |>
  dplyr::count(tx, state, .drop = FALSE, name = "n") |>
  dplyr::group_by(tx) |>
  dplyr::mutate(
    proportion = n / sum(n),
    source = "Observed raw data"
  ) |>
  dplyr::ungroup()

imputed_probabilities <- purrr::map_dfr(indices, function(index) {
  imputations[[index]] |>
    dplyr::mutate(state = factor(y_landmark, levels = 1:8)) |>
    dplyr::count(tx, state, .drop = FALSE, name = "n") |>
    dplyr::group_by(tx) |>
    dplyr::mutate(
      imputation = index,
      proportion = n / sum(n)
    ) |>
    dplyr::ungroup() |>
    dplyr::select(imputation, tx, state, proportion)
})

imputed_summary <- imputed_probabilities |>
  dplyr::group_by(tx, state) |>
  dplyr::summarise(proportion = mean(proportion), .groups = "drop") |>
  dplyr::mutate(
    n = NA_integer_,
    source = "Mean across completed datasets"
  )

model_draws <- purrr::map_dfr(indices, function(index) {
  model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", index)))
  data <- imputations[[index]]
  set.seed(analysis_seed(config, "supporting", index, offset = 700L))
  selected <- sample.int(
    nrow(model$draws),
    config$imputation$supporting$post_estimation_draws,
    replace = FALSE
  )
  model$draws <- model$draws[selected, , drop = FALSE]

  purrr::map_dfr(0:1, function(treatment) {
    newdata <- data |>
      dplyr::mutate(tx = treatment)
    probabilities <- predict(
      model,
      newdata = newdata,
      type = "fitted.ind",
      posterior.summary = "all"
    )
    standardized <- apply(probabilities, c(1, 3), mean)
    tidyr::expand_grid(
      draw = seq_len(nrow(standardized)),
      state = state_levels
    ) |>
      dplyr::mutate(
        imputation = index,
        tx = treatment,
        probability = as.vector(t(standardized))
      )
  })
})

model_summary <- model_draws |>
  dplyr::group_by(tx, state) |>
  dplyr::summarise(
    proportion = median(probability),
    q025 = quantile(probability, 0.025, names = FALSE),
    q975 = quantile(probability, 0.975, names = FALSE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    n = NA_integer_,
    source = "Median unadjusted proportional-odds prediction"
  )

plot_data <- dplyr::bind_rows(
  raw_counts |>
    dplyr::select(tx, state, n, proportion, source),
  imputed_summary |>
    dplyr::select(tx, state, n, proportion, source),
  model_summary |>
    dplyr::select(tx, state, n, proportion, source)
) |>
  dplyr::mutate(
    state = factor(state, levels = display_levels),
    tx = factor(tx, levels = 0:1, labels = c("Control", "Intervention")),
    source = factor(
      source,
      levels = c(
        "Observed raw data",
        "Mean across completed datasets",
        "Median unadjusted proportional-odds prediction"
      )
    )
  )

state_colours <- c(
  stats::setNames(grDevices::hcl.colors(8L, "viridis"), state_levels),
  Missing = "grey75"
)
state_plot <- ggplot2::ggplot(
  plot_data,
  ggplot2::aes(x = tx, y = proportion, fill = state)
) +
  ggplot2::geom_col(width = 0.72) +
  ggplot2::facet_wrap(ggplot2::vars(source), nrow = 1L) +
  ggplot2::scale_fill_manual(values = state_colours, drop = FALSE) +
  ggplot2::scale_y_continuous(labels = scales::label_percent()) +
  ggplot2::labs(
    x = NULL,
    y = "Proportion",
    fill = "QoL/death state"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.grid.major.x = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
  )

arrow::write_parquet(raw_summary, file.path(result_dir, "raw-summary.parquet"))
arrow::write_parquet(raw_counts, file.path(result_dir, "raw-state-counts.parquet"))
arrow::write_parquet(
  imputed_probabilities,
  file.path(result_dir, "imputed-state-probabilities.parquet")
)
arrow::write_parquet(
  model_summary,
  file.path(result_dir, "model-state-probabilities.parquet")
)
ggplot2::ggsave(
  file.path(figure_dir, "observed-imputed-model-state-distributions.png"),
  state_plot,
  width = 12,
  height = 5,
  dpi = 300
)
```

```{r}
#| label: tbl-supporting-qol-unadjusted-raw-summary
#| tbl-cap: "Availability of the raw QoL variables by randomized group"
#| echo: false

knitr::kable(raw_summary)
```

```{r}
#| label: fig-supporting-qol-unadjusted-state-distributions
#| fig-cap: "Observed, imputed, and unadjusted model-derived distributions of the six-month QoL/death state"
#| fig-alt: "Three faceted stacked bar charts compare QoL and death states 1 through 8 by treatment group. The raw-data panel also includes a grey missing category."
#| fig-width: 12
#| fig-height: 5
#| echo: false

state_plot
```
