---
title: "QoL unadjusted landmark — proportional-odds check"
subtitle: "Common and threshold-specific unadjusted treatment odds ratios"
format: html
execute:
  freeze: false
---

The proportional-odds assumption is examined in the same way as for the main
QoL analysis, but every model contains treatment only. The common treatment
log-odds ratio is compared with treatment effects from seven binary logistic
models. For threshold (k), the binary event is (Y_{182} > k), so positive
values indicate greater odds of a worse outcome under intervention. Estimates
and variances are pooled across primary imputations 1–5 using Rubin's rules.

This is a frequentist graphical model check, not a replacement for the
Bayesian supporting estimand.

```{r}
#| label: check-supporting-qol-unadjusted-proportional-odds
#| include: false

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
imputations <- readRDS(here::here(
  "artifacts", "primary", "quality-of-life", "imputations.rds"
))
indices <- config$imputation$supporting$indices
thresholds <- 1:7

estimates <- purrr::map_dfr(indices, function(index) {
  data <- imputations[[index]] |>
    dplyr::mutate(y_landmark = ordered(y_landmark, levels = 1:8))

  common_fit <- rms::lrm(
    y_landmark ~ tx,
    data = data,
    x = TRUE,
    y = TRUE
  )
  common_estimate <- tibble::tibble(
    imputation = index,
    threshold = NA_integer_,
    comparison = "Common proportional-odds effect",
    estimate = unname(stats::coef(common_fit)["tx"]),
    variance = unname(stats::vcov(common_fit)["tx", "tx"])
  )

  binary_estimates <- purrr::map_dfr(thresholds, function(threshold) {
    binary_data <- data |>
      dplyr::mutate(
        binary_worse = as.integer(as.integer(y_landmark) > threshold)
      )
    binary_fit <- rms::lrm(
      binary_worse ~ tx,
      data = binary_data,
      x = TRUE,
      y = TRUE
    )
    tibble::tibble(
      imputation = index,
      threshold = threshold,
      comparison = paste0("Y > ", threshold),
      estimate = unname(stats::coef(binary_fit)["tx"]),
      variance = unname(stats::vcov(binary_fit)["tx", "tx"])
    )
  })

  dplyr::bind_rows(common_estimate, binary_estimates)
})

pooled <- estimates |>
  dplyr::group_by(comparison) |>
  dplyr::group_modify(function(data, key) {
    m <- nrow(data)
    q_bar <- mean(data$estimate)
    u_bar <- mean(data$variance)
    between <- stats::var(data$estimate)
    total <- u_bar + (1 + 1 / m) * between
    degrees_freedom <- if (is.finite(between) && between > 0) {
      (m - 1) * (1 + u_bar / ((1 + 1 / m) * between))^2
    } else {
      Inf
    }
    critical_value <- if (is.finite(degrees_freedom)) {
      stats::qt(0.975, df = degrees_freedom)
    } else {
      stats::qnorm(0.975)
    }
    standard_error <- sqrt(total)
    tibble::tibble(
      threshold = data$threshold[[1L]],
      log_odds_ratio = q_bar,
      standard_error = standard_error,
      conf_low = q_bar - critical_value * standard_error,
      conf_high = q_bar + critical_value * standard_error,
      odds_ratio = exp(q_bar),
      odds_ratio_low = exp(q_bar - critical_value * standard_error),
      odds_ratio_high = exp(q_bar + critical_value * standard_error),
      imputations = m
    )
  }) |>
  dplyr::ungroup() |>
  dplyr::mutate(order = dplyr::coalesce(threshold, 0L)) |>
  dplyr::arrange(order)

binary_pooled <- pooled |>
  dplyr::filter(!is.na(threshold)) |>
  dplyr::mutate(comparison = factor(comparison, levels = rev(comparison)))
common_pooled <- pooled |>
  dplyr::filter(is.na(threshold))
common_repeated <- binary_pooled |>
  dplyr::mutate(
    log_odds_ratio = common_pooled$log_odds_ratio[[1L]],
    conf_low = common_pooled$conf_low[[1L]],
    conf_high = common_pooled$conf_high[[1L]]
  )

comparison_plot <- ggplot2::ggplot() +
  ggplot2::geom_vline(
    xintercept = common_pooled$log_odds_ratio[[1L]],
    colour = "grey65",
    linewidth = 0.5
  ) +
  ggplot2::geom_errorbarh(
    data = binary_pooled,
    ggplot2::aes(y = comparison, xmin = conf_low, xmax = conf_high),
    height = 0.12,
    position = ggplot2::position_nudge(y = 0.08),
    colour = "black"
  ) +
  ggplot2::geom_point(
    data = binary_pooled,
    ggplot2::aes(x = log_odds_ratio, y = comparison),
    position = ggplot2::position_nudge(y = 0.08),
    size = 2.5,
    colour = "black"
  ) +
  ggplot2::geom_errorbarh(
    data = common_repeated,
    ggplot2::aes(y = comparison, xmin = conf_low, xmax = conf_high),
    height = 0.12,
    position = ggplot2::position_nudge(y = -0.08),
    colour = "steelblue"
  ) +
  ggplot2::geom_point(
    data = common_repeated,
    ggplot2::aes(x = log_odds_ratio, y = comparison),
    position = ggplot2::position_nudge(y = -0.08),
    size = 2.5,
    colour = "steelblue"
  ) +
  ggplot2::labs(
    x = "Unadjusted treatment log odds ratio for a worse outcome",
    y = "Binary threshold"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(panel.grid.major.y = ggplot2::element_blank())

result_dir <- here::here(
  "results", "supporting", "qol-unadjusted-landmark",
  "proportional-odds-check"
)
figure_dir <- here::here(
  "artifacts", "supporting", "qol-unadjusted-landmark", "figures"
)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
arrow::write_parquet(
  estimates,
  file.path(result_dir, "imputation-estimates.parquet")
)
arrow::write_parquet(
  pooled,
  file.path(result_dir, "pooled-estimates.parquet")
)
ggplot2::ggsave(
  file.path(figure_dir, "common-vs-binary-log-odds-ratios.png"),
  comparison_plot,
  width = 8,
  height = 5.5,
  dpi = 300
)
```

```{r}
#| label: tbl-supporting-qol-unadjusted-proportional-odds-check
#| tbl-cap: "Pooled common and threshold-specific unadjusted treatment odds ratios for a worse six-month outcome"
#| echo: false

knitr::kable(pooled[c(
  "comparison", "odds_ratio", "odds_ratio_low", "odds_ratio_high",
  "log_odds_ratio", "conf_low", "conf_high"
)])
```

```{r}
#| label: fig-supporting-qol-unadjusted-proportional-odds-check
#| fig-cap: "Common and threshold-specific unadjusted treatment log odds ratios pooled across five completed datasets"
#| fig-alt: "A horizontal dot-and-interval plot compares the common unadjusted proportional-odds treatment coefficient with seven binary logistic treatment coefficients for progressively worse QoL thresholds."
#| fig-width: 8
#| fig-height: 5.5
#| echo: false

comparison_plot
```
