---
title: "Multiple Imputation Time Alive and Out of Hospital"
---

## Setup

```{r}
#| output: false
library(here)
library(tidyverse)
library(mice)
library(miceadds)
library(arrow)
library(data.table)
```

## Data

```{r}
data_wide <- read_parquet(here("data", "analysis-data", "liqplat_analysis_data_wide.parquet")) |>
  mutate(
    diagnosis = fct_lump_n(diagnosis, n = 4),
    diagnosis = factor(as.integer(diagnosis)),
    # category is very rare, lump wth category 3
    ecog_fstcnt = if_else(
        ecog_fstcnt == "4", "3", ecog_fstcnt
    ),
    ecog_fstcnt = factor(ecog_fstcnt, levels = 0:3, ordered = TRUE),
    lactate_dehydrogenase = log(lactate_dehydrogenase + 1)
    )

data_wide$na_est <- nelsonaalen(data_wide, timevar = surv_time, statusvar = status_death)

data_wide <- data_wide |>
    select(id, diagnosis, ecog_fstcnt, albumin, bilirubin, c_reactive_protein, lactate_dehydrogenase, na_est)


data_raw <- read_parquet(here("data", "analysis-data", "liqplat_analysis_data_long.parquet")) |>
    mutate(
        death = if_else(y_taooh == 5, 1, 0),
        tx = factor(tx),
        week = ceiling(day / 7),
        surv_time = ceiling(surv_time / 7),
    ) |>
    group_by(id, week) |>
    mutate(
        death = factor(max(death)),
        in_hospital = factor(max(in_hospital)),
        unplanned_stay = factor(max(unplanned_stay)),
        er_visit = factor(max(er_visit))) |>
    ungroup() |>
    filter(week > -2) |>
    select(-day, -y_qol, -y_taooh) |>
    distinct() |>
    arrange(id, week) |>
    select(-c(diagnosis, ecog_fstcnt)) |>
    left_join(data_wide, by = "id")
```

## Data Preprocessing

### Variable Selection and Transformations

```{r}
imputation_vars <- c(
  "id",
  "tx",
  "week",
  "gender",
  "age",
  "ecog_fstcnt",
  "diagnosis",
  "albumin",
  "bilirubin",
  "c_reactive_protein",
  "lactate_dehydrogenase",
  "in_hospital",
  "unplanned_stay",
  "er_visit",
  "death",
  "status_death",
  "na_est"
)

df <- data_raw |>
  select(all_of(imputation_vars))

# Model non-linear time effects using restricted cubic splines with 4 knots
# Time is never missing, so we can create the spline basis directly
basis_rcs <- rms::rcs(df$week, 4)

df$week <- as.vector(basis_rcs[, 1])
df$week_nl1 <- as.vector(basis_rcs[, 2])
df$week_nl2 <- as.vector(basis_rcs[, 3])
```

## Multiple Imputation Procedure {#sec-imputation}

### Predictor Matrix Configuration

The predictor matrix is configured to:

- Use `id` as the clustering variable (coded as $-2$)
- Exclude death-related variables from predictions (we use `na_est` instead)
- Prevent time variables from predicting static baseline variables

```{r}
#| output: false
md.pattern(df)

init <- mice(df, maxit = 0)
pred_matrix <- init$predictorMatrix
meth <- init$method

# Use patient ID as clustering variable
pred_matrix[, "id"] <- -2
pred_matrix["id", ] <- 0

# Exclude raw survival variables from predictions (we use na_est instead)
# This avoids aggregation artefacts when collapsing longitudinal data to cluster means
# status_death = whether some died (we use this in combination with na_est)
# death == 1 in the week someone died. We don't use this.
pred_matrix[, "death"] <- 0
pred_matrix["death", ] <- 0

# week shouldn't predict static baseline variables nor itself (nor na_est)
baseline_vars <- c("tx", "age", "gender", "diagnosis", "ecog_fstcnt", "albumin", "bilirubin",
                   "c_reactive_protein", "lactate_dehydrogenase", "na_est", "status_death")

pred_matrix[c(baseline_vars, "week", "week_nl1", "week_nl2"), c("week", "week_nl1", "week_nl2")] <- 0

# treatment, gender, age, and week are never missing
meth["tx"] <- ""
meth["gender"] <- ""
meth["age"] <- ""
meth["week"] <- ""
meth["week_nl1"] <- ""
meth["week_nl2"] <- ""


# use mean matching at the cluster level for baseline observations
# this ensures only one ecog per patient per imputation
meth["diagnosis"] <- "2lonly.pmm"
meth["ecog_fstcnt"] <- "2lonly.pmm"
meth["albumin"] <- "2lonly.pmm"
meth["bilirubin"] <- "2lonly.pmm"
meth["c_reactive_protein"] <- "2lonly.pmm"
meth["lactate_dehydrogenase"] <- "2lonly.pmm"

# Longitudinal variables: multilevel PMM
meth["in_hospital"] <- "2l.pmm"
meth["unplanned_stay"] <- "2l.pmm"
meth["er_visit"] <- "2l.pmm"


imputed_data <- mice(df,
                     m = 5,       # 50 in final analysis
                     maxit = 50,   # 50 in final analysis
                     predictorMatrix = pred_matrix,
                     method = meth,
                     donors = 5,
                     seed = 1234)
```

### Imputation Parameters

For the final trial analysis:

- Number of imputed datasets: $m = 50$
- Maximum iterations: 50
- Donor pool size: $k = 5$

## Convergence Diagnostics {#sec-diagnostics}

Convergence is assessed via trace plots of the mean and standard deviation of imputed values across iterations.

```{r trace-plots}
#| fig-cap: "Trace plots showing the mean and standard deviation of imputed values across iterations for each imputed dataset."
plot(imputed_data)
```

## Distribution Diagnostics {#sec-distribution}

We compare the distributions of observed and imputed values to assess plausibility.

```{r}
#| fig-cap: "Density plots comparing observed (blue) and imputed (red) values for all imputed variables."
densityplot(imputed_data)
```

```{r}
#| fig-cap: "Density plot for ECOG performance status."
densityplot(imputed_data, ~ecog_fstcnt)
```

```{r}
# set.seed(1234)
# original_data <- imputed_data$data
# missing_pat_ids <- unique(original_data$pat_id[is.na(original_data$in_hospital)])
# sample_ids <- sample(missing_pat_ids, 9)

# xyplot(imputed_data,
#        in_hospital ~ week | factor(pat_id),
#        subset = pat_id %in% sample_ids,
#        pch = c(1, 20),
#        type = c("p"),
#        lwd = 0,
#        layout = c(3, 3),
#        main = "Hospitalisation: Observed vs Imputed")

# xyplot(imputed_data,
#        unplanned_stay ~ week | factor(pat_id),
#        subset = pat_id %in% sample_ids,
#        pch = c(1, 20),
#        type = c("p"),
#        lwd = 0,
#        layout = c(3, 3),
#        main = "Unplanned Stay: Observed vs Imputed")

# xyplot(imputed_data,
#        er_visit ~ week | factor(pat_id),
#        subset = pat_id %in% sample_ids,
#        pch = c(1, 20),
#        type = c("p"),
#        lwd = 0,
#        layout = c(3, 3),
#        main = "ER Visit: Observed vs Imputed")
```

## Export

```{r}
for(i in 1:imputed_data$m){
    complete_data <- complete(imputed_data, i)

    # format for QoL analysis
    complete_data <- complete_data |>
      mutate(
      y = case_when(
        death == 1           ~ 5,
        unplanned_stay == 1  ~ 4,
        er_visit == 1        ~ 3,
        in_hospital == 1     ~ 2,
        TRUE                 ~ 1
      )) |>
      group_by(id) |>
      mutate(
      yprev = lag(y, n = 1, order_by = week),
      ypprev = lag(yprev, n = 1, order_by = week),
      yprev = factor(yprev, levels = 1:4, ordered = FALSE),
      ypprev = factor(ypprev, levels = 1:4, ordered = FALSE)) |>
    ungroup() |>
    filter(week > 0) |>
    relocate(id, tx, y, yprev, ypprev, week)

    write_parquet(complete_data, sink = here("0-main-analysis", "TAOOH", "imp-data", paste0("taooh-data-", i, ".parquet")))
}
```
