---
title: "TAOOH — primary imputation"
subtitle: "Patient-level baseline-covariate imputation"
params:
  horizon_days: 182
format: html
execute:
  freeze: false
---

TAOOH has no intermittent missing states. The observed weekly outcome is
therefore not imputed. Instead, incomplete baseline covariates are imputed in a
one-row-per-patient dataset using observed TAOOH occupancy proportions and
survival information as auxiliary predictors. Each completed baseline dataset
is then joined back to the unchanged, already lagged weekly history.

```{r}
#| label: impute-taooh
#| include: false
source(here::here("R", "taooh-helpers.R"))
horizon_days <- params$horizon_days
horizon_weeks <- taooh_horizon_weeks(horizon_days)

library(tidyverse)
library(mice)
source(here::here("R", "analysis-config.R"))


config <- read_analysis_config()
indices <- seq_len(config$imputation$main$m)
input_path <- taooh_path("artifacts", horizon_days, "prepared.parquet")
branch_dir <- taooh_path("artifacts", horizon_days, "imputations")
dir.create(branch_dir, recursive = TRUE, showWarnings = FALSE)

prepared <- arrow::read_parquet(input_path) |>
  as.data.frame() |>
  mutate(
    diagnosis = factor(diagnosis),
    ecog_fstcnt = factor(ecog_fstcnt, levels = 0:3, ordered = TRUE)
  )

taooh_summaries <- prepared |>
  group_by(id) |>
  summarise(
    n_alive = sum(y %in% 1:4, na.rm = TRUE),
    no_alive_history = as.integer(n_alive == 0L),
    prop_planned_admission = if_else(
      n_alive > 0,
      sum(y == 2L, na.rm = TRUE) / n_alive,
      0
    ),
    prop_er = if_else(
      n_alive > 0,
      sum(y == 3L, na.rm = TRUE) / n_alive,
      0
    ),
    prop_unplanned_admission = if_else(
      n_alive > 0,
      sum(y == 4L, na.rm = TRUE) / n_alive,
      0
    )
  )

transition_summaries <- prepared |> 
  group_by(id) |>
  summarise(
    n_from_home = sum(
      yprev == 1L & y %in% 1:4,
      na.rm = TRUE
    ),
    n_acute_from_home = sum(
      yprev == 1L & y %in% c(3L, 4L),
      na.rm = TRUE
    ),
    prop_acute_from_home = if_else(
      n_from_home > 0,
      n_acute_from_home / n_from_home,
      0
    ),
    .groups = "drop"
  )

patient_data <- prepared |>
  select(
    id, tx, age, gender, diagnosis, ecog_fstcnt, albumin, bilirubin,
    c_reactive_protein, lactate_dehydrogenase, survival_time_days,
    event_death
  ) |>
  distinct() |>
  left_join(taooh_summaries, by = "id") |>
  left_join(transition_summaries, by = "id") |> 
  mutate(
    tx = factor(tx),
    gender = factor(gender),
    event_death = as.integer(event_death)
  )

patient_data$na_est <- mice::nelsonaalen(
  patient_data,
  survival_time_days,
  event_death
)

patient_ids <- patient_data$id
imputation_data <- patient_data |>
  select(-survival_time_days, -n_acute_from_home, -n_alive)

mice::md.pattern(imputation_data)

methods <- mice::make.method(imputation_data)
methods[] <- ""
methods["ecog_fstcnt"] <- "polr"
methods[c(
  "albumin", "bilirubin", "c_reactive_protein",
  "lactate_dehydrogenase"
)] <- "pmm"

predictors <- mice::make.predictorMatrix(imputation_data)
predictors[, "id"] <- 0

imputation_fit <- mice::mice(
  data = imputation_data,
  method = methods,
  predictorMatrix = predictors,
  m = config$imputation$main$m,
  maxit = config$imputation$main$maxit,
  donors = config$imputation$main$donors,
  seed = analysis_seed(config, "imputation", 1L),
  printFlag = FALSE
)

saveRDS(
  imputation_fit,
  file.path(branch_dir, "imputation-fit.rds"),
  compress = "xz"
)

longitudinal_data <- prepared |>
  select(id, week, y, yprev, ypprev)

for (index in indices) {
  output_path <- file.path(branch_dir, sprintf("imputation-%02d.parquet", index))

  completed_baseline <- mice::complete(imputation_fit, index) |>
    select(
      id, tx, age, gender, diagnosis, ecog_fstcnt, albumin, bilirubin,
      c_reactive_protein, lactate_dehydrogenase
    )

  completed <- longitudinal_data |>
    left_join(completed_baseline, by = "id") |>
    mutate(
      yprev = factor(yprev, levels = 1:4),
      ypprev = factor(ypprev, levels = 1:4),
      id = factor(id)
    )

  arrow::write_parquet(completed, output_path, compression = "zstd")
}
```

```{r}
#| label: taooh-imputation-convergence
#| echo: false
#| fig-cap: "MICE convergence traces for imputed TAOOH baseline covariates."

plot(imputation_fit)
```

```{r}
#| label: taooh-imputation-distributions
#| echo: false
#| fig-cap: "Observed and imputed distributions of TAOOH baseline covariates."

densityplot(
  imputation_fit,
  ~ ecog_fstcnt + albumin + bilirubin +
    c_reactive_protein + lactate_dehydrogenase
)
```
