---
title: "TAOOH — primary diagnostics"
subtitle: "Full proportional-odds second-order diagnostics for 50 imputations"
params:
  horizon_days: 182
format: html
execute:
  freeze: false
---

```{r}
#| label: check-taooh-models
#| include: false
source(here::here("R", "taooh-helpers.R"))
horizon_days <- params$horizon_days
horizon_weeks <- taooh_horizon_weeks(horizon_days)

library(dplyr)
source(here::here("R", "analysis-config.R"))
source(here::here("R", "markov-helpers.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
indices <- seq_len(config$imputation$main$m)
taooh_diagnostic_thresholds <- config$diagnostics
taooh_diagnostic_thresholds$max_divergent_transitions <- 19L
model_dir <- taooh_path("artifacts", horizon_days, "models")
diagnostic_dir <- taooh_path("artifacts", horizon_days, "diagnostics")
stan_dx_dir <- file.path(diagnostic_dir, "stan-dx")
figure_dir <- taooh_path("artifacts", horizon_days, "figures")
result_path <- taooh_path("results", horizon_days, "diagnostics.parquet")
dir.create(diagnostic_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(stan_dx_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(result_path), recursive = TRUE, showWarnings = FALSE)

for (index in indices) {
  model_path <- file.path(model_dir, sprintf("model-%02d.rds", index))
  output_path <- file.path(
    diagnostic_dir,
    sprintf("diagnostics-%02d.parquet", index)
  )
  if (!force && taooh_outputs_current(output_path, model_path)) next

  diagnostics <- markov_diagnostics(
    readRDS(model_path),
    taooh_diagnostic_thresholds
  )
  assert_markov_diagnostics(diagnostics)
  output <- as.data.frame(diagnostics$summary)
  output$imputation <- index
  output$divergent_transitions <- diagnostics$divergent_transitions
  output$treedepth_exceeded <- diagnostics$treedepth_exceeded
  output$min_bfmi <- diagnostics$min_bfmi
  arrow::write_parquet(output, output_path)
}

diagnostic_paths <- file.path(
  diagnostic_dir,
  sprintf("diagnostics-%02d.parquet", indices)
)
diagnostics <- lapply(diagnostic_paths, function(path) {
  arrow::read_parquet(path) |>
    as.data.frame()
})
arrow::write_parquet(do.call(rbind, diagnostics), result_path)
```

The full `rmsb` Stan diagnostic report and `stanDxplot()` output are saved for
every fitted model. This replaces separate `mcmc_trace()` figures and keeps
divergences, treedepth, E-BFMI, R-hat, effective sample sizes, and chain
behavior together for review by imputation.

```{r}
#| label: save-taooh-stan-diagnostics
#| include: false

for (index in indices) {
  stan_dx_text_path <- file.path(
    stan_dx_dir,
    sprintf("stan-dx-%02d.txt", index)
  )
  stan_dx_plot_path <- file.path(
    stan_dx_dir,
    sprintf("stan-dx-%02d.png", index)
  )
  if (
    !force &&
      taooh_outputs_current(
        c(stan_dx_text_path, stan_dx_plot_path),
        file.path(model_dir, sprintf("model-%02d.rds", index))
      )
  ) next

  model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", index)))
  stan_dx_text <- capture.output(rmsb::stanDx(model))
  writeLines(stan_dx_text, stan_dx_text_path)

  stan_dx_plot <- rmsb::stanDxplot(model)
  ggplot2::ggsave(
    stan_dx_plot_path,
    stan_dx_plot,
    width = 11,
    height = max(6, 1.05 * length(unique(stan_dx_plot$data$param))),
    dpi = 180,
    limitsize = FALSE
  )
}
```

The empirical graphical checks use the observed weekly dataset with death
carried forward through the requested horizon. Model-derived checks use the first completed
dataset and its fitted model as a representative imputation-specific
diagnostic. SOPs, transition proportions, correlations, and variograms are
kept as separate files so they can be compared directly. Correlation plots are
used for TAOOH because its dense weekly observations make the empirical
time-by-time correlation structure estimable.

```{r}
#| label: derive-taooh-graphical-diagnostics
#| include: false

diagnostic_index <- indices[[1L]]
data <- arrow::read_parquet(file.path(
  taooh_path("artifacts", horizon_days, "imputations"),
  sprintf("imputation-%02d.parquet", diagnostic_index)
)) |>
  as.data.frame()
data$y <- factor(data$y, levels = 1:5, ordered = TRUE)
data$yprev <- factor(data$yprev, levels = 1:5)
data$ypprev <- factor(data$ypprev, levels = 1:5)

cohort <- arrow::read_parquet(here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
)) |>
  as.data.frame()
empirical_data <- arrow::read_parquet(here::here(
  "analysis", "00-data", "results", "analysis-taooh.parquet"
)) |>
  as.data.frame() |>
  taooh_empirical_history(horizon_days) |>
  left_join(cohort |> select(id, tx), by = "id") |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(
    y = as.integer(y_taooh),
    yprev = lag(y)
  ) |>
  ungroup() |>
  filter(between(week, 1L, horizon_weeks)) |>
  mutate(
    y = factor(y, levels = 1:5, ordered = TRUE),
    yprev = factor(yprev, levels = 1:5),
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Intervention"))
  )

model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", diagnostic_index)))
sops <- arrow::read_parquet(file.path(
  taooh_path("artifacts", horizon_days, "estimand"),
  sprintf("sops-draws-%02d.parquet", diagnostic_index)
))
diagnostic_times <- unique(as.integer(round(seq(1, horizon_weeks, length.out = 9))))
diagnostic_seed <- analysis_seed(config, "diagnostics", diagnostic_index)

empirical_sop_plot <- mostr::plot_sops(
  empirical_data,
  time_var = "week",
  y_var = "y",
  facet_var = NULL,
  linetype_var = "tx",
  geom = "line"
) +
  ggplot2::scale_color_brewer(palette = "Dark2")

empirical_transition_plot <- mostr::plot_transitions(
  empirical_data,
  times = diagnostic_times,
  time_var = "week",
  y_var = "y",
  p_var = "yprev",
  y_levels = 1:5,
  # facet_var = "tx",
  show_values = TRUE
)
empirical_transition_plot$layers[[2]]$aes_params$colour <- "white"

model_transition_plot <- mostr::plot_transitions(
  model,
  times = diagnostic_times,
  y_levels = 1:5,
  absorb = 5L,
  time_var = "week",
  p_var = "yprev",
  p2_var = "ypprev",
  id_var = "id",
  # facet_var = "tx",
  seed = diagnostic_seed,
  n_draws = 200L,
  show_values = TRUE
)

model_transition_plot$layers[[2]]$aes_params$colour <- "white"

empirical_correlation_plot <- mostr::plot_correlation(
  empirical_data,
  times = seq_len(horizon_weeks),
  y_levels = 1:5,
  id_var = "id",
  time_var = "week",
  y_var = "y",
  # facet_var = "tx",
  show_values = TRUE
)

model_correlation_plot <- mostr::plot_correlation(
  model,
  times = seq_len(horizon_weeks),
  y_levels = 1:5,
  absorb = 5L,
  id_var = "id",
  time_var = "week",
  p_var = "yprev",
  p2_var = "ypprev",
  # facet_var = "tx",
  seed = diagnostic_seed,
  n_draws = 50L,
  show_values = TRUE
)
empirical_variogram_plot <- mostr::plot_variogram(
  empirical_data,
  times = seq_len(horizon_weeks),
  y_levels = 1:5,
  id_var = "id",
  time_var = "week",
  y_var = "y",
  # facet_var = "tx"
)
model_variogram_plot <- mostr::plot_variogram(
  model,
  times = seq_len(horizon_weeks),
  y_levels = 1:5,
  absorb = 5L,
  id_var = "id",
  time_var = "week",
  p_var = "yprev",
  p2_var = "ypprev",
  # facet_var = "tx",
  seed = diagnostic_seed,
  n_draws = 50L
)

diagnostic_plots <- list(
  "empirical-sops.png" = empirical_sop_plot,
  # "model-sops.png" = model_sop_plot,
  "empirical-transitions.png" = empirical_transition_plot,
  "model-transitions.png" = model_transition_plot,
  "empirical-correlation.png" = empirical_correlation_plot,
  "model-correlation.png" = model_correlation_plot,
  "empirical-variogram.png" = empirical_variogram_plot,
  "model-variogram.png" = model_variogram_plot
)
for (filename in names(diagnostic_plots)) {
  ggplot2::ggsave(
    file.path(figure_dir, filename),
    diagnostic_plots[[filename]],
    width = if (grepl("transition", filename)) 12 else 8,
    height = if (grepl("transition", filename)) 9 else 5.5,
    dpi = 300,
    limitsize = FALSE
  )
}
```

```{r}
#| label: fig-taooh-sop-check
#| fig-cap: "Empirical and model-derived TAOOH state occupancy probabilities in the first completed dataset"
#| fig-subcap:
#|   - "Empirical state occupancy probabilities"
#|   - "Model-derived state occupancy probabilities"
#| layout-ncol: 1
#| fig-width: 8
#| fig-height: 5.5
#| echo: false

empirical_sop_plot
# model_sop_plot
```

```{r}
#| label: fig-taooh-correlation-check
#| fig-cap: "Empirical and second-order model-derived TAOOH correlation structures in the first completed dataset"
#| fig-subcap:
#|   - "Empirical correlations"
#|   - "Model-derived correlations"
#| layout-ncol: 2
#| fig-width: 8
#| fig-height: 5.5
#| echo: false

empirical_correlation_plot
model_correlation_plot
```
