---
title: "TAOOH — primary diagnostics"
subtitle: "Second-order Markov diagnostics for 50 completed datasets"
format: html
execute:
  freeze: false
---

```{r}
#| label: check-taooh-models
#| include: false
library(dplyr)
source(here::here("R", "analysis-config.R"))
source(here::here("R", "markov-helpers.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
indices <- seq_len(config$imputation$main$m)[1:2]
taooh_diagnostic_thresholds <- config$diagnostics
taooh_diagnostic_thresholds$max_divergent_transitions <- 19L
model_dir <- here::here("artifacts", "primary", "taooh", "models")
diagnostic_dir <- here::here("artifacts", "primary", "taooh", "diagnostics")
trace_dir <- file.path(diagnostic_dir, "trace-plots")
stan_dx_dir <- file.path(diagnostic_dir, "stan-dx")
figure_dir <- here::here("artifacts", "primary", "taooh", "figures")
result_path <- here::here("results", "primary", "taooh", "diagnostics.parquet")
dir.create(diagnostic_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(trace_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(stan_dx_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(result_path), recursive = TRUE, showWarnings = FALSE)

for (index in indices) {
  model_path <- file.path(model_dir, sprintf("model-%02d.rds", index))
  output_path <- file.path(
    diagnostic_dir,
    sprintf("diagnostics-%02d.parquet", index)
  )
  if (!force && file.exists(output_path)) next

  diagnostics <- markov_diagnostics(
    readRDS(model_path),
    taooh_diagnostic_thresholds
  )
  assert_markov_diagnostics(diagnostics)
  output <- as.data.frame(diagnostics$summary)
  output$imputation <- index
  output$divergent_transitions <- diagnostics$divergent_transitions
  output$treedepth_exceeded <- diagnostics$treedepth_exceeded
  output$min_bfmi <- diagnostics$min_bfmi
  arrow::write_parquet(output, output_path)
}

diagnostic_paths <- file.path(
  diagnostic_dir,
  sprintf("diagnostics-%02d.parquet", indices)
)
diagnostics <- lapply(diagnostic_paths, function(path) {
  arrow::read_parquet(path) |>
    as.data.frame()
})
arrow::write_parquet(do.call(rbind, diagnostics), result_path)
```

One trace plot is saved for every fitted model. The trace selection uses the
indexed raw Stan parameter names produced by `rmsb`, retaining ordinal
intercepts, regression coefficients, partial-proportional-odds coefficients,
and random-effect scale parameters while excluding pointwise log likelihoods.
Numerical diagnostics continue to cover the full posterior.

```{r}
#| label: save-taooh-trace-plots
#| include: false

for (index in indices) {
  trace_path <- file.path(trace_dir, sprintf("trace-%02d.png", index))
  if (!force && file.exists(trace_path)) next

  model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", index)))
  draws <- posterior::as_draws_array(rmsb::stanGet(model))
  trace_variables <- posterior::variables(draws)
  trace_variables <- trace_variables[grepl(
    "^(alpha|beta|tau|sigma|sd)(\\[|$)",
    trace_variables
  )]
  if (length(trace_variables) == 0L) {
    stop("No TAOOH alpha, beta, tau, or scale parameters were found for trace plotting.")
  }
  trace_plot <- bayesplot::mcmc_trace(draws, pars = trace_variables)
  ggplot2::ggsave(
    trace_path,
    trace_plot,
    width = 11,
    height = max(6, 1.05 * length(trace_variables)),
    dpi = 180,
    limitsize = FALSE
  )
}
```

The full `rmsb` Stan diagnostic report and diagnostic plot are saved for every
fitted model, allowing divergences, treedepth, E-BFMI, R-hat, effective sample
sizes, and chain behavior to be reviewed by imputation.

```{r}
#| label: save-taooh-stan-diagnostics
#| include: false

for (index in indices) {
  stan_dx_text_path <- file.path(
    stan_dx_dir,
    sprintf("stan-dx-%02d.txt", index)
  )
  stan_dx_plot_path <- file.path(
    stan_dx_dir,
    sprintf("stan-dx-%02d.png", index)
  )
  if (
    !force &&
      file.exists(stan_dx_text_path) &&
      file.exists(stan_dx_plot_path)
  ) next

  model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", index)))
  stan_dx_text <- capture.output(rmsb::stanDx(model))
  writeLines(stan_dx_text, stan_dx_text_path)

  stan_dx_plot <- rmsb::stanDxplot(model)
  ggplot2::ggsave(
    stan_dx_plot_path,
    stan_dx_plot,
    width = 11,
    height = max(6, 1.05 * length(unique(stan_dx_plot$data$param))),
    dpi = 180,
    limitsize = FALSE
  )
}
```

The empirical graphical checks use the observed weekly dataset with death
carried forward through week 26. Model-derived checks use the first completed
dataset and its fitted model as a representative imputation-specific
diagnostic. SOPs, transition proportions, correlations, and variograms are
kept as separate files so they can be compared directly. Correlation plots are
used for TAOOH because its dense weekly observations make the empirical
time-by-time correlation structure estimable.

```{r}
#| label: derive-taooh-graphical-diagnostics
#| include: false

diagnostic_index <- indices[[1L]]
data <- arrow::read_parquet(file.path(
  here::here("artifacts", "primary", "taooh", "imputations"),
  sprintf("imputation-%02d.parquet", diagnostic_index)
)) |>
  as.data.frame()
data$y <- factor(data$y, levels = 1:5, ordered = TRUE)
data$yprev <- factor(data$yprev, levels = 1:5)
data$ypprev <- factor(data$ypprev, levels = 1:5)

cohort <- arrow::read_parquet(here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
)) |>
  as.data.frame()
empirical_data <- arrow::read_parquet(here::here(
  "analysis", "00-data", "results", "analysis-taooh.parquet"
)) |>
  as.data.frame() |>
  left_join(cohort |> select(id, tx), by = "id") |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(
    y = as.integer(y_taooh),
    yprev = lag(y)
  ) |>
  ungroup() |>
  filter(between(week, 1L, 26L)) |>
  mutate(
    y = factor(y, levels = 1:5, ordered = TRUE),
    yprev = factor(yprev, levels = 1:5),
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Intervention"))
  )

model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", diagnostic_index)))
sops <- arrow::read_parquet(file.path(
  here::here("artifacts", "primary", "taooh", "estimand"),
  sprintf("sops-draws-%02d.parquet", diagnostic_index)
))
diagnostic_times <- c(1L, 3L, 6L, 9L, 12L, 16L, 20, 24, 26)
diagnostic_seed <- analysis_seed(config, "diagnostics", diagnostic_index)

empirical_sop_plot <- markov.misc::plot_sops(
  empirical_data,
  time_var = "week",
  y_var = "y",
  facet_var = NULL,
  linetype_var = "tx",
  geom = "line"
) +
  ggplot2::scale_color_brewer(palette = "Dark2")

empirical_transition_plot <- markov.misc::plot_transitions(
  empirical_data,
  times = diagnostic_times,
  time_var = "week",
  y_var = "y",
  p_var = "yprev",
  y_levels = 1:5,
  # facet_var = "tx",
  show_values = TRUE
)
empirical_transition_plot$layers[[2]]$aes_params$colour <- "white"

model_transition_plot <- markov.misc::plot_transitions(
  model,
  times = diagnostic_times,
  y_levels = 1:5,
  absorb = 5L,
  time_var = "week",
  p_var = "yprev",
  p2_var = "ypprev",
  id_var = "id",
  # facet_var = "tx",
  seed = diagnostic_seed,
  n_draws = 200L,
  show_values = TRUE
)

model_transition_plot$layers[[2]]$aes_params$colour <- "white"

empirical_correlation_plot <- markov.misc::plot_correlation(
  empirical_data,
  times = 1:26,
  y_levels = 1:5,
  id_var = "id",
  time_var = "week",
  y_var = "y",
  # facet_var = "tx",
  show_values = TRUE
)

model_correlation_plot <- markov.misc::plot_correlation(
  model,
  times = 1:26,
  y_levels = 1:5,
  absorb = 5L,
  id_var = "id",
  time_var = "week",
  p_var = "yprev",
  p2_var = "ypprev",
  # facet_var = "tx",
  seed = diagnostic_seed,
  n_draws = 50L,
  show_values = TRUE
)
empirical_variogram_plot <- markov.misc::plot_variogram(
  empirical_data,
  times = 1:26,
  y_levels = 1:5,
  id_var = "id",
  time_var = "week",
  y_var = "y",
  # facet_var = "tx"
)
model_variogram_plot <- markov.misc::plot_variogram(
  model,
  times = 1:26,
  y_levels = 1:5,
  absorb = 5L,
  id_var = "id",
  time_var = "week",
  p_var = "yprev",
  p2_var = "ypprev",
  # facet_var = "tx",
  seed = diagnostic_seed,
  n_draws = 50L
)

diagnostic_plots <- list(
  "empirical-sops.png" = empirical_sop_plot,
  # "model-sops.png" = model_sop_plot,
  "empirical-transitions.png" = empirical_transition_plot,
  "model-transitions.png" = model_transition_plot,
  "empirical-correlation.png" = empirical_correlation_plot,
  "model-correlation.png" = model_correlation_plot,
  "empirical-variogram.png" = empirical_variogram_plot,
  "model-variogram.png" = model_variogram_plot
)
for (filename in names(diagnostic_plots)) {
  ggplot2::ggsave(
    file.path(figure_dir, filename),
    diagnostic_plots[[filename]],
    width = if (grepl("transition", filename)) 12 else 8,
    height = if (grepl("transition", filename)) 9 else 5.5,
    dpi = 300,
    limitsize = FALSE
  )
}
```

```{r}
#| label: fig-taooh-sop-check
#| fig-cap: "Empirical and model-derived TAOOH state occupancy probabilities in the first completed dataset"
#| fig-subcap:
#|   - "Empirical state occupancy probabilities"
#|   - "Model-derived state occupancy probabilities"
#| layout-ncol: 1
#| fig-width: 8
#| fig-height: 5.5
#| echo: false

empirical_sop_plot
# model_sop_plot
```

```{r}
#| label: fig-taooh-correlation-check
#| fig-cap: "Empirical and second-order model-derived TAOOH correlation structures in the first completed dataset"
#| fig-subcap:
#|   - "Empirical correlations"
#|   - "Model-derived correlations"
#| layout-ncol: 2
#| fig-width: 8
#| fig-height: 5.5
#| echo: false

empirical_correlation_plot
model_correlation_plot
```
