---
title: "Multiple Imputation Strategy for Quality of Life (QoL) Outcome"
format: 
  typst:
    papersize: a4
    margin:
      x: 2cm
      y: 2.5cm
    toc: true
    columns: 1
    tbl-cap-location: bottom
---

## Overview

This document describes the multiple imputation strategy for the Quality of Life (QoL) outcome in the LIQPLAT trial. The approach uses Multiple Imputation by Chained Equations (MICE) with multilevel predictive mean matching (PMM) to handle missing data under the Missing At Random (MAR) assumption.

## Setup

```{r}
#| output: false
library(here)
library(tidyverse)
library(patchwork)
library(mice)
library(miceadds)
library(arrow)
```

## Data

The following code uses historical data from the Division of Medical Oncology at USB (first appointments from 2023 to mid-2024, prior to LIQPLAT initiation). These data are used to develop and validate the imputation procedure; actual trial data are not publicly available.

```{r}
set.seed(1234)
df_raw <- read_parquet(here("data", "alt_qol.parquet"))
```

## Data Preprocessing

### Inclusion Criteria

The substantive analysis model includes only patients who survived beyond 182 days. To maintain congruence between the imputation and analysis models, we restrict the imputation dataset to:

1.  Observations up to day 182 after random selection
2.  Patients who either have no recorded death or died after day 182

### Variable Transformations

Ordered factors (ECOG, QoL, distress) are converted to numeric format for imputation. This avoids sparse dummy variable matrices that can cause convergence issues in MICE. Predictive mean matching preserves the discrete nature of these variables by imputing only observed values.

```{r}
tx <- sample(unique(df_raw$pat_id), size = 0.5 * length(unique(df_raw$pat_id)), replace = FALSE)

imputation_vars <- c(
  "pat_id",
  "tx",
  "gender",
  "age", 
  "ecog_fstcnt",
  "diagnosis",
  "plan_fstcnt_coded",
  "week",
  "distress",
  "q30"
  # Variables available in trial data but not in historical dataset:
  # "albumin",
  # "c_reactive_protein"
)

df <- df_raw |> 
  filter(quest_day <= 182, (is.na(death_day) | death_day >= 182)) |>
  mutate(
    tx = if_else(pat_id %in% tx, "1", "0"),
    distress = as.numeric(as.factor(coalesce(distress_therm_2, distress_therm_v2, distress_therm_inpat)), ordered = TRUE),
    week = ceiling(quest_day / 7),
    diagnosis = fct_lump_n(diagnosis, n = 4),
    diagnosis = factor(as.integer(diagnosis)),
    plan_fstcnt_coded = fct_lump_n(plan_fstcnt_coded, n = 4),
    plan_fstcnt_coded = factor(as.integer(plan_fstcnt_coded)),
    q30       = as.numeric(as.character(q30)),    
    ecog_fstcnt = as.numeric(as.character(ecog_fstcnt)),
    tx = factor(tx)
    ) |> 
  rename(age = pat_age) |> 
  select(all_of(imputation_vars)) |> 
  arrange(pat_id, week)

# Extract baseline QoL: last observation at or before random selection (week <= 0)
baseline_qol <- df |> 
  filter(week <= 0) |> 
  select(pat_id, q30, week) |>
  group_by(pat_id) |>
  arrange(desc(week)) |>
  slice_head(n = 1) |> 
  ungroup() |> 
  select(pat_id, baseline_q30 = q30)

df <- df |>
  left_join(baseline_qol, by = "pat_id") |> 
  filter(week > 0) |> 
  relocate(baseline_q30, ecog_fstcnt, diagnosis, plan_fstcnt_coded, .after = pat_id)
  # group_by(pat_id) |>
  # mutate(
  #   yprev = lag(q30, n = 1L, order_by = week)) |> 
  # fill(yprev, .direction = "down") |>
  # mutate(
  #   yprev = if_else(is.na(yprev), baseline_q30, yprev)
  #   ) |> 
  # ungroup() 

# Model non-linear time effects using restricted cubic splines with 4 knots
basis_rcs <- rms::rcs(df$week, 4)

df$week <- as.vector(basis_rcs[, 1])
df$week_nl1 <- as.vector(basis_rcs[, 2])
df$week_nl2 <- as.vector(basis_rcs[, 3])
```

## Imputation Variables {#sec-variables}

The imputation model includes variables from the substantive analysis model and auxiliary variables that are predictive of missing values or their missingness mechanism \[Mathur2025-sl; van-Buuren2021-pc\].

### Analysis Variables

| Variable | Description | Level |
|------------------------|-------------------------------|------------------|
| `tx` | Treatment group allocation | Baseline |
| `age` | Age at date of random selection (years) | Baseline |
| `gender` | Patient sex | Baseline |
| `ecog_fstcnt` | ECOG performance status at baseline (0–4) | Baseline |
| `diagnosis` | Primary cancer diagnosis (5 categories) | Baseline |
| `baseline_q30` | EORTC QLQ-C30 Q30 at baseline | Baseline |
| `week` | Week of questionnaire completion (post-random selection) | Longitudinal |
| `q30` | EORTC QLQ-C30 question 30 (overall QoL, 1–7) | Longitudinal |

: Analysis variables included in the imputation model {#tbl-analysis-vars}

### Auxiliary Variables

| Variable | Description | Level |
|------------------------|-------------------------------|------------------|
| `distress` | Distress thermometer score (0–10) | Longitudinal |
| `plan_fstcnt_coded` | Treatment intent at baseline (5 categories) | Baseline |
| `albumin` | Serum albumin at baseline (g/L) | Baseline |
| `c_reactive_protein` | C-reactive protein at baseline (mg/L) | Baseline |

: Auxiliary variables included in the imputation model. Albumin and CRP are available in trial data but not in the historical dataset used here. {#tbl-auxiliary-vars}

## Multiple Imputation Procedure {#sec-imputation}

We employ multilevel multiple imputation to account for the hierarchical data structure, where repeated QoL measurements (Level 1) are nested within patients (Level 2). The patient identifier (`pat_id`) serves as the clustering variable.

### Imputation Methods

**Baseline variables** (Level 2) are imputed using `2lonly.pmm` from the `miceadds` package. This method:

1.  Aggregates longitudinal predictors to the cluster level by computing patient means

2.  Fits a linear regression model on the collapsed data: 
$$
    Y_j = \beta_0 + \beta_1 \mathbf{X}_j + \beta_2 \bar{\mathbf{Z}}_j + \epsilon_j
$$

    where $Y_j$ is the missing baseline variable for patient $j$, $\mathbf{X}_j$ represents other baseline predictors, and $\bar{\mathbf{Z}}_j$ represents aggregated means of longitudinal predictors

3.  Selects imputed values via predictive mean matching

**Longitudinal variables** (Level 1) are imputed using `2l.pmm`, which fits a linear mixed-effects model: 
$$
Y_{ij} = \beta_0 + \beta_1 \mathbf{X}_{ij} + u_{0j} + \epsilon_{ij}
$$

where $Y_{ij}$ is the outcome for observation $i$ of patient $j$, $\mathbf{X}_{ij}$ includes fixed effects (time-varying and static predictors), and $u_{0j} \sim N(0, \sigma^2_u)$ is the patient-specific random intercept.

For both methods, we use Type 1 predictive mean matching with a donor pool of $k = 5$ \[van-Buuren2021-pc, chapters 3.4, 7.3, 7.8\].

### Predictor Matrix Configuration

The predictor matrix is configured to:

-   Use `pat_id` as the clustering variable (coded as $-2$)
-   Prevent time variables from predicting static baseline variables
-   Allow baseline variables to predict longitudinal outcomes

```{r}
#| output: false
md.pattern(df)

init <- mice(df, maxit = 0)
pred_matrix <- init$predictorMatrix
meth <- init$method

# Set patient ID as clustering variable
pred_matrix[, "pat_id"] <- -2
pred_matrix["pat_id", ] <- 0

# Time variables should not predict static baseline variables
baseline_vars <- c("tx", "age", "gender", "diagnosis", "ecog_fstcnt", "plan_fstcnt_coded", "baseline_q30")
pred_matrix[c(baseline_vars, "week", "week_nl1", "week_nl2"), c("week", "week_nl1", "week_nl2")] <- 0

# Longitudinal variables: multilevel PMM
meth["q30"] <- "2l.pmm"
meth["distress"] <- "2l.pmm"
# meth["yprev"] <- ""

# Baseline variables: cluster-level PMM
meth["ecog_fstcnt"] <- "2lonly.pmm"
meth["plan_fstcnt_coded"] <- "2lonly.pmm"
meth["diagnosis"] <- "2lonly.pmm"
meth["baseline_q30"] <- "2lonly.pmm"


imputed_data <- mice(df,
                     m = 5,        # 50 in final analysis
                     maxit = 5,    # 50 in final analysis
                     predictorMatrix = pred_matrix,
                     method = meth,
                     donors = 5,
                     seed = 78109823)
```

### Imputation Parameters

For the final trial analysis:

-   Number of imputed datasets: $m = 50$
-   Maximum iterations: 50
-   Donor pool size: $k = 5$

## Convergence Diagnostics {#sec-diagnostics}

Convergence is assessed via trace plots of the mean and standard deviation of imputed values across iterations.

```{r trace-plots}
#| fig-cap: "Trace plots showing the mean and standard deviation of imputed values across iterations for each imputed dataset."
plot(imputed_data)
```

## Distribution Diagnostics {#sec-distribution}

We compare the distributions of observed and imputed values to assess plausibility.

```{r}
#| fig-cap: "Density plots comparing observed (blue) and imputed (red) values for baseline variables and QoL."
densityplot(imputed_data, ~baseline_q30 + q30 + ecog_fstcnt + diagnosis + plan_fstcnt_coded)
```

```{r}
#| fig-cap: "Strip plots of imputed values for baseline variables across imputed datasets."
stripplot(imputed_data, baseline_q30 + ecog_fstcnt + diagnosis + plan_fstcnt_coded ~ .imp)
```

```{r}
#| fig-cap: "Scatter plots of QoL against time, by imputed dataset and treatment group, and against baseline QoL."
xyplot(imputed_data, q30 ~ week | .imp, pch = c(1, 20))
xyplot(imputed_data, q30 ~ week | tx, pch = c(1, 20))
xyplot(imputed_data, q30 ~ baseline_q30 | .imp, pch = c(1, 20))
```

```{r}
#| fig-cap: "Scatter plot of QoL against distress thermometer score by imputed dataset."
xyplot(imputed_data, q30 ~ distress | .imp, pch = c(1, 20))
```

## Analysis of Trial Data

For the final trial analysis, all $m = 50$ imputed datasets will be analysed, and posterior draws will be stacked to propagate imputation uncertainty into the final inference.

## Simulation Study Dataset

For simulation studies, a single completed dataset is extracted. This is computationally necessary but does not reflect the full imputation uncertainty.

```{r}
#| fig-cap: "Distribution of QoL scores over time in the selected completed dataset."
complete_data <- complete(imputed_data, 4)

complete_data |> 
  mutate(q30 = as.numeric(q30)) |> 
  ggplot(aes(x = week, y = q30)) +
  geom_point(alpha = 0.5) +
  labs(x = "Week", y = "QoL (Q30)")
```

## Export

```{r}
write_parquet(complete_data, sink = here("data", "qol_complete.parquet"))
```