---
title: Simulating a super-population
format: 
  typst:
    papersize: a4
    margin:
      x: 2cm
      y: 2.5cm
    toc: false
    columns: 1
    tbl-cap-location: bottom
---

The historical data is unfortunately too sparse to derive plausible simulated datasets. Where therefore generate our own, using a latent brownian motion model.

## Setup

```{r}
#| output: false
library(here)
library(tidyverse)
library(arrow)
library(patchwork)
library(markov.misc)
library(data.table)
library(Hmisc)
library(rms)
library(VGAM)

set.seed(1234)
```

```{r}
data <- read_parquet(here("data", "qol_historic.parquet")) |> 
  mutate(
    yprev = factor(yprev, ordered = FALSE),
    ecog_fstcnt = factor(ecog_fstcnt)
  )
```

## Fit PO model

```{r}
# we ignore the gap in the historic data, because the data are too sparse and adding gap doesn't provide a better fit. 
# In LIQPLAT we have about three times the data, with shorter spacings, so we will add the gap then
# We use polynomial here, because it makes it easier to simulate data later, but results very similar to splines
m <- vglm(ordered(y) ~ pol(week, 3) + yprev + ecog_fstcnt + diagnosis, family = cumulative(reverse = TRUE, parallel = TRUE), data = data)

# we can use standardize_sops from markov misc to calculate marginal sops
# we don't include tx in the linear predictor, so control or treatment will be the same
bs_sops <- bootstrap_standardized_sops(m, data = data, n_boot = 96, workers = 8, parallel = TRUE, ylevels = factor(1:7), absorb = NULL, times = 1:26, varnames = list(tvarname = "week", pvarname = "yprev", id = "pat_id", tx = "tx", gap = NULL))

# it all seems quite compatible with constant QoL over time, maybe slight initialy until about 12 weeks, then slight deterioration again
plot_bootstrap_sops(bs_sops$sops) +
  coord_cartesian(ylim = c(0, 0.3))
```

## Simulate Data

### Markov Data

For setting a prior on the treatment effect, we simulate some data from the model fitted to historical data.

```{r}
ints <- rev(coef(m)[1:6])
extra_params <- coef(m)[-c(1:6)]
fu <- 26

N <- 1e4

baseline_data <- data |> 
  select(pat_id, baseline_q30, ecog_fstcnt, diagnosis, tx) |>
  distinct() |> 
  mutate(
    yprev = factor(baseline_q30, ordered = TRUE),
    week = 1) |> 
  slice_sample(n = N, replace = TRUE) |> 
  mutate(
    id = row_number()
  ) |> 
  select(-pat_id, -week)

g <- function(yprev, t, tx, ecog_fstcnt, diagnosis, parameter, extra_params) {
  tx_effect <- parameter * as.numeric(as.character(tx))
  time_effect <- extra_params["pol(week, 3)week"] * t + 
    extra_params["pol(week, 3)week^2"] * t^2 + 
    extra_params["pol(week, 3)week^3"] * t^3

   yprev_effect <- rep(0, length(yprev))
    for (state in unique(yprev)) {
        if (state != 1) {
            state_idx <- which(yprev == state) # use which because function is vectorized
            yprev_coef_name <- paste0("yprev", state)
            if (yprev_coef_name %in% names(extra_params)) {
                yprev_effect[state_idx] <- extra_params[[yprev_coef_name]]
            }
        }
    }
    diagnosis_effect <- rep(0, length(diagnosis))
    for (diag in unique(diagnosis)) {
        if (diag != levels(diagnosis)[1]) {
            diag_idx <- which(diagnosis == diag)
            diag_coef_name <- paste0("diagnosis", diag)
            if (diag_coef_name %in% names(extra_params)) {
                diagnosis_effect[diag_idx] <- extra_params[[diag_coef_name]]
            }
        }
    }
    ecog_effect <- ifelse(ecog_fstcnt == 2, extra_params["ecog_fstcnt2"], 0)
    lp <- as.numeric(tx_effect + time_effect + yprev_effect + diagnosis_effect + ecog_effect)
    lp
}
```

```{r}
# H0
d <- sim_trajectories_markov(
  baseline_data = baseline_data,
  follow_up_time = fu,
  intercepts = ints, 
  lp_function = g,
  extra_params = extra_params,
  parameter = log(1),
  absorbing_states = NULL,
  seed = 1234,
  covariate_names = c("tx", "ecog_fstcnt", "diagnosis")
)

plot_sops(d, time_var = "time", geom = "line", facet_var = NULL) +
  coord_cartesian(ylim = c(0, 0.3))

plot_sops(d, time_var = "time", facet_var = "tx") +
  scale_fill_brewer(palette = "Dark2")

calc_time_in_state_diff(d, target_state = 1:3)

# HA: largest plausible effect (only 10% of prior mass will be above this)
d_ha <- sim_trajectories_markov(
  baseline_data = baseline_data,
  follow_up_time = fu,
  intercepts = ints, 
  lp_function = g,
  extra_params = extra_params,
  parameter = log(0.53),
  absorbing_states = NULL,
  seed = 1234,
  covariate_names = c("tx", "ecog_fstcnt", "diagnosis")
)

plot_sops(d_ha, time_var = "time", facet_var = "tx") +
  scale_fill_brewer(palette = "Dark2")

calc_time_in_state_diff(d_ha, target_state = 1:3)

# Inverge of the largest plausible effect (because we have to set normal prior on log scale)
d_ha2 <- sim_trajectories_markov(
  baseline_data = baseline_data,
  follow_up_time = fu,
  intercepts = ints, 
  lp_function = g,
  extra_params = extra_params,
  parameter = log(1.9),
  absorbing_states = NULL,
  seed = 1234,
  covariate_names = c("tx", "ecog_fstcnt", "diagnosis")
)

plot_sops(d_ha2, time_var = "time", facet_var = "tx") +
  scale_fill_brewer(palette = "Dark2")

calc_time_in_state_diff(d_ha2, target_state = 1:3)
```

### Brownian Motion

We simulate data based on a brownian motion model, which very loosely approximates the data generated from the markov model. I.e., initial increase followed by decrease after about 15 weeks.

```{r}
int <- c(-2.5, -0.5, 1, 2.5, 5, 6.5)
K <- 7
fu <- 26
mu <- 0.15
mu_quad <- -0.011
sigma <- 0.5
N <- 2e5

super_pop_h0 <- sim_trajectories_brownian(
  n_patients = N,
  follow_up_time = fu,
  n_states = K,
  mu_drift = mu,
  mu_quad = mu_quad,
  sigma_rw = sigma,
  thresholds = int,
  absorbing_state = NULL,
  drift_change_times = c(20, 50)
  ) |> 
  mutate(week = time)

effect_h0 <- calc_time_in_state_diff(super_pop_h0, target_state = 1:3)$true_effect

plot_sops(super_pop_h0) +
  scale_fill_brewer(palette = "Dark2")

write_parquet(super_pop_h0, here("1-qol", "sim-data", sprintf("super_pop_h0_effect_%.2f.parquet", effect_h0)))
```

1 week improvement

```{r}
super_pop_ha1 <- sim_trajectories_brownian(
  n_patients = N,
  follow_up_time = fu,
  n_states = K,
  mu_drift = mu,
  mu_quad = mu_quad,
  mu_treatment_effect = -0.02,
  sigma_rw = sigma,
  thresholds = int,
  absorbing_state = NULL,
  drift_change_times = c(20, 50)
  ) |> 
  mutate(week = time)

effect_ha1 <- calc_time_in_state_diff(super_pop_ha1, target_state = 1:3)$true_effect
write_parquet(super_pop_ha1, here("1-qol", "sim-data", sprintf("super_pop_ha_effect_%.2f.parquet", effect_ha1)))

plot_sops(super_pop_ha1) +
  scale_fill_brewer(palette = "Dark2")

calc_time_in_state_diff(super_pop_ha1, target_state = 1:3)
```

2 week improvement

```{r}
super_pop_ha2 <- sim_trajectories_brownian(
  n_patients = N,
  follow_up_time = fu,
  n_states = K,
  mu_drift = mu,
  mu_quad = mu_quad,
  mu_treatment_effect = -0.042,
  sigma_rw = sigma,
  thresholds = int,
  absorbing_state = NULL,
  drift_change_times = c(20, 50)
  ) |> 
  mutate(week = time)

effect_ha2 <- calc_time_in_state_diff(super_pop_ha2, target_state = 1:3)$true_effect
write_parquet(super_pop_ha2, here("1-qol", "sim-data", sprintf("super_pop_ha_effect_%.2f.parquet", effect_ha2)))

plot_sops(super_pop_ha2) +
  scale_fill_brewer(palette = "Dark2")

calc_time_in_state_diff(super_pop_ha2, target_state = 1:3)
```

3 week improvement

```{r}
super_pop_ha3 <- sim_trajectories_brownian(
  n_patients = N,
  follow_up_time = fu,
  n_states = K,
  mu_drift = mu,
  mu_quad = mu_quad,
  mu_treatment_effect = -0.065,
  sigma_rw = sigma,
  thresholds = int,
  absorbing_state = NULL,
  drift_change_times = c(20, 50)
  ) |> 
  mutate(week = time)

effect_ha3 <- calc_time_in_state_diff(super_pop_ha3, target_state = 1:3)$true_effect
write_parquet(super_pop_ha3, here("1-qol", "sim-data", sprintf("super_pop_ha_effect_%.2f.parquet", effect_ha3)))

plot_sops(super_pop_ha3) +
  scale_fill_brewer(palette = "Dark2")

calc_time_in_state_diff(super_pop_ha3, target_state = 1:3)
```

## Plots

### Individual Trajectories

```{r}
#tile plot
ids <- super_pop_h0$id |> unique() |> sample(20)

fig_tile_random <- super_pop_h0 |> 
  filter(id %in% ids) |> 
  mutate(id = as.factor(as.numeric(as.factor((id))))) |> 
  ggplot(aes(y = id, x = time)) +
  geom_tile(mapping = aes(fill = factor(y)), width = 0.95, height = 0.4) +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = 1:26) +
  labs(title = "Individual Trajectories - No Treatment Effect",
       y = "Patient ID",
       x = "Study Week",
       fill = "State")

fig_tile_random
```

The serial correlation appears appear to be quite low.

### Transition Plot

```{r}
fu <- 26
# Create the transition data.frame for week = x
get_transitions_df <- function(df, x){
  
  
  subset <- df[df$week == x, ]
  subtable <- table(subset$y, subset$yprev, dnn = c("y", "yprev")) |> 
  as.data.table()

  subtable[, week := x]
  return(subtable)
}

# Get the transition data frame for each week and combine
# into a single data frame
df_transitions <- 
  map(2:fu, ~get_transitions_df(super_pop_h0, .x)) |>
  rbindlist()

# Compute the proportion of transitions relative to the number of
# transitions on each week
df_transitions <-
  df_transitions |>
    mutate(total = sum(N),
           p = N / total, .by=week)
```

```{r}
FONT_SIZE <- 10
df_transitions |>
    filter(week%%2 == 0) |> 
    mutate(yprev = as.factor(yprev), y = as.factor(y)) |>
    filter(p != 0) |>
    ggplot() +
    aes(x = yprev, y = y, size = p) +
    geom_point() +
    scale_size_area(breaks = seq(0, 1, by = 0.1), max_size = 4) +
    facet_wrap(~week,
               nrow = 4,
               scales = "free",
               labeller = labeller(week = \(x) glue::glue("week {as.integer(x)-1} to {as.integer(x)}"))) +
    scale_x_discrete(drop=FALSE) +
    scale_y_discrete(drop=FALSE) +
    theme(axis.ticks = element_blank()) +
    theme(aspect.ratio = 0.9) +
    labs(x = "Previous State",
         y = "Current State",
         size = "Proportion")
```

### Correlation Plot

```{r}
# Create correlation matrix
corr <-
  super_pop_h0 |> 
    select(id, week, y) |>
    arrange(week) |>
    mutate(y = as.integer(y)) |>
    pivot_wider(names_from = week, values_from = y) |>
    ungroup() |>
    select(-id) |>
    cor(method="spearman", use="pairwise.complete.obs")

corr |>
  as.data.frame() |>
  rownames_to_column() |>
  pivot_longer(-rowname) |>
  mutate(rowname = as.integer(rowname), name = as.integer(name)) |>
  filter(rowname < name) |> #remove tiles below diagonal, because redundant
  mutate(rowname = as.factor(rowname), name = as.factor(name)) |>
  ggplot() + 
    aes(x=rowname, y=name, fill=value) +
    geom_tile() +
    geom_text(mapping=aes(x=rowname, y=name, label=round(value, 2)),
              color="black",
              size = 3) +
    scale_fill_viridis_c(limits=c(-0.01,1), n.breaks = 10) +
    expand_limits(value = c(0, 1)) +
    # cowplot::theme_half_open(font_size=FONT_SIZE,
    #           font_family = "Source Sans Pro") +
    theme(plot.title = element_text(hjust = 0.5),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    theme(axis.line = element_blank(),
          axis.ticks = element_blank()) +
    theme(legend.key.height = unit(0.8, "in")) +
    labs(x="Study week",
         y="Study week",
         fill="")
```

### Variogram

Correlation deacreases over time, as expected.

```{r}
plotCorrM(corr)[[2]]
```