---
title: Simulation of different scenarios
format: 
  typst:
    papersize: a4
    margin:
      x: 2cm
      y: 2.5cm
    toc: false
    columns: 1
    tbl-cap-location: bottom
---



## Setup

```{r}
#| output: false
library(here)
library(tidyverse)
library(data.table)
library(tidybayes)
library(rmsb)
library(rms)
library(Hmisc)
library(arrow)
library(VGAM)
library(patchwork)

set.seed(1234)
```

## Load model and dataset

```{r}
df <- read_parquet(here("data", "censored_qol.parquet")) |> 
  mutate(
    ecog_fstcnt = if_else(ecog_fstcnt == "4" | ecog_fstcnt == "3", "3plus", ecog_fstcnt), 
    ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE)
  )

model <- readRDS(here("1-qol", "models", "qol_long_ord_pol.rds"))
model
```

## Extract median of posterior draws

```{r}
draws <- as.data.table(model$draws) 

coef <- draws |> 
  summarise_all(median)

coef
```

## Baseline-Covariates

```{r}
N <- 100000 # Simulation size

baseline_w <- df |> 
  select(-c(y, y.a, y.b, week, surv_weeks, tx, gap)) |> 
  group_by(pat_id) |> 
  slice_head(n = 1) |> 
  ungroup()

i <- sample(1:nrow(baseline_w), N, replace = TRUE)
sim_baseline_data <- baseline_w[i, ] |> 
  mutate(pat_id = row_number())

init <- sim_baseline_data$yprev # initial states to be passed to g function

baseline_formula <- ~ ecog_fstcnt + diagnosis + gender + pat_age + plan_fstcnt_coded
X_baseline <- model.matrix(baseline_formula, data = sim_baseline_data)[, -1] # Baseline values dummy coded, remove intercept

tx <- sample(0:1, N, replace = TRUE) # Sample treatment assignment
X <- cbind(X_baseline, tx) 
```

## Simulation Functions

### Function A

- Proportional Part: Effects that are constant across all k-1 logits (e.g., yprev, gap, baseline covariates, pol(week,2)).
- Non-Proportional Part: The week x f(y) term, which varies with the outcome level k. For our model cppo = function(y) y, this term is k * week * \gamma_{week}.

```{r}
ints <- unlist(coef[1:7])
extra <- unlist(coef[- (1:8)]) #remove tx coef as well

names(extra)[names(extra) == "ecog_fstcnt=2"] <- "ecog_fstcnt2"
names(extra)[names(extra) == "ecog_fstcnt=3plus"] <- "ecog_fstcnt3plus"
names(extra)[names(extra) == "diagnosis=2"] <- "diagnosis2"
names(extra)[names(extra) == "diagnosis=3"] <- "diagnosis3"
names(extra)[names(extra) == "diagnosis=4"] <- "diagnosis4"
names(extra)[names(extra) == "diagnosis=5"] <- "diagnosis5"
names(extra)[names(extra) == "plan_fstcnt_coded=2"] <- "plan_fstcnt_coded2"
names(extra)[names(extra) == "plan_fstcnt_coded=3"] <- "plan_fstcnt_coded3"
names(extra)[names(extra) == "plan_fstcnt_coded=4"] <- "plan_fstcnt_coded4"
names(extra)[names(extra) == "plan_fstcnt_coded=5"] <- "plan_fstcnt_coded5"
names(extra)[names(extra) == "gender=male"] <- "gendermale"

g <- function(yprev, t, gap, X, parameter = 0, extra) {

  # X is a vector for one patient, containing baseline covariates and tx (which we want to remove)
  # yprev is the previous state (character)
  # t is the current week (numeric)
  # gap is assumed to be 1 for this simulation
  gap <- 1 

  # 0. Treatment Effect
  # The treatment indicator 'tx' is the last element of the X vector.
  tx_value <- X[length(X)]
  tx_effect <- parameter * tx_value

  # 1. Baseline effect
  # X contains the non-treatment variables. We match its names to the 'extra' vector.
  X_nontx <- X[-length(X)] # X is a single vector. We remove tx which is as the end of this vector
  baseline_coef_names <- names(X_nontx)

  baseline_effect <- sum(extra[baseline_coef_names] * X_nontx)

  # 2.  time effect (proportional part) from pol(week, 2)
  time_effect <- extra[["week"]] * t + extra[["week^2"]] * (t^2)

  # 3. Gap effect
  gap_effect <- extra[["gap"]] * gap

  # This is if yprev is equal to 1 (the reference group)
  yprev_effect <- 0
  yprev_gap_effect <- 0

  # 4. Previous state effect
  yprev_coef_name <- paste0('yprev=', yprev)
  if (yprev_coef_name %in% names(extra)) {
      yprev_effect <- extra[[yprev_coef_name]]
  }
  
  yprev_gap_coef_name <- paste0('yprev=', yprev, ' * gap')
  if (yprev_gap_coef_name %in% names(extra)) {
      yprev_gap_effect <- extra[[yprev_gap_coef_name]] * gap
  }

  # Sum of all proportional effects
  lin_pred_prop <- tx_effect + baseline_effect + time_effect + gap_effect + yprev_effect + yprev_gap_effect

  # 5. Non-proportional effect of time
  # From cppo=function(y) y, the effect for P(Y>=k) is k * week * gamma
  # The columns of the final lp correspond to k = 2, 3, ..., 8
  nonprop_gamma <- extra[['week x f(y)']]
  k_levels <- 2:8 
  lin_pred_nonprop <- k_levels * t * nonprop_gamma

  # 6. Final linear predictor (vector of length 7)
  # This is the sum of the proportional part (a scalar) and non-proportional part (a vector)
  lp <- lin_pred_prop + lin_pred_nonprop
    
  return(lp)

}

formals(g)$extra <- extra
```

### Function B (no impact on survival)

- This is a somewhat implausible scenario
- For LIQPLAT we hypothesize that ctDNA might lead to earlier deescalation, less toxicity and in turn better QoL
- The quality of life is thus mediated by bodily changes, which may impact survival as well
- For simulation purposes, we could handle a situation where the treatment does not affect survival with a conservative prior on the non-proportionality of the treatment effect

```{r}
g2 <- function(yprev, t, gap, X, parameter = 0, extra) {

  # X is a vector for one patient, containing baseline covariates and tx
  # yprev is the previous state (character)
  # t is the current week (numeric)
  # the gaps between measurements gap is assumed to be 1 for this simulation
  gap <- 1 

  # 0. Treatment Effect
  # The treatment indicator 'tx' is the last element of the X vector.
  tx_value <- X[length(X)]
  tx_effect <- parameter * tx_value

  # 1. Baseline effect
  # X contains the non-treatment variables. We match its names to the 'extra' vector.
  X_nontx <- X[-length(X)] # We remove tx which is as the end of this vector
  baseline_coef_names <- names(X_nontx)

  baseline_effect <- sum(extra[baseline_coef_names] * X_nontx)

  # 2.  time effect (proportional part) from pol(week, 2)
  time_effect <- extra[["week"]] * t + extra[["week^2"]] * (t^2)

  # 3. Gap effect
  gap_effect <- extra[["gap"]] * gap

  # This is if yprev is equal to 1 (the reference group)
  yprev_effect <- 0
  yprev_gap_effect <- 0

  # 4. Previous state effect
  yprev_coef_name <- paste0('yprev=', yprev)
  if (yprev_coef_name %in% names(extra)) {
      yprev_effect <- extra[[yprev_coef_name]]
  }
  
  yprev_gap_coef_name <- paste0('yprev=', yprev, ' * gap')
  if (yprev_gap_coef_name %in% names(extra)) {
      yprev_gap_effect <- extra[[yprev_gap_coef_name]] * gap
  }

  # Sum of all proportional effects
  lin_pred_prop <- baseline_effect + time_effect + gap_effect + yprev_effect + yprev_gap_effect

  # 5. Non-proportional effect of time
  # From cppo=function(y) y, the effect for P(Y>=k) is k * week * gamma
  # The columns of the final lp correspond to k = 2, 3, ..., 8
  nonprop_gamma <- extra[['week x f(y)']]
  k_levels <- 2:8 
  lin_pred_nonprop <- k_levels * t * nonprop_gamma

  # 6. Final linear predictor (vector of length 7)
  # This is the sum of the proportional part (a scalar) and non-proportional part (a vector)
  lp <- lin_pred_prop + lin_pred_nonprop + c(rep(tx_effect, 6), 0) # tx not applied to last threshold
    
  return(lp)

}

formals(g2)$extra <- extra
```

## Simulations
### PO - OR 0.8

- Odds of 0.8 of moving to a higher (worse state)

```{r}
# Run the simulation
carried_forward_08 <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:8, ordered = TRUE)),                # 8 levels of the outcome
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 8,
                  intercepts = ints, 
                  g = g, 
                  parameter = log(0.8),
                  carry = TRUE)

not_carried_forward_08 <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:8, ordered = TRUE)),                # 8 levels of the outcome
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = '8',
                  intercepts = ints, 
                  g = g, 
                  parameter = log(0.8),
                  carry = FALSE)
```

#### Visualize empirical SOPs

```{r}
carried_forward |>               
  ggplot(aes(x = time, fill = factor(y))) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(~ tx) +
  labs(x = "week", y = "proportion", fill = "Y")
```

#### Check if OR of 0.8 is recovered

```{r}
sample_08 <- not_carried_forward |> 
    select(id, time, y, yprev, tx) |> 
    left_join(sim_baseline_data |> select(-yprev), by = c("id" = "pat_id"))

h <- vglm(ordered(y) ~ yprev + poly(time, 2) + ecog_fstcnt + diagnosis + tx,
                cumulative(reverse=TRUE, parallel=FALSE ~ pol(time, 2)),
                data=sample |> filter(id < 10000))

h
```

#### Calculate the SOPs
```{r}
sops <- carried_forward |> 
  group_by(time, tx) |> 
  count(y) |> 
  reframe(
    y = y,
    sop = n / sum(n))
```

#### Calculate the difference in weeks with >= Good QoL
```{r}
# Difference in mean days with excellent, very good or good quality of life
weeks_good_qol <- sops |> 
  filter(y == 1 | y == 2 | y == 3) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_good_qol

weeks_bad_qol <- sops |> 
  filter(y == 7 | y == 6 | y == 5 | y == 8) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_bad_qol
```

In this population an OR of 0.8 for a proportional treatment effect leads to a 1.1 week increase in the expected time spent with a quality of life in the highest three categories. The difference in weeks spent with bad quality of life or death is -0.9. 


### nonPO - OR 0.8

- Odds of 0.8 of moving to a higher (worse state), except for the absorbing state death, which has OR of 1.

```{r}
# Run the simulation
carried_forward_nonPO_08 <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:8, ordered = TRUE)), # 8 levels of the outcome
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 8,
                  intercepts = ints, 
                  g = g2, 
                  parameter = log(0.8),
                  carry = TRUE)

not_carried_forward_nonPO_08 <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:8, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = '8',
                  intercepts = ints, 
                  g = g2, 
                  parameter = log(0.8),
                  carry = FALSE)
```

#### Visualize empirical SOPs

```{r}
carried_forward_nonPO |>              
  ggplot(aes(x = time, fill = factor(y))) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(~ tx) +
  labs(x = "week", y = "proportion", fill = "Y")
```

#### Examine OR

```{r}
sample_nonPO_08 <- not_carried_forward_nonPO |> 
    select(id, time, y, yprev, tx) |> 
    left_join(sim_baseline_data |> select(-yprev), by = c("id" = "pat_id"))

h_nonPO <- vglm(ordered(y) ~ yprev + poly(time, 2) + ecog_fstcnt + diagnosis + tx,
                cumulative(reverse=TRUE, parallel=FALSE ~ pol(time, 2)+tx),
                data=sample_nonPO)

h_nonPO
```

VGAM estimated separate effects for each cumulative threshold. Correctly estimated 0.8 for all but threshold 7, which is estimated as 1.

#### Calculate the SOPs
```{r}
sops_nonPO <- carried_forward_nonPO |> 
  group_by(time, tx) |> 
  count(y) |> 
  reframe(
    y = y,
    sop = n / sum(n))
```

#### Calculate the difference in weeks with >= Good QoL
```{r}
# Difference in mean days with excellent, very good or good quality of life
weeks_good_qol_nonPO <- sops_nonPO |> 
  filter(y == 1 | y == 2 | y == 3) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_good_qol_nonPO

weeks_bad_qol_nonPO <- sops_nonPO |> 
  filter(y == 7 | y == 6 | y == 5) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_bad_qol_nonPO
```

If the treatment has no effect on mortality, the effect is reduced from a gain of 1.1 weeks week >= Good QoL to a gain of 0.9 weeks. 

### nonPO - OR 0.3

- Odds Ratio of 0.3 of moving to a higher (worse state), except for the absorbing state death

```{r}
# Run the simulation
carried_forward_nonPO <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:8, ordered = TRUE)), # 8 levels of the outcome
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 8,
                  intercepts = ints, 
                  g = g2, 
                  parameter = log(0.3),
                  carry = TRUE)

not_carried_forward_nonPO <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:8, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = '8',
                  intercepts = ints, 
                  g = g2, 
                  parameter = log(0.3),
                  carry = FALSE)
```

#### Visualize empirical SOPs

```{r}
carried_forward_nonPO |>              
  ggplot(aes(x = time, fill = factor(y))) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(~ tx) +
  labs(x = "week", y = "proportion", fill = "Y")
```

#### Examine OR

```{r}
sample_nonPO <- not_carried_forward_nonPO |> 
    select(id, time, y, yprev, tx) |> 
    left_join(sim_baseline_data |> select(-yprev), by = c("id" = "pat_id"))

h_nonPO <- vglm(ordered(y) ~ yprev + poly(time, 2) + ecog_fstcnt + diagnosis + tx,
                cumulative(reverse=TRUE, parallel=FALSE ~ pol(time, 2)+tx),
                data=sample_nonPO |> filter(id < 10000))

h_nonPO
```

VGAM estimated separate effects for each cumulative threshold. Correctly estimated 0.3 for all but threshold 7, which is estimated as 1.

#### Calculate the SOPs

In those that survive > 26 weeks (primary estimand).
```{r}
sops_nonPO <- carried_forward_nonPO |> 
  group_by(id) |> 
  filter(all(y != 8)) |> 
  ungroup() |>
  group_by(time, tx) |> 
  count(y) |> 
  reframe(
    y = y,
    sop = n / sum(n))
```

#### Calculate the difference in weeks with >= Good QoL
```{r}
# Difference in mean days with excellent, very good or good quality of life
weeks_good_qol_nonPO <- sops_nonPO |> 
  filter(y == 1 | y == 2 | y == 3) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_good_qol_nonPO

weeks_bad_qol_nonPO <- sops_nonPO |> 
  filter(y == 7 | y == 6 | y == 5) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_bad_qol_nonPO
```

If the treatment has no effect on mortality, the effect of an OR of 0.3 is a gain of 3 weeks with good QoL.

### nonOR - OR 3.33

- Odds Ratio of 3.33 of moving to a higher (worse state), except for the absorbing state death

```{r}
# Run the simulation
carried_forward_nonPO <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:8, ordered = TRUE)), # 8 levels of the outcome
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = 8,
                  intercepts = ints, 
                  g = g2, 
                  parameter = log(3.33),
                  carry = TRUE)

not_carried_forward_nonPO <- simMarkovOrd(n = N, 
                  y = levels(factor(na.omit(df)$y, levels = 1:8, ordered = TRUE)),
                  times = 1:26,
                  initial = init, 
                  X = X, 
                  absorb = '8',
                  intercepts = ints, 
                  g = g2, 
                  parameter = log(3.33),
                  carry = FALSE)
```

#### Visualize empirical SOPs

```{r}
carried_forward_nonPO |>              
  ggplot(aes(x = time, fill = factor(y))) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(~ tx) +
  labs(x = "week", y = "proportion", fill = "Y")
```

#### Examine OR

```{r}
sample_nonPO <- not_carried_forward_nonPO |> 
    select(id, time, y, yprev, tx) |> 
    left_join(sim_baseline_data |> select(-yprev), by = c("id" = "pat_id"))

h_nonPO <- vglm(ordered(y) ~ yprev + poly(time, 2) + ecog_fstcnt + diagnosis + tx,
                cumulative(reverse=TRUE, parallel=FALSE ~ pol(time, 2)+tx),
                data=sample_nonPO |> filter(id < 10000))

h_nonPO
```

VGAM estimated separate effects for each cumulative threshold. Correctly estimated 3.33 for all but threshold 7, which is estimated as 1.

#### Calculate the SOPs

In those that survive > 26 weeks (primary estimand).
```{r}
sops_nonPO <- carried_forward_nonPO |> 
  group_by(id) |> 
  filter(all(y != 8)) |> 
  ungroup() |>
  group_by(time, tx) |> 
  count(y) |> 
  reframe(
    y = y,
    sop = n / sum(n))
```

#### Calculate the difference in weeks with >= Good QoL
```{r}
# Difference in mean days with excellent, very good or good quality of life
weeks_good_qol_nonPO <- sops_nonPO |> 
  filter(y == 1 | y == 2 | y == 3) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_good_qol_nonPO

weeks_bad_qol_nonPO <- sops_nonPO |> 
  filter(y == 7 | y == 6 | y == 5) |> 
  group_by(tx) |> 
  summarise(mean_weeks = sum(sop))

weeks_bad_qol_nonPO
```

If the treatment has no effect on mortality, the effect of an OR of 3.3 is a loss of 6 weeks with good QoL.

### Create and export images for OR = 0.8

### State Occupancy Probabilities
```{r}
source(here("functions", "carry-death-forward.R"))

sample_cf_08 <- sample_08 |> 
  carry_death_forward() |> 
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment"))
  )

sops_line_08 <- sample_cf_08 |> 
  group_by(time, tx) |> 
  count(y) |> 
  mutate(
    total = sum(n),
    sop   = n / total,
    tx = factor(tx)) |>
  ungroup() |> 
  ggplot(aes(x = time, y = sop, color = y)) +
  geom_line(aes(linetype = tx)) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.1)) +
  coord_cartesian(ylim = c(0, 0.4)) +
  labs(x = "Study Week",
   y = "SOP",
   subtitle = "Empirical state occupancy probabilities (OR = 0.8)",
   linetype = "Treatment",
   color = "State")

sops_stacked_08 <- sample_cf_08 |>  
  ggplot(aes(x = time, fill = y)) +
  geom_bar(position = "fill")  +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.2)) +
  facet_wrap(~tx) +
  labs(x = "Study Week",
   y = "Cumulative Proportion",
   subtitle = "Distribution of ordinal states over time (OR = 0.8)",
   color = "State")

sops_line_08 / sops_stacked_08 +
  plot_annotation(tag_level = 'A', 
                  title = "PO Model - OR = 0.8")

ggsave(filename = here("img", "fig-sim-data-sop.svg"), width = 12, height = 10)
```

```{r}
# State Occupancy Probabilities - nonPO model
sample_nonPO_cf_08 <- sample_nonPO_08 |> 
  carry_death_forward() |> 
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment"))
  )

sops_line_nonPO_08 <- sample_nonPO_cf_08 |> 
  group_by(time, tx) |> 
  count(y) |> 
  mutate(
    total = sum(n),
    sop   = n / total,
    tx = factor(tx)) |>
  ungroup() |> 
  ggplot(aes(x = time, y = sop, color = y)) +
  geom_line(aes(linetype = tx)) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.1)) +
  coord_cartesian(ylim = c(0, 0.4)) +
  labs(x = "Study Week",
   y = "SOP",
   subtitle = "Empirical state occupancy probabilities (OR = 0.8, nonPO)",
   linetype = "Treatment",
   color = "State")

sops_stacked_nonPO_08 <- sample_nonPO_cf_08 |>  
  ggplot(aes(x = time, fill = y)) +
  geom_bar(position = "fill")  +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.2)) +
  facet_wrap(~tx) +
  labs(x = "Study Week",
   y = "Cumulative Proportion",
   subtitle = "Distribution of ordinal states over time (OR = 0.8, nonPO)",
   color = "State")

sops_line_nonPO_08 / sops_stacked_nonPO_08 +
  plot_annotation(tag_level = 'A',
                  title = "NonPO Model - OR = 0.8")

ggsave(filename = here("img", "fig-sim-data-sop-nonPO.svg"), width = 12, height = 10)
```

### Individual trajectories

```{r}
set.seed(4)
ids_08 <- sample_08$id |> unique() |> sample(20) 

fig_tile_random_08 <- sample_08 |> 
  filter(id %in% ids_08) |> 
  mutate(id = as.factor(as.numeric(as.factor((id))))) |> 
  ggplot(aes(y = id, x = time)) +
  geom_tile(mapping = aes(fill = y), width = 0.95, height = 0.4) +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = 1:26) +
  labs(title = "Individual Trajectories - OR = 0.8",
       y = "Patient ID",
       x = "Study Week",
       fill = "State")

ggsave(filename = here("img", "fig-sim-tile.svg"), width = 12, height = 6)
```

## Export simulataed datasets

```{r}
write_parquet(sample_08, here("1-qol", "sim-data", "sim_data_po.parquet"))
write_parquet(carried_forward_08, here("1-qol", "sim-data", "sim_data_po_forward.parquet"))
write_parquet(sample_nonPO_08, here("1-qol", "sim-data", "sim_data_nonpo.parquet"))
```
