---
title: QoL - Primary Estimand
format: 
  typst:
    papersize: a4
    margin:
      x: 2cm
      y: 2.5cm
    toc: false
    columns: 1
    tbl-cap-location: bottom
---

Here we analyse a random sample of the simulated data as we would analyze the true data of the liqplat trial.

## Setup

```{r}
#| output: false
library(arrow)
library(here)
library(tidyverse)
library(data.table)
library(rmsb)
library(ggplot2)
library(patchwork)
library(furrr)
library(markov.misc)
```

## Load data
```{r}
set.seed(1234)
# Load full dataset
full_data <- open_dataset(here::here("1-qol", "sim-data", "super_pop_ha_effect_1.96.parquet"))
generate_SOPS <- FALSE
fit <- FALSE
```

## Sample and format data
```{r}
# --- 2. Sample Patients ---
# we expect 20% of patients to die before week 26
ss_tx <- round(180*0.8)
ss_control <- round(90*0.8)

tx_ids <- sample(full_data |> filter(tx == 1) |> pull(id) |> unique(), ss_tx, replace = FALSE)
control_ids <- sample(full_data |> filter(tx == 0) |> pull(id) |> unique(), ss_control, replace = FALSE)

df <- full_data |>
    filter(id %in% c(tx_ids, control_ids)) |> 
    collect()

# difference in this sample happens to be 3.88 weeks
df |> calc_time_in_state_diff(target_state = 1:3)

# --- 3. Pre-process Data & Save Empirical SOPs ---
# Save empirical SOP image for this iteration's data
p_empirical <- plot_sops(df, linetype_var = "tx", facet_var = NULL, geom = "line") +
    scale_color_brewer(palette = "Dark2") +
    coord_cartesian(ylim = c(0, 0.5))


# --- 4. Create analysis dataset with missing follow-up ---
baseline <- df |>
    filter(time == 1) |>
    mutate(y = yprev, time = 0) |>
    select(-yprev)

follow_up <- df |>
    slice_sample(prop = 0.15) |>
    select(-yprev)

data_for_model <- baseline |>
    bind_rows(follow_up) |>
    arrange(id, time) |>
    group_by(id) |>
    mutate(
        yprev = lag(y, n = 1, order_by = time),
        yprev = factor(yprev, levels = 1:7, ordered = FALSE),
        timeprev = lag(time, n = 1, order_by = time),
        gap = time - timeprev,
        y = as.numeric(as.character(y))
    ) |>
    ungroup() |>
    filter(!is.na(yprev)) |>
    select(-timeprev)
```

## Visualize the follow-up pattern

```{r}
set.seed(12345)
ids <- data_for_model$id |> unique() |> sample(20) 

fig_tile_random <- data_for_model |> 
  filter(id %in% ids) |> 
  mutate(id = as.factor(as.numeric(as.factor((id))))) |> 
  ggplot(aes(y = id, x = time)) +
  geom_tile(mapping = aes(fill = factor(y)), width = 0.95, height = 0.4) +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = 1:26) +
  labs(fill = "y")

ggsave(filename = here::here("img", "fig-sparse-tile.svg"), width = 12, height = 6)
```

## Fit the model

```{r}
# brm_m <- brm(
#   formula = bf(
#     y ~ tx + s(time, bs = "cr") + yprev + (1 + tx | id)),
#   data = data_for_model,
#   family = cumulative(link = "logit"),
#   chains = 4, 
#   cores = 1,
#   iter = 1000
# )
```

```{r}
if(fit){
dd <- datadist(data_for_model)
# Set the global option so that rms/rmsb functions know to use 'dd'
options(datadist = 'dd')
options(mc.cores = 8)

Sys.setenv(RSTUDIO = 1)

pcontrast_arg <- list(
  c1 = list(tx = 1),
  c2 = list(tx = 0),
  contrast = expression(c1 - c2),
  mu = 0,
  sd = 0.5
)

model <-
    blrm(
      formula = y  ~ tx + rcs(time, 4) + yprev * gap,
      pcontrast = pcontrast_arg,
      data = data_for_model,
      ppo = ~ time,
      cppo = function(y) y,
      refresh = 5,
      iter = 2000,
      chains = 4,
      loo = FALSE,
      method = "sampling"
  )

stanDx(model)
stanDxplot(model)

Sys.setenv(RSTUDIO = 0)

saveRDS(model, here::here("1-qol", "models", "model-primary-estimand.rds"))} else{
  model <- readRDS(here::here("1-qol", "models", "model-primary-estimand.rds"))
}
```

### Recovering difference in weeks with >= good QoL

#### Marginalized State Occupancy Probabilities

```{r}
baseline_df <- data_for_model |> 
    group_by(id) |> 
    arrange(time) |> 
    slice_head(n = 1) |> 
    ungroup() |> 
    mutate(
        time = 1,
        gap = 1,
        y = NA) |> 
    as.data.table()
```

```{r}
write_SOP <- function(i, tx, baseline_df, path){
  
  row <- baseline_df[i]
  
  sops <-
  soprobMarkovOrdm(
    model,
    data = list(tx = tx, yprev=row$yprev, gap = row$gap),
    times = 1:26,
    ylevels = 1:7,
    absorb = NULL,
    tvarname = "time",
    pvarname = "yprev"
  )
  
  # Because of the size of the data, only use the first 500 MCMC draws
  sops <- sops[1:2000,,]
  
  sops <- as.data.table(sops)

  # Rename columns
  setnames(sops,
           old = c("V1", "V2", "V3", "value"),
           new = c("draw", "time", "state", "sop"))
  
  sops$time <- as.integer(sops$time)
  sops$state <- as.factor(sops$state)
  sops$tx <- tx
  sops$i <- i
  
  folder <- file.path(path, glue::glue("marginalized_sop_{tx}"), glue::glue("msop_{i}.parquet"))

  # Create the directory if it doesn't exist
  dir_to_create <- dirname(folder)
  if (!dir.exists(dir_to_create)) {
    dir.create(dir_to_create, recursive = TRUE)
  }
  
  arrow::write_parquet(
    x = sops,
    sink = folder
    )
}
```

#### Generate marginalized SOPs

```{r}
path <- here::here("1-qol", "output", "primary-estimand")
future::plan(future::multisession, workers=6)


# Control
if (generate_SOPS) {
    furrr::future_walk(1:nrow(baseline_df), \(x) write_SOP(x, 0, baseline_df, path),
                       .progress=TRUE,
                       .options = furrr_options(seed=TRUE))
}
# Treatment
if (generate_SOPS) {
    furrr::future_walk(1:nrow(baseline_df), \(x) write_SOP(x, 1, baseline_df, path),
                       .progress=TRUE,
                       .options = furrr_options(seed=TRUE))
}
```

#### Read maginalized SOPs

```{r}
# Get list of all files
files <- fs::dir_ls(here::here("1-qol", "output", "primary-estimand", "marginalized_sop_0"), glob = "*.parquet")

# Read in each parquet file as a data.frame and merge into a data.table
soc_df <- map(files, \(x) arrow::read_parquet(x), .progress=TRUE) |> rbindlist()

# Average SOPs over covariate settings -- for each state, day, and MCMC draw
soc_df <- soc_df[, .(sop = mean(sop)), by = .(state, time, draw, tx)]
soc_df[, state := as.factor(state)]
```

```{r}
# Get list of all files
files <- fs::dir_ls(here::here("1-qol", "output", "primary-estimand", "marginalized_sop_1"), glob = "*.parquet")

# Read in each parquet file as a data.frame and merge into a data.table
tx_df <- map(files, \(x) arrow::read_parquet(x), .progress=TRUE) |> rbindlist()

# Average SOPs over covariate settings -- for each state, day, and MCMC draw
tx_df <- tx_df[, .(sop = mean(sop)), by = .(state, time, draw, tx)]
tx_df[, state := as.factor(state)]
```

```{r}
sop_df <-
  bind_rows(soc_df, tx_df) |>
  mutate(tx = as.factor(tx))
```

### Plotting

#### Marginalized SOPs

```{r}
#| fig-width: 10
#| fig-height: 6
fig_marginalized <-
sop_df |>
ungroup() |> 
  ggplot() +
  aes(x = time, y = sop) +
  ggdist::stat_lineribbon(aes(fill = state, linetype = tx),
                          linewidth = 0.6,
                          alpha = 0.4,
                          .width = 0.95) +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.1)) +
  coord_cartesian(ylim = c(0, 0.5)) +
  labs(x = "Study Week",
   y = "SOP",
   subtitle = "Marginalized state occupancy probabilities",
   fill = "State",
   linetype = "Treatment")

ggsave(plot = fig_marginalized, here::here("img", "qol-primary-estimand-sop-line.svg"), width = 10, height = 6)


fig_marginalized_grouped <- sop_df |> 
  ungroup() |> 
  mutate(group = case_when(
    state == 1 | state == 2 | state == 3 ~ "Good",
    state == 7 | state == 6 | state == 5 ~ "Bad",
    TRUE ~ "Average"
  )) |> 
  group_by(group, draw, tx, time) |> 
  summarise(sop = sum(sop)) |> 
  ggplot() +
  aes(x = time, y = sop) +
  ggdist::stat_lineribbon(aes(fill = group, linetype = tx),
                          linewidth = 0.6,
                          alpha = 0.5,
                          .width = c(0.95)) +
  scale_color_brewer(palette = "Dark2", guide="none") +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq(1, 26, by=4)) +
  scale_y_continuous(breaks = seq(0, 1, by=0.1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Study Week",
   y = "SOP",
   subtitle = "Marginalized state occupancy probabilities",
   fill = "State",
   linetype = "Selected for Invitation to SAT") +
  theme(
    legend.position = "bottom"
  )

ggsave(plot = fig_marginalized_grouped, here::here("img", "qol-primary-estimand-sop-line-grouped.svg"), width = 10, height = 6)

p_empirical + fig_marginalized + plot_layout(guides = "collect")
```

#### Difference in Weeks with >= Good Quality of Life

In the large dataset the true difference between tx == 1 and tx == 0 is 1.5 weeks.
```{r}
#| fig-width: 8
#| fig-height: 5
# marginalized difference in weeks for >= good QoL
posterior <- sop_df |> 
  filter(state == "1" | state == "2" | state == "3") |> 
  group_by(tx, draw) |> 
  summarise(mean_weeks = sum(sop))  |> 
  pivot_wider(names_from = tx, values_from = mean_weeks, names_prefix = "tx_") |> 
  mutate(difference = tx_1 - tx_0)

# P benefit based on maringalized Sops
mean(posterior$difference > 0)

# P benefit based on coefficient (almost)
mean(model$draws[, 7] < 0)


plot_posterior <- posterior |> 
  ggplot(aes(x = difference)) +
  ggdist::stat_dots(side = "bottom") +
  ggdist::stat_halfeye() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
  x = bquote(paste("Treatment Difference in Weeks spent with ", "" >= "", " Good Quality of Life (SAT - UC)")),
  y = "Density"
)

plot_posterior

ggsave(here::here("img", "qol-primary-estimand-posterior.svg"), width = 8, height = 5)
```