---
title: "Bayesian Power Simulation Pipeline - QoL Analysis"
---

This document implements a Bayesian power simulation to compare three analysis 
strategies for Quality of Life outcomes:
  
1. **Markov Model** (`blrm`): Longitudinal ordinal model using all repeated measures
2. **6-Month Ordinal** (`blrm`): Cross-sectional ordinal model at 6 months only  
3. **Change from Baseline** (`brms`): Linear model for change from baseline at 6 months

All models use comparably skeptical priors centered at 0 for the treatment effect.

**Scenarios:**

- **H0**: effect_size = -0.05 (null hypothesis, treatment rerandomized)
- **HA**: effect_size = 0.98, 1.96, 2.96 (alternative hypotheses, no rerandomization)

The power for the 6-month ordinal models is an overestimate, because this analysis 
assumes we would have data from every patient at 6 months, which in routine care 
we would not.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(here)
library(arrow)
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(furrr)
library(fs)
library(glue)
library(data.table)
library(rmsb)
library(brms)
library(markov.misc)

# Configuration
N_SIMULATIONS <- 96
SAMPLE_SIZE <- 270 * 0.8       # Total patients to sample per iteration (20% die)
ALLOCATION_RATIO <- 2/3        # Proportion allocated to treatment (2:1 ratio)
N_CORES <- 6
SEED <- 123456
DECISION_THRESHOLD <- 0.95    # P(benefit) threshold for declaring success

# Paths
DATA_PATH <- here("1-qol", "sim-data")
OUTPUT_PATH <- here("1-qol", "output", "bayesian-power-sim")
MODEL_PATH <- here("1-qol", "models")

# Create output directory if needed
if (!dir.exists(OUTPUT_PATH)) dir.create(OUTPUT_PATH, recursive = TRUE)
```

## Step 1: Identify Scenarios

We scan for available superpopulation files and classify them as H0 or HA.

```{r identify-scenarios}
# Get all superpopulation files
h0_files <- list.files(DATA_PATH, pattern = "super_pop_h0_effect_.*\\.parquet", full.names = TRUE)
ha_files <- list.files(DATA_PATH, pattern = "super_pop_ha_effect_.*\\.parquet", full.names = TRUE)

# Extract effect sizes
h0_effect_sizes <- as.numeric(gsub(".*super_pop_h0_effect_([0-9.-]+)\\.parquet", "\\1", h0_files))
ha_effect_sizes <- as.numeric(gsub(".*super_pop_ha_effect_([0-9.-]+)\\.parquet", "\\1", ha_files))

# Create scenario tibble
scenarios <- bind_rows(
  tibble(
    scenario_type = "H0",
    effect_size = h0_effect_sizes,
    data_path = h0_files,
    rerandomize = TRUE,
    true_value = 0  # For H0, true effect under null is 0
  ),
  tibble(
    scenario_type = "HA",
    effect_size = ha_effect_sizes,
    data_path = ha_files,
    rerandomize = FALSE,
    true_value = NA_real_  # True value unknown for HA scenarios
  )
) |>
  mutate(
    scenario_label = glue("{scenario_type}_effect_{effect_size}"),
    output_path = file.path(OUTPUT_PATH, scenario_label)
  )

scenarios
```

## Step 2: Define Data Preparation Functions

```{r data-prep-functions}
#' Prepare data for Markov model (longitudinal analysis)
#' 
#' @param df Full dataset with repeated measures
#' @param sample_percentage Proportion of follow-up observations to sample
#' @return Data ready for blrm() Markov model
prepare_markov_data <- function(df, sample_percentage = 0.15) {
  # Create baseline observation
  baseline <- df |>
    filter(time == 1) |>
    mutate(y = yprev, time = 0) |>
    select(-yprev)
  
  # Sample follow-up observations (simulating sparse measurement)
  follow_up <- df |>
    slice_sample(prop = sample_percentage) |>
    select(-yprev)
  
  # Combine and create lagged variables
  data_for_model <- baseline |>
    bind_rows(follow_up) |>
    arrange(id, time) |>
    group_by(id) |>
    mutate(
      yprev = lag(y, n = 1, order_by = time),
      yprev = factor(yprev, levels = 1:7, ordered = FALSE),
      timeprev = lag(time, n = 1, order_by = time),
      gap = time - timeprev,
      y = as.numeric(as.character(y))
    ) |>
    ungroup() |>
    filter(!is.na(yprev)) |>
    select(-timeprev)
  
  data_for_model
}

#' Prepare data for 6-month ordinal model (cross-sectional)
#' 
#' @param df Full dataset with repeated measures
#' @return Data ready for blrm() ordinal model at 6 months
prepare_6m_ordinal_data <- function(df) {
  # Get baseline for each patient
  baseline <- df |>
    group_by(id) |>
    arrange(time) |> 
    slice_head(n = 1) |>
    mutate(ybaseline = as.numeric(as.character(yprev))) |>
    select(id, ybaseline)
  
  # Get last observation after week 14 (approximately 6 months)
  data_for_model <- df |>
    filter(time > 14) |>
    group_by(id) |>
    arrange(desc(time)) |>
    slice_head(n = 1) |>
    ungroup() |>
    left_join(baseline, by = "id") |>
    mutate(
      ybaseline = factor(ybaseline, levels = 1:7, ordered = FALSE),
      y = as.numeric(as.character(y))
    )
  
  data_for_model
}

#' Prepare data for change-from-baseline model
#' 
#' @param df Full dataset with repeated measures
#' @return Data ready for brms linear model
prepare_cfb_data <- function(df) {
  # Get baseline for each patient
  baseline <- df |>
    group_by(id) |>
    arrange(time) |> 
    slice_head(n = 1) |>
    mutate(ybaseline = as.numeric(as.character(yprev))) |>
    select(id, ybaseline)
  
  # Get last observation after week 14
  data_for_model <- df |>
    filter(time > 14) |>
    group_by(id) |>
    arrange(desc(time)) |>
    slice_head(n = 1) |>
    ungroup() |>
    left_join(baseline, by = "id") |>
    mutate(
      y = as.numeric(as.character(y)),
      y_delta = y - ybaseline
    )
  
  data_for_model
}
```

## Step 3: Load Pre-compiled BRMS Template

```{r load-brms-template}
brms_template <- readRDS(file.path(MODEL_PATH, "cfb_brms_template.rds"))
```

## Step 4: Define Bayesian Fitting Functions

Each function fits a Bayesian model and returns results in the format expected 
by `assess_operating_characteristics()`: a list where the first element is a 
summary tibble with required columns.

For Bayesian models, we use:
- `estimate`: posterior median
- `std_error`: posterior SD
- `statistic`: NA (not applicable)
- `p_value`: 1 - P(benefit) to work with standard "reject if p < alpha" logic
- `conf_low`, `conf_high`: 95% credible interval

The posterior draws are saved in the details for later analysis.

```{r fitting-functions}
#' Fit Markov model using blrm
#' 
#' @param data Sampled data from superpopulation
#' @param iter Iteration number
#' @return List with summary tibble and posterior draws
fit_markov_blrm <- function(data, iter) {
  # Prepare data for Markov model
  markov_data <- prepare_markov_data(data, sample_percentage = 0.15)
  
  # Set up datadist
  dd <- datadist(markov_data)
  assign("dd", dd, envir = .GlobalEnv)
  options(datadist = "dd")
  
  # Skeptical prior: normal(0, 0.5) on log-OR scale
  pcontrast_arg <- list(
    c1 = list(tx = 1),
    c2 = list(tx = 0),
    contrast = expression(c1 - c2),
    mu = 0,
    sd = 0.5
  )
  
  model <- tryCatch({
    blrm(
      formula = y ~ tx + rcs(time, 4) + yprev * gap,
      pcontrast = pcontrast_arg,
      data = markov_data,
      ppo = ~time,
      cppo = function(y) y,
      refresh = 0,
      iter = 2000,
      chains = 4,
      method = "sampling",
      seed = 1,
      cores = 1,
      loo = FALSE
    )
  }, error = function(e) {
    message(glue("ERROR in Markov iteration {iter}: {e$message}"))
    return(NULL)
  })
  
  if (is.null(model)) {
    return(list(
      tibble(
        iter = iter,
        analysis = "markov_blrm",
        se_type = "bayesian",
        term = "tx",
        estimate = NA_real_,
        std_error = NA_real_,
        statistic = NA_real_,
        p_value = NA_real_,
        conf_low = NA_real_,
        conf_high = NA_real_
      )
    ))
  }
  
  # Extract tx posterior
  tx_col <- which(colnames(model$draws) == "tx")
  if (length(tx_col) == 0) tx_col <- 7
  tx_draws <- as.numeric(model$draws[, tx_col])
  
  # Calculate P(benefit): lower is better (log-OR < 0 means treatment helps)
  p_benefit <- mean(tx_draws < 0)
  
  # Create summary in required format
  summary_tbl <- tibble(
    iter = iter,
    analysis = "markov_blrm",
    se_type = "bayesian",
    term = "tx",
    estimate = median(tx_draws),
    std_error = sd(tx_draws),
    statistic = NA_real_,
    p_value = 1 - p_benefit,  # Convert so p < 0.05 means success
    conf_low = quantile(tx_draws, 0.025),
    conf_high = quantile(tx_draws, 0.975)
  )
  
  # Return list: summary + details (posterior draws)
  list(
    summary_tbl,
    list(
      posterior_draws = tibble(
        iter = iter,
        analysis = "markov_blrm",
        draw_id = seq_along(tx_draws),
        tx_effect = tx_draws
      ),
      p_benefit = p_benefit
    )
  )
}

#' Fit 6-month ordinal model using blrm
#' 
#' @param data Sampled data from superpopulation
#' @param iter Iteration number
#' @return List with summary tibble and posterior draws
fit_6m_ordinal_blrm <- function(data, iter) {
  # Prepare data for 6-month ordinal model
  ordinal_data <- prepare_6m_ordinal_data(data)
  
  ordinal_data <- ordinal_data |> 
    mutate(ybaseline = factor(ybaseline, ordered = FALSE))
  
  dd <- datadist(ordinal_data)
  assign("dd", dd, envir = .GlobalEnv)
  options(datadist = "dd")
  
  # Skeptical prior: normal(0, 0.5) on log-OR scale
  pcontrast_arg <- list(
    c1 = list(tx = 1),
    c2 = list(tx = 0),
    contrast = expression(c1 - c2),
    mu = 0,
    sd = 0.5
  )
  
  model <- tryCatch({
    blrm(
      formula = y ~ tx + ybaseline,
      pcontrast = pcontrast_arg,
      data = ordinal_data,
      refresh = 0,
      iter = 2000,
      chains = 4,
      method = "sampling",
      seed = 1,
      cores = 1,
      loo = FALSE
    )
  }, error = function(e) {
    message(glue("ERROR in 6m ordinal iteration {iter}: {e$message}"))
    return(NULL)
  })
  
  if (is.null(model)) {
    return(list(
      tibble(
        iter = iter,
        analysis = "ordinal_6m_blrm",
        se_type = "bayesian",
        term = "tx",
        estimate = NA_real_,
        std_error = NA_real_,
        statistic = NA_real_,
        p_value = NA_real_,
        conf_low = NA_real_,
        conf_high = NA_real_
      )
    ))
  }
  
  # Extract tx posterior
  tx_col <- which(colnames(model$draws) == "tx")
  if (length(tx_col) == 0) tx_col <- 7
  tx_draws <- as.numeric(model$draws[, tx_col])
  
  p_benefit <- mean(tx_draws < 0)
  
  summary_tbl <- tibble(
    iter = iter,
    analysis = "ordinal_6m_blrm",
    se_type = "bayesian",
    term = "tx",
    estimate = median(tx_draws),
    std_error = sd(tx_draws),
    statistic = NA_real_,
    p_value = 1 - p_benefit,
    conf_low = quantile(tx_draws, 0.025),
    conf_high = quantile(tx_draws, 0.975)
  )
  
  list(
    summary_tbl,
    list(
      posterior_draws = tibble(
        iter = iter,
        analysis = "ordinal_6m_blrm",
        draw_id = seq_along(tx_draws),
        tx_effect = tx_draws
      ),
      p_benefit = p_benefit
    )
  )
}

#' Fit change-from-baseline model using brms
#' 
#' @param data Sampled data from superpopulation
#' @param iter Iteration number
#' @return List with summary tibble and posterior draws
fit_cfb_brms <- function(data, iter) {
  # Prepare data for CFB model
  cfb_data <- prepare_cfb_data(data)
  
  model <- tryCatch({
    update(
      brms_template,
      newdata = cfb_data,
      refresh = 0,
      seed = 1,
      cores = 1
    )
  }, error = function(e) {
    message(glue("ERROR in CFB iteration {iter}: {e$message}"))
    return(NULL)
  })
  
  if (is.null(model)) {
    return(list(
      tibble(
        iter = iter,
        analysis = "cfb_brms",
        se_type = "bayesian",
        term = "tx",
        estimate = NA_real_,
        std_error = NA_real_,
        statistic = NA_real_,
        p_value = NA_real_,
        conf_low = NA_real_,
        conf_high = NA_real_
      )
    ))
  }
  
  # Extract tx posterior
  draws_df <- brms::as_draws_df(model)
  tx_draws <- draws_df$b_tx
  
  p_benefit <- mean(tx_draws < 0)
  
  summary_tbl <- tibble(
    iter = iter,
    analysis = "cfb_brms",
    se_type = "bayesian",
    term = "tx",
    estimate = median(tx_draws),
    std_error = sd(tx_draws),
    statistic = NA_real_,
    p_value = 1 - p_benefit,
    conf_low = quantile(tx_draws, 0.025),
    conf_high = quantile(tx_draws, 0.975)
  )
  
  list(
    summary_tbl,
    list(
      posterior_draws = tibble(
        iter = iter,
        analysis = "cfb_brms",
        draw_id = seq_along(tx_draws),
        tx_effect = tx_draws
      ),
      p_benefit = p_benefit
    )
  )
}
```

## Step 5: Run Simulations for All Scenarios

```{r run-simulations}
# Define fit functions list - all use the same data path since they're 
# applied to the same sampled dataset
fit_functions <- list(
  markov_blrm = fit_markov_blrm,
  ordinal_6m_blrm = fit_6m_ordinal_blrm,
  cfb_brms = fit_cfb_brms
)

# Run simulations for each scenario
for (i in seq_len(nrow(scenarios))) {
  scenario <- scenarios[i, ]
  
  message(glue("\n=== Running scenario: {scenario$scenario_label} ==="))
  
  # Create output directory
  if (!dir.exists(scenario$output_path)) {
    dir.create(scenario$output_path, recursive = TRUE)
  }
  
  # Data paths for all analyses (same file)
  data_paths <- list(
    markov_blrm = scenario$data_path,
    ordinal_6m_blrm = scenario$data_path,
    cfb_brms = scenario$data_path
  )
  
  # Check for completed iterations
  summary_dir <- file.path(scenario$output_path, "summary")
  if (dir.exists(summary_dir)) {
    completed_files <- list.files(summary_dir, pattern = "_iter_\\d+\\.rds", 
                                   recursive = TRUE, full.names = TRUE)
    completed_iters <- as.integer(
      unique(gsub(".*_iter_(\\d+)\\.rds", "\\1", completed_files))
    )
  } else {
    completed_iters <- integer(0)
  }
  
  iters_to_run <- setdiff(1:N_SIMULATIONS, completed_iters)
  n_to_run <- length(iters_to_run)
  n_completed <- length(completed_iters)
  
  message(glue("{n_completed} of {N_SIMULATIONS} iterations already complete."))
  
  if (n_to_run > 0) {
    message(glue("Starting {n_to_run} new simulations on {N_CORES} cores..."))
    
    future::plan(future.callr::callr, workers = N_CORES)
    
    furrr::future_map(
      .x = iters_to_run,
      .f = ~ markov.misc::assess_operating_characteristics(
        iter_num = .x,
        data_paths = data_paths,
        output_path = scenario$output_path,
        fit_functions = fit_functions,
        sample_size = SAMPLE_SIZE,
        allocation_ratio = ALLOCATION_RATIO,
        seed = SEED,
        rerandomize = scenario$rerandomize
      ),
      .options = furrr_options(
        seed = TRUE,
        globals = c(
          # --- Helper Functions & Templates ---
          "prepare_markov_data", 
          "prepare_6m_ordinal_data", 
          "prepare_cfb_data", 
          "brms_template",
          
          # --- Configuration Constants ---
          "SEED",
          "SAMPLE_SIZE",
          "ALLOCATION_RATIO",
          "fit_functions",
          
          # --- Loop-Specific Variables ---
          "scenario",     # Needed for scenario$output_path and scenario$rerandomize
          "data_paths"    # Needed for the data_paths argument
        ),
        packages = c("rms", "rmsb", "brms", "dplyr", "tibble", "markov.misc")
      ),
      .progress = TRUE
    )
    
    future::plan("sequential")
    message(glue("Finished {n_to_run} simulations for {scenario$scenario_label}."))
  } else {
    message("All simulations already complete.")
  }
}

message("\n=== All scenarios completed ===")
```