---
Title: Results of Simulations
---

In this document we summarise and visualize the results of the simulations from step 6.

## Setup

### Load packages

```{r}
library(here)
library(ggplot2)
library(fs)
library(tidyverse)
library(markov.misc)
```

```{r}
DECISION_THRESHOLD <- 0.95 

DATA_PATH <- here("1-qol", "sim-data")
OUTPUT_PATH <- here("1-qol", "output", "bayesian-power-sim")
h0_files <- list.files(DATA_PATH, pattern = "super_pop_h0_effect_.*\\.parquet", full.names = TRUE)
ha_files <- list.files(DATA_PATH, pattern = "super_pop_ha_effect_.*\\.parquet", full.names = TRUE)

# Extract effect sizes
h0_effect_sizes <- as.numeric(gsub(".*super_pop_h0_effect_([0-9.-]+)\\.parquet", "\\1", h0_files))
ha_effect_sizes <- as.numeric(gsub(".*super_pop_ha_effect_([0-9.-]+)\\.parquet", "\\1", ha_files))

scenarios <- bind_rows(
  tibble(
    scenario_type = "H0",
    effect_size = h0_effect_sizes,
    data_path = h0_files,
    rerandomize = TRUE,
    true_value = 0  # For H0, true effect under null is 0
  ),
  tibble(
    scenario_type = "HA",
    effect_size = ha_effect_sizes,
    data_path = ha_files,
    rerandomize = FALSE,
    true_value = NA_real_  # True value unknown for HA scenarios
  )
) |>
  mutate(
    scenario_label = str_glue("{scenario_type}_effect_{effect_size}"),
    output_path = file.path(OUTPUT_PATH, scenario_label)
  )

scenarios
```

## Step 6: Aggregate Results Across All Scenarios

```{r aggregate-results}
#' Load and combine summary results from all iterations
#' 
#' @param output_path Path to scenario output directory
#' @return Combined tibble of all summary results
load_scenario_summaries <- function(output_path) {
  summary_dir <- file.path(output_path, "summary")
  
  if (!dir.exists(summary_dir)) {
    return(NULL)
  }
  
  summary_files <- list.files(summary_dir, pattern = "\\.rds$", 
                               recursive = TRUE, full.names = TRUE)
  
  if (length(summary_files) == 0) {
    return(NULL)
  }
  
  map_dfr(summary_files, readRDS)
}

#' Load p_benefit values from details files
#' 
#' @param output_path Path to scenario output directory
#' @return Tibble with iter, analysis, and p_benefit
load_p_benefit <- function(output_path) {
  details_dir <- file.path(output_path, "details")
  
  if (!dir.exists(details_dir)) {
    return(NULL)
  }
  
  details_files <- list.files(details_dir, pattern = "\\.rds$", 
                               recursive = TRUE, full.names = TRUE)
  
  if (length(details_files) == 0) {
    return(NULL)
  }
  
  map_dfr(details_files, function(f) {
    d <- readRDS(f)
    # Extract iteration from filename
    iter <- as.integer(gsub(".*_iter_(\\d+)\\.rds", "\\1", f))
    # Extract analysis from directory name
    analysis <- basename(dirname(f))
    
    tibble(
      iter = iter,
      analysis = analysis,
      p_benefit = d[[1]]$p_benefit
    )
  })
}

# Aggregate results for all scenarios
all_results <- map_dfr(seq_len(nrow(scenarios)), function(i) {
  scenario <- scenarios[i, ]
  
  summaries <- load_scenario_summaries(scenario$output_path)
  p_benefits <- load_p_benefit(scenario$output_path)
  
  if (is.null(summaries)) {
    return(NULL)
  }
  
  # Join p_benefit if available
  if (!is.null(p_benefits)) {
    summaries <- summaries |>
      left_join(p_benefits, by = c("iter", "analysis"))
  } else {
    # Calculate from p_value (p_value = 1 - p_benefit)
    summaries <- summaries |>
      mutate(p_benefit = 1 - p_value)
  }
  
  summaries |>
    mutate(
      scenario_type = scenario$scenario_type,
      effect_size = scenario$effect_size,
      scenario_label = scenario$scenario_label
    )
})

all_results <- all_results |> 
  mutate(effect_size = if_else(scenario_type == "H0", 0, effect_size)) # had to save probability of benefit in p-value for oper_char function. Is not p-value.
```

## Step 7: Summarize Operating Characteristics

```{r summarize-oc}
#' Summarize Bayesian operating characteristics with jackknife MCSE
#' 
#' @param results Combined results from all iterations
#' @param decision_threshold P(benefit) threshold for declaring success
#' @return Summary tibble with power and MCSE
summarize_bayesian_oc <- function(results, decision_threshold = 0.975) {
  results |>
    group_by(scenario_type, effect_size, scenario_label, analysis, term) |>
    summarise(
      n_iters = n(),
      mean_estimate = mean(estimate, na.rm = TRUE),
      sd_estimate = sd(estimate, na.rm = TRUE),
      mean_sd_posterior = mean(std_error, na.rm = TRUE),
      mean_p_benefit = mean(p_benefit, na.rm = TRUE),
      # Power/Type I error: proportion where P(benefit) > threshold
      power = mean(p_benefit > decision_threshold, na.rm = TRUE),
      # Jackknife MCSE for power
      power_mcse = markov.misc::jackknife_mcse(
        p_benefit, 
        statistic = function(x) mean(x > decision_threshold, na.rm = TRUE)
      ),
      # Coverage: proportion where 95% CI includes 0
      coverage_includes_null = mean(conf_low <= 0 & conf_high >= 0, na.rm = TRUE),
      .groups = "drop"
    )
}

# Generate summary
oc_summary <- summarize_bayesian_oc(all_results, decision_threshold = DECISION_THRESHOLD)

oc_summary |>
  select(scenario_type, effect_size, analysis, n_iters, 
         mean_estimate, power, power_mcse)
```

## Step 8: Visualize Power Across Effect Sizes

```{r power-plot}
#| fig-width: 10
#| fig-height: 6

if (nrow(oc_summary) > 0) {
  # Create power curve plot
  power_plot <- oc_summary |>
    mutate(
      analysis_label = case_when(
        analysis == "markov_blrm" ~ "Markov (longitudinal)",
        analysis == "ordinal_6m_blrm" ~ "Ordinal 6-month",
        analysis == "cfb_brms" ~ "Change from baseline",
        TRUE ~ analysis
      )
    ) |>
    ggplot(aes(x = effect_size, y = power, color = analysis_label, 
               group = analysis_label)) +
    geom_line(linewidth = 1) +
    geom_point(size = 3) +
    geom_errorbar(
      aes(ymin = pmax(0, power - 1.96 * power_mcse), 
          ymax = pmin(1, power + 1.96 * power_mcse)),
      width = 0.05,
      linewidth = 0.8
    ) +
    geom_hline(yintercept = 0.05, linetype = "dashed", color = "gray50") +
    geom_hline(yintercept = 0.80, linetype = "dotted", color = "gray50") +
    geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
    scale_y_continuous(
      limits = c(0, 1),
      breaks = seq(0, 1, 0.1),
      labels = scales::percent_format()
    ) +
    scale_color_brewer(palette = "Dark2") +
    labs(
      title = "Bayesian Power by Effect Size and Analysis Method",
      subtitle = glue("Decision threshold: P(benefit) > {DECISION_THRESHOLD}; ",
                      "Error bars: 95% CI"),
      x = "Effect Size (Difference in Weeks with Good QoL)",
      y = "Power / Type I Error Rate",
      color = "Analysis Method",
      caption = paste0(
        "Dashed line = 5% Type I error target;\n",
        "Dotted line = 80% power target"
      )
    ) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
  
  power_plot
  
  # Save plot
  ggsave(
    file.path(OUTPUT_PATH, "power_curve_by_effect_size.png"),
    power_plot,
    width = 10, height = 6, dpi = 150
  )
}
```

## Step 9: Summary Table

```{r summary-table}
if (nrow(oc_summary) > 0) {
  # Create formatted summary table
  summary_table <- oc_summary |>
    mutate(
      power_formatted = sprintf("%.1f%% (%.1f%%)", power * 100, power_mcse * 100),
      estimate_formatted = sprintf("%.3f (SD: %.3f)", mean_estimate, sd_estimate)
    ) |>
    select(
      Scenario = scenario_type,
      `Effect Size` = effect_size,
      Analysis = analysis,
      N = n_iters,
      `Mean Estimate (SD)` = estimate_formatted,
      `Power/T1E (MCSE)` = power_formatted
    )
  
  summary_table |>
    knitr::kable(digits = 3)
  
  # Save summary
  arrow::write_parquet(oc_summary, file.path(OUTPUT_PATH, "operating_characteristics_all.parquet"))
}
```

## Step 10: Posterior Distribution Comparison (Optional)

```{r posterior-comparison, eval=FALSE}
# Load posterior draws for detailed analysis
load_posterior_draws <- function(output_path) {
  details_dir <- file.path(output_path, "details")
  
  if (!dir.exists(details_dir)) return(NULL)
  
  details_files <- list.files(details_dir, pattern = "\\.rds$", 
                               recursive = TRUE, full.names = TRUE)
  
  map_dfr(details_files, function(f) {
    d <- readRDS(f)
    if (!is.null(d[[1]]$posterior_draws)) {
      d[[1]]$posterior_draws
    } else {
      NULL
    }
  })
}

# Example: Load and plot posterior draws for one scenario
if (nrow(scenarios) > 0) {
  example_scenario <- scenarios$output_path[2]
  draws <- load_posterior_draws(example_scenario)
  
  if (!is.null(draws) && nrow(draws) > 0) {
    draws |>
      filter(iter <= 10) |>  # Limit to first 10 iterations for clarity
      ggplot(aes(x = tx_effect, color = analysis)) +
      geom_density(alpha = 0.5) +
      # facet_wrap(~iter, scales = "free_y") +
      geom_vline(xintercept = 0, linetype = "dashed") +
      coord_cartesian(xlim = c(-1.5, 0.5)) +
      labs(
        title = "Posterior Distributions by Analysis Method",
        x = "Treatment Effect",
        y = "Density"
      ) +
      theme_minimal()
  }
}
```





















-------------------------------------------------

### Load Full Data

```{r}
df_full <- read_parquet("./1-qol/sim-data/sim_data_nonpo.parquet")
threshold <- 0.95
```

### Functions

```{r}
source("./functions/jackknife_mcse.R")
source("./functions/summarise-posterior.R")
source("./functions/summarise-iterations.R")
source("./functions/plot-rhat.R")
source("./functions/plot-p-benefit.R")
source("./functions/plot-posterior-distribution.R")
source("./functions/load-draws.R")
source("./functions/load-rhat-values.R")
```

## Markov Model

#### Load draws + diagnostics
```{r}
# H0
path <- "./1-qol/output/sim-principal-stratum-markov-H0/"
markov_h0 <- load_draws(path)
rhat_markov_h0 <- load_rhat_values(path = path)

#HA
path <- "./1-qol/output/sim-principal-stratum-markov-HA/"
markov_ha <- load_draws(path)
rhat_markov_ha <- load_rhat_values(path = path)
```

#### MCSE increase plot (power)

```{r}
iter_200 <- markov_ha |> 
    filter(iter <= 200) |> 
    group_by(iter) |> 
    summarise(p_benefit = mean(tx < 0)) |> 
    reframe(rejected  = as.numeric(p_benefit > threshold)) |> 
    summarise(`Monte Carlo SE for power based on posterior probability` = jackknife_mcse(rejected)) |> 
    mutate(`$n_\text{sim}$` = "200") |> 
    relocate(`$n_\text{sim}$`)


iter_400 <- markov_ha |> 
    filter(iter <= 400) |> 
    group_by(iter) |> 
    summarise(p_benefit = mean(tx < 0)) |> 
    reframe(rejected  = as.numeric(p_benefit > threshold)) |> 
    summarise(`Monte Carlo SE for power based on posterior probability` = jackknife_mcse(rejected)) |> 
    mutate(`$n_\text{sim}$` = "400")

iter_600 <- markov_ha |> 
    filter(iter <= 600) |> 
    group_by(iter) |> 
    summarise(p_benefit = mean(tx < 0)) |> 
    reframe(rejected  = as.numeric(p_benefit > threshold)) |> 
    summarise(`Monte Carlo SE for power based on posterior probability` = jackknife_mcse(rejected)) |> 
    mutate(`$n_\text{sim}$` = "600")

iter_800 <- markov_ha |> 
    filter(iter <= 800) |> 
    group_by(iter) |> 
    summarise(p_benefit = mean(tx < 0)) |> 
    reframe(rejected  = as.numeric(p_benefit > threshold)) |> 
    summarise(`Monte Carlo SE for power based on posterior probability` = jackknife_mcse(rejected)) |> 
    mutate(`$n_\text{sim}$` = "800")

iter_1000 <- markov_ha |> 
    filter(iter <= 1000) |> 
    group_by(iter) |> 
    summarise(p_benefit = mean(tx < 0)) |> 
    reframe(rejected  = as.numeric(p_benefit > threshold)) |> 
    summarise(`Monte Carlo SE for power based on posterior probability` = jackknife_mcse(rejected)) |> 
    mutate(`$n_\text{sim}$` = "1000")

mcse_evolution <- bind_rows(iter_200, iter_400, iter_600, iter_800, iter_1000)

saveRDS(mcse_evolution, file = "./1-qol/output/mcse-evolution.rds")
```

#### Convergence diagnostics
```{r}
rhat_plot_markov_ha <- rhat_markov_ha |> 
    plot_rhat()

rhat_plot_markov_h0 <- rhat_markov_h0 |> 
    plot_rhat()

rhat_plot_markov_h0 + rhat_plot_markov_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-rhat-markov.svg", width = 10, height = 5)
```

### Plot probability of benefit

```{r}
p1 <- markov_h0 |> 
    plot_p_benefit(param = tx)

p2 <- markov_ha |> 
    plot_p_benefit(param = tx)

p1 + p2 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-pbenefit-markov.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_markov_h0 <- markov_h0 |> 
  plot_posterior_distribution(x_variable = tx)

post_markov_ha <- markov_ha |> 
  plot_posterior_distribution(
    x_variable = tx, 
    subtitle_type = "HA", 
    hline_intercept = log(0.8))

post_markov_h0 + post_markov_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-posteriors-markov.svg", width = 10, height = 5)
```

```{r}
summary_markov_h0 <- markov_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = tx, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_markov_ha <- markov_ha |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = tx, 
        true_param_value = log(0.8), 
        threshold = 0.95) |> 
    summarise_iterations()

rm(markov_h0, markov_ha)
```

## Multilevel Model with RI

#### Load draws
```{r}
path <- "./1-qol/output/sim-principal-stratum-random-intercept-H0/"

ri_h0 <- load_draws(path)
rhat_ri_h0 <- load_rhat_values(path)

path <- "./1-qol/output/sim-principal-stratum-random-intercept-HA/"

ri_ha <- load_draws(path)
rhat_ri_ha <- load_rhat_values(path)
```

#### Plot

```{r}
# probability of benefit over time
p1 <- ri_h0 |> 
    plot_p_benefit(param = b_tx)

p2 <- ri_ha |> 
    plot_p_benefit(param = b_tx)

p1 + p2 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-pbenefit-ri.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_ri_h0 <- ri_h0 |> 
  plot_posterior_distribution(x_variable = b_tx)

post_ri_ha <- ri_ha |> 
  plot_posterior_distribution(
    x_variable = b_tx, 
    subtitle_type = "HA", 
    hline_intercept = log(0.8))

post_ri_h0 + post_ri_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-posteriors-ri.svg", width = 10, height = 5)
```

#### Converge diagnostics

```{r}
rhat_plot_ri_h0 <- rhat_ri_h0 |> 
    plot_rhat(b_tx)

rhat_plot_ri_ha <- rhat_ri_ha |> 
    plot_rhat(b_tx)

rhat_plot_ri_ha + rhat_plot_ri_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-rhat-ri.svg", width = 10, height = 5)
```

#### MCSE

```{r}
summary_ri_h0 <- ri_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_ri_ha <- ri_ha |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = log(0.8), 
        threshold = 0.95) |> 
    summarise_iterations()

rm(ri_h0, ri_ha)
```

## Change From Baseline Model

#### Load draws
```{r}
path <- "./1-qol/output/sim-principal-stratum-change-from-baseline-H0/"

cfb_h0 <- load_draws(path)
rhat_cfb_h0 <- load_rhat_values(path)

path <- "./1-qol/output/sim-principal-stratum-change-from-baseline-HA/"

cfb_ha <- load_draws(path)
rhat_cfb_ha <- load_rhat_values(path)

# 'true' change from baseline needs to be approximated from large dataset
mean_change_empirical <- df_full |> 
    filter(time == 1 | time == 26) |>
    mutate(y = if_else(time == 1, yprev, y)) |> 
    select(id, tx, time, y) |> 
    pivot_wider(names_from = time, values_from = y, names_prefix = "y_") |> 
    filter(y_26 != "8") |> # restrict to survivors
    mutate(delta_y = as.numeric(as.character(y_26)) - as.numeric(as.character(y_1))) |> 
    group_by(tx) |> 
    summarise(mean_change = mean(delta_y))

true_diff <- mean_change_empirical$mean_change[2] - mean_change_empirical$mean_change[1]
```

#### Converge diagnostics

```{r}
rhat_plot_cfb_h0 <- rhat_cfb_h0 |> 
    plot_rhat(b_tx)

rhat_plot_cfb_ha <- rhat_cfb_ha |> 
    plot_rhat(b_tx)

rhat_plot_cfb_h0 + rhat_plot_cfb_h0 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-rhat-cfb.svg", width = 10, height = 5)
```

```{r}
p1 <- cfb_h0 |> 
    plot_p_benefit(param = b_tx)

p2 <- cfb_ha |> 
    plot_p_benefit(param = b_tx)

p1 + p2 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-pbenefit-cfb.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_cfb_h0 <- cfb_h0 |> 
  plot_posterior_distribution(x_variable = b_tx)

post_cfb_ha <- cfb_ha |> 
  plot_posterior_distribution(
    x_variable = b_tx, 
    subtitle_type = "HA", 
    hline_intercept = true_diff)

post_cfb_h0 + post_cfb_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-posteriors-cfb.svg", width = 10, height = 5)
```

```{r}
summary_cfb_h0 <- cfb_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = 0, 
        threshold = 0.95) |> 
    summarise_iterations()

summary_cfb_ha <- cfb_ha |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = true_diff, 
        threshold = 0.95) |> 
    summarise_iterations()

rm(cfb_h0, cfb_ha)
```

## Time to Deterioation

#### Load data

```{r}
path <- "./1-qol/output/sim-principal-stratum-time-to-deterioration-H0/"

ttd_h0 <- load_draws(path)
rhat_ttd_h0 <- load_rhat_values(path)

path <- "./1-qol/output/sim-principal-stratum-time-to-deterioration-HA/"

ttd_ha <- load_draws(path)
rhat_ttd_ha <- load_rhat_values(path)
```

#### Plots

```{r}
p1 <- ttd_h0 |> 
    plot_p_benefit(param = b_tx)

p2 <- ttd_ha |> 
    plot_p_benefit(param = b_tx)

p1 + p2 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-pbenefit-ttd.svg", width = 10, height = 5)
```

```{r}
rhat_plot_ttd_ha <- rhat_ttd_ha |> 
    plot_rhat(b_tx)

rhat_plot_ttd_h0 <- rhat_ttd_h0 |> 
    plot_rhat(b_tx)

rhat_plot_ttd_h0 + rhat_plot_ttd_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-rhat-ttd.svg", width = 10, height = 5)
```

#### MCSE

```{r}
# we need to approximate 'true' HR from large dataset

ids_death <- df_full$id[df_full$y == '8']

df_filtered <- df_full |>
        filter(!(id %in% ids_death))

baseline <- df_filtered |>
        filter(time == 1) |>
        mutate(y = yprev, time = 0) |>
        select(id, ybaseline = y)

deteriorated <- df_filtered |> 
    select(-yprev) |> 
    left_join(baseline, by = "id") |> 
    mutate(
        ybaseline_numeric = as.numeric(as.character(ybaseline)),
        y_numeric         = as.numeric(as.character(y)),
        y_delta = y_numeric - ybaseline_numeric
    ) |> 
    filter(y_delta >= 1) |>
    group_by(id) |>
    arrange(time) |>
    slice_head(n = 1) |> 
    ungroup() |> 
    mutate(status = 1) |> 
    select(id, tx, time, status, ecog_fstcnt, diagnosis)

not_deteriorated <- df_filtered |> 
        filter(!(id %in% deteriorated$id)) |> 
        mutate(
            status = 0,
            time   = 26) |> 
        select(id, tx, time, status, ecog_fstcnt, diagnosis) |> 
        distinct()

cox_data <- bind_rows(deteriorated, not_deteriorated)


cox_model <- coxph(
  Surv(time, status) ~ tx + ecog_fstcnt + diagnosis,
  data = cox_data
)

true_hr <- cox_model$coefficients[1]

# get summary of sims

summary_ttd_h0 <- ttd_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_ttd_ha <- ttd_ha |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = true_hr, 
        threshold = 0.95) |> 
    summarise_iterations()
```

```{r}
post_ttd_h0 <- ttd_h0 |> 
  plot_posterior_distribution(x_variable = b_tx)

post_ttd_ha <- ttd_ha |> 
  plot_posterior_distribution(
    x_variable = b_tx, 
    subtitle_type = "HA", 
    hline_intercept = true_hr)

post_ttd_h0 + post_ttd_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-posteriors-ttd.svg", width = 10, height = 5)

rm(ttd_h0, ttd_ha)
```

## Cumulative Logit at 6M

#### Load draws
```{r}
path <- "./1-qol/output/sim-principal-stratum-6m-only-H0/"

month_6_h0 <- load_draws(path)
rhat_month_6_h0 <- load_rhat_values(path)

path <- "./1-qol/output/sim-principal-stratum-6m-only-HA/"

month_6_ha <- load_draws(path)
rhat_month_6_ha <- load_rhat_values(path)
```

#### Plot

```{r}
p1 <- month_6_h0 |> 
    plot_p_benefit(param = tx)

p2 <- month_6_ha |> 
    plot_p_benefit(param = tx)

p1 + p2 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-pbenefit-6m.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_month_6_h0 <- month_6_h0 |> 
  plot_posterior_distribution(x_variable = tx)

post_month_6_ha <- month_6_ha |> 
  plot_posterior_distribution(
    x_variable = tx, 
    subtitle_type = "HA", 
    hline_intercept = log(0.8))

post_month_6_h0 + post_month_6_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-posteriors-6m.svg", width = 10, height = 5)
```

```{r}
rhat_plot_month_6_h0 <- rhat_month_6_h0 |> 
    plot_rhat(tx)

rhat_plot_month_6_ha <- rhat_month_6_ha |> 
    plot_rhat(tx)

rhat_plot_month_6_h0 + rhat_plot_month_6_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-rhat-6m.svg", width = 10, height = 5)
```

#### MCSE

```{r}
summary_month_6_h0 <- month_6_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = tx, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_month_6_ha <- month_6_ha |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = tx, 
        true_param_value = log(0.8), 
        threshold = 0.95) |> 
    summarise_iterations()

rm(month_6_h0, month_6_ha)
```

## Combine results

```{r}
results_h0 <- bind_rows(
  "Markov Model" = summary_markov_h0,
  "Random Intercept" = summary_ri_h0,
  "Time-to-Deterioration" = summary_ttd_h0,
  "6-Month Ordinal" = summary_month_6_h0,
  "Change from Baseline" = summary_cfb_h0,
  .id = "Model"
)

results_ha <- bind_rows(
  "Markov Model" = summary_markov_ha,
  "Random Intercept" = summary_ri_ha,
  "Time-to-Deterioration" = summary_ttd_ha,
  "6-Month Ordinal" = summary_month_6_ha,
  "Change from Baseline" = summary_cfb_ha,
  .id = "Model"
)

format_est_mcse <- function(est, mcse, digits = 3) {
  paste0(
    format(round(est, digits), nsmall = digits),
    " (",
    format(round(mcse, digits + 1), nsmall = digits + 1),
    ")"
  )
}

summary_table <- tibble(
  Model = results_h0$Model,
  `Absolute Bias ($H_A$)` = format_est_mcse(results_ha$mean_bias, results_ha$mcse_bias),
  `Coverage ($H_A$)` = format_est_mcse(results_ha$mean_coverage, results_ha$mcse_coverage),
  `Absolute Bias ($H_0$)` = format_est_mcse(results_h0$mean_bias, results_h0$mcse_bias),
  `Coverage ($H_0$)` = format_est_mcse(results_h0$mean_coverage, results_h0$mcse_coverage),
  `Power` = format_est_mcse(results_ha$prob_rejection, results_ha$mcse_prob_rejection),
  `Type I Error` = format_est_mcse(results_h0$prob_rejection, results_h0$mcse_prob_rejection)
)

saveRDS(summary_table, file = "./1-qol/output/sim-summary.rds")
```
