---
Title: Results of Simulations
---

In this document we summarise and visualize the results of the simulations from step 6.

## Setup

### Load packages

```{r}
library(fs)
library(tidyverse)
library(arrow)
library(survival)
library(patchwork)
```

### Load Full Data

```{r}
df_full <- read_parquet("./1-qol/sim-data/sim_data_nonpo.parquet")
threshold <- 0.95
```

### Functions

```{r}
source("./functions/jackknife_mcse.R")
source("./functions/summarise-posterior.R")
source("./functions/summarise-iterations.R")
source("./functions/plot-rhat.R")
source("./functions/plot-p-benefit.R")
source("./functions/plot-posterior-distribution.R")
source("./functions/load-draws.R")
source("./functions/load-rhat-values.R")
```

## Markov Model

#### Load draws + diagnostics
```{r}
# H0
path <- "./1-qol/output/sim-principal-stratum-markov-H0/"
markov_h0 <- load_draws(path)
rhat_markov_h0 <- load_rhat_values(path = path)

#HA
path <- "./1-qol/output/sim-principal-stratum-markov-HA/"
markov_ha <- load_draws(path)
rhat_markov_ha <- load_rhat_values(path = path)
```

#### MCSE increase plot (power)

```{r}
iter_200 <- markov_ha |> 
    filter(iter <= 200) |> 
    group_by(iter) |> 
    summarise(p_benefit = mean(tx < 0)) |> 
    reframe(rejected  = as.numeric(p_benefit > threshold)) |> 
    summarise(`Monte Carlo SE for power based on posterior probability` = jackknife_mcse(rejected)) |> 
    mutate(`$n_\text{sim}$` = "200") |> 
    relocate(`$n_\text{sim}$`)


iter_400 <- markov_ha |> 
    filter(iter <= 400) |> 
    group_by(iter) |> 
    summarise(p_benefit = mean(tx < 0)) |> 
    reframe(rejected  = as.numeric(p_benefit > threshold)) |> 
    summarise(`Monte Carlo SE for power based on posterior probability` = jackknife_mcse(rejected)) |> 
    mutate(`$n_\text{sim}$` = "400")

iter_600 <- markov_ha |> 
    filter(iter <= 600) |> 
    group_by(iter) |> 
    summarise(p_benefit = mean(tx < 0)) |> 
    reframe(rejected  = as.numeric(p_benefit > threshold)) |> 
    summarise(`Monte Carlo SE for power based on posterior probability` = jackknife_mcse(rejected)) |> 
    mutate(`$n_\text{sim}$` = "600")

iter_800 <- markov_ha |> 
    filter(iter <= 800) |> 
    group_by(iter) |> 
    summarise(p_benefit = mean(tx < 0)) |> 
    reframe(rejected  = as.numeric(p_benefit > threshold)) |> 
    summarise(`Monte Carlo SE for power based on posterior probability` = jackknife_mcse(rejected)) |> 
    mutate(`$n_\text{sim}$` = "800")

iter_1000 <- markov_ha |> 
    filter(iter <= 1000) |> 
    group_by(iter) |> 
    summarise(p_benefit = mean(tx < 0)) |> 
    reframe(rejected  = as.numeric(p_benefit > threshold)) |> 
    summarise(`Monte Carlo SE for power based on posterior probability` = jackknife_mcse(rejected)) |> 
    mutate(`$n_\text{sim}$` = "1000")

mcse_evolution <- bind_rows(iter_200, iter_400, iter_600, iter_800, iter_1000)

saveRDS(mcse_evolution, file = "./1-qol/output/mcse-evolution.rds")
```

#### Convergence diagnostics
```{r}
rhat_plot_markov_ha <- rhat_markov_ha |> 
    plot_rhat()

rhat_plot_markov_h0 <- rhat_markov_h0 |> 
    plot_rhat()

rhat_plot_markov_h0 + rhat_plot_markov_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-rhat-markov.svg", width = 10, height = 5)
```

### Plot probability of benefit

```{r}
p1 <- markov_h0 |> 
    plot_p_benefit(param = tx)

p2 <- markov_ha |> 
    plot_p_benefit(param = tx)

p1 + p2 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-pbenefit-markov.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_markov_h0 <- markov_h0 |> 
  plot_posterior_distribution(x_variable = tx)

post_markov_ha <- markov_ha |> 
  plot_posterior_distribution(
    x_variable = tx, 
    subtitle_type = "HA", 
    hline_intercept = log(0.8))

post_markov_h0 + post_markov_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-posteriors-markov.svg", width = 10, height = 5)
```

```{r}
summary_markov_h0 <- markov_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = tx, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_markov_ha <- markov_ha |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = tx, 
        true_param_value = log(0.8), 
        threshold = 0.95) |> 
    summarise_iterations()

rm(markov_h0, markov_ha)
```

## Multilevel Model with RI

#### Load draws
```{r}
path <- "./1-qol/output/sim-principal-stratum-random-intercept-H0/"

ri_h0 <- load_draws(path)
rhat_ri_h0 <- load_rhat_values(path)

path <- "./1-qol/output/sim-principal-stratum-random-intercept-HA/"

ri_ha <- load_draws(path)
rhat_ri_ha <- load_rhat_values(path)
```

#### Plot

```{r}
# probability of benefit over time
p1 <- ri_h0 |> 
    plot_p_benefit(param = b_tx)

p2 <- ri_ha |> 
    plot_p_benefit(param = b_tx)

p1 + p2 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-pbenefit-ri.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_ri_h0 <- ri_h0 |> 
  plot_posterior_distribution(x_variable = b_tx)

post_ri_ha <- ri_ha |> 
  plot_posterior_distribution(
    x_variable = b_tx, 
    subtitle_type = "HA", 
    hline_intercept = log(0.8))

post_ri_h0 + post_ri_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-posteriors-ri.svg", width = 10, height = 5)
```

#### Converge diagnostics

```{r}
rhat_plot_ri_h0 <- rhat_ri_h0 |> 
    plot_rhat(b_tx)

rhat_plot_ri_ha <- rhat_ri_ha |> 
    plot_rhat(b_tx)

rhat_plot_ri_ha + rhat_plot_ri_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-rhat-ri.svg", width = 10, height = 5)
```

#### MCSE

```{r}
summary_ri_h0 <- ri_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_ri_ha <- ri_ha |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = log(0.8), 
        threshold = 0.95) |> 
    summarise_iterations()

rm(ri_h0, ri_ha)
```

## Change From Baseline Model

#### Load draws
```{r}
path <- "./1-qol/output/sim-principal-stratum-change-from-baseline-H0/"

cfb_h0 <- load_draws(path)
rhat_cfb_h0 <- load_rhat_values(path)

path <- "./1-qol/output/sim-principal-stratum-change-from-baseline-HA/"

cfb_ha <- load_draws(path)
rhat_cfb_ha <- load_rhat_values(path)

# 'true' change from baseline needs to be approximated from large dataset
mean_change_empirical <- df_full |> 
    filter(time == 1 | time == 26) |>
    mutate(y = if_else(time == 1, yprev, y)) |> 
    select(id, tx, time, y) |> 
    pivot_wider(names_from = time, values_from = y, names_prefix = "y_") |> 
    filter(y_26 != "8") |> # restrict to survivors
    mutate(delta_y = as.numeric(as.character(y_26)) - as.numeric(as.character(y_1))) |> 
    group_by(tx) |> 
    summarise(mean_change = mean(delta_y))

true_diff <- mean_change_empirical$mean_change[2] - mean_change_empirical$mean_change[1]
```

#### Converge diagnostics

```{r}
rhat_plot_cfb_h0 <- rhat_cfb_h0 |> 
    plot_rhat(b_tx)

rhat_plot_cfb_ha <- rhat_cfb_ha |> 
    plot_rhat(b_tx)

rhat_plot_cfb_h0 + rhat_plot_cfb_h0 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-rhat-cfb.svg", width = 10, height = 5)
```

```{r}
p1 <- cfb_h0 |> 
    plot_p_benefit(param = b_tx)

p2 <- cfb_ha |> 
    plot_p_benefit(param = b_tx)

p1 + p2 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-pbenefit-cfb.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_cfb_h0 <- cfb_h0 |> 
  plot_posterior_distribution(x_variable = b_tx)

post_cfb_ha <- cfb_ha |> 
  plot_posterior_distribution(
    x_variable = b_tx, 
    subtitle_type = "HA", 
    hline_intercept = true_diff)

post_cfb_h0 + post_cfb_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-posteriors-cfb.svg", width = 10, height = 5)
```

```{r}
summary_cfb_h0 <- cfb_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = 0, 
        threshold = 0.95) |> 
    summarise_iterations()

summary_cfb_ha <- cfb_ha |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = true_diff, 
        threshold = 0.95) |> 
    summarise_iterations()

rm(cfb_h0, cfb_ha)
```

## Time to Deterioation

#### Load data

```{r}
path <- "./1-qol/output/sim-principal-stratum-time-to-deterioration-H0/"

ttd_h0 <- load_draws(path)
rhat_ttd_h0 <- load_rhat_values(path)

path <- "./1-qol/output/sim-principal-stratum-time-to-deterioration-HA/"

ttd_ha <- load_draws(path)
rhat_ttd_ha <- load_rhat_values(path)
```

#### Plots

```{r}
p1 <- ttd_h0 |> 
    plot_p_benefit(param = b_tx)

p2 <- ttd_ha |> 
    plot_p_benefit(param = b_tx)

p1 + p2 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-pbenefit-ttd.svg", width = 10, height = 5)
```

```{r}
rhat_plot_ttd_ha <- rhat_ttd_ha |> 
    plot_rhat(b_tx)

rhat_plot_ttd_h0 <- rhat_ttd_h0 |> 
    plot_rhat(b_tx)

rhat_plot_ttd_h0 + rhat_plot_ttd_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-rhat-ttd.svg", width = 10, height = 5)
```

#### MCSE

```{r}
# we need to approximate 'true' HR from large dataset

ids_death <- df_full$id[df_full$y == '8']

df_filtered <- df_full |>
        filter(!(id %in% ids_death))

baseline <- df_filtered |>
        filter(time == 1) |>
        mutate(y = yprev, time = 0) |>
        select(id, ybaseline = y)

deteriorated <- df_filtered |> 
    select(-yprev) |> 
    left_join(baseline, by = "id") |> 
    mutate(
        ybaseline_numeric = as.numeric(as.character(ybaseline)),
        y_numeric         = as.numeric(as.character(y)),
        y_delta = y_numeric - ybaseline_numeric
    ) |> 
    filter(y_delta >= 1) |>
    group_by(id) |>
    arrange(time) |>
    slice_head(n = 1) |> 
    ungroup() |> 
    mutate(status = 1) |> 
    select(id, tx, time, status, ecog_fstcnt, diagnosis)

not_deteriorated <- df_filtered |> 
        filter(!(id %in% deteriorated$id)) |> 
        mutate(
            status = 0,
            time   = 26) |> 
        select(id, tx, time, status, ecog_fstcnt, diagnosis) |> 
        distinct()

cox_data <- bind_rows(deteriorated, not_deteriorated)


cox_model <- coxph(
  Surv(time, status) ~ tx + ecog_fstcnt + diagnosis,
  data = cox_data
)

true_hr <- cox_model$coefficients[1]

# get summary of sims

summary_ttd_h0 <- ttd_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_ttd_ha <- ttd_ha |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = b_tx, 
        true_param_value = true_hr, 
        threshold = 0.95) |> 
    summarise_iterations()
```

```{r}
post_ttd_h0 <- ttd_h0 |> 
  plot_posterior_distribution(x_variable = b_tx)

post_ttd_ha <- ttd_ha |> 
  plot_posterior_distribution(
    x_variable = b_tx, 
    subtitle_type = "HA", 
    hline_intercept = true_hr)

post_ttd_h0 + post_ttd_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-posteriors-ttd.svg", width = 10, height = 5)

rm(ttd_h0, ttd_ha)
```

## Cumulative Logit at 6M

#### Load draws
```{r}
path <- "./1-qol/output/sim-principal-stratum-6m-only-H0/"

month_6_h0 <- load_draws(path)
rhat_month_6_h0 <- load_rhat_values(path)

path <- "./1-qol/output/sim-principal-stratum-6m-only-HA/"

month_6_ha <- load_draws(path)
rhat_month_6_ha <- load_rhat_values(path)
```

#### Plot

```{r}
p1 <- month_6_h0 |> 
    plot_p_benefit(param = tx)

p2 <- month_6_ha |> 
    plot_p_benefit(param = tx)

p1 + p2 +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-pbenefit-6m.svg", width = 10, height = 5)
```

```{r}
#plotting stacked posteriors
post_month_6_h0 <- month_6_h0 |> 
  plot_posterior_distribution(x_variable = tx)

post_month_6_ha <- month_6_ha |> 
  plot_posterior_distribution(
    x_variable = tx, 
    subtitle_type = "HA", 
    hline_intercept = log(0.8))

post_month_6_h0 + post_month_6_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-posteriors-6m.svg", width = 10, height = 5)
```

```{r}
rhat_plot_month_6_h0 <- rhat_month_6_h0 |> 
    plot_rhat(tx)

rhat_plot_month_6_ha <- rhat_month_6_ha |> 
    plot_rhat(tx)

rhat_plot_month_6_h0 + rhat_plot_month_6_ha +
    plot_annotation(tag_levels = 'A')

ggsave(filename = "./img/fig-rhat-6m.svg", width = 10, height = 5)
```

#### MCSE

```{r}
summary_month_6_h0 <- month_6_h0 |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = tx, 
        true_param_value = log(1), 
        threshold = 0.95) |> 
    summarise_iterations()

summary_month_6_ha <- month_6_ha |> 
    group_by(iter) |> 
    summarise_posterior(
        param_name = tx, 
        true_param_value = log(0.8), 
        threshold = 0.95) |> 
    summarise_iterations()

rm(month_6_h0, month_6_ha)
```

## Combine results

```{r}
results_h0 <- bind_rows(
  "Markov Model" = summary_markov_h0,
  "Random Intercept" = summary_ri_h0,
  "Time-to-Deterioration" = summary_ttd_h0,
  "6-Month Ordinal" = summary_month_6_h0,
  "Change from Baseline" = summary_cfb_h0,
  .id = "Model"
)

results_ha <- bind_rows(
  "Markov Model" = summary_markov_ha,
  "Random Intercept" = summary_ri_ha,
  "Time-to-Deterioration" = summary_ttd_ha,
  "6-Month Ordinal" = summary_month_6_ha,
  "Change from Baseline" = summary_cfb_ha,
  .id = "Model"
)

format_est_mcse <- function(est, mcse, digits = 3) {
  paste0(
    format(round(est, digits), nsmall = digits),
    " (",
    format(round(mcse, digits + 1), nsmall = digits + 1),
    ")"
  )
}

summary_table <- tibble(
  Model = results_h0$Model,
  `Absolute Bias ($H_A$)` = format_est_mcse(results_ha$mean_bias, results_ha$mcse_bias),
  `Coverage ($H_A$)` = format_est_mcse(results_ha$mean_coverage, results_ha$mcse_coverage),
  `Absolute Bias ($H_0$)` = format_est_mcse(results_h0$mean_bias, results_h0$mcse_bias),
  `Coverage ($H_0$)` = format_est_mcse(results_h0$mean_coverage, results_h0$mcse_coverage),
  `Power` = format_est_mcse(results_ha$prob_rejection, results_ha$mcse_prob_rejection),
  `Type I Error` = format_est_mcse(results_h0$prob_rejection, results_h0$mcse_prob_rejection)
)

saveRDS(summary_table, file = "./1-qol/output/sim-summary.rds")
```
