---
title: "TAOOH first-order — diagnostics"
subtitle: "Full proportional-odds diagnostics for 50 imputations"
format: html
execute:
  freeze: false
---

```{r}
#| label: check-taooh-first-order-models
#| include: false

library(tidyverse)
source(here::here("R", "analysis-config.R"))
source(here::here("R", "markov-helpers.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
indices <- seq_len(config$imputation$main$m)
taooh_diagnostic_thresholds <- config$diagnostics
taooh_diagnostic_thresholds$max_divergent_transitions <- 19L
model_dir <- here::here(
  "artifacts", "supporting", "taooh-first-order", "models"
)
diagnostic_dir <- here::here(
  "artifacts", "supporting", "taooh-first-order", "diagnostics"
)
stan_dx_dir <- file.path(diagnostic_dir, "stan-dx")
figure_dir <- here::here(
  "artifacts", "supporting", "taooh-first-order", "figures"
)
result_path <- here::here(
  "results", "supporting", "taooh-first-order", "diagnostics.parquet"
)
dir.create(diagnostic_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(stan_dx_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(result_path), recursive = TRUE, showWarnings = FALSE)

for (index in indices) {
  model_path <- file.path(model_dir, sprintf("model-%02d.rds", index))
  output_path <- file.path(
    diagnostic_dir,
    sprintf("diagnostics-%02d.parquet", index)
  )
  if (!force && file.exists(output_path)) next

  diagnostics <- markov_diagnostics(
    readRDS(model_path),
    taooh_diagnostic_thresholds
  )
  assert_markov_diagnostics(diagnostics)
  output <- as.data.frame(diagnostics$summary)
  output$imputation <- index
  output$divergent_transitions <- diagnostics$divergent_transitions
  output$treedepth_exceeded <- diagnostics$treedepth_exceeded
  output$min_bfmi <- diagnostics$min_bfmi
  arrow::write_parquet(output, output_path)
}

diagnostic_paths <- file.path(
  diagnostic_dir,
  sprintf("diagnostics-%02d.parquet", indices)
)
diagnostics <- diagnostic_paths |>
  lapply(arrow::read_parquet) |>
  bind_rows()
arrow::write_parquet(diagnostics, result_path, compression = "zstd")
```

The full `rmsb` Stan diagnostic report and `stanDxplot()` output are stored for
every first-order fit. These replace separate `mcmc_trace()` figures.

```{r}
#| label: save-taooh-first-order-stan-diagnostics
#| include: false

for (index in indices) {
  stan_dx_text_path <- file.path(
    stan_dx_dir,
    sprintf("stan-dx-%02d.txt", index)
  )
  stan_dx_plot_path <- file.path(
    stan_dx_dir,
    sprintf("stan-dx-%02d.png", index)
  )
  if (
    !force &&
      file.exists(stan_dx_text_path) &&
      file.exists(stan_dx_plot_path)
  ) next

  model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", index)))
  stan_dx_text <- capture.output(rmsb::stanDx(model))
  writeLines(stan_dx_text, stan_dx_text_path)

  stan_dx_plot <- rmsb::stanDxplot(model)
  ggplot2::ggsave(
    stan_dx_plot_path,
    stan_dx_plot,
    width = 11,
    height = max(6, 1.05 * length(unique(stan_dx_plot$data$param))),
    dpi = 180,
    limitsize = FALSE
  )
}
```

The graphical checks use the first completed dataset and first-order fit as a
representative imputation-specific diagnostic. They use the same empirical
weekly dataset and plotting settings as the second-order diagnostics.

```{r}
#| label: derive-taooh-first-order-graphical-diagnostics
#| include: false

diagnostic_index <- indices[[1L]]
model <- readRDS(file.path(
  model_dir,
  sprintf("model-%02d.rds", diagnostic_index)
))
cohort <- arrow::read_parquet(here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
)) |>
  as.data.frame()
empirical_data <- arrow::read_parquet(here::here(
  "analysis", "00-data", "results", "analysis-taooh.parquet"
)) |>
  as.data.frame() |>
  left_join(cohort |> select(id, tx), by = "id") |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(
    y = as.integer(y_taooh),
    yprev = lag(y)
  ) |>
  ungroup() |>
  filter(between(week, 1L, 26L)) |>
  mutate(
    y = factor(y, levels = 1:5, ordered = TRUE),
    yprev = factor(yprev, levels = 1:5),
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Intervention"))
  )

diagnostic_times <- c(1L, 3L, 6L, 9L, 12L, 16L, 20L, 24L, 26L)
diagnostic_seed <- analysis_seed(
  config,
  "diagnostics",
  diagnostic_index,
  offset = 2000L
)

empirical_transition_plot <- mostr::plot_transitions(
  empirical_data,
  times = diagnostic_times,
  time_var = "week",
  y_var = "y",
  p_var = "yprev",
  y_levels = 1:5,
  show_values = TRUE
)
if (length(empirical_transition_plot$layers) >= 2L) {
  empirical_transition_plot$layers[[2L]]$aes_params$colour <- "white"
}

model_transition_plot <- mostr::plot_transitions(
  model,
  times = diagnostic_times,
  y_levels = 1:5,
  absorb = 5L,
  time_var = "week",
  p_var = "yprev",
  id_var = "id",
  seed = diagnostic_seed,
  n_draws = 200L,
  show_values = TRUE
)
if (length(model_transition_plot$layers) >= 2L) {
  model_transition_plot$layers[[2L]]$aes_params$colour <- "white"
}

empirical_correlation_plot <- mostr::plot_correlation(
  empirical_data,
  times = 1:26,
  y_levels = 1:5,
  id_var = "id",
  time_var = "week",
  y_var = "y",
  show_values = TRUE
)
model_correlation_plot <- mostr::plot_correlation(
  model,
  times = 1:26,
  y_levels = 1:5,
  absorb = 5L,
  id_var = "id",
  time_var = "week",
  p_var = "yprev",
  seed = diagnostic_seed,
  n_draws = 50L,
  show_values = TRUE
)
empirical_variogram_plot <- mostr::plot_variogram(
  empirical_data,
  times = 1:26,
  y_levels = 1:5,
  id_var = "id",
  time_var = "week",
  y_var = "y"
)
model_variogram_plot <- mostr::plot_variogram(
  model,
  times = 1:26,
  y_levels = 1:5,
  absorb = 5L,
  id_var = "id",
  time_var = "week",
  p_var = "yprev",
  seed = diagnostic_seed,
  n_draws = 50L
)

diagnostic_plots <- list(
  "empirical-transitions.png" = empirical_transition_plot,
  "model-transitions.png" = model_transition_plot,
  "empirical-correlation.png" = empirical_correlation_plot,
  "model-correlation.png" = model_correlation_plot,
  "empirical-variogram.png" = empirical_variogram_plot,
  "model-variogram.png" = model_variogram_plot
)
for (filename in names(diagnostic_plots)) {
  ggplot2::ggsave(
    file.path(figure_dir, filename),
    diagnostic_plots[[filename]],
    width = if (grepl("transition", filename)) 12 else 8,
    height = if (grepl("transition", filename)) 9 else 5.5,
    dpi = 300,
    limitsize = FALSE
  )
}
```

```{r}
#| label: fig-taooh-first-order-correlation-check
#| fig-cap: "Empirical and first-order model-derived TAOOH correlation structures"
#| fig-subcap:
#|   - "Empirical correlations"
#|   - "First-order model-derived correlations"
#| layout-ncol: 2
#| fig-width: 8
#| fig-height: 5.5
#| echo: false

empirical_correlation_plot
model_correlation_plot
```
