---
title: "TAOOH — first- versus second-order model fit"
subtitle: "PSIS-LOO comparison across 50 imputations"
format: html
execute:
  freeze: false
---

The first- and second-order models use the same completed datasets and
the same full proportional-odds specification. The first-order model omits
only the second lag and its interaction with week. PSIS-LOO is calculated
after fitting because including it in every sampling run is slow. The workflow
targets all 50 imputations and resumes from the stored LOO and model
checkpoints.

For each imputation, a difference in expected log predictive density (ELPD) is
calculated as second order minus first order. Positive values favor the
second-order model. We describe an improvement as substantial when the ELPD
difference is greater than twice its standard error. Results are kept separate by
imputation. We use
ordinary observation-level PSIS-LOO, which evaluates conditional weekly
transition prediction. We do not sum the conditional likelihood by patient:
with patient-specific random intercepts, valid leave-one-patient-out
prediction requires integrating those random effects out, and `rmsb` does not
currently provide that marginal likelihood. Simply grouping the stored
conditional likelihood would still reuse each patient's fitted random effect.

```{r}
#| label: compare-taooh-markov-order
#| include: false

library(tidyverse)
source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
indices <- seq_len(config$imputation$main$m)
second_order_model_dir <- here::here(
  "artifacts", "primary", "taooh", "models"
)
first_order_model_dir <- here::here(
  "artifacts", "supporting", "taooh-first-order", "models"
)
loo_dir <- here::here(
  "artifacts", "supporting", "taooh-first-order", "loo"
)
result_dir <- here::here("results", "supporting", "taooh-first-order")
dir.create(loo_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)

comparison_rows <- vector("list", length(indices))

for (position in seq_along(indices)) {
  index <- indices[[position]]
  second_order_model_path <- file.path(
    second_order_model_dir,
    sprintf("model-%02d.rds", index)
  )
  first_order_model_path <- file.path(
    first_order_model_dir,
    sprintf("model-%02d.rds", index)
  )
  second_order <- readRDS(second_order_model_path)
  first_order <- readRDS(first_order_model_path)

  if (
    !isTRUE(second_order$pppo == 0L) || !is.null(second_order$cppo) ||
      !isTRUE(first_order$pppo == 0L) || !is.null(first_order$cppo)
  ) {
    stop("LOO comparison requires full proportional odds in both models.")
  }

  second_terms <- attr(terms(formula(second_order)), "term.labels")
  first_terms <- attr(terms(formula(first_order)), "term.labels")
  second_order_only <- setdiff(second_terms, first_terms)
  first_order_only <- setdiff(first_terms, second_terms)
  if (
    !setequal(second_order_only, c("ypprev", "week %ia% ypprev")) ||
      length(first_order_only) > 0L
  ) {
    stop(
      "The compared models differ by terms other than the second-order ",
      "state and its interaction with week."
    )
  }

  second_order_loo_path <- file.path(
    loo_dir,
    sprintf("second-order-loo-%02d.rds", index)
  )
  first_order_loo_path <- file.path(
    loo_dir,
    sprintf("first-order-loo-%02d.rds", index)
  )

  if (!force && !is.null(second_order$loo)) {
    second_order_loo <- second_order$loo
  } else if (!force && file.exists(second_order_loo_path)) {
    second_order_loo <- readRDS(second_order_loo_path)
  } else {
    second_order_loo <- rmsb::stanGet(second_order)$loo()
    saveRDS(second_order_loo, second_order_loo_path, compress = "xz")
  }
  if (!file.exists(second_order_loo_path)) {
    saveRDS(second_order_loo, second_order_loo_path, compress = "xz")
  }
  if (!force && !is.null(first_order$loo)) {
    first_order_loo <- first_order$loo
  } else if (!force && file.exists(first_order_loo_path)) {
    first_order_loo <- readRDS(first_order_loo_path)
  } else {
    first_order_loo <- rmsb::stanGet(first_order)$loo()
    saveRDS(first_order_loo, first_order_loo_path, compress = "xz")
  }
  if (!file.exists(first_order_loo_path)) {
    saveRDS(first_order_loo, first_order_loo_path, compress = "xz")
  }

  # compareBmods() reads the $loo element from each rmsb model object. Store
  # each calculated LOO object with its fitted model as well as in the separate
  # cache so later inspection does not require recalculation.
  second_order$loo <- second_order_loo
  first_order$loo <- first_order_loo
  saveRDS(second_order, second_order_model_path, compress = "xz")
  saveRDS(first_order, first_order_model_path, compress = "xz")
  stacking_weights <- rmsb::compareBmods(
    first_order = first_order,
    second_order = second_order,
    method = "stacking"
  )
  loo_comparison <- loo::loo_compare(list(
    first_order = first_order_loo,
    second_order = second_order_loo
  ))

  first_pointwise <- first_order_loo$pointwise[, "elpd_loo"]
  second_pointwise <- second_order_loo$pointwise[, "elpd_loo"]
  if (length(first_pointwise) != length(second_pointwise)) {
    stop("The two LOO objects do not contain the same number of observations.")
  }
  pointwise_difference <- second_pointwise - first_pointwise
  delta_elpd <- sum(pointwise_difference)
  se_delta_elpd <- sqrt(
    length(pointwise_difference) * stats::var(pointwise_difference)
  )

  comparison_rows[[position]] <- tibble(
    imputation = index,
    n_observations = length(pointwise_difference),
    elpd_loo_first_order =
      first_order_loo$estimates["elpd_loo", "Estimate"],
    se_elpd_loo_first_order =
      first_order_loo$estimates["elpd_loo", "SE"],
    elpd_loo_second_order =
      second_order_loo$estimates["elpd_loo", "Estimate"],
    se_elpd_loo_second_order =
      second_order_loo$estimates["elpd_loo", "SE"],
    delta_elpd_second_minus_first = delta_elpd,
    se_delta_elpd = se_delta_elpd,
    delta_elpd_over_se = delta_elpd / se_delta_elpd,
    first_order_stacking_weight = unname(stacking_weights[["first_order"]]),
    second_order_stacking_weight = unname(stacking_weights[["second_order"]]),
    max_pareto_k_first_order = max(loo::pareto_k_values(first_order_loo)),
    max_pareto_k_second_order = max(loo::pareto_k_values(second_order_loo)),
    substantial_second_order_improvement =
      delta_elpd > 2 * se_delta_elpd,
    substantial_first_order_improvement =
      delta_elpd < -2 * se_delta_elpd
  )

  arrow::write_parquet(
    as.data.frame(loo_comparison) |>
      rownames_to_column("model"),
    file.path(loo_dir, sprintf("loo-compare-%02d.parquet", index)),
    compression = "zstd"
  )
}

model_comparison <- bind_rows(comparison_rows)
arrow::write_parquet(
  model_comparison,
  file.path(result_dir, "loo-model-comparison.parquet"),
  compression = "zstd"
)

n_substantial_second <- sum(
  model_comparison$substantial_second_order_improvement
)
n_substantial_first <- sum(
  model_comparison$substantial_first_order_improvement
)
n_favor_second <- sum(
  model_comparison$delta_elpd_second_minus_first > 0
)
pareto_warning <- any(
  model_comparison$max_pareto_k_first_order > 0.7 |
    model_comparison$max_pareto_k_second_order > 0.7
)

interpretation <- case_when(
  n_substantial_second == length(indices) ~ paste0(
    "The second-order model has a substantial predictive-fit advantage in ",
    "all completed datasets included in this run."
  ),
  n_substantial_second >= ceiling(length(indices) / 2) ~ paste0(
    "The second-order model has a substantial predictive-fit advantage in ",
    n_substantial_second,
    " of ", length(indices),
    " completed datasets, with positive ELPD differences in ",
    n_favor_second,
    " of ", length(indices),
    ". The advantage is directionally consistent but not uniform."
  ),
  n_substantial_first >= ceiling(length(indices) / 2) ~ paste0(
    "The first-order model has a substantial predictive-fit advantage in ",
    n_substantial_first,
    " of ", length(indices), " completed datasets."
  ),
  TRUE ~ paste0(
    "The comparison does not show a consistent substantial predictive-fit ",
    "advantage for either Markov order. The second-order ELPD is higher in ",
    n_favor_second,
    " of ", length(indices), " completed datasets."
  )
)
if (pareto_warning) {
  interpretation <- paste0(
    interpretation,
    " At least one fit has a Pareto-k value above 0.7, so the affected ",
    "PSIS-LOO comparison requires caution."
  )
}

interpretation_result <- tibble(
  imputations = length(indices),
  substantial_second_order = n_substantial_second,
  substantial_first_order = n_substantial_first,
  favor_second_order = n_favor_second,
  pareto_k_warning = pareto_warning,
  interpretation = interpretation
)
arrow::write_parquet(
  interpretation_result,
  file.path(result_dir, "loo-interpretation.parquet"),
  compression = "zstd"
)
```

```{r}
#| label: tbl-taooh-markov-order-comparison
#| tbl-cap: "PSIS-LOO comparison of first- and second-order TAOOH models"
#| echo: false

model_comparison |>
  select(
    imputation,
    delta_elpd_second_minus_first,
    se_delta_elpd,
    first_order_stacking_weight,
    second_order_stacking_weight,
    max_pareto_k_first_order,
    max_pareto_k_second_order
  ) |>
  knitr::kable(digits = 3)
```

```{r}
#| label: interpret-taooh-markov-order-comparison
#| results: asis
#| echo: false

cat(interpretation)
```
