---
title: "Quality of life: preparation"
subtitle: "Six-month death-inclusive landmark endpoint and weekly longitudinal grid"
format: html
---

- QoL states 1–7 run from best to worst; state 8 is death.

```{r}
#| label: setup
#| include: false

library(here)
library(arrow)
library(yaml)
library(tidyverse)

config <- read_yaml(here("config", "analysis.yml"))

results_dir <- here("analysis", "03-quality-of-life", "results")
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: read-data
#| include: false

cohort <- read_parquet(
  here("analysis", "00-data", "results", "analysis-cohort.parquet")
) |>
  as.data.frame()

qol <- read_parquet(
  here("analysis", "00-data", "results", "analysis-qol.parquet")
) |>
  as.data.frame()

landmark_columns <- c(
  "id", "tx", "age", "gender", "diagnosis", "ecog_fstcnt",
  "stage_binary", "baseline_q30", "event_death", "survival_time_days"
)

stopifnot(
  "The analysis cohort is missing a required QoL landmark variable." =
    all(landmark_columns %in% names(cohort)),
  "The QoL dataset lacks dates or Q30 values required for landmark preparation." =
    all(
      c("id", "assessment_id", "questionnaire_date", "day", "q30") %in% names(qol)
    ),
  "The analysis cohort must have one row per participant." =
    !anyDuplicated(cohort$id)
)

# The cohort currently carries the visit date as legacy `visit_date`.
visit_variable <- intersect(c("date_visit", "visit_date"), names(cohort))
stopifnot(
  "The analysis cohort must contain exactly one of `date_visit` or `visit_date`." =
    length(visit_variable) == 1L
)

cohort <- cohort |>
  mutate(date_visit = .data[[visit_variable]])
```

## Diagnosis groups

```{r}
#| label: diagnosis-groups
#| include: false

common_diagnoses <- cohort |>
  filter(!is.na(diagnosis)) |>
  count(diagnosis, name = "n") |>
  arrange(desc(n), as.character(diagnosis)) |>
  slice_head(n = 4L) |>
  pull(diagnosis) |>
  as.character()

cohort <- cohort |>
  mutate(
    diagnosis_group = case_when(
      is.na(diagnosis) ~ NA_character_,
      as.character(diagnosis) %in% common_diagnoses ~ as.character(diagnosis),
      TRUE ~ "Other"
    ),
    diagnosis_code = as.integer(factor(
      diagnosis_group,
      levels = c(common_diagnoses, "Other")
    ))
  )
```

## Baseline Q30 (landmark dataset)

- Fallback baseline questionnaire is removed from follow-up.

```{r}
#| label: baseline-q30
#| include: false

missing_baseline_ids <- cohort |>
  filter(is.na(baseline_q30)) |>
  pull(id)

fallback_baseline <- qol |>
  filter(
    id %in% missing_baseline_ids,
    !is.na(q30),
    !is.na(questionnaire_date)
  ) |>
  select(id, assessment_id, questionnaire_date, q30) |>
  inner_join(cohort |> select(id, date_visit), by = "id") |>
  mutate(distance_days = abs(as.integer(questionnaire_date - date_visit))) |>
  filter(!is.na(distance_days), distance_days <= 14L) |>
  arrange(id, distance_days, questionnaire_date, assessment_id) |>
  distinct(id, .keep_all = TRUE) |>
  transmute(
    id,
    fallback_assessment_id = assessment_id,
    fallback_q30 = as.integer(q30)
  )

baseline <- cohort |>
  transmute(id, baseline_q30 = as.integer(baseline_q30)) |>
  left_join(fallback_baseline, by = "id") |>
  mutate(
    baseline_assessment_id = if_else(
      is.na(baseline_q30),
      fallback_assessment_id,
      NA_character_
    ),
    baseline_q30 = coalesce(baseline_q30, fallback_q30)
  ) |>
  select(id, baseline_q30, baseline_assessment_id)
```

## Six-month landmark endpoint

- Censored before day 182: endpoint unknown until vital status is imputed.

```{r}
#| label: landmark-endpoint
#| include: false

latest <- qol |>
  filter(
    !is.na(q30),
    day > 0,
    day <= config$analysis$horizon_days
  ) |>
  select(id, assessment_id, day, q30) |>
  left_join(baseline |> select(id, baseline_assessment_id), by = "id") |>
  filter(is.na(baseline_assessment_id) | assessment_id != baseline_assessment_id) |>
  arrange(id, day) |>
  slice_tail(n = 1L, by = id) |>
  transmute(
    id,
    landmark_day = day,
    landmark_q30 = as.integer(q30)
  )

prepared_columns <- c(
  landmark_columns,
  "diagnosis_group", "diagnosis_code",
  intersect(c("plan_fstcnt_coded", "albumin", "c_reactive_protein"), names(cohort))
)

prepared <- cohort |>
  select(all_of(prepared_columns), -baseline_q30) |>
  left_join(baseline, by = "id") |>
  left_join(latest, by = "id") |>
  mutate(
    y_landmark = case_when(
      event_death == 1L ~ 8L,
      survival_time_days < config$analysis$horizon_days |
        is.na(landmark_q30) ~ NA_integer_,
      TRUE ~ landmark_q30
    )
  )

stopifnot(
  "Baseline Q30 produced a QoL state outside 1:7." =
    all(is.na(prepared$baseline_q30) | prepared$baseline_q30 %in% 1:7),
  "The six-month QoL endpoint contains a state outside 1:8." =
    all(is.na(prepared$y_landmark) | prepared$y_landmark %in% 1:8),
  "Every death by day 182 must be assigned the absorbing death state." =
    !any(prepared$event_death == 1L & prepared$y_landmark != 8L)
)

write_parquet(
  prepared,
  file.path(results_dir, "prepared.parquet"),
  compression = "zstd"
)
```

## Weekly death-inclusive longitudinal grid

- No ±14-day baseline fallback here; such a questionnaire stays in follow-up.
- Final partial week kept, as in TAOOH.

```{r}
#| label: longitudinal-grid
#| include: false

weekly_qol <- qol |>
  filter(
    !is.na(q30),
    day > 0,
    day <= config$analysis$horizon_days
  ) |>
  mutate(week = pmax(1L, pmin(26L, as.integer(ceiling(day / 7))))) |>
  arrange(id, week, day) |>
  slice_tail(n = 1L, by = c(id, week)) |>
  select(id, week, q30)

long_grid <- cohort |>
  select(
    id, tx, age, gender, diagnosis_code, ecog_fstcnt,
    stage_binary, plan_fstcnt_coded, albumin, c_reactive_protein,
    baseline_q30, event_death, survival_time_days
  ) |>
  cross_join(tibble(week = 1:26)) |>
  left_join(weekly_qol, by = c("id", "week")) |>
  mutate(
    death_week = if_else(
      event_death == 1L,
      pmax(1L, pmin(26L, as.integer(ceiling(survival_time_days / 7)))),
      NA_integer_
    ),
    known_alive = week <= ceiling(survival_time_days / 7) | !is.na(q30),
    after_death = event_death == 1L & week >= death_week
  ) |>
  filter(known_alive | after_death) |>
  mutate(
    y_death_inclusive = case_when(
      event_death == 1L & week >= death_week ~ 8L,
      is.na(q30) ~ NA_integer_,
      TRUE ~ as.integer(q30)
    )
  ) |>
  select(-known_alive, -after_death) |>
  arrange(id, week)

write_parquet(
  long_grid,
  file.path(results_dir, "death-inclusive-longitudinal-prepared.parquet"),
  compression = "zstd"
)
```
