---
title: "Secondary outcome: survival rates"
format: html
execute:
  freeze: false
  echo: false
---

# Survival rates and risk differences

At day 182 we use the primary adjusted OS models; at day 365 we use the
adjusted proportional-hazards models fitted to unrestricted follow-up.
Their saved posterior curves average patient-specific predictions over the
randomized cohort under each treatment. We interpolate within each curve,
pair intervention and control within posterior draw and imputation, and
pool equal numbers of draws from each completed dataset.

The survival risk difference is $S^1(t)-S^0(t)$: positive values favour
invitation. The 365-day estimate extends the fixed-time contrast using the
existing unrestricted-follow-up models. Predictions must cover the requested
day; the primary curves ending at day 182 cannot supply a 365-day estimate.

```{r}
#| label: derive-survival-rate-contrasts
#| include: false

library(dplyr)
source(here::here("R", "survival-helpers.R"))

result_dir <- here::here("results", "secondary", "survival-rates")
figure_dir <- here::here("artifacts", "secondary", "survival-rates", "figures")
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

primary_curves <- arrow::read_parquet(here::here(
  "results", "primary", "overall-survival", "survival-curves.parquet"
))
unrestricted_paths <- list.files(here::here(
  "artifacts", "supporting", "os-unrestricted-ph", "survival-curves"
), pattern = "^survival-curves-[0-9]+\\.parquet$", full.names = TRUE)
stopifnot(length(unrestricted_paths) > 0L)
unrestricted_curves <- purrr::map(unrestricted_paths, arrow::read_parquet) |>
  bind_rows()

rate_draws <- bind_rows(
  survival_risk_draws(primary_curves, horizon = 182) |>
    mutate(horizon_days = 182L, model = "Primary adjusted OS"),
  survival_risk_draws(unrestricted_curves, horizon = 365) |>
    mutate(horizon_days = 365L, model = "Unrestricted adjusted PH OS")
)
stopifnot("imputation" %in% names(rate_draws), setequal(rate_draws$tx, 0:1))
draw_counts <- rate_draws |> count(horizon_days, imputation, tx)
stopifnot(all((draw_counts |> summarise(
  equal = n_distinct(n) == 1L, .by = horizon_days
))$equal))

contrast_draws <- rate_draws |>
  tidyr::pivot_wider(names_from = tx, values_from = survival, names_prefix = "survival_") |>
  mutate(risk_difference = survival_1 - survival_0)
stopifnot(!anyNA(contrast_draws))

rate_summary <- rate_draws |>
  summarise(
    median = median(survival),
    lower_95 = quantile(survival, 0.025, names = FALSE),
    upper_95 = quantile(survival, 0.975, names = FALSE),
    .by = c(horizon_days, model, tx)
  )
contrast_summary <- contrast_draws |>
  summarise(
    median = median(risk_difference),
    lower_66 = quantile(risk_difference, 0.17, names = FALSE),
    upper_66 = quantile(risk_difference, 0.83, names = FALSE),
    lower_95 = quantile(risk_difference, 0.025, names = FALSE),
    upper_95 = quantile(risk_difference, 0.975, names = FALSE),
    probability_benefit = mean(risk_difference > 0),
    imputations = n_distinct(imputation),
    draws_per_imputation = n() / n_distinct(imputation),
    .by = c(horizon_days, model)
  )
arrow::write_parquet(rate_summary, file.path(result_dir, "rates-summary.parquet"))
arrow::write_parquet(contrast_summary, file.path(result_dir, "summary.parquet"))
arrow::write_parquet(contrast_draws, file.path(result_dir, "estimand-draws.parquet"))
```

```{r}
#| label: fig-survival-risk-difference-posterior
#| fig-cap: "Posterior survival risk differences at 182 and 365 days. Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "Posterior half-eye densities and dot plots of intervention-minus-control survival differences at two horizons, with a zero reference line. Positive differences favour invitation."
#| fig-width: 8
#| fig-height: 7

risk_difference_plot <- ggplot2::ggplot(
  contrast_draws, ggplot2::aes(x = 100 * risk_difference)
) +
  ggdist::stat_dots(side = "bottom") +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  ggplot2::geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  ggplot2::facet_wrap(~ horizon_days, ncol = 1, labeller = ggplot2::as_labeller(c(
    `182` = "Day 182: primary adjusted OS",
    `365` = "Day 365: unrestricted adjusted PH OS"
  ))) +
  ggplot2::labs(x = "Survival risk difference (intervention - control), percentage points", y = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor.y = ggplot2::element_blank()
  )
ggplot2::ggsave(file.path(figure_dir, "survival-risk-difference-posterior.png"),
  risk_difference_plot, width = 8, height = 7, dpi = 300)
risk_difference_plot
```
