---
title: "Quality of life — primary imputation"
subtitle: "Fifty MICE imputations of the six-month ordinal endpoint"
format: html
execute:
  freeze: false
---

Missing baseline and living-participant QoL responses are imputed on the
pre-derived analysis-state scale (1 = best, 7 = worst). For participants
censored before day 182, vital status at day 182 is drawn from a dedicated
Bayesian survival model conditional on survival to the recorded censoring day.
Drawn and observed deaths receive state 8; QoL is imputed only for participants
classified as alive at day 182.

```{r}
#| label: impute-primary-qol-landmark
#| include: false
source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
input_path <- here::here(
  "artifacts", "primary", "quality-of-life", "prepared.parquet"
)
output_dir <- here::here("artifacts", "primary", "quality-of-life")
output_path <- file.path(output_dir, "imputations.rds")
imputation_fits_path <- file.path(output_dir, "imputation-fits.rds")
vital_status_model_path <- file.path(output_dir, "vital-status-model.rds")
vital_status_draws_path <- file.path(output_dir, "vital-status-draws.parquet")
vital_status_diagnostics_path <- file.path(
  output_dir,
  "vital-status-diagnostics.parquet"
)
diagnostic_dir <- file.path(output_dir, "imputation-diagnostics")
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

prepared <- arrow::read_parquet(input_path) |>
  as.data.frame()
diagnosis_levels <- unique(prepared$diagnosis_group[
  order(prepared$diagnosis_code)
])
prepared$diagnosis_group <- factor(
  prepared$diagnosis_group,
  levels = diagnosis_levels
)
prepared$gender <- factor(prepared$gender)
prepared$age_scaled <- scale(prepared$age)
prepared$ecog_fstcnt <- ifelse(prepared$ecog_fstcnt == 4, 3, prepared$ecog_fstcnt)


fit_vital_status_model <- force || !file.exists(vital_status_model_path)

if (fit_vital_status_model) {
  vital_status_data <- prepared[prepared$survival_time_days > 0, ]
  
  vital_status_formula <- survival::Surv(
    survival_time_days,
    event_death
  ) ~ tx + age_scaled + gender + diagnosis_group + stage_binary
  
  vital_status_terms <- colnames(stats::model.matrix(
    ~ tx + age_scaled + gender + diagnosis_group + stage_binary,
    data = vital_status_data
  ))[-1L]
  vital_status_model <- rstanarm::stan_surv(
    formula = vital_status_formula,
    data = vital_status_data,
    basehaz = "ms",
    basehaz_ops = list(df = 5),
    prior = rstanarm::normal(
      location = rep(0, length(vital_status_terms)),
      scale = ifelse(vital_status_terms == "tx", 0.55, 2.5)
    ),
    prior_intercept = rstanarm::normal(0, 20),
    prior_aux = rstanarm::dirichlet(1),
    chains = 4L,
    cores = 4L,
    iter = 2000L,
    warmup = 1000L,
    seed = analysis_seed(config, "qol", offset = 100L)
  )
  saveRDS(vital_status_model, vital_status_model_path, compress = "xz")
}
vital_status_model <- readRDS(vital_status_model_path)

vital_status_diagnostics <- summary(vital_status_model) |>
  as.data.frame()

vital_status_diagnostics$parameter <- rownames(vital_status_diagnostics)

rownames(vital_status_diagnostics) <- NULL

finite_rhat <- vital_status_diagnostics$Rhat[
  is.finite(vital_status_diagnostics$Rhat)
]
finite_ess <- vital_status_diagnostics$n_eff[
  is.finite(vital_status_diagnostics$n_eff)
]

if (any(finite_rhat > config$diagnostics$max_rhat) ||
    any(finite_ess < config$diagnostics$min_ess_bulk)) {
  stop("The dedicated QoL vital-status survival model did not converge.")
}

arrow::write_parquet(
  vital_status_diagnostics,
  vital_status_diagnostics_path,
  compression = "zstd"
)

rerun_imputations <- force || fit_vital_status_model ||
  !all(file.exists(c(output_path, imputation_fits_path, vital_status_draws_path)))
if (rerun_imputations) {
  prepared$gender_code <- as.integer(factor(prepared$gender))
  prepared$plan_code <- as.integer(factor(prepared$plan_fstcnt_coded))

  completed_datasets <- vector("list", config$imputation$main$m)
  imputation_fits <- vector("list", config$imputation$main$m)
  vital_status_draws <- vector("list", config$imputation$main$m)
  unresolved <- prepared$event_death == 0L &
    prepared$survival_time_days < config$analysis$horizon_days

  for (index in seq_len(config$imputation$main$m)) {
    death_182 <- prepared$event_death
    conditional_survival <- numeric(sum(unresolved))
    if (any(unresolved)) {
      predicted_survival <- rstanarm::posterior_survfit(
        vital_status_model,
        newdata = prepared[unresolved, ],
        type = "surv",
        times = config$analysis$horizon_days,
        condition = TRUE,
        last_time = "survival_time_days",
        standardise = FALSE,
        extrapolate = FALSE,
        draws = 1L,
        seed = analysis_seed(config, "qol", index, offset = 120L),
        return_matrix = TRUE
      )
      conditional_survival <- as.numeric(predicted_survival[[1L]][1L, ])
      death_probability <- 1 - conditional_survival
      set.seed(analysis_seed(config, "qol", index, offset = 140L))
      death_182[unresolved] <- stats::rbinom(
        n = sum(unresolved),
        size = 1L,
        prob = death_probability
      )
      vital_status_draws[[index]] <- data.frame(
        id = prepared$id[unresolved],
        imputation = index,
        censoring_day = prepared$survival_time_days[unresolved],
        conditional_survival = conditional_survival,
        death_probability = death_probability,
        death_182 = death_182[unresolved],
        stringsAsFactors = FALSE
      )
    }

    imputation_columns <- intersect(c(
      "id", "tx", "age", "gender_code", "diagnosis_code", "ecog_fstcnt",
      "stage_binary", "plan_code", "albumin", "c_reactive_protein", 
      "baseline_q30", "landmark_q30"
    ), names(prepared))
    imputation_data <- prepared[imputation_columns]
    imputation_data$death_182 <- death_182
    ordinal_levels <- list(
      ecog_fstcnt = 0:3,
      baseline_q30 = 1:7,
      landmark_q30 = 1:7
    )
    ordinal_columns <- intersect(names(ordinal_levels), names(imputation_data))
    for (variable in ordinal_columns) {
      imputation_data[[variable]] <- ordered(
        imputation_data[[variable]],
        levels = ordinal_levels[[variable]]
      )
    }
    methods <- mice::make.method(imputation_data)
    # methods[] <- ""
    imputed_columns <- intersect(
      c(
        # "age", "gender_code", "diagnosis_code", "stage_binary", # are complete 
        "ecog_fstcnt", 
        "plan_code",
        "albumin", 
        "c_reactive_protein", 
        "baseline_q30", 
        "landmark_q30"
      ),
      names(imputation_data)
    )
    # methods[imputed_columns] <- "pmm"
    # methods[ordinal_columns] <- "polr"

    predictors <- mice::make.predictorMatrix(imputation_data)
    predictors[,"id"] <- 0L #don't impute anything with id
    # predictor_columns <- setdiff(names(imputation_data), "id")
    # predictors[imputed_columns, predictor_columns] <- 1L
    # diag(predictors) <- 0L
    # predictors[setdiff(imputed_columns, "landmark_q30"), "landmark_q30"] <- 0L

    # where <- matrix(
    #   FALSE,
    #   nrow = nrow(imputation_data),
    #   ncol = ncol(imputation_data),
    #   dimnames = dimnames(imputation_data)
    # )
    # for (variable in setdiff(imputed_columns, "landmark_q30")) {
    #   where[, variable] <- is.na(imputation_data[[variable]])
    # }
    # where[, "landmark_q30"] <- is.na(imputation_data$landmark_q30) &
    #   imputation_data$death_182 == 0L

    imputation_fit <- mice::mice(
      imputation_data,
      m = 1L,
      maxit = config$imputation$main$maxit,
      method = methods,
      predictorMatrix = predictors,
      # where = where,
      donors = config$imputation$main$donors,
      seed = analysis_seed(config, "imputation", index, offset = 200L),
      printFlag = FALSE
    )
    completed <- mice::complete(imputation_fit, 1L)
    completed$baseline_q30 <- as.integer(as.character(completed$baseline_q30))
    completed$landmark_q30 <- as.integer(as.character(completed$landmark_q30))
    completed$y_landmark <- ifelse(
      completed$death_182 == 1L,
      8L,
      completed$landmark_q30
    )
    
    if (anyNA(completed$baseline_q30) || anyNA(completed$y_landmark)) {
      stop("QoL imputation left a substantive-model variable missing.")
    }
    if (any(!completed$baseline_q30 %in% 1:7) ||
        any(!completed$y_landmark %in% 1:8)) {
      stop("QoL imputation produced an invalid ordinal state.")
    }
    completed_datasets[[index]] <- completed
    imputation_fits[[index]] <- imputation_fit
  }

  saveRDS(completed_datasets, output_path, compress = "xz")
  saveRDS(imputation_fits, imputation_fits_path, compress = "xz")
  arrow::write_parquet(
    do.call(rbind, vital_status_draws),
    vital_status_draws_path,
    compression = "zstd"
  )
}
```

```{r}
#| label: plot-primary-qol-imputation-diagnostics
#| echo: false
#| fig-cap: "MICE convergence traces and observed-versus-imputed distributions for the primary QoL analysis."

completed_datasets <- readRDS(output_path)
imputation_fits <- readRDS(imputation_fits_path)
dir.create(diagnostic_dir, recursive = TRUE, showWarnings = FALSE)

trace_variables <- c("baseline_q30", "landmark_q30", "ecog_fstcnt")
old_par <- graphics::par(mfrow = c(1, length(trace_variables)))
for (variable in trace_variables) {
  trace_values <- vapply(
    imputation_fits,
    function(fit) as.numeric(fit$chainMean[variable, , 1L]),
    numeric(config$imputation$main$maxit)
  )
  if (variable == "ecog_fstcnt") trace_values <- trace_values - 1
  graphics::matplot(
    seq_len(config$imputation$main$maxit),
    trace_values,
    type = "l",
    lty = 1,
    col = grDevices::adjustcolor("black", alpha.f = 0.2),
    xlab = "Iteration",
    ylab = "Mean imputed value",
    main = variable
  )
}
graphics::par(old_par)

imputation_labels <- sprintf(
  "Imputation %02d",
  seq_along(completed_datasets)
)

density_data <- do.call(
  rbind,
  lapply(trace_variables, function(variable) {
    observed <- data.frame(
      variable = variable,
      series = "Observed",
      value = as.numeric(as.character(
        prepared[[variable]][!is.na(prepared[[variable]])]
      )),
      stringsAsFactors = FALSE
    )

    imputed <- do.call(
      rbind,
      lapply(seq_along(completed_datasets), function(index) {
        data.frame(
          variable = variable,
          series = imputation_labels[index],
          value = as.numeric(as.character(
            completed_datasets[[index]][[variable]][
              is.na(prepared[[variable]])
            ]
          )),
          stringsAsFactors = FALSE
        )
      })
    )

    rbind(observed, imputed)
  })
)
density_data$series <- factor(
  density_data$series,
  levels = c("Observed", imputation_labels)
)

density_plot <- lattice::densityplot(
  ~ value | variable,
  groups = series,
  data = density_data,
  scales = "free",
  plot.points = FALSE,
  par.settings = list(
    superpose.line = list(
      col = c(
        "black",
        rep("#da545480", length(imputation_labels))
      ),
      lwd = c(2.5, rep(0.8, length(imputation_labels)))
    )
  ),
  key = list(
    columns = 2,
    lines = list(
      col = c("black", "#2878B580"),
      lwd = c(2.5, 0.8)
    ),
    text = list(c("Observed", "Each completed dataset"))
  )
)
print(density_plot)
```
