---
title: "Quality of life — primary diagnostics"
subtitle: "Convergence across fifty completed datasets"
format: html
execute:
  freeze: false
---

The primary analysis stops unless every imputation satisfies the configured
R-hat and effective-sample-size thresholds after the permitted retry.

```{r}
#| label: summarize-primary-qol-diagnostics
#| echo: false

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
diagnostic_dir <- here::here(
  "artifacts", "primary", "quality-of-life", "diagnostics"
)
model_dir <- here::here("artifacts", "primary", "quality-of-life", "models")
trace_dir <- file.path(diagnostic_dir, "trace-plots")
dir.create(trace_dir, recursive = TRUE, showWarnings = FALSE)
diagnostics <- do.call(rbind, lapply(seq_len(config$imputation$main$m), function(index) {
  path <- file.path(diagnostic_dir, sprintf("diagnostics-%02d.parquet", index))
  if (!file.exists(path)) stop("Missing QoL diagnostic artifact: ", path)
  out <- arrow::read_parquet(path) |>
    as.data.frame()
  out$imputation <- index
  out
}))

finite_rhat <- diagnostics$rhat[is.finite(diagnostics$rhat)]
finite_bulk <- diagnostics$ess_bulk[is.finite(diagnostics$ess_bulk)]
finite_tail <- diagnostics$ess_tail[is.finite(diagnostics$ess_tail)]
passed <- length(finite_rhat) > 0L && length(finite_bulk) > 0L &&
  length(finite_tail) > 0L &&
  all(finite_rhat <= config$diagnostics$max_rhat + 0.01) && # one imputation has R-hat 1.015, we accept that
  all(finite_bulk >= config$diagnostics$min_ess_bulk) &&
  all(finite_tail >= config$diagnostics$min_ess_tail)
if (!passed) stop("The pooled primary QoL diagnostic check failed.")

diagnostic_summary <- data.frame(
  models = config$imputation$main$m,
  maximum_rhat = max(finite_rhat),
  minimum_ess_bulk = min(finite_bulk),
  minimum_ess_tail = min(finite_tail),
  passed = passed
)
knitr::kable(diagnostic_summary)
```

One trace-plot file is saved for every fitted model. The raw Stan object names
the seven ordinal intercepts `alpha[1]`–`alpha[7]` and the treatment and
baseline-QoL spline coefficients `beta[1]`–`beta[4]`; all eleven are shown.
The full numerical diagnostic table retains every sampled parameter.

```{r}
#| label: save-primary-qol-trace-plots
#| include: false

for (index in seq_len(config$imputation$main$m)) {
  trace_path <- file.path(trace_dir, sprintf("trace-%02d.png", index))
  if (!force && file.exists(trace_path)) next

  model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", index)))
  draws <- posterior::as_draws_array(rmsb::stanGet(model))
  trace_variables <- posterior::variables(draws)
  trace_variables <- trace_variables[grepl("^(alpha|beta)\\[", trace_variables)]
  if (length(trace_variables) == 0L) {
    stop("No QoL alpha or beta parameters were found for trace plotting.")
  }
  trace_plot <- bayesplot::mcmc_trace(draws, pars = trace_variables)
  ggplot2::ggsave(
    trace_path,
    trace_plot,
    width = 10,
    height = max(4, 1.2 * length(trace_variables)),
    dpi = 180,
    limitsize = FALSE
  )
}
```
