---
title: "Quality of life — descriptive summaries"
subtitle: "Observed, imputed, and model-derived six-month state distributions"
format: html
execute:
  freeze: false
---

This stage saves descriptive summaries of the raw six-month endpoint and
compares its observed distribution with the distributions averaged across the
50 completed datasets and standardized from the fitted proportional-odds
models. Missing outcomes are retained as an explicit grey category in the raw
data display. The model-derived bars use the median of the stacked posterior
state-probability draws across imputations.

```{r}
#| label: describe-primary-qol
#| include: false

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
prepared_path <- here::here(
  "artifacts", "primary", "quality-of-life", "prepared.parquet"
)
imputation_path <- here::here(
  "artifacts", "primary", "quality-of-life", "imputations.rds"
)
model_dir <- here::here("artifacts", "primary", "quality-of-life", "models")
result_dir <- here::here("results", "primary", "quality-of-life", "descriptive")
figure_dir <- here::here("artifacts", "primary", "quality-of-life", "figures")
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

prepared <- arrow::read_parquet(prepared_path) |>
  as.data.frame()
imputations <- readRDS(imputation_path)
indices <- seq_len(config$imputation$main$m)
state_levels <- as.character(1:8)
display_levels <- c(state_levels, "Missing")

raw_summary <- do.call(rbind, lapply(split(prepared, prepared$tx), function(data) {
  data.frame(
    tx = data$tx[[1L]],
    n = nrow(data),
    deaths_by_day_182 = sum(data$event_death == 1L, na.rm = TRUE),
    baseline_qol_observed = sum(!is.na(data$baseline_q30)),
    baseline_qol_missing = sum(is.na(data$baseline_q30)),
    landmark_qol_observed = sum(!is.na(data$landmark_q30)),
    landmark_qol_missing = sum(is.na(data$landmark_q30)),
    composite_observed = sum(!is.na(data$y_landmark)),
    composite_missing = sum(is.na(data$y_landmark))
  )
}))
rownames(raw_summary) <- NULL

raw_state <- ifelse(is.na(prepared$y_landmark), "Missing", prepared$y_landmark)
raw_counts <- as.data.frame(table(
  tx = prepared$tx,
  state = factor(raw_state, levels = display_levels),
  useNA = "no"
))
raw_counts$proportion <- ave(raw_counts$Freq, raw_counts$tx, FUN = function(x) {
  x / sum(x)
})
raw_counts$source <- "Observed raw data"
names(raw_counts)[names(raw_counts) == "Freq"] <- "n"

imputed_probabilities <- do.call(rbind, lapply(indices, function(index) {
  data <- imputations[[index]]
  do.call(rbind, lapply(0:1, function(treatment) {
    counts <- table(factor(
      data$y_landmark[data$tx == treatment],
      levels = 1:8
    ))
    data.frame(
      imputation = index,
      tx = treatment,
      state = names(counts),
      proportion = as.numeric(counts) / sum(counts)
    )
  }))
}))
imputed_summary <- stats::aggregate(
  proportion ~ tx + state,
  data = imputed_probabilities,
  FUN = mean
)
imputed_summary$n <- NA_integer_
imputed_summary$source <- "Mean across completed datasets"

model_draws <- vector("list", length(indices) * 2L)
position <- 1L
for (index in indices) {
  model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", index)))
  data <- imputations[[index]]
  set.seed(analysis_seed(config, "qol", index, offset = 700L))
  selected <- sample.int(
    nrow(model$draws),
    config$markov$post_estimation_draws,
    replace = FALSE
  )
  model$draws <- model$draws[selected, , drop = FALSE]

  for (treatment in 0:1) {
    newdata <- data
    newdata$tx <- treatment
    probabilities <- predict(
      model,
      newdata = newdata,
      type = "fitted.ind",
      posterior.summary = "all"
    )
    standardized <- apply(probabilities, c(1, 3), mean)
    model_draws[[position]] <- data.frame(
      imputation = index,
      draw = rep(seq_len(nrow(standardized)), each = ncol(standardized)),
      tx = treatment,
      state = rep(state_levels, times = nrow(standardized)),
      probability = as.vector(t(standardized))
    )
    position <- position + 1L
  }
}
model_draws <- do.call(rbind, model_draws)
model_summary <- do.call(rbind, lapply(
  split(model_draws, interaction(model_draws$tx, model_draws$state, drop = TRUE)),
  function(data) {
    data.frame(
      tx = data$tx[[1L]],
      state = data$state[[1L]],
      proportion = median(data$probability),
      q025 = quantile(data$probability, 0.025, names = FALSE),
      q975 = quantile(data$probability, 0.975, names = FALSE)
    )
  }
))
rownames(model_summary) <- NULL
model_summary$n <- NA_integer_
model_summary$source <- "Median proportional-odds prediction"

plot_data <- rbind(
  raw_counts[c("tx", "state", "n", "proportion", "source")],
  imputed_summary[c("tx", "state", "n", "proportion", "source")],
  model_summary[c("tx", "state", "n", "proportion", "source")]
)
plot_data$state <- factor(plot_data$state, levels = display_levels)
plot_data$tx <- factor(plot_data$tx, levels = 0:1, labels = c("Control", "Intervention"))
plot_data$source <- factor(
  plot_data$source,
  levels = c(
    "Observed raw data",
    "Mean across completed datasets",
    "Median proportional-odds prediction"
  )
)
state_colours <- c(
  stats::setNames(grDevices::hcl.colors(8L, "viridis"), state_levels),
  Missing = "grey75"
)
state_plot <- ggplot2::ggplot(
  plot_data,
  ggplot2::aes(x = tx, y = proportion, fill = state)
) +
  ggplot2::geom_col(width = 0.72) +
  ggplot2::facet_wrap(ggplot2::vars(source), nrow = 1L) +
  ggplot2::scale_fill_manual(values = state_colours, drop = FALSE) +
  ggplot2::scale_y_continuous(labels = scales::label_percent()) +
  ggplot2::labs(
    x = NULL,
    y = "Proportion",
    fill = "QoL/death state"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.grid.major.x = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
  )

arrow::write_parquet(raw_summary, file.path(result_dir, "raw-summary.parquet"))
arrow::write_parquet(raw_counts, file.path(result_dir, "raw-state-counts.parquet"))
arrow::write_parquet(
  imputed_probabilities,
  file.path(result_dir, "imputed-state-probabilities.parquet")
)
arrow::write_parquet(
  model_summary,
  file.path(result_dir, "model-state-probabilities.parquet")
)
ggplot2::ggsave(
  file.path(figure_dir, "observed-imputed-model-state-distributions.png"),
  state_plot,
  width = 12,
  height = 5,
  dpi = 300
)
```

```{r}
#| label: tbl-qol-raw-summary
#| tbl-cap: "Availability of the raw QoL variables by randomized group"
#| echo: false

knitr::kable(raw_summary)
```

```{r}
#| label: fig-qol-state-distributions
#| fig-cap: "Observed, imputed, and proportional-odds model-derived distributions of the six-month QoL/death state"
#| fig-alt: "Three faceted stacked bar charts compare QoL and death states 1 through 8 by treatment group. The raw-data panel also includes a grey missing category."
#| fig-width: 12
#| fig-height: 5
#| echo: false

state_plot
```
