---
title: "Quality of life — proportional-odds check"
subtitle: "Common and threshold-specific treatment odds ratios"
format: html
execute:
  freeze: false
---

The proportional-odds assumption for the treatment effect is examined by
comparing the common treatment log-odds ratio with treatment effects from seven
binary logistic models. For threshold \(k\), the binary event is
\(Y_{182} > k\), so positive values consistently indicate greater odds of a
worse outcome under intervention. All models adjust for baseline QoL using the
same four-knot restricted cubic spline as the primary model. Estimates and
variances are pooled across the 50 completed datasets using Rubin's rules.

This is a frequentist graphical model check, not a replacement for the Bayesian
primary estimand.

```{r}
#| label: check-qol-proportional-odds
#| include: false

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
imputations <- readRDS(here::here(
  "artifacts", "primary", "quality-of-life", "imputations.rds"
))
indices <- seq_len(config$imputation$main$m)
thresholds <- 1:7

estimates <- vector("list", length(indices) * (length(thresholds) + 1L))
position <- 1L
for (index in indices) {
  data <- imputations[[index]]
  data$y_landmark <- ordered(data$y_landmark, levels = 1:8)
  dd <- rms::datadist(data)
  options(datadist = "dd")

  common_fit <- rms::lrm(
    y_landmark ~ tx + rms::rcs(baseline_q30, 4),
    data = data,
    x = TRUE,
    y = TRUE
  )
  estimates[[position]] <- data.frame(
    imputation = index,
    threshold = NA_integer_,
    comparison = "Common proportional-odds effect",
    estimate = unname(stats::coef(common_fit)["tx"]),
    variance = unname(stats::vcov(common_fit)["tx", "tx"])
  )
  position <- position + 1L

  for (threshold in thresholds) {
    data$binary_worse <- as.integer(as.integer(data$y_landmark) > threshold)
    binary_fit <- rms::lrm(
      binary_worse ~ tx + rms::rcs(baseline_q30, 4),
      data = data,
      x = TRUE,
      y = TRUE
    )
    estimates[[position]] <- data.frame(
      imputation = index,
      threshold = threshold,
      comparison = paste0("Y > ", threshold),
      estimate = unname(stats::coef(binary_fit)["tx"]),
      variance = unname(stats::vcov(binary_fit)["tx", "tx"])
    )
    position <- position + 1L
  }
}
options(datadist = NULL)
estimates <- do.call(rbind, estimates)

pooled <- do.call(rbind, lapply(split(estimates, estimates$comparison), function(data) {
  m <- nrow(data)
  q_bar <- mean(data$estimate)
  u_bar <- mean(data$variance)
  between <- stats::var(data$estimate)
  total <- u_bar + (1 + 1 / m) * between
  degrees_freedom <- if (is.finite(between) && between > 0) {
    (m - 1) * (1 + u_bar / ((1 + 1 / m) * between))^2
  } else {
    Inf
  }
  critical_value <- if (is.finite(degrees_freedom)) {
    stats::qt(0.975, df = degrees_freedom)
  } else {
    stats::qnorm(0.975)
  }
  data.frame(
    threshold = data$threshold[[1L]],
    comparison = data$comparison[[1L]],
    log_odds_ratio = q_bar,
    standard_error = sqrt(total),
    conf_low = q_bar - critical_value * sqrt(total),
    conf_high = q_bar + critical_value * sqrt(total),
    odds_ratio = exp(q_bar),
    odds_ratio_low = exp(q_bar - critical_value * sqrt(total)),
    odds_ratio_high = exp(q_bar + critical_value * sqrt(total)),
    imputations = m
  )
}))
rownames(pooled) <- NULL
pooled$order <- ifelse(is.na(pooled$threshold), 0L, pooled$threshold)
pooled <- pooled[order(pooled$order), ]

# Keep pooled unchanged for the results table.
# Create separate data frames just for plotting.
binary_pooled <- pooled[!is.na(pooled$threshold), ]
common_pooled <- pooled[is.na(pooled$threshold), ]

# Use only the seven binary thresholds on the y-axis
binary_pooled$comparison <- factor(
  binary_pooled$comparison,
  levels = rev(binary_pooled$comparison)
)

# Repeat the common proportional-odds estimate at every binary threshold
common_repeated <- binary_pooled

common_repeated$log_odds_ratio <- common_pooled$log_odds_ratio[[1L]]
common_repeated$conf_low <- common_pooled$conf_low[[1L]]
common_repeated$conf_high <- common_pooled$conf_high[[1L]]

comparison_plot <- ggplot2::ggplot() +
  ggplot2::geom_vline(
    xintercept = 0,
    colour = "grey65",
    linewidth = 0.5
  ) +

  # Threshold-specific binary ORs
  ggplot2::geom_errorbarh(
    data = binary_pooled,
    ggplot2::aes(
      y = comparison,
      xmin = conf_low,
      xmax = conf_high
    ),
    height = 0.12,
    position = ggplot2::position_nudge(y = 0.08),
    colour = "black"
  ) +
  ggplot2::geom_point(
    data = binary_pooled,
    ggplot2::aes(
      x = log_odds_ratio,
      y = comparison
    ),
    position = ggplot2::position_nudge(y = 0.08),
    size = 2.5,
    colour = "black"
  ) +

  # Common proportional-odds estimate, repeated at each threshold
  ggplot2::geom_errorbarh(
    data = common_repeated,
    ggplot2::aes(
      y = comparison,
      xmin = conf_low,
      xmax = conf_high
    ),
    height = 0.12,
    position = ggplot2::position_nudge(y = -0.08),
    colour = "steelblue"
  ) +
  ggplot2::geom_point(
    data = common_repeated,
    ggplot2::aes(
      x = log_odds_ratio,
      y = comparison
    ),
    position = ggplot2::position_nudge(y = -0.08),
    size = 2.5,
    colour = "steelblue"
  ) +

  ggplot2::labs(
    x = "Treatment log odds ratio for a worse outcome",
    y = "Binary threshold"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.grid.major.y = ggplot2::element_blank()
  )

result_dir <- here::here(
  "results", "primary", "quality-of-life", "proportional-odds-check"
)
figure_dir <- here::here("artifacts", "primary", "quality-of-life", "figures")
dir.create(result_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
arrow::write_parquet(estimates, file.path(result_dir, "imputation-estimates.parquet"))
arrow::write_parquet(pooled, file.path(result_dir, "pooled-estimates.parquet"))
ggplot2::ggsave(
  file.path(figure_dir, "common-vs-binary-log-odds-ratios.png"),
  comparison_plot,
  width = 8,
  height = 5.5,
  dpi = 300
)
```

```{r}
#| label: tbl-qol-proportional-odds-check
#| tbl-cap: "Pooled common and threshold-specific treatment odds ratios for a worse six-month outcome"
#| echo: false

knitr::kable(pooled[c(
  "comparison", "odds_ratio", "odds_ratio_low", "odds_ratio_high",
  "log_odds_ratio", "conf_low", "conf_high"
)])
```

```{r}
#| label: fig-qol-proportional-odds-check
#| fig-cap: "Common and threshold-specific treatment log odds ratios pooled across 50 completed datasets"
#| fig-alt: "A horizontal dot-and-interval plot compares the common proportional-odds treatment coefficient with seven binary logistic treatment coefficients for progressively worse QoL thresholds."
#| fig-width: 8
#| fig-height: 5.5
#| echo: false

comparison_plot
```
