---
title: "Overall survival — primary preparation"
subtitle: "Locked cohort and 182-day survival endpoint"
format: html
execute:
  freeze: false
---

This module constructs the schema-validated overall-survival analysis data. It
uses the fixed data lock from `config/analysis.yml`; it does not infer a lock
date from the machine clock. The notebook stops with an actionable error when
the private cohort input is absent or lacks a SAP covariate.

```{r}
#| label: setup
#| include: false
source(here::here("R", "analysis-config.R"))
source(here::here("R", "paths-artifacts.R"))
source(here::here("R", "checkpoints.R"))
source(here::here("R", "data-validation.R"))
source(here::here("R", "data-helpers.R"))
source(here::here("R", "imputation.R"))
source(here::here("R", "survival-helpers.R"))
source(here::here("analysis", "module-runtime.R"))

runtime <- module_runtime()
config <- runtime$config
input_path <- project_path("analysis", "00-data", "results", "analysis-cohort.parquet")
schema_path <- project_path("data", "schema", "cohort_follow_up.yml")
output_path <- project_path("artifacts", "primary", "overall-survival", "prepared.parquet")
stage_id <- "primary-os-preparation"
code_paths <- module_code_paths(
  here::here("R", "analysis-config.R"), here::here("R", "data-helpers.R"),
  here::here("R", "data-validation.R"), here::here("R", "survival-helpers.R"),
  here::here("R", "checkpoints.R"),
  here::here("analysis", "module-runtime.R"), knitr::current_input()
)
```

```{r}
#| label: prepare-os-data
#| include: false
module_branch(
  runtime = runtime,
  stage_id = stage_id,
  input_paths = c(input_path, schema_path),
  output_paths = output_path,
  code_paths = code_paths,
  metadata = list(endpoint = "overall_survival", horizon_days = 182L),
  worker = function() {
    cohort <- module_input(input_path, schema_path)
    module_require_columns(
      cohort,
      c("id", "randomization_date", "tx", "death_date", "follow_up_end_date",
        "ecog_fstcnt", "albumin", "c_reactive_protein", "stage_binary", "lung"),
      "cohort_follow_up"
    )
    if (!all(validate_binary(cohort$tx, "tx", allow_missing = FALSE) %in% c(0, 1))) {
      stop("cohort_follow_up$tx must use the SAP coding 0 = comparator, 1 = SAT invitation.")
    }
    if (any(!is.na(cohort$stage_binary) & !cohort$stage_binary %in% 0:1)) {
      stop("stage_binary must be coded 0/1.")
    }
    if (any(!is.na(cohort$lung) & !cohort$lung %in% 0:1)) {
      stop("lung must be coded 0/1.")
    }
    prepared <- derive_survival_endpoint(
      cohort,
      origin_col = "randomization_date",
      death_date_col = "death_date",
      follow_up_end_col = "follow_up_end_date",
      lock_date = parse_analysis_date(config$analysis$data_lock, "data_lock", allow_missing = FALSE),
      horizon_days = as.integer(config$analysis$horizon_days)
    )
    prepared$ecog_binary <- derive_ecog_binary(prepared$ecog_fstcnt)
    prepared$mgps <- derive_mgps(prepared$albumin, prepared$c_reactive_protein)
    assert_ecog_binary_consistent(prepared)
    assert_mgps_consistent(prepared)
    prepared <- prepared[
      , c("id", "tx", "survival_time_days", "event_death", "ecog_binary",
          "ecog_fstcnt", "stage_binary", "lung", "mgps", "age", "gender", "albumin",
          "c_reactive_protein"), drop = FALSE
    ]
    if (anyDuplicated(prepared$id)) stop("Prepared OS data contains duplicate patient IDs.")
    module_save_parquet(prepared, output_path)
  }
)
```

The prepared artifact is intentionally stored under ignored `artifacts/`; it
contains no fabricated patient records and is the only input accepted by the
downstream OS stages.
