---
title: "Overall survival — primary imputation"
subtitle: "One MICE run with 50 completed datasets"
format: html
execute:
  freeze: false
---

Albumin and C-reactive protein are imputed separately. mGPS is recalculated
passively from the completed laboratory values. The ordered ECOG score is
imputed with a proportional-odds model, after which binary ECOG is recalculated
passively from `ecog_fstcnt`.

```{r}
#| label: impute-os
#| include: false
library(mice)

source(here::here("R", "analysis-config.R"))
source(here::here("R", "survival-helpers.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
input_path <- here::here(
  "artifacts", "primary", "overall-survival", "prepared.parquet"
)
output_path <- here::here(
  "artifacts", "primary", "overall-survival", "imputations.rds"
)
dir.create(dirname(output_path), recursive = TRUE, showWarnings = FALSE)

if (force || !file.exists(output_path)) {
  prepared <- arrow::read_parquet(input_path) |>
    as.data.frame()
  prepared$ecog_fstcnt <- ordered(prepared$ecog_fstcnt, levels = 0:4)
  prepared$na_est <- mice::nelsonaalen(
    prepared,
    timevar = survival_time_days,
    statusvar = event_death
  )
  prepared <- prepared[c(
    setdiff(names(prepared), c("ecog_binary", "mgps")),
    "ecog_binary", "mgps"
  )]
  mice::md.pattern(prepared)

  methods <- mice::make.method(prepared)
  predictors <- mice::make.predictorMatrix(prepared)

  methods[c("id", "tx", "survival_time_days", "event_death", "na_est")] <- ""
  methods["ecog_fstcnt"] <- "polr"
  methods["ecog_binary"] <- "~I(derive_ecog_binary(ecog_fstcnt))"
  methods["mgps"] <- "~I(derive_mgps(albumin, c_reactive_protein))"
  predictors[c("id", "tx", "survival_time_days", "event_death", "na_est"), ] <- 0
  predictors[, c(
    "id", "survival_time_days", "ecog_binary", "mgps"
  )] <- 0

  imputation <- mice::mice(
    prepared,
    method = methods,
    predictorMatrix = predictors,
    m = config$imputation$main$m,
    maxit = config$imputation$main$maxit,
    donors = config$imputation$main$donors,
    seed = analysis_seed(config, "imputation"),
    printFlag = FALSE
  )
  saveRDS(imputation, output_path, compress = "xz")
}

imputation <- readRDS(output_path)
```


Check convergence

```{r}
#| label: fig-os-imputation-traces
#| fig-cap: "MICE trace plots for directly imputed overall-survival variables"
#| fig-width: 10
#| fig-height: 8
#| echo: false

imputed_variables <- names(imputation$method)[
  nzchar(imputation$method) & !startsWith(imputation$method, "~")
]

plot(
  imputation,
  # y = imputed_variables,
  layout = c(2, 5)
)
```

Compare the distributions of observed and imputed values. Density plots are
used for the continuous laboratory measurements, whereas the ordered ECOG
categories are compared with a strip plot.

```{r}
#| label: fig-os-imputation-distributions-continuous
#| fig-cap: "Observed and imputed distributions of OS imputation variables"
#| fig-width: 10
#| fig-height: 5
#| echo: false

densityplot(
  imputation,
  ~ albumin + bilirubin + c_reactive_protein + lactate_dehydrogenase +
    ecog_fstcnt + mgps,
  layout = c(2, 3),
  # plot.points = TRUE
)
```

```{r}
#| label: fig-os-imputation-distributions-ecog
#| fig-cap: "Observed and imputed distributions of the ordered ECOG score"
#| fig-width: 12
#| fig-height: 5
#| echo: false

stripplot(
  imputation,
  ecog_fstcnt ~ .imp,
  pch = 20,
  cex = 0.6
)
```
