---
title: "Overall survival — primary diagnostics"
subtitle: "Posterior diagnostics for 50 Bayesian survival fits"
format: html
execute:
  freeze: false
---

```{r}
#| label: check-os-models
#| include: false

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")
indices <- seq_len(config$imputation$main$m)
model_dir <- here::here("artifacts", "primary", "overall-survival", "models")
diagnostic_dir <- here::here(
  "artifacts", "primary", "overall-survival", "diagnostics"
)
trace_dir <- file.path(diagnostic_dir, "trace-plots")
figure_dir <- here::here("artifacts", "primary", "overall-survival", "figures")
result_path <- here::here(
  "results", "primary", "overall-survival", "diagnostics.parquet"
)
dir.create(diagnostic_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(trace_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(result_path), recursive = TRUE, showWarnings = FALSE)

for (index in indices) {
  model_path <- file.path(model_dir, sprintf("model-%02d.rds", index))
  output_path <- file.path(
    diagnostic_dir,
    sprintf("diagnostics-%02d.parquet", index)
  )
  if (!force && file.exists(output_path)) next

  diagnostics <- posterior::summarise_draws(
    posterior::as_draws_df(readRDS(model_path))
  ) |>
    as.data.frame()
  diagnostics$imputation <- index
  if (
    any(diagnostics$rhat[is.finite(diagnostics$rhat)] >
      config$diagnostics$max_rhat) ||
      any(diagnostics$ess_bulk[is.finite(diagnostics$ess_bulk)] <
        config$diagnostics$min_ess_bulk) ||
      any(diagnostics$ess_tail[is.finite(diagnostics$ess_tail)] <
        config$diagnostics$min_ess_tail)
  ) {
    stop("OS posterior diagnostics failed for imputation ", index, ".")
  }
  arrow::write_parquet(diagnostics, output_path)
}

diagnostic_paths <- file.path(
  diagnostic_dir,
  sprintf("diagnostics-%02d.parquet", indices)
)
diagnostics <- lapply(diagnostic_paths, function(path) {
  arrow::read_parquet(path) |>
    as.data.frame()
})
arrow::write_parquet(do.call(rbind, diagnostics), result_path)
```

One trace plot is saved for each fitted survival model. These files are retained
as diagnostic artifacts and are not automatically inserted into the main
report.

```{r}
#| label: save-os-trace-plots
#| include: false

for (index in indices) {
  trace_path <- file.path(trace_dir, sprintf("trace-%02d.png", index))
  if (!force && file.exists(trace_path)) next

  model <- readRDS(file.path(model_dir, sprintf("model-%02d.rds", index)))
  draws <- posterior::as_draws_array(model)
  trace_variables <- posterior::variables(draws)
  trace_variables <- trace_variables[grepl(
    "^(tx|ecog_binary|stage_binary|lung|mgps)",
    trace_variables
  )]
  if (length(trace_variables) == 0L) {
    stop("No OS regression parameters were found for trace plotting.")
  }
  trace_plot <- bayesplot::mcmc_trace(draws, pars = trace_variables)
  ggplot2::ggsave(
    trace_path,
    trace_plot,
    width = 10,
    height = max(5, 1.1 * length(trace_variables)),
    dpi = 180,
    limitsize = FALSE
  )
}
```

The observed Kaplan–Meier curves use the complete randomized cohort, retain
censoring marks, and are overlaid with the pooled standardized posterior
survival curves. Model-based diagnostics use participants with positive
follow-up time. The Cox–Snell check is
calculated directly from posterior survival probabilities from the first
imputation's `stan_surv` fit. A companion Cox model with the same linear
predictor is used only for supplemental scaled Schoenfeld and deviance-residual
checks. These checks do not replace or refit the Bayesian primary estimand.

```{r}
#| label: derive-os-graphical-diagnostics
#| include: false

prepared <- arrow::read_parquet(here::here(
  "artifacts", "primary", "overall-survival", "prepared.parquet"
)) |>
  as.data.frame()
curve_draws <- arrow::read_parquet(here::here(
  "results", "primary", "overall-survival", "survival-curves.parquet"
)) |>
  as.data.frame()

prepared$tx_plot <- factor(
  prepared$tx,
  levels = 0:1,
  labels = c("Control", "Intervention")
)
km_fit <- ggsurvfit::survfit2(
  survival::Surv(survival_time_days, event_death) ~ tx_plot,
  data = prepared,
  conf.type = "log-log"
)

curve_summary <- do.call(rbind, lapply(
  split(curve_draws, interaction(curve_draws$tx, curve_draws$time, drop = TRUE)),
  function(data) {
    data.frame(
      tx = data$tx[[1L]],
      time = data$time[[1L]],
      median = median(data$survival),
      lower = quantile(data$survival, 0.025, names = FALSE),
      upper = quantile(data$survival, 0.975, names = FALSE)
    )
  }
))
rownames(curve_summary) <- NULL
curve_summary$tx <- factor(
  curve_summary$tx,
  levels = 0:1,
  labels = c("Control", "Intervention")
)
risk_times <- c(
  seq(0, config$analysis$horizon_days - 30, by = 30),
  config$analysis$horizon_days
)

survival_plot <- ggsurvfit::ggsurvfit(
  km_fit,
  linewidth = 0.55,
  linetype = 2
) +
  ggplot2::geom_ribbon(
    data = curve_summary,
    ggplot2::aes(x = time, ymin = lower, ymax = upper, fill = tx),
    alpha = 0.16,
    inherit.aes = FALSE
  ) +
  ggplot2::geom_line(
    data = curve_summary,
    ggplot2::aes(x = time, y = median, colour = tx),
    linewidth = 0.9,
    inherit.aes = FALSE
  ) +
  ggsurvfit::add_censor_mark(size = 1.8) +
  ggplot2::scale_x_continuous(
    breaks = risk_times
  ) +
  ggplot2::coord_cartesian(
    xlim = c(0, config$analysis$horizon_days),
    ylim = c(0, 1)
  ) +
  ggplot2::labs(
    x = "Days since randomization",
    y = "Survival probability",
    colour = NULL,
    fill = NULL
  ) +
  ggplot2::theme_bw() +
  ggsurvfit::add_risktable(
    times = risk_times,
    risktable_stats = "n.risk",
    stats_label = list(n.risk = "At risk")
  )

imputations <- readRDS(here::here(
  "artifacts", "primary", "overall-survival", "imputations.rds"
))
diagnostic_data <- mice::complete(imputations, 1L)
diagnostic_data$mgps <- factor(diagnostic_data$mgps, levels = 0:2)
# Match the primary survival-model fitting population.
diagnostic_data <- diagnostic_data[diagnostic_data$survival_time_days > 0, ]
cox_model <- survival::coxph(
  survival::Surv(survival_time_days, event_death) ~
    tx + ecog_binary + stage_binary + lung + mgps,
  data = diagnostic_data,
  x = TRUE,
  y = TRUE
)
zph <- survival::cox.zph(cox_model, transform = "km")
zph_table <- as.data.frame(zph$table)
zph_table$term <- rownames(zph_table)
rownames(zph_table) <- NULL

posterior_model <- readRDS(file.path(model_dir, "model-01.rds"))
posterior_survival <- rstanarm::posterior_survfit(
  posterior_model,
  newdata = diagnostic_data,
  times = "survival_time_days",
  extrapolate = FALSE,
  draws = config$diagnostics$posterior_draws,
  seed = analysis_seed(config, "diagnostics", 1L),
  return_matrix = TRUE
)
survival_draws <- as.matrix(posterior_survival[[1L]])
if (ncol(survival_draws) != nrow(diagnostic_data)) {
  stop("Posterior survival predictions do not match the diagnostic dataset.")
}
cox_snell_residuals <- -log(pmax(survival_draws, .Machine$double.xmin))
residual_grid <- seq(
  0,
  stats::quantile(cox_snell_residuals, 0.99, names = FALSE),
  length.out = 200L
)

cox_snell_hazards <- vapply(
  seq_len(nrow(cox_snell_residuals)),
  function(draw) {
    residual_fit <- survival::survfit(
      survival::Surv(
        cox_snell_residuals[draw, ],
        diagnostic_data$event_death
      ) ~ 1,
      stype = 2,
      ctype = 1
    )
    stats::approx(
      x = c(0, residual_fit$time),
      y = c(0, residual_fit$cumhaz),
      xout = residual_grid,
      method = "constant",
      f = 0,
      rule = 2
    )$y
  },
  numeric(length(residual_grid))
)
cox_snell_data <- data.frame(
  residual = residual_grid,
  median = apply(cox_snell_hazards, 1L, stats::median),
  lower = apply(
    cox_snell_hazards,
    1L,
    stats::quantile,
    probs = 0.025,
    names = FALSE
  ),
  upper = apply(
    cox_snell_hazards,
    1L,
    stats::quantile,
    probs = 0.975,
    names = FALSE
  )
)
cox_snell_plot <- ggplot2::ggplot(
  cox_snell_data,
  ggplot2::aes(x = residual, y = median)
) +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = lower, ymax = upper),
    fill = "steelblue",
    alpha = 0.2
  ) +
  ggplot2::geom_line(colour = "steelblue4", linewidth = 0.8) +
  ggplot2::geom_abline(intercept = 0, slope = 1, linetype = 2, colour = "grey45") +
  ggplot2::labs(
    x = "Posterior Cox–Snell residual",
    y = "Nelson–Aalen cumulative hazard"
  ) +
  ggplot2::theme_bw()

deviance_data <- data.frame(
  linear_predictor = stats::predict(cox_model, type = "lp"),
  deviance_residual = stats::residuals(cox_model, type = "deviance")
)
deviance_plot <- ggplot2::ggplot(
  deviance_data,
  ggplot2::aes(x = linear_predictor, y = deviance_residual)
) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey65") +
  ggplot2::geom_point(alpha = 0.65) +
  ggplot2::geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  ggplot2::labs(x = "Companion Cox-model linear predictor", y = "Deviance residual") +
  ggplot2::theme_bw()

ggplot2::ggsave(
  file.path(figure_dir, "kaplan-meier-vs-posterior-model.png"),
  ggsurvfit::ggsurvfit_build(survival_plot),
  width = 8,
  height = 6.5,
  dpi = 300
)
ggplot2::ggsave(
  file.path(figure_dir, "cox-snell-residuals.png"),
  cox_snell_plot,
  width = 6.5,
  height = 5,
  dpi = 300
)
ggplot2::ggsave(
  file.path(figure_dir, "deviance-residuals.png"),
  deviance_plot,
  width = 6.5,
  height = 5,
  dpi = 300
)
grDevices::png(
  file.path(figure_dir, "scaled-schoenfeld-residuals.png"),
  width = 1800,
  height = 1800,
  res = 180
)
plot(zph)
grDevices::dev.off()
arrow::write_parquet(
  zph_table,
  here::here("results", "primary", "overall-survival", "ph-diagnostics.parquet")
)
arrow::write_parquet(
  cox_snell_data,
  here::here(
    "results", "primary", "overall-survival",
    "cox-snell-posterior-summary.parquet"
  )
)
```

```{r}
#| label: fig-os-km-model-check
#| fig-cap: "Observed Kaplan–Meier estimates, numbers at risk, and pooled standardized posterior survival curves"
#| fig-alt: "Survival plot by randomized group with an aligned number-at-risk table. Dashed steps show Kaplan–Meier estimates with censoring marks; solid lines and bands show median Bayesian model predictions and 95 percent credible intervals."
#| fig-width: 8
#| fig-height: 6.5
#| echo: false

survival_plot
```

```{r}
#| label: fig-os-residual-checks
#| fig-cap: "Bayesian survival-model Cox–Snell and supplemental companion-Cox deviance residual checks"
#| fig-subcap:
#|   - "Posterior Cox–Snell cumulative-hazard check"
#|   - "Companion-Cox deviance residuals against the linear predictor"
#| layout-ncol: 2
#| fig-width: 6.5
#| fig-height: 5
#| echo: false

cox_snell_plot
deviance_plot
```
