---
title: "LIQPLAT analysis datasets"
format: html
---


```{r}
#| label: setup
#| message: false

library(arrow)
library(here)
library(knitr)
library(yaml)
library(tidyverse)

config <- read_yaml(here("config", "analysis.yml"))
source(here("R", "taooh.R"))

private_dir <- here("data", "private")
results_dir <- here("analysis", "00-data", "results")
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

domains <- c(
  "consort_ipd",
  "recruitment",
  "cohort_follow_up",
  "qol",
  "taooh",
  "ctdna_samples",
  "ctdna_analyses",
  "alterations"
)
```

## Data lock and file set

```{r}
#| label: manifest

manifest <- read_yaml(here(private_dir, "manifest.yml"))

stopifnot(
  "The manifest data lock differs from config/analysis.yml." =
    identical(as.character(manifest$data_lock), config$analysis$data_lock),
  "The manifest must describe exactly the eight analysis datasets." =
    setequal(names(manifest$domains), domains),
  "The manifest marks a dataset as unavailable." =
    all(map_lgl(manifest$domains, \(record) isTRUE(record$available)))
)

observed_files <- list.files(
  private_dir,
  pattern = "\\.parquet$",
  ignore.case = TRUE
)

if (!setequal(observed_files, paste0(domains, ".parquet"))) {
  stop(
    "data/private must contain exactly: ",
    str_flatten_comma(paste0(domains, ".parquet")),
    ". Found: ",
    str_flatten_comma(observed_files)
  )
}
```

## Read the private datasets

```{r}
#| label: read

private <- domains |>
  set_names() |>
  map(\(domain) {
    read_parquet(here(private_dir, paste0(domain, ".parquet"))) |>
      as.data.frame()
  })

schemas <- domains |>
  set_names() |>
  map(\(domain) read_yaml(here("data", "schema", paste0(domain, ".yml"))))
```

## Column, key and identifier checks

```{r}
#| label: structure-checks

# Columns used in this notebook that the schemas do not mark as required.
build_columns <- list(
  recruitment = c("informed_consent_date", "randomization_date", "group_assignment"),
  cohort_follow_up = c(
    "id", "randomization_date", "tx", "diagnosis", "death_date",
    "follow_up_end_date", "status_death", "survival_time_days_unrestricted",
    "status_death_unrestricted", "potential_follow_up_days",
    "invitation_offered", "invitation_accepted", "enrolled_date",
    "last_known_alive_date", "mtb_date_first", "mtb_n_reports",
    "n_blood_products", "n_erythrocyte_concentrates",
    "n_thrombocyte_concentrates"
  ),
  qol = c("id", "assessment_id", "day", "q30", "baseline_q30"),
  taooh = c("id", "week", "y_taooh"),
  ctdna_samples = c("sample_id", "id", "sample_date", "valid_ctdna_result"),
  ctdna_analyses = c(
    "analysis_id", "sample_id", "id", "analysis_date", "lod_q1", "lod_q3",
    "valid_ctdna_result", "n_mut"
  ),
  alterations = c(
    "alteration_id", "analysis_id", "sample_id", "id",
    "found_in_solid_biopsy", "found_in_buffy_coat"
  )
)

walk(domains, \(domain) {
  data <- private[[domain]]
  schema <- schemas[[domain]]

  required <- schema$columns |>
    keep(\(column) isTRUE(column$required)) |>
    names() |>
    union(build_columns[[domain]])
  missing_columns <- setdiff(required, names(data))
  if (length(missing_columns) > 0) {
    stop(domain, " is missing: ", str_flatten_comma(missing_columns))
  }

  if (isTRUE(schema$exact_columns) && !setequal(names(data), names(schema$columns))) {
    stop(domain, " must contain exactly the schema columns.")
  }

  key <- as.character(schema$primary_key)
  if (length(key) > 0 &&
      (!all(complete.cases(data[key])) || anyDuplicated(data[key]) > 0)) {
    stop(domain, " has a missing or duplicated primary key.")
  }
})

recruitment <- private$recruitment
cohort <- private$cohort_follow_up
qol <- private$qol
taooh <- private$taooh
ctdna_samples <- private$ctdna_samples
ctdna_analyses <- private$ctdna_analyses
alterations <- private$alterations
```

## Content checks


```{r}
#| label: content-checks

stopifnot(
  "Recruitment group_assignment must be 0/1 or missing." =
    is.na(recruitment$group_assignment) | recruitment$group_assignment %in% 0:1
)

validate_taooh_follow_up(taooh, cohort)

stopifnot(
  "QoL export extends beyond the analysis horizon." =
    !any(qol$day > config$analysis$horizon_days, na.rm = TRUE)
)

blood_fields <- c(
  "n_blood_products",
  "n_erythrocyte_concentrates",
  "n_thrombocyte_concentrates"
)
blood_complete <- cohort[complete.cases(cohort[blood_fields]), ]

stopifnot(
  "Blood-product counts must be nonnegative integers or missing." =
    map_lgl(cohort[blood_fields], \(x) {
      is.numeric(x) && all(is.na(x) | (is.finite(x) & x >= 0 & x == floor(x)))
    }),
  "Combined blood-product counts differ from the sum of components." =
    blood_complete$n_blood_products ==
      blood_complete$n_erythrocyte_concentrates +
      blood_complete$n_thrombocyte_concentrates
)
```

## Cohort

```{r}
#| label: cohort

cohort_dates <- c(
  "randomization_date",
  "death_date",
  "enrolled_date",
  "last_known_alive_date",
  "mtb_date_first",
  "follow_up_end_date"
)

stopifnot(
  "Cohort date columns must be Date." =
    map_lgl(cohort[cohort_dates], \(x) inherits(x, "Date")),
  "randomization_date contains missing dates." =
    !anyNA(cohort$randomization_date),
  "follow_up_end_date contains missing dates." =
    !anyNA(cohort$follow_up_end_date),
  "tx must be coded 0/1 without missing values." =
    cohort$tx %in% 0:1,
  "status_death must be coded 0/1 without missing values." =
    cohort$status_death %in% 0:1
)

cohort <- cohort |>
  mutate(
    tx = as.numeric(tx),
    survival_time_days = pmin(
      survival_time_days_unrestricted,
      config$analysis$horizon_days
    ),
    event_death = as.numeric(
      status_death_unrestricted == 1 &
        survival_time_days_unrestricted <= config$analysis$horizon_days
    )
  )

exported_follow_up <- as.numeric(
  cohort$follow_up_end_date - cohort$randomization_date
)

stopifnot(
  "The derived and exported 182-day deaths differ." =
    cohort$event_death == cohort$status_death,
  "The derived 182-day follow-up is shorter than the exported one." =
    cohort$survival_time_days >= exported_follow_up
)
```

- 182-day endpoint truncates unrestricted follow-up; the export censored
  `r sum(cohort$survival_time_days > exported_follow_up)` of these participants earlier.

## Invitation

- Offered = answer recorded; the exported `invitation_offered` disagrees and is replaced.

```{r}
#| label: invitation

stopifnot(
  "Usual-care patients cannot have an invitation answer." =
    all(is.na(cohort$invitation_accepted[cohort$tx == 0]))
)

cohort <- cohort |>
  mutate(
    invitation_offered = if_else(
      tx == 1,
      as.integer(!is.na(invitation_accepted)),
      NA_integer_
    )
  )
```

## ctDNA analyses

```{r}
#| label: ctdna-analyses

stopifnot(
  "ctDNA sample and analysis dates must be Date." =
    inherits(ctdna_samples$sample_date, "Date") &&
      inherits(ctdna_analyses$analysis_date, "Date")
)

ctdna_analyses <- merge(
  ctdna_analyses,
  ctdna_samples |> select(sample_id, sample_date),
  by = "sample_id",
  all.x = TRUE,
  sort = FALSE
)

stopifnot(
  "Every ctDNA analysis must refer to a sample with a recorded sample date." =
    !anyNA(ctdna_analyses$sample_date)
)

ctdna_analyses <- ctdna_analyses |>
  mutate(
    attempted_sample = !is.na(sample_date),
    ctdna_error_derived = as.integer(
      attempted_sample & is.na(lod_q1) & is.na(lod_q3)
    ),
    valid_ctdna_result_derived = attempted_sample & ctdna_error_derived == 0L
  )

stopifnot(
  "Exported and LOD-derived ctDNA analysis validity flags disagree." =
    ctdna_analyses$valid_ctdna_result == ctdna_analyses$valid_ctdna_result_derived
)
```

## ctDNA samples

```{r}
#| label: ctdna-samples

analysis_by_sample <- ctdna_analyses |>
  group_by(sample_id) |>
  summarise(
    analysis_date = if (all(is.na(analysis_date))) {
      as.Date(NA)
    } else {
      min(analysis_date, na.rm = TRUE)
    },
    n_mut = sum(n_mut[valid_ctdna_result_derived], na.rm = TRUE),
    analysis_attempted = any(attempted_sample),
    analysis_valid = any(valid_ctdna_result_derived),
    ctdna_error_qc = if (length(unique(ctdna_error[!is.na(ctdna_error)])) == 1L) {
      as.integer(unique(ctdna_error[!is.na(ctdna_error)]))
    } else {
      NA_integer_
    },
    .groups = "drop"
  )

ctdna_samples <- ctdna_samples |>
  left_join(analysis_by_sample, by = "sample_id")

stopifnot(
  "Every ctDNA sample must have at least one analysis." =
    !anyNA(ctdna_samples$analysis_valid),
  "Exported and analysis-derived sample validity flags disagree." =
    ctdna_samples$valid_ctdna_result == ctdna_samples$analysis_valid
)

ctdna_samples <- ctdna_samples |>
  mutate(
    attempted_sample = !is.na(sample_date),
    valid_ctdna_result = analysis_valid,
    ctdna_error_derived = as.integer(attempted_sample & !valid_ctdna_result)
  ) |>
  select(-analysis_attempted, -analysis_valid)
```

## Alterations

```{r}
#| label: alterations

alterations <- merge(
  alterations,
  ctdna_analyses |>
    select(analysis_id, sample_id, id, valid_ctdna_result_derived),
  by = "analysis_id",
  all.x = TRUE,
  suffixes = c("", "_analysis"),
  sort = FALSE
)

stopifnot(
  "Every alteration must refer to a ctDNA analysis in ctdna_analyses.parquet." =
    !anyNA(alterations$valid_ctdna_result_derived),
  "Alteration sample/patient links disagree with ctdna_analyses." =
    alterations$sample_id == alterations$sample_id_analysis &
      alterations$id == alterations$id_analysis,
  "Exported alteration and analysis validity flags disagree." =
    alterations$valid_ctdna_result == alterations$valid_ctdna_result_derived,
  "Every alteration must refer to a ctDNA sample in ctdna_samples.parquet." =
    alterations$sample_id %in% ctdna_samples$sample_id
)

alterations <- alterations |>
  mutate(valid_ctdna_result = valid_ctdna_result_derived) |>
  select(-sample_id_analysis, -id_analysis, -valid_ctdna_result_derived)

# merge() sorts by sample_id here (ties keep the order above).
alterations <- merge(
  alterations,
  ctdna_samples |> select(sample_id, ctdna_detected_from_sample = ctdna_detected),
  by = "sample_id",
  all.x = TRUE
) |>
  mutate(ctdna_detected = ctdna_detected_from_sample) |>
  select(-ctdna_detected_from_sample)
```

## Write

```{r}
#| label: write

analysis_datasets <- list(
  "analysis-recruitment" = recruitment,
  "analysis-cohort" = cohort,
  "analysis-qol" = qol,
  "analysis-taooh" = taooh,
  "analysis-ctdna-samples" = ctdna_samples,
  "analysis-ctdna-analyses" = ctdna_analyses,
  "analysis-alterations" = alterations
)

iwalk(analysis_datasets, \(data, name) {
  write_parquet(
    data,
    here(results_dir, paste0(name, ".parquet")),
    compression = "zstd"
  )
})
```

```{r}
#| label: tbl-datasets
#| tbl-cap: "Analysis datasets"
#| echo: false

tibble(
  dataset = names(analysis_datasets),
  n_rows = map_int(analysis_datasets, nrow),
  n_columns = map_int(analysis_datasets, ncol)
) |>
  kable()
```
