---
title: "TAOOH first-order — computationally reduced supporting/sensitivity analysis (5 imputations)"
format: html
execute:
  freeze: false
---

This computationally reduced supporting/sensitivity analysis (5 imputations)
fits a first-order TAOOH transition model as a robustness check against the
primary second-order specification. It uses fixed imputations 1–5, 200 draws
per imputation, and posterior standardization over observed participants using
patient-specific random-intercept draws (not empirical Bayes).

```{r}
#| label: setup
#| include: false
source(here::here("R", "analysis-config.R"))
source(here::here("R", "paths-artifacts.R"))
source(here::here("R", "checkpoints.R"))
source(here::here("R", "markov-helpers.R"))
source(here::here("analysis", "module-runtime.R"))
runtime <- module_runtime()
config <- runtime$config
indices <- module_exact_indices(config, supporting = TRUE)
if (!identical(indices, 1:5)) stop("Supporting TAOOH first-order requires exactly imputations 1:5.")
data_dir <- project_path("artifacts", "primary", "taooh", "imputations")
output_dir <- ensure_directory(project_path("artifacts", "supporting", "taooh-first-order"))
code_paths <- module_code_paths(
  here::here("R", "analysis-config.R"), here::here("R", "markov-helpers.R"),
  here::here("R", "checkpoints.R"),
  here::here("analysis", "module-runtime.R"), knitr::current_input()
)
```

```{r}
#| label: fit-taooh-first-order
#| include: false
for (index in indices) {
  data_path <- file.path(data_dir, sprintf("imputation-%02d.parquet", index))
  model_path <- file.path(output_dir, sprintf("model-%02d.rds", index))
  sops_path <- file.path(output_dir, sprintf("sops-%02d.rds", index))
  comparison_path <- file.path(output_dir, sprintf("comparisons-%02d.rds", index))
  if (!file.exists(data_path)) stop("Missing fixed TAOOH imputation ", index, ".")
  module_branch(
    runtime = runtime,
    stage_id = sprintf("supporting-taooh-first-order-%02d", index),
    input_paths = data_path,
    output_paths = c(model_path, sops_path, comparison_path),
    code_paths = code_paths,
    metadata = list(label = "computationally reduced supporting/sensitivity analysis (5 imputations)",
                   imputation = index, imputations = 1:5, posterior_draws = 200L,
                   order = 1L, state = 1L,
                   standardization = "observed participants; patient-specific random-intercept draws"),
    worker = function() {
      data <- module_read_parquet(data_path)
      data$y <- factor(data$y, levels = 1:5, ordered = TRUE)
      data$yprev <- factor(data$yprev, levels = 1:5, ordered = FALSE)
      dd <- rms::datadist(data)
      old_datadist <- getOption("datadist")
      options(datadist = "dd")
      on.exit(options(datadist = old_datadist), add = TRUE)
      model_formula <- y ~ tx + rms::rcs(week, 4) + yprev +
        ecog_fstcnt + factor(diagnosis) + rms::rcs(albumin, 3) +
        rms::rcs(c_reactive_protein, 3)
      fit_seed <- analysis_seed(config, "supporting", index)
      fitted <- tryCatch(
        markov.misc::blrm_markov(
          formula = model_formula,
          ppo = ~week,
          cppo = function(y) y,
          data = data,
          id_var = "id",
          time_var = "week",
          random_intercept = "id",
          p_var = "yprev",
          chains = as.integer(config$markov$chains),
          iter = as.integer(config$markov$iterations),
          warmup = as.integer(config$markov$warmup),
          seed = fit_seed,
          refresh = 0L
        ),
        error = identity
      )
      diagnostics <- if (inherits(fitted, "error")) {
        fitted
      } else {
        tryCatch(markov_diagnostics(fitted, config$diagnostics), error = identity)
      }
      retry <- inherits(fitted, "error") || inherits(diagnostics, "error") ||
        !isTRUE(diagnostics$passed)
      if (retry) {
        if (!isTRUE(config$diagnostics$retry_on_failure)) {
          stop("Supporting TAOOH first-order fit failed and retries are disabled.")
        }
        fitted <- markov.misc::blrm_markov(
          formula = model_formula,
          ppo = ~week,
          cppo = function(y) y,
          data = data,
          id_var = "id",
          time_var = "week",
          random_intercept = "id",
          p_var = "yprev",
          chains = as.integer(config$markov$chains),
          iter = as.integer(config$markov$retry_iterations),
          warmup = as.integer(config$markov$retry_iterations / 2L),
          seed = as.integer(fit_seed + 500000L),
          refresh = 0L
        )
        diagnostics <- markov_diagnostics(fitted, config$diagnostics)
      }
      assert_markov_diagnostics(diagnostics)
      observed <- data[!duplicated(data$id), c("id", "tx", "diagnosis", "ecog_fstcnt", "albumin", "c_reactive_protein", "yprev"), drop = FALSE]
      observed$yprev <- factor(observed$yprev, levels = 1:5, ordered = FALSE)
      sops <- markov.misc::avg_sops(
        model = fitted,
        newdata = observed,
        variables = list(tx = c(0, 1)),
        times = 1:26,
        y_levels = 1:5,
        absorb = 5L,
        p_var = "yprev",
        id_var = "id",
        time_var = "week",
        include_re = TRUE,
        n_draws = 200L,
        seed = fit_seed,
        posterior_summary = "mean",
        return_draws = TRUE
      )
      comparison <- markov.misc::avg_comparisons(
        model = fitted,
        newdata = observed,
        variables = list(tx = c(0, 1)),
        estimand = "time_in_state",
        state_sets = 1L,
        times = 1:26,
        y_levels = 1:5,
        absorb = 5L,
        id_var = "id",
        time_var = "week",
        p_var = "yprev",
        include_re = TRUE,
        n_draws = 200L,
        seed = fit_seed,
        posterior_summary = "mean",
        return_draws = TRUE
      )
      module_save_rds(fitted, model_path)
      module_save_rds(sops, sops_path)
      module_save_rds(comparison, comparison_path)
    }
  )
}
```
