---
title: "QoL six-month landmark: imputation"
subtitle: "Vital status at day 182 and MICE, one run per imputation index"
format: html
---

- Censored before day 182: vital status drawn from the survival model, given survival to censoring.

```{r}
#| label: setup
#| include: false

library(here)
library(arrow)
library(yaml)
library(rstanarm)
library(mice)
library(tidyverse)
source(here("R", "plot-style.R"))

config <- read_yaml(here("config", "analysis.yml"))

results_dir <- here(
  "analysis", "03-quality-of-life", "02-six-month-landmark", "results"
)
imputation_dir <- file.path(results_dir, "imputations")
figure_dir <- file.path(results_dir, "figures")
dir.create(imputation_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

vital_status_model_path <- file.path(results_dir, "vital-status-model.rds")
```

## Covariates

- Diagnosis lumped with `fct_lump_n()` (alphabetical levels), unlike the preparation grouping; kept as legacy.

```{r}
#| label: read-prepared
#| include: false

prepared <- read_parquet(
  here("analysis", "03-quality-of-life", "results", "prepared.parquet")
) |>
  as.data.frame() |>
  mutate(
    gender = factor(gender),
    age_scaled = (age - mean(age)) / sd(age),
    ecog_fstcnt = factor(if_else(ecog_fstcnt == 4, 3, ecog_fstcnt)),
    plan_fstcnt_coded = fct_lump_n(factor(plan_fstcnt_coded), n = 4),
    diagnosis_group = fct_lump_n(factor(diagnosis), n = 4)
  )
```

## Vital-status model

```{r}
#| label: fit-vital-status-model
#| include: false

if (!file.exists(vital_status_model_path)) {
  vital_status_data <- prepared |>
    filter(survival_time_days > 0)

  vital_status_terms <- colnames(model.matrix(
    ~ tx + age_scaled + gender + diagnosis_group + stage_binary,
    data = vital_status_data
  ))[-1L]

  # survival::Surv stays qualified: OS 03-death-at-26-weeks predicts from this
  # saved model without attaching survival.
  vital_status_model <- stan_surv(
    formula = survival::Surv(survival_time_days, event_death) ~
      tx + age_scaled + gender + diagnosis_group + stage_binary,
    data = vital_status_data,
    basehaz = "ms",
    basehaz_ops = list(df = 5),
    prior = normal(
      location = rep(0, length(vital_status_terms)),
      scale = ifelse(vital_status_terms == "tx", 0.55, 2.5)
    ),
    prior_intercept = normal(0, 20),
    prior_aux = dirichlet(1),
    chains = 4L,
    cores = 4L,
    iter = 2000L,
    warmup = 1000L,
    seed = config$seeds$qol + 100
  )

  saveRDS(vital_status_model, vital_status_model_path, compress = "xz")
}

vital_status_model <- readRDS(vital_status_model_path)
```

```{r}
#| label: check-vital-status-model
#| include: false

vital_status_diagnostics <- summary(vital_status_model) |>
  as.data.frame() |>
  rownames_to_column("parameter") |>
  relocate(parameter, .after = last_col())

rhat <- vital_status_diagnostics$Rhat
n_eff <- vital_status_diagnostics$n_eff

if (any(rhat[is.finite(rhat)] > config$diagnostics$max_rhat) ||
    any(n_eff[is.finite(n_eff)] < config$diagnostics$min_ess_bulk)) {
  stop("The dedicated QoL vital-status survival model did not converge.")
}

write_parquet(
  vital_status_diagnostics,
  file.path(results_dir, "vital-status-diagnostics.parquet"),
  compression = "zstd"
)
```

## Imputation, one index at a time

- `landmark_q30` is also imputed for deaths; it only acts as a predictor, never enters `y_landmark`.

```{r}
#| label: impute
#| include: false

unresolved <- prepared |>
  filter(
    event_death == 0L,
    survival_time_days < config$analysis$horizon_days
  )

for (index in seq_len(config$imputation$m_main)) {
  mids_path <- file.path(imputation_dir, sprintf("imputation-%02d.rds", index))
  completed_path <- file.path(imputation_dir, sprintf("completed-%02d.rds", index))
  draws_path <- file.path(
    imputation_dir,
    sprintf("vital-status-draws-%02d.parquet", index)
  )
  if (all(file.exists(c(mids_path, completed_path, draws_path)))) next

  # Day-182 survival probability for each unresolved participant, from one
  # posterior draw, conditional on survival to the censoring day.
  predicted_survival <- posterior_survfit(
    vital_status_model,
    newdata = unresolved,
    type = "surv",
    times = config$analysis$horizon_days,
    condition = TRUE,
    last_time = "survival_time_days",
    standardise = FALSE,
    extrapolate = FALSE,
    draws = 1L,
    seed = config$seeds$qol + (index - 1) * 1000 + 120,
    return_matrix = TRUE
  )
  conditional_survival <- as.numeric(predicted_survival[[1L]][1L, ])

  set.seed(config$seeds$qol + (index - 1) * 1000 + 140)
  vital_status_draws <- tibble(
    id = unresolved$id,
    imputation = index,
    censoring_day = unresolved$survival_time_days,
    conditional_survival = conditional_survival,
    death_probability = 1 - conditional_survival,
    death_182 = rbinom(
      n = nrow(unresolved),
      size = 1L,
      prob = death_probability
    )
  )

  # Column order sets the MICE visit sequence; keep it.
  imputation_data <- prepared |>
    left_join(
      vital_status_draws |> select(id, drawn_death = death_182),
      by = "id"
    ) |>
    mutate(death_182 = coalesce(drawn_death, event_death)) |>
    select(
      id, tx, age_scaled, gender, diagnosis_group, ecog_fstcnt,
      plan_fstcnt_coded, albumin, c_reactive_protein, baseline_q30,
      landmark_q30, death_182
    ) |>
    mutate(
      baseline_q30 = factor(baseline_q30, levels = 1:7, ordered = TRUE),
      landmark_q30 = factor(landmark_q30, levels = 1:7, ordered = TRUE)
    )

  methods <- make.method(imputation_data)
  predictors <- make.predictorMatrix(imputation_data)
  predictors[, "id"] <- 0L

  imputation <- mice(
    imputation_data,
    m = 1L,
    maxit = config$imputation$maxit,
    method = methods,
    predictorMatrix = predictors,
    donors = config$imputation$donors,
    seed = config$seeds$imputation + (index - 1) * 1000 + 200,
    printFlag = FALSE
  )

  completed <- complete(imputation, 1L) |>
    mutate(
      baseline_q30 = as.integer(as.character(baseline_q30)),
      landmark_q30 = as.integer(as.character(landmark_q30)),
      y_landmark = if_else(death_182 == 1L, 8L, landmark_q30)
    )

  stopifnot(
    "QoL imputation left a substantive-model variable missing." =
      !anyNA(completed$baseline_q30) && !anyNA(completed$y_landmark),
    "QoL imputation produced an invalid ordinal state." =
      all(completed$baseline_q30 %in% 1:7) && all(completed$y_landmark %in% 1:8)
  )

  write_parquet(vital_status_draws, draws_path, compression = "zstd")
  saveRDS(imputation, mids_path, compress = "xz")
  saveRDS(completed, completed_path, compress = "xz")
}
```

```{r}
#| label: read-imputations
#| include: false

indices <- seq_len(config$imputation$m_main)

imputation_fits <- map(
  indices,
  \(index) readRDS(file.path(imputation_dir, sprintf("imputation-%02d.rds", index)))
)
completed_datasets <- map(
  indices,
  \(index) readRDS(file.path(imputation_dir, sprintf("completed-%02d.rds", index)))
)

vital_status_draws <- indices |>
  map(\(index) {
    read_parquet(file.path(
      imputation_dir,
      sprintf("vital-status-draws-%02d.parquet", index)
    ))
  }) |>
  list_rbind()

write_parquet(
  vital_status_draws,
  file.path(results_dir, "vital-status-draws.parquet"),
  compression = "zstd"
)
```

## MICE diagnostics

```{r}
#| label: fig-mice-convergence
#| echo: false
#| fig-cap: "MICE convergence traces for the six-month landmark imputation: mean and standard deviation of the imputed values, one line per completed dataset (one MICE run each). Factors are summarised by their integer codes."
#| fig-width: 8
#| fig-height: 9

mice_stats <- mice_chain_stats(imputation_fits)
mice_trace_plot <- plot_mice_traces(mice_stats)

save_figure(
  mice_trace_plot,
  file.path(figure_dir, "mice-convergence-traces.png"),
  height = 1 + 1.3 * n_distinct(mice_stats$variable)
)

mice_trace_plot
```

```{r}
#| label: fig-observed-vs-imputed
#| echo: false
#| fig-cap: "Category proportions among observed values (black) and imputed values of each completed dataset (pink) for the six-month landmark imputation."
#| fig-width: 8
#| fig-height: 3.5

diagnostic_variables <- c(
  "Baseline Q30" = "baseline_q30",
  "Six-month Q30" = "landmark_q30",
  ECOG = "ecog_fstcnt"
)

# The runs differ in the drawn vital status, so ibind() cannot combine them:
# proportions per run (one completed dataset each), observed values once (they
# are the same in every run).
category_proportions <- imap(imputation_fits, \(fit, index) {
  imputed_category_proportions(fit, diagnostic_variables) |>
    filter(source == "Imputed" | index == 1) |>
    mutate(dataset = if_else(source == "Imputed", as.character(index), dataset))
}) |>
  list_rbind() |>
  # Numeric category order across variables (ECOG starts at 0).
  mutate(category = fct_inseq(category))

category_plot <- plot_imputed_categories(category_proportions)

save_figure(
  category_plot,
  file.path(figure_dir, "observed-vs-imputed.png"),
  height = 3.5
)

category_plot
```
