---
title: "Multiple Imputation Strategy for Overall Survival (OS) Outcome"
format: 
  typst:
    papersize: a4
    margin:
      x: 2cm
      y: 2.5cm
    toc: true
    columns: 1
    tbl-cap-location: bottom
---

## Overview

This document describes the multiple imputation strategy for the Overall Survival (OS) outcome in the LIQPLAT trial. The approach uses Multiple Imputation by Chained Equations (MICE) to handle missing baseline covariate data under the Missing At Random (MAR) assumption.

Since OS data are structured in wide format (one row per patient), we use standard single-level imputation methods appropriate for each variable type.

## Setup

```{r}
#| output: false
library(tidyverse)
library(here)
library(mice)
library(arrow)
```

## Data

The following code uses historical data from the Division of Medical Oncology at USB (first appointments from 2023 to mid-2024, prior to LIQPLAT initiation). These data are used to develop and validate the imputation procedure; actual trial data are not publicly available.

```{r}
data_raw <- read_parquet(here("data", "wide_data_730.parquet"))
```

## Data Preprocessing

### Variable Transformations

Survival time is censored at 182 days (6 months) for the primary analysis. The death event indicator is derived accordingly. Diagnosis categories are collapsed to the 5 most frequent groups to avoid sparse categories that could cause convergence issues.

```{r}
data <- data_raw |> 
  mutate(
    diagnosis = fct_lump_n(diagnosis, n = 4),
    diagnosis = factor(as.integer(diagnosis)),
    event_death   = if_else(survival_time <= 182, 1, 0),
    survival_time = pmin(survival_time, 182),
    tx            = factor(if_else(randomselect_liqplat == "1", 1, 0)),
    ecog_fstcnt = factor(ecog_fstcnt, ordered = TRUE)
  ) |> 
  select(
    pat_id,
    gender,
    age,
    ecog_fstcnt,
    tx,
    albumin,
    bilirubin,
    c_reactive_protein,
    lactate_dehydrogenase,
    event_death,
    survival_time,
    diagnosis
  ) |> 
  relocate(pat_id, event_death, survival_time)
```

### Nelson-Aalen Cumulative Hazard Estimate

We include the Nelson-Aalen cumulative hazard estimate (`na_est`) as a predictor in the imputation model instead of survival time [van-Buuren2021-pc, chapter 9.1].

```{r}
data$na_est <- nelsonaalen(data, timevar = survival_time, statusvar = event_death)
```

## Imputation Variables {#sec-variables}

The imputation model includes variables from the substantive analysis model and auxiliary variables that are predictive of missing values or their missingness mechanism.

### Analysis Variables

| Variable | Description | Type |
|----------|-------------|------|
| `tx` | Treatment group allocation | Binary |
| `ecog_fstcnt` | ECOG performance status at baseline (0–4) | Ordered factor |
| `diagnosis` | Primary cancer diagnosis (5 categories) | Nominal factor |
| `albumin` | Serum albumin at baseline (g/L) | Continuous |
| `c_reactive_protein` | C-reactive protein at baseline (mg/L) | Continuous |
| `event_death` | Death indicator (within 182 days) | Binary |

: Analysis variables included in the imputation model {#tbl-analysis-vars}

### Auxiliary Variables

| Variable | Description | Type |
|----------|-------------|------|
| `age` | Age at date of randomisation (years) | Continuous |
| `gender` | Patient sex | Binary |
| `bilirubin` | Serum bilirubin at baseline (µmol/L) | Continuous |
| `lactate_dehydrogenase` | Lactate dehydrogenase at baseline (U/L) | Continuous |
| `na_est` | Nelson-Aalen cumulative hazard estimate | Continuous |

: Auxiliary variables included in the imputation model {#tbl-auxiliary-vars}

## Multiple Imputation Procedure {#sec-imputation}

### Imputation Methods

| Variable | Method | Description |
|----------|--------|-------------|
| `ecog_fstcnt` | `polr` | Proportional odds logistic regression for ordered factors |
| `diagnosis` | `polyreg` | Polytomous logistic regression for unordered factors |
| `albumin` | `pmm` | Predictive mean matching for continuous variables |
| `bilirubin` | `pmm` | Predictive mean matching for continuous variables |
| `c_reactive_protein` | `pmm` | Predictive mean matching for continuous variables |
| `lactate_dehydrogenase` | `pmm` | Predictive mean matching for continuous variables |

: Imputation methods by variable {#tbl-methods}

Predictive mean matching (PMM) is used for continuous variables to preserve the marginal distributions and avoid imputing implausible values. We use Type 1 PMM with a donor pool of $k = 5$ [van-Buuren2021-pc, chapter 3.4]. For ordered categorical variables (ECOG), proportional odds logistic regression (`polr`) maintains the ordinal structure. For unordered categorical variables (diagnosis), polytomous logistic regression (`polyreg`) is used.

### Predictor Matrix Configuration

The predictor matrix is configured to:

- Exclude `pat_id` from predictions (identifier only)
- Exclude `survival_time` from predictions (we use `na_est` instead to incorporate survival information)
- Allow all other variables to predict each other

```{r}
#| output: false
md.pattern(data)

init <- mice(data, maxit = 0)
meth <- init$method
pred_matrix <- init$predictorMatrix

# Patient ID should not be used for imputation
pred_matrix[, "pat_id"] <- 0
pred_matrix["pat_id", ] <- 0
meth["pat_id"] <- ""

# Use Nelson-Aalen estimate instead of raw survival time
pred_matrix[, "survival_time"] <- 0
pred_matrix["survival_time", ] <- 0
meth["survival_time"] <- ""

imputed_data <- mice(data,
                     m = 10,      # 50 in final analysis
                     maxit = 10,  # 50 in final analysis
                     predictorMatrix = pred_matrix,
                     method = meth,
                     donors = 5,
                     seed = 1234)
```

### Imputation Parameters

For the final trial analysis:

- Number of imputed datasets: $m = 50$
- Maximum iterations: 50
- Donor pool size: $k = 5$

## Convergence Diagnostics {#sec-diagnostics}

Convergence is assessed via trace plots of the mean and standard deviation of imputed values across iterations.

```{r trace-plots}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Trace plots showing the mean and standard deviation of imputed values across iterations for each imputed dataset."
plot(imputed_data)
```

## Distribution Diagnostics {#sec-distribution}

We compare the distributions of observed and imputed values to assess plausibility.

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Density plots comparing observed (blue) and imputed (red) values for all imputed variables."
densityplot(imputed_data)
```

```{r}
#| fig-cap: "Density plot for ECOG performance status."
densityplot(imputed_data, ~ecog_fstcnt)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Strip plots of imputed values for continuous laboratory variables across imputed datasets."
stripplot(imputed_data, albumin + bilirubin + c_reactive_protein + lactate_dehydrogenase ~ .imp)
```

```{r}
#| fig-cap: "Strip plot of imputed values for ECOG performance status across imputed datasets."
stripplot(imputed_data, ecog_fstcnt ~ .imp)
```

## Analysis of Trial Data

For the final trial analysis, all $m = 50$ imputed datasets will be analysed, and posterior draws will be stacked to propagate imputation uncertainty into the final inference.

## Simulation Study Dataset

For simulation studies, a single completed dataset is extracted. This is computationally necessary but does not reflect the full imputation uncertainty.

```{r}
complete_data <- complete(imputed_data, 4)
```

## Export

```{r}
write_parquet(complete_data, sink = here("data", "survival_complete.parquet"))
```