---
Title : Simulate Datasets
---

## Setup

```{r}
library(furrr)
library(tidyverse)
library(data.table)
library(arrow)
library(rms)
```

## Load data

```{r}
df_ypprev <- read_parquet("./data/hosp_long_daily.parquet") |> 
    mutate(
        tx = factor(tx),
        ecog_fstcnt = factor(ecog_fstcnt, ordered = FALSE),
        yprev  = factor(yprev,  levels = 1:4, ordered = FALSE),
        ypprev = factor(ypprev, levels = 1:4, ordered = FALSE),
        y = ordered(y)
    ) |> 
      select(pat_id, tx, y, yprev, ypprev, day, ecog_fstcnt, diagnosis, age, gender, albumin, c_reactive_protein)

d <- df_ypprev
```

## Model fitting

We fit a model without a treatment parameter. We will randomly assign a treatment effect during the simulation of the dataset.
We will use the same for model for HA, but simulate a treatment effect during the simulation of the dataset.

```{r}
fit_h0 <- orm(
    y ~ rcs(day, 5) + yprev + ypprev + day %ia% yprev + day %ia% ypprev + ecog_fstcnt + gender + diagnosis + rcs(albumin, 3) + rcs(c_reactive_protein, 3) + rcs(age, 3),
    data = d 
  )

fit_h0
```

## Simulations function

### Simulate baseline data

```{r}
N <- 30000
baseline_data <- d |> 
  filter(day == 1) |> 
  select(yprev, ypprev, ecog_fstcnt, gender, diagnosis, age, albumin, c_reactive_protein) |> 
  slice_sample(n = N, replace = TRUE) |> 
  mutate(
    pat_id = row_number(),
    tx = sample(c(0, 1), size = N, replace = TRUE),
    x = rnorm(N) # add patient specific offset
    ) |> 
  data.frame()
```

### Predict next states

```{r}
# Vectorized version: computes category probabilities for all rows in `newdata` at once
# and returns an integer vector of sampled categories (1..k+1).
predict_next_state_vec <- function(model, newdata, k,
                                   tx_effect = log(1), x_effect = log(1)) {
  N <- nrow(newdata)

  # Get cumulative-logit linear predictors for each cutpoint, for ALL rows at once.
  # Had simple loop before, but vapply is much faster
  # Each call returns a length-N numeric vector.
  eta_mat <- vapply(
    X = seq_len(k),
    FUN = function(i) {
      as.numeric(predict(model, newdata = newdata, type = "lp", kint = i))
    },
    FUN.VALUE = numeric(N)
  )
  # Now eta_mat is N x K (rows: patients, cols: cutpoints 1..K)

  # Add external shifts (same shift to all cutpoints per row)
  shift <- newdata$tx * tx_effect + newdata$x * x_effect
  eta_mat <- sweep(eta_mat, 1, shift, FUN = "+")

  # Add +/- Inf boundaries, convert to probs via logistic, then difference adjacent CDFs
  # to get category probabilities (N x (k+1))
  eta_ext <- cbind(Inf, eta_mat, -Inf)                        # N x (K+2)
  p_ext   <- plogis(eta_ext)
  sop     <- p_ext[, 1:(k+1), drop = FALSE] - p_ext[, 2:(k+2), drop = FALSE]  # N x (K+1)

  # Row-wise categorical sampling without loops: inverse-CDF method
  u <- runif(N) #between 0 and 1 by default
  # cumulative probabilities by row
  cs <- t(apply(sop, 1, cumsum))                              # N x (K+1)
  # first index where cumprob >= u
  picks <- 1L + rowSums(cs < u)                               # integer in 1..(k+1)
  return(picks)
}
```

```{r}
generate_trajectories <- function(model, baseline_data, times, k,
                                       absorbing_state = NULL,
                                       tx_effect = log(1), x_effect = log(1),
                                       seed = 1234, verbose = FALSE) {
  set.seed(seed)

  # Keep factors as factors; I only update their character levels once where needed.
  N <- nrow(baseline_data)

  # Validate factor coding just once
  yprev_levels <- levels(baseline_data$yprev)
  ypprev_levels <- levels(baseline_data$ypprev)
  if (is.null(yprev_levels) || is.null(ypprev_levels)) {
    stop("Error: 'yprev' and 'ypprev' must be factor variables.")
  }

  current <- baseline_data

  # Preallocate results list
  out <- vector("list", length(times))

  msg <- function(...) if (isTRUE(verbose)) message(sprintf(...))

  for (i in seq_along(times)) {
    t <- times[i]
    if (verbose) msg("day %s (%d/%d)", t, i, length(times))

    # Set the day column once for all rows this step
    current$day <- t

    # Absorption mask: absorbed if previous state equals absorbing_state
    if (!is.null(absorbing_state)) {
      absorbed <- current$yprev == as.character(absorbing_state)
    } else {
      absorbed <- rep(FALSE, N)
    }

    # Sample next states for the non-absorbed patients in one vectorized call
    y_new <- integer(N)

    if (any(!absorbed)) {
      picks <- predict_next_state_vec(
        model, current[!absorbed, , drop = FALSE], k = k,
        tx_effect = tx_effect, x_effect = x_effect
      )
      y_new[!absorbed] <- picks
    }
    if (any(absorbed)) {
      y_new[absorbed] <- absorbing_state
    }

    # Store results for this day (keep as integers; convert to factor later if needed)
    out[[i]] <- data.frame(
      pat_id = current$pat_id,
      day   = t,
      y      = y_new,
      stringsAsFactors = FALSE
    )

    # Roll the state forward:
    # ypprev <- old yprev ; yprev <- new y
    # (keep them as character to avoid refactor overhead inside the loop)
    current$ypprev <- as.character(current$yprev)
    current$yprev  <- as.character(y_new)
  }

  
  res <- do.call(rbind, out)
  res$y <- factor(res$y, levels = seq_len(k + 1), ordered = TRUE)

  res
}
```

## h0: No treatment effect

```{r}
patient_trajectories <- generate_trajectories(
  model = fit_h0,
  baseline_data = baseline_data,
  times = 1:182,
  k = 4,
  absorbing_state = 5,
  tx_effect = log(1),
  x_effect = log(1),
  seed = 1234,
  verbose = FALSE
)

h0_data_forward <- baseline_data |> 
  rename(
    yprev_base = yprev, 
    ypprev_base = ypprev,
    ) |> 
  left_join(patient_trajectories, by = "pat_id") |> 
  arrange(pat_id, day) |> 
  group_by(pat_id) |> 
  mutate(
    yprev  = lag(y, 1),
    ypprev = lag(y, 2),
    # For the first one or two days, seed lags from baseline
    yprev  = ifelse(is.na(yprev),  as.integer(as.character(yprev_base)),  as.integer(as.character(yprev))),
    ypprev = ifelse(is.na(ypprev), as.integer(as.character(ypprev_base)), as.integer(as.character(ypprev)))
  ) |> 
  mutate(dday = ifelse(any(y == 5), min(day[y == 5]), max(day))) |> 
  ungroup()

h0_data <- h0_data_forward |>
  filter(day <= dday) |> 
  ungroup() |> 
  mutate(
    y          = as.numeric(y),
    yprev  = factor(yprev,  levels = 1:4, ordered = FALSE),
    ypprev = factor(ypprev, levels = 1:4, ordered = FALSE)
  ) |> 
    select(-yprev_base, -ypprev_base, -dday, -x)


fit_test <- orm(
    y ~ rcs(day, 5) + yprev + ypprev + day %ia% yprev + day %ia% ypprev + ecog_fstcnt + gender + diagnosis + rcs(albumin, 3) + rcs(c_reactive_protein, 3) + rcs(age, 3),
    data = h0_data
  )

# we have very close agreement between model used to generate data and same model fit to generated data
# if we set the treatment effect and x effect to log(1) 
cbind(round(coef(fit_h0), 3), round(coef(fit_test), 3))
```

### export

```{r}
write_parquet(h0_data, "./3-survival/sim-data/h0.parquet")
write_parquet(h0_data_forward, "./3-survival/sim-data/h0_forward.parquet")
```

### Visualize simulated data

```{r}
baseline_data |> 
  left_join(patient_trajectories, by = "pat_id") |> 
  group_by(tx, day, y) |> 
  summarise(n = n(), .groups = "drop") |> 
  group_by(tx, day) |>
  mutate(prop = n / sum(n)) |> 
  ggplot(aes(x = day, y = prop)) +
  # stacked bar char
  geom_bar(aes(fill = factor(y)), stat = "identity", position = "fill") +
  facet_wrap(~ tx) +
  labs(
    title = "Simulated patient trajectories",
    x = "day",
    y = "Proportion",
    fill = "State",
    color = "State"
  )

baseline_data |> 
  left_join(patient_trajectories, by = "pat_id") |> 
  group_by(tx, day, y) |> 
  summarise(n = n(), .groups = "drop") |> 
  group_by(tx, day) |>
  mutate(prop = n / sum(n)) |> 
  mutate(tx = factor(tx)) |>
  ggplot(aes(x = day, y = prop, color = factor(y), linetype = tx)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Simulated patient trajectories",
    x = "day",
    y = "Proportion",
    fill = "State",
    color = "State"
  )
```

### Calculating Surv, Cum Haz, Haz

```{r}
haz <- baseline_data |> 
  left_join(patient_trajectories, by = "pat_id") |> 
  group_by(day, y) |> 
  summarise(n = n(), .groups = "drop") |> 
  group_by(day) |>
  mutate(prop = n / sum(n)) |> 
  filter(y == 5) |> 
  ungroup() |> 
  arrange(day) |>
  mutate(
    surv = 1 - prop,
    cum_haz = -log(surv),
    haz = cum_haz - lag(cum_haz, n = 1, order_by = day, default = 0))

# plot the hazard
haz |> 
  ggplot(aes(x = day, y = haz)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Hazard of death",
    x = "day",
    y = "Hazard",
    color = "Treatment"
  )

# plot the cumulative hazard
haz |> 
  ggplot(aes(x = day, y = cum_haz)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Cumulative hazard of death",
    x = "day",
    y = "Cumulative hazard",
    color = "Treatment"
  )
```

## HA: Treatment effect

## Or of 0.9 ~ Difference in RMST of 3 days

```{r}
patient_trajectories_09 <- generate_trajectories(
  model = fit_h0,
  baseline_data = baseline_data,
  times = 1:182,
  k = 4,
  absorbing_state = 5,
  tx_effect = log(0.9),
  x_effect = log(1),
  seed = 1234,
  verbose = FALSE
)

ha_data_forward_09 <- baseline_data |> 
  rename(
    yprev_base = yprev, 
    ypprev_base = ypprev,
    ) |> 
  left_join(patient_trajectories_09, by = "pat_id") |> 
  arrange(pat_id, day) |> 
  group_by(pat_id) |> 
  mutate(
    yprev  = lag(y, 1),
    ypprev = lag(y, 2),
    # For the first one or two days, seed lags from baseline
    yprev  = ifelse(is.na(yprev),  as.integer(as.character(yprev_base)),  as.integer(as.character(yprev))),
    ypprev = ifelse(is.na(ypprev), as.integer(as.character(ypprev_base)), as.integer(as.character(ypprev)))
  ) |> 
  mutate(dday = ifelse(any(y == 5), min(day[y == 5]), max(day)))

ha_data_09 <- ha_data_forward_09 |>
  filter(day <= dday) |> 
  ungroup() |> 
  mutate(
    y          = as.numeric(y),
    yprev  = factor(yprev,  levels = 1:4, ordered = FALSE),
    ypprev = factor(ypprev, levels = 1:4, ordered = FALSE)
  ) |> 
    select(-yprev_base, -ypprev_base, -dday, -x)

baseline_data |> 
  left_join(patient_trajectories_09, by = "pat_id") |> 
  group_by(tx, day, y) |> 
  summarise(n = n(), .groups = "drop") |> 
  group_by(tx, day) |>
  mutate(prop = n / sum(n)) |> 
  ggplot(aes(x = day, y = prop)) +
  # stacked bar char
  geom_bar(aes(fill = factor(y)), stat = "identity", position = "fill") +
  facet_wrap(~ tx) +
  labs(
    title = "Simulated patient trajectories",
    x = "day",
    y = "Proportion",
    fill = "State",
    color = "State"
  )
```

```{r}
sops <- baseline_data |> 
  left_join(patient_trajectories_09, by = "pat_id") |> 
  group_by(tx, day, y) |> 
  summarise(n = n(), .groups = "drop") |> 
  group_by(tx, day) |>
  mutate(prop = n / sum(n)) |> 
  mutate(tx = factor(tx))

# difference in time out of hospital
sops |> 
  group_by(tx) |> 
  filter(y == 5) |>
  summarise(rmst = 1 - sum(prop)) |> 
  pull(rmst) |> 
  diff()
```

## OR of 0.8 ~ 6 days RMST difference

```{r}
patient_trajectories_08 <- generate_trajectories(
  model = fit_h0,
  baseline_data = baseline_data,
  times = 1:182,
  k = 4,
  absorbing_state = 5,
  tx_effect = log(0.8),
  x_effect = log(1),
  seed = 1234,
  verbose = FALSE
)

ha_data_forward_08 <- baseline_data |> 
  rename(
    yprev_base = yprev, 
    ypprev_base = ypprev,
    ) |> 
  left_join(patient_trajectories_08, by = "pat_id") |> 
  arrange(pat_id, day) |> 
  group_by(pat_id) |> 
  mutate(
    yprev  = lag(y, 1),
    ypprev = lag(y, 2),
    # For the first one or two days, seed lags from baseline
    yprev  = ifelse(is.na(yprev),  as.integer(as.character(yprev_base)),  as.integer(as.character(yprev))),
    ypprev = ifelse(is.na(ypprev), as.integer(as.character(ypprev_base)), as.integer(as.character(ypprev)))
  ) |> 
  mutate(dday = ifelse(any(y == 5), min(day[y == 5]), max(day)))

ha_data_08 <- ha_data_forward_08 |>
  filter(day <= dday) |> 
  ungroup() |> 
  mutate(
    y          = as.numeric(y),
    yprev  = factor(yprev,  levels = 1:4, ordered = FALSE),
    ypprev = factor(ypprev, levels = 1:4, ordered = FALSE)
  ) |> 
    select(-yprev_base, -ypprev_base, -dday, -x)
```

```{r}
sops <- baseline_data |> 
  left_join(patient_trajectories_08, by = "pat_id") |> 
  group_by(tx, day, y) |> 
  summarise(n = n(), .groups = "drop") |> 
  group_by(tx, day) |>
  mutate(prop = n / sum(n)) |> 
  mutate(tx = factor(tx))

# difference in time out of hospital
sops |> 
  group_by(tx) |> 
  filter(y == 5) |>
  summarise(rmst = 1 - sum(prop)) |> 
  pull(rmst) |> 
  diff()
```

## OR of 0.7 ~ 9 days of RMST difference 

```{r}
patient_trajectories_07 <- generate_trajectories(
  model = fit_h0,
  baseline_data = baseline_data,
  times = 1:182,
  k = 4,
  absorbing_state = 5,
  tx_effect = log(0.7),
  x_effect = log(1),
  seed = 1234,
  verbose = FALSE
)

ha_data_forward_07 <- baseline_data |> 
  rename(
    yprev_base = yprev, 
    ypprev_base = ypprev,
    ) |> 
  left_join(patient_trajectories_07, by = "pat_id") |> 
  arrange(pat_id, day) |> 
  group_by(pat_id) |> 
  mutate(
    yprev  = lag(y, 1),
    ypprev = lag(y, 2),
    # For the first one or two days, seed lags from baseline
    yprev  = ifelse(is.na(yprev),  as.integer(as.character(yprev_base)),  as.integer(as.character(yprev))),
    ypprev = ifelse(is.na(ypprev), as.integer(as.character(ypprev_base)), as.integer(as.character(ypprev)))
  ) |> 
  mutate(dday = ifelse(any(y == 5), min(day[y == 5]), max(day)))

ha_data_07 <- ha_data_forward_07 |>
  filter(day <= dday) |> 
  ungroup() |> 
  mutate(
    y          = as.numeric(y),
    yprev  = factor(yprev,  levels = 1:4, ordered = FALSE),
    ypprev = factor(ypprev, levels = 1:4, ordered = FALSE)
  ) |> 
    select(-yprev_base, -ypprev_base, -dday, -x)
```

```{r}
sops <- baseline_data |> 
  left_join(patient_trajectories_07, by = "pat_id") |> 
  group_by(tx, day, y) |> 
  summarise(n = n(), .groups = "drop") |> 
  group_by(tx, day) |>
  mutate(prop = n / sum(n)) |> 
  mutate(tx = factor(tx))

# difference in time out of hospital
sops |> 
  group_by(tx) |> 
  filter(y == 5) |>
  summarise(rmst = 1- sum(prop)) |> 
  pull(rmst) |> 
  diff()
```

### Export

```{r}
write_parquet(ha_data_forward_07, "./3-survival/sim-data/ha_07_forward.parquet")
write_parquet(ha_data_07, "./3-survival/sim-data/ha_07.parquet")
write_parquet(ha_data_forward_08, "./3-survival/sim-data/ha_08_forward.parquet")
write_parquet(ha_data_08, "./3-survival/sim-data/ha_08.parquet")
write_parquet(ha_data_forward_09, "./3-survival/sim-data/ha_09_forward.parquet")
write_parquet(ha_data_09, "./3-survival/sim-data/ha_09.parquet")
```
