---
title: "Bayesian Power Simulation Pipeline - Survival Analysis"
---

This document implements a Bayesian power simulation for survival outcomes using 
`rstanarm::stan_surv()` with M-spline baseline hazard.

**Analyses:**

1. **RMST (full)**: Marginalized RMST difference via posterior survival curves (slower)
2. **Treatment coefficient**: Posterior draws of the treatment log-HR only (faster)

**Scenarios:**

- **H0**: OR = 1.0 (null hypothesis, treatment rerandomized, RMST diff ≈ 0 days)
- **HA_09**: OR = 0.9 (RMST diff ≈ 3 days)
- **HA_08**: OR = 0.8 (RMST diff ≈ 6 days)
- **HA_07**: OR = 0.7 (RMST diff ≈ 9 days)
- **HA_06**: OR = 0.6 (RMST diff ≈ 12 days)

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(here)
library(arrow)
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(furrr)
library(fs)
library(glue)
library(data.table)
library(rstanarm)
library(markov.misc)

# Configuration
N_SIMULATIONS <- 1000
SAMPLE_SIZE <- 270              # Total patients: 180 tx + 90 soc
ALLOCATION_RATIO <- 2/3         # Proportion allocated to treatment (2:1 ratio)
N_CORES <- 7
SEED <- 456
DECISION_THRESHOLD <- 0.95      # P(benefit) threshold for declaring success

# Paths
DATA_PATH <- here("3-survival", "sim-data")
OUTPUT_PATH <- here("3-survival", "output", "bayesian-power-sim")

# Create output directory if needed
if (!dir.exists(OUTPUT_PATH)) dir.create(OUTPUT_PATH, recursive = TRUE)


```

## Load All Datasets and Calculate True RMST Differences

```{r load-datasets}
# Load all simulation datasets
h0_data <- read_parquet(file.path(DATA_PATH, "h0_forward.parquet"))
ha_09_data <- read_parquet(file.path(DATA_PATH, "ha_09_forward.parquet"))
ha_08_data <- read_parquet(file.path(DATA_PATH, "ha_08_forward.parquet"))
ha_07_data <- read_parquet(file.path(DATA_PATH, "ha_07_forward.parquet"))
ha_06_data <- read_parquet(file.path(DATA_PATH, "ha_06_forward.parquet"))

# Calculate true RMST differences for each dataset
# Target states 1:4 (alive states), excluding 5 (death)
true_rmst_h0 <- h0_data |>
  select(id, time = day, y, tx) |>
  calc_time_in_state_diff(target_state = 1:4) |> 
  pull(true_effect)

true_rmst_09 <- ha_09_data |>
  select(id, time = day, y, tx) |>
  calc_time_in_state_diff(target_state = 1:4) |> 
  pull(true_effect)

true_rmst_08 <- ha_08_data |>
  select(id, time = day, y, tx) |>
  calc_time_in_state_diff(target_state = 1:4) |> 
  pull(true_effect)

true_rmst_07 <- ha_07_data |>
  select(id, time = day, y, tx) |>
  calc_time_in_state_diff(target_state = 1:4) |> 
  pull(true_effect)

true_rmst_06 <- ha_06_data |>
  select(id, time = day, y, tx) |>
  calc_time_in_state_diff(target_state = 1:4) |> 
  pull(true_effect)

# Display true RMST differences
cat("True RMST differences (days):\n")
cat(sprintf("H0 (OR=1.0): %.2f\n", true_rmst_h0))
cat(sprintf("HA (OR=0.9): %.2f\n", true_rmst_09))
cat(sprintf("HA (OR=0.8): %.2f\n", true_rmst_08))
cat(sprintf("HA (OR=0.7): %.2f\n", true_rmst_07))
cat(sprintf("HA (OR=0.6): %.2f\n", true_rmst_06))

rm(h0_data, ha_09_data, ha_08_data, ha_07_data, ha_06_data)
```

## Step 1: Identify Scenarios

We define four scenarios based on the available superpopulation files.

```{r identify-scenarios}
# Define scenarios using calculated true RMST differences
scenarios <- tibble(
  scenario_type = c("H0", "HA", "HA", "HA", "HA"),
  effect_label = c("OR_1.0", "OR_0.9", "OR_0.8", "OR_0.7", "OR_0.6"),
  rmst_diff_true = c(0, true_rmst_09, true_rmst_08, true_rmst_07, true_rmst_06),  # True RMST difference from data
  data_file = c("h0.parquet", "ha_09.parquet", "ha_08.parquet", "ha_07.parquet", "ha_06.parquet"),
  rerandomize = c(TRUE, FALSE, FALSE, FALSE, FALSE)
) |>
  mutate(
    data_path = file.path(DATA_PATH, data_file),
    scenario_label = glue("{scenario_type}_{effect_label}"),
    output_path = file.path(OUTPUT_PATH, scenario_label)
  )

scenarios
```

## Step 2: Data Preparation Function

```{r data-prep-function}
#' Prepare survival data from raw simulation data
#' 
#' @param df Full dataset from superpopulation
#' @return Data ready for stan_surv() model
prepare_survival_data <- function(df) {
  df |>
    filter(y == 5 | day == 182) |>
    group_by(id) |>
    arrange(desc(y)) |>
    slice_head(n = 1) |>
    ungroup() |>
    mutate(death = ifelse(y == 5, 1, 0)) |>
    select(id, tx, day, death)
}
```

## Step 3: Define Bayesian Fitting Functions

Each function fits a Bayesian model and returns results in the format expected 
by `assess_operating_characteristics()`: a list where the first element is a 
summary tibble with required columns, and subsequent elements contain details.

For Bayesian models, we use:
- `estimate`: posterior median
- `std_error`: posterior SD
- `conf_low`, `conf_high`: 95% credible interval
- `p_benefit`: stored in details list for power/type I error calculations

```{r fitting-functions}
#' Fit survival model and extract treatment coefficient only (fast)
#' 
#' @param data Sampled data from superpopulation
#' @param iter Iteration number
#' @return List with summary tibble and posterior draws
fit_surv_coef <- function(data, iter) {
  # Prepare survival data
  surv_data <- prepare_survival_data(data)

  prior <- normal(
    location = 0,
    scale = 0.55  
  )
 
  model <- tryCatch({
    stan_surv(
      formula = Surv(day, death) ~ tx,
      data = surv_data,
      basehaz = "ms",
      basehaz_ops = list(df = 5),
      prior = prior,
      chains = 4,
      cores = 1,
      seed = 1234,
      iter = 2000,
      refresh = 0
    )
  }, error = function(e) {
    message(glue("ERROR in surv_coef iteration {iter}: {e$message}"))
    return(NULL)
  })
  
  if (is.null(model)) {
    return(list(
      tibble(
        iter = iter,
        analysis = "surv_coef",
        se_type = "bayesian",
        term = "tx",
        estimate = NA_real_,
        std_error = NA_real_,
        conf_low = NA_real_,
        conf_high = NA_real_
      )
    ))
  }
  
  # Extract tx posterior draws (log-HR scale)
  draws_df <- as.data.frame(model)
  tx_draws <- draws_df$tx
  
  # Calculate P(benefit): negative log-HR means treatment reduces hazard
  p_benefit <- mean(tx_draws < 0)
  
  # Extract diagnostics
  model_summary <- summary(model)
  rhat_values <- model_summary[, "Rhat"]
  n_eff_values <- model_summary[, "n_eff"]
  
  # Create summary in required format
  summary_tbl <- tibble(
    iter = iter,
    analysis = "surv_coef",
    se_type = "bayesian",
    term = "tx",
    estimate = median(tx_draws),
    std_error = sd(tx_draws),
    conf_low = quantile(tx_draws, 0.025),
    conf_high = quantile(tx_draws, 0.975)
  )
  
  # Return list: summary + details
  list(
    summary_tbl,
    list(
      posterior_draws = tibble(
        iter = iter,
        analysis = "surv_coef",
        draw_id = seq_along(tx_draws),
        tx_effect = tx_draws
      ),
      p_benefit = p_benefit,
      rhat = as.list(rhat_values),
      n_eff = as.list(n_eff_values)
    )
  )
}

#' Fit survival model and calculate marginalized RMST difference (slow)
#' 
#' @param data Sampled data from superpopulation
#' @param iter Iteration number
#' @return List with summary tibble and posterior draws including RMST
fit_surv_rmst <- function(data, iter) {
  # Prepare survival data
  surv_data <- prepare_survival_data(data)
  
  prior <- normal(
    location = 0,
    scale = 0.55  
  )
 
  model <- tryCatch({
    stan_surv(
      formula = Surv(day, death) ~ tx,
      data = surv_data,
      basehaz = "ms",
      basehaz_ops = list(df = 5),
      prior = prior,
      chains = 4,
      cores = 1,
      seed = 1234,
      iter = 2000,
      refresh = 0
    )
  }, error = function(e) {
    message(glue("ERROR in surv_rmst iteration {iter}: {e$message}"))
    return(NULL)
  })
  
  if (is.null(model)) {
    return(list(
      tibble(
        iter = iter,
        analysis = "surv_rmst",
        se_type = "bayesian",
        term = "rmst_diff",
        estimate = NA_real_,
        std_error = NA_real_,
        conf_low = NA_real_,
        conf_high = NA_real_
      )
    ))
  }
  
  # Extract tx posterior draws (log-HR scale)
  draws_df <- as.data.frame(model)
  tx_draws <- draws_df$tx
  
  # Extract diagnostics
  model_summary <- summary(model)
  rhat_values <- model_summary[, "Rhat"]
  n_eff_values <- model_summary[, "n_eff"]
  
  # Calculate marginalized RMST via posterior survival curves
  rmst_result <- tryCatch({
    # Get posterior survival curves for tx=0
    ps_0_list <- posterior_survfit(
      model,
      newdata = surv_data |> mutate(tx = 0),
      times = 0,
      extrapolate = TRUE,
      standardise = TRUE,
      control = list(edist = 182, epoints = 200),
      return_matrix = TRUE,
      draws = 4000
    )
    std_surv_matrix_0 <- do.call(cbind, ps_0_list)
    time_points <- sapply(ps_0_list, attr, "times")
    
    # Get posterior survival curves for tx=1
    ps_1_list <- posterior_survfit(
      model,
      newdata = surv_data |> mutate(tx = 1),
      times = 0,
      extrapolate = TRUE,
      standardise = TRUE,
      control = list(edist = 182, epoints = 200),
      return_matrix = TRUE,
      draws = 4000
    )
    std_surv_matrix_1 <- do.call(cbind, ps_1_list)
    
    # Calculate RMST for tx=0 using trapezoidal rule
    delta_t <- diff(time_points)
    s_t1_0 <- std_surv_matrix_0[, -ncol(std_surv_matrix_0)]
    s_t2_0 <- std_surv_matrix_0[, -1]
    trapezoid_areas_0 <- 0.5 * (s_t1_0 + s_t2_0) * matrix(delta_t, nrow = nrow(s_t1_0), ncol = length(delta_t), byrow = TRUE)
    rmst_0 <- rowSums(trapezoid_areas_0)
    
    # Calculate RMST for tx=1
    s_t1_1 <- std_surv_matrix_1[, -ncol(std_surv_matrix_1)]
    s_t2_1 <- std_surv_matrix_1[, -1]
    trapezoid_areas_1 <- 0.5 * (s_t1_1 + s_t2_1) * matrix(delta_t, nrow = nrow(s_t1_1), ncol = length(delta_t), byrow = TRUE)
    rmst_1 <- rowSums(trapezoid_areas_1)
    
    # Calculate the difference
    rmst_diff <- rmst_1 - rmst_0
    
    list(rmst_0 = rmst_0, rmst_1 = rmst_1, rmst_diff = rmst_diff)
  }, error = function(e) {
    message(glue("ERROR calculating RMST in iteration {iter}: {e$message}"))
    return(NULL)
  })
  
  if (is.null(rmst_result)) {
    return(list(
      tibble(
        iter = iter,
        analysis = "surv_rmst",
        se_type = "bayesian",
        term = "rmst_diff",
        estimate = NA_real_,
        std_error = NA_real_,
        conf_low = NA_real_,
        conf_high = NA_real_
      )
    ))
  }
  
  rmst_diff <- rmst_result$rmst_diff
  
  # Calculate P(benefit): positive RMST diff means treatment extends survival

  p_benefit <- mean(rmst_diff > 0)
  
  # Create summary in required format
  summary_tbl <- tibble(
    iter = iter,
    analysis = "surv_rmst",
    se_type = "bayesian",
    term = "rmst_diff",
    estimate = median(rmst_diff),
    std_error = sd(rmst_diff),
    conf_low = quantile(rmst_diff, 0.025),
    conf_high = quantile(rmst_diff, 0.975)
  )
  
  # Return list: summary + details
  list(
    summary_tbl,
    list(
      posterior_draws = tibble(
        iter = iter,
        analysis = "surv_rmst",
        draw_id = seq_along(rmst_diff),
        rmst_diff = rmst_diff,
        rmst_0 = rmst_result$rmst_0,
        rmst_1 = rmst_result$rmst_1
      ),
      tx_draws = tibble(
        iter = iter,
        draw_id = seq_along(tx_draws),
        tx_effect = tx_draws
      ),
      p_benefit = p_benefit,
      rhat = as.list(rhat_values),
      n_eff = as.list(n_eff_values)
    )
  )
}
```

## Step 4: Run Simulations for All Scenarios

```{r run-simulations}
# Define fit functions list
fit_functions <- list(
  surv_coef = fit_surv_coef
#   surv_rmst = fit_surv_rmst
)

# Run simulations for each scenario
for (i in seq_len(nrow(scenarios))) {
  scenario <- scenarios[i, ]


  message(glue("\n=== Running scenario: {scenario$scenario_label} ==="))
  message(glue("True RMST difference: {round(scenario$rmst_diff_true, 2)} days"))
  
  # Create output directory
  if (!dir.exists(scenario$output_path)) {
    dir.create(scenario$output_path, recursive = TRUE)
  }
  
  # Data paths for all analyses (same file for both)
  data_paths <- list(
    surv_coef = scenario$data_path
    # surv_rmst = scenario$data_path
  )
  
  # Check for completed iterations
  summary_dir <- file.path(scenario$output_path, "summary")
  if (dir.exists(summary_dir)) {
    completed_files <- list.files(summary_dir, pattern = "_iter_\\d+\\.rds",
                                   recursive = TRUE, full.names = TRUE)
    completed_iters <- as.integer(
      unique(gsub(".*_iter_(\\d+)\\.rds", "\\1", completed_files))
    )
  } else {
    completed_iters <- integer(0)
  }
  
  iters_to_run <- setdiff(1:N_SIMULATIONS, completed_iters)
  n_to_run <- length(iters_to_run)
  n_completed <- length(completed_iters)
  
  message(glue("{n_completed} of {N_SIMULATIONS} iterations already complete."))
  
  if (n_to_run > 0) {
    message(glue("Starting {n_to_run} new simulations on {N_CORES} cores..."))
    
    t1 <- Sys.time()
    future::plan(future.callr::callr, workers = N_CORES)
    
    furrr::future_map(
      .x = iters_to_run,
      .f = ~ markov.misc::assess_operating_characteristics(
        iter_num = .x,
        data_paths = data_paths,
        output_path = scenario$output_path,
        fit_functions = fit_functions,
        sample_size = SAMPLE_SIZE,
        allocation_ratio = ALLOCATION_RATIO,
        seed = SEED,
        rerandomize = scenario$rerandomize
      ),
      .options = furrr_options(
        seed = TRUE,
        globals = c(
          # --- Helper Functions ---
          "prepare_survival_data",
          
          # --- Configuration Constants ---
          "SEED",
          "SAMPLE_SIZE",
          "ALLOCATION_RATIO",
          "fit_functions",
          
          # --- Loop-Specific Variables ---
          "scenario",
          "data_paths"
        ),
        packages = c("rstanarm", "dplyr", "tibble", "glue", "markov.misc")
      ),
      .progress = TRUE
    )
    
    future::plan("sequential")
    t2 <- Sys.time()
    message(glue("Finished {n_to_run} simulations for {scenario$scenario_label}."))
    message(glue("Duration: {round(difftime(t2, t1, units = 'mins'), 1)} minutes"))
  } else {
    message("All simulations already complete.")
  }
}

message("\n=== All scenarios completed ===")
```