---
Title : Estimand
---

## Setup

```{r}
library(here)
library(tidyverse)
library(data.table)
library(arrow)
library(rstanarm)
library(glue)
```

## Load data

```{r}
set.seed(123)
data_path <- here("3-survival", "sim-data", "ha_09.parquet")

patient_info <- arrow::open_dataset(data_path) |>
        select(pat_id, tx) |>
        distinct() |>
        collect()

tx_ids <- sample(patient_info$pat_id[patient_info$tx == 1], size = 180, replace = FALSE)
soc_ids <- sample(patient_info$pat_id[patient_info$tx == 0], size = 90, replace = FALSE)
selected_pat_ids <- c(tx_ids, soc_ids)

# --- 3. Load ONLY the required subset of data ---
data <- arrow::open_dataset(data_path) |>
    filter(pat_id %in% selected_pat_ids) |>
    collect() |>
    rename(id = pat_id) |>
    filter(y == 5 | day == 182) |>
    group_by(id) |>
    arrange(desc(y)) |>
    slice_head(n = 1) |>
    ungroup() |>
    mutate(death = ifelse(y == 5, 1, 0)) |>
    select(id, tx, day, death, ecog_fstcnt, diagnosis)
```

## Estimand

We use the potential outcomes framework to define the estimand. Let $T_i^a$ be the potential time-to-death for individual $i$ under treatment assignment $a$, where $a=1$ is invitation to the single-arm trial (SAT) and $a=0$ is the standard of care (SoC).

The Restricted Mean Survival Time (RMST) under treatment $a$ up to a pre-specified time horizon of $t_{max}=182$ days is the area under the survival curve, $S^a(t) = P(T^a > t)$:

$$
\text{RMST}^a(t_{max}) = \int_0^{t_{max}} S^a(t) dt
$$

Our primary estimand is the Average Treatment Effect (ATE) on the RMST, denoted $\tau$.

$$
\tau = \text{RMST}^1(t_{max}) - \text{RMST}^0(t_{max})
$$

### Identification

The estimand is identified by calculating the marginal survival curve under each treatment, averaged over the covariate distribution of sample for increased precision. Because the survival data are from a death registry, we do not expect informative censoring.

## Mathematical Model Notation

The model is defined by the hazard rate $h_i(t)$ for each individual $i$ at time $t$.

$$
\begin{aligned}
    \text{day}_i, \text{death}_i &\sim \text{Survival}(h_i(t)) \\
    h_i(t) &= h_0(t) \cdot \exp(\eta_i) \\
    \eta_i &= \beta_0 + \beta_1 \cdot \text{tx}_i + \sum_{j=2}^{3} \beta_{ecog, j} \cdot \text{ecog}_{ij} + \sum_{k=2}^{5} \beta_{diag, k} \cdot \text{diag}_{ik} \\
    h_0(t) &= \sum_{k=1}^{5} \gamma_k \cdot M_k(t) \\
\\
    \beta_0 &\sim \text{Normal}(0, 20) \\
    \beta_1 &\sim \text{Normal}(0, 0.55) \\
    \beta_{\text{ecog},j \in 2,3} &\sim \text{Normal}(0, 2.5) \\
    \beta_{\text{diag},k \in 2..5} &\sim \text{Normal}(0, 2.5) \\
    \gamma &\sim \text{Dirichlet}(\mathbf{1}_5)
\end{aligned}
$$

This is a **proportional hazards survival model**. We assume the observed data for individual $i$, which consists of a time-to-event indicator ($\text{day}$), and an event indicator ($\text{death}$) follows a survival process which is defined by the hazard function $h_i(t)$.

The hazard rate $h_i(t)$ for an individual is the product of a flexible **baseline hazard**, $h_0(t)$, and the exponentiated effect of their covariates, $\exp(\eta_i)$.

-   **Baseline Hazard**: $h_0(t)$ is modeled using **M-splines** with 5 degrees of freedom. $M_k(t)$ are non-negative M-spline basis functions.
-   **Linear Predictor**: $\eta_i$ contains the intercept ($\beta_0$), the effect of treatment ($\beta_1 \cdot \text{tx}_i$), and the effects of the categorical predictions, which are represented by sums over their dummy variables.
-   **Priors**: The model specifies Normal priors for the regression coefficients ($\beta_1, \beta_{ecog, j}, \beta_{diag, k}$). A Dirichlet prior is used for the M-spline coefficients, $\gamma$. The Dirichlet prior is used because the spline coefficients are constrained by `rstanarm` to be positive and sum to 1 for identification.

## Fit the model

Prior: We deem it unlikely that invitation to the single arm trial (SAT) could increase or decrease hazard ratio (HR) by more than a factor of 2, i.e., a HR of $0.5$ or $2.0$. To place ca. 80% of the prior probability mass within this range we use a normal prior on the log-hazard ratio scale with mean $0$ and standard deviation $0.55$.

```{r}
# all priors except for the tx prior are default rstanarm priors
tx_scale <- 0.55
default_scale <- 2.5
my_prior <- normal(
  location = c(0, 0, 0, 0, 0, 0, 0),
  scale = c(tx_scale, default_scale, default_scale, default_scale, default_scale, default_scale, default_scale)
)

model <- stan_surv(
            formula = Surv(day, death) ~ tx + ecog_fstcnt + diagnosis,
            data = data,
            basehaz = "ms",
            basehaz_ops = list(df = 5),
            prior = my_prior,
            prior_intercept = normal(0, 20),
            prior_aux = dirichlet(1),
            chains = 4,
            cores = 1,
            seed = 1234,
            iter = 2000
        )
```

## Inspect Baseline Hazard

```{r}
plotfun <- function(model, title) {
  plot(model, plotfun = "basehaz") + # plot baseline hazard
    coord_cartesian(ylim = c(0, 0.007)) + # adjust y-axis limits
    labs(title = title) + # add plot title
    theme(plot.title = element_text(hjust = 0.5)) # centre plot title
}

plotfun(model, "Baseline Hazard Function")
```

## Calculate RMST

```{r}
# --- 6b. Calculate and Save RMST ---
# Get posterior survival curves for tx=0
ps_0_list <- posterior_survfit(
    model,
    newdata = data_for_model |> mutate(tx = 0),
    times = 0,
    extrapolate = TRUE,
    standardise = TRUE,
    control = list(edist = 182, epoints = 200),
    return_matrix = TRUE,
    draws = 4000
)

std_surv_matrix_0 <- do.call(cbind, ps_0_list)
time_points <- sapply(ps_0_list, attr, "times")

# Get posterior survival curves for tx=1
ps_1_list <- posterior_survfit(
    model,
    newdata = data_for_model |> mutate(tx = 1),
    times = 0,
    extrapolate = TRUE,
    standardise = TRUE,
    control = list(edist = 182, epoints = 200),
    return_matrix = TRUE,
    draws = 4000
)
std_surv_matrix_1 <- do.call(cbind, ps_1_list)

# Calculate RMST for tx=0 using trapezoidal rule
delta_t <- diff(time_points)
s_t1_0 <- std_surv_matrix_0[, -ncol(std_surv_matrix_0)]
s_t2_0 <- std_surv_matrix_0[, -1]
trapezoid_areas_0 <- 0.5 * (s_t1_0 + s_t2_0) * matrix(delta_t, nrow = nrow(s_t1_0), ncol = length(delta_t), byrow = TRUE)
rmst_0 <- rowSums(trapezoid_areas_0)

# Calculate RMST for tx=1
s_t1_1 <- std_surv_matrix_1[, -ncol(std_surv_matrix_1)]
s_t2_1 <- std_surv_matrix_1[, -1]
trapezoid_areas_1 <- 0.5 * (s_t1_1 + s_t2_1) * matrix(delta_t, nrow = nrow(s_t1_1), ncol = length(delta_t), byrow = TRUE)
rmst_1 <- rowSums(trapezoid_areas_1)

# Calculate the difference
rmst_diff <- rmst_1 - rmst_0

# Combine and save results
rmst_results <- data.frame(rmst_0, rmst_1, rmst_diff)

ggplot(rmst_results, (aes(x = rmst_diff))) +
    ggdist::stat_dots(side = "bottom") +
    ggdist::stat_halfeye() +
    labs(
        title = "Posterior Distribution of RMST Difference (SAT - SoC)",
        x = "RMST Difference (Days)",
        y = "Density"
    )
```

## Survival Curves

```{r}
long_data_0 <- std_surv_matrix_0 |>
  as_tibble(.name_repair = ~ as.character(time_points)) |>
  mutate(.draw = 1:n()) |>
  pivot_longer(
    cols = !.draw,
    names_to = "time",
    values_to = "survival",
    names_transform = as.numeric
  ) |>
  mutate(treatment = "SoC")

long_data_1 <- std_surv_matrix_1 |>
  as_tibble(.name_repair = ~ as.character(time_points)) |>
  mutate(.draw = 1:n()) |>
  pivot_longer(
    cols = !.draw,
    names_to = "time",
    values_to = "survival",
    names_transform = as.numeric
  ) |>
  mutate(treatment = "SAT")

# plot
bind_rows(long_data_0, long_data_1) |> 
  ggplot(aes(x = time, y = survival, fill = treatment, color = treatment)) +
  ggdist::stat_lineribbon(.width = 0.95, alpha = 0.5, linewidth = 0.8) +
  coord_cartesian(ylim = c(0, 1), xlim = c(0, 182), expand = FALSE) +
  labs(
    title = "Standardised Survival Curves by Treatment Group",
    x = "Time in Days",
    y = "Survival Probability",
    color = "Treatment Group",
    fill = "Treatment Group"
  ) +
  theme(
      legend.position = "bottom"
    #   plot.title = element_text(face = "bold"),
    #   panel.grid.minor = element_blank()
      )
```