---
title: "Imaging — model fitting"
format: html
execute:
  echo: true
  warning: false
  message: false
---

Following the SAP (*Comparative Clinical Outcomes — Imaging Count*), the total
number of imaging examinations is modeled with a Bayesian negative binomial
regression with a `log(exposure)` offset, adjusting for randomized group,
linear age, and ECOG (0 as reference, 1, 2, 3+). 

We also fit the model without covariates (randomized
group and offset only).

The regression coefficients get weakly informative $\mathcal{N}(0, 2.5)$
priors; the intercept and the negative binomial shape keep the `brms`
defaults.

```{r}
#| label: setup

library(tidyverse)

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")

result_dir <- here::here("results", "secondary", "imaging")

imaging_data <- arrow::read_parquet(file.path(result_dir, "analysis-data.parquet"))
completed_datasets <- readRDS(file.path(result_dir, "completed-datasets.rds"))
```

## Model formulas

```{r}
#| label: model-formulas

imaging_formulas <- list(
  adjusted = brms::bf(
    n_imaging_total ~ tx + age + ecog_group + offset(log(exposure_days))
  ),
  unadjusted = brms::bf(
    n_imaging_total ~ tx + offset(log(exposure_days))
  )
)

coefficient_prior <- brms::prior(normal(0, 2.5), class = "b")
```

## Fitting

The adjusted model is fitted to every completed dataset with
`brm_multiple()`, keeping one fit per imputation so that predictions can use
each imputation's own covariates. The unadjusted model needs no imputation and
is fitted once to the observed data. Both are stored as a list of fits.

```{r}
#| label: fit-models

fit_imaging_model <- function(adjustment, seed_offset) {
  model_path <- file.path(result_dir, paste0("model-", adjustment, ".rds"))

  if (force || !file.exists(model_path)) {
    seed <- analysis_seed(config, "supporting", offset = seed_offset)

    fits <- if (adjustment == "adjusted") {
      brms::brm_multiple(
        imaging_formulas$adjusted,
        data = completed_datasets,
        family = brms::negbinomial(),
        prior = coefficient_prior,
        chains = 4L,
        iter = 2000L,
        cores = 1L,
        init = 0,
        seed = seed,
        combine = FALSE,
        refresh = 0L
      )
    } else {
      list(brms::brm(
        imaging_formulas$unadjusted,
        data = imaging_data,
        family = brms::negbinomial(),
        prior = coefficient_prior,
        chains = 4L,
        iter = 2000L,
        cores = 1L,
        init = 0,
        seed = seed,
        refresh = 0L
      ))
    }

    saveRDS(fits, model_path, compress = "xz")
  }

  readRDS(model_path)
}

imaging_fits <- tibble(
  adjustment = c("adjusted", "unadjusted"),
  seed_offset = c(2601L, 2602L)
) |>
  mutate(fits = map2(adjustment, seed_offset, fit_imaging_model))
```

## Diagnostics

```{r}
#| label: tbl-model-diagnostics
#| tbl-cap: "Largest R-hat and smallest bulk effective sample size per model and imputation"

imaging_fits |>
  mutate(diagnostics = map(fits, \(fits) {
    tibble(
      imputation = seq_along(fits),
      max_rhat = map_dbl(fits, \(fit) max(brms::rhat(fit), na.rm = TRUE)),
      min_ess_bulk = map_dbl(fits, \(fit) {
        min(posterior::summarise_draws(fit, "ess_bulk")$ess_bulk, na.rm = TRUE)
      })
    )
  })) |>
  select(adjustment, diagnostics) |>
  unnest(diagnostics) |>
  knitr::kable(digits = 3)
```

```{r}
#| label: tbl-adjusted-coefficients
#| tbl-cap: "Posterior coefficients of the adjusted imaging model (first imputation)"

imaging_fits |>
  filter(adjustment == "adjusted") |>
  pull(fits) |>
  pluck(1, 1) |>
  brms::fixef() |>
  knitr::kable(digits = 3)
```
