---
title: "Multiple Imputation Quality of Life"
---

## Setup

```{r}
#| output: false
library(here)
library(tidyverse)
library(patchwork)
library(mice)
library(miceadds)
library(arrow)
```

## Data

```{r}
data_raw <- read_parquet(here("data", "analysis-data", "liqplat_analysis_data_long.parquet"))

data_wide <- read_parquet(here("data", "analysis-data", "liqplat_analysis_data_wide.parquet")) |>
  select(id, age, albumin, c_reactive_protein) |>
  mutate(
    albumin_std = as.numeric(scale(albumin)),
    crp_std = as.numeric(scale(c_reactive_protein)),
    age_std = as.numeric(scale(age))
    )
```

Invert QoL Data for modelling

```{r}
data_raw <- data_raw |>
  mutate(
    baseline_q30 = case_when(
      baseline_q30 == 1 ~ 7,
      baseline_q30 == 2 ~ 6,
      baseline_q30 == 3 ~ 5,
      baseline_q30 == 4 ~ 4,
      baseline_q30 == 5 ~ 3,
      baseline_q30 == 6 ~ 2,
      baseline_q30 == 7 ~ 1,
      TRUE ~ baseline_q30
    ),
    y_qol = case_when(
      y_qol == 1 ~ 7,
      y_qol == 2 ~ 6,
      y_qol == 3 ~ 5,
      y_qol == 4 ~ 4,
      y_qol == 5 ~ 3,
      y_qol == 6 ~ 2,
      y_qol == 7 ~ 1,
      TRUE ~ y_qol
    )
  )
```

## Data Preprocessing

### Inclusion Criteria

The substantive analysis model includes only patients who survived beyond 182 days.

### Variable Transformations

```{r}
imputation_vars <- c(
  "id",
  "tx",
  "y_qol",
  "baseline_q30",
  "gender",
  "age_std",
  "ecog_fstcnt",
  "diagnosis",
  # "plan_fstcnt_coded",
  "week",
  "albumin_std",
  "crp_std"
)

data <- data_raw |>
  left_join(data_wide, by = "id") |>
  filter(day >0, status_death == 0, !is.na(y_qol)) |>
  mutate(
    week = ceiling(day / 7),
    diagnosis = fct_lump_n(diagnosis, n = 4),
    diagnosis = factor(as.integer(diagnosis)),
    baseline_q30 = factor(baseline_q30),
    y_qol       = as.factor(as.character(y_qol)),
    ecog_fstcnt = as.factor(as.character(ecog_fstcnt)),
    tx = factor(tx)
    ) |>
  select(all_of(imputation_vars)) |>
  arrange(id, week)

# Model non-linear time effects using restricted cubic splines with 4 knots
basis_rcs <- rms::rcs(data$week, 4)

data$week <- as.vector(basis_rcs[, 1])
data$week_nl1 <- as.vector(basis_rcs[, 2])
data$week_nl2 <- as.vector(basis_rcs[, 3])
```

## Multiple Imputation

### Predictor Matrix Configuration

```{r}
#| output: false
md.pattern(data)

init <- mice(data, maxit = 0)
pred_matrix <- init$predictorMatrix
meth <- init$method

# Set patient ID as clustering variable
pred_matrix[, "id"] <- -2
pred_matrix["id", ] <- 0

# Time variables should not predict static baseline variables
baseline_vars <- c("tx", "age_std", "gender", "diagnosis", "ecog_fstcnt", "baseline_q30")
pred_matrix[c(baseline_vars, "week", "week_nl1", "week_nl2"), c("week", "week_nl1", "week_nl2")] <- 0

# Longitudinal variables: multilevel PMM
meth["y_qol"] <- "2l.pmm"
# meth["yprev"] <- ""

# Baseline variables: cluster-level PMM
meth["ecog_fstcnt"] <- "2lonly.pmm"
# meth["plan_fstcnt_coded"] <- "2lonly.pmm"
meth["diagnosis"] <- "2lonly.pmm"
meth["baseline_q30"] <- "2lonly.pmm"
meth["albumin_std"] <- "2lonly.pmm"
meth["crp_std"] <- "2lonly.pmm"

imputed_data <- mice(data,
                     m = 10,        # 50 in final analysis
                     maxit = 50,    # 50 in final analysis
                     predictorMatrix = pred_matrix,
                     method = meth,
                     donors = 5,
                     seed = 78109823)
```

### Imputation Parameters

For the final trial analysis:

-   Number of imputed datasets: $m = 50$
-   Maximum iterations: $50$
-   Donor pool size: $k = 5$

## Convergence Diagnostics {#sec-diagnostics}

Convergence is assessed via trace plots of the mean and standard deviation of imputed values across iterations.

```{r trace-plots}
#| fig-cap: "Trace plots showing the mean and standard deviation of imputed values across iterations for each imputed dataset."
plot(imputed_data)
```

## Distribution Diagnostics {#sec-distribution}

We compare the distributions of observed and imputed values to assess plausibility.

```{r}
#| fig-cap: "Density plots comparing observed (blue) and imputed (red) values for baseline variables and QoL."
densityplot(imputed_data, ~baseline_q30 + ecog_fstcnt)
```

```{r}
#| fig-cap: "Strip plots of imputed values for baseline variables across imputed datasets."
stripplot(imputed_data, baseline_q30 + ecog_fstcnt + diagnosis ~ .imp)
```

```{r}
#| fig-cap: "Scatter plots of QoL against time, by imputed dataset and treatment group, and against baseline QoL."
# xyplot(imputed_data, y_qol ~ week | .imp, pch = c(1, 20))
# xyplot(imputed_data, y_qol ~ week | tx, pch = c(1, 20))
# xyplot(imputed_data, y_qol ~ baseline_q30 | .imp, pch = c(1, 20))
```

## Export

```{r}
for(i in 1:imputed_data$m){
    complete_data <- complete(imputed_data, i)

    # format for QoL analysis
    complete_data <- complete_data |>
      rename(y = y_qol) |>
      arrange(id, week) |>
      group_by(id) |>
      mutate(
          yprev = lag(y, n = 1, order_by = week),
          yprev = if_else(is.na(yprev), baseline_q30, yprev),
          yprev = factor(yprev, levels = 1:7, ordered = FALSE),
          timeprev = lag(week, n = 1, order_by = week),
          timeprev = if_else(is.na(timeprev), 0, timeprev),
          gap = week - timeprev,
          y = as.numeric(as.character(y)),
          ecog_fstcnt = factor(ecog_fstcnt, levels = 0:3),
          tx = as.numeric(as.character(tx))
      ) |>
      ungroup() |>
      select(-timeprev)

    write_parquet(complete_data, sink = here("0-main-analysis", "QOL", "imp-data", paste0("qol-data-", i, ".parquet")))
}
```
