---
title: "Sensitivity Analysis Including Death - MICE"
---

## Setup

```{r}
#| output: false
#| label: setup
library(here)
library(tidyverse)
library(arrow)
library(patchwork)
library(markov.misc)
library(data.table)
library(Hmisc)
library(rms)
library(VGAM)
library(mice)
library(miceadds)
```

## Load Data

```{r}
data_raw <- read_parquet(here("data", "analysis-data", "liqplat_analysis_data_long.parquet")) |>
  mutate(
    death = if_else(y_taooh == 5, 1, 0)
  )

data_wide <- read_parquet(here("data", "analysis-data", "liqplat_analysis_data_wide.parquet")) |>
  select(id, age, albumin, c_reactive_protein, status_death, surv_time) |>
  mutate(
    albumin_std = as.numeric(scale(albumin)),
    crp_std = as.numeric(scale(c_reactive_protein)),
    age_std = as.numeric(scale(age))
    )

data_wide$na_est <- nelsonaalen(data_wide, timevar = surv_time, statusvar = status_death)
data_wide <- data_wide |> select(-status_death, surv_time)
```

Invert QoL Data for modelling

```{r}
data_raw <- data_raw |>
  mutate(
    baseline_q30 = case_when(
      baseline_q30 == 1 ~ 7,
      baseline_q30 == 2 ~ 6,
      baseline_q30 == 3 ~ 5,
      baseline_q30 == 4 ~ 4,
      baseline_q30 == 5 ~ 3,
      baseline_q30 == 6 ~ 2,
      baseline_q30 == 7 ~ 1,
      TRUE ~ baseline_q30
    ),
    y_qol = case_when(
      y_qol == 1 ~ 7,
      y_qol == 2 ~ 6,
      y_qol == 3 ~ 5,
      y_qol == 4 ~ 4,
      y_qol == 5 ~ 3,
      y_qol == 6 ~ 2,
      y_qol == 7 ~ 1,
      TRUE ~ y_qol
    )
  )
```

## Data Preprocessing

### Inclusion Criteria

The substantive analysis model includes only patients who survived beyond 182 days.

### Variable Transformations

```{r}
imputation_vars <- c(
  "id",
  "tx",
  "y_qol",
  "yprev",
  "baseline_q30",
  "gender",
  "age_std",
  "ecog_fstcnt",
  "diagnosis",
  # "plan_fstcnt_coded",
  "week",
  "crp_std",
  "albumin_std",
  "status_death",
  "death",
  "na_est"
)

data <- data_raw |>
  left_join(data_wide, by = "id") |>
  filter(day > 0) |>
  mutate(
    week = ceiling(day / 7),
    diagnosis = fct_lump_n(diagnosis, n = 4),
    diagnosis = factor(as.integer(diagnosis)),
    y_qol       = factor(y_qol), # try keeping as factor
    tx = factor(tx),
    gender = factor(gender)
    ) |>
  group_by(id, week) |>
  mutate(death = max(death)) |>
  arrange(desc(y_qol)) |>
  slice_head(n = 1) |>
  distinct() |>
  group_by(id) |>
  mutate(
    yprev = as.numeric(as.character(lag(y_qol, 1, order_by = week))),
    yprev = if_else(week == 1, baseline_q30, yprev)) |>
  fill(yprev, .direction = "down") |>
  ungroup() |>
  select(all_of(imputation_vars))

# Model non-linear time effects using restricted cubic splines with 4 knots
basis_rcs <- rms::rcs(data$week, 4)

data$week <- as.vector(basis_rcs[, 1])
data$week_nl1 <- as.vector(basis_rcs[, 2])
data$week_nl2 <- as.vector(basis_rcs[, 3])

# model yprev non-linearly (but not as dummy variable, because that causes convergences issues)
basis_yprev <- rms::rcs(data$yprev, 3)
data$yprev <- as.vector(basis_yprev[, 1])
data$yprev_nl1 <- as.vector(basis_yprev[, 2])
```

## MICE

## Multiple Imputation

### Predictor Matrix Configuration

```{r}
#| output: false
md.pattern(data)

init <- mice(data, maxit = 0)
pred_matrix <- init$predictorMatrix
meth <- init$method

# Don't use death event to predict anything (we just use na_est and status_death)
pred_matrix["death", ] <- 0
pred_matrix[ ,"death"] <- 0
# Set patient ID as clustering variable
pred_matrix["id", ] <- 0
pred_matrix[, "id"] <- -2

# Time variables should not predict static baseline variables
baseline_vars <- c("tx", "age_std", "gender", "diagnosis", "ecog_fstcnt", "baseline_q30")
pred_matrix[c(baseline_vars, "week", "week_nl1", "week_nl2"), c("week", "week_nl1", "week_nl2")] <- 0

# Longitudinal variables: multilevel PMM
meth["y_qol"] <- "2l.pmm"
meth["yprev"] <- "2l.pmm"
meth["yprev_nl1"] <- "2l.pmm"

# Baseline variables: cluster-level PMM
meth["ecog_fstcnt"] <- "2lonly.pmm"
meth["diagnosis"] <- "2lonly.pmm"
meth["baseline_q30"] <- "2lonly.pmm"
meth["albumin_std"] <- "2lonly.pmm"
meth["crp_std"] <- "2lonly.pmm"

imputed_data <- mice(data,
                     m = 10,        # 50 in final analysis
                     maxit = 50,    # 50 in final analysis
                     predictorMatrix = pred_matrix,
                     method = meth,
                     donors = 10, # for this analysis we choose then to increase variability
                     seed = 78109823)
```


## Convergence Diagnostics

Convergence is assessed via trace plots of the mean and standard deviation of imputed values across iterations.

```{r trace-plots}
#| fig-cap: "Trace plots showing the mean and standard deviation of imputed values across iterations for each imputed dataset."
plot(imputed_data)
```

## Distribution Diagnostics

```{r}
week_bin <- cut(data$week,
                breaks = seq(0, max(data$week) + 5, by = 5),
                include.lowest = TRUE)

# 2. Add this new variable to your mice object
# mice::cbind works safely with mids objects
imp_with_bins <- cbind(imputed_data, week_bin = week_bin)

# 3. Plot using the new variable
densityplot(imp_with_bins, ~ y_qol | week_bin,
            main = "Imputed y_qol by 5-Week Packets")
```

We compare the distributions of observed and imputed values to assess plausibility.

```{r}
#| fig-cap: "Density plots comparing observed (blue) and imputed (red) values for baseline variables and QoL."
densityplot(imputed_data)
```

```{r}
#| fig-cap: "Strip plots of imputed values for baseline variables across imputed datasets."
stripplot(imputed_data, ecog_fstcnt + diagnosis ~ .imp)
```

```{r}
#| fig-cap: "Scatter plots of QoL against time, by imputed dataset and treatment group, and against baseline QoL."
xyplot(imputed_data, y_qol ~ week | .imp, pch = c(1, 20))
xyplot(imputed_data, y_qol ~ week | tx, pch = c(1, 20))
```

## Convert MICE Output

```{r}
# Process imputations efficiently
process_imputation <- function(i) {
  completed_qol <- complete(imputed_data, i)
  completed_qol$ecog_fstcnt <- factor(completed_qol$ecog_fstcnt)
  completed_qol$diagnosis <- factor(completed_qol$diagnosis)
  completed_qol |>
    select(
      id,
      tx,
      week,
      y = y_qol,
      death,
      baseline_q30,
      ecog_fstcnt,
      diagnosis) |>
    group_by(id) |>
    mutate(
      y = if_else(death == 1, "8", y),
      yprev = as.numeric(as.character(lag(y, 1, order_by = week))),
      yprev = if_else(week == 1, baseline_q30, yprev)
    ) |>
    ungroup() |>
    mutate(.imp = i)  # Add imputation identifier
}

# Combine all imputations
all_imputations <- bind_rows(lapply(1:imputed_data$m, process_imputation))

# For visualization, use the first imputation
df_cf <- all_imputations |>
  group_by(.imp, id) |>
  complete(week = 1:26) |>
  fill(c(y, tx), .direction = "down") |>
  filter(!(is.na(yprev) & y != 8)) |>
  ungroup() |>
  mutate(.imp = factor(.imp))

df_cf |>
  group_by(.imp, week, tx) |>
  count(y) |>
  mutate(
    prop = n / sum(n)
  ) |>
  group_by(week, tx, y) |>
  mutate(
    max = quantile(prop, 0.9),
    min = quantile(prop, 0.1)
  ) |>
  ungroup() |>
  mutate(tx = factor(tx)) |>
  ggplot(aes(x = week, fill = y, color = y)) +
  geom_ribbon(aes(ymin = min, ymax = max, group = interaction(y, tx)), alpha = 0.2) +
  facet_wrap(~y)

markov.misc::plot_sops(df_cf, time_var = "week", facet_var = ".imp", linetype_var = "tx", geom = "line") +
  scale_fill_brewer(palette = "Dark2")
```

```{r}
for(i in 1:imputed_data$m){
    complete_data <- complete(imputed_data, i)

    # format for QoL analysis
    complete_data <- complete_data |>
      rename(y = y_qol) |>
      arrange(id, week) |>
      group_by(id) |>
      mutate(
          y = as.numeric(as.character(y)),
          y = if_else(death == 1, 8, y),
          yprev = lag(y, n = 1, order_by = week),
          yprev = if_else(is.na(yprev), baseline_q30, yprev),
          yprev = factor(yprev, levels = 1:7, ordered = FALSE),
          ecog_fstcnt = factor(ecog_fstcnt, levels = 0:4),
          tx = as.numeric(as.character(tx))
      ) |>
      ungroup()

    write_parquet(complete_data, sink = here("0-main-analysis", "QOL", "sens-1", "imp-data", paste0("qol-data-", i, ".parquet")))
}
```
