---
title: "QoL death-inclusive Markov — imputation"
subtitle: "Two-stage imputation of the weekly QoL data"
format: html
---

```{r}
#| label: setup
#| include: false

library(mice)
library(Hmisc)
library(arrow)
library(here)
library(yaml)
library(tidyverse)

config <- read_yaml(here("config", "analysis.yml"))
n_imputations <- config$imputation$m_supporting
indices <- seq_len(n_imputations)

input_path <- here(
  "analysis", "03-quality-of-life", "results",
  "death-inclusive-longitudinal-prepared.parquet"
)
results_dir <- here(
  "analysis", "03-quality-of-life", "03-longitudinal-with-death", "results"
)
imputation_dir <- file.path(results_dir, "imputations")
baseline_imputation_path <- file.path(results_dir, "baseline-imputations.rds")

dir.create(imputation_dir, recursive = TRUE, showWarnings = FALSE)
```

## Prepare the weekly imputation data

```{r}
#| label: prepare-weekly-data
#| include: false

prepared_full <- read_parquet(input_path) |>
  as.data.frame() |>
  arrange(id, week)

required_columns <- c(
  "id", "tx", "week", "q30", "baseline_q30", "event_death",
  "survival_time_days", "death_week", "ecog_fstcnt", "diagnosis_code"
)
stopifnot(
  all(required_columns %in% names(prepared_full)),
  !anyNA(prepared_full$id),
  !anyDuplicated(prepared_full[c("id", "week")])
)

patient_survival <- prepared_full |>
  distinct(id, survival_time_days, event_death)

patient_survival$na_est <- nelsonaalen(
  patient_survival,
  timevar = survival_time_days,
  statusvar = event_death
)

mean_followup_qol <- prepared_full |>
  filter(!is.na(q30), q30 %in% 1:7) |>
  summarise(mean_observed_q30 = mean(as.numeric(q30)), .by = id)

week_basis <- rcspline.eval(1:26, nk = 4, inclx = TRUE) |>
  as.data.frame() |>
  set_names(c("week", "week_rcs_1", "week_rcs_2"))

prepared_full <- prepared_full |>
  left_join(patient_survival |> select(id, na_est), by = "id") |>
  left_join(week_basis, by = "week") |>
  mutate(
    gender = factor(gender),
    age_scaled = (age - mean(age, na.rm = TRUE)) / sd(age, na.rm = TRUE),
    ecog_fstcnt = factor(
      if_else(ecog_fstcnt == 4, 3, ecog_fstcnt),
      ordered = TRUE
    ),
    plan_fstcnt_coded = fct_lump_n(factor(plan_fstcnt_coded), n = 4),
    diagnosis_code = factor(diagnosis_code),
    stage_binary = factor(stage_binary),
    baseline_q30 = factor(baseline_q30, levels = 1:7, ordered = TRUE)
  )
```

```{r}
#| label: split-living-and-death-rows
#| include: false

prepared <- prepared_full |>
  filter(event_death == 0L | week < death_week) |>
  select(-y_death_inclusive)

death_template <- prepared_full |>
  filter(event_death == 1L, week == death_week) |>
  select(-y_death_inclusive)

decedent_ids <- prepared_full |>
  filter(event_death == 1L) |>
  distinct(id) |>
  pull(id)

stopifnot(
  "The death template must contain exactly one death row per decedent" =
    !anyNA(death_template$id) &&
      !anyDuplicated(death_template[c("id", "week")]) &&
      nrow(death_template) == length(decedent_ids) &&
      setequal(death_template$id, decedent_ids)
)

write_parquet(
  prepared,
  file.path(results_dir, "prepared.parquet"),
  compression = "zstd"
)
write_parquet(
  death_template,
  file.path(results_dir, "death-template.parquet"),
  compression = "zstd"
)
```

## Impute patient-level variables

```{r}
#| label: impute-baseline
#| include: false

baseline_variables <- c(
  "age_scaled", "gender", "ecog_fstcnt", "diagnosis_code", "stage_binary",
  "plan_fstcnt_coded", "albumin", "c_reactive_protein", "baseline_q30"
)
baseline_targets <- c(
  "ecog_fstcnt", "diagnosis_code", "plan_fstcnt_coded", "albumin",
  "c_reactive_protein", "baseline_q30", "mean_observed_q30"
)
completed_baseline_variables <- setdiff(baseline_targets, "mean_observed_q30")

# Column order drives the MICE visit sequence; keep it.
baseline_data <- prepared_full |>
  distinct(id, .keep_all = TRUE) |>
  select(id, tx, all_of(baseline_variables), event_death, na_est) |>
  left_join(mean_followup_qol, by = "id")

if (file.exists(baseline_imputation_path)) {
  baseline_imputations <- readRDS(baseline_imputation_path)
} else {
  methods <- make.method(baseline_data)
  methods[setdiff(names(methods), baseline_targets)] <- ""

  predictors <- make.predictorMatrix(baseline_data)
  predictors[, "id"] <- 0L

  baseline_imputations <- mice(
    baseline_data,
    m = n_imputations,
    maxit = config$imputation$maxit,
    method = methods,
    predictorMatrix = predictors,
    donors = config$imputation$donors,
    seed = config$seeds$supporting + 500,
    printFlag = TRUE
  )

  saveRDS(baseline_imputations, baseline_imputation_path, compress = "xz")
}

stopifnot(
  "The cached baseline MICE object has the wrong number of datasets" =
    baseline_imputations$m == n_imputations
)
```

## Configure the longitudinal QoL imputation

- Previous-state predictor: observed Q30 carried forward; completed baseline Q30 until the first one.
- Nelson--Aalen estimate and death indicator are predictors: informative about missing Q30.

```{r}
#| label: configure-longitudinal-imputation
#| include: false

history_proxy <- prepared_full |>
  select(id, week, q30) |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(yprev_imputation = lag(q30)) |>
  fill(yprev_imputation, .direction = "down") |>
  ungroup() |>
  select(id, week, yprev_imputation)

q30_predictors <- c(
  "tx", "age_scaled", "gender", "ecog_fstcnt", "diagnosis_code",
  "stage_binary", "plan_fstcnt_coded", "albumin", "c_reactive_protein",
  "baseline_q30", "event_death", "na_est", "week", "week_rcs_1",
  "week_rcs_2", "yprev_imputation"
)

analysis_baseline_variables <- c(
  "tx", "age", "gender", "diagnosis_code", "stage_binary", "ecog_fstcnt",
  "plan_fstcnt_coded", "albumin", "c_reactive_protein"
)
```

## Impute weekly QoL and add death

- Death is never a MICE target; the observed death row is reattached afterwards.

```{r}
#| label: impute-weekly-qol
#| include: false

expected_keys <- bind_rows(
  prepared |> select(id, week),
  death_template |> select(id, week)
)

run_summary <- list()

for (index in indices) {
  long_imputation_path <- file.path(
    imputation_dir,
    sprintf("long-imputation-%02d.rds", index)
  )

  completed_baseline <- complete(baseline_imputations, index) |>
    as.data.frame() |>
    select(id, all_of(completed_baseline_variables))

  stopifnot(
    !anyNA(completed_baseline$id),
    !anyDuplicated(completed_baseline$id),
    setequal(completed_baseline$id, baseline_data$id)
  )

  if (file.exists(long_imputation_path)) {
    long_imputation <- readRDS(long_imputation_path)
  } else {
    imputation_data <- prepared |>
      select(-all_of(completed_baseline_variables)) |>
      left_join(completed_baseline, by = "id", relationship = "many-to-one") |>
      left_join(
        history_proxy,
        by = c("id", "week"),
        relationship = "one-to-one"
      ) |>
      mutate(
        yprev_imputation = coalesce(
          yprev_imputation,
          as.numeric(as.character(baseline_q30))
        ),
        yprev_imputation = factor(yprev_imputation, levels = 1:7),
        q30 = factor(q30, levels = 1:7, ordered = TRUE),
        tx = as.numeric(as.character(tx))
      )

    stopifnot(
      !anyNA(imputation_data$id),
      !anyDuplicated(imputation_data[c("id", "week")])
    )

    missing_predictors <- imputation_data[q30_predictors] |>
      keep(anyNA) |>
      names()
    if (length(missing_predictors) > 0L) {
      stop(
        "Longitudinal imputation ", index,
        " has missing values in completed predictors: ",
        str_flatten_comma(missing_predictors)
      )
    }

    methods <- make.method(imputation_data)
    methods[] <- ""
    methods["q30"] <- "polr"

    predictors <- make.predictorMatrix(imputation_data)
    predictors[, ] <- 0L
    predictors["q30", q30_predictors] <- 1L

    long_imputation <- mice(
      imputation_data,
      m = 1L,
      maxit = config$imputation$maxit,
      method = methods,
      predictorMatrix = predictors,
      seed = config$seeds$supporting + (index - 1) * 1000 + 550,
      polr.to.loggedEvents = TRUE,
      printFlag = TRUE
    )

    saveRDS(long_imputation, long_imputation_path, compress = "xz")
  }

  living_completed <- complete(long_imputation, 1L) |>
    as.data.frame() |>
    mutate(
      q30 = as.integer(as.character(q30)),
      y = q30
    )

  stopifnot(
    "MICE did not preserve the living patient-week rows" =
      !anyNA(living_completed$id) &&
        !anyNA(living_completed$y) &&
        !anyDuplicated(living_completed[c("id", "week")]) &&
        setequal(
          select(living_completed, id, week),
          select(prepared, id, week)
        )
  )

  death_rows <- death_template |>
    select(-all_of(completed_baseline_variables)) |>
    left_join(completed_baseline, by = "id", relationship = "many-to-one") |>
    mutate(q30 = NA_integer_, y = 8L)

  stopifnot(
    "The death rows do not match the observed death template" =
      !anyNA(death_rows$id) &&
        nrow(death_rows) == nrow(death_template) &&
        setequal(
          select(death_rows, id, week),
          select(death_template, id, week)
        )
  )

  completed <- bind_rows(living_completed, death_rows) |>
    arrange(id, week) |>
    group_by(id) |>
    mutate(
      yprev = coalesce(
        lag(y),
        as.integer(as.character(first(baseline_q30)))
      ),
      yprev = factor(yprev, levels = 1:7)
    ) |>
    ungroup() |>
    select(
      id, y, q30, yprev, week, baseline_q30,
      all_of(analysis_baseline_variables)
    )

  stopifnot(
    "A completed dataset failed the patient, week, outcome or death checks" =
      !anyNA(completed$id) &&
        !anyNA(completed$y) &&
        !anyDuplicated(completed[c("id", "week")]) &&
        setequal(select(completed, id, week), expected_keys) &&
        setequal(completed$id, prepared_full$id) &&
        setequal(completed$id[completed$y == 8L], death_template$id) &&
        all(completed$y %in% 1:8)
  )

  write_parquet(
    completed,
    file.path(imputation_dir, sprintf("imputation-%02d.parquet", index)),
    compression = "zstd"
  )

  run_summary[[index]] <- tibble(
    imputation = index,
    rows = nrow(completed),
    participants = n_distinct(completed$id),
    imputed_q30 = nrow(long_imputation$imp$q30),
    deaths = sum(completed$y == 8L),
    logged_events = if (is.null(long_imputation$loggedEvents)) {
      0L
    } else {
      nrow(long_imputation$loggedEvents)
    }
  )
}

run_summary <- list_rbind(run_summary)

write_parquet(
  run_summary,
  file.path(results_dir, "imputation-run-summary.parquet"),
  compression = "zstd"
)
```
