---
title: "QoL death-inclusive Markov — imputation diagnostics"
subtitle: "Baseline checks and longitudinal Q30 variability"
format:
  html:
    toc: true
---

```{r}
#| label: setup
#| include: false

library(mice)
library(knitr)
library(scales)
library(arrow)
library(here)
library(yaml)
library(tidyverse)
source(here("R", "plot-style.R"))

config <- read_yaml(here("config", "analysis.yml"))
n_imputations <- config$imputation$m_supporting
indices <- seq_len(n_imputations)

results_dir <- here(
  "analysis", "03-quality-of-life", "03-longitudinal-with-death", "results"
)
imputation_dir <- file.path(results_dir, "imputations")
table_dir <- file.path(results_dir, "imputation-diagnostics")
figure_dir <- file.path(results_dir, "figures", "imputation")

dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

prepared <- read_parquet(file.path(results_dir, "prepared.parquet"))
death_template <- read_parquet(file.path(results_dir, "death-template.parquet"))
baseline_imputations <- readRDS(
  file.path(results_dir, "baseline-imputations.rds")
)
run_summary <- read_parquet(
  file.path(results_dir, "imputation-run-summary.parquet")
)

completed_long <- indices |>
  map(\(index) {
    read_parquet(file.path(
      imputation_dir,
      sprintf("imputation-%02d.parquet", index)
    )) |>
      mutate(imputation = index)
  }) |>
  list_rbind()

stopifnot(
  baseline_imputations$m == n_imputations,
  identical(run_summary$imputation, indices)
)
```

## Completion and death checks

```{r}
#| label: tbl-completion-checks
#| tbl-cap: "Patient, outcome, MICE event and deterministic death checks for all completed datasets"
#| echo: false

completion_qc <- completed_long |>
  summarise(
    rows = n(),
    participants = n_distinct(id),
    missing_id = sum(is.na(id)),
    duplicate_patient_weeks = sum(duplicated(paste(id, week))),
    missing_outcomes = sum(is.na(y)),
    death_rows = sum(y == 8L),
    death_participants = n_distinct(id[y == 8L]),
    .by = imputation
  ) |>
  left_join(
    run_summary |> select(imputation, imputed_q30, logged_events),
    by = "imputation",
    relationship = "one-to-one"
  )

death_qc <- completed_long |>
  filter(y == 8L) |>
  select(imputation, id, week) |>
  left_join(
    death_template |> select(id, expected_death_week = week),
    by = "id",
    relationship = "many-to-one"
  ) |>
  summarise(
    death_rows = n(),
    expected_death_rows = nrow(death_template),
    missing_expected_id = sum(is.na(expected_death_week)),
    wrong_death_week = sum(week != expected_death_week, na.rm = TRUE),
    .by = imputation
  )

stopifnot(
  "A completed dataset failed its ID, outcome or death checks" =
    all(completion_qc$missing_id == 0L) &&
      all(completion_qc$duplicate_patient_weeks == 0L) &&
      all(completion_qc$missing_outcomes == 0L) &&
      all(death_qc$missing_expected_id == 0L) &&
      all(death_qc$wrong_death_week == 0L) &&
      all(death_qc$death_rows == death_qc$expected_death_rows)
)

write_parquet(
  completion_qc,
  file.path(table_dir, "completion-qc.parquet"),
  compression = "zstd"
)
write_parquet(
  death_qc,
  file.path(table_dir, "death-qc.parquet"),
  compression = "zstd"
)

completion_qc |>
  left_join(death_qc, by = "imputation", relationship = "one-to-one") |>
  kable()
```

## Baseline imputation

- Baseline Q30 is shown because it sets the first previous state, though it is not a covariate.

```{r}
#| label: fig-baseline-imputation
#| fig-cap: "Baseline MICE traces and observed-versus-completed distributions"
#| fig-subcap:
#|   - "MICE chain means and standard deviations for imputed baseline Q30 and ECOG"
#|   - "Category proportions among observed values (black) and imputed values of each completed dataset (pink)"
#| fig-alt: "The first panel shows one MICE chain per completed dataset over 50 iterations for the mean and standard deviation of baseline Q30 and ECOG. The second compares the category proportions of observed and imputed baseline Q30 and ECOG."
#| layout-ncol: 1
#| fig-width: 8
#| fig-height: 4.5
#| echo: false

baseline_labels <- c(
  baseline_q30 = "Baseline Q30",
  ecog_fstcnt = "Baseline ECOG"
)

baseline_trace_plot <- mice_chain_stats(baseline_imputations) |>
  filter(variable %in% names(baseline_labels)) |>
  mutate(variable = unname(baseline_labels[variable])) |>
  plot_mice_traces()

# fct_inseq(): numeric category order across variables (ECOG starts at 0).
baseline_distribution_plot <- baseline_imputations |>
  imputed_category_proportions(set_names(names(baseline_labels), baseline_labels)) |>
  mutate(category = fct_inseq(category)) |>
  plot_imputed_categories()

save_figure(
  baseline_trace_plot,
  file.path(figure_dir, "baseline-traces.png"),
  height = 4.5
)
save_figure(
  baseline_distribution_plot,
  file.path(figure_dir, "baseline-distributions.png"),
  height = 4
)

baseline_trace_plot
baseline_distribution_plot
```

## Longitudinal Q30 convergence

```{r}
#| label: fig-longitudinal-traces
#| fig-cap: "Longitudinal Q30 MICE traces across independent completions"
#| fig-alt: "Two panels show the mean and standard deviation of imputed Q30 over 50 MICE iterations, with one line for each independent completion."
#| fig-width: 8
#| fig-height: 3
#| echo: false

# One single-imputation MICE run per completed dataset.
long_trace_plot <- indices |>
  map(\(index) {
    readRDS(file.path(
      imputation_dir,
      sprintf("long-imputation-%02d.rds", index)
    ))
  }) |>
  mice_chain_stats() |>
  filter(variable == "q30") |>
  mutate(variable = "Q30") |>
  plot_mice_traces()

save_figure(
  long_trace_plot,
  file.path(figure_dir, "longitudinal-q30-traces.png"),
  height = 3
)

long_trace_plot
```

## Observed and imputed Q30 distributions

```{r}
#| label: fig-longitudinal-distributions
#| fig-cap: "Observed and imputed longitudinal Q30 distributions"
#| fig-subcap:
#|   - "Smoothed densities: observed Q30 in black and imputed Q30 by completed dataset in pink"
#|   - "Exact ordinal state proportions: observed Q30 in black and imputed Q30 by completed dataset in pink"
#|   - "Death-inclusive model-row proportions for living Q30 states 1--7 and observed death state 8, one line per completed dataset"
#| fig-alt: "Three panels compare observed Q30 with imputed values from all completed datasets using smoothed densities and exact state proportions, then show the final model-row distribution including death state eight."
#| layout-ncol: 1
#| fig-width: 8
#| fig-height: 4.5
#| echo: false

living_values <- completed_long |>
  filter(y != 8L) |>
  select(imputation, id, week, tx, q30) |>
  left_join(
    prepared |> transmute(id, week, observed_q30 = as.integer(q30)),
    by = c("id", "week"),
    relationship = "many-to-one"
  ) |>
  mutate(source = if_else(is.na(observed_q30), "Imputed", "Observed"))

observed_q30 <- prepared |>
  filter(!is.na(q30)) |>
  transmute(
    source = "Observed",
    imputation = NA_integer_,
    value = as.numeric(q30)
  )
imputed_q30 <- living_values |>
  filter(source == "Imputed") |>
  transmute(source, imputation, value = as.numeric(q30))

q30_density_plot <- plot_imputed_densities(
  observed_q30 |> mutate(variable = "Q30"),
  imputed_q30 |> mutate(variable = "Q30", dataset = imputation)
) +
  scale_x_continuous(breaks = 1:7) +
  labs(x = "Q30 state (1 = excellent, 7 = very poor)")

q30_state_proportions <- bind_rows(observed_q30, imputed_q30) |>
  count(source, imputation, state = value, name = "n") |>
  mutate(proportion = n / sum(n), .by = c(source, imputation))

q30_state_proportion_plot <- q30_state_proportions |>
  transmute(
    variable = "Q30",
    category = factor(state),
    source,
    dataset = imputation,
    proportion
  ) |>
  plot_imputed_categories() +
  labs(x = "Q30 state (1 = excellent, 7 = very poor)")

model_state_proportion_plot <- completed_long |>
  count(imputation, state = y, name = "n") |>
  mutate(proportion = n / sum(n), .by = imputation) |>
  ggplot(aes(x = state, y = proportion, group = imputation)) +
  geom_line(colour = imputation_colours[["Imputed"]], alpha = 0.7) +
  geom_point(colour = imputation_colours[["Imputed"]], size = 1.5, alpha = 0.7) +
  scale_x_continuous(breaks = 1:8, labels = c(1:7, "Death")) +
  labs(
    x = "Model outcome state",
    y = "Proportion of transition rows"
  )

save_figure(
  q30_density_plot,
  file.path(figure_dir, "longitudinal-q30-density.png"),
  height = 4.5
)
save_figure(
  q30_state_proportion_plot,
  file.path(figure_dir, "longitudinal-q30-state-proportions.png"),
  height = 4.5
)
save_figure(
  model_state_proportion_plot,
  file.path(figure_dir, "death-inclusive-model-state-proportions.png"),
  height = 4.5
)

q30_density_plot
q30_state_proportion_plot
model_state_proportion_plot
```

## How variable are the imputations?

- Only cells where living Q30 was originally missing.

```{r}
#| label: tbl-imputation-variability
#| tbl-cap: "Between-imputation variability for originally missing living Q30 values"
#| echo: false

imputed_cells <- living_values |>
  filter(source == "Imputed") |>
  select(imputation, id, week, tx, q30)

cell_variability <- imputed_cells |>
  summarise(
    imputations = n(),
    distinct_states = n_distinct(q30),
    state_range = max(q30) - min(q30),
    state_sd = sd(q30),
    entropy = {
      probabilities <- table(q30) / n()
      -sum(probabilities * log(probabilities)) / log(7)
    },
    .by = c(id, week, tx)
  )

stopifnot(
  "Every missing living Q30 cell must be present in all imputations" =
    all(cell_variability$imputations == n_imputations)
)

participant_variability <- imputed_cells |>
  summarise(mean_imputed_q30 = mean(q30), .by = c(id, imputation)) |>
  summarise(
    between_imputation_sd = sd(mean_imputed_q30),
    between_imputation_range = max(mean_imputed_q30) - min(mean_imputed_q30),
    .by = id
  )

variability_summary <- cell_variability |>
  summarise(
    missing_patient_weeks = n(),
    proportion_identical = mean(distinct_states == 1L),
    proportion_variable = mean(distinct_states > 1L),
    median_distinct_states = median(distinct_states),
    q90_distinct_states = quantile(distinct_states, 0.90, names = FALSE),
    mean_state_range = mean(state_range),
    mean_normalized_entropy = mean(entropy),
    median_participant_mean_sd = median(
      participant_variability$between_imputation_sd
    ),
    q90_participant_mean_sd = quantile(
      participant_variability$between_imputation_sd,
      0.90,
      names = FALSE
    )
  )

wide_cells <- imputed_cells |>
  select(id, week, imputation, q30) |>
  pivot_wider(
    names_from = imputation,
    values_from = q30,
    names_prefix = "imp_"
  )

pairwise_agreement <- combn(indices, 2L, simplify = FALSE) |>
  map(\(pair) {
    first <- wide_cells[[paste0("imp_", pair[[1L]])]]
    second <- wide_cells[[paste0("imp_", pair[[2L]])]]
    tibble(
      imputation_1 = pair[[1L]],
      imputation_2 = pair[[2L]],
      exact_agreement = mean(first == second),
      mean_absolute_difference = mean(abs(first - second))
    )
  }) |>
  list_rbind()

weekly_variability <- living_values |>
  summarise(
    mean_q30 = mean(q30),
    proportion_good = mean(q30 <= 3L),
    .by = c(imputation, tx, week)
  ) |>
  summarise(
    mean_q30_sd = sd(mean_q30),
    mean_q30_min = min(mean_q30),
    mean_q30_max = max(mean_q30),
    proportion_good_sd = sd(proportion_good),
    proportion_good_min = min(proportion_good),
    proportion_good_max = max(proportion_good),
    mean_q30 = mean(mean_q30),
    proportion_good = mean(proportion_good),
    .by = c(tx, week)
  ) |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Usual care", "ctDNA"))
  )

write_parquet(
  cell_variability,
  file.path(table_dir, "q30-cell-variability.parquet"),
  compression = "zstd"
)
write_parquet(
  variability_summary,
  file.path(table_dir, "q30-variability-summary.parquet"),
  compression = "zstd"
)
write_parquet(
  pairwise_agreement,
  file.path(table_dir, "q30-pairwise-agreement.parquet"),
  compression = "zstd"
)
write_parquet(
  weekly_variability,
  file.path(table_dir, "q30-weekly-variability.parquet"),
  compression = "zstd"
)

variability_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

- Cells receiving more than one state: `r percent(variability_summary$proportion_variable, accuracy = 0.1)`
- Median distinct states per cell: `r variability_summary$median_distinct_states`
- Mean pairwise exact agreement: `r percent(mean(pairwise_agreement$exact_agreement), accuracy = 0.1)`
- Median participant-level between-imputation SD of mean Q30: `r round(variability_summary$median_participant_mean_sd, 2)`

```{r}
#| label: fig-cell-variability
#| fig-cap: "Patient-week variability in imputed Q30"
#| fig-subcap:
#|   - "Number of distinct Q30 states assigned to each missing patient-week"
#|   - "Mean normalized imputation entropy by week; ribbons show the interquartile range across missing patient-weeks"
#| fig-alt: "A bar chart counts how many distinct Q30 states each missing patient-week receives across imputations. A second chart shows normalized entropy over weeks by randomized group."
#| layout-ncol: 1
#| fig-width: 8
#| fig-height: 5
#| echo: false

cell_variability_plot <- ggplot(
  cell_variability,
  aes(x = factor(distinct_states))
) +
  geom_bar(colour = "white") +
  labs(
    x = "Distinct imputed Q30 states across datasets",
    y = "Missing patient-week cells"
  )

entropy_plot <- cell_variability |>
  summarise(
    mean_entropy = mean(entropy),
    q25 = quantile(entropy, 0.25, names = FALSE),
    q75 = quantile(entropy, 0.75, names = FALSE),
    .by = c(tx, week)
  ) |>
  mutate(tx = group_factor(tx)) |>
  ggplot(aes(x = week, y = mean_entropy, colour = tx, fill = tx)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_colour_group() +
  scale_fill_group() +
  labs(
    x = "Week after random selection",
    y = "Normalized entropy of imputed Q30",
    colour = NULL,
    fill = NULL
  )

save_figure(
  cell_variability_plot,
  file.path(figure_dir, "q30-distinct-states.png")
)
save_figure(entropy_plot, file.path(figure_dir, "q30-entropy-by-week.png"))

cell_variability_plot
entropy_plot
```

```{r}
#| label: fig-weekly-variability
#| fig-cap: "Between-imputation variability in weekly Q30 summaries"
#| fig-subcap:
#|   - "Mean living Q30 state; ribbons span the minimum to maximum across completed datasets"
#|   - "Proportion with good-or-better living Q30; ribbons span the minimum to maximum across completed datasets"
#| fig-alt: "Two line charts show weekly mean living Q30 and the weekly proportion in states one through three by randomized group. Shaded bands show the full range across completed datasets."
#| layout-ncol: 1
#| fig-width: 8
#| fig-height: 5
#| echo: false

# ctDNA first, as in every group plot.
weekly_plot_data <- weekly_variability |>
  mutate(tx = fct_relevel(tx, "ctDNA"))

weekly_mean_plot <- ggplot(
  weekly_plot_data,
  aes(x = week, y = mean_q30, colour = tx, fill = tx)
) +
  geom_ribbon(
    aes(ymin = mean_q30_min, ymax = mean_q30_max),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 0.8) +
  scale_colour_group() +
  scale_fill_group() +
  labs(
    x = "Week after random selection",
    y = "Mean living Q30 state",
    colour = NULL,
    fill = NULL
  )

weekly_good_plot <- ggplot(
  weekly_plot_data,
  aes(x = week, y = proportion_good, colour = tx, fill = tx)
) +
  geom_ribbon(
    aes(ymin = proportion_good_min, ymax = proportion_good_max),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 0.8) +
  scale_colour_group() +
  scale_fill_group() +
  labs(
    x = "Week after random selection",
    y = "Proportion with good-or-better living Q30",
    colour = NULL,
    fill = NULL
  )

save_figure(weekly_mean_plot, file.path(figure_dir, "q30-weekly-mean-range.png"))
save_figure(weekly_good_plot, file.path(figure_dir, "q30-weekly-good-range.png"))

weekly_mean_plot
weekly_good_plot
```

```{r}
#| label: fig-pairwise-agreement
#| fig-cap: "Pairwise exact agreement of imputed Q30 values"
#| fig-alt: "A heat map gives the proportion of originally missing patient-week Q30 values assigned the same state in each pair of completed datasets."
#| fig-width: 8
#| fig-height: 5
#| echo: false

# Mirror the upper triangle and add the diagonal for a full square heat map.
agreement_matrix <- bind_rows(
  pairwise_agreement |> select(imputation_1, imputation_2, exact_agreement),
  pairwise_agreement |>
    transmute(
      imputation_1 = imputation_2,
      imputation_2 = imputation_1,
      exact_agreement
    ),
  tibble(imputation_1 = indices, imputation_2 = indices, exact_agreement = 1)
)

agreement_plot <- ggplot(
  agreement_matrix,
  aes(
    x = factor(imputation_1),
    y = factor(imputation_2),
    fill = exact_agreement
  )
) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", exact_agreement)), size = 3) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  coord_equal() +
  labs(x = "Imputation", y = "Imputation", fill = "Exact agreement")

save_figure(agreement_plot, file.path(figure_dir, "q30-pairwise-agreement.png"))

agreement_plot
```

## Raw observed trajectories

- Lines join actual assessments only; baseline is the day-0 cohort Q30.

```{r}
#| label: fig-raw-trajectories
#| fig-cap: "Raw observed longitudinal Q30 values by randomized group"
#| fig-alt: "Two panels show observed Q30 states from excellent to very poor over 182 days. Transparent lines connect repeated observations from the same participant and points mark actual questionnaire dates."
#| fig-width: 8
#| fig-height: 4.5
#| echo: false

cohort <- read_parquet(
  here("analysis", "00-data", "results", "analysis-cohort.parquet")
)
qol <- read_parquet(
  here("analysis", "00-data", "results", "analysis-qol.parquet")
)

baseline_qol <- cohort |>
  transmute(id, assessment_id = NA_character_, day = 0, q30 = baseline_q30) |>
  filter(!is.na(q30))

followup_qol <- qol |>
  filter(!is.na(q30), day > 0, day <= config$analysis$horizon_days) |>
  select(id, assessment_id, day, q30)

raw_qol <- bind_rows(baseline_qol, followup_qol) |>
  left_join(cohort |> select(id, tx), by = "id") |>
  distinct(id, assessment_id, day, q30, .keep_all = TRUE) |>
  arrange(id, day) |>
  mutate(tx = group_factor(tx))

raw_trajectory_plot <- ggplot(raw_qol, aes(x = day, y = q30, group = id)) +
  geom_line(colour = "grey35", alpha = 0.20, linewidth = 0.45) +
  geom_point(
    colour = "black",
    fill = "grey70",
    shape = 21,
    size = 1.8,
    stroke = 0.35,
    alpha = 0.50,
    position = position_jitter(width = 0, height = 0.035)
  ) +
  facet_wrap(vars(tx), nrow = 1) +
  scale_x_continuous(
    breaks = c(0, 28, 56, 84, 112, 140, 168, 182),
    limits = c(0, config$analysis$horizon_days)
  ) +
  scale_y_continuous(breaks = 1:7, limits = c(0.8, 7.2)) +
  labs(
    x = "Day since first contact",
    y = "Q30 state (1 = excellent, 7 = very poor)"
  )

save_figure(
  raw_trajectory_plot,
  file.path(figure_dir, "raw-qol-trajectories.png"),
  height = 4.5
)

raw_trajectory_plot
```
