---
title: "Blood products — data and imputation"
format: html
execute:
  echo: true
  warning: false
  message: false
---

The outcomes are the numbers of erythrocyte concentrates (EK), thrombocyte
concentrates (TK), and their sum (all blood products) administered during the
first 182 days after randomization. Exposure is the time each patient was
observed during that period, from randomization to `follow_up_end_date`, which
already includes the 182-day censoring and death.

```{r}
#| label: setup

library(tidyverse)
library(mice)

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")

cohort_path <- here::here(
  "analysis", "00-data", "results", "analysis-cohort.parquet"
)
result_dir <- here::here("results", "secondary", "blood-products")
figure_dir <- file.path(result_dir, "figures")
imputation_path <- file.path(result_dir, "imputation.rds")
completed_path <- file.path(result_dir, "completed-datasets.rds")

dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
```

## Data

The export leaves the combined count `n_blood_products` empty when a patient
received no blood products, so we rebuild it from the two component counts,
which are complete.

```{r}
#| label: blood-data

cohort <- arrow::read_parquet(cohort_path)

blood_data <- cohort |>
  mutate(
    n_blood_products = n_erythrocyte_concentrates + n_thrombocyte_concentrates,
    exposure_days = as.numeric(follow_up_end_date - randomization_date),
    ecog_group = case_when(
      ecog_fstcnt <= 1 ~ "0-1",
      ecog_fstcnt == 2 ~ "2",
      ecog_fstcnt >= 3 ~ "3+"
    ),
    ecog_group = factor(ecog_group, levels = c("0-1", "2", "3+"), ordered = TRUE)
  ) |>
  select(
    id, tx,
    n_blood_products, n_erythrocyte_concentrates, n_thrombocyte_concentrates,
    exposure_days,
    age, hemoglobin, platelet_count, creatinine, ecog_group
  )

stopifnot(!anyNA(blood_data$n_blood_products))
```

Patients whose follow-up ended on the day of randomization have no time at
risk and cannot contribute to a rate, so they are excluded.

```{r}
#| label: exclude-zero-exposure

n_zero_exposure <- sum(blood_data$exposure_days == 0)

blood_data <- blood_data |>
  filter(exposure_days > 0)

arrow::write_parquet(
  blood_data,
  file.path(result_dir, "analysis-data.parquet"),
  compression = "zstd"
)

n_zero_exposure
```

## Descriptive summary

```{r}
#| label: tbl-blood-descriptive
#| tbl-cap: "Blood products during the first 182 days by randomized group"

descriptive_summary <- blood_data |>
  mutate(group = if_else(tx == 1, "Invitation group", "External control")) |>
  summarise(
    patients = n(),
    patients_transfused = sum(n_blood_products > 0),
    erythrocyte_concentrates = sum(n_erythrocyte_concentrates),
    thrombocyte_concentrates = sum(n_thrombocyte_concentrates),
    blood_products = sum(n_blood_products),
    exposure_days = sum(exposure_days),
    .by = group
  ) |>
  mutate(n_excluded_zero_exposure = n_zero_exposure)

arrow::write_parquet(
  descriptive_summary,
  file.path(result_dir, "descriptive-summary.parquet"),
  compression = "zstd"
)

descriptive_summary |>
  knitr::kable()
```

## Imputation of baseline covariates

Hemoglobin, platelet count, creatinine, and ECOG are partly missing.

```{r}
#| label: missing-covariates

blood_data |>
  summarise(across(
    c(hemoglobin, platelet_count, creatinine, ecog_group),
    \(x) sum(is.na(x))
  ))
```

## Spline knots

The SAP models hemoglobin, platelet count, age, and creatinine with restricted
cubic splines with three knots. We use natural splines, `splines::ns()`, with
the interior knot at the median and the boundary knots at the 10th and 90th
percentiles of the observed values. These are the knots `rms::rcs()` would use
for three knots, and the two `ns()` basis functions span exactly the same
restricted cubic spline. The knots are fixed here and reused by the imputation
and analysis models.

```{r}
#| label: spline-knots

spline_variables <- c("age", "hemoglobin", "platelet_count", "creatinine")

spline_knots <- blood_data |>
  select(all_of(spline_variables)) |>
  map(\(x) quantile(x, c(0.1, 0.5, 0.9), na.rm = TRUE, names = FALSE))

saveRDS(spline_knots, file.path(result_dir, "spline-knots.rds"))

spline_knots
```

`ns_call()` writes the spline of one variable as code, so that the same call
can be used for passive imputation and in the analysis formula.

```{r}
#| label: ns-call

ns_call <- function(variable) {
  knots <- spline_knots[[variable]]

  sprintf(
    "splines::ns(%s, knots = %s, Boundary.knots = c(%s, %s))",
    variable, knots[2], knots[1], knots[3]
  )
}

ns_call("hemoglobin")
```

## Imputation model

As for the other supporting analyses, we create the supporting number of MICE
datasets. The imputation model contains only the variables of the analysis
models: the covariates, randomized group, both outcome components, and the log
exposure. Laboratory values use predictive mean matching and ECOG uses
proportional odds imputation.

To match the analysis model, the covariates enter each other's imputation
models through the same spline basis. The two basis columns of every spline
variable (`*_ns1`, `*_ns2`) are added to the data and, for incomplete
variables, are updated by passive imputation after each imputed value. Each
incomplete variable is then imputed from the spline bases of the other spline
variables, instead of from their raw values.

```{r}
#| label: imputation-data

add_spline_basis <- function(data, variable) {
  basis <- eval(parse(text = ns_call(variable)), envir = data)

  data[[paste0(variable, "_ns1")]] <- basis[, 1]
  data[[paste0(variable, "_ns2")]] <- basis[, 2]
  data
}

imputation_data <- blood_data |>
  mutate(log_exposure = log(exposure_days)) |>
  select(
    tx,
    n_erythrocyte_concentrates, n_thrombocyte_concentrates, log_exposure,
    all_of(spline_variables), ecog_group
  )

for (variable in spline_variables) {
  imputation_data <- add_spline_basis(imputation_data, variable)
}
```

```{r}
#| label: imputation-methods

incomplete_labs <- c("hemoglobin", "platelet_count", "creatinine")

methods <- make.method(imputation_data)
methods[] <- ""
methods[incomplete_labs] <- "pmm"
methods["ecog_group"] <- "polr"

# Passive imputation keeps the basis columns in line with the imputed values.
for (variable in incomplete_labs) {
  for (column in 1:2) {
    methods[paste0(variable, "_ns", column)] <- sprintf(
      "~I(%s[, %d])", ns_call(variable), column
    )
  }
}

methods[methods != ""]
```

```{r}
#| label: imputation-predictors

basis_columns <- function(variables) {
  paste0(rep(variables, each = 2), "_ns", 1:2)
}

common_predictors <- c(
  "tx", "n_erythrocyte_concentrates", "n_thrombocyte_concentrates",
  "log_exposure"
)

predictors <- make.predictorMatrix(imputation_data)
predictors[] <- 0

for (variable in incomplete_labs) {
  other_splines <- setdiff(spline_variables, variable)

  predictors[variable, c(common_predictors, "ecog_group",
                         basis_columns(other_splines))] <- 1
}

predictors["ecog_group", c(common_predictors,
                           basis_columns(spline_variables))] <- 1

predictors[c(incomplete_labs, "ecog_group"), colSums(predictors) > 0]
```

```{r}
#| label: impute-covariates

if (force || !file.exists(imputation_path)) {
  imputation <- mice(
    imputation_data,
    method = methods,
    predictorMatrix = predictors,
    m = config$imputation$supporting$m,
    maxit = config$imputation$supporting$maxit,
    donors = config$imputation$supporting$donors,
    seed = analysis_seed(config, "supporting", offset = 2400L),
    printFlag = FALSE
  )

  saveRDS(imputation, imputation_path, compress = "xz")
}

imputation <- readRDS(imputation_path)

c(m = imputation$m, iterations = imputation$iteration)
```

```{r}
#| label: fig-imputation-traces
#| fig-cap: "MICE trace plots for the imputed baseline covariates"

plot(imputation, c(incomplete_labs, "ecog_group"))
```

## Completed datasets

The basis columns are dropped again, because the analysis models build the same
splines with `ns()` in their formulas.

```{r}
#| label: completed-datasets

completed_datasets <- complete(imputation, action = "all") |>
  map(\(data) {
    data |>
      mutate(
        n_blood_products = n_erythrocyte_concentrates + n_thrombocyte_concentrates,
        exposure_days = exp(log_exposure),
        ecog_group = factor(ecog_group, ordered = FALSE)
      ) |>
      select(-log_exposure, -ends_with(c("_ns1", "_ns2")))
  })

saveRDS(completed_datasets, completed_path, compress = "xz")
```
