---
title: "Blood products — model fitting"
format: html
execute:
  echo: true
  warning: false
  message: false
---

Following the SAP (*Comparative Clinical Outcomes — Blood Product
Transfusions*), each count is modeled with a Bayesian negative binomial
regression with a `log(exposure)` offset, adjusting for randomized group,
ECOG (0–1, 2, 3+), and restricted cubic splines with three knots for baseline
hemoglobin, platelet count, age, and creatinine.

The prespecified outcome is the combined count of all blood products. The same
model is fitted separately to erythrocyte (EK) and thrombocyte (TK)
concentrates as additional, not prespecified analyses. For transparency, every
outcome is also fitted without covariates (randomized group and offset only).

```{r}
#| label: setup

library(tidyverse)

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()
force <- identical(Sys.getenv("LIQPLAT_FORCE"), "1")

result_dir <- here::here("results", "secondary", "blood-products")

blood_data <- arrow::read_parquet(file.path(result_dir, "analysis-data.parquet"))
completed_datasets <- readRDS(file.path(result_dir, "completed-datasets.rds"))

blood_outcomes <- c(
  blood_products = "n_blood_products",
  erythrocytes = "n_erythrocyte_concentrates",
  thrombocytes = "n_thrombocyte_concentrates"
)
```

## Model formulas

The splines use the knots fixed in the imputation step, written into the
formula as numbers so that predictions on new data use the same basis.

```{r}
#| label: spline-calls

spline_knots <- readRDS(file.path(result_dir, "spline-knots.rds"))

ns_call <- function(variable) {
  knots <- spline_knots[[variable]]

  sprintf(
    "splines::ns(%s, knots = %s, Boundary.knots = c(%s, %s))",
    variable, knots[2], knots[1], knots[3]
  )
}
```

```{r}
#| label: model-formulas-adjusted

adjusted_terms <- paste(
  "tx",
  ns_call("hemoglobin"),
  ns_call("platelet_count"),
  ns_call("age"),
  ns_call("creatinine"),
  "ecog_group",
  "offset(log(exposure_days))",
  sep = " + "
)

unadjusted_terms <- "tx + offset(log(exposure_days))"

blood_formula <- function(outcome, adjustment) {
  terms <- if (adjustment == "adjusted") adjusted_terms else unadjusted_terms

  brms::bf(as.formula(paste(blood_outcomes[[outcome]], "~", terms)))
}
```

## Fitting

Adjusted models are fitted to every completed dataset with `brm_multiple()`,
keeping one fit per imputation so that predictions can use each imputation's
own covariates. Unadjusted models need no imputation and are fitted once to the
observed data. Both are stored as a list of fits.

```{r}
#| label: fit-models

fit_blood_model <- function(outcome, adjustment, seed_offset) {
  model_path <- file.path(
    result_dir, paste0("model-", outcome, "-", adjustment, ".rds")
  )

  if (force || !file.exists(model_path)) {
    seed <- analysis_seed(config, "supporting", offset = seed_offset)

    fits <- if (adjustment == "adjusted") {
      brms::brm_multiple(
        blood_formula(outcome, adjustment),
        data = completed_datasets,
        family = brms::negbinomial(),
        chains = 4L,
        iter = 2000L,
        cores = 1L,
        seed = seed,
        combine = FALSE,
        refresh = 0L
      )
    } else {
      list(brms::brm(
        blood_formula(outcome, adjustment),
        data = blood_data,
        family = brms::negbinomial(),
        chains = 4L,
        iter = 2000L,
        cores = 1L,
        seed = seed,
        refresh = 0L
      ))
    }

    saveRDS(fits, model_path, compress = "xz")
  }

  readRDS(model_path)
}

model_grid <- expand_grid(
  outcome = names(blood_outcomes),
  adjustment = c("adjusted", "unadjusted")
) |>
  mutate(seed_offset = 2400L + row_number())

blood_fits <- model_grid |>
  mutate(fits = pmap(list(outcome, adjustment, seed_offset), fit_blood_model))
```

## Diagnostics

```{r}
#| label: tbl-model-diagnostics
#| tbl-cap: "Largest R-hat and smallest bulk effective sample size per model and imputation"

blood_fits |>
  mutate(diagnostics = map(fits, \(fits) {
    tibble(
      imputation = seq_along(fits),
      max_rhat = map_dbl(fits, \(fit) max(brms::rhat(fit), na.rm = TRUE)),
      min_ess_bulk = map_dbl(fits, \(fit) {
        min(posterior::summarise_draws(fit, "ess_bulk")$ess_bulk, na.rm = TRUE)
      })
    )
  })) |>
  select(outcome, adjustment, diagnostics) |>
  unnest(diagnostics) |>
  knitr::kable(digits = 3)
```

```{r}
#| label: tbl-combined-coefficients
#| tbl-cap: "Posterior coefficients of the adjusted combined blood-product model (first imputation)"

blood_fits |>
  filter(outcome == "blood_products", adjustment == "adjusted") |>
  pull(fits) |>
  pluck(1, 1) |>
  brms::fixef() |>
  knitr::kable(digits = 3)
```
