---
title: "Blood products — marginal rates, rate ratio, and rate difference"
format: html
execute:
  echo: true
  warning: false
  message: false
---

For each outcome and imputation, every patient's expected count is predicted
under both randomized groups with the exposure set to 30 days, and these
predictions are averaged over all patients. This gives the marginal rate per
30 days in each group, their ratio, and their difference. The first
`post_estimation_draws` posterior draws of each imputation are then stacked.
Unadjusted models are summarized the same way, using the observed data.

```{r}
#| label: setup

library(tidyverse)

source(here::here("R", "analysis-config.R"))

config <- read_analysis_config()

result_dir <- here::here("results", "secondary", "blood-products")
figure_dir <- file.path(result_dir, "figures")
completed_path <- file.path(result_dir, "completed-datasets.rds")
blood_draws_path <- file.path(result_dir, "draws.parquet")
blood_summary_path <- file.path(result_dir, "summary.parquet")

blood_data <- arrow::read_parquet(file.path(result_dir, "analysis-data.parquet"))
completed_datasets <- readRDS(completed_path)

blood_fits <- expand_grid(
  outcome = c("blood_products", "erythrocytes", "thrombocytes"),
  adjustment = c("adjusted", "unadjusted")
) |>
  mutate(fits = map2(outcome, adjustment, \(outcome, adjustment) {
    readRDS(file.path(
      result_dir, paste0("model-", outcome, "-", adjustment, ".rds")
    ))
  }))

rate_days <- 30

outcome_labels <- c(
  blood_products = "All blood products (prespecified)",
  erythrocytes = "Erythrocyte concentrates",
  thrombocytes = "Thrombocyte concentrates"
)
```

## Marginal rates

```{r}
#| label: marginal-rates

predict_marginal_rates <- function(fit, data) {
  fit <- marginaleffects::set_modeldata(fit, data)

  counterfactual_data <- bind_rows(
    data |> mutate(tx = 0),
    data |> mutate(tx = 1)
  ) |>
    mutate(exposure_days = rate_days)

  marginaleffects::avg_predictions(
    fit, newdata = counterfactual_data, by = "tx", type = "response"
  ) |>
    marginaleffects::get_draws() |>
    mutate(drawid = as.integer(as.character(drawid))) |>
    filter(drawid <= config$imputation$supporting$post_estimation_draws) |>
    select(tx, drawid, draw)
}

predict_model_rates <- function(fits, adjustment) {
  prediction_data <- if (adjustment == "adjusted") {
    completed_datasets
  } else {
    list(blood_data)
  }

  map2(fits, prediction_data, predict_marginal_rates) |>
    list_rbind(names_to = "imputation")
}

rate_draws <- blood_fits |>
  mutate(draws = map2(fits, adjustment, predict_model_rates)) |>
  select(outcome, adjustment, draws) |>
  unnest(draws)
```

```{r}
#| label: contrasts

blood_draws <- rate_draws |>
  pivot_wider(names_from = tx, values_from = draw, names_prefix = "rate_tx") |>
  mutate(
    rate_ratio = rate_tx1 / rate_tx0,
    rate_difference = rate_tx1 - rate_tx0
  )

arrow::write_parquet(blood_draws, blood_draws_path, compression = "zstd")
```

## Summary

`probability_lower_rate` is the posterior probability that the invitation
group has a lower rate than the external control.

```{r}
#| label: tbl-blood-summary
#| tbl-cap: "Posterior marginal rates per 30 days, rate ratio, and rate difference (intervention - control)"

blood_summary <- blood_draws |>
  pivot_longer(
    c(rate_tx0, rate_tx1, rate_ratio, rate_difference),
    names_to = "quantity", values_to = "draw"
  ) |>
  summarise(
    posterior_median = median(draw),
    lower_95 = quantile(draw, 0.025),
    upper_95 = quantile(draw, 0.975),
    probability_lower_rate = case_when(
      quantity[1] == "rate_ratio" ~ mean(draw < 1),
      quantity[1] == "rate_difference" ~ mean(draw < 0)
    ),
    .by = c(outcome, adjustment, quantity)
  )

arrow::write_parquet(blood_summary, blood_summary_path, compression = "zstd")

blood_summary |>
  knitr::kable(digits = 3)
```

## Figures

The adjusted (prespecified) and unadjusted estimates are shown next to each
other for each outcome.

```{r}
#| label: plot-data

blood_plot_draws <- blood_draws |>
  mutate(
    outcome = factor(outcome_labels[outcome], levels = outcome_labels),
    adjustment = factor(
      str_to_sentence(adjustment), levels = c("Unadjusted", "Adjusted")
    )
  )

comparison_theme <- theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

```{r}
#| label: fig-blood-rate-ratio
#| fig-cap: "Posterior marginal rate ratio (intervention / control), adjusted and unadjusted, for all blood products and for erythrocyte and thrombocyte concentrates separately. Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "Three panels, one per outcome (all blood products, erythrocyte concentrates, thrombocyte concentrates), each with half-eye densities of the marginal rate ratio, intervention over control, on a log scale for the adjusted and unadjusted model, with thick and thin intervals for the 66 and 95 percent credible intervals and a dashed vertical line at a ratio of one."
#| fig-width: 8
#| fig-height: 9

rate_ratio_plot <- ggplot(blood_plot_draws, aes(x = rate_ratio, y = adjustment)) +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  scale_x_log10() +
  facet_wrap(~ outcome, ncol = 1) +
  labs(x = "Rate ratio (intervention / control)", y = NULL) +
  comparison_theme

ggsave(
  file.path(figure_dir, "blood-products-rate-ratio.png"),
  rate_ratio_plot, width = 8, height = 9, dpi = 300
)

rate_ratio_plot
```

```{r}
#| label: fig-blood-rate-difference
#| fig-cap: "Posterior marginal difference in products per 30 days (intervention - control), adjusted and unadjusted, for all blood products and for erythrocyte and thrombocyte concentrates separately. Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "Three panels, one per outcome (all blood products, erythrocyte concentrates, thrombocyte concentrates), each with half-eye densities of the marginal difference in products per 30 days, intervention minus control, for the adjusted and unadjusted model, with thick and thin intervals for the 66 and 95 percent credible intervals and a dashed vertical line at zero."
#| fig-width: 8
#| fig-height: 9

rate_difference_plot <- ggplot(
  blood_plot_draws, aes(x = rate_difference, y = adjustment)
) +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ outcome, ncol = 1, scales = "free_x") +
  labs(x = "Difference in products per 30 days (intervention - control)", y = NULL) +
  comparison_theme

ggsave(
  file.path(figure_dir, "blood-products-rate-difference.png"),
  rate_difference_plot, width = 8, height = 9, dpi = 300
)

rate_difference_plot
```

```{r}
#| label: fig-blood-marginal-rates
#| fig-cap: "Posterior marginal rate per 30 days by randomized group from the adjusted models. Thick and thin intervals show 66% and 95% credible intervals."
#| fig-alt: "Three panels, one per outcome (all blood products, erythrocyte concentrates, thrombocyte concentrates), each with half-eye densities of the marginal number of products per 30 days for the external control and the invitation group from the adjusted model, with thick and thin intervals for the 66 and 95 percent credible intervals."
#| fig-width: 8
#| fig-height: 9

marginal_rates_plot <- rate_draws |>
  filter(adjustment == "adjusted") |>
  mutate(
    outcome = factor(outcome_labels[outcome], levels = outcome_labels),
    group = if_else(tx == 1, "Invitation group", "External control")
  ) |>
  ggplot(aes(x = draw, y = group)) +
  ggdist::stat_halfeye(.width = c(0.66, 0.95), point_interval = ggdist::median_qi) +
  facet_wrap(~ outcome, ncol = 1, scales = "free_x") +
  labs(x = "Products per 30 days", y = NULL) +
  comparison_theme

ggsave(
  file.path(figure_dir, "blood-products-marginal-rates.png"),
  marginal_rates_plot, width = 8, height = 9, dpi = 300
)

marginal_rates_plot
```
