---
title: "Multiple Imputation Overall Survival"
---

## Setup

```{r}
#| output: false
library(tidyverse)
library(here)
library(mice)
library(arrow)
```

## Data

```{r}
data_raw <- read_parquet(here("data", "analysis-data", "liqplat_analysis_data_wide.parquet"))
```

## Data Preprocessing

### Variable Transformations

- Stage is currently missing

```{r}
data <- data_raw |>
  mutate(
    lung = if_else(diagnosis == "SNM_93880001", 1, 0) #SNM Code for Lung Ca
  ) |>
  select(
    id,
    status_death,
    surv_time,
    tx,
    gender,
    age,
    ecog_fstcnt,
    lung,
    albumin,
    bilirubin,
    c_reactive_protein,
    lactate_dehydrogenase
  )
```

### Nelson-Aalen Cumulative Hazard Estimate

We include the Nelson-Aalen cumulative hazard estimate (`na_est`) as a predictor in the imputation model instead of survival time [van-Buuren2021-pc, chapter 9.1].

```{r}
data$na_est <- nelsonaalen(data, timevar = surv_time, statusvar = status_death)
```

## Imputation

For the final trial analysis:

- Number of imputed datasets: $m = 50$
- Maximum iterations: 50
- Donor pool size: $k = 5$

```{r}
#| output: false
md.pattern(data)

init <- mice(data, maxit = 0)
meth <- init$method
pred_matrix <- init$predictorMatrix

# Patient ID should not be used for imputation
pred_matrix[, "id"] <- 0
pred_matrix["id", ] <- 0
meth["id"] <- ""

# Use Nelson-Aalen estimate instead of raw survival time
pred_matrix[, "surv_time"] <- 0
pred_matrix["surv_time", ] <- 0
meth["surv_time"] <- ""

imputed_data <- mice(data,
                     m = 10,      # 50 in final analysis
                     maxit = 10,  # 50 in final analysis
                     predictorMatrix = pred_matrix,
                     method = meth,
                     donors = 5,
                     seed = 78109823)
```

## Convergence Diagnostics

Convergence is assessed via trace plots of the mean and standard deviation of imputed values across iterations.

```{r trace-plots}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Trace plots showing the mean and standard deviation of imputed values across iterations for each imputed dataset."
plot(imputed_data)
```

## Distribution Diagnostics

We compare the distributions of observed and imputed values to assess plausibility.

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Density plots comparing observed (blue) and imputed (red) values for all imputed variables."
densityplot(imputed_data)
```

```{r}
#| fig-cap: "Density plot for ECOG performance status."
densityplot(imputed_data, ~ecog_fstcnt)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Strip plots of imputed values for continuous laboratory variables across imputed datasets."
stripplot(imputed_data, albumin + bilirubin + c_reactive_protein + lactate_dehydrogenase ~ .imp)
```

```{r}
#| fig-cap: "Strip plot of imputed values for ECOG performance status across imputed datasets."
stripplot(imputed_data, ecog_fstcnt ~ .imp)
```

## Export

```{r}
for(i in 1:imputed_data$m){
    complete_data <- complete(imputed_data, i)
    write_parquet(complete_data, sink = here("0-main-analysis", "OS", "imp-data", paste0("os-data-", i, ".parquet")))
}
```
