---
title: "OS Model Fitting"
---

## Setup

```{r}
#| output: false
library(fs)
library(here)
library(tidyverse)
library(arrow)
library(rstanarm)
library(ggdist)
library(survminer)
library(patchwork)
```

## Load and wrangle data

RMST

```{r}
post_stack <- data.frame()

f <- dir_ls(here("0-main-analysis", "OS", "results", "rmst"))

for (i in 1:length(f)){
    post_stack <- post_stack |>
        bind_rows(read_parquet(f[i]))
}
```

Survival curves

```{r}
g <- dir_ls(here("0-main-analysis", "OS", "results", "survival_curves"))

long_data_0 <- data.frame()
long_data_1 <- data.frame()


for (i in 1:length(g)) {
    surv_matrix <- read_rds(g[i])

    long_data_i <- surv_matrix$surv_tx0 |>
    as_tibble(.name_repair = ~ as.character(surv_matrix$time_points)) |>
    mutate(.draw = 1:n()) |>
    pivot_longer(
        cols = !.draw,
        names_to = "time",
        values_to = "survival",
        names_transform = as.numeric
    ) |>
    mutate(
        treatment = "UC",
        .imp = i)

    long_data_0 <- bind_rows(long_data_0, long_data_i)

    long_data_i <- surv_matrix$surv_tx1 |>
    as_tibble(.name_repair = ~ as.character(surv_matrix$time_points)) |>
    mutate(.draw = 1:n()) |>
    pivot_longer(
        cols = !.draw,
        names_to = "time",
        values_to = "survival",
        names_transform = as.numeric
    ) |>
    mutate(
        treatment = "SAT",
        .imp = i)


    long_data_1 <- bind_rows(long_data_1, long_data_i)
}

surv_curv_0 <- long_data_0 |>
    group_by(time, treatment) |>
    summarize(
        median = median(survival),
        mean = mean(survival),
        conf.low.95 = quantile(survival, 0.025),
        conf.high.95 = quantile(survival, 0.975),
        conf.high.66 = quantile(survival, 0.83),
        conf.low.66 = quantile(survival, 0.17),
    ) |>
    ungroup()

surv_curv_1 <- long_data_1 |>
    group_by(time, treatment) |>
    summarize(
        median = median(survival),
        mean = mean(survival),
        conf.low.95 = quantile(survival, 0.025),
        conf.high.95 = quantile(survival, 0.975),
        conf.high.66 = quantile(survival, 0.83),
        conf.low.66 = quantile(survival, 0.17),
    ) |>
    ungroup()

surv_curv <- bind_rows(surv_curv_0, surv_curv_1)

risk <- long_data_1 |>
    bind_rows(long_data_0) |>
    pivot_wider(
        id_cols = c(time, .draw, .imp),
        values_from = "survival",
        names_from = "treatment",
        names_prefix = "surv_"
    ) |>
    mutate(
        risk_SAT = 1 - surv_SAT,
        risk_UC = 1 - surv_UC,
        RR = risk_SAT / risk_UC,
        RD = risk_SAT - risk_UC
    )

conf_width <- risk |>
    group_by(time) |>
    summarise(
        width = quantile(surv_SAT - surv_UC, 0.975) - quantile(surv_SAT - surv_UC, 0.025)
    )

conf_band <- surv_curv |>
    select(time, treatment, median) |>
    pivot_wider(
        id_cols = time,
        names_from = treatment,
        values_from = median
    ) |>
    mutate(
        middle = UC + (SAT - UC)/2
    ) |>
    left_join(conf_width, by = "time") |>
    mutate(
        band.low = middle - width/4,
        band.high = middle + width/4)
```

Kaplan-Meier Curves

```{r}
analysis_data <- read_parquet(here("data", "analysis-data", "liqplat_analysis_data_wide.parquet"))

# Prepare data with treatment labels
analysis_data <- analysis_data |>
  mutate(
    treatment = ifelse(tx == 0, "UC", "SAT")
  )

km_fit <- survival::survfit(survival::Surv(surv_time, status_death) ~ treatment, data = analysis_data)

# Extract KM data including censoring information
km_data <- data.frame(
  time = km_fit$time,
  survival = km_fit$surv,
  n.risk = km_fit$n.risk,
  n.event = km_fit$n.event,
  n.censor = km_fit$n.censor,
  strata = rep(names(km_fit$strata), km_fit$strata)
) |>
  mutate(
    treatment = str_remove(strata, "treatment=")
  )

# Extract censoring times for marks
censor_data <- analysis_data |>
  filter(status_death == 0) |>
  select(surv_time, treatment) |>
  left_join(
    km_data |> select(time, survival, treatment),
    by = c("surv_time" = "time", "treatment")
  ) |>
  filter(!is.na(survival))

# Create risk table data
risk_table_times <- seq(0, 180, by = 20)
risk_table_data <- map_dfr(risk_table_times, function(t) {
  summary_fit <- summary(km_fit, times = t, extend = TRUE)
  data.frame(
    time = t,
    n.risk = summary_fit$n.risk,
    treatment = str_remove(summary_fit$strata, "treatment=")
  )
})
```

## Plot Difference in RMST

```{r}
plot_rmst_diff <- ggplot(post_stack, (aes(x = rmst_diff))) +
    ggdist::stat_dots(side = "bottom") +
    ggdist::stat_halfeye() +
    labs(
        title = "",
        x = "Difference in the restricted mean survival time (SAT - UC) [Days]",
        y = "Density"
    )

ggsave(
    here("0-main-analysis", "img", "os_rmst_difference.svg"),
    plot = plot_rmst_diff,
    width = 8,
    height = 5,
    dpi = 300
)

mean(post_stack$rmst_diff > 0)

rmst_tbl <- post_stack |>
  summarise(
    UC_mean = mean(rmst_0),
    UC_sd = sd(rmst_0),
    UC_median = median(rmst_0),
    UC_lower = quantile(rmst_0, 0.025),
    UC_upper = quantile(rmst_0, 0.975),

    SAT_mean = mean(rmst_1),
    SAT_sd = sd(rmst_1),
    SAT_median = median(rmst_1),
    SAT_lower = quantile(rmst_1, 0.025),
    SAT_upper = quantile(rmst_1, 0.975),

    diff_mean = mean(rmst_diff),
    diff_sd = sd(rmst_diff),
    diff_median = median(rmst_diff),
    diff_lower = quantile(rmst_diff, 0.025),
    diff_upper = quantile(rmst_diff, 0.975),

    p_benefit = mean(rmst_diff > 0)
  ) |>
  mutate(
    Outcome = "RMST",
    `UC Mean (SD)` = sprintf("%.1f (%.1f)", UC_mean, UC_sd),
    `UC Median [95%% CrI]` = sprintf("%.1f [%.1f, %.1f]", UC_median, UC_lower, UC_upper),
    `SAT Mean (SD)` = sprintf("%.1f (%.1f)", SAT_mean, SAT_sd),
    `SAT Median [95%% CrI]` = sprintf("%.1f [%.1f, %.1f]", SAT_median, SAT_lower, SAT_upper),
    `SAT - UC Mean (SD)` = sprintf("%.1f (%.1f)", diff_mean, diff_sd),
    `SAT - UC Median [95%% CrI]` = sprintf("%.1f [%.1f, %.1f]", diff_median, diff_lower, diff_upper),
    `Posterior Pr(SAT > UC)` = sprintf("%.1f%%", 100 * p_benefit),
    .keep = "none"
  )

rmst_tbl

# save table
write_parquet(rmst_tbl, here("0-main-analysis", "OS", "results", "rmst_summary.parquet"))
```

## Survival Curves

With confidence band (i.e., +- half width of the credible interval of the difference in the survival times for each time point, centered at the middle between the two curves). If it intersects one cruve, will intersect the other as well. If the band and the lines do not intercept > 97.5% probability of benefit for the superior treatment.

```{r}
plot_surv_confband <- ggplot() +
  geom_ribbon(data = conf_band, aes(x = time, ymin = band.low, ymax = band.high, color = NULL), fill = "grey", alpha = 0.2) +
#   geom_line(data = conf_band, aes(x = time, y = middle), color = "black", linetype = "dotted") +
  geom_line(data = surv_curv, aes(y = median, x = time, color = treatment)) +
  scale_color_brewer(palette = "Dark2") +
#   geom_step(data = km_data, aes(x = time, y = survival), linetype = "dashed", linewidth = 0.8) +
  coord_cartesian(ylim = c(0.5, 1), xlim = c(0, 182), expand = FALSE) +
  scale_x_continuous(breaks = seq(0, 182, by = 20)) +
  labs(
    x = "Day",
    y = "Survival Probability",
    color = "",
    fill = ""
  ) +
  theme_light() +
  theme(
      legend.position = "bottom"
    #   plot.title = element_text(face = "bold"),
    #   panel.grid.minor = element_blank()
      )

plot_surv_confband

ggsave(
    here("0-main-analysis", "img", "os_survival_curves_confband.svg"),
    plot = plot_surv_confband,
    width = 8,
    height = 5,
    dpi = 300
)
```

With credible bands around each survival curve, not around the difference + KM curves overlaid

```{r}
plot_surv_credband <- surv_curv |>
  ggplot(aes(x = time, fill = treatment, color = treatment)) +
  geom_ribbon(aes(ymin = conf.low.95, ymax = conf.high.95, color = NULL), alpha = 0.1) +
  geom_line(aes(y = median)) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  geom_step(data = km_data, aes(x = time, y = survival), linetype = "dashed", linewidth = 0.8, alpha = 0.5) +
  geom_point(data = censor_data, aes(x = surv_time, y = survival),
             shape = 3, size = 2, stroke = 1.2, alpha = 0.7) +
  coord_cartesian(ylim = c(0.5, 1), xlim = c(0, 182), expand = FALSE) +
  scale_x_continuous(breaks = seq(0, 182, by = 20), limits = c(0, 182), expand = c(0, 0.05)) +
  labs(
    x = "Day",
    y = "Survival Probability",
    color = "",
    fill = ""
  ) +
  theme_light() +
  theme(
      legend.position = "bottom",
      plot.margin = margin(5, 5, 5, 5, "pt")
      )

# Create risk table plot
risk_table_plot <- risk_table_data |>
  mutate(treatment = factor(treatment, levels = c("UC", "SAT"))) |>
  ggplot(aes(x = time, y = treatment, label = n.risk, color = treatment)) +
  geom_text(size = 3, fontface = "bold") +
  scale_color_brewer(palette = "Dark2", direction = -1) + # need to switch if I want SAT to be on top
  scale_x_continuous(breaks = seq(0, 180, by = 20), limits = c(0, 182), expand = c(0, 0.05)) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL, title = "Number at Risk") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "",
    plot.margin = margin(5, 5, 5, 5, "pt")
  ) +
    guides(color = "none")

combined_plot <- plot_surv_credband / risk_table_plot +
 plot_layout(
    heights = c(6, 1),
    guides = "collect" # <--- This collects the legend from the plot
  ) &
  theme(legend.position = "bottom")

ggsave(
    here("0-main-analysis", "img", "os_survival_curves_credband.svg"),
    plot = combined_plot,
    width = 8,
    height = 6,
    dpi = 300
)
```

## Risk Difference over Time

```{r}
plot_risk_diff <- risk |>
    group_by(time) |>
    summarise(
        mean = mean(RD),
        median = median(RD),
        conf.low = quantile(RD, 0.025),
        conf.high = quantile(RD, 0.975)
    ) |>
    ggplot(aes(x = time, y = median)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
    scale_x_continuous(breaks = seq(0, 182, by = 20)) +
    labs(
    x = "Day",
    y = "Risk Difference",
    color = "",
    fill = ""
    ) +
    theme_light() +
    theme(
        legend.position = "bottom"
        #   plot.title = element_text(face = "bold"),
        #   panel.grid.minor = element_blank()
        )

ggsave(
    here("0-main-analysis", "img", "os_risk_difference.svg"),
    plot = plot_risk_diff,
    width = 8,
    height = 5,
    dpi = 300
)
```

## Risk Ratio over Time

```{r}
plot_risk_ratio <- risk |>
    filter(time != 0) |>
    group_by(time) |>
    summarise(
        mean = mean(RR),
        median = median(RR),
        conf.low = quantile(RR, 0.025),
        conf.high = quantile(RR, 0.975)
    ) |>
    ggplot(aes(x = time, y = median)) +
    geom_line() +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
    scale_x_continuous(breaks = seq(0, 182, by = 20)) +
    labs(
    x = "Day",
    y = "Risk Ratio",
    color = "",
    fill = ""
    ) +
    theme_light() +
    theme(
        legend.position = "bottom"
        #   plot.title = element_text(face = "bold"),
        #   panel.grid.minor = element_blank()
        )

ggsave(
    here("0-main-analysis", "img", "os_risk_ratio.svg"),
    plot = plot_risk_ratio,
    width = 8,
    height = 5,
    dpi = 300
)
```
