---
title: "Best supportive care — unadjusted and adjusted results"
format: html
---

- Cause-specific HR: BSC rate among patients alive and BSC-free; deaths are censored
- Composite HR: first of BSC or death; neither is a subdistribution HR
- Adjusted HRs are conditional, so each adjustment targets a different HR
- Standardized curves and day-182 differences are marginal over the cohort

```{r}
#| label: setup
#| include: false
library(rstanarm)
library(posterior)
library(ggsurvfit)
library(survival)
library(here)
library(arrow)
library(yaml)
library(knitr)
library(tidyverse)

config <- read_yaml(here("config", "analysis.yml"))
source(here("R", "survival.R"))
source(here("R", "plot-style.R"))

results_dir <- here("analysis", "06-best-supportive-care", "results")
model_dir <- file.path(results_dir, "models")
figure_dir <- file.path(results_dir, "figures")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

m <- config$imputation$m_supporting
post_draws <- as.integer(config$posterior_draws$supporting)

prepared <- read_parquet(file.path(results_dir, "prepared.parquet"))
max_follow_up <- max(prepared$time_to_bsc_days)

adjustment_labels <- c(
  unadjusted = "Unadjusted",
  fixed = "Adjusted (fixed)",
  hierarchical = "Adjusted (hierarchical)"
)
endpoint_labels <- c(
  cause_specific = "Cause-specific BSC",
  composite = "BSC or death"
)
```

## Fitted models

```{r}
#| label: load-fits
#| include: false
fit_index <- expand_grid(
  endpoint = names(endpoint_labels),
  adjustment = names(adjustment_labels),
  imputation = 0:m
) |>
  filter((adjustment == "unadjusted") == (imputation == 0L))

fits <- pmap(fit_index, function(endpoint, adjustment, imputation) {
  model_file <- if (adjustment == "unadjusted") {
    paste0("unadjusted-", endpoint, ".rds")
  } else {
    sprintf("%s-%s-%02d.rds", adjustment, endpoint, imputation)
  }

  model <- readRDS(file.path(model_dir, model_file))

  list(
    model = model,
    data = model$data,
    endpoint = endpoint,
    adjustment = adjustment,
    imputation = imputation
  )
})
```

## Sampler diagnostics

```{r}
#| label: diagnostics
#| include: false
diagnostics <- fits |>
  map(function(fit) {
    parameters <- summarise_draws(as_draws_array(fit$model))
    sampler <- rstan::get_sampler_params(fit$model$stanfit, inc_warmup = FALSE)

    tibble(
      endpoint = fit$endpoint,
      adjustment = unname(adjustment_labels[fit$adjustment]),
      imputation = fit$imputation,
      max_rhat = max(parameters$rhat, na.rm = TRUE),
      min_ess_bulk = min(parameters$ess_bulk, na.rm = TRUE),
      min_ess_tail = min(parameters$ess_tail, na.rm = TRUE),
      divergences = sum(map_dbl(sampler, \(x) sum(x[, "divergent__"]))),
      treedepth_hits = sum(map_dbl(sampler, \(x) sum(x[, "treedepth__"] >= 10))),
      min_bfmi = min(map_dbl(
        sampler,
        \(x) mean(diff(x[, "energy__"])^2) / var(x[, "energy__"])
      ))
    )
  }) |>
  list_rbind() |>
  mutate(
    diagnostics_passed = max_rhat <= config$diagnostics$max_rhat &
      min_ess_bulk >= config$diagnostics$min_ess_bulk &
      min_ess_tail >= config$diagnostics$min_ess_tail &
      divergences <= config$diagnostics$max_divergent_transitions &
      treedepth_hits <= config$diagnostics$max_treedepth_exceeded &
      min_bfmi >= config$diagnostics$min_bfmi
  )

stopifnot(
  nrow(diagnostics) == 22L,
  all(diagnostics$diagnostics_passed)
)
```

## Hazard ratios

```{r}
#| label: hazard-ratios
#| include: false
hr_draws <- fits |>
  map(function(fit) {
    log_hr <- as.numeric(as.matrix(fit$model, pars = "tx"))

    if (fit$imputation > 0L) {
      seed <- if (fit$adjustment == "fixed") {
        config$seeds$supporting + (fit$imputation - 1) * 1000 + 2000
      } else {
        3000L + fit$imputation
      }
      set.seed(seed)
      log_hr <- sample(log_hr, post_draws)
    }

    if (fit$adjustment == "hierarchical") {
      stopifnot(
        identical(fit$model$cnms$diagnosis, "(Intercept)"),
        length(grep("^Sigma", colnames(as.matrix(fit$model)))) == 1L
      )
    }

    tibble(
      endpoint = fit$endpoint,
      adjustment = unname(adjustment_labels[fit$adjustment]),
      imputation = fit$imputation,
      draw = seq_along(log_hr),
      log_hazard_ratio = log_hr,
      hazard_ratio = exp(log_hr)
    )
  }) |>
  list_rbind()

hr_summary <- hr_draws |>
  summarise(
    median = median(hazard_ratio),
    lower_95 = quantile(hazard_ratio, 0.025),
    upper_95 = quantile(hazard_ratio, 0.975),
    probability_hr_below_one = mean(hazard_ratio < 1),
    imputations = sum(unique(imputation) > 0),
    posterior_draws = n(),
    .by = c(endpoint, adjustment)
  ) |>
  left_join(
    diagnostics |>
      summarise(
        diagnostics_passed = all(diagnostics_passed),
        .by = c(endpoint, adjustment)
      ),
    by = c("endpoint", "adjustment")
  )

draw_counts <- count(hr_draws, endpoint, adjustment, imputation)

stopifnot(
  nrow(hr_summary) == 6L,
  all(draw_counts$n == if_else(draw_counts$imputation == 0L, 4000L, post_draws))
)
```

## Standardized survival curves

```{r}
#| label: standardized-curves
#| include: false
curves <- fits |>
  map(function(fit) {
    posterior_standardized_survival_draws(
      fit$model,
      fit$data,
      horizon = max_follow_up,
      n_points = 200L,
      n_draws = post_draws,
      seed = 3000L + fit$imputation
    ) |>
      mutate(
        endpoint = fit$endpoint,
        adjustment = unname(adjustment_labels[fit$adjustment]),
        imputation = fit$imputation
      )
  }) |>
  list_rbind()

stopifnot(
  min(curves$time) == 0,
  max(curves$time) == max_follow_up,
  all(is.finite(curves$survival)),
  all(between(curves$survival, 0, 1))
)

curve_summary <- curves |>
  summarise(
    median = median(survival),
    lower_95 = quantile(survival, 0.025),
    upper_95 = quantile(survival, 0.975),
    .by = c(endpoint, adjustment, tx, time)
  )
```

## Prediction checks

```{r}
#| label: prediction-checks
#| include: false
checks <- fits |>
  map(function(fit) {
    newdata <- fit$data |> mutate(tx = 1)

    individual <- posterior_survfit_compatible(
      fit$model,
      newdata = newdata,
      times = 182,
      extrapolate = FALSE,
      standardise = FALSE,
      return_matrix = TRUE,
      draws = 100L,
      seed = 3001L + fit$imputation
    )[[1]]

    standardized <- posterior_survfit_compatible(
      fit$model,
      newdata = newdata,
      times = 182,
      extrapolate = FALSE,
      standardise = TRUE,
      return_matrix = TRUE,
      draws = 100L,
      seed = 3001L + fit$imputation
    )[[1]]

    error <- max(abs(rowMeans(individual) - as.numeric(standardized)))
    diagnosis_range <- NA_real_

    if (fit$adjustment == "hierarchical") {
      contrast_data <- newdata[rep(1L, nlevels(newdata$diagnosis)), ]
      contrast_data$diagnosis <- factor(
        levels(newdata$diagnosis),
        levels = levels(newdata$diagnosis)
      )

      predictions <- posterior_survfit_compatible(
        fit$model,
        newdata = contrast_data,
        times = 182,
        extrapolate = FALSE,
        standardise = FALSE,
        return_matrix = TRUE,
        draws = 100L,
        seed = 3001L + fit$imputation
      )[[1]]

      diagnosis_range <- max(apply(predictions, 1, \(x) diff(range(x))))
      stopifnot(diagnosis_range > 1e-8)
    }

    stopifnot(error < 1e-10)

    tibble(
      endpoint = fit$endpoint,
      adjustment = unname(adjustment_labels[fit$adjustment]),
      imputation = fit$imputation,
      maximum_averaging_error = error,
      maximum_diagnosis_contrast = diagnosis_range
    )
  }) |>
  list_rbind()
```

## Day-182 survival

```{r}
#| label: day-182
#| include: false
landmark_draws <- fits |>
  map(function(fit) {
    predictions <- map(0:1, function(treatment) {
      newdata <- fit$data |> mutate(tx = treatment)

      posterior_survfit_compatible(
        fit$model,
        newdata = newdata,
        times = 182,
        extrapolate = FALSE,
        standardise = TRUE,
        return_matrix = TRUE,
        draws = post_draws,
        seed = 3000L + fit$imputation
      )[[1]] |>
        as.numeric()
    })

    contrast <- predictions[[2]] - predictions[[1]]
    stopifnot(length(contrast) == post_draws, sd(contrast) > 0)

    tibble(
      endpoint = fit$endpoint,
      adjustment = unname(adjustment_labels[fit$adjustment]),
      imputation = fit$imputation,
      draw = seq_len(post_draws),
      usual_care = predictions[[1]],
      ctdna = predictions[[2]],
      contrast = contrast
    )
  }) |>
  list_rbind()

stopifnot(
  nrow(landmark_draws) == 22L * post_draws,
  all(count(landmark_draws, endpoint, adjustment, imputation)$n == post_draws)
)

landmarks <- landmark_draws |>
  summarise(
    usual_care = median(usual_care),
    ctdna = median(ctdna),
    difference = median(contrast),
    lower_95 = quantile(contrast, 0.025),
    upper_95 = quantile(contrast, 0.975),
    probability_difference_positive = mean(contrast > 0),
    .by = c(endpoint, adjustment)
  )
```

```{r}
#| label: save-results
#| include: false
write_parquet(hr_draws, file.path(results_dir, "hazard-ratio-draws.parquet"))
write_parquet(hr_summary, file.path(results_dir, "summary.parquet"))
write_parquet(landmark_draws, file.path(results_dir, "day-182-draws.parquet"))
write_parquet(landmarks, file.path(results_dir, "day-182-summary.parquet"))
write_parquet(checks, file.path(results_dir, "prediction-checks.parquet"))
write_parquet(diagnostics, file.path(results_dir, "diagnostics.parquet"))
write_parquet(curves, file.path(results_dir, "standardized-survival-draws.parquet"))
```

## Results

```{r}
#| label: tbl-bsc-hazard-ratios
#| tbl-cap: "ctDNA-versus-usual-care hazard ratios for BSC and BSC-or-death"
#| echo: false
hr_summary |>
  mutate(
    Endpoint = unname(endpoint_labels[endpoint]),
    Adjustment = adjustment
  ) |>
  select(
    Endpoint, Adjustment, median, lower_95, upper_95, probability_hr_below_one
  ) |>
  kable(digits = 3)
```

```{r}
#| label: fig-bsc-hazard-ratios
#| fig-cap: "Unadjusted, Adjusted (fixed), and Adjusted (hierarchical) BSC hazard ratios with 95% credible intervals"
#| fig-alt: "Treatment hazard ratios for cause-specific BSC and BSC or death, comparing unadjusted, fixed-covariate, and diagnosis-random-intercept models."
#| fig-width: 8
#| fig-height: 4
#| echo: false
hr_plot <- hr_summary |>
  mutate(
    endpoint = factor(
      unname(endpoint_labels[endpoint]),
      levels = unname(endpoint_labels)
    ),
    adjustment = factor(adjustment, levels = unname(adjustment_labels))
  ) |>
  ggplot(aes(x = median, y = adjustment, xmin = lower_95, xmax = upper_95)) +
  geom_pointrange(orientation = "y") +
  geom_vline(xintercept = 1, linetype = "dashed", colour = reference_line_colour) +
  facet_wrap(~ endpoint) +
  scale_x_log10() +
  labs(x = "Hazard ratio (ctDNA / usual care)", y = NULL)

save_figure(hr_plot, file.path(figure_dir, "hazard-ratios.png"), height = 4)

hr_plot
```

- Cause-specific curves and day-182 values are net survival (deaths censored), not cumulative incidence
- Composite curves and day-182 values are the probability of staying alive and BSC-free

```{r}
#| label: tbl-bsc-day-182
#| tbl-cap: "Day-182 standardized survival and ctDNA-minus-usual-care difference"
#| echo: false
landmarks |>
  mutate(
    Outcome = unname(endpoint_labels[endpoint]),
    Adjustment = adjustment
  ) |>
  select(
    Outcome, Adjustment, usual_care, ctdna, difference, lower_95,
    upper_95, probability_difference_positive
  ) |>
  kable(digits = 3)
```

```{r}
#| label: fig-bsc-standardized-survival
#| fig-cap: "Standardized BSC survival by endpoint, adjustment, and randomized group"
#| fig-alt: "Standardized cause-specific BSC net survival and BSC-or-death survival curves for usual-care and ctDNA under unadjusted, fixed-covariate, and diagnosis-random-intercept adjustment."
#| fig-width: 8
#| fig-height: 6
#| echo: false
plot_data <- curve_summary |>
  mutate(
    endpoint = factor(
      unname(endpoint_labels[endpoint]),
      levels = unname(endpoint_labels)
    ),
    adjustment = factor(adjustment, levels = unname(adjustment_labels)),
    tx = group_factor(tx)
  )

survival_plot <- ggplot(plot_data, aes(time, median, colour = tx, fill = tx)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), alpha = 0.15, colour = NA) +
  geom_line() +
  facet_grid(endpoint ~ adjustment) +
  scale_x_continuous(breaks = seq(0, max_follow_up, by = 200)) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_colour_group() +
  scale_fill_group() +
  labs(
    x = "Days since random selection",
    y = "Standardized survival probability",
    colour = NULL,
    fill = NULL
  )

save_figure(
  survival_plot,
  file.path(figure_dir, "standardized-survival.png"),
  height = 6
)

survival_plot
```

```{r}
#| label: fig-bsc-composite-km
#| fig-cap: "Unadjusted Kaplan-Meier probability of remaining alive and BSC-free"
#| fig-alt: "Kaplan-Meier curves for the randomized groups from randomization until BSC or death, with censor marks and numbers at risk."
#| fig-width: 8
#| fig-height: 6
#| echo: false
observed <- prepared |>
  mutate(tx_label = group_factor(tx))

composite_fit <- survfit2(
  Surv(time_to_bsc_days, event_bsc_or_death) ~ tx_label,
  data = observed
)

# theme = NULL: the report theme instead of ggsurvfit's own.
composite_plot <- ggsurvfit(composite_fit, theme = NULL) +
  add_confidence_interval() +
  add_censor_mark() +
  scale_x_continuous(breaks = seq(0, max_follow_up, by = 100)) +
  scale_colour_group() +
  scale_fill_group() +
  add_risktable(
    times = seq(0, max_follow_up, by = 100),
    risktable_stats = c("n.risk", "cum.event"),
    stats_label = list(n.risk = "At risk", cum.event = "Events")
  ) +
  labs(x = "Days since random selection", y = "Alive and BSC-free probability")

save_figure(
  ggsurvfit_build(composite_plot),
  file.path(figure_dir, "composite-kaplan-meier.png"),
  height = 6
)

composite_plot
```

```{r}
#| label: fig-bsc-cumulative-hazard
#| fig-cap: "Unadjusted cumulative cause-specific hazard of BSC, censoring death before BSC"
#| fig-alt: "Nelson–Aalen cumulative BSC hazard curves by randomized group; death before BSC is treated as censoring."
#| fig-width: 8
#| fig-height: 5
#| echo: false
cause_fit <- survfit(Surv(time_to_bsc_days, event_bsc) ~ tx_label, data = observed)

hazard_data <- tibble(
  time = cause_fit$time,
  cumulative_hazard = cause_fit$cumhaz,
  group = str_remove(
    rep(names(cause_fit$strata), cause_fit$strata),
    fixed("tx_label=")
  )
) |>
  bind_rows(tibble(
    time = 0,
    cumulative_hazard = 0,
    group = levels(observed$tx_label)
  )) |>
  arrange(group, time)

hazard_plot <- hazard_data |>
  mutate(group = factor(group, levels = levels(observed$tx_label))) |>
  ggplot(aes(time, cumulative_hazard, colour = group)) +
  geom_step() +
  scale_colour_group() +
  labs(
    x = "Days since random selection",
    y = "Cumulative cause-specific BSC hazard",
    colour = NULL
  )

save_figure(
  hazard_plot,
  file.path(figure_dir, "cause-specific-hazard.png"),
  height = 5
)

hazard_plot
```

## Diagnostics and checks

```{r}
#| label: tbl-bsc-diagnostics
#| tbl-cap: "Posterior diagnostics for all BSC models"
#| echo: false
kable(diagnostics, digits = 3)
```

```{r}
#| label: tbl-bsc-prediction-checks
#| tbl-cap: "Prediction checks at day 182 for all BSC models (NA: diagnosis contrast not applicable)"
#| echo: false
kable(checks, digits = 10)
```
